---
title: "Semi-manuel Annotation Subset 1"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
# Goal
The aim of this script is subcluster questionable annotation of T and NK cells and to adapt Hao and Hao et. al annotation where required.

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(readxl)
library(ArchR)
library(tidyverse)
library(RColorBrewer)
library(ggrepel)
library(SeuratDisk)
library(Seurat)
library(glue)
```

```{r Run_parameters}
set.seed(1)
addArchRThreads(threads = 1)
```

# Set paths
```{r define_directories }
source("config_paths.R")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "ArchR_celltype_annotation")
RDS_dir <- file.path(analysis_dir, "RDS_Objects")
ArchR_dir <- file.path("ArchR_Subset4")

ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")
ArchR_result_dir <- file.path(ArchR_dir_Annotation, ArchR_dir)
plotting_dir <- file.path(ArchR_dir_Annotation, "Plots")
subset_result_dir <- file.path(ArchR_dir_Annotation, "ArchR_Healthy_Subset4")
subset_nb_result_dir <- file.path(ArchR_dir_Annotation, "ArchR_Healthy_NonB_Subset4")
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r Load_ArchR_Project }
setwd(file.path(ArchR_result_dir))
CLLProj <- loadArchRProject(path = ArchR_result_dir)

if (!file.exists("scRNA-PBMcs_Hao_and_Hao_et_2021.h5seurat")) {
  dir.create(file.path(data_dir, "reference_genomes"))
  download.file(
    url = "https://atlas.fredhutch.org/data/nygc/multimodal/pbmc_multimodal.h5seurat",
    destfile = file.path(data_dir, "reference_genomes/Hao_and_Hao_et_al_2021_PBMC/pbmc_multimodal.h5seurat")
  )
}

Hao_seRNA <- LoadH5Seurat(file.path(data_dir, "reference_genomes/Hao_and_Hao_et_al_2021_PBMC/pbmc_multimodal.h5seurat"))
Hao_seRNA
```

# Analysis
## Hao and Hao et al Subcluster annotation
```{r Create_RNA_reference_assay, fig.width=2, fig.height=2}
# As this dataset does not have an RNA assay, the non-SCT values need to be reconstructed to an RNA assay for the reference
rna_counts <- CreateAssayObject(Hao_seRNA@assays[["SCT"]]@counts)

NormalizeData(
  rna_counts,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  margin = 1,
  verbose = TRUE
)

Hao_seRNA@assays[["RNA"]] <- rna_counts
```

## Subclustering Healthy cells
```{r Subset_ArchR_project_healthy_cells, fig.width=10, fig.height=10}
plotEmbedding(
  plotAs = "points",
  size = 0.5,
  embedding = "UMAP_LSI_0.4",
  CLLProj,
  colorBy = "cellColData",
  name = "Clusters_0.3_IterativeLSI_0.4",
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

plotEmbedding(
  plotAs = "points",
  size = 0.5,
  embedding = "UMAP_LSI_0.4",
  CLLProj,
  colorBy = "cellColData",
  name = "Sample",
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )


subset_healthy_pass <- which(CLLProj@cellColData$Clusters_0.3_IterativeLSI_0.4 %in% c("C9", "C13", "C14", "C15", "C16"))
subset_healthy_cells_pass <- CLLProj$cellNames[subset_healthy_pass]

CLLProj_Subset <- subsetArchRProject(
  ArchRProj = CLLProj,
  cells = subset_healthy_cells_pass,
  outputDirectory = subset_result_dir,
  dropCells = TRUE
)

CLLProj_Subset <- loadArchRProject(path = subset_result_dir)
```

### Iterative LSI
```{r LSI_0.2_and_0.4}
CLLProj_Subset <- addIterativeLSI(
  ArchRProj = CLLProj_Subset,
  useMatrix = "TileMatrix",
  name = "Subset_IterativeLSI",
  iterations = 2,
  clusterParams = list(
    resolution = c(0.2),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30,
  threads = 1
)

CLLProj_Subset <- addIterativeLSI(
  ArchRProj = CLLProj_Subset,
  useMatrix = "TileMatrix",
  name = "Subset_IterativeLSI_0.4",
  iterations = 2,
  clusterParams = list(
    resolution = c(0.4),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30,
  threads = 1
)

saveArchRProject(ArchRProj = CLLProj_Subset, outputDirectory = subset_result_dir, load = FALSE, overwrite = TRUE)
```

### Add Clusters of different resolution
```{r Clustering_0.2_to_0.6}
# Try combinations between different LSI resolutions and clustering resolutions
param_combinations_subset <- tidyr::expand_grid(
  dimred = c("Subset_IterativeLSI", "Subset_IterativeLSI_0.4"),
  resolution = c(0.2, 0.3, 0.4, 0.6)
)

for (row in 1:nrow(param_combinations_subset)) {
  resolution <- param_combinations_subset$resolution[[row]]
  dimred <- param_combinations_subset$dimred[[row]]

  CLLProj_Subset <- addClusters(
    input = CLLProj_Subset,
    reducedDims = dimred,
    method = "Seurat",
    name = glue("Subset_Clusters_{resolution}_{dimred}"),
    resolution = resolution, force = TRUE
  )
}

saveArchRProject(ArchRProj = CLLProj_Subset, outputDirectory = subset_result_dir, load = FALSE, overwrite = TRUE)
```

### Add UMAP embedding
```{r UMAP}
CLLProj_Subset <- addUMAP(
  ArchRProj = CLLProj_Subset,
  reducedDims = "Subset_IterativeLSI",
  name = "Subset_UMAP_Subset_IterativeLSI",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine"
)

CLLProj_Subset <- addUMAP(
  ArchRProj = CLLProj_Subset,
  reducedDims = "Subset_IterativeLSI_0.4",
  name = "Subset_UMAP_Subset_IterativeLSI_0.4",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine"
)

saveArchRProject(ArchRProj = CLLProj_Subset, outputDirectory = subset_result_dir, load = FALSE, overwrite = TRUE)
```

### Plot embedding
```{r}
sample_IterativeLSI <- plotEmbedding(
  plotAs = "points",
  size = 0.5,
  embedding = "Subset_UMAP_Subset_IterativeLSI",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "Sample",
  pal = c(my_paired_colours_named, my_unpaired_colours_named)
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

sample_IterativeLSI_0.4 <- plotEmbedding(
  plotAs = "points",
  size = 0.5,
  embedding = "Subset_UMAP_Subset_IterativeLSI_0.4",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "Sample",
  pal = c(my_paired_colours_named, my_unpaired_colours_named)
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
```

```{r Plot_clusterings, fig.width=30, fig.height=12}
all_cluster_plots <- list()

for (row in 1:nrow(param_combinations_subset)) {
  resolution <- param_combinations_subset$resolution[[row]]
  dimred <- param_combinations_subset$dimred[[row]]
  print(resolution)
  print(dimred)

  cluster_plots <- plotEmbedding(
    ArchRProj = CLLProj_Subset,
    colorBy = "cellColData",
    name = glue("Subset_Clusters_{resolution}_{dimred}"),
    embedding = glue("Subset_UMAP_{dimred}"),
    plotAs = "points",
    size = 3
  ) +
    labs(x = "UMAP1", y = "UMAP2") +
    theme_pub() +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal"
    )
  # print(cluster_plots)
  all_cluster_plots[[row]] <- cluster_plots
}

cowplot::plot_grid(ggarrange(sample_IterativeLSI, sample_IterativeLSI_0.4, nrow = 2),
  ggarrange(plotlist = all_cluster_plots, nrow = 2, ncol = 4),
  ncol = 2, rel_widths = c(1, 4)
)

cowplot::plot_grid(ggarrange(sample_IterativeLSI, sample_IterativeLSI_0.4, nrow = 2),
  ggarrange(plotlist = all_cluster_plots, nrow = 2, ncol = 4),
  ncol = 2, rel_widths = c(1, 4)
) %>%
  ggsave(
    plot = .,
    filename = paste0(plotting_dir, "Plot-UMAP-Healthy-Cells-Subclustering-Res-0.2-0.6.png"),
    width = 90,
    height = 30,
    unit = "cm"
  )

#-->resolution of 0.2 for clustering using a LSI of 0.2 was sufficient
```

### Bridge Integration at different LSI resolutions
```{r Bridge_Integration_Subset}
CLLProj_Subset <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj_Subset,
  useMatrix = "GeneScoreMatrix",
  matrixName = "Subset_GeneIntegrationMatrix_IterativeLSI",
  reducedDims = "Subset_IterativeLSI",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "Subset_predictedCell_Un_Hao_IterativeLSI_,",
  nameGroup = "Subset_predictedGroup_Un_Hao_IterativeLSI",
  nameScore = "Subset_predictedScore_Un_Hao_IterativeLSI",
  threads = 1
)

CLLProj_Subset <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj_Subset,
  useMatrix = "GeneScoreMatrix",
  matrixName = "Subset_GeneIntegrationMatrix_IterativeLSI_0.4",
  reducedDims = "Subset_IterativeLSI_0.4",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "Subset_predictedCell_Un_Hao_IterativeLSI_0.4,",
  nameGroup = "Subset_predictedGroup_Un_Hao_IterativeLSI_0.4",
  nameScore = "Subset_predictedScore_Un_Hao_IterativeLSI_0.4",
  threads = 1
)

saveArchRProject(ArchRProj = CLLProj_Subset, outputDirectory = subset_result_dir, load = FALSE, overwrite = TRUE)
```

### Plot annotations
```{r Plot_annotion_LSI_0.2, fig.height=8, fig.width=20}
annotation_IterativeLSI_subset <- plotEmbedding(
  plotAs = "points",
  size = 2,
  embedding = "Subset_UMAP_Subset_IterativeLSI",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "Subset_predictedGroup_Un_Hao_IterativeLSI",
  pal = annotation_colours_named
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggAlignPlots(sample_IterativeLSI, all_cluster_plots[[1]], annotation_IterativeLSI_subset, type = "h")
```

```{r Plot_annotation_LSI_0.4, fig.height=8, fig.width=20}
annotation_IterativeLSI_0.4_subset <- plotEmbedding(
  plotAs = "points",
  size = 2,
  embedding = "Subset_UMAP_Subset_IterativeLSI_0.4",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "Subset_predictedGroup_Un_Hao_IterativeLSI_0.4",
  pal = annotation_colours_named
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggAlignPlots(sample_IterativeLSI_0.4, all_cluster_plots[[5]], annotation_IterativeLSI_0.4_subset, type = "h")
```

```{r, Compare_LSI_resolutions_fig.width=70, fig.height=15}
cowplot::plot_grid(ggarrange(sample_IterativeLSI, all_cluster_plots[[1]], annotation_IterativeLSI_subset, ncol = 3),
  ggarrange(sample_IterativeLSI_0.4, all_cluster_plots[[5]], annotation_IterativeLSI_0.4_subset, ncol = 3),
  nrow = 2
)

cowplot::plot_grid(ggarrange(sample_IterativeLSI, all_cluster_plots[[1]], annotation_IterativeLSI_subset, ncol = 3),
  ggarrange(sample_IterativeLSI_0.4, all_cluster_plots[[5]], annotation_IterativeLSI_0.4_subset, ncol = 3),
  nrow = 2
) %>%
  ggsave(
    plot = ., filename = paste0(plotting_dir, "Plot-UMAP-Healthy-Cells-Subclustering-Annotation.pdf"),
    width = 50,
    height = 30,
    unit = "cm"
  )
```


## Subclustering non-B cell healthy immune cells
```{r Subset_ArchR_proj_to_non_healthy_Bcells}
subset_healthy_nb_pass <- which(CLLProj@cellColData$Clusters_0.3_IterativeLSI_0.4 %in% c("C13", "C14", "C15", "C16"))
subset_healthy_nb_cells_pass <- CLLProj$cellNames[subset_healthy_nb_pass]

CLLProj_Subset_nb <- subsetArchRProject(
  ArchRProj = CLLProj,
  cells = subset_healthy_nb_cells_pass,
  outputDirectory = subset_nb_result_dir,
  dropCells = TRUE
)

# CLLProj_Subset_nb <- loadArchRProject(path = subset_nb_result_dir)
```

### Iterative LSI
```{r Subset_nb_LSIs}
CLLProj_Subset_nb <- addIterativeLSI(
  ArchRProj = CLLProj_Subset_nb,
  useMatrix = "TileMatrix",
  name = "Subset_nB_IterativeLSI",
  iterations = 2,
  clusterParams = list(
    resolution = c(0.2),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30,
  threads = 1
)

CLLProj_Subset_nb <- addIterativeLSI(
  ArchRProj = CLLProj_Subset_nb,
  useMatrix = "TileMatrix",
  name = "Subset_nB_IterativeLSI_0.4",
  iterations = 2,
  clusterParams = list(
    resolution = c(0.4),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30,
  threads = 1
)

saveArchRProject(ArchRProj = CLLProj_Subset_nb, outputDirectory = subset_nb_result_dir, load = FALSE, overwrite = TRUE)
```

### Add Clusters of different resolution
```{r Clustering_subset_nb}
# Try combinations between different LSI resolutions and clustering resolutions
param_combinations_subset_nb <- tidyr::expand_grid(
  dimred = c("Subset_nB_IterativeLSI", "Subset_nB_IterativeLSI_0.4"),
  resolution = c(0.2, 0.3, 0.4, 0.6)
)

for (row in 1:nrow(param_combinations_subset_nb)) {
  resolution <- param_combinations_subset_nb$resolution[[row]]
  dimred <- param_combinations_subset_nb$dimred[[row]]

  CLLProj_Subset_nb <- addClusters(
    input = CLLProj_Subset_nb,
    reducedDims = dimred,
    method = "Seurat",
    name = glue("Subset_nB_Clusters_{resolution}_{dimred}"),
    resolution = resolution
  )
}

saveArchRProject(ArchRProj = CLLProj_Subset_nb, outputDirectory = subset_nb_result_dir, load = FALSE, overwrite = TRUE)
```

### Add UMAP embedding
```{r UMAP_subset_nb}
CLLProj_Subset_nb <- addUMAP(
  ArchRProj = CLLProj_Subset_nb,
  reducedDims = "Subset_nB_IterativeLSI",
  name = "Subset_nB_UMAP_Subset_nB_IterativeLSI",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine"
)

CLLProj_Subset_nb <- addUMAP(
  ArchRProj = CLLProj_Subset_nb,
  reducedDims = "Subset_nB_IterativeLSI_0.4",
  name = "Subset_nB_UMAP_Subset_nB_IterativeLSI_0.4",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine"
)

saveArchRProject(ArchRProj = CLLProj_Subset_nb, outputDirectory = subset_nb_result_dir, load = FALSE, overwrite = TRUE)
```

### Plot embedding

```{r Plot_subset_nb_sample}
subset_nb_sample_IterativeLSI <- plotEmbedding(
  plotAs = "points",
  embedding = "Subset_nB_UMAP_Subset_nB_IterativeLSI",
  CLLProj_Subset_nb,
  colorBy = "cellColData",
  name = "Sample",
  pal = c(my_paired_colours_named, my_unpaired_colours_named)
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

subset_nb_sample_IterativeLSI_0.4 <- plotEmbedding(
  plotAs = "points",
  embedding = "Subset_nB_UMAP_Subset_nB_IterativeLSI_0.4",
  CLLProj_Subset_nb,
  colorBy = "cellColData",
  name = "Sample",
  pal = c(my_paired_colours_named, my_unpaired_colours_named)
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
```

```{r Plot_Clustering_subset_nb, fig.width=30, fig.height=12}
all_subcluster_plots <- list()

for (row in 1:nrow(param_combinations_subset_nb)) {
  resolution <- param_combinations_subset_nb$resolution[[row]]
  dimred <- param_combinations_subset_nb$dimred[[row]]
  print(resolution)
  print(dimred)

  subcluster_plots <- plotEmbedding(
    ArchRProj = CLLProj_Subset_nb,
    colorBy = "cellColData",
    name = glue("Subset_nB_Clusters_{resolution}_{dimred}"),
    embedding = glue("Subset_nB_UMAP_{dimred}"),
    plotAs = "points"
  ) +
    geom_point(size = 2) +
    labs(x = "UMAP1", y = "UMAP2") +
    theme_pub() +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal"
    )

  all_subcluster_plots[[row]] <- subcluster_plots
}

cowplot::plot_grid(ggarrange(subset_nb_sample_IterativeLSI, subset_nb_sample_IterativeLSI_0.4, nrow = 2),
  ggarrange(plotlist = all_subcluster_plots, nrow = 2, ncol = 4),
  ncol = 2, rel_widths = c(1, 4)
)

cowplot::plot_grid(ggarrange(subset_nb_sample_IterativeLSI, subset_nb_sample_IterativeLSI_0.4, nrow = 2),
  ggarrange(plotlist = all_subcluster_plots, nrow = 2, ncol = 4),
  ncol = 2, rel_widths = c(1, 4)
) %>%
  ggsave(
    plot = ., filename = file.path(ArchR_result_dir, "Plots/Plot-UMAP-Healthy-NonB-Cells-Subclustering-Res-0.2-0.6.pdf"),
    width = 90,
    height = 30,
    unit = "cm"
  )
```

### Bridge Integration at different LSI resolutions
```{r Bridge_integration_subset_nb}
CLLProj_Subset_nb <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj_Subset_nb,
  useMatrix = "GeneScoreMatrix",
  matrixName = "Subset_nB_GeneIntegrationMatrix_IterativeLSI",
  reducedDims = "Subset_nB_IterativeLSI",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "Subset_nB_predictedCell_Un_Hao_IterativeLSI",
  nameGroup = "Subset_nB_predictedGroup_Un_Hao_IterativeLSI",
  nameScore = "Subset_nB_predictedScore_Un_Hao_IterativeLSI",
  threads = 1
)

CLLProj_Subset_nb <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj_Subset_nb,
  useMatrix = "GeneScoreMatrix",
  matrixName = "Subset_nB_GeneIntegrationMatrix_IterativeLSI_0.4",
  reducedDims = "Subset_nB_IterativeLSI_0.4",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "Subset_nB_predictedCell_Un_Hao_IterativeLSI_0.4,",
  nameGroup = "Subset_nB_predictedGroup_Un_Hao_IterativeLSI_0.4",
  nameScore = "Subset_nB_predictedScore_Un_Hao_IterativeLSI_0.4",
  threads = 1
)

saveArchRProject(ArchRProj = CLLProj_Subset_nb, outputDirectory = subset_nb_result_dir, load = FALSE, overwrite = TRUE)
```

### Plot annotations
```{r Plot_annotation_subset_nb_LSI, fig.height=8, fig.width=20}
annotation_IterativeLSI_subset_nb <- plotEmbedding(
  plotAs = "points",
  embedding = "Subset_nB_UMAP_Subset_nB_IterativeLSI",
  CLLProj_Subset_nb,
  colorBy = "cellColData",
  name = "Subset_nB_predictedGroup_Un_Hao_IterativeLSI",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggAlignPlots(subset_nb_sample_IterativeLSI, all_subcluster_plots[[1]], annotation_IterativeLSI_subset_nb, type = "h")
```

```{r Plot_annotation_subset_nb_LSI_0.4, fig.height=8, fig.width=20}
annotation_IterativeLSI_0.4_subset_nb <- plotEmbedding(
  plotAs = "points",
  embedding = "Subset_nB_UMAP_Subset_nB_IterativeLSI_0.4",
  CLLProj_Subset_nb,
  colorBy = "cellColData",
  name = "Subset_nB_predictedGroup_Un_Hao_IterativeLSI_0.4",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggAlignPlots(subset_nb_sample_IterativeLSI_0.4, all_subcluster_plots[[5]], annotation_IterativeLSI_0.4_subset_nb, type = "h")
```

```{r, fig.width=30, fig.height=10}
cowplot::plot_grid(ggarrange(subset_nb_sample_IterativeLSI, all_subcluster_plots[[1]], annotation_IterativeLSI_subset_nb, ncol = 3),
  ggarrange(subset_nb_sample_IterativeLSI_0.4, all_subcluster_plots[[5]], annotation_IterativeLSI_0.4_subset_nb, ncol = 3),
  nrow = 2
)

cowplot::plot_grid(ggarrange(subset_nb_sample_IterativeLSI, all_subcluster_plots[[1]], annotation_IterativeLSI_subset_nb, ncol = 3),
  ggarrange(subset_nb_sample_IterativeLSI_0.4, all_subcluster_plots[[5]], annotation_IterativeLSI_0.4_subset_nb, ncol = 3),
  nrow = 2
) %>%
  ggsave(
    plot = ., filename = file.path(ArchR_result_dir, "Plots/Plot-UMAP-Healthy-NonB-Cells-Subclustering-Annotation.pdf"),
    width = 50,
    height = 30,
    unit = "cm"
  )
```

```{r Plot_Subset_Annotation_in_initial_UMAP_context, fig.width=20, fig.height=10}
annotation_full_set <- plotEmbedding(
  plotAs = "points",
  embedding = "UMAP_LSI_0.4",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "predictedGroup_Un_Hao",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

annotation_full_set_LSI_0.4 <- plotEmbedding(
  plotAs = "points",
  embedding = "UMAP_LSI_0.4",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "predictedGroup_Un_Hao_IterativeLSI_0.4",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

annotation_subset <- plotEmbedding(
  plotAs = "points",
  embedding = "UMAP_LSI_0.4",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "Subset_predictedGroup_Un_Hao_IterativeLSI",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

annotation_subset_LSI_0.4 <- plotEmbedding(
  plotAs = "points",
  embedding = "UMAP_LSI_0.4",
  CLLProj_Subset,
  colorBy = "cellColData",
  name = "Subset_predictedGroup_Un_Hao_IterativeLSI_0.4",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

annotation_subset_nb <- plotEmbedding(
  plotAs = "points",
  embedding = "UMAP_LSI_0.4",
  CLLProj_Subset_nb,
  colorBy = "cellColData",
  name = "Subset_nB_predictedGroup_Un_Hao_IterativeLSI",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

annotation_subset_nb_LSI_0.4 <- plotEmbedding(
  plotAs = "points",
  embedding = "UMAP_LSI_0.4",
  CLLProj_Subset_nb,
  colorBy = "cellColData",
  name = "Subset_nB_predictedGroup_Un_Hao_IterativeLSI_0.4",
  pal = annotation_colours_named
) +
  geom_point(size = 2) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggarrange(annotation_full_set, annotation_subset, annotation_subset_nb,
  annotation_full_set_LSI_0.4, annotation_subset_LSI_0.4, annotation_subset_nb_LSI_0.4,
  nrow = 2, ncol = 3
)

# The initial annotation does not have B cells in these cluster, so I keep it
CLLProj_Subset_nb@cellColData$predictedGroup_Un_Hao %>% table()
CLLProj_Subset_nb@cellColData$Subset_predictedGroup_Un_Hao_IterativeLSI %>% table()
CLLProj_Subset_nb@cellColData$Subset_nB_predictedGroup_Un_Hao_IterativeLSI %>% table()
```

```{r}
ggarrange(annotation_full_set, annotation_subset, annotation_subset_nb,
  annotation_full_set_LSI_0.4, annotation_subset_LSI_0.4, annotation_subset_nb_LSI_0.4,
  nrow = 2,
  ncol = 3
) %>%
  ggsave(
    plot = ., filename = file.path(ArchR_result_dir, "Plots/Plot-UMAP-Healthy-NonB-Cells-Subclustering-Annotation_Projection.pdf"),
    width = 50,
    height = 30,
    unit = "cm"
  )
```

## Annotate
```{r Distinguish_benign_malignant_cells, fig.width=8}
# Check where non-CLL/MBL cells are distributed
MBL_CLL_index <- which(CLLProj$Clusters_0.3_IterativeLSI_0.4 %in% c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C10", "C11", "C12"))
MBL_CLL <- CLLProj$cellNames[MBL_CLL_index]

CLLProj@cellColData$MBL_CLL_cells <- CLLProj$cellNames %in% MBL_CLL

plotEmbedding(
  plotAs = "point",
  size = 1,
  embedding = "UMAP_LSI_0.4",
  CLLProj,
  colorBy = "cellColData",
  name = "MBL_CLL_cells"
)

plotEmbedding(
  plotAs = "point",
  size = 1,
  embedding = "UMAP_LSI_0.4",
  CLLProj,
  colorBy = "cellColData",
  name = "Clusters_0.3_IterativeLSI_0.4"
)

plotEmbedding(
  plotAs = "point",
  size = 1,
  embedding = "UMAP_LSI_0.4",
  CLLProj,
  colorBy = "cellColData",
  name = "Sample"
)
```


```{r Define_annotation}
healthy_clusters <- c("C9", "C13", "C14", "C15", "C16")

My_Annotation <- tibble(
  label_cellbarcode = CLLProj$cellNames,
  `Clusters_0.3_IterativeLSI_0.4` = CLLProj$`Clusters_0.3_IterativeLSI_0.4`
) %>%
  dplyr::left_join(tibble(
    label_cellbarcode = CLLProj_Subset$cellNames,
    Subset_predictedGroup_Un_Hao_IterativeLSI_0.4 = CLLProj_Subset$Subset_predictedGroup_Un_Hao_IterativeLSI_0.4
  )) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  mutate(My_Annotation = case_when(str_ends(label, "\\.1") & !(Clusters_0.3_IterativeLSI_0.4 %in% healthy_clusters) ~ "MBL_B_Cells",
    str_starts(label, paste0(c("MBL4"), collapse = "|")) & !(Clusters_0.3_IterativeLSI_0.4 %in% healthy_clusters) ~ "CLL_B_Cells",
    str_starts(label, paste0(paste0("MBL", c(10:15, "7dup")), collapse = "|")) & !(Clusters_0.3_IterativeLSI_0.4 %in% healthy_clusters) ~ "MBL_B_Cells",
    str_ends(label, "\\.2") & !(Clusters_0.3_IterativeLSI_0.4 %in% healthy_clusters) ~ "CLL_B_Cells",
    .default = Subset_predictedGroup_Un_Hao_IterativeLSI_0.4
  ))
```

```{r Add_annotation_column}
# Assign annotations to new column
CLLProj@cellColData$My_Annotation <- My_Annotation$My_Annotation

# Summarize CD4 and CD8-Tcells to only T-cells
CLLProj@cellColData$My_Annotation <- case_when(CLLProj@cellColData$My_Annotation %in% c("CD4 T", "CD8 T", "other T") ~ "T",
  .default = CLLProj@cellColData$My_Annotation
)

saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, overwrite = TRUE, load = FALSE)

final_annotation <- plotEmbedding(
  plotAs = "point",
  size = 1,
  embedding = "UMAP_LSI_0.4",
  CLLProj,
  colorBy = "cellColData",
  name = "My_Annotation",
  pal = annotation_colours_named
) +
  geom_point(size = 0.5) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

final_annotation
```

```{r Export_full_annotation}
plotPDF(final_annotation,
  name = "Plot-UMAP-RNA-Hao-Manual_Integration.png",
  ArchRProj = CLLProj,
  addDOC = FALSE,
  width = 5,
  height = 5
)
```
