---
title: "Investigate MBL/CLL data"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

# Load libraries

```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(ArchR)
library(readxl)
library(tidyverse)
library(glue)
library(ggpubr)
```

```{r Run_parameters}
set.seed(1)
addArchRThreads(threads = 1)
addArchRGenome("hg38")
```

# Set paths
```{r }
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "ATAC")
result_dir <- file.path(project_dir, "results")

# subset
# ArchR_dir <- file.path("ArchR_Subset1")
# ArchR_dir <- file.path("ArchR_Subset2")
# ArchR_dir <- file.path("ArchR_Subset3")
ArchR_dir <- file.path("ArchR_Subset4")

ArchR_result_dir <- file.path(result_dir, ArchR_dir)
RDS_dir <- file.path(ArchR_result_dir, "RDS_objects")
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r }
library_metadata <- read_xlsx(paste0(data_dir, "20240324_MBL_CLL_library_metadata.xlsx"), .name_repair = "unique_quiet")
```

```{r Load_all_samples}
setwd(result_dir)

# Get all libraries files and paths
# Subset1
# sample_subset <- c(
#   "CLL1.1",
#   "CLL1.2",
#   "CLL2.1",
#   "CLL2.2",
#   "CLL3.1",
#   "CLL3.2",
#   "MBL1",
#   "MBL2",
#   "MBL3"
# )
# Subset2
# sample_subset <- c(
#   "CLL4.1",
#   "CLL4.2",
#   "CLL5.1",
#   "CLL5.2",
#   "CLL6.1",
#   "CLL6.2",
#   "MBL4",
#   "MBL6",
#   "MBL7"
# )
# Subset3
# sample_subset <- c("CLL7.1",
#                    "CLL7.2",
#                    "CLL8.1",
#                    "CLL8.2",
#                    "CLL9.1",
#                    "CLL9.2",
#                    "CLL10.1",
#                    "CLL10.2",
#                    "MBL9",
#                    "MBL8")

# Subset4
sample_subset <- c(
  "MBL7dup",
  "MBL10",
  "MBL11",
  "MBL12",
  "MBL13",
  "MBL14",
  "MBL15"
)

# Get all libraries and samples names
sample_names <- library_metadata %>%
  dplyr::filter(`Data type` == "ATAC") %>%
  dplyr::select(Label, Pool) %>%
  dplyr::filter(Label %in% sample_subset) %>%
  dplyr::arrange(Pool) %>%
  dplyr::pull(var = Label, name = Pool)

# Select library_file_paths based on sample_names
library_file_paths <- purrr::map(names(sample_names), ~ paste0(library_path, .x, "/fragments.tsv.gz")) %>%
  unlist()

# Create ArrowFiles for all samples
# ArrowFiles <- createArrowFiles(
#   inputFiles = library_file_paths,
#   sampleNames = sample_names,
#   minTSS = 4,
#   minFrags = 1000,
#   addTileMat = TRUE,
#   addGeneScoreMat = TRUE
# )

ArrowFiles <- file.path(result_dir, paste0(sample_names, ".arrow"))
```
# Analysis
## Doublet Inference
```{r Doublet_Inference}
setwd(result_dir)

doubScores <- addDoubletScores(
  input = ArrowFiles,
  k = 10, # Refers to how many cells near a "pseudo-doublet" to count.
  knnMethod = "UMAP", # Refers to the embedding to use for nearest neighbor search with doublet projection.
  LSIMethod = 1
)
dir.create(file.path(RDS_dir))
saveRDS(doubScores, file = file.path(RDS_dir, "doubScores.RDS"))
```

## Create ArchR Project
```{r Make_ArchRProject}
setwd(result_dir)

CLLProj <- ArchRProject(
  ArrowFiles = ArrowFiles,
  outputDirectory = ArchR_dir,
  copyArrows = TRUE,
  threads = 1
)

saveArchRProject(
  ArchRProj = CLLProj,
  outputDirectory = ArchR_result_dir,
  load = FALSE,
  overwrite = TRUE
)

CLLProj <- loadArchRProject(path = ArchR_result_dir)
```

## Remove doublets
```{r Ignore_doublets}
setwd(result_dir)

CLLProj <- filterDoublets(ArchRProj = CLLProj)
saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, load = FALSE, overwrite = TRUE)
```

## LSI Dimensionality Reduction
```{r LSI_Dimensionality_Reduction}
CLLProj <- addIterativeLSI(
  seed = 1,
  ArchRProj = CLLProj,
  useMatrix = "TileMatrix",
  name = "IterativeLSI",
  iterations = 2,
  clusterParams = list( # See Seurat::FindClusters
    resolution = c(0.2),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30
)

CLLProj <- addIterativeLSI(
  seed = 1,
  ArchRProj = CLLProj,
  useMatrix = "TileMatrix",
  name = "IterativeLSI_0.4",
  iterations = 2,
  clusterParams = list( # See Seurat::FindClusters
    resolution = c(0.4),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30
)

saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, load = FALSE, overwrite = TRUE)
```

## Seurat Clustering
```{r}
setwd(result_dir)

# Try combinations between different LSI resolutions and clustering resolutions
param_combinations <- tidyr::expand_grid(
  dimred = c("IterativeLSI_0.4", "IterativeLSI"),
  resolution = c(0.2, 0.3, 0.4, 0.6)
)

for (row in 1:nrow(param_combinations)) {
  resolution <- param_combinations$resolution[[row]]
  dimred <- param_combinations$dimred[[row]]

  CLLProj <- addClusters(
    input = CLLProj,
    seed = 1,
    reducedDims = dimred,
    method = "Seurat",
    name = glue("Clusters_{resolution}_{dimred}"),
    resolution = resolution
  )
}
saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, load = FALSE, overwrite = TRUE)
```

```{r Count_cells_per_cluster}
# Count number of cells per cluster
table(CLLProj$Clusters_0.6_IterativeLSI)
```


```{r Create_confusion_matrix}
# Create confusion matrix
cM <- confusionMatrix(paste0(CLLProj$Clusters_0.6_IterativeLSI), paste0(CLLProj$Sample))
cM
```


```{r Heatmap_confusion_matrix}
# Show confusion matrix as heatmap
cM <- confusionMatrix(paste0(CLLProj$Clusters_0.6_IterativeLSI), paste0(CLLProj$Sample))
cM
```

## UMAP
```{r UMAP_Embedding}
setwd(paste0(project_dir))

CLLProj <- addUMAP(
  seed = 1,
  ArchRProj = CLLProj,
  reducedDims = "IterativeLSI",
  name = "UMAP",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine"
)

CLLProj <- addUMAP(
  seed = 1,
  ArchRProj = CLLProj,
  reducedDims = "IterativeLSI_0.4",
  name = "UMAP_LSI_0.4",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine"
)

saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, load = FALSE, overwrite = TRUE)
```

## Assign Gene Scores to Clusters
## Identify Marker Genes
```{r}
# Add Gene Scores to Clusters
markersGS <- getMarkerFeatures(
  threads = 1,
  ArchRProj = CLLProj,
  useMatrix = "GeneScoreMatrix",
  groupBy = "Clusters_0.2_IterativeLSI",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

saveRDS(markersGS, file.path(RDS_dir, "markerGS.RDS"))

markersGS_0.3 <- getMarkerFeatures(
  threads = 1,
  ArchRProj = CLLProj,
  useMatrix = "GeneScoreMatrix",
  groupBy = "Clusters_0.3_IterativeLSI",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

saveRDS(markersGS_0.3, file.path(RDS_dir, "markersGS_0.3.RDS"))

markersGS_0.4 <- getMarkerFeatures(
  threads = 1,
  ArchRProj = CLLProj,
  useMatrix = "GeneScoreMatrix",
  groupBy = "Clusters_0.4_IterativeLSI",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

saveRDS(markersGS_0.4, file.path(RDS_dir, "markersGS_0.4.RDS"))

markersGS_0.6 <- getMarkerFeatures(
  threads = 1,
  ArchRProj = CLLProj,
  useMatrix = "GeneScoreMatrix",
  groupBy = "Clusters_0.6_IterativeLSI",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

saveRDS(markersGS_0.6, file.path(RDS_dir, "markersGS_0.6.RDS"))

markersGS_0.4_0.4 <- getMarkerFeatures(
  threads = 1,
  ArchRProj = CLLProj,
  useMatrix = "GeneScoreMatrix",
  groupBy = "Clusters_0.4_IterativeLSI_0.4",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

saveRDS(markersGS_0.4_0.4, file.path(RDS_dir, "markersGS_0.4_0.4.RDS"))

markerList <- getMarkers(markersGS, cutOff = "FDR <= 0.01 & Log2FC >= 1.25")
markerList$C3
```

## Impute Weights
```{r}
setwd(project_dir)
CLLProj <- addImputeWeights(CLLProj,
  seed = 1
)
saveArchRProject(
  ArchRProj = CLLProj,
  outputDirectory = ArchR_result_dir,
  load = FALSE
)
```

## QC
```{r TSS_enrichment}
TSS_enrichment <- getCellColData(CLLProj, select = c("log10(nFrags)", "TSSEnrichment"))
```

```{r TSS_enrichment_plot}
TSS_plot <- ggPoint(
  x = TSS_enrichment[, 1],
  y = TSS_enrichment[, 2],
  colorDensity = TRUE,
  continuousSet = "sambaNight",
  xlabel = "Log10 Unique Fragments",
  ylabel = "TSS Enrichment",
  xlim = c(log10(500), quantile(TSS_enrichment[, 1], probs = 0.99)),
  ylim = c(0, quantile(TSS_enrichment[, 2], probs = 0.99))
) +
  geom_hline(yintercept = 4, lty = "dashed") +
  geom_vline(xintercept = 3, lty = "dashed")

TSS_plot

plotPDF(TSS_plot,
  name = "TSS-vs-Frags.pdf",
  ArchRProj = CLLProj,
  addDOC = FALSE
)
```

```{r Ridge_plot_per_sample}
ridge_plot_sample <- plotGroups(
  ArchRProj = CLLProj,
  groupBy = "Sample",
  colorBy = "cellColData",
  name = "TSSEnrichment",
  plotAs = "ridges"
) +
  scale_fill_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
    panel.border = element_blank(),
    axis.line.x = element_line(linetype = "solid", colour = "black"),
    axis.line.y = element_line(linetype = "solid", colour = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "none"
  )

ridge_plot_sample
```

```{r Boxplot_sample}
boxplot_sample <- plotGroups(
  ArchRProj = CLLProj,
  groupBy = "Sample",
  colorBy = "cellColData",
  name = "TSSEnrichment",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
) +
  scale_colour_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  scale_fill_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
    panel.border = element_blank(),
    axis.line.x = element_line(linetype = "solid", colour = "black"),
    axis.line.y = element_line(linetype = "solid", colour = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "none"
  )
boxplot_sample
```

```{r Ridge_plot_sample_log10}
ridge_plot_sample_log10 <- plotGroups(
  ArchRProj = CLLProj,
  groupBy = "Sample",
  colorBy = "cellColData",
  name = "log10(nFrags)",
  plotAs = "ridges"
) +
  scale_fill_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
    panel.border = element_blank(),
    axis.line.x = element_line(linetype = "solid", colour = "black"),
    axis.line.y = element_line(linetype = "solid", colour = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "none"
  )

ridge_plot_sample_log10
```

```{r boxplot_sample_log10}
boxplot_sample_log10 <- plotGroups(
  ArchRProj = CLLProj,
  groupBy = "Sample",
  colorBy = "cellColData",
  name = "log10(nFrags)",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
) +
  scale_colour_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  scale_fill_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
    panel.border = element_blank(),
    axis.line.x = element_line(linetype = "solid", colour = "black"),
    axis.line.y = element_line(linetype = "solid", colour = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "none"
  )

boxplot_sample_log10
```

```{r Save_TSS_plots}
plotPDF(ridge_plot_sample,
  ridge_plot_sample_log10,
  boxplot_sample,
  boxplot_sample_log10,
  name = "QC-Sample-Statistics.pdf",
  ArchRProj = CLLProj,
  addDOC = FALSE,
  width = 7,
  height = 7
)
```

```{r fragment_size_plot}
fragementsize_plot <- plotFragmentSizes(
  threads = 1,
  ArchRProj = CLLProj
) +
  scale_colour_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
    panel.border = element_blank(),
    axis.line.x = element_line(linetype = "solid", colour = "black"),
    axis.line.y = element_line(linetype = "solid", colour = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
fragementsize_plot
```

```{r TSS_enrichment_profile}
TSS_enrichment_profile <- plotTSSEnrichment(
  threads = 1,
  ArchRProj = CLLProj
) +
  scale_colour_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
    panel.border = element_blank(),
    axis.line.x = element_line(linetype = "solid", colour = "black"),
    axis.line.y = element_line(linetype = "solid", colour = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
TSS_enrichment_profile

plotPDF(fragementsize_plot,
  TSS_enrichment_profile,
  name = "QC-Sample-FragSizes-TSSProfile.pdf",
  ArchRProj = CLLProj,
  addDOC = FALSE,
  width = 7,
  height = 7
)
```

## Save and Load Project
```{r Loading_of_ArchRProject}
setwd(paste0(project_dir))
saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, load = FALSE, overwrite = TRUE)
CLLProj <- loadArchRProject(path = ArchR_result_dir)
```

# Visualise Results

## UMAP
```{r Plot_UMAP_Sample, fig.width=15, fig.height=8}
umap_samples <- plotEmbedding(
  ArchRProj = CLLProj,
  colorBy = "cellColData",
  name = "Sample",
  embedding = "UMAP",
  plotAs = "points",
  pal = c(my_paired_colours_named, my_unpaired_colours_named)
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = rel(1.2)),
    strip.background = element_rect(fill = "white", colour = NA),
    panel.border = element_blank(),
    axis.line = element_line(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.text = element_text(size = 10),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  guides(colour = guide_legend(override.aes = list(size = rel(2))))

umap_samples
saveRDS(umap_samples, paste0(project_dir, "/RDS_Objects/umap_samples_subset4.RDS"))

umap_samples_LSI_0.4 <- plotEmbedding(
  ArchRProj = CLLProj,
  colorBy = "cellColData",
  name = "Sample",
  embedding = "UMAP_LSI_0.4",
  plotAs = "points",
  pal = c(my_paired_colours_named, my_unpaired_colours_named)
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = rel(1.2)),
    strip.background = element_rect(fill = "white", colour = NA),
    panel.border = element_blank(),
    axis.line = element_line(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.text = element_text(size = 10),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  guides(colour = guide_legend(override.aes = list(size = rel(2))))

umap_samples_LSI_0.4

saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, load = FALSE, overwrite = TRUE)

ggAlignPlots(umap_samples, umap_samples_LSI_0.4,
  type = "h"
)
```

```{r, fig.width=25, fig.height=10}
# Try combinations between different LSI resolutions and clustering resolutions
param_combinations <- tidyr::expand_grid(
  dimred = c("IterativeLSI", "IterativeLSI_0.4"),
  resolution = c(0.2, 0.3, 0.4, 0.6)
) %>% mutate("umap" = case_when(
  dimred == "IterativeLSI" ~ "",
  dimred == "IterativeLSI_0.4" ~ "_LSI_0.4"
))

create_clusters <- function() {
  all_plots <- list()
  for (row in 1:nrow(param_combinations)) {
    resolution <- param_combinations$resolution[[row]]
    dimred <- param_combinations$dimred[[row]]
    umap <- param_combinations$umap[[row]]

    plot <- plotEmbedding(
      ArchRProj = CLLProj,
      colorBy = "cellColData",
      name = glue("Clusters_{resolution}_{dimred}"),
      embedding = glue("UMAP{umap}"),
      plotAs = "points"
    ) +
      labs(x = "UMAP1", y = "UMAP2") +
      scale_color_discrete(
        name = "Cluster",
        type = cluster_colours
      ) +
      theme_bw() +
      theme(
        strip.text = element_text(face = "bold", size = rel(1.2)),
        strip.background = element_rect(fill = "white", colour = NA),
        panel.border = element_blank(),
        axis.line = element_line(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_text(face = "bold"),
        panel.grid = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = "bottom",
        legend.direction = "horizontal"
      ) +
      guides(colour = guide_legend(override.aes = list(size = rel(2))))

    print(plot)
    all_plots[[row]] <- plot
  }

  return(all_plots)
}

all_clusters <- create_clusters()

cowplot::plot_grid(ggpubr::ggarrange(umap_samples, umap_samples_LSI_0.4, nrow = 2, ncol = 1),
  ggpubr::ggarrange(plotlist = all_clusters, nrow = 2, ncol = 4),
  ncol = 2, rel_widths = c(1, 4)
) %>%
  ggsave(.,
    filename = file.path(ArchR_result_dir, "Plots/Plot-UMAP-Samples_Clusters_Res_0.2_0.6.pdf"),
    width = 80,
    height = 30,
    unit = "cm"
  )

cowplot::plot_grid(ggpubr::ggarrange(umap_samples, umap_samples_LSI_0.4, nrow = 2, ncol = 1),
  ggpubr::ggarrange(plotlist = all_clusters, nrow = 2, ncol = 4),
  ncol = 2, rel_widths = c(1, 4)
)
```

## Heatmap Marker Genes
```{r}
markerGenes <- c(
  "CD34", # Early Progenitor
  "GATA1", # Erythroid
  "PAX5", "MS4A1", "EBF1", "MME", # B-Cell Trajectory
  "CD14", "CEBPB", "MPO", # Monocytes
  "IRF8",
  "CD3D", "CD8A", "TBX21", "IL7R", # TCells
  "CD19", "CD5", "FCER2", "CD79b", # CLL B-Cell Marker
  "GNLY", "NKG7", "NCAM1" # NK cells
)
```

```{r}
heatmapGS <- plotMarkerHeatmap(
  seMarker = markersGS,
  cutOff = "FDR <= 0.01 & Log2FC >= 1.25",
  labelMarkers = markerGenes,
  transpose = TRUE
)
heatmapGS
saveRDS(heatmapGS, file = file.path(RDS_dir, "heatmapGS.RDS"))
```

```{r}
heatmapGS_0.3 <- plotMarkerHeatmap(
  seMarker = markersGS_0.3,
  cutOff = "FDR <= 0.01 & Log2FC >= 1.25",
  labelMarkers = markerGenes,
  transpose = TRUE
)
heatmapGS_0.3
saveRDS(heatmapGS_0.3, file = file.path(RDS_dir, "heatmapGS_0.3.RDS"))
```

```{r}
heatmapGS_0.4 <- plotMarkerHeatmap(
  seMarker = markersGS_0.4,
  cutOff = "FDR <= 0.01 & Log2FC >= 1.25",
  labelMarkers = markerGenes,
  transpose = TRUE
)
heatmapGS_0.4
saveRDS(heatmapGS_0.4, file = file.path(RDS_dir, "heatmapGS_0.4.RDS"))
```

```{r}
heatmapGS_0.6 <- plotMarkerHeatmap(
  seMarker = markersGS_0.6,
  cutOff = "FDR <= 0.01 & Log2FC >= 1.25",
  labelMarkers = markerGenes,
  transpose = TRUE
)
heatmapGS_0.6
saveRDS(heatmapGS_0.6, file = file.path(RDS_dir, "heatmapGS_0.6.RDS"))
```

```{r}
plotPDF(heatmapGS, heatmapGS_0.3, heatmapGS_0.4, heatmapGS_0.6,
  name = "GeneScores-Marker-Heatmap-ClusterRes_0.2_0.6",
  width = 8,
  height = 6,
  ArchRProj = CLLProj,
  addDOC = FALSE
)
```

```{r}
ComplexHeatmap::draw(heatmapGS, heatmap_legend_side = "bot", annotation_legend_side = "bot")

ComplexHeatmap::draw(heatmapGS_0.3, heatmap_legend_side = "bot", annotation_legend_side = "bot")

ComplexHeatmap::draw(heatmapGS_0.4, heatmap_legend_side = "bot", annotation_legend_side = "bot")

ComplexHeatmap::draw(heatmapGS_0.6, heatmap_legend_side = "bot", annotation_legend_side = "bot")
```

## Visualise Marker Genes on an Embedding
### UMAP WO Imputation weights
```{r}
UMAP_marker_genes <- plotEmbedding(
  ArchRProj = CLLProj,
  colorBy = "GeneScoreMatrix",
  name = markerGenes,
  embedding = "UMAP",
  quantCut = c(0.01, 0.95),
  imputeWeights = NULL,
  plotAs = "points"
)
```

```{r fig.width=7, fig.height=15}
UMAP_marker_genes_figure <- lapply(UMAP_marker_genes, function(x) {
  x + guides(color = "none", fill = "none") +
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
})

do.call(cowplot::plot_grid, c(list(ncol = 3), UMAP_marker_genes_figure)) %>%
  ggsave(.,
    filename = file.path(ArchR_result_dir, "Plots/Plot-UMAP-Marker-Genes-WO-Imputation.pdf"),
    unit = "cm",
    width = 17,
    height = 40
  )

do.call(cowplot::plot_grid, c(list(ncol = 3), UMAP_marker_genes_figure))
```

### UMAP W  Imputation weights
```{r }
UMAP_marker_genes_imputation <- plotEmbedding(
  ArchRProj = CLLProj,
  colorBy = "GeneScoreMatrix",
  name = markerGenes,
  embedding = "UMAP_LSI_0.4",
  imputeWeights = getImputeWeights(CLLProj),
  plotAs = "points"
)
```

```{r fig.width=7, fig.height=15}
UMAP_marker_genes_imputation_figure <- lapply(UMAP_marker_genes_imputation, function(x) {
  x + guides(color = FALSE, fill = FALSE) +
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3), UMAP_marker_genes_imputation_figure)) %>%
  ggsave(.,
    filename = file.path(ArchR_result_dir, "Plots/Plot-UMAP-Marker-Genes-W-Imputation.pdf"),
    unit = "cm",
    width = 17,
    height = 40
  )

do.call(cowplot::plot_grid, c(list(ncol = 3), UMAP_marker_genes_imputation_figure))
```

###Track plotting
```{r}
tracks <- plotBrowserTrack(
  ArchRProj = CLLProj,
  groupBy = "Clusters_0.6_IterativeLSI",
  geneSymbol = markerGenes,
  upstream = 50000,
  downstream = 50000
)
saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_result_dir, load = FALSE, overwrite = TRUE)
```

```{r}
plotPDF(
  plotList = tracks,
  name = "Plot-Tracks-Marker-Genes.pdf",
  ArchRProj = CLLProj,
  addDOC = FALSE,
  width = 5,
  height = 5
)

for (track in tracks) {
  grid::grid.newpage()
  grid::grid.draw(track)
}
```
