---
title: "Compare_BCR_mtDNA_clones"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
# Goal
The aim of this script is to extract mitochondrial mutations for all samples from scRNA data and compare them to BCR clones of the same samples

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(Signac)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(SummarizedExperiment)
library(ggalluvial)
library(ggrepel)
library(Seurat)
library(circlize)
library(ComplexHeatmap)
library(Matrix)
```

# Set paths
```{r define_directories }
source("config_paths.R")

data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "GEX")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "mtDNA_BCR_analysis")
plotting_dir <- file.path(analysis_dir, "Plots")
RDS_dir <- file.path(analysis_dir, "RDS_objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")

ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")
GEX_analysis_RDS_dir <- file.path(result_dir, "GEX_analysis/RDS_objects")
VDJ_analysis_RDS_dir <- file.path(result_dir, "VDJ_analysis/RDS_Objects")
```

# Load data
```{r}
source(file.path(project_dir, "/scripts/publication_theme.R"))
```

```{r}
metadata_all <- read_tsv(file = file.path(data_dir, "20240723_Tidy_MBL_CLL_metadata.tsv")) 
gex_scaled_all <- readRDS(file.path(GEX_analysis_RDS_dir, "gex_scaled_all.RDS"))
new_BCR_paired <- readRDS(file.path(VDJ_analysis_RDS_dir, "new_BCR_paired.RDS"))
```

```{r}
BCR_clonotype_annotation <- new_BCR_paired %>%
  bind_rows() %>%
  dplyr::select(label_cellbarcode, new_common_ct_id) %>%
  dplyr::rename("BCR_clonotype" = new_common_ct_id) %>%
  dplyr::mutate(BCR_clonotype = as.factor(str_remove(BCR_clonotype, "_clonotype"))) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#") %>%
  tidyr::unite("label_cellbarcode", label:cellbarcode, sep = "_") %>%
  with(setNames(BCR_clonotype, label_cellbarcode))

gex_scaled_all <- AddMetaData(gex_scaled_all, 
                              metadata = BCR_clonotype_annotation, 
                              col.name = "BCR_clonotype")
```

```{r}
matched_samples <- metadata_all %>%
  dplyr::filter(data_type %in% c("GEX", "BCR")) %>%
  dplyr::select(label, patient_id, plotting_id, MBL_subtype, data_type) %>%
  tidyr::pivot_wider(id_cols = c(label, patient_id, plotting_id, MBL_subtype), names_from = data_type, values_from = data_type) %>%
  dplyr::filter(!(is.na(GEX) | is.na(BCR))) %>%
  pull(label)

mtDNA_samples <- metadata_all %>%
  dplyr::filter(
    data_type == "GEX",
    label %in% matched_samples
  ) %>%
  dplyr::select(pool, label) %>%
  deframe()

BCR_samples <- metadata_all %>%
  dplyr::filter(
    data_type == "BCR",
    label %in% matched_samples
  ) %>%
  dplyr::select(pool, label) %>%
  deframe()

metadata_matched_samples <- metadata_all %>%
  dplyr::filter(pool %in% c(names(mtDNA_samples), names(BCR_samples)))

# mgatk on GEX data
mgatk_file_paths <- file.path(library_path, paste0(names(mtDNA_samples), ".mgatk"))
```

```{r}
# previously called clonal variants on ATAC data
ATAC_MBL_cln_variants <- c(readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/cln.variants.MBL.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/cln.variants.MBL.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/cln.variants.MBL.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset4/RDS_objects/cln.variants.MBL.RDS")))

ATAC_cln_variants <- c(readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/cln.variants.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/cln.variants.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/cln.variants.RDS")))
```

```{r}
# VAFs of previously called clonal variants on ATAC data
ATAC_VAFs_cln_MBL <- readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset4/RDS_objects/VAFs.cln.MBL.RDS"))
```

```{r}
# previously called clonotypes on ATAC data
ATAC_MBL_clonotypes <- c(readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/clonotypes_MBL.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/clonotypes_MBL.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/clonotypes_MBL.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset4/RDS_objects/clonotypes_MBL.RDS")))

ATAC_clonotypes <- c(readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/clonotypes.RDS")),
                     readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/clonotypes.RDS")),
                         readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/clonotypes.RDS")))
```

```{r}
pool2label <- metadata_matched_samples %>%
  select(pool, label) %>%
  deframe()

# Individual mgatk files per sample
mito.data.idv <- mgatk_file_paths %>%
  map(~ ReadMGATK(.)) %>%
  set_names(mtDNA_samples)

saveRDS(mito.data.idv, file = file.path(RDS_dir, "mito.data.idv.RDS"))
```

```{r}
# New filtering function also accounting for mtseq depth of min 10!
 filter_depth_from_mitodata <- function(mgatk_idv_list, archrproj) {
   # Filter cells
   for (patient in unique(mtDNA_samples)) {
     keep_col_depth <- mgatk_idv_list[[patient]]$depth %>%
       dplyr::filter(mito.depth >= 10) %>%
       rownames()

     mgatk_idv_list[[patient]]$counts <- mgatk_idv_list[[patient]]$counts[, keep_col_depth]

     mgatk_idv_list[[patient]]$depth <- mgatk_idv_list[[patient]]$depth[keep_col_depth, , drop = FALSE]
   }

   return(mgatk_idv_list)
 }

 mito.data.idv <- filter_depth_from_mitodata(mito.data.idv, CLLProj)
 saveRDS(mito.data.idv, file = file.path(RDS_dir, "mito.data.idv.RDS"))
```

```{r}
# Paired mgatk files per patient
mito.data <- list()

sample_subset <- metadata_matched_samples %>%
  dplyr::filter(
    label %in% mtDNA_samples,
    str_starts(label, "CLL"),
    data_type == "GEX"
  ) %>%
  pull(patient_id) %>%
  unique()

sample_subset_sample <- metadata_matched_samples %>%
  dplyr::filter(
    patient_id %in% sample_subset,
    str_starts(label, "CLL"),
    data_type == "GEX"
  ) %>%
  pull(label)

mito.data <- list()
for (i in sample_subset) {
  subset_names <- str_subset(names(mito.data.idv), i)
  subset_names_index <- match(subset_names, names(mito.data.idv))

  df_1 <- mito.data.idv[[subset_names_index[[1]]]]$counts
  df_2 <- mito.data.idv[[subset_names_index[[2]]]]$counts

  colnames(df_1) <- paste(subset_names[1], colnames(df_1), sep = "#")
  colnames(df_2) <- paste(subset_names[2], colnames(df_2), sep = "#")

  subset_df <- cbind(df_1, df_2)
  rm(df_1, df_2)
  mito.data <- append(mito.data, subset_df)
}
names(mito.data) <- sample_subset

saveRDS(mito.data, file = file.path(RDS_dir, "mito.data.RDS"))
mito.data <- readRDS(file = file.path(RDS_dir, "mito.data.RDS"))
```

```{r}
# Individual mgatk per patient
sample_subset_MBL <- metadata_matched_samples %>%
  dplyr::filter(
    label %in% mtDNA_samples,
    str_starts(label, "MBL"),
    data_type == "GEX"
  ) %>%
  pull(patient_id) %>%
  unique()

mito.data.MBL <- list()
for (i in sample_subset_MBL) {
  subset_names <- str_subset(names(mito.data.idv), i)
  subset_names_index <- match(subset_names, names(mito.data.idv))

  df_1 <- mito.data.idv[[subset_names_index[[1]]]]$counts
  colnames(df_1) <- paste(subset_names[1], colnames(df_1), sep = "#")
  mito.data.MBL <- append(mito.data.MBL, df_1)
}
names(mito.data.MBL) <- sample_subset_MBL
saveRDS(mito.data.MBL, file = file.path(RDS_dir, "mito.data.MBL.RDS"))
mito.data.MBL <- readRDS(file = file.path(RDS_dir, "mito.data.MBL.RDS"))
```

# Analysis
## Variants identification
```{r Get_mt_variants}
# Get variants cumulative per patient over both samples
variant.df <- mito.data %>%
  map(~ IdentifyVariants(., refallele = mito.data.idv[[1]]$refallele)) %>%
  imap( ~ .x %>%
    dplyr::mutate(sample_id = .y, .before = "nucleotide"))

# Export_mt_variants
variant.df %>%
  imap(~ write_tsv(.x,
    quote = "none",
    file = file.path(analysis_dir, paste0("MitochondrialVariants/Variants_", .y, ".csv"))
  ))

saveRDS(variant.df, file = file.path(RDS_dir, "variant.df.RDS"))
variant.df <- readRDS(file = file.path(RDS_dir, "variant.df.RDS"))

variant.df.MBL <- mito.data.MBL %>%
  map(~ IdentifyVariants(., refallele = mito.data.idv[[1]]$refallele)) %>%
  imap(, ~ .x %>%
    dplyr::mutate(sample_id = .y, .before = "nucleotide"))

variant.df.MBL %>%
  imap(~ write_tsv(.x,
    quote = "none",
    file = file.path(analysis_dir, paste0("MitochondrialVariants/Variants_", .y, ".csv"))
  ))

saveRDS(variant.df.MBL, file = file.path(RDS_dir, "variant.df.MBL.RDS"))
variant.df.MBL <- readRDS(file = file.path(RDS_dir, "variant.df.MBL.RDS"))
```

```{r Plot_mt_variants}
all.variant.df.plots <- c(variant.df, variant.df.MBL) %>%
  map(VariantPlot) %>%
  iwalk(~ ggsave(
    filename = file.path(analysis_dir, paste0("MitochondrialVariants/Plot_Variance_Mean_Ratio_", .y, ".pdf")),
    plot = .x,
    width = 5,
    height = 3
  ))
```

## Filtering clonal mtDNA mutations
```{r Filter_clonal_mt_variants}
# Get clonal variants
cln.variants <- variant.df %>%
  imap(~dplyr::filter(., variant %in% ATAC_cln_variants[[.y]][["variant"]]))
  
# # Export_cln_mito_variants
cln.variants %>%
  imap(~ write_tsv(.x, file = file.path(analysis_dir, paste0("MitochondrialVariants/Clonal_Variants_", .y, ".tsv"))))

saveRDS(cln.variants, file = file.path(RDS_dir, "cln.variants.RDS"))

# Get clonal variants
cln.variants.MBL <- variant.df.MBL %>%
  imap(~dplyr::filter(., variant %in% ATAC_MBL_cln_variants[[.y]][["variant"]]))

# Export_cln_mito_variants
cln.variants.MBL %>%
  imap(~ write_tsv(.x, file = paste0(analysis_dir, "/MitochondrialVariants/Clonal_Variants_", .y, ".tsv")))

saveRDS(cln.variants.MBL, file = file.path(RDS_dir, "cln.variants.MBL.RDS"))
```

### Heteroplasmy calculation
```{r VAFs_clonal_mt_variants}
# Get clonal VAFs
VAFs.cln <- mito.data %>%
  map2(cln.variants, ~ AlleleFreq(.,
    variants = .y$variant,
    assay = NULL
  ))

VAFs.cln %>%
  imap( ~ writeMM(., file = file.path(analysis_dir, paste0("MitochondrialVariants/VAFs_Variants_sparse_", .y, ".mtx"))))

saveRDS(VAFs.cln, file = file.path(RDS_dir, "VAFs.cln.RDS"))

VAFs.cln.MBL <- mito.data.MBL%>%
  map2(cln.variants.MBL, ~ AlleleFreq(.,
    variants = .y$variant,
    assay = NULL
  ))

# export in sparse format
VAFs.cln.MBL %>%
  imap( ~ writeMM(., file = file.path(analysis_dir, paste0("MitochondrialVariants/VAFs_Variants_sparse_", .y, ".mtx"))))

saveRDS(VAFs.cln.MBL, file = file.path(RDS_dir, "VAFs.cln.MBL.RDS"))
```


```{r Summarise_mutation_counts}
 bind_rows(bind_rows(VAFs.cln.MBL[-c(1,4)] %>%
  map(~ t(.) %>% as.matrix() %>%
    .[as.logical(rowSums(. != 0)), as.logical(colSums(. != 0))] %>%
    as_tibble(rownames = "label_cellbarcode") %>% 
    pivot_longer(cols=-label_cellbarcode, names_to = "variant_id", values_to = "VAF")),
  .id = "patient_id"),
 bind_rows(VAFs.cln %>%
  map(~ t(.) %>% as.matrix() %>%
    .[as.logical(rowSums(. != 0)), as.logical(colSums(. != 0))] %>%
    as_tibble(rownames = "label_cellbarcode") %>% 
    pivot_longer(cols=-label_cellbarcode, names_to = "variant_id", values_to = "VAF")),
  .id = "patient_id")) %>% 
  left_join(metadata_all %>% dplyr::filter(data_type=="GEX") %>%
              dplyr::select(plotting_id, patient_id) %>% unique) %>% 
  dplyr::filter(plotting_id !="P9dup") %>% 
  dplyr::select(plotting_id, variant_id) %>% 
  unique() %>% 
  group_by(plotting_id) %>% 
  dplyr::summarise(variant_count=n()) %>% 
  dplyr::mutate(plotting_id=fct_reorder(plotting_id, -variant_count)) %>% 
  ggplot2::ggplot(aes(x=plotting_id, y=variant_count))+
  geom_bar(stat="identity",fill="midnightblue")+
  labs(y="mtDNA mutation counts [#]", x="")+
  scale_y_continuous(expand = expansion(mult = c(0, NA)))+
  theme_pub()
```

## Coverage mtDNA variants
```{r}
#Calculates means per position across all cells --> then calculates rolling mean across positions
pull_coverage <- function(SE, cells, resolution = 5){
  zoo::rollmean(rowMeans(assays(SE)[['coverage']]), resolution)
}

mtmut_positions_patient <- bind_rows(bind_rows(VAFs.cln.MBL %>%
  map(~ as.matrix(.) %>%
    as_tibble(rownames = "variant_id") %>%
    .[, as.logical(colSums(. != 0))]), .id = "patient_id"),
 bind_rows(VAFs.cln %>%
  map(~ as.matrix(.) %>%
    as_tibble(rownames = "variant_id") %>%
    .[, as.logical(colSums(. != 0))]), .id = "patient_id")) %>% 
  left_join(metadata_all %>% dplyr::filter(data_type=="GEX") %>%
              dplyr::select(plotting_id, patient_id) %>% unique) %>% 
  dplyr::select(patient_id, variant_id) %>% 
  tidyr::separate(variant_id, into = c("position", NA), sep = "(A|C|G|T)", remove = FALSE)

mtmut_positions_label <- mtmut_positions_patient %>% 
  dplyr::filter(str_starts(patient_id, "CLL")) %>% 
  dplyr::mutate(label=paste0(patient_id, ".2")) %>% 
  bind_rows(mtmut_positions_patient %>% 
              dplyr::mutate(label=if_else(str_starts(patient_id, "CLL"), paste0(patient_id, ".1"), patient_id))) %>% 
  arrange(label) %>% 
  dplyr::mutate(position=as.numeric(position)) %>% 
  split(~label)

coverage_all_samples <- file.path(library_path, paste0(names(mtDNA_samples), ".mgatk"), paste0(names(mtDNA_samples), ".rds")) %>% 
  map2(mtDNA_samples, ~data.frame(
  position = zoo::rollmean(1:16569, 5),
  coverage = pull_coverage(readRDS(.))) %>% 
  dplyr::left_join(mtmut_positions_label[[.y]])
  ) %>% 
  set_names(mtDNA_samples)

write_tsv(bind_rows(coverage_all_samples, .id = "label"),
          file=file.path(tsv_dir, "mtDNA_coverage_GEX_samples.tsv"))
```

```{r}
coverage_plots_variants <- coverage_all_samples %>%
  imap(~ggplot(., aes(x = position, y = coverage, label=variant_id)) +
  geom_line(color = "black") +
  geom_label_repel(label.size = NA, 
                   fill=NA, max.overlaps = 47,
                   nudge_y = 40, 
                   colour = "darkred")+
  scale_y_continuous(expand = expansion(mult=c(0, 0.05)))+
  scale_x_continuous(expand = expansion(mult=c(0, 0.01)))+
  theme_pub()+
  labs(x = "Position on mtDNA chromosome", 
       y = "Rolling mean coverage across cells",
       title=.y))
coverage_plots_variants
```

```{r fig.width=10}
cowplot::plot_grid(plotlist = coverage_plots_variants[16:17], cols = 2, rows = 1)
```

## FindClonotypes 
```{r }
# Cluster together per patient and filter for non-zero VAF cells exclude patients with variants "2617A>G", "2617A>T" as present in every patient seem to be artefacts
mito_VAFs_MBL <- VAFs.cln.MBL[-c(4)] %>%
  map(~ as.matrix(.) %>%
    as_tibble(rownames = "variant_id") %>%
    column_to_rownames(var = "variant_id") %>%
    .[, as.logical(colSums(. != 0))] %>%
    .[,colSums(is.na(.))<nrow(.)] %>%
    dplyr::mutate(across(where(is.numeric), ~ ifelse(. < 10e-6 & .>0, 10e-6, .))) %>%
    as.matrix() %>%
    as("sparseMatrix") %>%
    Seurat::CreateSeuratObject(assay = "alleles"))

mito_VAFs <- VAFs.cln %>%
  map(~ as.matrix(.) %>%
    as_tibble(rownames = "variant_id") %>%
    column_to_rownames(var = "variant_id") %>%
    .[, as.logical(colSums(. != 0))] %>%
    .[,colSums(is.na(.))<nrow(.)] %>%
    dplyr::mutate(across(where(is.numeric), ~ ifelse(. < 10e-6 & .>0, 10e-6, .))) %>%
    as.matrix() %>%
    as("sparseMatrix") %>%
    CreateSeuratObject(assay = "alleles"))

clonotypes_MBL_new <- mito_VAFs_MBL[-c(1,2,7,12)] %>%
  map(~ FindClonotypes(.))

saveRDS(object = clonotypes_MBL, file = file.path(RDS_dir, "clonotypes_MBL.RDS"))


clonotypes <- mito_VAFs[-c(5)] %>%
  map(~ FindClonotypes(.))

saveRDS(object = clonotypes, file = file.path(RDS_dir, "clonotypes.RDS"))

find_clonotype_clusters_MBL <- clonotypes_MBL %>%
  map(~ .@active.ident %>%
    data.frame() %>%
    rownames_to_column("label_cellbarcode") %>%
    dplyr::rename(Cluster = ".")) %>%
  imap( ~ write_tsv(.x, file = file.path(analysis_dir, paste0("MitochondrialVariants/FindClonotypesClusters_", .y, ".tsv"))))

find_clonotype_clusters <- clonotypes %>%
  map(~ .@active.ident %>%
    data.frame() %>%
    rownames_to_column("label_cellbarcode") %>%
    dplyr::rename(Cluster = ".")) %>%
  imap (~write_tsv(.x, file = file.path(analysis_dir, paste0("MitochondrialVariants/FindClonotypesClusters_", .y, ".tsv"))))
```

```{r fig.width=6}
bind_rows(bind_rows(new_BCR_paired) %>% 
  dplyr::select(patient_id, new_common_ct_id) %>% 
  left_join(metadata_all %>% dplyr::filter(data_type=="GEX") %>%
              dplyr::select(plotting_id, patient_id) %>% unique) %>%
  unique() %>% 
  dplyr::filter(plotting_id !="P9dup") %>% 
  group_by(plotting_id) %>% 
  dplyr::summarise(clone_counts=n()) %>%
  dplyr::mutate("method"="BCR"),
find_clonotype_clusters_MBL %>% 
  bind_rows(.id = "patient_id") %>% dplyr::select(patient_id, Cluster) %>%  
  left_join(metadata_all %>% dplyr::filter(data_type=="GEX") %>%
              dplyr::select(plotting_id, patient_id) %>% unique) %>%
  dplyr::filter(plotting_id !="P9dup") %>% 
  unique() %>% 
  group_by(plotting_id) %>% 
  dplyr::summarise(clone_counts=n())%>%
  dplyr::mutate("method"="mtDNA mutation"),
find_clonotype_clusters %>% 
  bind_rows(.id = "patient_id") %>% dplyr::select(patient_id, Cluster) %>%  
  left_join(metadata_all %>% dplyr::filter(data_type=="GEX") %>%
              dplyr::select(plotting_id, patient_id) %>% unique) %>%
  dplyr::filter(plotting_id !="P9dup") %>% 
  unique() %>% 
  group_by(plotting_id) %>% 
  dplyr::summarise(clone_counts=n())%>%
  dplyr::mutate("method"="mtDNA mutation"))%>% 
  dplyr::mutate(plotting_id=fct_reorder(plotting_id, -clone_counts)
                ) %>% 
  ggplot2::ggplot(aes(x=plotting_id, y=clone_counts, fill=method))+
  geom_bar(stat="identity", position=position_dodge2(width = 0.9, preserve = "single"))+
  labs(y="clone counts [#]", x="")+
  scale_y_log10(expand = expansion(mult = c(0, NA)))+
  scale_fill_manual(name="Tracing method", values = c("BCR"="lightgoldenrod3", 
                                                      "mtDNA mutation"="indianred"))+
  theme_pub()+
  theme(axis.text.x=element_text(angle=90))
```

## UMAP mt clones
```{r fig.width=10, fig.height=10}
gex_scaled_all <- AddMetaData(gex_scaled_all,
  metadata = c(find_clonotype_clusters, find_clonotype_clusters_MBL) %>% 
    map(~dplyr::mutate(., label_cellbarcode = str_replace(label_cellbarcode, "#", "_"))) %>%
    bind_rows() %>% 
    deframe(),
  col.name = "mt_clonotypes")

matched_mt_BCR_clonotypes <- gex_scaled_all@meta.data %>%
  as_tibble(rownames = NA) %>%
  dplyr::filter(
    !is.na(BCR_clonotype),
    !is.na(mt_clonotypes)
  ) %>%
  rownames()

gex_scaled_clonotypes <- subset(gex_scaled_all, 
                                cells = matched_mt_BCR_clonotypes)

sample_umap <- DimPlot(
  object = gex_scaled_clonotypes,
  reduction = "umap",
  group.by = "orig.ident",
  split.by = "plotting_id",
  pt.size = 0.001, 
  ncol = 5) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.line = element_line(),
    legend.position = "none",
    aspect.ratio = 1,
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

mt_clontoype_umap <- DimPlot(
  object = gex_scaled_clonotypes,
  reduction = "umap",
  group.by = "mt_clonotypes",
  split.by = "plotting_id",
  cols = annotation_colours_named,
  ncol = 5,
  shuffle = FALSE,
  pt.size = 0.001, 
) +
  scale_colour_manual(values = sample(colorRampPalette(brewer.pal(8, "Set2"))(length(unique(gex_scaled_clonotypes@meta.data$mt_clonotypes))))) +
  labs(x = "UMAP1", y = "UMAP2", title = "mt clones") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.line = element_line(),
    legend.position = "right",
    aspect.ratio = 1,
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_text(hjust = 0.5, 
                                size = 10,
                                face = "plain")
  )

saveRDS(mt_clontoype_umap, file.path(RDS_dir, "mt_clontoype_umap.RDS"))

BCR_clontoype_umap <- DimPlot(gex_scaled_clonotypes,
  reduction = "umap",
  raster = FALSE,
  shuffle = FALSE,
  split.by = "plotting_id",
  cols = annotation_colours_named,
  ncol = 5,
  group.by = "BCR_clonotype",
  pt.size=0.001
) +
  scale_colour_manual(values = sample(colorRampPalette(brewer.pal(12, "Set2"))(length(unique(gex_scaled_clonotypes@meta.data$BCR_clonotype))))) +
  labs(x = "UMAP1", y = "UMAP2", title = "BCR clones") +
  theme_pub() +
  theme(
    legend.position = "none",
    aspect.ratio = 1
  ) +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.line = element_line(),
    legend.position = "none",
    aspect.ratio = 1,
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_text(hjust = 0.5, 
                                size = 10,
                                face = "plain")
  )
saveRDS(BCR_clontoype_umap, file.path(RDS_dir, "BCR_clontoype_umap.RDS"))

sample_umap
mt_clontoype_umap
BCR_clontoype_umap
```

### MBL11 
```{r}
matched_mt_BCR_clonotypes_MBL11 <- gex_scaled_all@meta.data %>%
  as_tibble(rownames = NA) %>%
  dplyr::filter(
    !is.na(BCR_clonotype),
    !is.na(mt_clonotypes),
    patient_id == "MBL11",
    My_annotation =="MBL B Cell"
  ) %>%
  rownames()

gex_scaled_clonotypes_MBL11 <- subset(gex_scaled_all, cells = matched_mt_BCR_clonotypes_MBL11)

sample_umap_MBL11 <- DimPlot(
  object = gex_scaled_clonotypes_MBL11,
  reduction = "umap",
  cols="antiquewhite2",
  group.by = "orig.ident",
  pt.size=0.001) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.line = element_line(),
    legend.position = "none",
    aspect.ratio = 1,
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_text(hjust = 0.5, 
                                size = 10,
                                face = "plain")
  )

set.seed(128)
mt_colours <- sample(colorRampPalette(brewer.pal(12, "Set2"))(length(unique(gex_scaled_clonotypes_MBL11@meta.data$mt_clonotypes))))
names(mt_colours) <- gex_scaled_clonotypes_MBL11@meta.data$mt_clonotypes %>% unique()
  
mt_clontoype_umap_MBL11 <- DimPlot(
  object = gex_scaled_clonotypes_MBL11,
  reduction = "umap",
  group.by = "mt_clonotypes",
  cols = annotation_colours_named,
  shuffle = FALSE,
  pt.size = 0.001, 
) +
  scale_colour_manual(values = mt_colours) +
  labs(x = "UMAP1", 
       y = "UMAP2", 
       title = paste0("mt clones\n(n=", gex_scaled_clonotypes_MBL11@meta.data$mt_clonotypes %>% unique() %>% length, ")")) +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.line = element_line(),
    legend.position = "none",
    aspect.ratio = 1,
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_text(hjust = 0.5, 
                                size = 10,
                                face = "plain")
  )

saveRDS(mt_clontoype_umap_MBL11, file.path(RDS_dir, "mt_clontoype_umap_MBL11.RDS"))

BCR_colours <-rep("black", length(unique(gex_scaled_clonotypes@meta.data$BCR_clonotype))) %>% 
  set_names(unique(gex_scaled_clonotypes@meta.data$BCR_clonotype)) %>% 
  enframe(name="BCR_clone", value="colour") %>% 
  dplyr::mutate(colour=dplyr::if_else(BCR_clone =="907","#ef8fd5", "#c9c9c9")) %>% 
  deframe()

BCR_clonotype_umap_MBL11 <- DimPlot(gex_scaled_clonotypes_MBL11,
  reduction = "umap",
  raster = FALSE,
  shuffle = FALSE,
  cols = annotation_colours_named,
  group.by = "BCR_clonotype",
  pt.size=0.001
) +
  scale_colour_manual(values = BCR_colours) +
  labs(x = "UMAP1", 
       y = "UMAP2", 
       title = paste0("BCR clones\n(n=", gex_scaled_clonotypes_MBL11@meta.data$BCR_clonotype %>% unique() %>% length, ")")) +
  theme_pub() +
  theme(
    legend.position = "none",
    aspect.ratio = 1
  ) +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.line = element_line(),
    legend.position = "none",
    aspect.ratio = 1,
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_text(hjust = 0.5, 
                                size = 10,
                                face = "plain")
  )
saveRDS(BCR_clonotype_umap_MBL11, file.path(RDS_dir, "BCR_clontoype_umap_MBL11.RDS"))
```


```{r, fig.height=6, fig.width=6}
# Filtered out all clonotype matchings which were supported by only one cell
input_alluvial_MBL11 <- gex_scaled_clonotypes_MBL11@meta.data %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, mt_clonotypes, BCR_clonotype) %>%
  dplyr::mutate(combination = paste0(mt_clonotypes, "_", BCR_clonotype)) %>%
  group_by(combination) %>%
  dplyr::summarise(counts = n()) %>%
  dplyr::arrange(desc(counts)) %>%
  tidyr::separate(combination, into = c("mt_clonotype", "BCR_clonotype"), sep = "_", remove = FALSE) %>%
  rowid_to_column("id") %>%
  mutate(freq = counts / sum(counts) * 100) %>%
  dplyr::mutate(
    BCR_clonotype = paste0("BCR_clone_", BCR_clonotype)) %>% 
  dplyr::mutate(BCR_clonotype=fct_reorder(BCR_clonotype, freq),
                mt_clonotype=factor(mt_clonotype, levels=rev(c(0, 5,3, 1, 2,4,6,7,8,9,10,11,12,13, 14))))
  

clonotype_matching_plot_MBL11 <- ggplot(
  input_alluvial_MBL11 %>% 
    dplyr::arrange(desc(BCR_clonotype)),
  aes(y = freq, 
      axis1 = BCR_clonotype, 
      axis2 = mt_clonotype)
) +
  geom_alluvium(aes(fill = mt_clonotype), width = 1 / 12) +
  geom_stratum(size=0.1,
    width = 1 / 12,
    color = "lightgrey"
  ) +
  scale_x_discrete(
    limits = c("BCR clones", "mt clones"),
    expand = c(.04, .04)
  ) +
  scale_y_continuous(expand = expansion(mult=0, 0))+
  scale_fill_manual(values = mt_colours) +
  labs(y = "Frequency [%]", x = "", title = paste0("n= ", sum(input_alluvial_MBL11$counts), " cells" )) +
  theme_pub() +
  theme(legend.position = "none",
        plot.title = element_text(face="plain",
                                  hjust=0.5))

saveRDS(clonotype_matching_plot_MBL11, file.path(RDS_dir, "clonotype_matching_plot_MBL11.RDS"))
```

```{r}
# 28 907 BCR clones get filtered out
input_alluvial_wo_major_MBL11 <- gex_scaled_clonotypes_MBL11@meta.data %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, mt_clonotypes, BCR_clonotype) %>%
  dplyr::mutate(combination = paste0(mt_clonotypes, "_", BCR_clonotype)) %>%
  group_by(combination) %>%
  dplyr::summarise(counts = n()) %>%
  dplyr::arrange(desc(counts)) %>%
  tidyr::separate(combination, into = c("mt_clonotype", "BCR_clonotype"), sep = "_", remove = FALSE) %>%
  dplyr::filter(BCR_clonotype != "907") %>%
  rowid_to_column("id") %>%
  mutate(freq = counts / sum(counts) * 100) %>%
  dplyr::mutate(
    BCR_clonotype = paste0("BCR_clone_", BCR_clonotype)
  ) %>% 
  dplyr::mutate(BCR_clonotype=fct_reorder(BCR_clonotype, freq),
                mt_clonotype=factor(mt_clonotype, levels=rev(c(0, 5,3, 1, 2,4,6,7,8,9,10,11,12,13, 14))))

set.seed(14)
clonotype_matching_plot_wo_major_MBL11 <- ggplot(
  input_alluvial_wo_major_MBL11,
  aes(y = freq, 
      axis1 = BCR_clonotype, 
      axis2 = mt_clonotype)
) +
  geom_alluvium(aes(fill = mt_clonotype), width = 1 / 12) +
  geom_stratum(size=0.1,
    width = 1 / 12,
    alpha=0.8,
    color = "azure4"
  ) +
  scale_x_discrete(
    limits = c("BCR clones", "mt clones"),
    expand = c(.04, .04)
  ) +
  scale_y_continuous(expand = expansion(mult=0, 0))+
  scale_fill_manual(values = mt_colours) +
  labs(y = "Frequency [%]", x = "", title = "Without BCR major clone\n(n=788 cells)") +
  theme_pub() +
  theme(legend.position = "none")
```

## Clone annotation heteroplasmy heatmap 
```{r}
# Add cell type annotation to seurat clonotype object
clonotypes_MBL$MBL11 <- AddMetaData(
  object = clonotypes_MBL$MBL11,
  metadata = gex_scaled_all@meta.data %>%
    as_tibble(rownames = NA) %>%
    rownames_to_column("label_cellbarcode") %>%
    dplyr::select(label_cellbarcode, My_annotation) %>%
    dplyr::mutate(label_cellbarcode = str_replace(label_cellbarcode, pattern = "_", replacement = "#")) %>%
    dplyr::filter(label_cellbarcode %in% str_replace(colnames(clonotypes_MBL$MBL11), pattern = "_", replacement = "#")) %>%
    deframe(),
  col.name = "My_annotation"
)

# filter only cells with BCR annotation
MBL11_barcodes <- new_BCR_paired$individual %>%
  unique() %>%
  dplyr::filter(label == "MBL11") %>%
  dplyr::filter(!is.na(new_common_ct_id)) %>%
  pull(label_cellbarcode) %>%
  unique()

clonotypes_MBL$MBL11 <- subset(
  x = clonotypes_MBL$MBL11,
  cells = MBL11_barcodes
)
```

```{r}
# Adapted from https://gist.github.com/armish/564a65ab874a770e2c26
memoSort <- function(M, seurat_obj) {
  geneOrder <- sort(rowSums(M), decreasing = TRUE, index.return = TRUE)$ix

  scoreCol <- function(x) {
    score <- 0
    for (i in 1:length(x)) {
      if (x[i]) {
        score <- score + 2^(length(x) - i)
      }
    }
    return(score)
  }
  
  scores <- apply(M[geneOrder, ], 2, scoreCol)
  sampleOrder <- sort(scores, decreasing = TRUE, index.return = TRUE)$ix
  
  barcode2mt_clone <- tibble(
    "label_cellbarcode" = colnames(seurat_obj),
    "seurat_clusters" = seurat_obj@meta.data$seurat_clusters
  ) %>%
    deframe()
  
  mt_clone_frequency <- enframe(barcode2mt_clone,
                      name="label_cellbarcode",
                      value = "mt_clone") %>%
  dplyr::group_by(mt_clone) %>%
  dplyr::summarise(frequency_mt=n()) %>%
  dplyr::arrange(desc(frequency_mt)) %>%
  dplyr::mutate(mt_clone=factor(mt_clone, levels=c(0,5,3,1,2,4,6,7,8,9,10,11,12,13,14)))%>%
  pull(mt_clone)
  
  barcodes_mt_ordered <- tibble("label_cellbarcode_M"=colnames(M),
         "scores_memosort"=scores) %>% 
    left_join(enframe(barcode2mt_clone, 
                      name="label_cellbarcode", 
                      value = "mt_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(mt_clone=factor(mt_clone, levels=levels(mt_clone_frequency))) %>% 
    dplyr::group_by(mt_clone) %>% 
    dplyr::arrange(mt_clone, desc(scores_memosort)) %>% 
    pull(label_cellbarcode_M)
  
  return(list(geneOrder, barcodes_mt_ordered))
}
```

```{r fig.width=10, fig.height=7}
# plotting function
plot_heatmap_MBL11 <- function(pat, celltype, suf) {
  # subset and logtransform data
  cells <- subset(x = clonotypes_MBL[[pat]], subset = My_annotation == celltype)
  cell_matrix <- GetAssayData(object = cells, slot = "counts") %>%
    .[rowSums(.) != 0,] %>% 
    .[, as.logical(colSums(. != 0))] %>%
    .[,colSums(is.na(.))<nrow(.)] %>%
    .[c("7161G>A",
        "1966G>A",
        "1793G>A",
        "2333G>A",
        "9481T>C",
        "7368T>C",
        "13539A>G",
        "9725T>C" ,
        "8696T>C" ,
        "8563A>G",
        "2833A>G",
        "3022G>A",
        "2623A>G",
        "2916G>A",
        "2702G>A" ,
        "10675G>A",
        "6149A>C",
        "1797G>A",
        "2592G>A" ,
        "14423G>A",
        "13453C>T",
        "8737A>G"),]
    # dplyr::mutate(across(where(is.numeric), ~ ifelse(. < 10e-6 & .>0, 10e-6, .)))

  
  ## Ordering of rows and columns
  ordering <- memoSort(cell_matrix, cells)
  
      
  cell_matrix <- -log10(cell_matrix)

  #reconcatenate columns and replace 0 with NA for complex heatmap
  cell_matrix <- as.matrix(cell_matrix)
  cell_matrix[cell_matrix == Inf] <- NA
  
  # get all cellbarcodes from matrix
  cell_barcodes <- colnames(cell_matrix)
  
  
  ## Prepare mtclones
  #assign barcode zu mt clone
  mt_clones <- tibble("mt_clone"=clonotypes_MBL[[pat]]@meta.data$seurat_clusters,
       "label_cellbarcode"=colnames(clonotypes_MBL[[pat]])) %>% 
  dplyr::filter(label_cellbarcode %in% cell_barcodes) %>% 
  dplyr::mutate(mt_clone=as.numeric(as.character(mt_clone)))

  #count mt clone number for colour palette
  mt_clones_n <- mt_clones$mt_clone %>% 
    unique() %>% 
    length()
  
  mt_clones2colour <- mt_colours
  
  #named vector of barcode to mtclone
  barcode2mtclone <- mt_clones %>% 
    dplyr::select(label_cellbarcode, mt_clone) %>% 
    deframe
  
  ## Prepare BCRclones
  #assign barcode zu BCR clone
  BCR_clones <- new_BCR_paired$individual %>%
    dplyr::filter(label == pat) %>%   
    dplyr::select(label_cellbarcode,new_common_ct_id) %>% 
    dplyr::filter(label_cellbarcode %in% cell_barcodes) %>% 
    unique() %>% 
    dplyr::rename("BCR_clone"=new_common_ct_id)

  #count BCR clone number for colour palette
  BCR_clones_n <- BCR_clones$BCR_clone %>% 
    unique() %>% 
    length()
  
  #assign colour to BCR clone
  BCR_clones2colour <- tibble(
      "clonotype" = BCR_clones$BCR_clone %>% 
    unique(),
      "colour" = sample(get_palette(palette = "Set3", k = BCR_clones_n))
    ) %>%
      deframe()
  
  #named vector of barcode to BCR clone
  barcode2BCRclone <- BCR_clones %>% 
    dplyr::select(label_cellbarcode, BCR_clone) %>% 
    deframe
  
  colours <- list(
    "mt clone" = mt_clones2colour,
    "BCR clone" = BCR_colours
  )

  anno_df <- tibble(
      "mt clone" = barcode2mtclone[colnames(cell_matrix)],
      "BCR clone" = barcode2BCRclone[colnames(cell_matrix)]
    )

  column_ha <- HeatmapAnnotation(
    df = anno_df, # data frame with all clonotype label versions
    col = colours, # is list of named vector
    show_legend = FALSE,
    border = TRUE,
    height = unit(0.1, "cm")
  )

  
  # Colouring
  # edit legend manually to >6 as max cut off
  col_fun <- colorRamp2(c(0, 6), c("darkred", "yellow"))

  hm <- Heatmap(# input functions 
                matrix = cell_matrix,
                row_order=c(
                            "7161G>A",
                            "1966G>A",
                            "1793G>A",
                            "2333G>A",
                            "9481T>C",
                            "7368T>C",
                            "13539A>G",
                            "9725T>C" ,
                            "8696T>C" ,
                            "8563A>G",
                            "2833A>G",
                            "3022G>A",
                            "2623A>G",
                            "2916G>A",
                            "2702G>A" ,
                            "10675G>A",
                            "6149A>C",
                            "1797G>A",
                            "2592G>A" ,
                            "14423G>A",
                            "13453C>T",
                            "8737A>G"#,
                            # "3591G>A",
                            # "9921G>A",
                            # "13718G>A",
                            # "7138T>C"  , 
                            # "1082A>G",
                            # "709G>A"  ,
                            # "13369T>C" ,
                            # "3147G>A",
                            # "13063G>A" ,
                            # "12997G>A",
                            # "14831G>A" ,
                            # "16390G>A",
                            # "16527C>T" ,
                            # "5205T>C"   ,
                            # "3085A>G"  ,
                            # "3244G>A" ,
                            # "5970G>A" ,
                            # "9305G>A"  ,
                            # "11711G>A",
                            # "1200G>A"  ,  
                            # "11914G>A" ,
                            # "11404A>G",
                            # "204T>C"
                            ),
                column_order = ordering[[2]],
                top_annotation = column_ha,
                col = col_fun,
                na_col = "white",
                border_gp = gpar(col = "black", lty = 1),
                raster_by_magick = TRUE,
                use_raster = TRUE,
                raster_quality = 6,
                
                #column parameters
                column_title = "",
                column_title_gp = gpar(fontsize = 5),
                show_column_names = FALSE,
                
                #row parameters
                row_title = "mtDNA mutation",
                row_title_gp = gpar(fontsize = 10),
                show_row_names = FALSE,

                #legend parameters
                heatmap_legend_param = list(
                  title = "Heteroplasmy",
                  #border="black",
                  direction = "vertical",
                  legend_direction = "vertical",
                  at = c(6, 4, 2, 0),
                  labels = c("≤ 10e-6", "10e-4", "10e-2", "1"),
                  title_gp = gpar(fontsize = 10, fontface = "plain") ,
                  labels_gp = gpar(
                       fontsize = 10,
                       fontface = "plain"
                  )
                )
              )

  # needs to be a grob for heatmap to be placable in cowplot grid
  hm_grob <- grid::grid.grabExpr(draw(hm,
     heatmap_legend_side = "right"
   ))


  cairo_pdf(
    file = file.path(plotting_dir, paste0("Heteroplasmy_heatmap_matched_mt_BCR_clones_2", pat, suf, ".pdf")),
    width = 7.26,
    height = 4  )
  
  draw(hm,
    column_title = pat,
    heatmap_legend_side = "right",
    column_title_gp = grid::gpar(
      fontsize = 10,
      fontface = "plain"
    ),
    annotation_legend_side = "right"
  )
  dev.off()
  return(hm_grob)
}

ordered_MBL11_heatmap <- plot_heatmap_MBL11(pat = "MBL11", 
                                            celltype = "MBL B Cell", 
                                            suf = "")

saveRDS(ordered_MBL11_heatmap, file.path(RDS_dir, "ordered_MBL11_heatmap.RDS"))

cowplot::plot_grid(ordered_MBL11_heatmap,
  nrow = 1,
  ncol = 1
)
```

```{r, fig.height=7, fig.width=9}
cowplot::plot_grid(
  cowplot::plot_grid(
    mt_clontoype_umap_MBL11,
    BCR_clonotype_umap_MBL11,NULL,
    nrow = 1
  ),
  cowplot::plot_grid(clonotype_matching_plot_MBL11,
    ordered_MBL11_heatmap,
    nrow = 1, rel_widths = c(1,  2)
  ),
  nrow = 2,
  ncol = 1
) %>%
  ggsave(.,
    file = file.path(plotting_dir, "Compare_clones_Figure.png"),
    width = 11,
    height = 7,
    units = "in",
    bg = "white",
    dpi = 600
  )
```

Pie charts
```{r}
input_alluvial_MBL11 %>%
  dplyr::mutate(malignance=if_else(condition=BCR_clonotype =="BCR_clone_907", true="malignant", false="non-malignant")) %>% 
  dplyr::group_by(malignance, mt_clonotype) %>% 
  summarise(cell_counts=sum(counts)) %>% 
  split(~malignance) %>% imap(~
ggplot(., aes(x="", y=cell_counts, fill=mt_clonotype)) +
  geom_bar(stat="identity", width=0.01) +
  coord_polar("y", start=0)+
  labs(title=.y)+
  scale_fill_manual(values = mt_colours)+
  theme_pub()+
  theme(axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "none",
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        )) %>% 
  ggarrange(plotlist = .) %>% 
  ggsave(., file=file.path(plotting_dir, "MBL11_mt_clones_piechart.pdf"))
```

```{r}
input_alluvial_MBL11 %>%
  dplyr::mutate(malignance=if_else(condition=BCR_clonotype =="BCR_clone_907", true="malignant", false="non-malignant")) %>% 
  dplyr::group_by(malignance, BCR_clonotype) %>% 
  summarise(cell_counts=sum(counts)) %>% 
  dplyr::mutate(BCR_clonotype=str_split_i(BCR_clonotype, pattern="_", i=3 )) %>% 
  split(~malignance) %>% 
  imap(~
ggplot(., aes(x="", y=cell_counts, fill=BCR_clonotype)) +
  geom_bar(stat="identity", width=0.01, colour="darkgrey") +
  coord_polar("y", start=0)+
  labs(title=.y)+
  scale_fill_manual(values = BCR_colours)+
  theme_pub()+
  theme(axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "none",
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        )) %>% 
  ggarrange(plotlist = .) %>% 
  ggsave(., file=file.path(plotting_dir, "MBL11_BCR_clones_piechart.pdf"))
```

## Coverage mtDNA variants MBL11
```{r}
# find out which mutations contribute to clonotypes in MBL11
MBL11_mt_variants <-c(
  "7161G>A",
  "1966G>A",
  "1793G>A",
  "2333G>A",
  "9481T>C",
  "7368T>C",
  "13539A>G",
  "9725T>C",
  "8696T>C",
  "8563A>G",
  "2833A>G",
  "3022G>A",
  "2623A>G",
  "2916G>A",
  "2702G>A",
  "10675G>A",
  "6149A>C",
  "1797G>A",
  "2592G>A",
  "14423G>A",
  "13453C>T",
  "8737A>G"
)

mtmut_positions_MBL11 <- bind_rows(ATAC_VAFs_cln_MBL %>%
  map(~ as.matrix(.) %>%
    as_tibble(rownames = "variant_id") %>%
    .[, as.logical(colSums(. != 0))]), .id = "patient_id") %>% 
  dplyr::filter(patient_id =="MBL11") %>% 
  left_join(metadata_all %>% 
              dplyr::filter(data_type=="GEX") %>%
              dplyr::select(plotting_id, patient_id) %>% unique) %>% 
  dplyr::select(patient_id, variant_id) %>% 
  tidyr::separate(variant_id, into = c("position", NA), 
                  sep = "(A|C|G|T)", 
                  remove = FALSE) %>% 
  dplyr::filter(variant_id %in% MBL11_mt_variants)

mtmut_positions_label_MBL11 <- mtmut_positions_MBL11 %>% 
  dplyr::filter(str_starts(patient_id, "CLL")) %>% 
  dplyr::mutate(label=paste0(patient_id, ".2")) %>% 
  bind_rows(mtmut_positions_MBL11 %>% 
              dplyr::mutate(label=if_else(str_starts(patient_id, "CLL"), paste0(patient_id, ".1"), patient_id))) %>% 
  arrange(label) %>% 
  dplyr::mutate(position=as.numeric(position)) 

coverage_MBL11 <- file.path(library_path, paste0(names(mtDNA_samples[27]), ".mgatk"), paste0(names(mtDNA_samples[27]), ".rds")) %>% 
  map( ~data.frame(
  position = zoo::rollmean(1:16569, 5),
  coverage = pull_coverage(readRDS(.x))) %>% 
  dplyr::left_join(mtmut_positions_label_MBL11, by = "position")
  )
```

```{r}
coverage_MBL11_plot <- coverage_MBL11[[1]] %>% 
  ggplot(aes(x = position, y = coverage, label=variant_id)) +
  geom_line(colour="thistle4") +
  geom_label_repel(label.size = NA, 
                   fill=NA, 
                   nudge_y = 500,
                   max.overlaps = 30,
                   colour = "darkred")+
  scale_y_continuous(expand = expansion(mult=c(0, 0.05)))+
  scale_x_continuous(expand = expansion(mult=c(0, 0.01)))+
  labs(x = "mtDNA position [bp]", 
       y = "Rolling mean coverage across cells",
       title="P3\n(LC-MBL)")+
  theme_pub()+
  theme(legend.position = "none")

write_tsv(coverage_MBL11_plot$data, file.path(tsv_dir, "mtDNA_coverage_GEX_MBL11.tsv"))
```