---
title: "Differential Chromatin Accessibility Analysis Full Cohort but separated by batch"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

GOAL: The aim of this script is to find out if there are recurrent differentially accessible chromatin regions between MBL and CLL samples across patients regardless to their sequencing batches . 

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(ArchR)
library(readxl)
library(ggrepel)
library(tidyverse)
library(glue)
library(ggpubr)
library(SeuratDisk)
library(Seurat)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Cairo)
library(sva)
library(scales)
library(chromVAR)
library(patchwork)
library(circlize)
library(ComplexHeatmap)
library(magick)
register(MulticoreParam(8))
```

```{r Run_parameters}
set.seed(1)
addArchRThreads(threads = 1)
addArchRGenome("hg38")
```

## Set paths

```{r }
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "DCA_analysis")

plotting_dir <- file.path(analysis_dir, "Plots")
RDS_dir <- file.path(analysis_dir, "RDS_objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")

ArchR_project_Blike <- "ArchR_35_samples/ArchR_Blike"
ArchR_dir_Blike <- file.path(result_dir, ArchR_project_Blike)
plotting_dir_Blike <- file.path(ArchR_dir_Blike, "Plots")
RDS_dir_Blike <- file.path(ArchR_dir_Blike, "RDS_objects")
```

```{r}
# local
source(file.path(project_dir, "scripts/publication_theme.R"))
source(file.path(project_dir, "scripts/heatmap_plotting_functions.R"))
```

## Metadata

```{r }
metadata <- read_tsv(file = file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata.tsv"))

patient_id2plotting_id <- metadata %>%
  dplyr::select(patient_id, plotting_id) %>%
  unique() %>%
  deframe()

label2plotting_id <- metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, plotting_id) %>%
  deframe()
```

## Load ArchRproj to subset from
```{r}
CLLProj_Blike <- loadArchRProject(ArchR_dir_Blike)
```

## Subset ArchR project by batch
```{r}
# Add sequencing date annotation
seqdate_annotation <- tibble("label_cellbarcode" = CLLProj_Blike$cellNames) %>%
  tidyr::separate(label_cellbarcode, into = c("label", "cellbarcode"), sep = "#", remove = F) %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "ATAC") %>% dplyr::select(label, sequencing_date)) %>%
  dplyr::mutate(sequencing_date = as.numeric(as.factor(sequencing_date))) %>%
  dplyr::select(label_cellbarcode, sequencing_date) %>%
  deframe()

# CLLProj_Blike <- addCellColData(
#   ArchRProj = CLLProj_Blike,
#   data = seqdate_annotation,
#   cells = names(seqdate_annotation),
#   name = "sequencing_date"
# )

idxBatch_Blike <- c(1:4) %>%
  map(~ BiocGenerics::which(CLLProj_Blike@cellColData$sequencing_date == .x)) %>%
  set_names(1:4)

cellsSample_Blike <- idxBatch_Blike %>%
  map(~ CLLProj_Blike$cellNames[.x])

batch_output_directories <- file.path(project_dir_ext, paste0("ArchR_Blike_seqbatch", names(cellsSample_Blike)))

# CLLProj_Blike_batches <- cellsSample_Blike %>%
#   map2(batch_output_directories, ~subsetArchRProject(
#   ArchRProj = CLLProj_Blike,
#   cells = .x,
#   outputDirectory = .y,
#   dropCells = TRUE))

CLLProj_Blike_batches <- map(batch_output_directories, ~ loadArchRProject(.x))
```

### Add pseudobulk replicates

```{r}
# CLLProj_Blike_batches <- CLLProj_Blike_batches %>%
#   map(~ addGroupCoverages(
#     ArchRProj = .x,
#     groupBy = "B_like_annotation"
#   ))
```

### Peak calling

```{r}
CLLProj_Blike_batches <- CLLProj_Blike_batches %>%
  map(~ addReproduciblePeakSet(
    ArchRProj = .x,
    groupBy = "B_like_annotation",
    pathToMacs2 = pathToMacs3,
  ))

CLLProj_Blike_batches %>%
  map2(
    batch_output_directories,
    ~ saveArchRProject(
      ArchRProj = .x,
      outputDirectory = .y,
      load = FALSE,
      overwrite = TRUE
    )
  )
```

```{r}
CLLProj_Blike_batches <- CLLProj_Blike_batches %>%
  map(~ addPeakMatrix(.x))

CLLProj_Blike_batches %>%
  map2(
    batch_output_directories,
    ~ saveArchRProject(
      ArchRProj = .x,
      outputDirectory = .y,
      load = FALSE,
      overwrite = TRUE
    )
  )

CLLProj_Blike_batches %>%
  map(~ getAvailableMatrices(.))
```

### Marker Peaks

```{r}
# Gets peaks which are unique to cell types and account for sequencing date bias
markersPeaks_Blike_batches <- CLLProj_Blike_batches %>%
  map(~ getMarkerFeatures(
    ArchRProj = .x,
    useMatrix = "PeakMatrix",
    groupBy = "B_like_annotation",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    testMethod = "wilcoxon"  ))

markerList <- markersPeaks_Blike_batches %>%
  map(~ getMarkers(.x, cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1"))
```

### Volcano Plots MarkerPeaks

```{r, fig.width=20,fig.height=20}
# unique peaks for MBL or CLL cells compared to all other cell types
volcanoplot_MP_B <- markersPeaks_Blike_batches %>%
  imap(~ plotMarkers(
    seMarker = .x,
    name = "B",
    cutOff = "FDR <= 0.05 & Log2FC >= 1",
    plotAs = "Volcano"
  ) +
    geom_point(size = 1) +
    scale_colour_manual(
      name = "DAR",
      values = c("Down-Regulated" = "gray2", "Up-Regulated" = "darkred", "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "LC-MBL and CLL cells",
        "Up-Regulated" = "Physiologic B cells",
        "Not-Differential" = "not significant"
      )
    ) +
    theme_pub() %>%
    set_names(.y))

volcanoplot_MP_LC_MBL <-
  plotMarkers(
    seMarker = markersPeaks_Blike_batches[[4]],
    name = "LC_MBL_B_Cells",
    cutOff = "FDR <= 0.05 & Log2FC >= 1",
    plotAs = "Volcano"
  ) +
  geom_point(size = 1) +
  scale_colour_manual(
    name = "DAR",
    values = c("Down-Regulated" = "gray2", "Up-Regulated" = "#fcbda4", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "Physiologic and CLL cells",
      "Up-Regulated" = "LC-MBL cells",
      "Not-Differential" = "not significant"
    )
  ) +
  theme_pub()

volcanoplot_MP_HC_MBL <- markersPeaks_Blike_batches[1:3] %>%
  map(~ plotMarkers(
    seMarker = .x,
    name = "HC_MBL_B_Cells",
    cutOff = "FDR <= 0.05 & Log2FC >= 1",
    plotAs = "Volcano"
  ) +
    geom_point(size = 3) +
    scale_colour_manual(
      name = "DAR",
      values = c("Down-Regulated" = "gray2", "Up-Regulated" = "#FC8D62", "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "Physiologic and CLL cells",
        "Up-Regulated" = "HC-MBL cells",
        "Not-Differential" = "not significant"
      )
    ) +
    theme_pub())

volcanoplot_MP_CLL <- markersPeaks_Blike_batches[1:3] %>%
  map(~ plotMarkers(
    seMarker = .x,
    name = "CLL_B_Cells",
    cutOff = "FDR <= 0.05 & Log2FC >= 1",
    plotAs = "Volcano"
  ) +
    geom_point(size = 1) +
    scale_colour_manual(
      name = "DAR",
      values = c("Down-Regulated" = "gray2", "Up-Regulated" = "#66C2A5", "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "Physiologic B and LC-MBL cells",
        "Up-Regulated" = "CLL cells",
        "Not-Differential" = "not significant"
      )
    ) +
    theme_pub())

ggarrange(ggarrange(plotlist = volcanoplot_MP_B, ncol = 4, nrow = 1, labels = paste0("Batch", c(1:4)), common.legend = TRUE, legend = "bottom"),
  ggarrange(NULL, NULL, NULL, volcanoplot_MP_LC_MBL, ncol = 4, nrow = 1, labels = paste0("Batch", c(4)), common.legend = TRUE, legend = "bottom"),
  ggarrange(plotlist = volcanoplot_MP_HC_MBL, ncol = 4, nrow = 1, labels = paste0("Batch", c(1:3)), common.legend = TRUE, legend = "bottom"),
  ggarrange(plotlist = volcanoplot_MP_CLL, ncol = 4, nrow = 1, labels = paste0("Batch", c(1:3)), common.legend = TRUE, legend = "bottom"),
  nrow = 4, ncol = 1
) %>%
  ggsave(
    plot = .,
    filename = file.path(plotting_dir, "Batch_correction/Unique_Markers_Blike_Volcanoplot_batches.pdf"),
    width = 20,
    height = 20,
    unit = "in"
  )
```

### Pairwise peaks

```{r }
markerTest_HC_MBL_CLL_batches <- CLLProj_Blike_batches[1:3] %>%
  imap(~ getMarkerFeatures(
    ArchRProj = .x,
    useMatrix = "PeakMatrix",
    groupBy = "B_like_annotation",
    testMethod = "wilcoxon",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    useGroups = "CLL_B_Cells",
    bgdGroups = "HC_MBL_B_Cells"
  ))

# unique peaks between LC-MBL and B cells only!
markerTest_B_LC_MBL_batches <- CLLProj_Blike_batches[4:4] %>%
  imap(~ getMarkerFeatures(
    ArchRProj = .x,
    useMatrix = "PeakMatrix",
    groupBy = "B_like_annotation",
    testMethod = "wilcoxon",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    useGroups = "LC_MBL_B_Cells",
    bgdGroups = "B"
  ))

markerTest_B_HC_MBL_batches <- CLLProj_Blike_batches[1:3] %>%
  imap(~ getMarkerFeatures(
    ArchRProj = .x,
    useMatrix = "PeakMatrix",
    groupBy = "B_like_annotation",
    testMethod = "wilcoxon",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    useGroups = "HC_MBL_B_Cells",
    bgdGroups = "B"
  ))

markerTest_B_CLL_batches <- CLLProj_Blike_batches[1:3] %>%
  imap(~ getMarkerFeatures(
    ArchRProj = .x,
    useMatrix = "PeakMatrix",
    groupBy = "B_like_annotation",
    testMethod = "wilcoxon",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    useGroups = "CLL_B_Cells",
    bgdGroups = "B"
  ))
```

### Volcano plots Pairwise testing

```{r fig.width=20, fig.height=20}
volcanoplot_PWT_HC_MBL_CLL <- markerTest_HC_MBL_CLL_batches[1:3] %>%
  imap(~ plotMarkers(
    seMarker = .x,
    name = "CLL_B_Cells",
    cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
    plotAs = "Volcano",
    rastr = TRUE
  ) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
    scale_y_continuous(
      name = "-Log10(FDR)",
      breaks = seq(0, 30, 10),
      limits = c(0, 30),
      # expand = expansion(mult = c(0, 0.01)),
      oob = scales::oob_squish_infinite
    ) +
    scale_colour_manual(
      name = "Peak accessible in",
      values = c("Down-Regulated" = "#FC8D62", "Up-Regulated" = "#66C2A5", "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "HC-MBL cells",
        "Up-Regulated" = "CLL cells",
        "Not-Differential" = "Not significant"
      )
    ) +
    theme_pub())

volcanoplot_PWT_B_LC_MBL <- markerTest_B_LC_MBL_batches[1:1] %>%
  imap(~ plotMarkers(
    seMarker = .x,
    name = "LC_MBL_B_Cells",
    cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
    plotAs = "Volcano",
    rastr = TRUE
  ) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
    scale_y_continuous(
      name = "-Log10(FDR)",
      breaks = seq(0, 30, 10),
      limits = c(0, 30),
      oob = scales::oob_squish_infinite
    ) +
    scale_colour_manual(
      name = "Peak accessible in",
      values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#fcbda4", "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "Physiologic B cells",
        "Up-Regulated" = "LC-MBL cells",
        "Not-Differential" = "Not significant"
      )
    ) +
    theme_pub())

volcanoplot_PWT_B_HC_MBL <- markerTest_B_HC_MBL_batches[1:3] %>%
  imap(~ plotMarkers(
    seMarker = .x,
    name = "HC_MBL_B_Cells",
    cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
    plotAs = "Volcano",
    rastr = TRUE
  ) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
    scale_y_continuous(
      name = "-Log10(FDR)",
      breaks = seq(0, 30, 10),
      limits = c(0, 30),
      oob = scales::oob_squish_infinite
    ) +
    scale_colour_manual(
      name = "Peak accessible in",
      values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#FC8D62", "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "Physiologic B cells",
        "Up-Regulated" = "HC-MBL cells",
        "Not-Differential" = "Not significant"
      )
    ) +
    theme_pub())

volcanoplot_PWT_B_CLL <- markerTest_B_CLL_batches[1:3] %>%
  imap(~ plotMarkers(
    seMarker = .x,
    name = "CLL_B_Cells",
    cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
    plotAs = "Volcano",
    rastr = TRUE
  ) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
    scale_y_continuous(
      name = "-Log10(FDR)",
      breaks = seq(0, 30, 10),
      limits = c(0, 30),
      oob = scales::oob_squish_infinite
    ) +
    scale_colour_manual(
      name = "Peak accessible in",
      values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#66C2A5", "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "Physiologic B cells",
        "Up-Regulated" = "CLL cells",
        "Not-Differential" = "Not significant"
      )
    ) +
    theme_pub())

ggarrange(ggarrange(plotlist = c(volcanoplot_PWT_B_LC_MBL), nrow = 1, ncol = 4, labels = paste0("Batch", 4), common.legend = TRUE, legend = "bottom"),
  ggarrange(plotlist = volcanoplot_PWT_B_HC_MBL, ncol = 4, nrow = 1, labels = paste0("Batch", 1:4), common.legend = TRUE, legend = "bottom"),
  ggarrange(plotlist = volcanoplot_PWT_B_CLL, ncol = 4, nrow = 1, labels = paste0("Batch", 1:4), common.legend = TRUE, legend = "bottom"),
  nrow = 3, ncol = 1
)

ggarrange(plotlist = c(volcanoplot_PWT_LC_MBL_HC_MBL, volcanoplot_PWT_LC_MBL_CLL, volcanoplot_PWT_HC_MBL_CLL), ncol = 4, nrow = 3, labels = rep(paste0("Batch", 1:4), 4))

ggarrange(volcanoplot_PWT_B_LC_MBL, volcanoplot_PWT_B_HC_MBL, volcanoplot_PWT_B_CLL, nrow=1) %>%
  ggsave(file=file.path(plotting_dir_Blike, "Peaks_volcanoplots.png"),
units = "in",
width=12 ,
height =4 )
```

```{r }
ggsave(
  plot = ggarrange(plotlist = c(volcanoplot_PWT_B_LC_MBL), nrow = 1, ncol = 1, labels = paste0("Batch", 4), common.legend = TRUE, legend = "bottom"),
  filename = file.path(plotting_dir, "Batch_correction/Upregulation_LC_MBL_Volcanoplot_batches.pdf"),
  width = 5,
  height = 5,
  unit = "in"
)

ggsave(
  plot = ggarrange(plotlist = volcanoplot_PWT_B_HC_MBL, ncol = 3, nrow = 1, labels = paste0("Batch", 1:3), common.legend = TRUE, legend = "bottom"),
  filename = file.path(plotting_dir, "Batch_correction/Upregulation_HC_MBL_Volcanoplot_batches.pdf"),
  width = 15,
  height = 5,
  unit = "in"
)

ggsave(
  plot = ggarrange(plotlist = volcanoplot_PWT_B_CLL, ncol = 3, nrow = 1, labels = paste0("Batch", 1:3), common.legend = TRUE, legend = "bottom"),
  filename = file.path(plotting_dir, "Batch_correction/Upregulation_CLL_Volcanoplot_batches.pdf"),
  width = 15,
  height = 5,
  unit = "in"
)

saveRDS(volcanoplot_PWT_B_LC_MBL, file.path(RDS_dir, "volcanoplot_PWT_B_LC_MBL.RDS"))
saveRDS(volcanoplot_PWT_B_HC_MBL, file.path(RDS_dir, "volcanoplot_PWT_B_HC_MBL.RDS"))
saveRDS(volcanoplot_PWT_B_CLL, file.path(RDS_dir, "volcanoplot_PWT_B_CLL.RDS"))
```

### Summary volcanoplots
```{r}
peak_summary <- list()
peak_summary[["HC_MBL_CLL"]] <- markerTest_HC_MBL_CLL_batches[1:3] %>%
  purrr::imap(.f = function(x, y) {
    bind_cols(
      assays(x)[["Log2FC"]],
      assays(x)[["FDR"]]
    ) %>%
      dplyr::rename(
        "Log2FC" = `CLL_B_Cells...1`,
        "FDR" = `CLL_B_Cells...2`
      ) %>%
      bind_cols(rowData(x) %>%
        as_tibble() %>%
        tidyr::unite(col = "peak", c(seqnames, start, end), sep = "_") %>%
        dplyr::select(peak))
  }) %>%
  bind_rows(.id = "batch") %>%
  as_tibble()

peak_summary[["B_LC_MBL"]] <- markerTest_B_LC_MBL_batches[1:1] %>%
  purrr::imap(.f = function(x, y) {
    bind_cols(
      assays(x)[["Log2FC"]],
      assays(x)[["FDR"]]
    ) %>%
      dplyr::rename(
        "Log2FC" = `LC_MBL_B_Cells...1`,
        "FDR" = `LC_MBL_B_Cells...2`
      ) %>%
      bind_cols(rowData(x) %>%
        as_tibble() %>%
        tidyr::unite(col = "peak", c(seqnames, start, end), sep = "_") %>%
        dplyr::select(peak))
  }) %>%
  bind_rows(.id = "batch") %>%
  as_tibble()

peak_summary[["B_HC_MBL"]] <- markerTest_B_HC_MBL_batches[1:3] %>%
  purrr::imap(.f = function(x, y) {
    bind_cols(
      assays(x)[["Log2FC"]],
      assays(x)[["FDR"]]
    ) %>%
      dplyr::rename(
        "Log2FC" = `HC_MBL_B_Cells...1`,
        "FDR" = `HC_MBL_B_Cells...2`
      ) %>%
      bind_cols(rowData(x) %>%
        as_tibble() %>%
        tidyr::unite(col = "peak", c(seqnames, start, end), sep = "_") %>%
        dplyr::select(peak))
  }) %>%
  bind_rows(.id = "batch") %>%
  as_tibble()

peak_summary[["B_CLL"]] <- markerTest_B_CLL_batches[1:3] %>%
  purrr::imap(.f = function(x, y) {
    bind_cols(
      assays(x)[["Log2FC"]],
      assays(x)[["FDR"]]
    ) %>%
      dplyr::rename(
        "Log2FC" = `CLL_B_Cells...1`,
        "FDR" = `CLL_B_Cells...2`
      ) %>%
      bind_cols(rowData(x) %>%
        as_tibble() %>%
        tidyr::unite(col = "peak", c(seqnames, start, end), sep = "_") %>%
        dplyr::select(peak))
  }) %>%
  bind_rows(.id = "batch") %>%
  as_tibble()


summary_volcano_HC_MBL_CLL <- peak_summary[["HC_MBL_CLL"]] %>%
  dplyr::mutate("significance" = case_when(FDR <= 0.05 & Log2FC >= 1 ~ "CLL",
    FDR <= 0.05 & Log2FC <= 1 ~ "HC-MBL",
    .default = NA
  )) %>%
  ggplot(aes(x = Log2FC, y = -log10(FDR), shape = batch, colour = significance)) +
  geom_point(size = 1) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  scale_x_continuous(
    name = "Log2FC",
    breaks = seq(-8, 8, 4),
    limits = c(-8, 8)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 35),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak enrichment",
    values = c("CLL" = "#66C2A5", "HC-MBL" = "#FC8D62"), na.value = "lightgrey"
  ) +
  scale_shape_manual(name = "Batch", values = c("1" = 15, "2" = 17, "3" = 18, "4" = 16)) +
  labs(title = paste0(length(unique(peak_summary[["HC_MBL_CLL"]]$peak)), " peaks")) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

summary_volcano_B_LC_MBL <- peak_summary[["B_LC_MBL"]] %>%
  dplyr::mutate(
    "batch" = "4",
    "significance" = case_when(FDR <= 0.05 & Log2FC >= 1 ~ "LC-MBL",
      FDR <= 0.05 & Log2FC <= 1 ~ "B",
      .default = NA
    )
  ) %>%
  ggplot(aes(x = Log2FC, y = -log10(FDR), shape = batch, colour = significance)) +
  geom_point(size = 1) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  scale_x_continuous(
    name = "Log2FC",
    breaks = seq(-8, 8, 4),
    limits = c(-8, 8)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 35),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak enrichment",
    values = c("B" = "darkred", "LC-MBL" = "#fcbda4"), na.value = "lightgrey"
  ) +
  scale_shape_manual(name = "Batch", values = c("1" = 15, "2" = 17, "3" = 18, "4" = 16)) +
  labs(title = paste0(length(unique(peak_summary[["B_LC_MBL"]]$peak)), " peaks")) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )


summary_volcano_B_HC_MBL <- peak_summary[["B_HC_MBL"]] %>%
  dplyr::mutate("significance" = case_when(FDR <= 0.05 & Log2FC >= 1 ~ "HC-MBL",
    FDR <= 0.05 & Log2FC <= 1 ~ "B",
    .default = NA
  )) %>%
  ggplot(aes(x = Log2FC, y = -log10(FDR), shape = batch, colour = significance)) +
  geom_point(size = 1) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  scale_x_continuous(
    name = "Log2FC",
    breaks = seq(-8, 8, 4),
    limits = c(-8, 8)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 35),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak enrichment",
    values = c("B" = "darkred", "HC-MBL" = "#FC8D62"), na.value = "lightgrey"
  ) +
  scale_shape_manual(name = "Batch", values = c("1" = 15, "2" = 17, "3" = 18, "4" = 16)) +
  labs(title = paste0(length(unique(peak_summary[["B_HC_MBL"]]$peak)), " peaks")) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

summary_volcano_B_CLL <- peak_summary[["B_CLL"]] %>%
  dplyr::mutate("significance" = case_when(FDR <= 0.05 & Log2FC >= 1 ~ "CLL",
    FDR <= 0.05 & Log2FC <= 1 ~ "B",
    .default = NA
  )) %>%
  ggplot(aes(x = Log2FC, y = -log10(FDR), shape = batch, colour = significance)) +
  geom_point(size = 1) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  scale_x_continuous(
    name = "Log2FC",
    breaks = seq(-8, 8, 4),
    limits = c(-8, 8)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 35),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak enrichment",
    values = c("B" = "darkred", "CLL" = "#66C2A5"), na.value = "lightgrey"
  ) +
  scale_shape_manual(name = "Batch", values = c("1" = 15, "2" = 17, "3" = 18, "4" = 16)) +
  labs(title = paste0(length(unique(peak_summary[["B_CLL"]]$peak)), " peaks")) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggarrange(summary_volcano_B_LC_MBL + theme(legend.position = "none"), summary_volcano_B_HC_MBL + theme(legend.position = "none"), summary_volcano_B_CLL + theme(legend.position = "none"), nrow = 1, ncol = 3) %>%
  ggsave(
    file = file.path(plotting_dir, "Batch_peak_summary.pdf"),
    width = 7.33,
    height = 1.71,
    units = "in"
  )
```


```{r}
CLLProj_Blike_batches %>%
  map2(
    batch_output_directories,
    ~ saveArchRProject(
      ArchRProj = .x,
      outputDirectory = .y,
      load = FALSE,
      overwrite = TRUE
    )
  )
```

### Add TF Mofif Annotations

```{r}
CLLProj_Blike_batches <- CLLProj_Blike_batches %>%
  map(~ addMotifAnnotations(
    ArchRProj = .x,
    motifSet = "cisbp",
    annoName = "Motif"
  ))
```

```{r}
# background peaks added
CLLProj_Blike_batches <- CLLProj_Blike_batches %>%
  map(~ addBgdPeaks(.x, method = "chromVAR"))
```

```{r}
# motif deviation matrix added
CLLProj_Blike_batches <- CLLProj_Blike_batches %>%
  map(~ addDeviationsMatrix(
    ArchRProj = .x,
    bgdPeaks = getBgdPeaks(.x,
      method = "chromVAR"
    ),
    peakAnnotation = "Motif",
    threads = 1,
    force = TRUE
  ))
```

```{r}
plotVarDev <- CLLProj_Blike_batches %>%
  imap(~ getVarDeviations(.x,
    name = "MotifMatrix",
    plot = TRUE
  )) %>%
  set_names(.y)
```

### Get Matrix from project

```{r}
CLLProj_Blike_batches <- list(
  "1" = loadArchRProject(file.path(project_dir_ext, "ArchR_Blike_seqbatch1/")),
  "2" = loadArchRProject(file.path(project_dir_ext, "ArchR_Blike_seqbatch2/")),
  "3" = loadArchRProject(file.path(project_dir_ext, "ArchR_Blike_seqbatch3/")),
  "4" = loadArchRProject(file.path(project_dir_ext, "ArchR_Blike_seqbatch4/"))
)

MotifMatrix_tmp_Blike_batches <- CLLProj_Blike_batches %>%
  map(~ getMatrixFromProject(
    ArchRProj = .x,
    useMatrix = "MotifMatrix",
    useSeqnames = NULL,
    verbose = TRUE,
    binarize = FALSE,
    threads = 1
  ))

zscore_matrix_Blike <- MotifMatrix_tmp_Blike_batches %>%
  map(~ as.matrix(assays(.x)$z) %>%
    as.data.frame())
```

```{r}
condition_Blike <- CLLProj_Blike_batches %>%
  map(~ tibble(
    "label_cellbarcode" = .x$cellNames,
    "B_like_annotation" = .x$B_like_annotation
  ))

B_cells_Blike <- condition_Blike %>%
  map(~ dplyr::filter(.x, B_like_annotation == "B") %>%
    pull(label_cellbarcode))

LC_MBL_cells_Blike <- condition_Blike %>%
  map(~ dplyr::filter(.x, B_like_annotation == "LC_MBL_B_Cells") %>%
    pull(label_cellbarcode))

HC_MBL_cells_Blike <- condition_Blike %>%
  map(~ dplyr::filter(.x, B_like_annotation == "HC_MBL_B_Cells") %>%
    pull(label_cellbarcode))

CLL_cells_Blike <- condition_Blike %>%
  map(~ dplyr::filter(.x, B_like_annotation == "CLL_B_Cells") %>%
    pull(label_cellbarcode))

# # calculate data for volcano plot
input_zscore_matrix_Blike <- zscore_matrix_Blike

# If no NA values are found in zcore matrix calculate statistics
condition_list <- pmap(.l = list(
  ..1 = B_cells_Blike,
  ..2 = LC_MBL_cells_Blike,
  ..3 = HC_MBL_cells_Blike,
  ..4 = CLL_cells_Blike
), .f = ~ list(
  "B_vs_LC_MBL" = list(..2, ..1),
  "B_vs_HC_MBL" = list(..3, ..1),
  "B_vs_CLL" = list(..4, ..1),
  "HC_MBL_vs_CLL" = list(..4, ..3)
))

# filter condition list manually for possible comparisons
condition_list[[1]]$B_vs_LC_MBL <- NULL
condition_list[[2]]$B_vs_LC_MBL <- NULL
condition_list[[3]]$B_vs_LC_MBL <- NULL
condition_list[[4]]$B_vs_HC_MBL <- NULL
condition_list[[4]]$B_vs_CLL <- NULL
condition_list[[4]]$HC_MBL_vs_CLL <- NULL


volcano.data.all.comparisons <- list(
  "B_vs_LC_MBL" = list(data.frame(), data.frame(), data.frame(), data.frame()),
  "B_vs_HC_MBL" = list(data.frame(), data.frame(), data.frame(), data.frame()),
  "B_vs_CLL" = list(data.frame(), data.frame(), data.frame(), data.frame()),
  "HC_MBL_vs_CLL" = list(data.frame(), data.frame(), data.frame(), data.frame())
)

batch <- 0
for (batch in 1:4) {
  print(paste0("Current batch ", batch))
  j <- 0
  for (j in 1:length(condition_list[[batch]])) {
    print(paste0("Current comparison: ", names(condition_list[[batch]][j])))
    comparison <- names(condition_list[[batch]][j])
    i <- 0
    for (TF in rownames(input_zscore_matrix_Blike[[batch]])) {
      i <- i + 1
      print(i)
      print(TF)
      if (length(which(is.na(input_zscore_matrix_Blike[[batch]][TF, ]))) == 0) {
        p.value <- wilcox.test(as.numeric(input_zscore_matrix_Blike[[batch]][TF, condition_list[[batch]][[j]][[1]]]), as.numeric(input_zscore_matrix_Blike[[batch]][TF, condition_list[[batch]][[j]][[2]]]))$p.value
        diff.accessibility <- median(as.numeric(input_zscore_matrix_Blike[[batch]][TF, condition_list[[batch]][[j]][[1]]])) - median(as.numeric(input_zscore_matrix_Blike[[batch]][TF, condition_list[[batch]][[j]][[2]]]))
        volcano.data.all.comparisons[[comparison]][[batch]] <- rbind(volcano.data.all.comparisons[[comparison]][[batch]], data.frame(TF = TF, p.value = p.value, diff.accessibility = diff.accessibility))
      }
    }
    volcano.data.all.comparisons[[comparison]][[batch]]$p.adj <- p.adjust(volcano.data.all.comparisons[[comparison]][[batch]]$p.value)
    volcano.data.all.comparisons[[comparison]][[batch]]$p.adj[which(volcano.data.all.comparisons[[comparison]][[batch]]$p.adj == 0)] <- min(volcano.data.all.comparisons[[comparison]][[batch]]$p.adj[which(volcano.data.all.comparisons[[comparison]][[batch]]$p.adj != 0)])
    volcano.data.all.comparisons[[comparison]][[batch]]$p.adj <- as.numeric(volcano.data.all.comparisons[[comparison]][[batch]]$p.adj)
    volcano.data.all.comparisons[[comparison]][[batch]]$diff.accessibility <- as.numeric(volcano.data.all.comparisons[[comparison]][[batch]]$diff.accessibility)
    volcano.data.all.comparisons[[comparison]][[batch]]$significant <- ifelse(abs(volcano.data.all.comparisons[[comparison]][[batch]]$diff.accessibility) > 1.5 & volcano.data.all.comparisons[[comparison]][[batch]]$p.adj <= 0.05,
      "significant",
      "not significant"
    )
    volcano.data.all.comparisons[[comparison]][[batch]] <- volcano.data.all.comparisons[[comparison]][[batch]] %>%
      dplyr::mutate(
        "significant" = case_when(
          p.adj >= 0.05 ~ "not significant",
          abs(diff.accessibility) < 1.5 ~ "not significant",
          diff.accessibility > 1.5 & p.adj < 0.05 ~ "upregulated",
          diff.accessibility < -1.5 & p.adj < 0.05 ~ "downregulated"
        ),
        "label" = if_else(significant != "not significant", stringr::str_split_fixed(TF, pattern = "_", n = 2)[, 1], NA)
      )
  }
}
saveRDS(volcano.data.all.comparisons, file = file.path(RDS_dir, "volcano.data.all.comparisons_batches.RDS"))
```

### Volcanoplots Differential TF enrichment

```{r, fig.width=15, fig.height=15}
volcano.plot_B_LC_MBL <- volcano.data.all.comparisons$B_vs_LC_MBL[[4]] %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-8, 8, 4),
    limits = c(-8, 8)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#fcbda4", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "LC-MBL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

volcano.plot_B_HC_MBL <- volcano.data.all.comparisons$B_vs_HC_MBL[1:3] %>%
  map(~ ggplot(., aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
    geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
    geom_point(size = 1) +
    geom_label_repel(
      size = 2.5,
      label.size = NA,
      label.padding = 0.1,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = 40,
      colour = "black"
    ) +
    scale_x_continuous(
      name = "Difference median z-score",
      breaks = seq(-4, 4, 2),
      limits = c(-4, 4)
    ) +
    scale_y_continuous(
      name = "-log10(FDR)",
      limits = c(0, 300),
      expand = expansion(mult = c(0, 0.1)),
      oob = squish_infinite
    ) +
    scale_colour_manual(
      name = "Differential TF motif enrichment",
      values = c("upregulated" = "#FC8D62", "downregulated" = "darkred", "not significant" = "lightgrey"),
      labels = c("upregulated" = "HC-MBL Cells", "downregulated" = "Physiologic B Cells")
    ) +
    theme_pub() +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal"
    ))

volcano.plot_B_CLL <- volcano.data.all.comparisons$B_vs_CLL[1:3] %>%
  map(~ ggplot(., aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
    geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
    geom_point(size = 1) +
    geom_label_repel(
      size = 2.5,
      label.size = NA,
      label.padding = 0.1,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = 40,
      colour = "black"
    ) +
    scale_x_continuous(
      name = "Difference median z-score",
      breaks = seq(-6, 6, 2),
      limits = c(-6, 6)
    ) +
    scale_y_continuous(
      name = "-log10(FDR)",
      limits = c(0, 300),
      expand = expansion(mult = c(0, 0.1)),
      oob = squish_infinite
    ) +
    scale_colour_manual(
      name = "Differential TF motif enrichment",
      values = c("upregulated" = "#66C2A5", "downregulated" = "darkred", "not significant" = "lightgrey"),
      labels = c("upregulated" = "CLL Cells", "downregulated" = "Physiologic B Cells")
    ) +
    theme_pub() +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal"
    ))


volcano.plot_HC_MBL_CLL <- volcano.data.all.comparisons$HC_MBL_vs_CLL[1:3] %>%
  map(~ ggplot(., aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
    geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
    geom_point(size = 1) +
    geom_label_repel(
      size = 2.5,
      label.size = NA,
      label.padding = 0.1,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = 40,
      colour = "black"
    ) +
    scale_x_continuous(
      name = "Difference median z-score",
      breaks = seq(-6, 6, 2),
      limits = c(-6, 6)
    ) +
    scale_y_continuous(
      name = "-log10(FDR)",
      limits = c(0, 300),
      expand = expansion(mult = c(0, 0.1)),
      oob = squish_infinite
    ) +
    scale_colour_manual(
      name = "Differential TF motif enrichment",
      values = c("upregulated" = "#66C2A5", "downregulated" = "#FC8D62", "not significant" = "lightgrey"),
      labels = c("upregulated" = "CLL Cells", "downregulated" = "HC-MBL")
    ) +
    theme_pub() +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal"
    ))

ggarrange(ggarrange(NULL, NULL, NULL, volcano.plot_B_LC_MBL, nrow = 1, ncol = 4, labels = c("", "", "", "Batch4")),
  ggarrange(plotlist = volcano.plot_B_HC_MBL, nrow = 1, ncol = 4, labels = c("Batch1", "Batch2", "Batch3", "")),
  ggarrange(plotlist = volcano.plot_B_CLL, nrow = 1, ncol = 4),
  ggarrange(plotlist = volcano.plot_HC_MBL_CLL, nrow = 1, ncol = 4),
  nrow = 4, ncol = 1
)

saveRDS(volcano.plot_B_LC_MBL, file.path(RDS_dir, "volcanoplot_TF_enrichment_pb_B_LC_MBL_batches.RDS"))
saveRDS(volcano.plot_B_HC_MBL, file.path(RDS_dir, "volcanoplot_TF_enrichment_pb_B_HC_MBL_batches.RDS"))
saveRDS(volcano.plot_B_CLL, file.path(RDS_dir, "volcanoplot_TF_enrichment_pb_B_CLL_batches.RDS"))
```

## Aggregate volcanoplots
```{r fig.width=15, fig.height=5}
volcano.plot_B_LC_MBL

volcano_HC <- volcano.data.all.comparisons$B_vs_HC[1:3] %>%
  bind_rows(.id = "batch") %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label, shape = batch)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 3),
    limits = c(-6, 6)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#FC8D62", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "HC-MBL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  scale_shape_manual(name = "Batch", values = c("1" = 15, "2" = 17, "3" = 18, "4" = 16)) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
saveRDS(volcano_HC, file.path(RDS_dir, "volcanoplot_TF_enrichment_pb_B_HC_MBL_batches_summary.RDS"))
```


```{r fig.width=15, fig.height=5}
volcano_CLL <- volcano.data.all.comparisons$B_vs_CLL[1:3] %>%
  bind_rows(.id = "batch") %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label, shape = batch)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 3),
    limits = c(-6, 6)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#66C2A5", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "CLL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  scale_shape_manual(name = "Batch", values = c("1" = 15, "2" = 17, "3" = 18, "4" = 16)) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
saveRDS(volcano_CLL, file.path(RDS_dir, "volcanoplot_TF_enrichment_pb_B_CLL_batches_summary.RDS"))
```

```{r fig.width=15, fig.height=5}
volcano_HC_CLL <- volcano.data.all.comparisons$HC_MBL_vs_CLL[1:3] %>%
  bind_rows(.id = "batch") %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label, shape = batch)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 3),
    limits = c(-6, 6)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#66C2A5", "downregulated" = "#FC8D62", "not significant" = "lightgrey"),
    labels = c("upregulated" = "CLL Cells", "downregulated" = "HC-MBL Cells")
  ) +
  scale_shape_manual(name = "Batch", values = c("1" = 15, "2" = 17, "3" = 18, "4" = 16)) +
  labs(title="HC-MBL vs. CLL")+
  theme_pub() +
  theme(
    legend.position = "none",
    legend.direction = "horizontal"
  )
saveRDS(volcano_HC_CLL, file.path(RDS_dir, "volcanoplot_TF_enrichment_pb_HC_MBL_CLL_batches_summary.RDS"))

ggsave(plot = volcano_HC_CLL, filename = file.path(plotting_dir, "volcanoplot_TF_enrichment_pb_HC_CLL_batches_summary.png"),
       width = 2.8, height = 2)
```


### Pull Deviation-scores for LC-MBL vs CLL differentially enrichted TF motifs

```{r}
comp_genes <- volcano.data.all.comparisons %>%
  bind_rows() %>%
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  str_split(string = , pattern = "_") %>%
  map_chr(~ .[[1]]) %>%
  unique() %>%
  sort()
```

```{r, fig.width=12, fig.height=6}
zscore_matrix_hm_input <- zscore_matrix_Blike %>%
  map(~ .x[, colnames(.x) %>% 
             keep(~ !str_starts(., "CLL6."))])


zscore_matrix_hm_input <- map(zscore_matrix_hm_input, ~ {
  rownames(.) <- rownames(.) %>%
    str_split_i("_", 1)
  return(.)
})
```

### Heatmap all batches full cell counts
```{r All_batches_full_cellcounts, fig.width=20, fig.height=7}
barcodes_ordered <- zscore_matrix_hm_input %>%
  imap(~ colnames(.x) %>%
    as_tibble() %>%
    dplyr::rename("label_cellbarcode" = value) %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
    dplyr::left_join(tibble(
      "label_cellbarcode" = CLLProj_Blike_batches[[.y]]$cellNames,
      "B_like_annotation" = CLLProj_Blike_batches[[.y]]@cellColData$B_like_annotation
    )) %>%
    dplyr::mutate(
        B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
        B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
        B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
        B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
        .default = B_like_annotation
      ),
      B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
      B_like_annotation_label = paste0(B_like_annotation, "_", label),
      B_like_annotation_label = factor(B_like_annotation_label,
        levels = c(
          paste0("Physiologic B Cells_", c("CLL8.1", "CLL7.2", "MBL9", "MBL13", "CLL10.2", "MBL14", "CLL4.1", "CLL3.1", "CLL10.1", "CLL1.2", "CLL6.2", "CLL5.1", "MBL3", "MBL6", "CLL8.2", "CLL9.2", "CLL9.1", "MBL7", "CLL3.2", "CLL2.1", "MBL10", "CLL6.1", "MBL12", "MBL11", "MBL1", "CLL5.2", "MBL2", "CLL7.1", "MBL8", "CLL2.2", "CLL1.1", "MBL4", "MBL15")),
          paste0("LC-MBL Cells_", c("MBL15", "MBL13", "MBL14", "MBL11", "MBL12", "MBL10")),
          paste0("HC-MBL Cells_", c("CLL8.1", "CLL9.1", "MBL9", "CLL10.1", "CLL1.1", "CLL2.1", "CLL5.1", "MBL8", "CLL4.1", "MBL7", "MBL3", "MBL1", "MBL2", "CLL7.1", "MBL6", "CLL3.1")),
          paste0("CLL Cells_", c("CLL10.2", "CLL9.2", "CLL5.2", "CLL2.2", "CLL1.2", "CLL7.2", "CLL4.2", "CLL3.2", "MBL4", "CLL8.2", "CLL6.2"))
        )
      )
    ) %>%
    arrange(B_like_annotation_label))

barcodes <- zscore_matrix_hm_input %>%
  imap(~ colnames(.x) %>%
    as_tibble() %>%
    dplyr::rename("label_cellbarcode" = value) %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
    dplyr::left_join(tibble(
      "label_cellbarcode" = CLLProj_Blike_batches[[.y]]$cellNames,
      "B_like_annotation" = CLLProj_Blike_batches[[.y]]@cellColData$B_like_annotation
    )) %>%
    dplyr::mutate(
      B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
        B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
        B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
        B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
        .default = B_like_annotation
      )) %>%
    arrange(B_like_annotation, label) %>%
    left_join(tibble(
      "B_like_annotation" = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"),
      "Colour" = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5")
    )))

batch_barcodes <- barcodes %>% 
  bind_rows() %>% 
  dplyr::mutate(B_like_annotation=factor(B_like_annotation, levels=c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>% 
  split(.$B_like_annotation)

zscore_matrix_hm_input_aggregated <- map(zscore_matrix_hm_input, ~.[comp_genes,]) %>% 
  bind_cols()


hm <- build_celltype_heatmap(bcs_per_celltype = batch_barcodes,
                             data_matrix =zscore_matrix_hm_input_aggregated,
                             metadata_all=metadata,
                             modality = "ATAC", 
                             include_all_metadata = TRUE, 
                             show_heatmap_legend = TRUE, 
                             show_annotation_legend = TRUE, 
                             absolute_capping=FALSE,
                             quantile_capping = 0.93, 
                             high_score_colour ="#783f0b", 
                             mid_score_colour = "white", 
                             low_score_colour = "#0c544c", 
                             heatmap_legend_name = "Deviation")

hm$`Physiologic B Cells`+hm$`LC-MBL Cells`+hm$`HC-MBL Cells`+hm$`CLL Cells`

hm %>%
  map(~ draw(.x, heatmap_legend_side = "bottom", annotation_legend_side = "bottom"))

hm_grob <- hm %>%
  map(~ grid::grid.grabExpr(draw(.x, heatmap_legend_side = "bottom")))
# cowplot::plot_grid(hm_grob)
# hm$`CLL Cells`
```

### Heatmaps per batch
```{r, fig.width=12, fig.height=6}
barcodes_ordered <- zscore_matrix_hm_input %>%
  imap(~ colnames(.x) %>%
    as_tibble() %>%
    dplyr::rename("label_cellbarcode" = value) %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
    dplyr::left_join(tibble(
      "label_cellbarcode" = CLLProj_Blike_batches[[.y]]$cellNames,
      "B_like_annotation" = CLLProj_Blike_batches[[.y]]@cellColData$B_like_annotation
    )) %>%
    dplyr::mutate(
        B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
        B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
        B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
        B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
        .default = B_like_annotation
      ),
      B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
      B_like_annotation_label = paste0(B_like_annotation, "_", label),
      B_like_annotation_label = factor(B_like_annotation_label,
        levels = c(
          paste0("Physiologic B Cells_", c("CLL8.1", "CLL7.2", "MBL9", "MBL13", "CLL10.2", "MBL14", "CLL4.1", "CLL3.1", "CLL10.1", "CLL1.2", "CLL6.2", "CLL5.1", "MBL3", "MBL6", "CLL8.2", "CLL9.2", "CLL9.1", "MBL7", "CLL3.2", "CLL2.1", "MBL10", "CLL6.1", "MBL12", "MBL11", "MBL1", "CLL5.2", "MBL2", "CLL7.1", "MBL8", "CLL2.2", "CLL1.1", "MBL4", "MBL15")),
          paste0("LC-MBL Cells_", c("MBL15", "MBL13", "MBL14", "MBL11", "MBL12", "MBL10")),
          paste0("HC-MBL Cells_", c("CLL8.1", "CLL9.1", "MBL9", "CLL10.1", "CLL1.1", "CLL2.1", "CLL5.1", "MBL8", "CLL4.1", "MBL7", "MBL3", "MBL1", "MBL2", "CLL7.1", "MBL6", "CLL3.1")),
          paste0("CLL Cells_", c("CLL10.2", "CLL9.2", "CLL5.2", "CLL2.2", "CLL1.2", "CLL7.2", "CLL4.2", "CLL3.2", "MBL4", "CLL8.2", "CLL6.2"))
        )
      )
    ) %>%
    arrange(B_like_annotation_label))


# map cell type annotation to this cell barcode
# map colour to barcode
barcodes <- zscore_matrix_hm_input %>%
  imap(~ colnames(.x) %>%
    as_tibble() %>%
    dplyr::rename("label_cellbarcode" = value) %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
    dplyr::left_join(tibble(
      "label_cellbarcode" = CLLProj_Blike_batches[[.y]]$cellNames,
      "B_like_annotation" = CLLProj_Blike_batches[[.y]]@cellColData$B_like_annotation
    )) %>%
    dplyr::mutate(
      B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
        B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
        B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
        B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
        .default = B_like_annotation
      ),
      B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
      label = factor(label, levels = c("MBL13", "MBL11", "MBL12", "MBL15", "MBL14", "MBL10", "CLL6.2", "CLL1.2", "CLL7.2", "CLL2.2", "CLL5.2", "CLL8.2", "CLL4.2", "CLL10.2", "CLL9.2", "CLL3.2", "MBL4"))
    ) %>%
    arrange(B_like_annotation, label) %>%
    left_join(tibble(
      "B_like_annotation" = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"),
      "Colour" = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5")
    )))

barcodes2annotation <- barcodes %>%
  map(~ dplyr::select(.x, label_cellbarcode, B_like_annotation) %>%
    dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
    deframe())

barcodes2plottingid <- barcodes %>%
  map2(zscore_matrix_hm_input, ~ left_join(.x, tibble(
    "label_cellbarcode" = colnames(.y),
    "plotting_id" = label2plotting_id[colnames(.y) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)]
  )) %>%
    dplyr::select(label_cellbarcode, plotting_id) %>%
    dplyr::mutate(plotting_id = factor(plotting_id, levels = paste0("P", 1:24))) %>%
    deframe())

colours <- barcodes %>% map(~ list(
  "Cell type" = .x %>%
    dplyr::select(B_like_annotation, Colour) %>%
    unique() %>%
    deframe(),
  "Patient" = plottingid2colours
))

anno_df <- list(
  zscore_matrix_hm_input,
  barcodes2annotation,
  barcodes2plottingid
) %>%
  pmap(~ tibble(
    "Cell type" = ..2[colnames(..1)],
    "Patient" = ..3[colnames(..1)]
  ))

column_ha <- anno_df %>%
  map2(colours, ~ HeatmapAnnotation(
    df = .x,
    col = .y,
    show_legend = FALSE,
    annotation_name_side = "left",
    annotation_name_gp = gpar(fontsize = 8),
      border = TRUE
  ))

hm_LC_CLL <- list(
  ..1 = zscore_matrix_hm_input,
  ..2 = barcodes_ordered,
  ..3 = column_ha
) %>%
  pmap(~
    Heatmap(as.matrix(..1[comp_genes, ]),
      column_order = ..2 %>% pull(label_cellbarcode),
      top_annotation = ..3,
      col = colorRamp2(breaks = c(-quantile(as.matrix(..1[comp_genes, ]), 0.93), 0, quantile(as.matrix(..1[comp_genes, ]), 0.93)), c("#0c544c", "white", "#783f0b")),
      na_col = "black",
      use_raster = TRUE,
      cluster_rows = TRUE,
      raster_by_magick = TRUE,
      cluster_columns = FALSE,
      show_column_names = FALSE,
      show_row_names = TRUE,
      show_row_dend = TRUE,
      show_column_dend = FALSE,
      show_heatmap_legend = TRUE,
      row_title_side = "left",
      row_names_side = "left",
      row_names_gp = gpar(fontsize = 8),
      row_title = "TF motif",
      row_title_gp = gpar(fontsize = 8),
      border_gp = gpar(col = "black", lty = 1),
      heatmap_legend_param = list(
        title = "Deviation-score",
        direction = "horizontal",
        legend_direction = "horizontal",
        legend_gp = gpar(
          fontsize = 8,
          fontface = "plain"
        )
      )
    ))

hm_LC_CLL %>%
  map(~ draw(.x, heatmap_legend_side = "bottom", 
             annotation_legend_side = "bottom"))

hm_LC_CLL_grob <- hm_LC_CLL %>%
  map(~ grid::grid.grabExpr(draw(.x, 
                                 heatmap_legend_side = "bottom")))
```

### Downsampled heatmap per batch
```{r downsample_barcodes}
# downsample cell barcodes to equal number per celltype
downsample_cells <- function(df, max_cells) {
  set.seed(123) # Set seed for reproducibility
  num_patients <- n_distinct(df$label)
  cells_per_patient <- max_cells %/% num_patients # integer division to get the cells per patient
  remaining_cells <- max_cells %% num_patients # remaining cells after even distribution

  # Sample cells per patient with even distribution
  downsampled <- df %>%
    group_by(label) %>%
    sample_n(size = min(cells_per_patient, n()), replace = FALSE) %>%
    ungroup()

  # If there are remaining cells, sample them from the already downsampled data
  if (remaining_cells > 0) {
    remaining <- df %>%
      anti_join(downsampled, by = "label_cellbarcode") %>%
      sample_n(size = min(remaining_cells, n()), replace = FALSE)

    downsampled <- bind_rows(downsampled, remaining)
  }

  return(downsampled)
}


downsample_parameters <- tibble(
  batch = c(1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4),
  celltype = c(rep(c("Physiologic B Cells", "HC-MBL Cells", "CLL Cells"), 3), c("Physiologic B Cells", "LC-MBL Cells")),
  downsample = c(
    c(200000, 113, 113),
    c(200000, 1776, 1776),
    c(200000, 501, 501),
    c(200000, 2457)
  )
)

# downsampling number for healthy B is so high, that really all cells are considered and no downsampling is performed in this case
celltype_barcodes_batch <- list()
for (batchnumber in 1:length(barcodes_ordered)) {
  downsample_parameters_batch <- downsample_parameters %>%
    dplyr::filter(batch %in% batchnumber)
  print(downsample_parameters_batch)

  barcodes_celltype <- list()
  for (row in 1:length(downsample_parameters_batch$celltype)) {
    print(row)
    barcodes_celltype[[downsample_parameters_batch$celltype[[row]]]] <- barcodes_ordered[[batchnumber]] %>%
      dplyr::filter(B_like_annotation == downsample_parameters_batch$celltype[[row]]) %>%
      downsample_cells(max_cells = downsample_parameters_batch$downsample[[row]])
  }
  celltype_barcodes_batch[[batchnumber]] <- barcodes_celltype
}
```

```{r cluster_patients}
# per patient calculate mean value per TF and run hierarchical clustering
clustering_input <- list()

for (batch in 1:length(celltype_barcodes_batch)) {
  clustering_input_batch <- celltype_barcodes_batch[[batch]] %>%
    map(~ zscore_matrix_hm_input[[batch]][comp_genes, .$label_cellbarcode] %>%
      as_tibble(rownames = NA) %>%
      rownames_to_column("TF") %>%
      pivot_longer(cols = -TF, names_to = "label_cellbarcode", values_to = "z_score") %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#") %>%
      group_by(TF, label) %>%
      summarise("mean_z_score" = mean(`z_score`)) %>%
      ungroup() %>%
      pivot_wider(id_cols = label, names_from = TF, values_from = mean_z_score) %>%
      column_to_rownames(var = "label")) %>%
    set_names(names(celltype_barcodes_batch[[batch]]))
  clustering_input[[batch]] <- clustering_input_batch
}

distance_mat <- list()
clusters <- list()
patients_ordered <- list()

for (batch in 1:length(clustering_input)) {
  set.seed(20)
  distance_mat_batch <- clustering_input[[batch]] %>%
    map(~ dist(., method = "euclidean"))

  clusters_batch <- distance_mat_batch %>%
    map(~ hclust(., method = "average"))

  patients_ordered_batch <- clustering_input[[batch]] %>%
    map2(clusters_batch, ~ rownames(.x)[.y %>%
      as.dendrogram() %>%
      order.dendrogram()])

  distance_mat[[batch]] <- distance_mat_batch
  clusters[[batch]] <- clusters_batch
  patients_ordered[[batch]] <- patients_ordered_batch
}
```

```{r column2_annotations}
celltype_barcodes_ordered <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  celltype_barcodes_ordered_batch <- celltype_barcodes_batch[[batch]] %>%
    map2(
      patients_ordered[[batch]],
      ~ dplyr::mutate(.x, label = factor(label, levels = .y)) %>%
        arrange(label)
    )

  celltype_barcodes_ordered[[batch]] <- celltype_barcodes_ordered_batch
}

barcodes2annotation <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  barcodes2annotation_batch <- celltype_barcodes_batch[[batch]] %>%
    map(~ dplyr::select(., label_cellbarcode, B_like_annotation) %>%
      dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
      deframe())

  barcodes2annotation[[batch]] <- barcodes2annotation_batch
}

barcodes2label <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  barcodes2label_batch <- celltype_barcodes_batch[[batch]] %>%
    map2(patients_ordered[[batch]], ~ left_join(.x, tibble(
      "label_cellbarcode" = colnames(zscore_matrix_hm_input[[batch]][, .x$label_cellbarcode]),
      "label" = colnames(zscore_matrix_hm_input[[batch]][, .x$label_cellbarcode]) %>%
        as_tibble() %>%
        dplyr::rename("label_cellbarcode" = value) %>%
        tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
        pull(label)
    )) %>%
      dplyr::select(label_cellbarcode, label) %>%
      dplyr::mutate(label = factor(label, levels = .y)) %>%
      deframe())
  barcodes2label[[batch]] <- barcodes2label_batch
}

barcodes2IGHV <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  barcodes2IGHV_batches <- celltype_barcodes_batch[[batch]] %>%
    map2(patients_ordered[[batch]], ~ left_join(.x, metadata %>%
      dplyr::filter(data_type == "ATAC") %>%
      dplyr::select("label", "IGHV")) %>%
      dplyr::mutate(
        label = factor(label, levels = .y),
        IGHV = if_else(is.na(IGHV), "NA", IGHV)
      ) %>%
      dplyr::select(label_cellbarcode, IGHV) %>%
      deframe())
  barcodes2IGHV[[batch]] <- barcodes2IGHV_batches
}

barcode2processing_date <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  barcode2processing_date_batches <- celltype_barcodes_batch[[batch]] %>%
    map2(patients_ordered[[batch]], ~ left_join(.x, metadata %>%
      dplyr::filter(data_type == "ATAC") %>%
      dplyr::select("label", "processing_date")) %>%
      dplyr::mutate(
        label = factor(label, levels = .y),
        processing_date = as.character(processing_date)
      ) %>%
      dplyr::select(label_cellbarcode, processing_date) %>%
      deframe())
  barcode2processing_date[[batch]] <- barcode2processing_date_batches
}

barcode2technology <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  barcode2technology_batches <- celltype_barcodes_batch[[batch]] %>%
    map2(patients_ordered[[batch]], ~ left_join(.x, metadata %>%
      dplyr::filter(data_type == "ATAC") %>%
      dplyr::select("label", "technology")) %>%
      dplyr::mutate(label = factor(label, levels = .y)) %>%
      dplyr::select(label_cellbarcode, technology) %>%
      deframe())
  barcode2technology[[batch]] <- barcode2technology_batches
}

barcode2sequencing_date <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  barcode2sequencing_date_batches <- celltype_barcodes_batch[[batch]] %>%
    map2(patients_ordered[[batch]], ~ left_join(.x, metadata %>%
      dplyr::filter(data_type == "ATAC") %>%
      dplyr::select("label", "sequencing_date")) %>%
      dplyr::mutate(
        label = factor(label, levels = .y),
        sequencing_date = as.character(sequencing_date)
      ) %>%
      dplyr::select(label_cellbarcode, sequencing_date) %>%
      deframe())
  barcode2sequencing_date[[batch]] <- barcode2sequencing_date_batches
}

barcode2sex <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  barcode2sex_batches <- celltype_barcodes_batch[[batch]] %>%
    map2(patients_ordered[[batch]], ~ left_join(.x, metadata %>%
      dplyr::filter(data_type == "ATAC") %>%
      dplyr::select("label", "sex")) %>%
      dplyr::mutate(label = factor(label, levels = .y)) %>%
      dplyr::select(label_cellbarcode, sex) %>%
      deframe())
  barcode2sex[[batch]] <- barcode2sex_batches
}

"barcodes2sequencing_date_?" <- list()
for (batch in 1:length(celltype_barcodes_batch)) {
  "barcodes2sequencing_date_?_batches" <- celltype_barcodes_batch[[batch]] %>%
    map2(patients_ordered[[batch]], ~ left_join(.x, metadata %>%
      dplyr::filter(data_type == "ATAC") %>%
      dplyr::select("label", "sequencing_date_?")) %>%
      dplyr::mutate(
        label = factor(label, levels = .y),
        sequencing_date = as.character(`sequencing_date_?`)
      ) %>%
      dplyr::select(label_cellbarcode, `sequencing_date_?`) %>%
      deframe())
  `barcodes2sequencing_date_?`[[batch]] <- `barcodes2sequencing_date_?_batches`
}


label2colours <- c(my_paired_colours_named, my_unpaired_colours_named)

colours <- list(
  "Cell type" = c("Physiologic B Cells" = "darkred", "LC-MBL Cells" = "#fcbda4", "HC-MBL Cells" = "#FC8D62", "CLL Cells" = "#66C2A5"),
  "Patient" = label2colours,
  "IGHV" = c("mutated" = "steelblue", "unmutated" = "orange", "NA" = "lightgrey"),
  "processing_date" = c(
    "2022-01-10" = "#1B9E77", "2022-10-20" = "#D95F02",
    "2022-11-04" = "#7570B3", "2023-03-24" = "#66A61E",
    "2024-03-05" = "#A6761D"
  ),
  "technology" = c("v2 scATAC-seq" = "#666666", "multiome kit" = "#E7298A"),
  "sequencing_date" = c("2022-02-26" = "#E6AB02", "2022-11-18" = "#7FC97F", "2023-05-02" = "#FFFF99", "2024-03-18" = "#386CB0"),
  "sequencing_date_?" = c("2022-01-14" = "midnightblue", "2022-02-26" = "darkgreen", "2022-10-20" = "steelblue4", "2022-10-26" = "magenta4", "2022-10-27" = "purple4", "2024-03-18" = "#386CB0", "2023-05-02" = "violet", "2022-11-18" = "plum4"),
  "sex" = c("male" = "lightblue", "female" = "lightpink")
)

anno_df <- list()
for (batch in 1:length(zscore_matrix_hm_input)) {
  anno_df_batches <- list(
    barcodes2annotation[[batch]],
    barcodes2label[[batch]],
    barcodes2IGHV[[batch]],
    barcode2processing_date[[batch]],
    barcode2technology[[batch]],
    barcode2sequencing_date[[batch]],
    barcode2sequencing_date[[batch]],
    barcode2sex[[batch]],
    celltype_barcodes_batch[[batch]]
  ) %>%
    pmap(~ tibble(
      "Cell type" = ..1[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
      "Patient" = ..2[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
      "IGHV" = ..3[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
      "processing_date" = ..4[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
      "technology" = ..5[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
      "sequencing_date" = ..6[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
      "sequencing_date_?" = ..7[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
      "sex" = ..8[colnames(zscore_matrix_hm_input[[batch]][, ..9$label_cellbarcode])],
    ))
  anno_df[[batch]] <- anno_df_batches
}

column_ha <- list()
for (batch in 1:length(anno_df)) {
  column_ha_batches <- anno_df[[batch]] %>%
    map(~ HeatmapAnnotation(
      df = .,
      col = colours,
      show_legend = FALSE,
      annotation_name_side = "left",
      annotation_name_gp = gpar(fontsize = 8)
    ))
  column_ha[[batch]] <- column_ha_batches
}
```

```{r, plot_heatmap}
hm <- list()
for (batch in 1:4) {
  print(batch)
  hm_batches <- list(
    celltype_barcodes_batch[[batch]],
    column_ha[[batch]],
    celltype_barcodes_ordered[[batch]] %>%
      map(~ pull(.x, label_cellbarcode))
  ) %>%
    pmap(~ Heatmap(as.matrix(zscore_matrix_hm_input[[batch]][comp_genes, ..1$label_cellbarcode]),
      col = colorRamp2(breaks = c(-quantile(as.matrix(zscore_matrix_hm_input[[batch]][comp_genes, ]), 0.93), 0, quantile(as.matrix(zscore_matrix_hm_input[[batch]][comp_genes, ]), 0.93)), c("#0c544c", "white", "#783f0b")),
      top_annotation = ..2,
      column_order = ..3,
      row_order = c("EBF1", "NFE2", "NFE2L2", "BATF", "MAFF", "MAFK", "POU2F2", "SPI1", "SPIB", "SPIC", "SMARCC1", "FOSL1", "JUNB", "FOSL2", "JUND", "FOS", "JUN", "FOSB", "BACH1", "BACH2", "JDP2", "NFATC2", "NFATC3", "CTCF", "TCF4"),
      use_raster = TRUE,
      cluster_rows = FALSE,
      cluster_columns = FALSE,
      raster_by_magick = TRUE,
      show_column_names = FALSE,
      show_heatmap_legend = FALSE,
      na_col = "black",
      row_title_side = "left",
      row_names_side = "left",
      row_names_gp = gpar(fontsize = 8),
      row_title_gp = gpar(fontsize = 8),
      heatmap_legend_param = list(
        title = "Z-score",
        direction = "horizontal",
        legend_direction = "vertical",
        legend_gp = gpar(
          fontsize = 8,
          fontface = "plain"
        )
      )
    ))
  hm[[batch]] <- hm_batches
}
```

```{r, fig.width=10, fig.height=7}
hm_LC_CLL_grob_downsampled <- list()
for (batch in 1:length(hm)) {
  draw(hm[[batch]]$`Physiologic B Cells` + hm[[batch]]$`LC-MBL Cells` + hm[[batch]]$`HC-MBL Cells` + hm[[batch]]$`CLL Cells`)

  hm_LC_CLL_grob_downsampled_batches <- grid::grid.grabExpr(draw(hm[[batch]]$`Physiologic B Cells` + hm[[batch]]$`LC-MBL Cells` + hm[[batch]]$`HC-MBL Cells` + hm[[batch]]$`CLL Cells`))

  hm_LC_CLL_grob_downsampled[[batch]] <- hm_LC_CLL_grob_downsampled_batches

  cowplot::plot_grid(hm_LC_CLL_grob_downsampled_batches)
}
# saveRDS(hm_LC_CLL_grob_downsampled, file = file.path(project_dir, "RDS_Objects/TF_enrichment_B_LC_MBL_CLL_marker_heatmap_downsampled.RDS"))
cowplot::plot_grid(TF_enrichment_B_LC_MBL_CLL_marker_heatmap_downsampled)
```

## Batch correlation analysis
```{r}
zscore_matrix_hm_input_long <- pmap(
  list(
    ..1 = zscore_matrix_hm_input[1:4],
    ..2 = barcodes2annotation[1:4]
  ),
  .f = ~ rownames_to_column(..1, var = "TF") %>%
    tidyr::pivot_longer(-TF, names_to = "label_cellbarcode", values_to = "z_score") %>%
    left_join(..2 %>% unlist() %>% enframe(name = "label_cellbarcode", value = "celltype") %>% 
                tidyr::separate("label_cellbarcode", into = c(NA, "label_cellbarcode"), sep = "\\.")) %>%
    group_by(TF, celltype) %>%
    dplyr::summarize(
      mean_z_score = mean(z_score),
      median_z_score = median(z_score)
    )
) %>%
  bind_rows(.id = "batch") %>%
  dplyr::select(TF, celltype, mean_z_score, median_z_score, batch) %>%
  pivot_wider(
    id_cols = c(TF, celltype),
    names_from = c(batch, batch),
    values_from = c(mean_z_score, median_z_score)
  ) %>%
  dplyr::mutate(
    TF_oi = if_else(condition = TF %in% comp_genes, true = celltype, false = NA),
    TF_oi_label = if_else(condition = TF %in% comp_genes, true = TF, false = NA)
  )
```

```{r fig.width=15, fig.height=16}
b1vsb2_mean <- zscore_matrix_hm_input_long %>%
  dplyr::filter(celltype != "LC-MBL Cells") %>%
  ggplot(aes(x = mean_z_score_2, y = mean_z_score_1, colour = TF_oi)) +
  geom_smooth(aes(colour = TF_oi, fill = TF_oi), method = "lm", alpha = 0.4, fullrange = TRUE) +
  geom_point(size = 0.8) +
  geom_rug() +
  geom_text_repel(aes(label = TF_oi_label, color = TF_oi), size = 3, max.overlaps = 30) +
  stat_cor(aes(color = TF_oi), method = "pearson", p.accuracy = 0.001, r.accuracy = 0.01) +
  scale_colour_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_fill_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_y_continuous(name = "Batch1 mean Z-score", limits = c(-1.5, 2.5)) +
  scale_x_continuous(name = "Batch2 mean Z-score", limits = c(-1.5, 2.5)) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  theme_pub() +
  facet_wrap(~celltype)

b2vsb3_mean <- zscore_matrix_hm_input_long %>%
  dplyr::filter(celltype != "LC-MBL Cells") %>%
  ggplot(aes(x = mean_z_score_3, y = mean_z_score_2, colour = TF_oi)) +
  geom_smooth(aes(colour = TF_oi, fill = TF_oi), method = "lm", alpha = 0.4, fullrange = TRUE) +
  geom_point(size = 0.8) +
  geom_rug() +
  geom_text_repel(aes(label = TF_oi_label, color = TF_oi), size = 3, max.overlaps = 30) +
  stat_cor(aes(color = TF_oi), method = "pearson", p.accuracy = 0.001, r.accuracy = 0.01) +
  scale_colour_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_fill_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_y_continuous(name = "Batch2 mean Z-score", limits = c(-2, 5)) +
  scale_x_continuous(name = "Batch3 mean Z-score", limits = c(-2, 5)) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  theme_pub() +
  facet_wrap(~celltype)

b1vsb3_mean <- zscore_matrix_hm_input_long %>%
  dplyr::filter(celltype != "LC-MBL Cells") %>%
  ggplot(aes(x = mean_z_score_3, y = mean_z_score_1, colour = TF_oi)) +
  geom_smooth(aes(colour = TF_oi, fill = TF_oi), method = "lm", alpha = 0.4, fullrange = TRUE) +
  geom_point(size = 0.8) +
  geom_rug() +
  geom_text_repel(aes(label = TF_oi_label, color = TF_oi), size = 3, max.overlaps = 30) +
  stat_cor(aes(color = TF_oi), method = "pearson", p.accuracy = 0.001, r.accuracy = 0.01) +
  scale_colour_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_fill_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_y_continuous(name = "Batch1 mean Z-score", limits = c(-4, 4)) +
  scale_x_continuous(name = "Batch3 mean Z-score", limits = c(-4, 4)) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  theme_pub() +
  facet_wrap(~celltype)

ggarrange(b1vsb2_mean, b2vsb3_mean, b1vsb3_mean, nrow = 3, ncol = 1, common.legend = TRUE, legend = "bottom")
```

```{r fig.width=15, fig.height=16}
b1vsb2_median <- zscore_matrix_hm_input_long %>%
  dplyr::filter(celltype != "LC-MBL Cells") %>%
  ggplot(aes(x = median_z_score_2, y = median_z_score_1, colour = TF_oi)) +
  geom_smooth(aes(colour = TF_oi, fill = TF_oi), method = "lm", alpha = 0.4, fullrange = TRUE) +
  geom_point(size = 0.8) +
  geom_rug() +
  geom_text_repel(aes(label = TF_oi_label, color = TF_oi), size = 3, max.overlaps = 30) +
  stat_cor(aes(color = TF_oi), method = "pearson", p.accuracy = 0.001, r.accuracy = 0.01) +
  scale_colour_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_fill_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_y_continuous(name = "Batch1 median Z-score", limits = c(-1.5, 2.5)) +
  scale_x_continuous(name = "Batch2 median Z-score", limits = c(-1.5, 2.5)) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  theme_pub() +
  facet_wrap(~celltype)

b2vsb3_median <- zscore_matrix_hm_input_long %>%
  dplyr::filter(celltype != "LC-MBL Cells") %>%
  ggplot(aes(x = median_z_score_3, y = median_z_score_2, colour = TF_oi)) +
  geom_smooth(aes(colour = TF_oi, fill = TF_oi), method = "lm", alpha = 0.4, fullrange = TRUE) +
  geom_point(size = 0.8) +
  geom_rug() +
  geom_text_repel(aes(label = TF_oi_label, color = TF_oi), size = 3, max.overlaps = 30) +
  stat_cor(aes(color = TF_oi), method = "pearson", p.accuracy = 0.001, r.accuracy = 0.01) +
  scale_colour_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_fill_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_y_continuous(name = "Batch2 median Z-score", limits = c(-2, 5)) +
  scale_x_continuous(name = "Batch3 median Z-score", limits = c(-2, 5)) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  theme_pub() +
  facet_wrap(~celltype)

b1vsb3_median <- zscore_matrix_hm_input_long %>%
  dplyr::filter(celltype != "LC-MBL Cells") %>%
  ggplot(aes(x = median_z_score_3, y = median_z_score_1, colour = TF_oi)) +
  geom_smooth(aes(colour = TF_oi, fill = TF_oi), method = "lm", alpha = 0.4, fullrange = TRUE) +
  geom_point(size = 0.8) +
  geom_rug() +
  geom_text_repel(aes(label = TF_oi_label, color = TF_oi), size = 3, max.overlaps = 30) +
  stat_cor(aes(color = TF_oi), method = "pearson", p.accuracy = 0.001, r.accuracy = 0.01) +
  scale_colour_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_fill_manual(name = "Significant TF motif", values = c(
    "LC-MBL Cells" = "#fcbda4",
    "HC-MBL Cells" = "#FC8D62",
    "CLL Cells" = "#66C2A5",
    "Physiologic B Cells" = "darkred"
  ), na.value = "lightgrey") +
  scale_y_continuous(name = "Batch1 median Z-score", limits = c(-4, 4)) +
  scale_x_continuous(name = "Batch3 median Z-score", limits = c(-4, 4)) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  theme_pub() +
  facet_wrap(~celltype)

ggarrange(b1vsb2_median, b2vsb3_median, b1vsb3_median, nrow = 3, ncol = 1, common.legend = TRUE, legend = "bottom")
```


## ComBat correction
```{r fig.width=10, fig.height=10}
barcode2covariate <- tibble(
  "label_cellbarcode" = CLLProj_Blike$cellNames,
  "My_annotation" = CLLProj_Blike$My_annotation,
  "TSSEnrichment" = CLLProj_Blike$TSSEnrichment,
  "log10(nFrags)" = log10(CLLProj_Blike$nFrags)
) %>%
  dplyr::mutate(My_annotation_LC = case_when(str_starts(label_cellbarcode, pattern = "MBL10|MBL11|MBL12|MBL13|MBL14|MBL15") & My_annotation == "MBL_B_Cells" ~ "LC_MBL_B_Cells",
    str_starts(label_cellbarcode, pattern = "MBL10|MBL11|MBL12|MBL13|MBL14|MBL15", negate = TRUE) & My_annotation == "MBL_B_Cells" ~ "HC_MBL_B_Cells",
    .default = My_annotation
  )) %>%
  dplyr::mutate(covariate = case_when(
    My_annotation_LC == "B" ~ 1,
    My_annotation_LC == "LC_MBL_B_Cells" ~ 2,
    My_annotation_LC == "HC_MBL_B_Cells" ~ 3,
    My_annotation_LC == "CLL_B_Cells" ~ 4
  )) %>%
  dplyr::select(label_cellbarcode, TSSEnrichment, `log10(nFrags)`, covariate)

# concatenate batch data; this matrix still holds a lot of NA values, as ENSG00000250542_156 is not present in the LC-MBL dataset --> will be excluded
combat_input_tmp <- zscore_matrix_Blike %>%
  map(~ rownames_to_column(.x, "TF") %>%
    dplyr::filter(TF != "ENSG00000250542_156"))

combat_input <- combat_input_tmp[[1]] %>%
  full_join(combat_input_tmp[[2]], by = "TF") %>%
  full_join(combat_input_tmp[[3]], by = "TF") %>%
  full_join(combat_input_tmp[[4]], by = "TF") %>%
  column_to_rownames("TF") %>%
  t() %>%
  as.data.frame()

barcode2batch <- rownames(combat_input) %>%
  tibble::enframe(value = "label_cellbarcode", name = NULL) %>%
  tidyr::separate(col = "label_cellbarcode", into = c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(label, sequencing_date)) %>%
  dplyr::mutate(sequencing_date = as.character(sequencing_date)) %>%
  dplyr::mutate(batch = case_when(sequencing_date == "2022-02-26" ~ 1,
    sequencing_date == "2022-11-18" ~ 2,
    sequencing_date == "2023-05-02" ~ 3,
    sequencing_date == "2024-03-18" ~ 4,
    .default = NA
  )) %>%
  dplyr::select(label_cellbarcode, batch) %>%
  tibble::deframe()

# subset with all batch levels and all covariate values
# combat_input_subset <- combat_input[c(0:100, 20000:20100, 96500:96600, 140000:140100), 771:870] %>%
#   t()
# mod.combat <- model.matrix(~-1,data = combat_input[c(0:100, 20000:20100, 96500:96600, 140000:140100), 771:871] )

combat_input_nona <- t(as.matrix(combat_input))[, colSums(is.na(t(as.matrix(combat_input)))) == 0]
batch <- barcode2batch[colnames(combat_input_nona)]

# column names need to be cellbarcodes, rownames need to be TF names and covariate; nFrags and celltype are confounded with the batch
result <- ComBat(
  dat = combat_input_nona,
  # mod = mod.combat,
  batch = batch,
  par.prior = TRUE,
  prior.plots = FALSE
)

#saveRDS(result, file = file.path(RDS_dir, "combat_result.RDS"))
result <- readRDS(file = file.path(RDS_dir, "combat_result.RDS"))
```

### PCAs ComBat
```{r}
pca_raw <- as.matrix(t(combat_input_nona)) %>%
  prcomp(scale = TRUE)

pca_combat <- t(result) %>%
  prcomp(scale = TRUE)
```


```{r fig.width=30, fig.height=10}
cowplot::plot_grid(
  as_tibble(pca_raw$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    ggplot(aes(x = PC1, y = PC2, colour = batch)) +
    geom_density_2d() +
    labs(title = "Pre ComBat", colour = "Batch") +
    scale_colour_manual(values = c("1" = "#E6AB02", "2" = "#7FC97F", "3" = "#FFFF99", "4" = "#386CB0")) +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  as_tibble(pca_combat$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    ggplot(aes(x = PC1, y = PC2, colour = batch)) +
    geom_density_2d() +
    scale_colour_manual(values = c("1" = "#E6AB02", "2" = "#7FC97F", "3" = "#FFFF99", "4" = "#386CB0")) +
    labs(title = "Post ComBat", colour = "Batch") +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  as_tibble(pca_raw$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    ggplot(aes(x = PC1, y = PC2, colour = TSSEnrichment)) +
    geom_point(shape = 16, size = 0.5) +
    scale_colour_distiller(palette = "YlGnBu", direction = 1, limits = c(0, 60)) +
    labs(title = "Pre ComBat") +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  as_tibble(pca_combat$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    ggplot(aes(x = PC1, y = PC2, colour = TSSEnrichment)) +
    geom_point(shape = 16, size = 0.5) +
    scale_colour_distiller(palette = "YlGnBu", direction = 1, limits = c(0, 60)) +
    labs(title = "Post ComBat") +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  as_tibble(pca_raw$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    ggplot(aes(x = PC1, y = PC2, colour = as.factor(covariate))) +
    geom_density_2d() +
    scale_colour_manual(name = "Cell type", values = c("1" = "darkred", "2" = "#fcbda4", "3" = "#FC8D62", "4" = "#66C2A5")) +
    labs(title = "Pre ComBat", colour = "Cell type") +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  as_tibble(pca_combat$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    ggplot(aes(x = PC1, y = PC2, colour = as.factor(covariate))) +
    geom_density_2d() +
    scale_colour_manual(name = "Cell type", values = c("1" = "darkred", "2" = "#fcbda4", "3" = "#FC8D62", "4" = "#66C2A5")) +
    labs(title = "Post ComBat", colour = "Cell type") +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  as_tibble(pca_raw$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    dplyr::mutate(nFrags = `log10(nFrags)`) %>%
    ggplot(aes(x = PC1, y = PC2, colour = nFrags)) +
    geom_point(shape = 16, size = 0.5) +
    scale_color_distiller(name = "log10(nFrags)", palette = "YlOrRd", direction = 1, limits = c(2.5, 5)) +
    labs(title = "Post ComBat") +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  as_tibble(pca_combat$x[, 1:2]) %>%
    cbind(label_cellbarcode = colnames(combat_input_nona)) %>%
    dplyr::mutate(batch = as.factor(barcode2batch[label_cellbarcode])) %>%
    dplyr::left_join(barcode2covariate) %>%
    dplyr::mutate(nFrags = `log10(nFrags)`) %>%
    ggplot(aes(x = PC1, y = PC2, colour = nFrags)) +
    geom_point(shape = 16, size = 0.5) +
    scale_color_distiller(name = "log10(nFrags)", palette = "YlOrRd", direction = 1, limits = c(2.5, 5)) +
    labs(title = "Post ComBat") +
    theme_pub() +
    theme(aspect.ratio = 1, legend.position = "bottom") +
    facet_wrap(~batch),
  nrow = 2,
  ncol = 4,
  byrow = FALSE
)
```

### Heatmap
```{r}
zscore_matrix_combat_hm_input <- result[, colnames(result) %>%
  keep(~ !str_starts(., "CLL6."))]

rownames(zscore_matrix_combat_hm_input) <- rownames(zscore_matrix_combat_hm_input) %>%
  str_split(string = , pattern = "_") %>%
  map_chr(~ .[[1]])

barcodes_ordered <- colnames(zscore_matrix_combat_hm_input) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = CLLProj_Blike$cellNames,
    "B_like_annotation" = CLLProj_Blike@cellColData$B_like_annotation
  )) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
      B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
      B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
    B_like_annotation_label = paste0(B_like_annotation, "_", label),
    B_like_annotation_label = factor(B_like_annotation_label,
      levels = c(
        paste0("Physiologic B Cells_", c("CLL8.1", "CLL7.2", "MBL9", "MBL13", "CLL10.2", "MBL14", "CLL4.1", "CLL3.1", "CLL10.1", "CLL1.2", "CLL6.2", "CLL5.1", "MBL3", "MBL6", "CLL8.2", "CLL9.2", "CLL9.1", "MBL7", "CLL3.2", "CLL2.1", "MBL10", "CLL6.1", "MBL12", "MBL11", "MBL1", "CLL5.2", "MBL2", "CLL7.1", "MBL8", "CLL2.2", "CLL1.1", "MBL4", "MBL15")),
        paste0("LC-MBL Cells_", c("MBL15", "MBL13", "MBL14", "MBL11", "MBL12", "MBL10")),
        paste0("HC-MBL Cells_", c("CLL8.1", "CLL9.1", "MBL9", "CLL10.1", "CLL1.1", "CLL2.1", "CLL5.1", "MBL8", "CLL4.1", "MBL7", "MBL3", "MBL1", "MBL2", "CLL7.1", "MBL6", "CLL3.1")),
        paste0("CLL Cells_", c("CLL10.2", "CLL9.2", "CLL5.2", "CLL2.2", "CLL1.2", "CLL7.2", "CLL4.2", "CLL3.2", "MBL4", "CLL8.2", "CLL6.2"))
      )
    )
  ) %>%
  arrange(B_like_annotation_label)


# map cell type annotation to this cell barcode
# map colour to barcode
barcodes <- colnames(zscore_matrix_combat_hm_input) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = CLLProj_Blike$cellNames,
    "B_like_annotation" = CLLProj_Blike@cellColData$B_like_annotation
  )) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
      B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
      B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
    label = factor(label, levels = c("MBL13", "MBL11", "MBL12", "MBL15", "MBL14", "MBL10", "CLL6.2", "CLL1.2", "CLL7.2", "CLL2.2", "CLL5.2", "CLL8.2", "CLL4.2", "CLL10.2", "CLL9.2", "CLL3.2", "MBL4"))
  ) %>%
  arrange(B_like_annotation, label) %>%
  left_join(tibble(
    "B_like_annotation" = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"),
    "Colour" = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5")
  ))

barcodes2annotation <- barcodes %>%
  dplyr::select(label_cellbarcode, B_like_annotation) %>%
  dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
  deframe()

barcodes2plottingid <- barcodes %>%
  left_join(tibble(
    "label_cellbarcode" = colnames(zscore_matrix_combat_hm_input),
    "plotting_id" = label2plotting_id[colnames(zscore_matrix_combat_hm_input) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)]
  )) %>%
  dplyr::select(label_cellbarcode, plotting_id) %>%
  dplyr::mutate(plotting_id = factor(plotting_id, levels = paste0("P", 1:24))) %>%
  deframe()

plottingid2colours <- c(
  "P15" = "darkgoldenrod1",
  "P17" = "darkorange1",
  "P23" = "indianred1",
  "P20" = "plum1",
  "P18" = "orchid",
  "P14" = "cadetblue1",
  "P16" = "turquoise",
  "P19" = "darkseagreen1",
  "P22" = "burlywood1",
  "P21" = "lightsalmon",
  "P8" = "darkslategrey",
  "P12" = "rosybrown3",
  "P10" = "thistle4",
  "P24" = "olivedrab",
  "P7" = "steelblue4",
  "P9" = "slateblue2",
  "P9dup" = "lavender",
  "P11" = "lightcoral",
  "P13" = "yellow3",
  "P6" = "lightgoldenrod1",
  "P3" = "antiquewhite2",
  "P2" = "azure4",
  "P1" = "aquamarine4",
  "P5" = "darkolivegreen1",
  "P4" = "lightsteelblue"
)

colours <- list(
  "Cell type" = barcodes %>%
    dplyr::select(B_like_annotation, Colour) %>%
    unique() %>%
    deframe(),
  "Patient" = plottingid2colours
)

anno_df <- tibble(
  "Cell type" = barcodes2annotation[colnames(zscore_matrix_combat_hm_input)],
  "Patient" = barcodes2plottingid[colnames(zscore_matrix_combat_hm_input)]
)

column_ha <- HeatmapAnnotation(
  df = anno_df,
  col = colours,
  show_legend = FALSE,
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8)
)

hm_combat_LC_CLL <- Heatmap(as.matrix(zscore_matrix_combat_hm_input),
  column_order = barcodes_ordered %>% pull(label_cellbarcode),
  top_annotation = column_ha,
  # col = colorRamp2(breaks = c(-quantile(as.matrix(zscore_matrix_combat_hm_input), 0.93), 0, quantile(as.matrix(zscore_matrix_combat_hm_input), 0.93)), c("#0c544c", "white", "#783f0b")),
  na_col = "black",
  use_raster = TRUE,
  cluster_rows = TRUE,
  raster_by_magick = TRUE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  show_row_names = TRUE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  show_heatmap_legend = TRUE,
  row_title_side = "left",
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),
  row_title = "TF motif",
  row_title_gp = gpar(fontsize = 8),
  heatmap_legend_param = list(
    title = "Z-score",
    direction = "horizontal",
    legend_direction = "horizontal",
    legend_gp = gpar(
      fontsize = 8,
      fontface = "plain"
    )
  )
)

draw(hm_combat_LC_CLL, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")

hm_LC_CLL_grob <- grid::grid.grabExpr(draw(hm_combat_LC_CLL, heatmap_legend_side = "bottom"))
```

### Downsampled heatmap separated by celltype

```{r}
# remove low quality and multiome samples from  zscore matrix
zscore_matrix_combat_hm_input <- result[, colnames(result) %>%
  keep(~ !str_starts(., "CLL6."))]

rownames(zscore_matrix_combat_hm_input) <- rownames(zscore_matrix_combat_hm_input) %>%
  str_split(string = , pattern = "_") %>%
  map_chr(~ .[[1]])
```

```{r}
# find out how many cells there are per celltype
# 4683 cells are physiologic B cells
barcodes_ordered <- colnames(zscore_matrix_combat_hm_input) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = CLLProj_Blike$cellNames,
    "B_like_annotation" = CLLProj_Blike@cellColData$B_like_annotation
  )) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
      B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
      B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
    B_like_annotation_label = paste0(B_like_annotation, "_", label),
    B_like_annotation_label = factor(B_like_annotation_label,
      levels = c(
        paste0("Physiologic B Cells_", c("CLL8.1", "CLL7.2", "MBL9", "MBL13", "CLL10.2", "MBL14", "CLL4.1", "CLL3.1", "CLL10.1", "CLL1.2", "CLL6.2", "CLL5.1", "MBL3", "MBL6", "CLL8.2", "CLL9.2", "CLL9.1", "MBL7", "CLL3.2", "CLL2.1", "MBL10", "CLL6.1", "MBL12", "MBL11", "MBL1", "CLL5.2", "MBL2", "CLL7.1", "MBL8", "CLL2.2", "CLL1.1", "MBL4", "MBL15")),
        paste0("LC-MBL Cells_", c("MBL15", "MBL13", "MBL14", "MBL11", "MBL12", "MBL10")),
        paste0("HC-MBL Cells_", c("CLL8.1", "CLL9.1", "MBL9", "CLL10.1", "CLL1.1", "CLL2.1", "CLL5.1", "MBL8", "CLL4.1", "MBL7", "MBL3", "MBL1", "MBL2", "CLL7.1", "MBL6", "CLL3.1")),
        paste0("CLL Cells_", c("CLL10.2", "CLL9.2", "CLL5.2", "CLL2.2", "CLL1.2", "CLL7.2", "CLL4.2", "CLL3.2", "MBL4", "CLL8.2", "CLL6.2"))
      )
    )
  ) %>%
  arrange(B_like_annotation_label)
```

```{r downsample_barcodes}
# downsample cell barcodes to equal number per celltype
downsample_cells <- function(df, max_cells) {
  set.seed(123) # Set seed for reproducibility
  num_patients <- n_distinct(df$label)
  cells_per_patient <- max_cells %/% num_patients # integer division to get the cells per patient
  remaining_cells <- max_cells %% num_patients # remaining cells after even distribution

  # Sample cells per patient with even distribution
  downsampled <- df %>%
    group_by(label) %>%
    sample_n(size = min(cells_per_patient, n()), replace = FALSE) %>%
    ungroup()

  # If there are remaining cells, sample them from the already downsampled data
  if (remaining_cells > 0) {
    remaining <- df %>%
      anti_join(downsampled, by = "label_cellbarcode") %>%
      sample_n(size = min(remaining_cells, n()), replace = FALSE)

    downsampled <- bind_rows(downsampled, remaining)
  }

  return(downsampled)
}

# downsampling number for healthy B is so high, that really all cells are considered and no downsampling is performed in this case
celltype_barcodes <- map2(
  c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"), c(200000, 4357, 4295, 4295),
  ~ barcodes_ordered %>%
    dplyr::filter(B_like_annotation == .x) %>%
    downsample_cells(max_cells = .y)
) %>%
  set_names(c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))
```

```{r cluster_patients}
# per patient calculate mean value per TF and run hierarchical clustering
clustering_input <- celltype_barcodes %>%
  map(~ zscore_matrix_combat_hm_input[comp_genes, .$label_cellbarcode] %>%
    as_tibble(rownames = NA) %>%
    tibble::rownames_to_column("TF") %>%
    pivot_longer(cols = -TF, names_to = "label_cellbarcode", values_to = "z_score") %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#") %>%
    dplyr::group_by(TF, label) %>%
    dplyr::summarise("mean_z_score" = mean(`z_score`)) %>%
    ungroup() %>%
    pivot_wider(id_cols = label, names_from = TF, values_from = mean_z_score) %>%
    column_to_rownames(var = "label")) %>%
  set_names(names(celltype_barcodes))

set.seed(20)
distance_mat <- clustering_input %>%
  map(~ dist(., method = "euclidean"))
clusters <- distance_mat %>%
  map(~ hclust(., method = "average"))

patients_ordered <- clustering_input %>%
  map2(clusters, ~ rownames(.x)[.y %>%
    as.dendrogram() %>%
    order.dendrogram()])
```

```{r column2_annotations}
celltype_barcodes_ordered <- celltype_barcodes %>%
  map2(patients_ordered, ~ dplyr::mutate(.x, label = factor(label, levels = .y)) %>%
    arrange(label))

barcodes2annotation <- celltype_barcodes %>%
  map(~ dplyr::select(., label_cellbarcode, B_like_annotation) %>%
    dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
    deframe())

barcodes2label <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, tibble(
    "label_cellbarcode" = colnames(zscore_matrix_combat_hm_input[, .x$label_cellbarcode]),
    "label" = colnames(zscore_matrix_combat_hm_input[, .x$label_cellbarcode]) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)
  )) %>%
    dplyr::select(label_cellbarcode, label) %>%
    dplyr::mutate(label = factor(label, levels = .y)) %>%
    deframe())

barcodes2patient <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, tibble(
    "label_cellbarcode" = colnames(zscore_matrix_combat_hm_input[, .x$label_cellbarcode]),
    "label" = colnames(zscore_matrix_combat_hm_input[, .x$label_cellbarcode]) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)
  )) %>%
    left_join(metadata %>%
      dplyr::filter(data_type == "ATAC") %>%
      dplyr::select("label", "patient_id")) %>%
    dplyr::select(label_cellbarcode, patient_id) %>%
    deframe())

barcodes2IGHV <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "IGHV")) %>%
    dplyr::mutate(
      label = factor(label, levels = .y),
      IGHV = if_else(is.na(IGHV), "NA", IGHV)
    ) %>%
    dplyr::select(label_cellbarcode, IGHV) %>%
    deframe())

barcode2processing_date <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "processing_date")) %>%
    dplyr::mutate(
      label = factor(label, levels = .y),
      processing_date = as.character(processing_date)
    ) %>%
    dplyr::select(label_cellbarcode, processing_date) %>%
    deframe())

barcode2technology <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "technology")) %>%
    dplyr::mutate(label = factor(label, levels = .y)) %>%
    dplyr::select(label_cellbarcode, technology) %>%
    deframe())

barcode2sequencing_date <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "sequencing_date")) %>%
    dplyr::mutate(
      label = factor(label, levels = .y),
      sequencing_date = as.character(sequencing_date)
    ) %>%
    dplyr::select(label_cellbarcode, sequencing_date) %>%
    deframe())

barcode2sex <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "sex")) %>%
    dplyr::mutate(label = factor(label, levels = .y)) %>%
    dplyr::select(label_cellbarcode, sex) %>%
    deframe())

"barcodes2sequencing_date_?" <- celltype_barcodes %>%
  map2(patients_ordered, ~ left_join(.x, metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "sequencing_date_?")) %>%
    dplyr::mutate(
      label = factor(label, levels = .y),
      sequencing_date = as.character(`sequencing_date_?`)
    ) %>%
    dplyr::select(label_cellbarcode, `sequencing_date_?`) %>%
    deframe())

colours <- list(
  "Cell type" = c("Physiologic B Cells" = "darkred", "LC-MBL Cells" = "#fcbda4", "HC-MBL Cells" = "#FC8D62", "CLL Cells" = "#66C2A5"),
  "Patient" = my_unpaired_colours_named,
  "IGHV" = c("mutated" = "steelblue", "unmutated" = "orange", "NA" = "lightgrey"),
  "processing_date" = c(
    "2022-01-10" = "#1B9E77", "2022-10-20" = "#D95F02",
    "2022-11-04" = "#7570B3", "2023-03-24" = "#66A61E",
    "2024-03-05" = "#A6761D"
  ),
  "technology" = c("v2 scATAC-seq" = "#666666", "multiome kit" = "#E7298A"),
  "sequencing_date" = c("2022-02-26" = "#E6AB02", "2022-11-18" = "#7FC97F", "2023-05-02" = "#FFFF99", "2024-03-18" = "#386CB0"),
  "sequencing_date_?" = c("2022-01-14" = "midnightblue", "2022-10-20" = "steelblue4", "2022-10-26" = "magenta4", "2022-10-27" = "purple4", "2024-03-18" = "#386CB0", "2023-05-02" = "violet", "2022-11-18" = "plum4"),
  "sex" = c("male" = "lightblue", "female" = "lightpink")
)

anno_df <- list(
  barcodes2annotation,
  barcodes2patient,
  barcodes2IGHV,
  barcode2processing_date,
  barcode2technology,
  barcode2sequencing_date,
  `barcodes2sequencing_date_?`,
  barcode2sex,
  celltype_barcodes
) %>%
  pmap(~ tibble(
    "Cell type" = ..1[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "Patient" = ..2[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "IGHV" = ..3[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    # "processing_date"=..4[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    # "technology"=..5[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    # "sequencing_date"=..6[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    # "sequencing_date_?"=..7[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    # "sex"=..8[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
  ))


column_ha <- anno_df %>%
  map(~ HeatmapAnnotation(
    df = .,
    col = colours,
    show_legend = FALSE,
    annotation_name_side = "left",
    border=TRUE,
    annotation_name_gp = gpar(fontsize = 8)
  ))
```

```{r, plot_heatmap}
hm <- list(
  celltype_barcodes,
  column_ha,
  celltype_barcodes_ordered %>% map(~ pull(., label_cellbarcode))
) %>%
  pmap(~ Heatmap(as.matrix(zscore_matrix_combat_hm_input[comp_genes, ..1$label_cellbarcode]),
    col = colorRamp2(breaks = c(
      -quantile(as.matrix(zscore_matrix_combat_hm_input[comp_genes, ]), 0.95),
      0,
      quantile(as.matrix(zscore_matrix_combat_hm_input[comp_genes, ]), 0.95)
    ), c("#0c544c", "white", "#783f0b")),
    top_annotation = ..2,
    column_order = ..3,
    use_raster = TRUE,
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    raster_by_magick = TRUE,
    show_column_names = FALSE,
    show_heatmap_legend = FALSE,
    na_col = "black",
    row_title_side = "left",
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 8),
    border_gp = gpar(col = "black", lty = 1),
    heatmap_legend_param = list(
      title = "Z-score",
      direction = "horizontal",
      legend_direction = "vertical",
      legend_gp = gpar(
        fontsize = 8,
        fontface = "plain"
      )
    )
  ))
```

```{r, fig.width=20, fig.height=9}
hm$`Physiologic B Cells` + hm$`LC-MBL Cells` + hm$`HC-MBL Cells` + hm$`CLL Cells`
hm_LC_CLL_grob_downsampled <- grid::grid.grabExpr(draw(hm$`Physiologic B Cells` + hm$`LC-MBL Cells` + hm$`HC-MBL Cells` + hm$`CLL Cells`))
cowplot::plot_grid(hm_LC_CLL_grob_downsampled)
saveRDS(hm_LC_CLL_grob_downsampled, file = file.path(project_dir, "RDS_Objects/TF_enrichment_B_LC_MBL_CLL_marker_heatmap_downsampled_batch_corrected_by_celltype.RDS"))
```

### Downsampled non-separated heatmap
```{r}
celltype_barcodes_unlist <- celltype_barcodes %>%
  bind_rows()

barcodes2annotation_unlist <- celltype_barcodes_unlist %>%
  dplyr::select(label_cellbarcode, B_like_annotation) %>%
  dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
  deframe()

barcodes2label_unlist <- celltype_barcodes_unlist %>%
  left_join(tibble(
    "label_cellbarcode" = colnames(zscore_matrix_combat_hm_input[, .$label_cellbarcode]),
    "label" = colnames(zscore_matrix_combat_hm_input[, .$label_cellbarcode]) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)
  )) %>%
  dplyr::select(label_cellbarcode, label) %>%
  deframe()


barcodes2IGHV_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "IGHV")) %>%
  dplyr::mutate(IGHV = if_else(is.na(IGHV), "NA", IGHV)) %>%
  dplyr::select(label_cellbarcode, IGHV) %>%
  deframe()

barcode2processing_date_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "processing_date")) %>%
  dplyr::mutate(
    processing_date = as.character(processing_date)
  ) %>%
  dplyr::select(label_cellbarcode, processing_date) %>%
  deframe()


barcode2technology_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "technology")) %>%
  dplyr::select(label_cellbarcode, technology) %>%
  deframe()

barcode2sequencing_date_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "sequencing_date")) %>%
  dplyr::mutate(
    sequencing_date = as.character(sequencing_date)
  ) %>%
  dplyr::select(label_cellbarcode, sequencing_date) %>%
  deframe()


barcode2sex_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "sex")) %>%
  dplyr::select(label_cellbarcode, sex) %>%
  deframe()

label2colours <- c(
  "CLL1.1" = "darkgoldenrod1", "CLL1.2" = "darkgoldenrod3",
  "CLL2.1" = "darkorange1", "CLL2.2" = "darkorange4",
  "CLL3.1" = "indianred1", "CLL3.2" = "indianred3",
  "CLL4.1" = "plum1", "CLL4.2" = "plum4",
  "CLL5.1" = "orchid", "CLL5.2" = "orchid4",
  "CLL6.1" = "cadetblue1", "CLL6.2" = "cadetblue4",
  "CLL7.1" = "turquoise", "CLL7.2" = "turquoise4",
  "CLL8.1" = "darkseagreen1", "CLL8.2" = "darkseagreen4",
  "CLL9.1" = "burlywood1", "CLL9.2" = "burlywood4",
  "CLL10.1" = "lightsalmon", "CLL10.2" = "lightsalmon4",
  "MBL1" = "darkslategrey",
  "MBL2" = "rosybrown3",
  "MBL3" = "thistle4",
  "MBL4" = "olivedrab",
  "MBL6" = "steelblue4",
  "MBL7" = "slateblue2",
  "MBL7dup" = "lavender",
  "MBL8" = "lightcoral",
  "MBL9" = "yellow3",
  "MBL10" = "lightgoldenrod1",
  "MBL11" = "antiquewhite2",
  "MBL12" = "azure4",
  "MBL13" = "aquamarine4",
  "MBL14" = "darkolivegreen1",
  "MBL15" = "lightsteelblue"
)

colours <- list(
  "Cell type" = c("Physiologic B Cells" = "darkred", "LC-MBL Cells" = "#fcbda4", "HC-MBL Cells" = "#FC8D62", "CLL Cells" = "#66C2A5"),
  "Patient" = label2colours,
  "IGHV" = c("mutated" = "steelblue", "unmutated" = "orange", "NA" = "lightgrey"),
  "processing_date" = c(
    "2022-01-10" = "#1B9E77", "2022-10-20" = "#D95F02",
    "2022-11-04" = "#7570B3", "2023-03-24" = "#66A61E",
    "2024-03-05" = "#A6761D"
  ),
  "technology" = c("v2 scATAC-seq" = "#666666", "multiome kit" = "#E7298A"),
  "sequencing_date" = c("2022-02-26" = "#E6AB02", "2022-11-18" = "#7FC97F", "2023-05-02" = "#FFFF99", "2024-03-18" = "#386CB0"),
  "sequencing_date_?" = c("2022-01-14" = "midnightblue", "2022-10-20" = "steelblue4", "2022-10-26" = "magenta4", "2022-10-27" = "purple4", "2024-03-18" = "#386CB0", "2023-05-02" = "violet", "2022-11-18" = "plum4"),
  "sex" = c("male" = "lightblue", "female" = "lightpink")
)

anno_df_unlist <- list(
  list(barcodes2annotation_unlist),
  list(barcodes2label_unlist),
  list(barcodes2IGHV_unlist),
  list(barcode2processing_date_unlist),
  list(barcode2technology_unlist),
  list(barcode2sequencing_date_unlist),
  list(barcode2sequencing_date_unlist),
  list(barcode2sex_unlist),
  list(celltype_barcodes_unlist)
) %>%
  pmap(~ tibble(
    "Cell type" = ..1[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "Patient" = ..2[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "IGHV" = ..3[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "processing_date" = ..4[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "technology" = ..5[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "sequencing_date" = ..6[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
    "sex" = ..8[colnames(zscore_matrix_combat_hm_input[, ..9$label_cellbarcode])],
  ))


column_ha_unlist <- anno_df_unlist %>%
  map(~ HeatmapAnnotation(
    df = .,
    col = colours,
    show_legend = FALSE,
    annotation_name_side = "left",
    border = TRUE,
    annotation_name_gp = gpar(fontsize = 8)
  ))
```

```{r}
hm_unlist <- list(
  list(celltype_barcodes_unlist),
  column_ha_unlist
) %>%
  pmap(~ Heatmap(as.matrix(zscore_matrix_combat_hm_input[comp_genes, ..1$label_cellbarcode]),
    col = colorRamp2(breaks = c(-quantile(as.matrix(zscore_matrix_combat_hm_input[comp_genes, ]), 0.98), 0, quantile(as.matrix(zscore_matrix_combat_hm_input[comp_genes, ]), 0.98)), c("#0c544c", "white", "#783f0b")),
    top_annotation = ..2,
    use_raster = TRUE,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    raster_by_magick = TRUE,
    show_column_names = FALSE,
    show_heatmap_legend = FALSE,
    na_col = "black",
    row_title_side = "left",
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 8),
    row_title_gp = gpar(fontsize = 8),
    border_gp = gpar(col = "black", lty = 1),
    heatmap_legend_param = list(
      title = "Z-score",
      direction = "horizontal",
      legend_direction = "vertical",
      legend_gp = gpar(
        fontsize = 8,
        fontface = "plain"
      )
    )
  ))
```

## DCA Scheme
```{r, fig.width=9, fig.height=4}
WBC_summary <- metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, patient_id, `WBC_[cells/nl]`, MBL_subtype) %>%
  dplyr::mutate(MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype)) %>%
  dplyr::select(patient_id, `WBC_[cells/nl]`, MBL_CLL_subtype) %>%
  group_by(MBL_CLL_subtype) %>%
  dplyr::summarise(
    mean_WBC = mean(`WBC_[cells/nl]`),
    median_WBC = median(`WBC_[cells/nl]`)
  )

LC_TFs <- volcano.data.all.comparisons[["B_vs_LC_MBL"]][[4]] %>%
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  str_split_i(pattern = "_", i = 1) %>%
  unlist()

HC_TFs <- volcano.data.all.comparisons[["B_vs_HC_MBL"]][1:3] %>%
  bind_rows() %>%
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  str_split_i(pattern = "_", i = 1) %>%
  unlist()

CLL_TFs <- volcano.data.all.comparisons[["B_vs_CLL"]][1:3] %>%
  bind_rows() %>%
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  str_split_i(pattern = "_", i = 1) %>%
  unlist()

barcodes <- colnames(zscore_matrix_combat_hm_input) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = CLLProj_Blike$cellNames,
    "B_like_annotation" = CLLProj_Blike@cellColData$B_like_annotation
  )) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
      B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
      B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
    label = factor(label, levels = c("MBL13", "MBL11", "MBL12", "MBL15", "MBL14", "MBL10", "CLL6.2", "CLL1.2", "CLL7.2", "CLL2.2", "CLL5.2", "CLL8.2", "CLL4.2", "CLL10.2", "CLL9.2", "CLL3.2", "MBL4"))
  ) %>%
  arrange(B_like_annotation, label) %>%
  left_join(tibble(
    "B_like_annotation" = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"),
    "Colour" = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5")
  ))

barcodes2annotation <- barcodes %>%
  dplyr::select(label_cellbarcode, B_like_annotation) %>%
  dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
  deframe()

deviation_summary_tmp <- zscore_matrix_combat_hm_input[comp_genes, ] %>%
  as_tibble(rownames = "TF") %>%
  pivot_longer(-TF, names_to = "label_cellbarcode", values_to = "deviation") %>%
  tidyr::separate(label_cellbarcode, into = c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(label, patient_id, MBL_subtype)) %>%
  dplyr::mutate(
    celltype = barcodes2annotation[label_cellbarcode],
    MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype)
  ) %>%
  dplyr::filter((MBL_CLL_subtype == "LC" & TF %in% LC_TFs) |
    (MBL_CLL_subtype == "HC" & TF %in% HC_TFs) |
    (MBL_CLL_subtype == "CLL" & TF %in% CLL_TFs)) %>%
  dplyr::group_by(TF, celltype) %>%
  dplyr::summarise(
    mean_deviation = mean(deviation),
    median_deviation = median(deviation)
  ) %>%
  dplyr::mutate(direction = case_when(mean_deviation & median_deviation >= 0 ~ "positive",
    mean_deviation & median_deviation <= 0 ~ "negative",
    .default = "ambiguous"
  )) %>%
  ungroup() %>%
  group_by(celltype, direction) %>%
  dplyr::summarise(
    celltype_mean_deviation = mean(mean_deviation),
    celltype_median_deviation = median(median_deviation)
  )


deviation_summary <- deviation_summary_tmp %>%
  group_by(direction) %>%
  filter(celltype != "Physiologic B Cells") %>%
  # Join with Physiologic B Cells on the same direction
  left_join(
    deviation_summary_tmp %>%
      filter(celltype == "Physiologic B Cells"),
    by = "direction",
    suffix = c("_current", "_reference")
  ) %>%
  # Calculate ratios
  dplyr::mutate( # mean_deviation_ratio = celltype_mean_deviation_current / celltype_mean_deviation_reference,
    mean_deviation_difference = celltype_mean_deviation_reference - celltype_mean_deviation_current,
    # median_deviation_ratio = celltype_median_deviation_current / celltype_median_deviation_reference,
    median_deviation_difference = celltype_median_deviation_reference - celltype_median_deviation_current
  ) %>%
  # Select relevant columns for output
  dplyr::select(
    celltype_current, direction, # mean_deviation_ratio, median_deviation_ratio,
    mean_deviation_difference, median_deviation_difference
  )

scheme_input <- deviation_summary %>%
  dplyr::mutate(MBL_CLL_subtype = case_when(
    celltype_current == "LC-MBL Cells" ~ "LC",
    celltype_current == "HC-MBL Cells" ~ "HC",
    celltype_current == "CLL Cells" ~ "CLL"
  )) %>%
  left_join(WBC_summary) %>%
  bind_rows(tibble(
    celltype_current = c("Physiologic B Cells", "Physiologic B Cells", "Physiologic B Cells", "Physiologic B Cells"),
    MBL_CLL_subtype = c(NA, NA, NA, NA),
    mean_WBC = c(0, 0, 5, 5),
    median_WBC = c(0, 0, 5, 5),
    mean_deviation_difference = c(0, 0, 0, 0),
    median_deviation_difference = c(0, 0, 0, 0),
    direction = c("positive", "negative", "positive", "negative")
  ))

rect_data <- tibble(
  ymin = rep(-2, 4),
  ymax = rep(2, 4),
  xmin = c(0, 5.48, 9.96, 18.35),
  xmax = c(5.48, 9.96, 18.35, 25)
)

scheme <- scheme_input %>%
  ggplot(aes(x = median_WBC, y = median_deviation_difference, group = direction)) +
  geom_rect(
    data = rect_data,
    mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    alpha = 0.2,
    fill = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5"),
    inherit.aes = FALSE
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "lightgrey ") +
  geom_line(colour = "indianred", linewidth = 1.5) +
  geom_point(colour = "indianred", size = 3) +
  scale_y_continuous(limits = c(-2, 2), expand = expansion(mult = c(0, 0))) +
  scale_x_continuous(limits = c(0, 25), expand = expansion(mult = c(0, 0))) +
  labs(title = "TF motif enrichment", x = "Median WBCC [cells/ml]", y = "Median TF motif enrichment change") +
  theme_pub() +
  theme(plot.title = element_text(size = 10))

write_csv(scheme_input, file = file.path(tsv_dir, "TF_motif_scheme_input.csv"))
saveRDS(object = scheme, file = file.path(RDS_dir, "TF_motif_scheme.RDS"))
```

## TF motif enrichment violins
```{r fig.width=8, fig.height=3}
# LC-MBL B cells vs. LC-MBL cells
comp_genes_raw_LC <- volcano.data.all.comparisons[["B_vs_LC_MBL"]] %>% bind_rows() %>% 
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  map_chr(~ .[[1]]) %>%
  unique() %>%
  sort()

enrichment_LC_violinplot <- bind_rows(result[comp_genes_raw_LC, B_cells_Blike[["4"]]] %>% 
  as.tibble(rownames="gene") %>% 
  pivot_longer(-gene, names_to = "label_cellbarcode", values_to = "enrichment") %>% mutate(celltype="B"),
result[comp_genes_raw_LC, LC_MBL_cells_Blike[["4"]]] %>% 
  as.tibble(rownames="gene") %>% 
  pivot_longer(-gene, names_to = "label_cellbarcode", values_to = "enrichment") %>% mutate(celltype="LC-MBL")) %>% 
  ggplot(aes(x=  str_split(string =gene , pattern = "_") %>%
  map_chr(~ .[[1]]) , y=enrichment, fill=celltype))+
  geom_violin(position=position_dodge(1))+
  geom_boxplot(aes(group = interaction(celltype, gene)), 
               colour="black",fill = NA, 
               position=position_dodge(1),width=0.2, outlier.shape = NA )+
  labs(x="TF", y="Enrichment over background peaks", fill="Cell type")+
  scale_fill_manual(values=my_MBL_CLL_colours_named)+
  theme_pub()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))

write_tsv(enrichment_LC_violinplot$data, file = file.path(tsv_dir, "enrichment_LC_violinplot_data.tsv"))
```

```{r fig.width=8, fig.height=3}
# HC-MBL/CLL B cells vs. HC-MBL cells
comp_genes_raw_HC <- volcano.data.all.comparisons[["B_vs_HC_MBL"]] %>% bind_rows() %>% 
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  map_chr(~ .[[1]]) %>%
  unique() %>%
  sort()

enrichment_HC_violinplot <- bind_rows(result[comp_genes_raw_HC, c(unlist(B_cells_Blike[1:3]))] %>% 
  as.tibble(rownames="gene") %>% 
  pivot_longer(-gene, names_to = "label_cellbarcode", values_to = "enrichment") %>% mutate(celltype="B"),
result[comp_genes_raw_HC, c(unlist(HC_MBL_cells_Blike))] %>% 
  as.tibble(rownames="gene") %>% 
  pivot_longer(-gene, names_to = "label_cellbarcode", values_to = "enrichment") %>% mutate(celltype="HC-MBL")) %>% 
  ggplot(aes(x=  str_split(string =gene , pattern = "_") %>%
  map_chr(~ .[[1]]) , y=enrichment, fill=celltype))+
  geom_violin(position=position_dodge(1))+
  geom_boxplot(aes(group = interaction(celltype, gene)), 
               colour="black",fill = NA, 
               position=position_dodge(1),width=0.2, outlier.shape = NA )+
  labs(x="TF", y="Enrichment over background peaks", fill="Cell type")+
  scale_fill_manual(values=my_MBL_CLL_colours_named)+
  theme_pub()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))

write_tsv(enrichment_HC_violinplot$data, file = file.path(tsv_dir, "enrichment_HC_violinplot_data.tsv"))
```

```{r fig.width=8, fig.height=3}
# HC-MBL/CLL B cells vs. CLL cells
comp_genes_raw_CLL <- volcano.data.all.comparisons[["B_vs_CLL"]] %>% bind_rows() %>% 
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  map_chr(~ .[[1]]) %>%
  unique() %>%
  sort()

enrichment_CLL_violinplot <- bind_rows(result[comp_genes_raw_CLL, c(unlist(B_cells_Blike[1:3]))] %>% 
  as.tibble(rownames="gene") %>% 
  pivot_longer(-gene, names_to = "label_cellbarcode", values_to = "enrichment") %>% mutate(celltype="B"),
result[comp_genes_raw_CLL, c(unlist(CLL_cells_Blike))] %>% 
  as.tibble(rownames="gene") %>% 
  pivot_longer(-gene, names_to = "label_cellbarcode", values_to = "enrichment") %>% mutate(celltype="CLL")) %>% 
  ggplot(aes(x=  str_split(string =gene , pattern = "_") %>%
  map_chr(~ .[[1]]) , y=enrichment, fill=celltype))+
  geom_violin(position=position_dodge(1))+
  geom_boxplot(aes(group = interaction(celltype, gene)), 
               colour="black",fill = NA, 
               position=position_dodge(1),width=0.2, outlier.shape = NA )+
  labs(x="TF", y="Enrichment over background peaks", fill="Cell type")+
  scale_fill_manual(values=my_MBL_CLL_colours_named)+
  theme_pub()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))

write_tsv(enrichment_CLL_violinplot$data, file = file.path(tsv_dir, "enrichment_CLL_violinplot_data.tsv"))
```

