---
title: "Mitochondrial_mutations_MBL.Rmd"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

#GOAL
The aim of this script is to extract mitochondrial mutations for all individual MBL samples and characterize clonotypes and their dynamics within the cohort

# Load Libraries
```{r Load_libraries}
library(Signac)
library(readxl)
library(ArchR)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(ggrepel)
library(cluster)
library(Seurat)
library(cowplot)
```


# Set paths
```{r}
source("config_paths.R")

data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "ATAC")
result_dir <- file.path(project_dir, "results")

ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")
ArchR_dir <- paste0("ArchR_Subset", 1:4) %>% as.list()
result_dir_subsets <- file.path(ArchR_dir_Annotation, ArchR_dir) %>% setNames(paste0("Subset", 1:4))
RDS_dir_subsets <- file.path(result_dir_subsets, "RDS_objects") %>% setNames(paste0("Subset", 1:4))
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r}
metadata <- read_tsv(file = file.path(data_dir, "20240530_Tidy_MBL_CLL_metadata.tsv"))
```

## Load ArchRProj
```{r}
CLLProj <- result_dir_subsets %>% 
  map(~loadArchRProject(path = paste0(.x, "/"))) %>% 
  purrr::set_names(paste0("Subset", 1:4))
```

## Load Samples
```{r}
sample_subset <- list("Subset1" = c("CLL1.1","CLL1.2", "CLL2.1","CLL2.2","CLL3.1", "CLL3.2","MBL1","MBL2","MBL3"),
                      "Subset2" = c("CLL4.1","CLL4.2","CLL5.1","CLL5.2","CLL6.1","CLL6.2","MBL4","MBL6","MBL7"),
  "Subset3" = c("CLL7.1","CLL7.2","CLL8.1","CLL8.2","CLL9.1","CLL9.2","CLL10.1","CLL10.2","MBL9","MBL8"),
  "Subset4" = c("MBL7dup","MBL10","MBL11","MBL12","MBL13","MBL14","MBL15"))
```

## MGATK Files
```{r Unpaired_MBL_files}
mito_data_idv <- RDS_dir_subsets %>% 
  map(~readRDS(file.path(.x, "mito_data_idv.RDS")))

# Unpaired mgatk files per patient
sample_subset_individual <- map(sample_subset, ~dplyr::filter(metadata, label %in% .x & str_starts(label, "MBL")) %>%
  pull(patient_id) %>%
  unique())
```

```{r}
# New filtering function also accounting for mtseq depth of min 10!
filter_qc_from_mitodata <- function(mgatk_idv_list, archrproj) {
  mito_subset <- list()

  # Get cell barcodes from singlets identified from ATACdata
  label_cellbarcode_df <- archrproj@cellColData@rownames %>%
    str_split_fixed("#", n = 2) %>%
    `colnames<-`(c("label", "cellbarcode")) %>%
    as_tibble()

  # Filter cells
  for (patient in unique(label_cellbarcode_df$label)) {
    barcodes <- label_cellbarcode_df %>%
      dplyr::filter(label == patient) %>%
      pull(cellbarcode)

    # Adapt here for mtseq depth
    keep_col_doublets <- mgatk_idv_list[[patient]]$counts@Dimnames[[2]] %in% barcodes
    keep_col_depth <- mgatk_idv_list[[patient]]$depth %>%
      dplyr::filter(mito.depth >= 10) %>%
      rownames()
    keep_col <- intersect(mgatk_idv_list[[patient]]$counts@Dimnames[[2]][keep_col_doublets], keep_col_depth)

    mgatk_idv_list[[patient]]$counts <- mgatk_idv_list[[patient]]$counts[, keep_col]

    mgatk_idv_list[[patient]]$depth <- mgatk_idv_list[[patient]]$depth[keep_col, , drop = FALSE]
  }

  return(mgatk_idv_list)
}

mito.data.idv <- map2(.x = mito.data.idv,
                      .y = CLLProj,
                      .f = ~filter_qc_from_mitodata(.x, .y))

mito_data_idv %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "mito.data.idv_new.RDS")))
```

```{r}
mito.data.MBL <- list()

for ( i in 1:4){ 
for (i in sample_subset_individual[[i]]) {
  subset_names <- str_subset(names(mito.data.idv), i)
  subset_names_index <- match(subset_names, names(mito.data.idv))

  df_1 <- mito.data.idv[[subset_names_index[[1]]]]$counts
  colnames(df_1) <- paste(subset_names[1], colnames(df_1), sep = "#")
  mito.data.MBL[[i]]<- append(mito.data.MB[[i]], df_1)
}
  names(mito.data.MBL[[i]]) <- sample_subset_individual[[i]]
}  
  
mito.data.MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "mito.data.MBL.RDS")))

mito.data.MBL <- RDS_dir_subsets %>% 
  map(~readRDS(file.path(.x, "mito.data.MBL.RDS")))
```

# Analysis
## Variant Calling for unpaired samples
```{r}
variant.df.MBL <- list()
for (i in 1:4){
variant.df.MBL[[names(sample_subset_individual[i])]] <- mito.data.MBL[[names(sample_subset_individual[i])]] %>%
  map(~ IdentifyVariants(., refallele = mito.data.MBL[[names(sample_subset_individual[i])]][[1]]$refallele)) %>%
  imap(~ .x %>%
    dplyr::mutate(sample_id = .y, .before = "nucleotide"))
}

for (i in 1:4){
  variant.df.MBL [[names(sample_subset_individual[i])]] %>%
  imap(~ write_tsv(.x,
    quote = "none",
    file = file.path(result_dir_subsets, "MitochondrialVariants", paste0("Variants_", .y, ".csv"))
  ))
}

variant.df.MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "variant.df.MBL")))
```

```{r}
# Plot all variants
all.variant.df.plots <- list()
  
for (i in 1:4){
  variant.df.MBL[[i]] %>%
 map(VariantPlot)%>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[i]], "MitochondrialVariants", paste0("Plot_Variance_Mean_Ratio_", .y, ".png")),
    plot = .x,
    width = 5,
    height = 3
  ))
}
```

## Maternal variants
### Filter maternal variants

```{r}
mat.variants <- list()

for (i in 1:3){
mat.variants[[names(sample_subset_paired[i])]] <- variant.df.MBL[[i]] %>%
  map(~ subset(.,
  subset = n_cells_conf_detected >= 5 & strand_correlation >= 0.65 & vmr < 0.01
))
}

mat.variants %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "mat.variants.RDS")))

mat.variants <- RDS_dir_subsets[1:4] %>% 
  map(~readRDS(file.path(.x, "mat.variants.RDS")))
```

### Get maternal VAFs

```{r}
# Get maternal VAFs
# VAFs.mat.MBL <- list()
# 
# for (i in 1:3){
# VAFs.mat.MBL[[names(sample_subset_paired[i])]] <- mito.data.MBL %>%
#   map2(mat.variants.MBL, ~ AlleleFreq(.,
#     variants = .y$variant,
#     assay = NULL
#   ))
# }
# 
# VAFs.mat.MBL %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "VAFs.mat.MBL.RDS")))

VAFs.mat.MBL <-file.path(RDS_dir_subsets[1:4], "VAFs.mat.MBL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset[1:4]))
```

### Project VAFs onto UMAP Plot
```{r}
umap_embedding <- CLLProj %>% 
                        map(~.x@embeddings$UMAP$df %>%
  rownames_to_column("id") %>%
  separate(id, c("label", "cell_barcode"), sep = "#") %>%
  separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE))
```


```{r}
# Inner join is used to consider only cells which have been used in the umap (here doublets are removed!)
umap_embedding_mito_mat_MBL <- list()
for (i in 1:3){
umap_embedding_mito_mat_MBL[[names(sample_subset_paired[i])]] <- map_dfr(VAFs.mat.MBL[[i]], ~ as.matrix(.) %>%
  as_tibble(rownames = "variant_id") %>%
  pivot_longer(-variant_id,
    names_to = "cell_barcode",
    values_to = "VAF"
  ), .id = "patient_id") %>%
  separate(cell_barcode, c("label", "cell_barcode"), sep = "#") %>%
  inner_join(umap_embedding[[names(sample_subset_paired[i])]], by = join_by(patient_id, label, cell_barcode)) %>%
  dplyr::rename(
    UMAP1 = "IterativeLSI#UMAP_Dimension_1",
    UMAP2 = "IterativeLSI#UMAP_Dimension_2"
  )
}

umap_embedding_mito_mat_MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "umap_embedding_mito_mat_MBL.RDS")))

umap_embedding_mito_mat_MBL <- file.path(RDS_dir_subsets[1:4], "umap_embedding_mito_mat_MBL.RDS")%>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset[1:4]))

```

```{r, fig.width=20, fig.height=10}
# Careful! group_by internally orders column before split which causes different order in split list
umap_embedding_mito_mat_plot_MBL <- list()

for (i in 1:3){
umap_embedding_mito_mat_plot_MBL[[names(sample_subset_individual[[i]])]] <- umap_embedding_mito_mat_plot_MBL[[i]] %>%
  split(.$patient_id) %>% 
  imap(~
    arrange(., label, -VAF) %>%
      dplyr::mutate(VAF=if_else(VAF==0, NA, VAF)) %>% 
      ggplot(aes(x = UMAP1, y = UMAP2, color = VAF)) +
      geom_point(size = 0.5, alpha = 0.2, stroke = NA) +
      coord_equal() +
      scale_colour_gradient(
        name = "Heteroplasmy",
        low = "yellow",
        high = "darkred",
        trans = "log10",
        breaks=10^seq(from=-6, to=0, by=2),
        limits = c(1e-06, 1e+00),
        oob=scales::oob_squish,
        labels=c("<=1e-06",  "1e-04", "0.01",  "1")) +
      theme_pub() +
      labs(title = .y) +
      theme(
        strip.background = element_rect(
          fill = "white",
          colour = NA),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.ticks = element_blank()
      ) +
      facet_wrap(~variant_id, ncol = 10)) %>%
  
iwalk(~ ggsave(
  filename = file.path(result_dir_subsets[[i]], paste0("Plots/UMAPs_maternal_mito_VAFs/Plot-UMAP-maternal-mitochondrial-VAF_", .y, ".png")),
  plot = .x,
  width = 20,
  height = 10))
}

umap_embedding_mito_mat_plot_MBL
```

## Clonal variants
### Filter clonal variants

```{r}
cln.variants.MBL <- list()
for (i in 1:4){
cln.variants.MBL[[names(sample_subset_individual[i])]] <- variant.df.MBL[[i]] %>%
  map(~ subset(., subset = n_cells_conf_detected >= 5 & strand_correlation >= 0.65 & vmr > 0.01))
}

# Export_cln_mito_variants
for (i in 1:4){
  cln.variants.MBL[[names(sample_subset_individual[i])]] %>%
  imap(~ write_tsv(.x,
    quote = "none",
    file = file.path(result_dir_subsets[[names(sample_subset_individual[i])]], "MitochondrialVariants", paste0("Clonal_Variants_", .y, ".csv"))
  ))
}

cln.variants.MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "cln.variants.MBL.RDS")))

cln.variants.MBL <- file.path(RDS_dir_subsets[1:4], "cln.variants.MBL.RDS")%>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_individual[1:4]))
```

### Get clonal VAFs
```{r}
# Get clonal VAFs
VAFs.cln.MBL <- list()

for (i in 1:3){
VAFs.cln.MBL[[names(sample_subset_individual[i])]] <- mito.data.MBL %>%
  map2(cln.variants.MBL, ~ AlleleFreq(.,
    variants = .y$variant,
    assay = NULL
  ))
}

VAFs.cln.MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "VAFs.cln.MBL.RDS")))

VAFs.cln.MBL <-file.path(RDS_dir_subsets[1:3], "VAFs.cln.MBL.RDS") %>%
  map(~readRDS(.x)) %>%
  purrr::set_names(names(sample_subset_individual[1:3]))

# export in sparse format
for (i in 1:3){
  VAFs.cln.MBL [[names(sample_subset_individual[i])]] %>%
  imap(~ write_tsv(.x,
    quote = "none",
    file = file.path(result_dir_subsets, "MitochondrialVariants", paste0("VAFs_Variants_sparse_", .y, "mtx"))
  ))
}

VAFs.cln.MBL <-file.path(RDS_dir_subsets[1:4], "VAFs.cln.MBL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:4]))
```

### UMAP VAF Plot
```{r}
# Inner join is used to consider only cells which have been used in the umap (here doublets are removed!)
umap_embedding_mito_cln_MBL <- list()
for (i in 1:4){
umap_embedding_mito_cln_MBL[[names(sample_subset_individual[i])]]  <- map_dfr(VAFs.cln.MBL[[i]], ~ as.matrix(.) %>%
  as_tibble(rownames = "variant_id") %>%
  pivot_longer(-variant_id,
    names_to = "cell_barcode",
    values_to = "VAF"), .id = "patient_id") %>%
  separate(cell_barcode, c("label", "cell_barcode"), sep = "#") %>%
  arrange(label, -VAF) %>%
  inner_join(umap_embedding[[names(sample_subset_individual[i])]]) %>%
  dplyr::rename(
    UMAP1 = "IterativeLSI#UMAP_Dimension_1",
    UMAP2 = "IterativeLSI#UMAP_Dimension_2"
  )
}

umap_embedding_mito_cln_MBL %>%  
   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "umap_embedding_mito_cln_MBL.RDS")))


umap_embedding_mito_cln_MBL <-file.path(RDS_dir_subsets[1:4], "umap_embedding_mito_cln_MBL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:4]))

for (i in 1:4){
  umap_embedding_mito_cln_MBL[[i]] %>%
  split(.$patient_id) %>%
  imap(~ write_tsv(.x,
    file = file.path(result_dir_subsets[[i]], "MitochondrialVariants", paste0("VAFs_Variants_UMAP_MBL_", .y, ".tsv"))))
}
```

```{r, fig.width=20, fig.height=10}
umap_embedding_mito_cln_MBL_plot <- list()

for (i in 1:4){
umap_embedding_mito_cln_MBL_plot[[names(sample_subset_individual[i])]] <- umap_embedding_mito_cln_MBL[[i]] %>%
  split(.$patient_id) %>% 
  imap(~ arrange(., label, VAF) %>%
    dplyr::mutate(VAF = if_else(VAF == 0, NA, VAF)) %>%
    ggplot(aes(x = UMAP1, y = UMAP2, color = VAF)) +
    geom_point(size = 0.5, alpha = 0.8, stroke = NA) +
    coord_equal() +
    scale_colour_gradient(
        name = "Heteroplasmy",
        low = "yellow",
        high = "darkred",
        na.value = "lightgrey",
        trans = "log10",
        breaks=10^seq(from=-6, to=0, by=2),
        limits = c(1e-06, 1e+00),
        oob=scales::oob_squish,
        labels=c("<=1e-06",  "1e-04", "0.01",  "1")) +
    theme_pub() +
    theme(
      strip.text = element_text(face = "bold", size = rel(1.2)),
      strip.background = element_rect(fill = "white", colour = NA),
      panel.border = element_blank(),
      axis.line.x = element_line(),
      axis.line.y = element_line(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    ) +
    facet_wrap(~variant_id, ncol=10)) %>%
  
iwalk(~ ggsave(
  filename = paste0(result_dir_subsets[[i]], "Plots/UMAPs_clonal_mito_VAFs/Plot-UMAP-clonal-mitochondrial-VAF_", .y, ".png"),
  plot = .x,
  width =20,
  height = 10
))
}

umap_embedding_mito_cln_MBL_plot
```

### Heatmap VAFs

```{r}
# Prepare heatmap VAF input data as matrix with only VAFs unequal 0
VAF_mat_MBL <- list()

for (i in 1:3){
VAF_mat_MBL[[names(sample_subset_individual[i])]] <- umap_embedding_mito_cln_MBL[[i]] %>%
  dplyr::filter(!(variant_id == "310T>C")) %>%
  split(.$patient_id) %>%
  map(~tidyr::unite(., "label_cell_barcode", label:cell_barcode, sep = "#", remove = FALSE) %>%
      dplyr::select(label_cell_barcode, VAF, variant_id) %>%
      tidyr::pivot_wider(values_from = VAF, names_from = label_cell_barcode) %>%
      column_to_rownames("variant_id") %>%
      drop_na() %>%
      dplyr::select(-where(~ all(. == 0))) %>%
      as.matrix())
}

VAF_mat_MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "VAF_mat_MBL.RDS")))


VAF_mat_MBL <-file.path(RDS_dir_subsets[1:4], "VAF_mat_MBL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_individual[1:4]))
```

```{r}
# Prepare annotation labels
column_annotation <- list()
col_annot <- list()
for (i in 1:4){
column_annotation[[names(sample_subset_individual[i])]] <- CLLProj[[i]]@cellColData %>%
  as.data.frame() %>%
  dplyr::select(Clusters_0.6_IterativeLSI) %>%
  dplyr::mutate(Label = rownames(.) %>% str_split_fixed("#", 2) %>% .[, 1]) %>%
  tidyr::separate(Label, c("patient_id", "sample"), remove = FALSE) %>%
  split(.$patient_id) %>%
  map(~ dplyr::select(., -c("sample", "patient_id")))

# Select only paired samples
column_annotation[[names(sample_subset_individual[i])]] <- column_annotation[[names(sample_subset_individual[i])]][grepl("^MBL", names(column_annotation[[names(sample_subset_individual[i])]]))]

# Prepare annotation colours
col_annot[[names(sample_subset_individual[i])]] <- column_annotation[[names(sample_subset_individual[i])]][[1]]
}
```


#### Seurat Find Clonotypes Function
```{r}
#Cluster together per patient and filter for non-zero VAF cells
mito_VAFs_MBL <- list()
clonotypes_MBL <- list()

for (i in 1:4){
mito_VAFs_MBL[[names(sample_subset_individual[i])]] <- VAFs.cln.MBL[[i]] %>%
  map(~ as.matrix(.x) %>%
  as_tibble(rownames = "variant_id") %>%
  dplyr::filter(variant_id != "310T>C") %>%
  column_to_rownames(var = "variant_id") %>%
  .[, as.logical(colSums(. != 0))] %>%
  as.matrix() %>%
  as("sparseMatrix") %>%
  CreateSeuratObject(assay = "alleles"))

clonotypes_MBL[[names(sample_subset_individual[i])]] <- mito_VAFs_MBL[[names(sample_subset_individual[i])]] %>%
  map(~ FindClonotypes(.))

}

clonotypes_MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "clonotypes_MBL.RDS")))

find_clonotype_clusters <- list()
for (i in 1:4){
find_clonotype_clusters[[names(sample_subset_individual[i])]] <- clonotypes_MBL[[i]] %>%
  map(~ .@active.ident %>%
    data.frame() %>%
    rownames_to_column("label_cellbarcode") %>%
    dplyr::rename(Cluster = ".")) %>%
  imap(~write_tsv(.x, file = file.path(result_dir, "MitochondrialVariants", paste0("FindClonotypesClusters_", .y, ".tsv"))))
}

clonotypes_MBL <- file.path(RDS_dir_subsets[1:4], "clonotypes_MBL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_individual[1:4]))
```

```{r fig.width=30, fig.height=6}
MBL.clonotypes.plot <- list()

for (i in 1:3){

MBL_clonotypes.plot[[names(sample_subset_individual[i])]] <- clonotypes_MBL[[i]] %>%
  imap(~ DoHeatmap(.x,
    features = VariableFeatures(.),
    slot = "data",
    group.colors = get_palette(palette = "Set3", k = length(unique(.@meta.data[["seurat_clusters"]])))
  ) +
    scale_fill_distiller(
      palette = "YlOrRd",
      direction = 1,
      trans = "log",
      breaks = 10^c(-12:0),
      name = "VAF",
      na.value = "#082567"
    ) +
    labs(title = .y) +
    theme(
      plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
      axis.text = element_text(face = "bold")
    ) +
    guides(colour = "none")) %>%
  iwalk(~ ggsave(
    filename = paste0(result_dir_subsets[[names(sample_subset_individual[[i]])]], "/Plots/VAF_Heatmap_Find_Clonotype/VAF_Heatmap_Find_Clonotype_", .y, ".png"),
    plot = .x,
    width = 15,
    height = 10
  ))
}
```

##### Quantify clusters
```{r}
# Count how many cells are within each cluster per disease condition
clonotype_cell_counts_MBL <- list()

for (i in 1:4){
clonotype_cell_counts_MBL[[names(sample_subset_individual[[i]])]] <- find_clonotype_clusters[[i]] %>%
  map(~ tidyr::separate(., "label_cellbarcode", c("label", "cell_barcode"), sep = "#") %>%
    split(.$label) %>%
    imap( ~ group_by(.x, Cluster) %>%
      dplyr::summarise("cell counts" = n()) %>%
      mutate(
        "Clonotype frequency [% cells]" = round(`cell counts` / sum(`cell counts`) * 100, 3),
        "label" = .y
      ) %>%
      relocate(label, .before = (Cluster))) %>%
    bind_rows())
}

clonotype_cell_counts_MBL %>%
  imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "clonotype_cell_counts_MBL.RDS")))

clonotype_cell_counts_MBL <- file.path(RDS_dir_subsets[1:4], "clonotype_cell_counts_MBL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_individual[1:4]))
```

```{r, fig.width=16, fig.height=3}
quantify_cluster_histogram <- function(cluster_counts, sample_name) {
  colourCountCluster <- length(unique(cluster_counts$Cluster))
  getPalette <- colorRampPalette(brewer.pal(colourCountCluster, "Set3"))

  pm <- cluster_counts %>% 
    arrange(desc(`Clonotype frequency [% cells]`)) %>% 
    dplyr::mutate(Cluster=factor(Cluster, Cluster)) %>% 
  ggplot(aes(x = Cluster, y = `Clonotype frequency [% cells]`, fill = `Cluster`)) +
    geom_bar(stat = "identity", position = "dodge2") +
    theme_pub() +
    labs(x="Clonotype", title = sample_name)+
    scale_y_continuous(name="Frequency [%]", expand = expansion(mult = c(0, NA))) +
    scale_fill_manual(values = getPalette(colourCountCluster + 1)) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45),
      legend.position = "none"
    )

  return(pm)
}
clonotype_quantification_plots <- list()

for (i in 1:4){
  clonotype_quantification_plots[[i]] <- clonotype_cell_counts_MBL[[names(sample_subset_individual[[i]])]] %>% 
  imap(~ quantify_cluster_histogram(.x, .y))
  
  if (i ==1){
    cowplot::plot_grid(plotlist =clonotype_quantification_plots[[i]], 
                       nrow = 1,
                       rel_widths = c(0.9, 0.9, 2.2)) %>% 
    iwalk(ggsave(filename = file.path(RDS_dir_subsets[[i]], "Plots/Clonotype_quantification_plots_MBL.png"), width=14, height=6))
  }
    if (i ==2){
  plot_grid(plotlist =clonotype_quantification_plots[[i]],  
            nrow = 1,
            rel_widths = c(1, 1, 2)) %>% 
        ggsave(filename = file.path(RDS_dir_subsets[[i]], "Plots/Clonotype_quantification_plots_MBL.png"), width=9, height=6)
    }
    if (i ==3){
  cowplot::plot_grid(plotlist = clonotype_quantification_plots[[i]],  
                     nrow = 1, 
                     rel_widths = c(1, 1)) %>% 
        ggsave(filename = file.path(RDS_dir_subsets[[i]], "Plots/Clonotype_quantification_plots_MBL.png"), width = 7, height = 4.5)
    }
    if (i ==4){
  cowplot::plot_grid(plotlist = clonotype_quantification_plots[[i]],  
                     nrow = 1, 
                     rel_widths = c(1, 1)) %>% 
    ggsave(filename = file.path(RDS_dir_subsets[[i]], "Plots/Clonotype_quantification_plots_MBL.png"), width = 16, height = 4.5)
    }    
}
```

### Jitter violin plots
```{r, fig.width=10, fig.height=10}
# Log jitter violin plot
VAF_min_non0 <- list()
jitterviolin_log_mutations <- list()

for (i in 1:4){
VAF_min_non0[[names(sample_subset_individual[i])]] <- as.vector(umap_embedding_mito_cln_MBL[[i]]$VAF) %>%
  unique() %>%
  sort() %>%
  .[[2]]

jitterviolin_log_mutations[[names(sample_subset_individual[i])]] <- umap_embedding_mito_cln_MBL[[i]] %>%
  dplyr::mutate(VAF = case_when(VAF == 0 ~ VAF + VAF_min_non0, .default = VAF)) %>%
  split(.$patient_id)%>%
  map(~tidyr::unite(., "label_cell_barcode", label:cell_barcode, sep = "#", remove = FALSE) %>%
      dplyr::select(label_cell_barcode, `VAF`, variant_id) %>%
      tidyr::pivot_wider(values_from = `VAF`, names_from = label_cell_barcode) %>%
      dplyr::select(-where(~ all(. == 0))) %>%
      tidyr::pivot_longer(-variant_id, names_to = "label_cell_barcode", values_to = "Heteroplasmy") %>%
      tidyr::separate(label_cell_barcode, c("label", "cell_barcode"), sep = "#", remove = FALSE) %>%
      ggplot(., aes(x = label, y = `Heteroplasmy`, colour = label)) +
      facet_wrap(~variant_id, scales ="free_x") +
      geom_jitter(size = 0.1, position = position_jitterdodge()) +
      geom_violin(scale = "width", show.legend = FALSE, color = "lightgrey", fill = NA) +
      scale_y_log10(limits = c(1e-04, 1e+01)) +
      theme_pub() +
      scale_colour_manual(values = my_MBL_CLL_colours, labels = c("MBL", "CLL"), name = "Condition") +
      theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(),
        axis.title.y = element_text(),
        axis.title.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()
      ) +
      guides(colour = guide_legend(override.aes = list(size = 4)))) %>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[names((sample_subset_individual[i]))]], paste0("Plots/Log_Jitterviolinplots/Log_Jitterviolinplot_", .y, ".png")),
    plot = .x,
    width = 10,
    height = 10
  ))
}
```
