---
title: "Get mtDNA mutations"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
# GOAL
The aim of this script is to extract mitochondrial mutations for samples from CLL patients with treatments and characterize clonotypes and their dynamics within the cohort. Initially for only 1 patient, later for all patients. I therefore investigate which clonotypes are unique to the different treatment stages, which increase or decrease throughout the coarse of the disease.


# Load libraries

```{r Load_libraries, warning=FALSE, message=FALSE}
library(Signac)
library(readxl)
library(ArchR)
library(tidyverse)
library(ggforce)
library(ggpubr)
library(RColorBrewer)
library(ggalluvial)
library(ggrepel)
library(VennDiagram)
library(factoextra)
library(pheatmap)
library(ggfortify)
library(Seurat)
library(cluster)
```

## Set paths

```{r }
source(config_paths.R)

library_path <- file.path(project_dir_Penter_Gohil, "data/ATAC")
result_dir <- file.path(project_dir_Penter_Gohil, "results")
ArchR_dir <- "ArchR_Subset1"
```

# Load data

```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```


```{r}
metadata <- read_tsv(file.path(project_dir_Penter_Gohil, "data/20240209_Tidy_Penter_Gohil_metadata.tsv"))
```

```{r Load_ArchR_Project }
# setwd(file.path(project_dir_Penter_Gohil, "results"))
# 
# ## Load data
# CLLProj <- loadArchRProject(path = result_dir)
```

# Analysis

```{r Get_mgatk_paths}
sample_subset <- c("CLL101.2",
                   "CLL101.3",
                   "CLL101.4",
                   "CLL101.5",
                   "CLL101.6")

library_names <- metadata %>%
  dplyr::filter(label %in% sample_subset) %>%
  pull(pool)

mgatk_file_paths <- file.path(library_path, paste0(library_names, ".mgatk/")) %>%
  as.list() %>%
  unlist()
```

```{r Load_individual_mgatk_files}
pool2label <- metadata %>%
  select(pool, label) %>%
  deframe()

#Individual mgatk files per sample
# mito.data.idv <- mgatk_file_paths %>%
#   map(~ ReadMGATK(.)) %>%
#   set_names(pool2label[unlist(library_names)])
# 
# saveRDS(mito.data.idv, file = file.path(result_dir, "RDS_objects/mito.data.idv.RDS"))
```

```{r}
filter_doublets_from_mitodata <- function(mgatk_idv_list){ #, archrproj) {
  mito_subset <-list()
  
  #Get cell barcodes from singlets identified from ATACdata
  # label_cellbarcode_df <- archrproj@cellColData@rownames %>%
  #   str_split_fixed("#", n = 2) %>%
  #   `colnames<-`(c("label", "cellbarcode")) %>%
  #   as_tibble()
  
  #Filter cells 
  for (patient in names(mgatk_idv_list)) {
    # barcodes <- label_cellbarcode_df %>%
    #   filter(label == patient) %>%
    #   pull(cellbarcode)
    
    # Adapt here for mtseq depth
    keep_col_doublets <- mgatk_idv_list[[patient]]$counts@Dimnames[[2]] #%in% barcodes
    keep_col_depth <-  mgatk_idv_list[[patient]]$depth %>% 
      dplyr::filter(mito.depth >=10) %>% 
      rownames()
    #keep_col <- intersect(mgatk_idv_list[[patient]]$counts@Dimnames[[2]][keep_col_doublets], keep_col_depth)

    mgatk_idv_list[[patient]]$counts <- mgatk_idv_list[[patient]]$counts[, keep_col_depth]

    mgatk_idv_list[[patient]]$depth <- mgatk_idv_list[[patient]]$depth[keep_col, , drop = FALSE]
    }

  return(mgatk_idv_list)
}

#mito.data.idv <- filter_doublets_from_mitodata(mito.data.idv)
#saveRDS(mito.data.idv, file = file.path(result_dir, "RDS_objects/mito.data.idv.RDS"))
```

```{r Load_paired_mgatk_files}
# Paired mgatk files per patient
sample_subset <- dplyr::filter(metadata, label %in% sample_subset & str_starts(label, "CLL")) %>%
  pull(patient_id) %>%
  unique()

sample_subset_sample <- dplyr::filter(metadata, patient_id %in% sample_subset & str_starts(label, "CLL")) %>%
  pull(label)

mito.data <- list()
for (i in sample_subset) {
  print(i)
  subset_names <- str_subset(names(mito.data.idv), i)
  subset_names_index <- match(subset_names, names(mito.data.idv))
  
  all_df <- list()
  for (j in subset_names_index){
    print(j)
       df <- mito.data.idv[[subset_names_index[[j]]]]$counts
       colnames(df) <- paste(subset_names[j], colnames(df), sep = "#")
       all_df[[j]] <- df
  }
  mito.data[[i]] <- do.call(cbind, all_df)
  rm(all_df)
}
names(mito.data) <- sample_subset

#saveRDS(mito.data, file = file.path(result_dir, "RDS_objects/mito.data.RDS"))
mito.data<- readRDS(file = file.path(result_dir, "RDS_objects/mito.data.RDS"))
```

# Analysis

```{r Get_mt_variants}
# Get variants cumulative per patient over both samples
variant.df <- mito.data %>%
  map(~ IdentifyVariants(., refallele = mito.data.idv[[1]]$refallele)) %>%
  map2(names(.), ~ .x %>%
    dplyr::mutate(sample_id = .y, .before = "nucleotide"))

# Export_mt_variants
dir.create(file.path(result_dir, "MitochondrialVariants"))

variant.df %>%
  imap(~ write_tsv(.x,
    quote = "none",
    file = file.path(result_dir, paste0("MitochondrialVariants/Variants_", .y, ".csv"))
  ))

saveRDS(variant.df, file = file.path(result_dir, "RDS_objects/variant.df.RDS"))
```

```{r Plot_mt_variants}
all.variant.df.plots <- variant.df %>%
  map(VariantPlot) %>%
  iwalk(~ ggsave(
    filename = file.path(result_dir, paste0("MitochondrialVariants/Plot_Variance_Mean_Ratio_", .y, ".pdf")),
    plot = .x,
    width = 5,
    height = 3
  ))
```

## Maternal mitochondrial variants

```{r Filter_maternal_mt_variants}
mat.variants <- variant.df %>% map(~ subset(.,
  subset = n_cells_conf_detected >= 5 & strand_correlation >= 0.65 & vmr < 0.01
))

saveRDS(mat.variants, file = file.path(result_dir, "RDS_objects/mat.variants.RDS"))
```

### VAFs

```{r VAFs_mat_mt_variants}
# Get maternal VAFs
VAFs.mat <- mito.data %>%
  map2(mat.variants, ~ AlleleFreq(.,
    variants = .y$variant,
    assay = NULL
  ))

saveRDS(VAFs.mat, file = file.path(result_dir, "RDS_objects/VAFs.mat.RDS"))
```

### UMAP VAF Plot

```{r Extract_umap_embedding, fig.height=50, fig.width=10}
# umap_embedding <- CLLProj@embeddings$UMAP$df %>%
#   rownames_to_column("id") %>%
#   separate(id, c("label", "cell_barcode"), sep = "#") %>%
#   separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE)
``` 

```{r Plot_UMAP_mat_mt_variants_VAFs, fig.height=60, fig.width=5}
# Inner join is used to consider only cells which have been used in the umap (here doublets are removed!)
# umap_embedding_mito_mat <- map_dfr(VAFs.mat, ~ as.matrix(.) %>%
#   as_tibble(rownames = "variant_id") %>%
#   pivot_longer(-variant_id,
#     names_to = "cell_barcode",
#     values_to = "VAF"
#   ), .id = "patient_id") %>%
#   separate(cell_barcode, c("label", "cell_barcode"), sep = "#") %>%
#   # arrange(label, -VAF) %>%
#   inner_join(umap_embedding) %>%
#   dplyr::rename(
#     UMAP1 = "IterativeLSI#UMAP_Dimension_1",
#     UMAP2 = "IterativeLSI#UMAP_Dimension_2"
#   )
# 
# saveRDS(umap_embedding_mito_mat, file = paste0(result_dir, "/RDS_objects/umap_embedding_mito_mat.RDS"))
# 
# YlOrRd <- c("snow2", brewer.pal(n = 9, name = "YlOrRd"))
# 
# # Careful! group_by internally orders column before split which causes different order in split list
# umap_embedding_mito_mat_plot <- umap_embedding_mito_mat %>%
#   group_by(patient_id) %>%
#   group_split() %>%
#   set_names(sort(sample_subset)) %>%
#   map(~
#     arrange(., label, -VAF) %>%
#       ggplot(aes(x = UMAP1, y = UMAP2, color = VAF)) +
#       facet_grid(variant_id ~ label) +
#       geom_point(size = 1, alpha = 0.2, stroke = NA) +
#       coord_equal() +
#       scale_colour_gradientn(colors = YlOrRd) +
#       theme_bw() +
#       theme(
#         strip.text = element_text(face = "bold", size = rel(1.2)),
#         strip.background = element_rect(fill = "white", colour = NA),
#         plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
#         panel.border = element_blank(),
#         axis.line.x = element_line(),
#         axis.line.y = element_line(),
#         axis.text.x = element_blank(),
#         axis.ticks.x = element_blank(),
#         axis.text.y = element_blank(),
#         axis.ticks.y = element_blank(),
#         panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank()
#       )) %>%
#   iwalk(~ ggsave(
#     filename = paste0(result_dir, "/Plots/UMAPs_maternal_mito_VAFs/Plot-UMAP-maternal-mitochondrial-VAF_", .y, ".jpg"),
#     plot = .x,
#     width = 5,
#     height = 49
#   ))
# 
# umap_embedding_mito_mat_plot
```

## Clonal mitochondrial variants

```{r Filter_clonal_mt_variants}
# Get clonal variants
cln.variants <- variant.df %>%
  map(~ subset(., subset = n_cells_conf_detected >= 5 & strand_correlation >= 0.65 & vmr > 0.01))

# Export_cln_mito_variants
cln.variants %>%
  map2(names(.), ~ write_tsv(.x, file = file.path(result_dir, paste0("MitochondrialVariants/Clonal_Variants_", .y, ".tsv"))))

saveRDS(cln.variants, file = file.path(result_dir, "RDS_objects/cln.variants.RDS"))
```


### VAFs

```{r VAFs_clonal_mt_variants}
# Get clonal VAFs
VAFs.cln <- mito.data %>%
  map2(cln.variants, ~ AlleleFreq(.,
    variants = .y$variant,
    assay = NULL
  ))

# export in sparse format
VAFs.cln %>%
  map2(names(.), ~ writeMM(., file = file.path(result_dir, paste0("MitochondrialVariants/VAFs_Variants_sparse_", .y, ".mtx"))))

saveRDS(VAFs.cln, file = file.path(result_dir, "RDS_objects/VAFs.cln.RDS"))
```

### UMAP VAF Plot

```{r Plot_UMAP_cln_mt_variants_VAFs, fig.height=60, fig.width=5}
# Inner join is used to consider only cells which have been used in the umap (here doublets are removed!)
umap_embedding_mito_cln <- map_dfr(VAFs.cln, ~ as.matrix(.) %>%
  as_tibble(rownames = "variant_id") %>%
  pivot_longer(-variant_id,
    names_to = "cell_barcode",
    values_to = "VAF"
  ), .id = "patient_id") %>%
  separate(cell_barcode, c("label", "cell_barcode"), sep = "#") %>%
  arrange(label, -VAF) 
```  

### Heatmap VAFs

```{r Heatmap_clonotypes_VAF, fig.width=10, fig.height=7}
# Prepare heatmap VAF input data as matrix with only VAFs unequal 0
VAF_mat <- umap_embedding_mito_cln %>%
  dplyr::filter(!(variant_id == "310T>C")) %>%
  dplyr::group_by(patient_id) %>%
  dplyr::group_split() %>%
  set_names(sort(sample_subset)) %>%
  map(~
    tidyr::unite(., "label_cell_barcode", label:cell_barcode, sep = "#", remove = FALSE) %>%
      dplyr::select(label_cell_barcode, VAF, variant_id) %>%
      tidyr::pivot_wider(values_from = VAF, names_from = label_cell_barcode) %>%
      column_to_rownames("variant_id") %>%
      drop_na() %>%
      dplyr::select(-where(~ all(. == 0))) %>%
      as.matrix())

saveRDS(VAF_mat, file = file.path(result_dir, "RDS_objects/VAF_mat.RDS"))
```


### Signac FindClonotype Function

```{r }
# Cluster together per patient and filter for non-zero VAF cells
mito_VAFs <- VAFs.cln %>% map(~ as.matrix(.) %>%
  as_tibble(rownames = "variant_id") %>%
  filter(variant_id != "310T>C") %>%
  column_to_rownames(var = "variant_id") %>%
  .[, as.logical(colSums(. != 0))] %>%
  as.matrix() %>%
  as("sparseMatrix") %>%
  CreateSeuratObject(assay = "alleles"))

clonotypes <- mito_VAFs %>% map(~ FindClonotypes(.))

saveRDS(object = clonotypes, file = file.path(result_dir, "RDS_objects/clonotypes.RDS"))

find_clonotype_clusters <- clonotypes %>%
  map(~ .@active.ident %>%
    data.frame() %>%
    rownames_to_column("label_cellbarcode") %>%
    dplyr::rename(Cluster = ".")) %>%
  map2(names(.), ~ write_tsv(.x, file = file.path(result_dir, paste0("MitochondrialVariants/FindClonotypesClusters_", .y, ".tsv"))))
```
