---
title: "Mitochondrial mutations"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
# Goal
The aim of this script is to extract mitochondrial mutations for all paired samples and characterize clonotypes and their dynamics within the cohort. 

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(Signac)
library(readxl)
library(ArchR)
library(ggfortify)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(ggrepel)
library(Seurat)
library(cluster)
```

# Set paths
```{r define_directories }
source("config_paths.R")

data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "ATAC")
result_dir <- file.path(project_dir, "results")
ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")
ArchR_dir <- paste0("ArchR_Subset", 1:4) %>% as.list()
result_dir_subsets <- file.path(ArchR_dir_Annotation, ArchR_dir) %>% setNames(paste0("Subset", 1:4))
RDS_dir_subsets <- file.path(result_dir_subsets, "RDS_objects") %>% setNames(paste0("Subset", 1:4))
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r Load_metadata }
metadata <- read_tsv(file = file.path(data_dir, "20240530_Tidy_MBL_CLL_metadata_atac.tsv"))
```

```{r Load_ArchR_Project }
CLLProj <- result_dir_subsets %>% 
  map(~loadArchRProject(path = paste0(.x, "/"))) %>% 
  purrr::set_names(paste0("Subset", 1:4))
```

```{r Get_mgatk_paths}
sample_subset <- list("Subset1" = c("CLL1.1","CLL1.2", "CLL2.1","CLL2.2","CLL3.1", "CLL3.2","MBL1","MBL2","MBL3"),
                      "Subset2" = c("CLL4.1","CLL4.2","CLL5.1","CLL5.2","CLL6.1","CLL6.2","MBL4","MBL6","MBL7"),
  "Subset3" = c("CLL7.1","CLL7.2","CLL8.1","CLL8.2","CLL9.1","CLL9.2","CLL10.1","CLL10.2","MBL9","MBL8"),
  "Subset4" = c("MBL7dup","MBL10","MBL11","MBL12","MBL13","MBL14","MBL15"))

library_names <- sample_subset %>% 
  map(~metadata %>%
  dplyr::filter(label %in% .x) %>%
  pull(pool))

mgatk_file_paths <- library_names %>% 
  map(~file.path(library_path, paste0(.x, ".mgatk")) %>%
  as.list() %>%
  unlist())
```

```{r}
pool2label <- metadata %>%
  select(pool, label) %>%
  deframe()
```

```{r Load_individual_mgatk_files}
#Individual mgatk files per sample
# mito.data.idv <- list()
# 
# for (i in 1:length(mgatk_file_paths)){
#   mito.data.idv[[names(mgatk_file_paths[i])]] <- mgatk_file_paths[[i]] %>%
#                         map(~ ReadMGATK(.x)) %>%
#                         set_names(pool2label[unlist(library_names[[i]])])
# }
# 
# mito_data_idv %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "mito.data.idv.RDS")))
```

```{r}
# New filtering function filtering out cells with low quality and also accounting for mtseq depth of min 10!
# filter_qc_from_mitodata <- function(mgatk_idv_list, archrproj) {
#   mito_subset <-list()
#   
#   #Get cell barcodes from singlets identified from ATACdata
#   label_cellbarcode_df <- archrproj@cellColData@rownames %>%
#     str_split_fixed("#", n = 2) %>%
#     `colnames<-`(c("label", "cellbarcode")) %>%
#     as_tibble()
#   
#   #Filter cells 
#   for (patient in unique(label_cellbarcode_df$label)) {
#     barcodes <- label_cellbarcode_df %>%
#       filter(label == patient) %>%
#       pull(cellbarcode)
#     
#     # Adapt here for mtseq depth
#     keep_col_doublets <- mgatk_idv_list[[patient]]$counts@Dimnames[[2]] %in% barcodes
#     keep_col_depth <-  mgatk_idv_list[[patient]]$depth %>% 
#       dplyr::filter(mito.depth >=10) %>% 
#       rownames()
#     keep_col <- intersect(mgatk_idv_list[[patient]]$counts@Dimnames[[2]][keep_col_doublets], keep_col_depth)
# 
#     mgatk_idv_list[[patient]]$counts <- mgatk_idv_list[[patient]]$counts[, keep_col]
# 
#     mgatk_idv_list[[patient]]$depth <- mgatk_idv_list[[patient]]$depth[keep_col, , drop = FALSE]
#     }
# 
#   return(mgatk_idv_list)
# }
# 
# mito.data.idv <- map2(.x = mito.data.idv,
#                       .y = CLLProj, 
#                       .f = ~filter_qc_from_mitodata(.x, .y))
# 
# mito_data_idv %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "mito.data.idv.RDS")))

mito_data_idv <- RDS_dir_subsets %>% 
  map(~readRDS(file.path(.x, "mito_data_idv.RDS")))
```

```{r Load_paired_mgatk_files}
# Paired mgatk files per patient
sample_subset_paired <- sample_subset %>% 
  map(~dplyr::filter(metadata, label %in% .x & str_starts(label, "CLL")) %>%
  pull(patient_id) %>%
  unique())

sample_subset_sample <- sample_subset_paired %>% 
  map(~dplyr::filter(metadata, patient_id %in% .x & str_starts(label, "CLL")) %>%
  pull(label))

# mito.data <- list()
# 
# for ( i in 1:3){ 
#   for (j in sample_subset_paired[[i]]) {
#     subset_names <- str_subset(names(mito.data.idv), j)
#     subset_names_index <- match(subset_names, names(mito.data.idv))
#   
#     df_1 <- mito.data.idv[[subset_names_index[[1]]]]$counts
#     df_2 <- mito.data.idv[[subset_names_index[[2]]]]$counts
#   
#     colnames(df_1) <- paste(subset_names[1], colnames(df_1), sep = "#")
#     colnames(df_2) <- paste(subset_names[2], colnames(df_2), sep = "#")
#   
#     subset_df <- cbind(df_1, df_2)
#     rm(df_1, df_2)
#     mito.data[[i]] <- append(mito.data[[i]], subset_df)
#   }
#   
# names(mito.data[[i]]) <- sample_subset_paired[[i]]
# }
# 
# mito_data %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "mito.data.RDS")))

mito_data <- RDS_dir_subsets %>% 
  map(~readRDS(file.path(.x, "mito.data.RDS")))
```

# Analysis

```{r Get_mt_variants}
# Get variants cumulative per patient over both samples
# variant.df <- list()
# for (i in 1:3){
# variant.df[[names(sample_subset_paired[i])]] <- mito.data[[names(sample_subset_paired[i])]] %>%
#   map(~ IdentifyVariants(., refallele = mito.data.idv[[names(sample_subset_paired[i])]][[1]]$refallele)) %>%
#   imap(~ .x %>%
#     dplyr::mutate(sample_id = .y, .before = "nucleotide"))
# }
# 
# for (i in 1:3){
#   variant.df [[names(sample_subset_paired[i])]] %>%
#   imap(~ write_tsv(.x,
#     quote = "none",
#     file = file.path(result_dir_subsets, "MitochondrialVariants", paste0("Variants_", .y, ".csv"))
#   ))
# }
# 
# variant.df %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "variant.df.RDS")))

variant.df <- RDS_dir_subsets[1:3] %>% 
  map(~readRDS(file.path(.x, "variant.df.RDS"))) %>% 
  purrr::set_names(names(sample_subset_paired)[1:3])
```

```{r Plot_mt_variants}
all.variant.df.plots <- list()
  
for (i in 1:3){
  variant.df[[i]] %>%
 map(VariantPlot)%>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[i]], "MitochondrialVariants", paste0("Plot_Variance_Mean_Ratio_", .y, ".png")),
    plot = .x,
    width = 5,
    height = 3
  ))
}
```

## Maternal mitochondrial variants
```{r Filter_maternal_mt_variants}
# mat.variants <- list()
# 
# for (i in 1:3){
# mat.variants[[names(sample_subset_paired[i])]] <- variant.df[[i]] %>% 
#   map(~ subset(.,
#   subset = n_cells_conf_detected >= 5 & strand_correlation >= 0.65 & vmr < 0.01
# ))
# }
# 
# mat.variants %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "mat.variants.RDS")))

mat.variants <- RDS_dir_subsets[1:3] %>% 
  map(~readRDS(file.path(.x, "mat.variants.RDS")))
```

### VAFs
```{r VAFs_mat_mt_variants}
# Get maternal VAFs
# VAFs.mat <- list()
# 
# for (i in 1:3){
# VAFs.mat[[names(sample_subset_paired[i])]] <- mito.data %>%
#   map2(mat.variants, ~ AlleleFreq(.,
#     variants = .y$variant,
#     assay = NULL
#   ))
# }
# 
# VAFs.mat %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "VAFs.mat.RDS")))

VAFs.mat <-file.path(RDS_dir_subsets[1:3], "VAFs.mat.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))
```

### UMAP VAF Plot
```{r Extract_umap_embedding, fig.height=50, fig.width=10}
umap_embedding <- CLLProj %>% 
                        map(~.x@embeddings$UMAP$df %>%
  rownames_to_column("id") %>%
  separate(id, c("label", "cell_barcode"), sep = "#") %>%
  separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE))
``` 

```{r}
# Inner join is used to consider only cells which have been used in the umap (here doublets are removed!)
# umap_embedding_mito_mat <- list()
# for (i in 1:3){
# umap_embedding_mito_mat[[names(sample_subset_paired[i])]] <- map_dfr(VAFs.mat[[i]], ~ as.matrix(.) %>%
#   as_tibble(rownames = "variant_id") %>%
#   pivot_longer(-variant_id,
#     names_to = "cell_barcode",
#     values_to = "VAF"
#   ), .id = "patient_id") %>%
#   separate(cell_barcode, c("label", "cell_barcode"), sep = "#") %>%
#   inner_join(umap_embedding[[names(sample_subset_paired[i])]], by = join_by(patient_id, label, cell_barcode)) %>%
#   dplyr::rename(
#     UMAP1 = "IterativeLSI#UMAP_Dimension_1",
#     UMAP2 = "IterativeLSI#UMAP_Dimension_2"
#   )
# }
# 
# umap_embedding_mito_mat %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "umap_embedding_mito_mat.RDS")))

umap_embedding_mito_mat <- file.path(RDS_dir_subsets[1:3], "umap_embedding_mito_mat.RDS")%>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))
```


```{r Plot_UMAP_mat_mt_variants_VAFs, fig.height=60, fig.width=5}
# Careful! group_by internally orders column before split which causes different order in split list
umap_embedding_mito_mat_plot <- list()

for (i in 1:3){
umap_embedding_mito_mat_plot[[names(sample_subset_paired[i])]] <- umap_embedding_mito_mat[[i]] %>%
  split(.$patient_id) %>% 
  imap(~
      dplyr::arrange(., label, variant_id, -VAF) %>% 
      dplyr::mutate(VAF=if_else(VAF==0, NA, VAF)) %>% 
      ggplot(aes(x = UMAP1, y = UMAP2, color = VAF)) +
      geom_point(size = 1, alpha = 0.2, stroke = NA) +
      coord_equal() +
      scale_colour_gradient(
        name = "Heteroplasmy",
        low = "yellow",
        high = "darkred",
        na.value = "lightgrey",
        trans = "log10",
        limits = c(1e-02, 1e+00),
        oob=scales::oob_squish,
        breaks=10^seq(from=-2, to=0, by=1),
        labels=c("≤0.01", "0.1",  "1")) +
      theme_pub() +
      labs(title = .y) +
      theme(
        strip.background = element_rect(
          fill = "white",
          colour = NA),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.ticks = element_blank()
      ) +
      facet_grid(variant_id~label)) %>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[i]], "Plots/UMAPs_maternal_mito_VAFs", paste0("Plot-UMAP-maternal-mitochondrial-VAF_", .y, ".png")),
    plot = .x,
    width = 5,
    height = 49
  ))
  
}

umap_embedding_mito_mat_plot
```

## Clonal mitochondrial variants
```{r Filter_clonal_mt_variants}
# # Get clonal variants
# cln.variants <- list()
# for (i in 1:3){
# cln.variants[[names(sample_subset_paired[i])]] <- variant.df[[i]] %>%
#   map(~ subset(., subset = n_cells_conf_detected >= 5 & strand_correlation >= 0.65 & vmr > 0.01))
# }
# 
# # Export_cln_mito_variants
# for (i in 1:3){
#   cln.variants [[names(sample_subset_paired[i])]] %>%
#   imap(~ write_tsv(.x,
#     quote = "none",
#     file = file.path(result_dir_subsets, "MitochondrialVariants", paste0("Clonal_Variants_", .y, ".csv"))
#   ))
# }
# 
# cln.variants %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "cln.variants.RDS")))

cln.variants <- file.path(RDS_dir_subsets[1:3], "cln.variants.RDS")%>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))
```


### VAFs
```{r VAFs_clonal_mt_variants}
# Get clonal VAFs
# VAFs.cln <- list()
# 
# for (i in 1:3){
# VAFs.cln[[names(sample_subset_paired[i])]] <- mito.data %>%
#   map2(cln.variants, ~ AlleleFreq(.,
#     variants = .y$variant,
#     assay = NULL
#   ))
# }
# 
# VAFs.cln %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "VAFs.cln.RDS")))
# 
# VAFs.cln <-file.path(RDS_dir_subsets[1:3], "VAFs.cln.RDS") %>% 
#   map(~readRDS(.x)) %>% 
#   purrr::set_names(names(sample_subset_paired[1:3]))
# 
# # export in sparse format
# for (i in 1:3){
#   VAFs.cln [[names(sample_subset_paired[i])]] %>%
#   imap(~ write_tsv(.x,
#     quote = "none",
#     file = file.path(result_dir_subsets, "MitochondrialVariants", paste0("VAFs_Variants_sparse_", .y, "mtx"))
#   ))
# }
# 

VAFs.cln <-file.path(RDS_dir_subsets[1:3], "VAFs.cln.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))
```

### UMAP VAF Plot
```{r, fig.width=10, fig.height=100}
# Inner join is used to consider only cells which have been used in the umap (here doublets are removed!)
# umap_embedding_mito_cln <- list()
# for (i in 1:3){
# umap_embedding_mito_cln[[names(sample_subset_paired[i])]] <- map_dfr(VAFs.cln[[i]], ~ as.matrix(.) %>%
#   as_tibble(rownames = "variant_id") %>%
#   pivot_longer(-variant_id,
#     names_to = "cell_barcode",
#     values_to = "VAF"
#   ), .id = "patient_id") %>%
#   separate(cell_barcode, c("label", "cell_barcode"), sep = "#") %>%
#   inner_join(umap_embedding[[names(sample_subset_paired[i])]], by = join_by(patient_id, label, cell_barcode)) %>%
#   dplyr::rename(
#     UMAP1 = "IterativeLSI#UMAP_Dimension_1",
#     UMAP2 = "IterativeLSI#UMAP_Dimension_2"
#   )
# }
# 
# umap_embedding_mito_cln %>%  
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "umap_embedding_mito_cln.RDS")))

umap_embedding_mito_cln <-file.path(RDS_dir_subsets[1:3], "umap_embedding_mito_cln.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))

# export with umap coordinates

# for (i in 1:3){
#   umap_embedding_mito_cln[[i]] %>% 
#   split(.$patient_id) %>%
#   imap(~ write_tsv(.x,
#     file = file.path(result_dir_subsets[[i]], "MitochondrialVariants", paste0("VAFs_Variants_UMAP_", .y, ".tsv"))))
# }
```

```{r, fig.width=10, fig.height=100}
umap_embedding_mito_cln_plot <- list()
  
for (i in 1:3){
  umap_embedding_mito_cln_plot[[names(sample_subset_paired[i])]] <- umap_embedding_mito_cln[[i]] %>%
  split(.$patient_id) %>%
  imap(~
      dplyr::arrange(.x, label, variant_id, VAF) %>% 
      dplyr::mutate(VAF=if_else(VAF==0, NA, VAF)) %>%
      ggplot(aes(x = UMAP1, y = UMAP2, color = VAF)) +
      geom_point(size = 1, alpha = 0.2, stroke = NA) +
      coord_equal() +
      scale_colour_gradient(
        name = "Heteroplasmy",
        low = "yellow",
        high = "darkred",
        na.value = "lightgrey",
        trans = "log10",
        limits = c(1e-02, 1e+00),
        oob=scales::oob_squish,
        breaks=10^seq(from=-2, to=0, by=1),
        labels=c("≤0.01", "0.1",  "1")) +
      theme_pub() +
      labs(title = .y) +
      theme(
        strip.background = element_rect(
          fill = "white",
          colour = NA),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.ticks = element_blank()
      ) +
      facet_grid(variant_id~label)) %>%
  iwalk(~ ggsave(
    filename = paste0(result_dir_subsets[[i]], "Plots/UMAPs_clonal_mito_VAFs/Plot-UMAP-clonal-mitochondrial-VAF_", .y, ".png"),
    plot = .x,
    width = 5,
    height = 49
  ))
}
umap_embedding_mito_cln_plot
```

### Heatmap VAFs
```{r}
# Prepare heatmap VAF input data as matrix with only VAFs unequal 0
# VAF_mat <- list()
# 
# for (i in 1:3){
# VAF_mat[[names(sample_subset_paired[i])]] <- umap_embedding_mito_cln[[i]] %>%
#   dplyr::filter(!(variant_id == "310T>C")) %>%
#   split(.$patient_id) %>%
#   map(~
#     tidyr::unite(., "label_cell_barcode", label:cell_barcode, sep = "#", remove = FALSE) %>%
#       dplyr::select(label_cell_barcode, VAF, variant_id) %>%
#       tidyr::pivot_wider(values_from = VAF, names_from = label_cell_barcode) %>%
#       column_to_rownames("variant_id") %>%
#       drop_na() %>%
#       dplyr::select(-where(~ all(. == 0))) %>%
#       as.matrix())
# }
# 
# VAF_mat %>% 
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "VAF_mat.RDS")))


VAF_mat <-file.path(RDS_dir_subsets[1:3], "VAF_mat.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))
```

```{r }
YlOrRd <- c("snow2", brewer.pal(n = 9, name = "YlOrRd"))

# Prepare annotation labels
column_annotation <- list()
col_annot <- list()
for (i in 1:3){
column_annotation[[names(sample_subset_paired[i])]] <- CLLProj[[i]]@cellColData %>%
  as.data.frame() %>%
  dplyr::select(Clusters_0.6_IterativeLSI) %>%
  dplyr::mutate(Label = rownames(.) %>% str_split_fixed("#", 2) %>% .[, 1]) %>%
  tidyr::separate(Label, c("patient_id", "sample"), remove = FALSE) %>%
  split(.$patient_id) %>%
  map(~ dplyr::select(., -c("sample", "patient_id")))

# Select only paired samples
column_annotation[[names(sample_subset_paired[i])]] <- column_annotation[[names(sample_subset_paired[i])]][grepl("^CLL", names(column_annotation[[names(sample_subset_paired[i])]]))]

# Prepare annotation colours
col_annot[[names(sample_subset_paired[i])]] <- column_annotation[[names(sample_subset_paired[i])]][[1]]
}

# Make function for heatmap
build_heatmap <- function(VAFs, patient_name, col_annot, subset) {
  # Prepare annotation colours
  colours_annotation <- imap(col_annot, unique) %>%
    map(~ set_names(brewer.pal(length(.), "Set3")[1:length(.)], .))

  hm <- pheatmap::pheatmap(
    mat = VAFs,
    color = colorRampPalette(YlOrRd)(10),
    annotation_col = col_annot,
    annotation_colors = colours_annotation,
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    show_colnames = FALSE,
    legend_breaks = c(0, 0.5, 1),
    main = paste0("Mitochondrial Clonotypes Individual ", patient_name),
    legend_labels = c("0", "0.5", "1"),
    legend = TRUE
  )

  ggsave(
    device = "jpg",
    filename = file.path(subset, "Plots/VAF_heatmaps_Seurat_Clusters", paste0("VAF_heatmap_seurat_clusters_", patient_name, ".png")),
    plot = hm,
    width = 10,
    height = 7
  )
  return(hm)
}

# Plot heatmap
patient_heatmap <- list()
for (i in 1:3){
 patient_heatmap[[names(sample_subset_paired[i])]]<- list(VAF_mat[[i]], 
                                                          sort(sample_subset_paired[[i]]), 
                                                          column_annotation[[i]],
                                                          subset=result_dir_subsets[names(sample_subset_paired[i])]) %>%
  pmap(build_heatmap)
}
```

### Test PCA
```{r}
umap_embedding_mito_cln %>%
  bind_rows() %>% 
  dplyr::select(patient_id, label, cell_barcode, VAF, variant_id) %>%
  split(.$patient_id) %>%
  map(~ tidyr::unite(., "label_cell_barcode", label:cell_barcode, sep = "#") %>%
    dplyr::select(-patient_id) %>%
    tidyr::pivot_wider(names_from = variant_id, values_from = VAF) %>%
    na.omit() %>%
    tibble::column_to_rownames("label_cell_barcode") %>%
    as.matrix() %>% 
    prcomp(scale = TRUE) %>%
    autoplot())
```


### Signac FindClonotype Function
Different plot types compare the VAF between MBL and CLL sample of a patient
```{r }
# Cluster together per patient and filter for non-zero VAF cells
# mito_VAFs <- list()
# clonotypes <- list()
# 
# for (i in 1:3){
# mito_VAFs[[names(sample_subset_paired[i])]] <- VAFs.cln[[i]] %>% 
#   map(~ as.matrix(.x) %>%
#   as_tibble(rownames = "variant_id") %>%
#   dplyr::filter(variant_id != "310T>C") %>%
#   column_to_rownames(var = "variant_id") %>%
#   .[, as.logical(colSums(. != 0))] %>%
#   as.matrix() %>%
#   as("sparseMatrix") %>%
#   CreateSeuratObject(assay = "alleles"))
# 
# clonotypes[[names(sample_subset_paired[i])]] <- mito_VAFs[[names(sample_subset_paired[i])]] %>% 
#   map(~ FindClonotypes(.))
# 
# }
# 
# clonotypes %>%
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "clonotypes.RDS")))
# 
# find_clonotype_clusters <- list()
# for (i in 1:3){
# find_clonotype_clusters[[names(sample_subset_paired[i])]] <- clonotypes[[i]] %>%
#   map(~ .@active.ident %>%
#     data.frame() %>%
#     rownames_to_column("label_cellbarcode") %>%
#     dplyr::rename(Cluster = ".")) %>%
#   imap(~write_tsv(.x, file = file.path(result_dir, "MitochondrialVariants", paste0("FindClonotypesClusters_", .y, ".tsv"))))
# }

clonotypes <- file.path(RDS_dir_subsets[1:3], "clonotypes.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))
```

```{r}
# # Subset MBL and CLL as seperate Seurat objects
# MBL.clonotypes.logical <- list()
# MBL.clonotypes <- list()
# CLL.clonotypes.logical <- list()
# CLL.clonotypes <- list()
# 
# for (i in 1:3){
# MBL.clonotypes.logical[[names(sample_subset_paired[i])]] <- clonotypes[[i]] %>% 
#   map(~ Cells(.) %>% 
#         str_detect("CLL\\d+\\.1"))
# 
# MBL.clonotypes[[names(sample_subset_paired[i])]] <- clonotypes[[i]] %>% 
#   map2(MBL.clonotypes.logical[[names(sample_subset_paired[i])]], ~ .x[, .y])
# 
# 
# CLL.clonotypes.logical[[names(sample_subset_paired[i])]] <- clonotypes[[i]] %>% 
#   map(~ Cells(.) %>% str_detect("CLL\\d+\\.2"))
# 
# CLL.clonotypes[[names(sample_subset_paired[i])]] <- clonotypes[[i]] %>% 
#   map2(CLL.clonotypes.logical[[names(sample_subset_paired[i])]], ~ .x[, .y])
# }
# 
# MBL.clonotypes %>% 
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "MBL.clonotypes.RDS")))
# 
# CLL.clonotypes %>% 
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "CLL.clonotypes.RDS")))

MBL.clonotypes <- file.path(RDS_dir_subsets[1:3], "MBL.clonotypes.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))

CLL.clonotypes <- file.path(RDS_dir_subsets[1:3], "CLL.clonotypes.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))

```

```{r fig.width=26.66, fig.height=9}
# Plot seperately per disease for ALL cells (unfiltered for MBL, CLL or healthy B cells)
MBL.clonotypes.plot <- list()
CLL.clonotoypes.plot <- list()

for (i in 1:3){
MBL.clonotypes.plot[[names(sample_subset_paired[i])]] <- MBL.clonotypes[[i]] %>%
  imap(~ DoHeatmap(.x,
    features = VariableFeatures(.),
    slot = "data",
    group.colors = get_palette(palette = "Set3", k = length(unique(.@meta.data[["seurat_clusters"]])))
  ) +
    scale_fill_distiller(
      palette = "YlOrRd",
      direction = 1,
      trans = "log",
      breaks = 10^c(-12:0),
      name = "VAF",
      na.value = "#082567"
    ) +
    labs(title = .y) +
    theme(
      plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
      axis.text = element_text(face = "bold")
    ) +
    guides(colour = "none")) %>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[names(sample_subset_paired[i])]], "Plots/VAF_Heatmap_Find_Clonotype/", paste0("VAF_Heatmap_Find_Clonotype_MBL_", .y, ".jpg")),
    plot = .x,
    width = 15,
    height = 10
  ))

CLL.clonotoypes.plot[[names(sample_subset_paired[i])]] <- CLL.clonotypes[[i]] %>%
  imap(~ DoHeatmap(.x,
    features = VariableFeatures(.),
    slot = "data",
    group.colors = get_palette(palette = "Set3", k = length(unique(.@meta.data[["seurat_clusters"]])))
  ) +
    scale_fill_distiller(
      palette = "YlOrRd",
      direction = 1,
      trans = "log",
      breaks = 10^c(-12:0),
      name = "VAF",
      na.value = "#082567"
    ) +
    labs(title = .y) +
    theme(
      plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
      axis.text = element_text(face = "bold")
    ) +
    guides(colour = "none")) %>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[names(sample_subset_paired[i])]], "Plots/VAF_Heatmap_Find_Clonotype/", paste0("VAF_Heatmap_Find_Clonotype_CLL_", .y, ".jpg")),
    plot = .x,
    width = 15,
    height = 10
  ))
}

ggarrange(plotlist=c(MBL.clonotypes.plot, CLL.clonotoypes.plot), ncol=4, nrow=2)
```

#### Quantify clusters
```{r}
# Count how many cells are within each cluster per disease condition
#For MBL or CLL cells
# my_annotation <- list()
# my_annotation_df <- list()
# find_clonotype_clusters <- list()
# clonotype_cell_counts_MBL_CLL <- list()
# clonotype_cell_counts_B <- list()
# 
# for (i in 1:3){
# my_annotation[[names(sample_subset_paired[i])]] <- tibble("label_cellbarcode" = CLLProj[[i]]@cellColData@rownames, "My_annotation" = CLLProj[[i]]@cellColData$My_Annotation) %>%
#   with(setNames(My_annotation, label_cellbarcode))
# 
# my_annotation_df[[names(sample_subset_paired[i])]] <- enframe(my_annotation[[i]], name = "label_cellbarcode", value = "My_annotation")
# 
# find_clonotype_clusters[[names(sample_subset_paired[i])]] <- find_clonotype_clusters[[i]] %>% 
#   map(~left_join(., my_annotation_df[[i]]))
# 
# clonotype_cell_counts_MBL_CLL[[names(sample_subset_paired[i])]] <- find_clonotype_clusters[[i]] %>% 
#   map(~ tidyr::separate(., "label_cellbarcode", c("label", "cell_barcode"), sep = "#", remove = FALSE) %>%
#   dplyr::filter(`My_annotation` %in% c("MBL_B_Cells", "CLL_B_Cells")) %>%
#   split(.$label) %>%
#   imap(~ group_by(.x, Cluster) %>%
#     dplyr::summarise("cell counts" = n()) %>%
#     dplyr::mutate(
#       "Clonotype frequency [% cells]" = round(`cell counts` / sum(`cell counts`) * 100, digits = 3),
#       "label" = .y
#     ) %>%
#     dplyr::relocate(label, .before = (Cluster))) %>%
#   bind_rows())
# 
# #For healthy B cells
# clonotype_cell_counts_B[[names(sample_subset_paired[i])]] <- find_clonotype_clusters[[i]] %>% 
#   map(~ tidyr::separate(., "label_cellbarcode", c("label", "cell_barcode"), sep = "#", remove = FALSE) %>%
#   dplyr::filter(`My_annotation` %in% c("B")) %>%
#   split(.$label) %>%
#   map2(names(.), ~ group_by(.x, Cluster) %>%
#     dplyr::summarise("cell counts" = n()) %>%
#     dplyr::mutate(
#       "Clonotype frequency [% cells]" = round(`cell counts` / sum(`cell counts`) * 100, 3),
#       "label" = .y
#     ) %>%
#     dplyr::relocate(label, .before = (Cluster))) %>%
#   bind_rows())
# }
# 
# clonotype_cell_counts_MBL_CLL %>% 
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "clonotype_cell_counts_MBL_CLL.RDS")))
# 
# clonotype_cell_counts_B %>% 
#   imap(~saveRDS(.x, file = file.path(RDS_dir_subsets[[.y]], "clonotype_cell_counts_B.RDS")))

clonotype_cell_counts_MBL_CLL <- file.path(RDS_dir_subsets[1:3], "clonotype_cell_counts_MBL_CLL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))

clonotype_cell_counts_MBL_CLL <- file.path(RDS_dir_subsets[1:3], "clonotype_cell_counts_MBL_CLL.RDS") %>% 
  map(~readRDS(.x)) %>% 
  purrr::set_names(names(sample_subset_paired[1:3]))
```

```{r, fig.width=8, fig.height=4}
quantify_cluster_plot <- function(cluster_counts, sample_name) {
  colourCountCluster <- length(unique(cluster_counts$Cluster))
  getPalette <- colorRampPalette(brewer.pal(colourCountCluster, "Set3"))

  pm <- ggplot(cluster_counts, aes(x = label, y = `Clonotype frequency [% cells]`, group = Cluster, colour = `Cluster`)) +
    geom_point() +
    geom_path() +
    theme_bw() +
    scale_colour_manual(values = getPalette(colourCountCluster + 1)) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(),
      axis.title.y = element_text(face = "bold"),
      axis.title.x = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.text = element_text(size = 8),
      legend.position = "right",
      legend.direction = "vertical"
    ) +
    scale_x_discrete(labels = c("MBL", "CLL")) +
    labs(title = sample_name) +
    guides(colour = guide_legend(override.aes = list(size = 4)))

  return(pm)
}

clonotype_quantification_plots <- list()

for (i in 1:3){
clonotype_quantification_plots[[names(sample_subset_paired[i])]] <- clonotype_cell_counts_MBL_CLL[[i]] %>% 
  imap(~ quantify_cluster_plot(.x, .y))
}

clonotype_quantification_plots %>% 
  imap(~ggarrange(plotlist = .x, nrow = 1) %>% 
  ggsave(filename = file.path(result_dir_subsets[[names(clonotype_quantification_plots[[i]])]], "/Plots/Clonotype_quantification_plots.jpg"), width = 18, height = 6))
```

```{r}
quantify_cluster_histogram <- function(cluster_counts, sample_name) {
  pm <- ggplot(cluster_counts, aes(x = Cluster, y = `Clonotype frequency [% cells]`, fill = `label`, group = `label`)) +
    geom_bar(stat = "identity", position = "dodge2") +
    theme_bw() +
    scale_fill_manual(values = my_MBL_CLL_colours, labels = c("MBL", "CLL"), name = "Disease") +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(),
      axis.title.y = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, face = "bold"),
      axis.text.y = element_text(face = "bold"),
      axis.title.x = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.text = element_text(size = 8)
    ) +
    labs(title = sample_name)

  return(pm)
}

clonotype_quantification_histograms <- list()
for (i in 1:3){
clonotype_quantification_histograms[[names(sample_subset_paired[i])]] <- clonotype_cell_counts_MBL_CLL[[i]] %>% 
  imap(~ quantify_cluster_histogram(.x, .y))

ggarrange(plotlist =clonotype_quantification_histograms[[names(sample_subset_paired[i])]], nrow=1, common.legend = TRUE, legend="right") %>% 
  ggsave(filename = paste0(result_dir_subsets[[names((sample_subset_paired[i]))]], "/Plots/Clonotype_quantification_histograms.jpg"), 
         width = 16, 
         height=4)
}
```

### Plots per mutation 
#### Jitter violin plots
```{r, fig.width=10, fig.height=8}
# Log jitter violin plot
VAF_min_non0 <- list()
jitterviolin_log_mutations <- list()

for (i in 1:3){
VAF_min_non0[[names(sample_subset_paired[i])]] <- as.vector(umap_embedding_mito_cln[[i]]$VAF) %>%
  unique() %>%
  sort() %>%
  .[[2]]

jitterviolin_log_mutations[[names(sample_subset_paired[i])]] <- umap_embedding_mito_cln[[i]] %>%
  dplyr::mutate(VAF = case_when(VAF == 0 ~ VAF + VAF_min_non0, .default = VAF)) %>%
  split(.$patient_id)%>%
  map(~
    tidyr::unite(., "label_cell_barcode", label:cell_barcode, sep = "#", remove = FALSE) %>%
      dplyr::select(label_cell_barcode, `VAF`, variant_id) %>%
      tidyr::pivot_wider(values_from = `VAF`, names_from = label_cell_barcode) %>%
      dplyr::select(-where(~ all(. == 0))) %>%
      tidyr::pivot_longer(-variant_id, names_to = "label_cell_barcode", values_to = "VAF") %>%
      tidyr::separate(label_cell_barcode, c("label", "cell_barcode"), sep = "#", remove = FALSE) %>%
      ggplot(., aes(x = label, y = `VAF`, colour = label)) +
      facet_wrap(~variant_id) +
      geom_jitter(size = 0.1, position = position_jitterdodge()) +
      geom_violin(scale = "width", show.legend = FALSE, color = "lightgrey", fill = NA) +
      scale_y_log10(limits = c(1e-04, 1e+01)) +
      stat_compare_means(
        method = "wilcox.test",
        paired = FALSE,
        hide.ns = TRUE,
        aes(label = ..p.signif..),
        label.x = 1.5,
        label.y = 0,
        show.legend = FALSE
      ) +
      theme_pub() +
      scale_colour_manual(values = my_MBL_CLL_colours, labels = c("MBL", "CLL"), name = "Disease") +
      theme(
        strip.text = element_text(face = "bold", size = rel(1.2)),
        strip.background = element_rect(fill = "white", colour = NA),
        plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
        panel.border = element_blank(),
        axis.line.x = element_line(linetype = "solid", colour = "black"),
        axis.line.y = element_line(linetype = "solid", colour = "black"),
        axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()
      ) +
      guides(colour = guide_legend(override.aes = list(size = 4)))) %>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[names((sample_subset_paired[i]))]], paste0("Plots/Log_Jitterviolinplots", paste0("Log_Jitterviolinplot_", .y, ".jpg"))),
    plot = .x,
    width = 15,
    height = 15
  ))

}
```

#### Volcano plots
```{r Volcano_plots, fig.height=5, fig.width=10}
# Calculate p-values for clonal mitochondrial variants per sample using unpaired wilcoxon rank sum test (no p-value correction so far!)
p_values <- list()
mean_per_sample <- list()
stats_volcano <- list()
volcano_input <- list()
volcano_plot <- list()

for (i in 1:3){
p_values[[names(sample_subset_paired[i])]] <- umap_embedding_mito_cln[[i]] %>%
  dplyr::filter(variant_id != "310T>C") %>%
  split(.$patient_id)%>%
  map(~ tidyr::pivot_wider(., names_from = `label`, values_from = `VAF`) %>%
    dplyr::rename(
      MBL = ends_with(".1"),
      CLL = ends_with(".2")
    ) %>%
    dplyr::group_by(variant_id) %>%
    replace_na(list(MBL = 0, CLL = 0)) %>%
    do(p_value = wilcox.test(.$MBL, .$CLL, paired = FALSE, data = .)$p.value) %>%
    dplyr::mutate(p_value = unlist(p_value)))

# Calculate means per sample and mutation
mean_per_sample[[names(sample_subset_paired[i])]] <- umap_embedding_mito_cln[[i]] %>%
  dplyr::filter(variant_id != "310T>C") %>%
  split(.$patient_id)%>%
  map(~ tidyr::pivot_wider(., names_from = `label`, values_from = `VAF`) %>%
    dplyr::rename(
      MBL = ends_with(".1"),
      CLL = ends_with(".2")
    ) %>%
    dplyr::group_by(variant_id) %>%
    replace_na(list(MBL = 0, CLL = 0)) %>%
    dplyr::summarise_at(vars(MBL, CLL), list(`Mean heteroplasmy per sample [%]` = mean)) %>%
    dplyr::mutate(difference_CLL_MBL = `CLL_Mean heteroplasmy per sample [%]` - `MBL_Mean heteroplasmy per sample [%]`) %>%
    dplyr::arrange(-difference_CLL_MBL))

# Summarize stats and calculate logs
stats_volcano[[names(sample_subset_paired[i])]] <- mean_per_sample[[i]] %>%
  map2(p_values[[names(sample_subset_paired[i])]], ~ dplyr::left_join(.x, .y) %>%
    dplyr::mutate(
      fold_change = `CLL_Mean heteroplasmy per sample [%]` / `MBL_Mean heteroplasmy per sample [%]`,
      log2_fold_change = log2(`CLL_Mean heteroplasmy per sample [%]` / `MBL_Mean heteroplasmy per sample [%]`),
      "-log10_p_value" = -log10(p_value),
      p_value_significant = case_when(p_value < 0.05 ~ TRUE, .default = FALSE)
    )) %>%
  set_names(sort(sample_subset_paired))

saveRDS(stats_volcano[[names(sample_subset_paired[i])]], file = file.path(RDS_dir_subsets[[names((sample_subset_paired[i]))]], "stats_volcano.RDS"))

# Plot only variants which have non-zero heteroplasmy in MBL state
volcano_input[[names(sample_subset_paired[i])]] <- stats_volcano[[i]] %>%
  map(~ filter(.x, log2_fold_change != Inf))

volcano_plot[[names(sample_subset_paired[i])]] <- volcano_input[[i]] %>%
  imap(~ ggplot(.x, aes(x = log2_fold_change, y = `-log10_p_value`, color = p_value_significant)) +
      geom_point(size = 3) +
      geom_text_repel(aes(label = variant_id),
        data = subset(.x, p_value_significant == "TRUE"),
        show.legend = FALSE
      ) +
      lims(x = c(-(max(abs(.x$log2_fold_change)) + 1), max(abs(.x$log2_fold_change)) + 1)) +
      theme_bw() +
      labs(
        title = .y,
        colour = "Significance",
        x = "log2(CLL/MBL)",
        y = "-log10(p-value)"
      ) +
      scale_colour_manual(values = c("TRUE" = "darkred", "FALSE" = "lightgrey"), labels = c( "TRUE" = "< 0.05", "FALSE"="\u2265 0.05")) +
      theme(
        plot.title = element_text(hjust = 0.5, size = rel(2), face = "bold"),
        panel.border = element_blank(),
        axis.line.x = element_line(linetype = "solid", colour = "black"),
        axis.line.y = element_line(linetype = "solid", colour = "black"),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold")
      )
  ) %>%
  iwalk(~ ggsave(
    filename = file.path(result_dir_subsets[[names((sample_subset_paired[i]))]], "/Plots/Log_Volcanoplots/Log_Volcanoplot_", .y, ".jpg"),
    plot = .x,
  ))
}
```
