---
title: "Heteroplasmy distribution"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
GOAL: The aim of this script is to investigate what the distribution of heteroplasmy looks like for each mutation in each patient sample

# Load libraries
```{r}
library(tidyverse)
library(ArchR)
library(RColorBrewer)
library(ggpubr)
library(vctrs)
library(ggrepel)
library(cowplot)
```

# Set paths
```{r}
source("config_paths.R")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "Heteroplasmy_analysis")
summary_dir <- file.path(project_dir, "Summary_Plots")
RDS_dir <- file.path(analysis_dir, "RDS_Objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")
plotting_dir <- file.path(analysis_dir, "Plots")

data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "ATAC")

ArchR_dir <- paste0("ArchR_Subset", 1:4)
ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")

result_dir_subset1 <- file.path(ArchR_dir_Annotation, ArchR_dir[1])
result_dir_subset2 <- file.path(ArchR_dir_Annotation, ArchR_dir[2])
result_dir_subset3 <- file.path(ArchR_dir_Annotation, ArchR_dir[3])
result_dir_subset4 <- file.path(ArchR_dir_Annotation, ArchR_dir[4])
```

```{r}
CLLProj_Subset1 <- loadArchRProject(path = result_dir_subset1)
CLLProj_Subset2 <- loadArchRProject(path = result_dir_subset2)
CLLProj_Subset3 <- loadArchRProject(path = result_dir_subset3)
CLLProj_Subset4 <- loadArchRProject(path = result_dir_subset4)
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r}
# Annotate all cells from ArchRprojects
my_annotation <- c(
  tibble("label_cellbarcode" = CLLProj_Subset1@cellColData@rownames, "My_annotation" = CLLProj_Subset1@cellColData$My_Annotation) %>%
    with(setNames(My_annotation, label_cellbarcode)),
  tibble("label_cellbarcode" = CLLProj_Subset2@cellColData@rownames, "My_annotation" = CLLProj_Subset2@cellColData$My_Annotation) %>%
    with(setNames(My_annotation, label_cellbarcode)),
  tibble("label_cellbarcode" = CLLProj_Subset3@cellColData@rownames, "My_annotation" = CLLProj_Subset3@cellColData$My_Annotation) %>%
    with(setNames(My_annotation, label_cellbarcode)),
  tibble("label_cellbarcode" = CLLProj_Subset4@cellColData@rownames, "My_annotation" = CLLProj_Subset4@cellColData$My_Annotation) %>%
    with(setNames(My_annotation, label_cellbarcode))
)

my_annotation_df <- enframe(my_annotation, name = "label_cellbarcode", value = "My_annotation")

heteroplasmy_long <- list(
  readRDS(file.path(result_dir_subset1, "RDS_objects/umap_embedding_mito_cln.RDS")),
  readRDS(file.path(result_dir_subset1, "RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
  readRDS(file.path(result_dir_subset2, "RDS_objects/umap_embedding_mito_cln.RDS")),
  readRDS(file.path(result_dir_subset2, "RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
  readRDS(file.path(result_dir_subset3, "RDS_objects/umap_embedding_mito_cln.RDS")),
  readRDS(file.path(result_dir_subset3, "RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
  readRDS(file.path(result_dir_subset4, "RDS_objects/umap_embedding_mito_cln_MBL.RDS"))
) %>%
  bind_rows() %>%
  dplyr::select(patient_id, label, variant_id, cell_barcode, VAF) %>%
  tidyr::unite("label_cellbarcode", c(label, cell_barcode), sep = "#", remove = FALSE) %>%
  left_join(my_annotation_df, by = join_by(label_cellbarcode)) %>%
  dplyr::rename(
    "Heteroplasmy" = VAF,
    "Variant" = variant_id
  ) %>%
  dplyr::filter(
    Variant != "310T>C",
    patient_id != "MBL7dup",
    My_annotation %in% c("MBL_B_Cells", "CLL_B_Cells"),
    !(patient_id == "CLL3" & Variant == "11711G>A"|
                patient_id == "MBL15" & Variant == "13886T>C"|
                patient_id == "CLL1" & Variant == "2857T>C"|
                patient_id == "CLL7" & Variant == "11435G>A"|
                patient_id == "CLL8" & Variant == "16519T>C")) %>% 
  dplyr::select(-My_annotation) %>%
  split(f = .$patient_id)

heteroplasmy_B <- list(
  readRDS(file.path(result_dir_subset1, "RDS_objects/umap_embedding_mito_cln.RDS")),
  readRDS(file.path(result_dir_subset1, "RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
  readRDS(file.path(result_dir_subset2, "RDS_objects/umap_embedding_mito_cln.RDS")),
  readRDS(file.path(result_dir_subset2, "RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
  readRDS(file.path(result_dir_subset3, "RDS_objects/umap_embedding_mito_cln.RDS")),
  readRDS(file.path(result_dir_subset3, "RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
  readRDS(file.path(result_dir_subset4, "RDS_objects/umap_embedding_mito_cln_MBL.RDS"))
) %>%
  bind_rows() %>%
  dplyr::select(patient_id, label, variant_id, cell_barcode, VAF) %>%
  tidyr::unite("label_cellbarcode", c(label, cell_barcode), sep = "#", remove = FALSE) %>%
  left_join(my_annotation_df, by = join_by(label_cellbarcode)) %>%
  dplyr::rename(
    "Heteroplasmy" = VAF,
    "Variant" = variant_id
  ) %>%
  dplyr::filter(
    Variant != "310T>C",
    patient_id != "MBL7dup",
    My_annotation %in% c("MBL_B_Cells", "CLL_B_Cells", "B"),
    !(patient_id == "CLL3" & Variant == "11711G>A"|
                patient_id == "MBL15" & Variant == "13886T>C"|
                patient_id == "CLL1" & Variant == "2857T>C"|
                patient_id == "CLL7" & Variant == "11435G>A"|
                patient_id == "CLL8" & Variant == "16519T>C")) %>%   
  split(f = .$patient_id)


metadata <- read_tsv(file = file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata.tsv"))
```

```{r}
patient_id2plotting_id <- metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(patient_id, plotting_id) %>%
  deframe()

label2plotting_id <- metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, plotting_id) %>%
  deframe()
```

# Plot distributions
## Individual plot per mtSNV
```{r, fig.width=15, fig.height=50}
variant_colours <- colorRampPalette(brewer.pal(12, "Set3"))(124)
```

```{r, fig.width=8, fig.height=40}
# find smallest non-zero value and set all zeros to this value * 10^-1
# convert_zeros_by_decade <- function(heteroplasmy_vector) {
#   heteroplasmy_vector %>%
#     sort() %>%
#     unique() %>%
#     .[. > 0] %>%
#     .[1] * 10^-1
# }

# what is smallest value across all patients and variants other than zero
heteroplasmy_long %>%
  bind_rows() %>%
  pull(Heteroplasmy) %>%
  unique() %>%
  sort() %>%
  .[. > 0]

zero_replacement <- 0.0001
```

```{r}
total_mtDNA_mutation_count_paired <- heteroplasmy_long[paste0("CLL", c(1:5, 7:10))] %>% 
  bind_rows() %>%
  dplyr::select(Variant) %>% 
  unique() %>% 
  pull(Variant) %>% 
  length

total_mtDNA_mutation_count_paired
```


```{r, fig.width=15, fig.height=15}
paired_het_plots <- list()
for (label in names(heteroplasmy_long[1:10])) {
  print(label)

  plot_length <- heteroplasmy_long[[label]]$Variant %>%
    unique() %>%
    length()

  distribution_paired <- heteroplasmy_long[[label]] %>%
    dplyr::mutate(Heteroplasmy = if_else(condition = Heteroplasmy == 0, true = zero_replacement, false = Heteroplasmy)) %>%
    dplyr::left_join(metadata %>% dplyr::filter(data_type == "ATAC") %>% dplyr::select(label, disease)) %>%
    ggplot(aes(x = Heteroplasmy, colour = disease)) +
    geom_freqpoly(linewidth = 1, binwidth = 0.1) +
    labs(y = "counts") +
    theme_bw() +
    scale_x_log10(label = scales::comma) +
    scale_y_log10() +
    scale_colour_manual(values = c("MBL" = "#FC8D62", "CLL" = "#66C2A5")) +
    theme_pub() +
    theme(legend.position = "none") +
    guides(colour = guide_legend(ncol = 1)) +
    facet_wrap(~Variant, scales = "free")

  paired_het_plots[[label]] <- distribution_paired

  # ggsave(
  #   plot = distribution_paired,
  #   path = file.path(plotting_dir, "Heteroplasmy_distributions"),
  #   filename = paste0("Heteroplasmy_distributions_", label, ".png"),
  #   width = 8 * 6,
  #   height = 2.42 * plot_length / 6,
  #   limitsize = FALSE,
  #   units = "cm"
  # )
}
paired_het_plots
```

```{r, fig.width=8, fig.height=40}
unpaired_het_plots <- list()
for (label in names(heteroplasmy_long[11:24])) {
  print(label)

  plot_length <- heteroplasmy_long[[label]]$Variant %>%
    unique() %>%
    length()

  distribution_unpaired <- heteroplasmy_long[[label]] %>%
    dplyr::mutate(Heteroplasmy = if_else(condition = Heteroplasmy == 0, true = zero_replacement, false = Heteroplasmy)) %>%
    dplyr::left_join(metadata %>% dplyr::filter(data_type == "ATAC") %>% 
                       dplyr::select(label, disease)) %>%
    ggplot(aes(x = Heteroplasmy, colour = disease)) +
    geom_freqpoly(size = 1, binwidth = 0.05) +
    labs(y = "counts") +
    scale_x_log10(label = scales::comma) +
    scale_y_log10() +
    theme_bw() +
    scale_colour_manual(values = c("MBL" = "#FC8D62", "CLL" = "#66C2A5")) +
    theme_pub() +
    theme(legend.position = "none") +
    guides(colour = guide_legend(ncol = 9)) +
    facet_wrap(Variant ~ ., scales = "free", ncol = 1, strip.position = "top")

  unpaired_het_plots[[label]] <- distribution_unpaired

  # ggsave(
  #   plot = distribution_unpaired,
  #   path = file.path(summary_dir, "Heteroplasmy_distributions"),
  #   filename = paste0("Heteroplasmy_distributions_", label, ".png"),
  #   width = 8,
  #   height = 2.42 * plot_length,
  #   units = "cm", limitsize = FALSE
  # )
}
unpaired_het_plots
```

# Count mtSNVs in paired and unpaired MBLs
```{r}
mtSNV_counts <- heteroplasmy_long %>%
  map(~ dplyr::select(., Variant) %>%
    unique() %>%
    pull() %>%
    length()) %>%
  bind_rows() %>%
  bind_cols(tibble(first_column = "counts"))

mtSNV_counts_plot <- mtSNV_counts %>%
  pivot_longer(-first_column, names_to = "patient_id", values_to = "mtSNVs [#]") %>%
  dplyr::select(-first_column) %>%
  dplyr::mutate(
    patient_id = factor(patient_id, level = c(paste0("MBL", 1:15), paste0("CLL", 1:10))),
    progression = if_else(str_detect(patient_id, "CLL"), "progressive", "not progressive")
  ) %>%
  ggplot(aes(x = patient_id, y = `mtSNVs [#]`, fill = progression)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, NA))) +
  scale_fill_manual(name = "Stage", values = c("progressive" = "#66C2A5", "not progressive" = "#FC8D62")) +
  labs(x = "", y = "mt mutation counts [#]", title = "mt mutation counts in disease progression") +
  theme_pub() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    legend.position = "right"
  )

ggsave(mtSNV_counts_plot,
  path = summary_dir,
  filename = "mtSNV_counts.png",
  width = 15,
  height = 8,
  units = "cm"
)
```

```{r}
mtSNV_counts %>%
  pivot_longer(-first_column, names_to = "patient_id", values_to = "mt mutation counts [#]") %>%
  dplyr::select(-first_column) %>%
  dplyr::mutate(
    patient_id = factor(patient_id, level = c(paste0("MBL", 1:15), paste0("CLL", 1:10))),
    progression = if_else(str_detect(patient_id, "CLL"), "progressive", "not progressive")
  ) %>%
  group_by(progression) %>%
  dplyr::summarize(Mean = mean(`mt mutation counts [#]`))

mtSNV_counts_prog <- mtSNV_counts %>%
  pivot_longer(-first_column, names_to = "patient_id", values_to = "mt mutation counts [#]") %>%
  dplyr::select(-first_column) %>%
  dplyr::mutate(
    patient_id = factor(patient_id, level = c(paste0("MBL", 1:15), paste0("CLL", 1:10))),
    progression = if_else(str_detect(patient_id, "CLL"), "progressive", "not progressive")
  ) %>%
  dplyr::filter(progression == "progressive") %>%
  pull(`mt mutation counts [#]`)

mtSNV_counts_np <- mtSNV_counts %>%
  pivot_longer(-first_column, names_to = "patient_id", values_to = "mt mutation counts [#]") %>%
  dplyr::select(-first_column) %>%
  dplyr::mutate(
    patient_id = factor(patient_id, level = c(paste0("MBL", 1:15), paste0("CLL", 1:10))),
    progression = if_else(str_detect(patient_id, "CLL"), "progressive", "not progressive")
  ) %>%
  dplyr::filter(progression == "not progressive") %>%
  pull(`mt mutation counts [#]`)

wilcox.test(x = mtSNV_counts_prog, y = mtSNV_counts_np, alternative = "two.sided")$p.value
```

## Evaluate heteroplasmy distributions
### QQ plots
```{r QQ_plot_function}
# split data into 100 percentiles
gg_qq_empirical <- function(a, b, quantiles = seq(0, 1, 0.01)) {
  df <- tibble(
    quantile = seq(0, 100, 1),
    MBL = quantile(b, quantiles),
    CLL = quantile(a, quantiles)
  )

  ggplot(
    data = df,
    mapping = aes(
      x = MBL,
      y = CLL,
      colour = quantile
    )
  ) +
    geom_abline(aes(slope = 1, intercept = 0), linetype = 2) +
    geom_point() +
    # scale_colour_distiller(palette = "Spectral") +
    scale_colour_gradientn(
      colours = c("#D9D9D9", "#B3B3B3", "#808080", "#666666", "#404040"),
      breaks = seq(0, 100, 25),
      labels = seq(0, 100, 25),
      limits = c(0, 100)
    ) +
    # scale_x_continuous(
    #   labels = c("0", "25", "50", "75", "100"),
    #   limits = c(0, 1),
    #   breaks = seq(0, 1, 0.25)
    # ) +
    scale_x_log10(
      labels = 10^seq(-3, 2, 1),
      limits = c(0.0001, 1),
      breaks = 10^seq(-5, 0, 1),
      expand = expansion(mult = c(0.05, 0.05))
    ) +
    # scale_y_continuous(
    #   labels = c("0", "25", "50", "75", "100"),
    #   limits = c(0, 1),
    #   breaks = seq(0, 1, 0.25)
    # ) +
    # ) +
    scale_y_log10(
      labels = 10^seq(-3, 2, 1),
      limits = c(0.0001, 1),
      breaks = 10^seq(-5, 0, 1),
      expand = expansion(mult = c(0.05, 0.05))
    ) +
    theme_pub() +
    theme(
      plot.title = element_blank(),
      legend.position = c(0.9, 0.35),
      aspect.ratio = 1
    ) +
    guides(color = guide_colorbar(barheight = 3, barwidth = 0.5)) +
    labs(
      colour = "Percentile",
      x = "MBL heteroplasmy [%]",
      y = "CLL heteroplasmy [%]"
    )
}
```

```{r het_filter_function}
# Filter heteroplasmy for only non-zero values for QQplot representation
filter_heteroplasmy <- function(het_df, disease_condition, variant_id, lower_het_threshold) {
  if (missing(lower_het_threshold)) {
    dplyr::mutate(het_df, label = case_when(str_detect(label, "CLL\\d+\\.1") ~ "MBL",
      str_detect(label, "CLL\\d+\\.2") ~ "CLL",
      .default = label
    )) %>%
      dplyr::filter(label == disease_condition & Variant == var) %>%
      dplyr::pull(Heteroplasmy)
  } else {
    dplyr::mutate(het_df, label = case_when(str_detect(label, "CLL\\d+\\.1") ~ "MBL",
      str_detect(label, "CLL\\d+\\.2") ~ "CLL",
      .default = label
    )) %>%
      dplyr::filter(label == disease_condition & Variant == variant_id & Heteroplasmy > lower_het_threshold) %>%
      dplyr::pull(Heteroplasmy)
  }
}
```

```{r QQ_plots_significant}
all_patient_var_plots <- list()
all_patient_statistics <- list()
for (patient in names(heteroplasmy_long)[c(1:6, 8:10)]) {
  # for (patient in names(heteroplasmy_long)[c(5,8)]) {
  print(patient)
  all_variants <- unique(heteroplasmy_long[[patient]]$Variant)
  all_var_plots <- list()
  all_var_statistics <- tibble(
    patient_id = as.character(),
    Variant = as.character(),
    mean_VAF_MBL = as.numeric(),
    total_cell_counts_MBL = as.numeric(),
    mean_VAF_CLL = as.numeric(),
    total_cell_counts_CLL = as.numeric(),
    p_value = as.numeric(),
    mean_VAF_MBL_non_zero = as.numeric(),
    cell_counts_MBL_non_zero = as.numeric(),
    mean_VAF_CLL_non_zero = as.numeric(),
    cell_counts_CLL_non_zero = as.numeric()
  )
  for (var in all_variants) {
    print(var)
    MBL_heteroplasmy <- filter_heteroplasmy(het_df = heteroplasmy_long[[patient]], disease_condition = "MBL", variant_id = var)
    mean_MBL_heteroplasmy <- mean(MBL_heteroplasmy)
    n_MBL_heteroplasmy <- length(MBL_heteroplasmy)

    CLL_heteroplasmy <- filter_heteroplasmy(het_df = heteroplasmy_long[[patient]], disease_condition = "CLL", variant_id = var)
    mean_CLL_heteroplasmy <- mean(CLL_heteroplasmy)
    n_CLL_heteroplasmy <- length(CLL_heteroplasmy)



    MBL_heteroplasmy_capped <- filter_heteroplasmy(het_df = heteroplasmy_long[[patient]], disease_condition = "MBL", variant_id = var, lower_het_threshold = 0)
    mean_MBL_heteroplasmy_capped <- mean(MBL_heteroplasmy_capped)
    n_MBL_heteroplasmy_capped <- length(MBL_heteroplasmy_capped)

    CLL_heteroplasmy_capped <- filter_heteroplasmy(het_df = heteroplasmy_long[[patient]], disease_condition = "CLL", variant_id = var, lower_het_threshold = 0)
    mean_CLL_heteroplasmy_capped <- mean(CLL_heteroplasmy_capped)
    n_CLL_heteroplasmy_capped <- length(CLL_heteroplasmy_capped)

    # perform kolmogorov smirnov test on full data distribution including all zeros
    ks_result_full_distribution <- ks.test(MBL_heteroplasmy, CLL_heteroplasmy)

    new_statistics <- tibble(
      patient_id = as.character(patient),
      Variant = as.character(var),
      mean_VAF_MBL = as.numeric(mean_MBL_heteroplasmy),
      total_cell_counts_MBL = as.numeric(n_MBL_heteroplasmy),
      mean_VAF_CLL = as.numeric(mean_CLL_heteroplasmy),
      total_cell_counts_CLL = as.numeric(n_CLL_heteroplasmy),
      p_value = as.numeric(ks_result_full_distribution$p.value),
      ks_statistic = as.numeric(ks_result_full_distribution$statistic),
      mean_VAF_MBL_non_zero = as.numeric(mean_MBL_heteroplasmy_capped),
      cell_counts_MBL_non_zero = as.numeric(n_MBL_heteroplasmy_capped),
      mean_VAF_CLL_non_zero = as.numeric(mean_CLL_heteroplasmy_capped),
      cell_counts_CLL_non_zero = as.numeric(n_CLL_heteroplasmy_capped)
    )

    all_var_statistics <- all_var_statistics %>%
      rbind(new_statistics)

    # create QQ plot only for mutations for which significant p value in ks test and at least 101 cells in one disease state or the other
    if ((n_MBL_heteroplasmy > 100 | n_CLL_heteroplasmy > 100) & ks_result_full_distribution$p.value <= 0.05) {
      qq <- gg_qq_empirical(a = CLL_heteroplasmy, b = MBL_heteroplasmy) +
        labs(
          title = paste0("QQ-Plot ", patient, "\n", var)
        )
      all_var_plots[[var]] <- qq
    }
  }
  all_patient_statistics[[patient]] <- all_var_statistics %>%
    dplyr::mutate("p_adjust" = p.adjust(p_value, method = "BH", n = length(.$p_value)))

  significant_vars <- all_var_statistics %>%
    dplyr::mutate("p_adjust" = p.adjust(p_value, method = "BH", n = length(.$p_value))) %>%
    dplyr::filter(p_adjust <= 0.05) %>%
    pull(Variant)

  all_patient_var_plots[[patient]] <- all_var_plots[significant_vars]
}

saveRDS(all_patient_var_plots$CLL4$`3424G>A`, file.path(RDS_dir, "all_patient_var_plots_CLL4_3424G>A.RDS"))
saveRDS(all_patient_var_plots$CLL9$`3777T>C`, file.path(RDS_dir, "all_patient_var_plots_CLL9_3777T>C.RDS"))
saveRDS(all_patient_var_plots, file = file.path(RDS_dir, "heteroplasmy_qqplots.RDS"))
saveRDS(all_patient_statistics, file = file.path(RDS_dir, "heteroplasmy_distribution_statistics.RDS"))

all_patient_statistics %>%
  map(~ dplyr::mutate(., "p_adjust" = p.adjust(p_value, method = "BH", n = length(.$p_value))) %>%
    dplyr::filter((cell_counts_MBL_non_zero > 100 | cell_counts_CLL_non_zero > 100) & p_adjust <= 0.05)) %>% bind_rows
```

```{r}
qqplot_data <- all_patient_var_plots %>% keep(~length(.) > 0) %>% 
  purrr::list_flatten() %>% 
  imap_dfr( ~{
  .x$data %>% 
    dplyr::mutate("patient_id_variant_id"=.y)
}) %>% 
  tidyr::separate(patient_id_variant_id, into = c("patient_id", "variant_id"), sep = "_",remove = FALSE)

write_tsv(x = qqplot_data, file=file.path(tsv_dir, "qqplot_data.tsv"), col_names = TRUE)
```

```{r save_QQplots, fig.width=6}
all_patient_var_plots <- all_patient_var_plots %>%
  list_drop_empty()

all_patient_var_plots %>%
  imap(safely(~ ggarrange(
    plotlist = .x,
    ncol = 2,
    nrow = ceiling(length(.x) / 2)
  ))) %>%
  purrr::transpose() %>%
  .$result %>%
  imap(safely(~ ggsave(.x,
    path = file.path(plotting_dir, "QQ_Plots"),
    filename = paste0("QQ_plots_", .y, ".png"),
    width = 12,
    limitsize = FALSE,
    height = ceiling(length(all_patient_var_plots[[.y]]) / 2) * 5
  )))
```

# Statistically significant variants vs. non-significant
### Evaluate KS statistics
```{r}
patientvariant2variant <- all_patient_statistics %>%
  imap(~ dplyr::select(
    .x, patient_id,
    Variant,
    p_adjust,
    ks_statistic,
    total_cell_counts_MBL,
    total_cell_counts_CLL
  ) %>%
    mutate(significant = if_else((total_cell_counts_MBL > 100 | total_cell_counts_CLL > 100) & p_adjust <= 0.05, "TRUE", "FALSE")) %>%
    dplyr::filter(significant == "TRUE")) %>%
  bind_rows() %>%
  dplyr::left_join(metadata %>%
    dplyr::select(patient_id, plotting_id) %>%
    unique()) %>%
  dplyr::select(plotting_id, patient_id, Variant, ks_statistic, p_adjust) %>%
  arrange(-ks_statistic) %>%
  dplyr::mutate(
    id = paste0(patient_id, "_", Variant),
    id = factor(id, levels = unique(id))
  ) %>%
  dplyr::select(id, Variant) %>%
  deframe()


D_value_signif_mtDNA_mutations <- all_patient_statistics %>%
  imap(
    ~ dplyr::select(
      .x, patient_id,
      Variant,
      p_adjust,
      ks_statistic,
      total_cell_counts_MBL,
      total_cell_counts_CLL
    ) %>%
      mutate(significant = if_else((total_cell_counts_MBL > 100 | total_cell_counts_CLL > 100) & p_adjust <= 0.05, "TRUE", "FALSE"))
    # %>%
    #   dplyr::filter(significant == "TRUE")
  ) %>%
  bind_rows() %>%
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id) %>% unique()) %>%
  dplyr::select(plotting_id, significant, patient_id, Variant, ks_statistic, p_adjust) %>%
  arrange(-ks_statistic) %>%
  dplyr::mutate(
    id = paste0(patient_id, "_", Variant),
    id = factor(id, levels = unique(id)),
    significant = factor(significant, levels = c("TRUE", "FALSE"))
  ) %>%
  dplyr::rename("ID" = patient_id, "KS statistic" = ks_statistic, "p-value" = p_adjust) %>%
  ggplot2::ggplot(aes(x = id, y = `KS statistic`, fill = significant)) +
  geom_bar(stat = "Identity") +
  labs(x = "mtDNA mutations") +
  scale_fill_manual(
    name = "KS test significance",
    values = c("TRUE" = "#956A77", "FALSE" = "lightgrey"),
    labels = c("TRUE" = "signficant", "FALSE" = "insignificant")
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, NA))) +
  scale_x_discrete(labels = patientvariant2variant) +
  theme_pub() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.9, 0.35)
  )
```

```{r fig.width=3, fig.height=3}
D_value_signif_mtDNA_mutations_scatter <- all_patient_statistics %>%
  imap(~ dplyr::select(.x, patient_id, Variant, p_adjust, ks_statistic, total_cell_counts_MBL, total_cell_counts_CLL) %>%
    mutate(significant = if_else((total_cell_counts_MBL > 100 | total_cell_counts_CLL > 100) & p_adjust <= 0.05, "TRUE", "FALSE"))) %>%
  bind_rows() %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(patient_id, plotting_id) %>% unique()) %>%
  dplyr::select(plotting_id, significant, patient_id, Variant, ks_statistic, p_adjust) %>%
  dplyr::mutate(
    significant = factor(significant, levels = c("TRUE", "FALSE")),
    label = if_else(significant == "TRUE", true = paste0(plotting_id, " ", Variant), false = NA)
  ) %>%
  ggplot(aes(x = ks_statistic, y = -log10(p_adjust), colour = significant, label = label)) +
  geom_point(size = 2) +
  geom_text_repel(colour = "black") +
  scale_y_continuous(expand = expansion(0, 0.1), oob = scales::squish_infinite) +
  scale_colour_manual(
    name = "KS test\nsignificance",
    values = c("TRUE" = "#956A77", "FALSE" = "lightgrey"),
    labels = c("TRUE" = "signficant", "FALSE" = "insignificant")
  ) +
  labs(x = "KS statistic", y = "-log10(FDR)") +
  theme_pub() +
  theme(legend.position = c(0.8, 0.2))

saveRDS(D_value_signif_mtDNA_mutations_scatter, file.path(RDS_dir, "D_value_signif_mtDNA_mutations.RDS"))
```

### Distributions of interest
```{r, fig.width=30, fig.height=12}
all_variants <- all_patient_statistics %>%
  bind_rows(.id = "patient_id")

variants_pass <- all_patient_statistics %>%
  map(~ dplyr::filter(., total_cell_counts_MBL > 100 | total_cell_counts_CLL > 100)) %>%
  bind_rows(.id = "patient_id")

variants_signif <- all_patient_statistics %>%
  map(~ dplyr::filter(., (total_cell_counts_MBL > 100 | total_cell_counts_CLL > 100) & p_adjust <= 0.05)) %>%
  bind_rows(.id = "patient_id") %>%
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id) %>% unique())


variants_insignif <- all_patient_statistics %>%
  map(~ dplyr::filter(., (total_cell_counts_MBL > 100 | total_cell_counts_CLL > 100) & p_adjust > 0.05)) %>%
  bind_rows(.id = "patient_id") %>%
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id))

print(paste0(
  "Out of ",
  dim(variants_pass)[1],
  " variants which had sufficient cell counts, ",
  dim(variants_signif)[1],
  " variant distributions seemed to originate from different distributions, which makes ",
  round(dim(variants_signif)[1] / dim(variants_pass)[1] * 100, digits = 2),
  "%."
))
```

```{r, fig.width=4, fig.height=1}
filter_input <- tibble(
  "filters" = c("all variants", "> 100  cell counts", "FDR <= 0.05"),
  "Variant counts [#]" = c(
    length(all_variants$Variant),
    dim(variants_pass)[1],
    dim(variants_signif)[1]
  ),
  "Percentage [%]" = c(
    100,
    round(dim(variants_pass)[1] / length(all_variants$Variant) * 100, digits = 2),
    round(dim(variants_signif)[1] / length(all_variants$Variant) * 100, digits = 2)
  )
)

filter_statistics <- ggtexttable(filter_input,
  theme = ttheme("light",
    base_size = 8,
    padding = unit(c(2, 2), "mm")
  ),
  rows = NULL
) +
  theme(plot.margin = margin(.1, .1, .1, .1, "cm"))

filter_statistics

saveRDS(object = filter_statistics, file = file.path(RDS_dir, "filter_statistics.RDS"))
```


```{r, fig.width=20, fig.height=9}
distributions_signif <- list()

for (row in 1:nrow(variants_signif)) {
  label <- variants_signif[row, ]$patient_id
  var <- variants_signif[row, ]$Variant
  plot_id <- variants_signif[row, ]$plotting_id

  xmin <- rep(zero_replacement, 2)
  xmax <- rep(1.1, 2)
  ymin <- rep(0, 2)
  ymax <- rep(8000, 2)

  distribution_paired <- heteroplasmy_long[[label]] %>%
    dplyr::filter(Variant == var) %>%
    dplyr::mutate(
      Heteroplasmy = if_else(condition = Heteroplasmy == 0, true = zero_replacement, false = Heteroplasmy),
      label = case_when(str_detect(label, "CLL\\d+\\.1") ~ "MBL",
        str_detect(label, "CLL\\d+\\.2") ~ "CLL",
        .default = label
      )
    ) %>%
    dplyr::mutate(label = factor(label, levels = c("MBL", "CLL"))) %>%
    ggplot(aes(x = Heteroplasmy * 100, colour = label)) +
    # annotate("rect", xmin=xmin*100, xmax=xmax*100, ymin=ymin*100, ymax=ymax*100, alpha=0.5, fill="lightgrey")+
    geom_freqpoly(linewidth = 1, binwidth = 0.05, alpha = 0.5) +
    labs(y = "cell counts", x = "Heteroplasmy [%]", title = paste0(plot_id, " ", var)) +
    scale_x_log10(labels = scales::label_comma()) +
    scale_y_log10(labels = scales::label_comma()) +
    theme_pub() +
    theme(legend.position = "none") +
    scale_colour_manual(values = c("MBL" = "#FC8D62", "CLL" = "#66C2A5")) +
    guides(colour = guide_legend(ncol = 9))

  distributions_signif[[paste0(label, "_", var)]] <- distribution_paired
}

ggarrange(plotlist = distributions_signif, nrow = 2, ncol = 4)

write_tsv(distributions_signif %>% 
            map(~ .$data) %>% 
  bind_rows(), file.path(tsv_dir, "distributions_signif_data.tsv"))
```

Normalised frequency and density of heteroplasmy
```{r, fig.width=20, fig.height=9}
density_distributions <- list()
for (row in 1:nrow(variants_signif)) {
  label <- variants_signif[row, ]$patient_id
  var <- variants_signif[row, ]$Variant
  plot_id <- variants_signif[row, ]$plotting_id

  density_distributions_variant <-
    heteroplasmy_long[[label]] %>%
    dplyr::mutate(Heteroplasmy = if_else(condition = Heteroplasmy == 0, true = zero_replacement, false = Heteroplasmy)) %>%
    dplyr::mutate(label = case_when(str_detect(label, "CLL\\d+\\.1") ~ "MBL",
      str_detect(label, "CLL\\d+\\.2") ~ "CLL",
      .default = label
    )) %>%
    dplyr::filter(Variant == var) %>%
    dplyr::mutate(label = factor(label, levels = c("CLL", "MBL"))) %>%
    ggplot(aes(
      x = Heteroplasmy * 100,
      colour = label
    )) +
    geom_density(bw = 0.03, linewidth = 1) +
    labs(y = "Density", x = "Heteroplasmy [%]", title = paste0(plot_id, " ", var)) +
    scale_x_log10(expand = expansion(mult = c(0.01, 0.01)), labels = scales::comma, breaks = c(0.001, 0.01, 0.1, 1, 10, 100)) +
    scale_y_log10(limits = c(10^(-9), 10), expand = expansion(mult = c(0, 0.01))) +
    theme_pub() +
    theme(legend.position = "bottom") +
    scale_colour_manual(values = c("CLL" = "#66C2A580", "MBL" = "#FC8D6280")) +
    guides(colour = guide_legend(ncol = 9))

  density_distributions[[paste0(label, "_", var)]] <- density_distributions_variant
}
ggarrange(plotlist = density_distributions, nrow = 2, ncol = 4, common.legend = T, legend = "bottom")

saveRDS(density_distributions$`CLL4_3424G>A`, file.path(RDS_dir, "density_distributions_CLL4_3424G>A.RDS"))
saveRDS(density_distributions$`CLL9_3777T>C`, file.path(RDS_dir, "density_distributions_CLL9_3777T>C.RDS"))

write_tsv(density_distributions %>% 
            map(~ .$data) %>% 
  bind_rows(), file.path(tsv_dir, "density_distribution_data.tsv"))
```

```{r, fig.width=20, fig.height=9}
# frequpoly normalises the area to 1 if the density is set as y value
normalised_distributions <- list()
for (row in 1:nrow(variants_signif)) {
  label <- variants_signif[row, ]$patient_id
  var <- variants_signif[row, ]$Variant
  plot_id <- variants_signif[row, ]$plotting_id

  normalised_distributions_variant <-
    heteroplasmy_long[[label]] %>%
    dplyr::mutate(Heteroplasmy = if_else(condition = Heteroplasmy == 0, true = zero_replacement, false = Heteroplasmy)) %>%
    dplyr::mutate(
      label = case_when(str_detect(label, "CLL\\d+\\.1") ~ "MBL",
        str_detect(label, "CLL\\d+\\.2") ~ "CLL",
        .default = label
      ),
      label = factor(label, levels = c("CLL", "MBL"))
    ) %>%
    dplyr::filter(Variant == var) %>%
    ggplot(aes(
      x = Heteroplasmy * 100,
      y = after_stat(density),
      colour = label
    )) +
    geom_freqpoly(binwidth = 0.08, linewidth = 1, alpha = 0.8) +
    labs(y = "Normalised counts", x = "Heteroplasmy [%]", title = paste0(plot_id, " ", var)) +
    scale_x_log10(expand = expansion(mult = c(NA, 0.01)), labels = scales::comma, breaks = c(0.001, 0.01, 0.1, 1, 10, 100)) +
    scale_y_log10(limits = c(0.000001, 100), expand = expansion(mult = c(0, 0.01))) +
    theme_pub() +
    theme(legend.position = "bottom") +
    scale_colour_manual(values = c("CLL" = "#66C2A5", "MBL" = "#FC8D62")) +
    guides(colour = guide_legend(ncol = 9))

  normalised_distributions[[paste0(label, "_", var)]] <- normalised_distributions_variant
}
ggarrange(plotlist = normalised_distributions, nrow = 2, ncol = 4, common.legend = T, legend = "bottom")
ggarrange(all_patient_var_plots$CLL4$`3424G>A`, all_patient_var_plots$CLL7$`11435G>A`, nrow = 1, ncol = 2)

saveRDS(normalised_distributions$`CLL4_3424G>A`, file.path(RDS_dir, "normalised_distributions_CLL4_3424G>A.RDS"))
saveRDS(normalised_distributions$`CLL7_11435G>A`, file.path(RDS_dir, "normalised_distributions_CLL7_11435G>A.RDS"))
```

## Comparison with clinical data
```{r}
WBCC_data <- metadata %>%
  dplyr::filter(
    data_type == "ATAC",
    str_detect(patient_id, "CLL")
  ) %>%
  split(.$plotting_id) %>%
  imap(~ dplyr::summarize(.x, WBCC_diff = .[.$disease == "CLL", ] %>%
    pull(`WBC_[cells/nl]`) - .[.$disease == "MBL", ] %>%
    pull(`WBC_[cells/nl]`)) %>%
    dplyr::mutate(plotting_id = .y)) %>%
  bind_rows()

sample_time <- metadata %>%
  dplyr::filter(
    data_type == "ATAC",
    str_detect(
      patient_id, "CLL"
    )
  ) %>%
  dplyr::select(plotting_id, disease, sample_date) %>%
  tidyr::pivot_wider(names_from = disease, values_from = sample_date) %>%
  dplyr::rename(
    "sample_date_MBL" = MBL,
    "sample_date_CLL" = CLL
  ) %>%
  dplyr::mutate("time between samples" = lubridate::time_length(difftime(sample_date_CLL, sample_date_MBL), "years"))

variants_signif_meta <- variants_signif %>%
  left_join(WBCC_data) %>%
  left_join(sample_time) %>%
  tidyr::unite(patient_id_variant, patient_id:Variant, sep = "_", remove = FALSE) %>%
  dplyr::mutate(patient_id = factor(patient_id, levels = paste0("CLL", 1:10))) %>%
  unique()
```

### Compare difference in WBCC and non-zero means to see if mean heteroplasmy changes with disease severity
```{r}
all_variants_long <- all_patient_statistics %>%
  map(~ dplyr::mutate(., significance = case_when((total_cell_counts_MBL > 100 | total_cell_counts_CLL > 100) & p_adjust <= 0.05 ~ TRUE, .default = NA))) %>%
  bind_rows() %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(patient_id, plotting_id) %>%
    unique()) %>%
  left_join(WBCC_data) %>%
  left_join(sample_time) %>%
  tidyr::unite(patient_id_variant, patient_id:Variant, sep = "_", remove = FALSE) %>%
  dplyr::mutate(patient_id = factor(patient_id, levels = paste0("CLL", 1:10))) %>%
  unique() %>%
  dplyr::mutate(
    label_colour = if_else(significance == TRUE, true = patient_id, false = NA),
    label_name = if_else(significance == TRUE, true = Variant, false = NA)
  )
```

```{r fig.height=4, fig.width=4}
WBCC_scatter_fc <- all_variants_long %>%
  dplyr::mutate(
    diff_means = mean_VAF_CLL - mean_VAF_MBL,
    fc_means = mean_VAF_CLL / mean_VAF_MBL,
    significance=case_when(significance == TRUE ~"significant",.default = "insignificant"),
    significance=factor(significance, levels=c("significant", "insignificant")),
    label=if_else(significance =="significant", paste0(patient_id2plotting_id[patient_id], " ", Variant), NA)) %>%
  dplyr::select(patient_id, Variant, significance, WBCC_diff, diff_means, fc_means, label) %>%
  unique() %>%
  dplyr::arrange(desc(significance)) %>% 
ggplot2::ggplot(aes(x = WBCC_diff, y = fc_means, colour = significance, label=label))+
  geom_point(size = 0.5) +
  geom_smooth(
    method = lm,
    colour = "black",
    fill = "lightgrey",
    linetype = 2,
    se = TRUE,
    alpha = 0.3,
    level = 0.95
  ) +
  geom_text_repel(colour="black", size = 2.5,
    fontface = "bold", force=900) +
  scale_y_continuous(limits = c(-10, 60), 
                     breaks = seq(-20, 60, 20), 
                     expand = expansion(mult = c(0.01, 0.01))) +
  scale_x_continuous(breaks = seq(0, 80, 20), 
                     expand = expansion(mult = c(0.1, 0.05))) +
  scale_colour_manual(name = "Significance",
                      values = c("significant"="#956A77", "insignificant"="lightgray")) +
  labs(y = "Fold change heteroplasmy (CLL / MBL)", x = "WBC increase [cells/nl]") +
  theme_pub() +
  theme(aspect.ratio = 1) +
  guides(colour = guide_legend(override.aes = list(size = 2.5)))
WBCC_scatter_fc

write_tsv(WBCC_scatter_fc$data, file.path(tsv_dir, "WBCC_scatter.tsv"))
```

```{r}
duration_scatter <-all_variants_long %>%
  dplyr::mutate(
    diff_means = mean_VAF_CLL - mean_VAF_MBL,
    fc_means = mean_VAF_CLL / mean_VAF_MBL,
    significance=case_when(significance == TRUE ~"significant",.default = "insignificant"),
    significance=factor(significance, levels=c("significant", "insignificant")),
    label=if_else(significance =="significant", paste0(patient_id2plotting_id[patient_id], " ", Variant), NA))%>%
  dplyr::select(`time between samples`, diff_means, fc_means, label_name, label_colour, significance, label) %>%
  unique() %>%
  dplyr::arrange(desc(significance)) %>% 
ggplot2::ggplot(aes(x = `time between samples`, y = fc_means, colour = significance, label=label))+
  geom_point(size = 0.5) +
  geom_smooth(
    method = lm,
    colour = "black",
    fill = "lightgrey",
    linetype = 2,
    se = TRUE,
    alpha = 0.3,
    level = 0.95
  ) +
  geom_text_repel(colour="black", size = 2.5,
    fontface = "bold", force=900) +
  scale_y_continuous(limits = c(-10, 40),
                     breaks = seq(0, 40, 20),
                     expand = expansion(mult = c(0.01, 0.01))) +
  scale_x_continuous(breaks = seq(0, 8, 2),
                     expand = expansion(mult = c(0.1, 0.05))) +
  scale_colour_manual(name = "Significance",
                      values = c("significant"="#956A77", "insignificant"="lightgray")) +
  labs(y = "Fold change heteroplasmy (CLL / MBL)", x = "Sampling period [years]") +
  theme_pub() +
  theme(aspect.ratio = 1) +
  guides(colour = guide_legend(override.aes = list(size = 2.5)))

duration_scatter
write_tsv(duration_scatter$data, file.path(tsv_dir, "duration_scatter.tsv"))
```

### Corresponding coverage of selected distributions of interest
```{r}
library <- metadata %>%
  dplyr::filter(
    data_type == "ATAC",
    str_starts(label, "CLL")
  ) %>%
  select(label, pool) %>%
  deframe()

all_cells_coverage <- list()
selected_variants <- variants_signif %>%
  dplyr::filter(Variant == "3424G>A" & patient_id == "CLL4" |
    Variant == "3777T>C" & patient_id == "CLL9")

for (var in selected_variants$Variant) {
  var <- var
  pat <- selected_variants %>%
    dplyr::filter(Variant == var) %>%
    pull(patient_id)

  labels <- paste0(pat, c(".1", ".2"))
  pos <- str_extract(var, "\\d+")

  for (lab in labels) {
    lib <- read_csv(file.path(project_dir, "data/ATAC", paste0(library[lab], ".mgatk"), paste0(library[lab], ".coverage.txt.gz")), col_names = c("position", "cellbarcode", "coverage")) %>%
      dplyr::filter(position == pos) %>%
      dplyr::mutate(
        cellbarcode = paste0(lab, "#", cellbarcode),
        label = lab,
        variant = var
      ) %>%
      dplyr::filter(cellbarcode %in% heteroplasmy_long[[pat]]$label_cellbarcode)

    all_cells_coverage[[lab]] <- lib
  }
}
all_cells_coverage <- bind_rows(all_cells_coverage)
saveRDS(all_cells_coverage, file.path(RDS_dir, "all_cells_coverage.RDS"))
```

#### Coverage plots
```{r}
coverage_paired <- all_cells_coverage %>%
  dplyr::mutate(
    disease = case_when(str_detect(label, "CLL\\d+\\.1") ~ "MBL",
      str_detect(label, "CLL\\d+\\.2") ~ "CLL",
      .default = label
    ),
    disease = factor(disease, levels = c("CLL", "MBL"))
  ) %>%
  tidyr::separate(label, c("patient_id", NA), sep = "\\.", remove = FALSE) %>%
  split(.$patient_id) %>%
  imap(~ ggplot(.x, aes(x = coverage, colour = disease)) +
    geom_density(linewidth = 1, bw = 1) +
    labs(y = "Density", x = "sequencing depth", title = .y) +
    theme_pub() +
    scale_y_continuous(expand = expansion(mult = c(0, NA))) +
    scale_x_continuous(expand = expansion(mult = c(0, NA))) +
    theme(legend.position = "none") +
    scale_colour_manual(values = c("MBL" = "#FC8D62", "CLL" = "#66C2A5")) +
    facet_wrap(~variant, ncol = 1))

coverage_paired %>%
  ggarrange(
    plotlist = .,
    ncol = 1
  ) %>%
  ggsave(
    plot = .,
    filename = "Coverage_mutations_oi.png",
    path = summary_dir,
    width = 10,
    height = 35
  )

saveRDS(coverage_paired$CLL4, file.path(RDS_dir, "coverage_paired_CLL4.RDS"))
saveRDS(coverage_paired$CLL9, file.path(RDS_dir, "coverage_paired_CLL9.RDS"))
```

# Mutations between B and (pre-)malignant cells
## Unique mutations
```{r}
heteroplasmy_B_samples <- heteroplasmy_B %>%
  map(~ dplyr::filter(., Heteroplasmy > 0) %>%
    group_by(patient_id, label, My_annotation) %>%
    dplyr::summarize(counts = n())) %>%
  bind_rows() %>%
  ungroup() %>%
  group_by(label) %>%
  dplyr::filter(sum(counts[My_annotation == "B"]) >= 100) %>%
  pull(label) %>%
  unique()
```

```{r}
celltype_heteroplasmy_samples <- heteroplasmy_B %>%
  map(~ 
    dplyr::filter(.x, Heteroplasmy > 0) %>%
    dplyr::filter(
      label %in% heteroplasmy_B_samples,
      My_annotation %in% c("MBL_B_Cells", "B", "CLL_B_Cells")
    )) %>%
  keep(function(x) nrow(x) > 0) %>%
  map(function(x) {
    split_df <- split(x, x$My_annotation)

    unique_B <- base::setdiff(split_df[["B"]][["Variant"]], split_df[["MBL_B_Cells"]][["Variant"]])
    unique_MBL <- base::setdiff(split_df[["MBL_B_Cells"]][["Variant"]], split_df[["B"]][["Variant"]])
    unique_CLL <- base::setdiff(split_df[["CLL_B_Cells"]][["Variant"]], split_df[["B"]][["Variant"]])
    return(list(B = unique_B, MBL_B_Cells = unique_MBL, CLL_B_Cells = unique_CLL))
  })


MBL4 <- heteroplasmy_B %>%
  map(~ 
    dplyr::filter(.x, Heteroplasmy > 0) %>%
    dplyr::filter(
      label %in% heteroplasmy_B_samples,
      My_annotation %in% c("MBL_B_Cells", "B", "CLL_B_Cells")
    ))
```

```{r}
unique_mtDNA_mutations <- celltype_heteroplasmy_samples %>%
  map(purrr::discard, is.null) %>%
  map_dfr(map_dbl, length, .id = "patient_id") %>%
  dplyr::mutate()
```

```{r}
unique_mtDNA_mutations %>%
  tidyr::pivot_longer(-patient_id, names_to = "cell_type", values_to = "unique_mutations") %>%
  dplyr::mutate(
    cell_type = factor(cell_type, levels = c("B", "MBL_B_Cells", "CLL_B_Cells")),
    plotting_id = patient_id2plotting_id[patient_id],
    plotting_id = factor(plotting_id, levels = paste0("P", 1:24))
  ) %>%
  arrange(plotting_id) %>%
  dplyr::mutate(patient_id = factor(patient_id, unique(patient_id))) %>%
  ggplot(aes(x = patient_id, y = unique_mutations, fill = cell_type)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(values = annotation_colours_named, labels = c("B" = "B", "MBL_B_Cells" = "MBL", "CLL_B_Cells" = "CLL")) +
  scale_x_discrete(labels = patient_id2plotting_id) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  labs(x = "Patient", y = "Unique mutation counts", fill = "Cell type") +
  theme_pub() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5),
    axis.title.x = element_blank()
  )
```
