---
title: "Surface markers from TotalSeqA"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
GOAL: The aim of this script is to investigate the surface marker expression of patients sequenced with ASAPseq.

# Load libraries
```{r}
library(tidyverse)
library(ArchR)
library(gplots)
library(Seurat)
library(RColorBrewer)
library(ggpubr)
```

# Set paths
```{r}
source("config_paths.R")
data_dir <- file.path(project_dir_ext, "kallisto_BUS_files")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "Surface_marker_analysis")
plotting_dir <- file.path(analysis_dir, "Plots")
RDS_dir <- file.path(analysis_dir, "RDS_Objects")

ArchR_dir <- "ArchR_35_samples/ArchR_all_samples_subsetted"
result_dir_summaries <- file.path(project_dir, "Summary_Plots")
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r}
CLLProj <- loadArchRProject(path = file.path(project_dir_ext, ArchR_dir))
metadata <- read_tsv(file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata.tsv"))
metadata_tsa <- read_tsv(file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata_tsa.tsv"))
TSA_targets <- read_csv(file.path(data_dir, "TSA_barcodes/TSA_universal_cocktail_reference.csv"), col_names = FALSE)[, 1] %>% 
  pull(X1)

label2plotting_id <- metadata_tsa %>% 
  dplyr::select(label, plotting_id) %>% 
  deframe()
```

```{r}
libraries <- metadata_tsa %>%
  dplyr::filter(!(technology == "multiome kit")) %>%
  select(pool, label) %>%
  deframe()

import_kite_counts <- function(dir) {
  mtx <- readMM(file.path(dir, "featurecounts.mtx"))
  colnames(mtx) <- readLines(file.path(dir, "featurecounts.genes.txt"))
  rownames(mtx) <- readLines(file.path(dir, "featurecounts.barcodes.txt"))
  return(t(mtx))
}

all_TSA_data <- list()
for (lib in metadata_tsa %>%
  dplyr::filter(!(technology == "multiome kit")) %>%
  pull(pool)) {
  print(libraries[lib])

  TSA_raw <- import_kite_counts(file.path(data_dir, lib, "featurecounts"))

  TSA_so <- CreateSeuratObject(TSA_raw, assay = "HTO")
  TSA_so <- NormalizeData(TSA_so, assay = "HTO", normalization.method = "CLR")
  TSA_so <- ScaleData(TSA_so, assay = "HTO")

  TSA_values <- as_tibble(t(GetAssayData(TSA_so)), rownames = NA) %>%
    rownames_to_column("cellbarcode") %>%
    dplyr::mutate(cellbarcode = paste0(libraries[lib], "#", cellbarcode, "-1")) %>%
    column_to_rownames("cellbarcode")

  all_TSA_data[[libraries[lib]]] <- TSA_values
}

all_TSA_data <- bind_rows(all_TSA_data)
```

Add TSA values as metadata_tsa to ArchR project
```{r}
sample_cells <- rownames(all_TSA_data)
for (ab in colnames(all_TSA_data)) {
  CLLProj <- addCellColData(
    ArchRProj = CLLProj,
    data = all_TSA_data[, ab],
    cells = sample_cells,
    name = ab
  )
}
```

Visualise target proteins
```{r}
TSA_targets_filtered <- colnames(CLLProj@cellColData)[34:197]

umap_coordinates <- CLLProj@embeddings@listData[["UMAP_IterativeLSI"]]@listData[["df"]] %>%
  as_tibble(rownames = "label_cellbarcode") %>%
  dplyr::left_join(as.data.frame(CLLProj@cellColData) %>%
    dplyr::select(any_of(c(TSA_targets_filtered))) %>%
    rownames_to_column("label_cellbarcode")) %>%
  tidyr::separate(label_cellbarcode, c("label", NA), sep = "#", remove = FALSE) %>%
  tidyr::separate(label, c("patient_id", "sample_id"), sep = "\\.", remove = FALSE) %>%
  tidyr::pivot_longer(
    cols = -c("label_cellbarcode", "label", "patient_id", "sample_id", "IterativeLSI#UMAP_Dimension_1", "IterativeLSI#UMAP_Dimension_2"),
    names_to = "target",
    values_to = "value"
  ) %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "GEX") %>%
    dplyr::select(label, patient_id, disease))
```

UMAPs
```{r}
# TSA_umaps <- umap_coordinates %>%
#   arrange(value) %>%
#   dplyr::mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>%
#   ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = value)) +
#   geom_point(size = .1, stroke = NA) +
#   coord_equal() +
#   scale_colour_distiller(
#     name = "",
#     palette = "YlOrRd",
#     na.value = "lightgrey",
#     direction = 1,
#     limits = c(0, 10),
#     breaks = c(0, 2, 4, 6, 8, 10)
#   ) +
#   theme_pub() +
#   theme(
#     legend.position = "bottom",
#     legend.direction = "horizontal",
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     axis.line = element_line(),
#     aspect.ratio = 1,
#     panel.grid = element_blank(),
#     panel.border = element_blank()
#   ) +
#   labs(x = "UMAP1", y = "UMAP2") +
#   facet_grid(cols = vars(disease), rows = vars(target))

# ggsave(
#   plot = TSA_umaps,
#   path = plotting_dir,
#   filename = "TSA_umaps.png",
#   width = 5,
#   height = 134,
#   limitsize = FALSE
# )
```

```{r}
TSA_selected_umaps <- umap_coordinates %>%
  dplyr::filter(target %in% c("CD79b", "CD19", "CD5", "IgM")) %>%
  dplyr::mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>%
  arrange(value) %>%
  ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = value)) +
  geom_point(size = .1, stroke = NA) +
  coord_equal() +
  scale_colour_distiller(
    name = "",
    palette = "YlOrRd",
    na.value = "lightgrey",
    direction = 1,
    limits = c(0, 10),
    breaks = c(0, 2, 4, 6, 8, 10)
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.line = element_line(),
    aspect.ratio = 1,
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  labs(x = "UMAP1", y = "UMAP2") +
  facet_grid(rows = vars(disease), cols = vars(target))
```

Violinplots
```{r, fig.width=25, fig.height=20}
TSA_vplots <- umap_coordinates %>%
  split(.$patient_id) %>%
  # .[-c(2,3)] %>%
  .[c(12, 13, 14, 15, 16, 17, 23)] %>%
  map(~ dplyr::mutate(., disease = factor(disease, levels = c("MBL", "CLL"))) %>%
    dplyr::filter(value != 0 & !is.na(value)) %>%
    ggplot(aes(x = target, y = value, fill = disease)) +
    geom_violin() +
    coord_equal() +
    scale_y_continuous(expand = expansion(mult = c(0, NA))) +
    scale_fill_manual(
      name = "Condition",
      values = my_MBL_CLL_colours_named
    ) +
    theme_pub() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, vjust = 0.9, hjust = 1),
      legend.direction = "horizontal"
    ) +
    labs(x = "", y = "Abundance"))

TSA_vplots %>%
  imap(~ ggsave(
    plot = .,
    path = plotting_dir,
    filename = paste0("TSA_vplots_", .y, ".png"),
    width = 100,
    height = 3,
    limitsize = FALSE
  ))
```

## Differential surface marker expression healthy and premalignant B cells
```{r}
B_cell_barcodes <- tibble(
  "label_cellbarcode" = CLLProj$cellNames,
  "label" = CLLProj$Sample,
  "My_annotation" = CLLProj$My_annotation
) %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(label, MBL_subtype)) %>%
  dplyr::filter(
    MBL_subtype == "LC",
    My_annotation == "B"
  ) %>%
  pull(label_cellbarcode)

premalignant_barcodes <- tibble(
  "label_cellbarcode" = CLLProj$cellNames,
  "label" = CLLProj$Sample,
  "My_annotation" = CLLProj$My_annotation
) %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(label, MBL_subtype)) %>%
  dplyr::filter(
    MBL_subtype == "LC",
    My_annotation == "MBL_B_Cells"
  ) %>%
  pull(label_cellbarcode)
```

```{r}
fc_expression_pp <- all_TSA_data %>%
  rownames_to_column("label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, into = c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::filter(label_cellbarcode %in% c(B_cell_barcodes, premalignant_barcodes)) %>%
  dplyr::mutate(celltype = case_when(
    label_cellbarcode %in% B_cell_barcodes ~ "B",
    label_cellbarcode %in% premalignant_barcodes ~ "LC-MBL"
  )) %>%
  tidyr::pivot_longer(-c(label_cellbarcode, label, celltype), names_to = "surface_protein", values_to = "value") %>%
  dplyr::group_by(surface_protein, label, celltype) %>%
  dplyr::summarise_at(vars(value), list(`Mean expression` = mean)) %>%
  tidyr::pivot_wider(names_from = celltype, values_from = `Mean expression`) %>%
  dplyr::mutate(fold_change = `LC-MBL` / `B`)

fc_expression <- all_TSA_data %>%
  rownames_to_column("label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, into = c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::filter(label_cellbarcode %in% c(B_cell_barcodes, premalignant_barcodes)) %>%
  dplyr::mutate(celltype = case_when(
    label_cellbarcode %in% B_cell_barcodes ~ "B",
    label_cellbarcode %in% premalignant_barcodes ~ "LC-MBL"
  )) %>%
  tidyr::pivot_longer(-c(label_cellbarcode, label, celltype), names_to = "surface_protein", values_to = "value") %>%
  dplyr::group_by(surface_protein, celltype) %>%
  dplyr::summarise_at(vars(value), list(`Mean expression` = mean)) %>%
  tidyr::pivot_wider(names_from = celltype, values_from = `Mean expression`) %>%
  dplyr::mutate(fold_change = `LC-MBL` / `B`)

# Calculate p-values between LC-MBL and B cell surface marker expression
p_values_expression_surface_marker_pp <- all_TSA_data %>%
  rownames_to_column("label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, into = c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::filter(label_cellbarcode %in% c(B_cell_barcodes, premalignant_barcodes)) %>%
  dplyr::mutate(celltype = case_when(
    label_cellbarcode %in% B_cell_barcodes ~ "B",
    label_cellbarcode %in% premalignant_barcodes ~ "LC-MBL"
  )) %>%
  tidyr::pivot_longer(-c(label_cellbarcode, label, celltype), names_to = "surface_protein", values_to = "value") %>%
  dplyr::group_by(label, surface_protein) %>%
  do(p_value = wilcox.test(dplyr::filter(., celltype == "LC-MBL") %>% pull(value),
    dplyr::filter(., celltype == "B") %>% pull(value),
    paired = FALSE
  )$p.value) %>%
  dplyr::mutate(
    p_value = unlist(p_value)
  ) %>%
  dplyr::mutate("p_adjust" = p.adjust(p_value, method = "BH", n = length(.$p_value)))

p_values_expression_surface_marker <- all_TSA_data %>%
  rownames_to_column("label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, into = c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::filter(label_cellbarcode %in% c(B_cell_barcodes, premalignant_barcodes)) %>%
  dplyr::mutate(celltype = case_when(
    label_cellbarcode %in% B_cell_barcodes ~ "B",
    label_cellbarcode %in% premalignant_barcodes ~ "LC-MBL"
  )) %>%
  tidyr::pivot_longer(-c(label_cellbarcode, label, celltype), names_to = "surface_protein", values_to = "value") %>%
  dplyr::group_by(surface_protein) %>%
  do(p_value = wilcox.test(dplyr::filter(., celltype == "LC-MBL") %>% pull(value),
    dplyr::filter(., celltype == "B") %>% pull(value),
    paired = FALSE
  )$p.value) %>%
  dplyr::mutate(
    p_value = unlist(p_value)
  ) %>%
  dplyr::mutate("p_adjust" = p.adjust(p_value, method = "BH", n = length(.$p_value)))

statistics_tsa_surface_markers <- fc_expression %>%
  left_join(p_values_expression_surface_marker, by = join_by(surface_protein)) %>%
  dplyr::mutate(
    significance = case_when(p_adjust <= 1e-100 & log2(fold_change) >= 0.5~ "LC-MBL",
                             p_adjust <= 1e-100 & log2(fold_change) <= -0.5 ~ "B",
                             .default = "not significant"),
    surface_marker_label = case_when(significance != "not significant" ~ surface_protein, .default = NA)
  )

statistics_tsa_surface_markers_pp <- fc_expression_pp %>%
  left_join(p_values_expression_surface_marker_pp, by = join_by(label, surface_protein)) %>%
  dplyr::mutate(
    significance = case_when(p_adjust <= 1e-100 & log2(fold_change) >= 0.5 ~ "LC-MBL",
                             p_adjust <= 1e-100 & log2(fold_change) <= -0.5 ~ "B",
                             .default = "not significant"
    ),
    surface_marker_label = case_when(significance != "not significant" ~ surface_protein, .default = NA)
  )
```

```{r}
volcanoplot_summary_TSA_pp <- statistics_tsa_surface_markers_pp %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(label, plotting_id) %>% unique()) %>%
  dplyr::mutate(surface_marker_label=case_when(!is.na(surface_marker_label) ~ paste(surface_marker_label, "_", plotting_id), .default = NA)) %>% 
  ggplot(aes(x = log2(fold_change), y = -log10(p_adjust), colour = significance, label = surface_marker_label)) +
  geom_vline(xintercept = c(-0.5, 0.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = -log10(1e-100), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 200,
    colour = "black"
  ) +
  scale_x_continuous("Median difference",
  oob = scales::squish_infinite,
  limits = c(-2.5, 2.5)
  ) +
  scale_y_continuous("-log10(FDR)",
    limits = c(0, 330),
    oob = scales::squish_infinite,
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_colour_manual(
    name = "Differentially expressed cell surface marker",
    values = c("B" = "darkred", "LC-MBL" = "#fcbda4", "not significant" = "#D3D3D3")
  ) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_summary_TSA <- statistics_tsa_surface_markers %>%
  # dplyr::mutate(surface_marker_label=case_when(!is.na(surface_marker_label) ~ paste(surface_marker_label), .default = NA)) %>% 
  ggplot(aes(x = log2(fold_change), y = -log10(p_adjust), colour = significance, label = surface_marker_label)) +
  geom_vline(xintercept = c(-0.5, 0.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(1e-100)), color = "grey", linetype = "dashed") +
  geom_point(size=1) +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 200,
    colour = "black"
  ) +
  scale_x_continuous("log2(Fold change)",
     oob = scales::squish_infinite,
    limits = c(-2.5, 2.5),
    breaks=c(-4,-2,0,2,4)
  ) +
  scale_y_continuous("-log10(FDR)",
    limits = c(0, 330),
    oob = scales::squish_infinite,
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_colour_manual(
    name = "Differentially expressed cell surface marker",
    values = c("B" = "darkred", "LC-MBL" = "#fcbda4", "not significant" = "#D3D3D3"),
    labels=c("B"="Physiologic B Cells", "LC-MBL"="LC-MBL Cells")
  ) +
  labs(title="B vs. LC-MBL")+
  theme_pub() +
  theme(legend.position = "bottom",
        aspect.ratio = 1,
        plot.title = element_text(face="plain"))

volcanoplot_summary_TSA
volcanoplot_summary_TSA_pp

ggsave(volcanoplot_summary_TSA_pp,
       filename=file.path(plotting_dir, "TSA_volcanoplots.png"),
       width = 10, 
       height = 4,
       unit="in")
```

```{r}
saveRDS(volcanoplot_summary_TSA, file.path(RDS_dir, "volcanoplot_TSA_LC_MBL_B.RDS"))
```

## Surface marker expression for DEPs across all patients
```{r}
DEPs <- statistics_tsa_surface_markers %>% 
  dplyr::filter((p_adjust<=1e-100 & log2(fold_change) >=0.5) | (p_adjust <= 1e-100 & log2(fold_change) <= -0.5 )) %>% 
  dplyr::pull(surface_marker_label) %>% 
  unique()

#range of values -2.1715	0.4178442
#FDR range 1 3.20e-114       1
```

```{r, fig.width=5.5, fig.height=4}
dotplot_surfacemarker <- statistics_tsa_surface_markers_pp %>% 
  dplyr::filter(surface_protein %in% DEPs | surface_protein %in% c( "CD19")) %>% 
  dplyr::select(surface_protein, label, fold_change, p_adjust) %>% 
  ungroup() %>% 
  dplyr::arrange(fold_change, p_adjust) %>% 
  dplyr::filter(fold_change != "NaN") %>% 
  dplyr::mutate(surface_protein=factor(surface_protein, unique(surface_protein))) %>% 
  ggplot(aes(x=surface_protein, y=label, colour= log2(fold_change), size=-log10(`p_adjust`)))+
  geom_point()+
  scale_color_gradient2(name="Expression",
                        high="darkmagenta",
                        mid="white",
                        low=  "darkolivegreen",
                        midpoint=0,
                        breaks=c(-3, -1.5, 0,1.5,3),
                        limits=c(-3.1,3.1),
                        oob=scales::squish_infinite)+
  scale_y_discrete(labels=label2plotting_id)+
  scale_size_area(name="-log10(FDR)",
                  max_size=3,
                  limits=c(NA,10),
                  oob=scales::squish,
                  labels=c("0.0", "2.5", "5.0", "7.5", "≥ 10")
                  )+
  labs(y="", x="")+
  theme_pub()+
  theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1),
        panel.background = element_rect(fill="white"),
        plot.title = element_text(hjust=0.5),
        legend.position = c(0.5, 0.5))

saveRDS(dotplot_surfacemarker, file.path(RDS_dir, "Dotplot_surfacemarker.RDS"))
# 
# ggsave(plot = dotplot_surfacemarker, 
#        file=file.path(plotting_dir, "Dotplot_surfacemarker.png"), 
#        width = 5.5, 
#        height = 4)
```

```{r fig.width=10, fig.height=4}
cowplot::plot_grid(volcanoplot_summary_TSA+
                     theme(legend.position = "bottom"), 
                   dotplot_surfacemarker,
                   rel_widths = c(1,2)) %>% 
  ggsave(.,
       file=file.path(plotting_dir, "Surfacemarker_figure.png"), 
       width = 10, 
       height = 4, 
       dpi = "print")
```

## Celltype quantification
```{r}
cell_annotation <- all_TSA_data %>%
  rownames_to_column("label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, into = c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::filter(label_cellbarcode %in% c(B_cell_barcodes, premalignant_barcodes)) %>%
  dplyr::mutate(celltype = case_when(
    label_cellbarcode %in% B_cell_barcodes ~ "B",
    label_cellbarcode %in% premalignant_barcodes ~ "LC-MBL"
  )) %>% 
  dplyr::select(label_cellbarcode, celltype)


cell_type_counts <- cell_annotation %>% 
  tidyr::separate(label_cellbarcode, into=c("label", "cellbarcode"), remove = FALSE, sep="#") %>% 
  dplyr::left_join(metadata_tsa %>% dplyr::select(MBL_subtype, label)) %>% 
  group_by(label, celltype) %>% 
  dplyr::summarize(counts=n())
```

```{r, fig.height=4, fig.width=4}
cell_type_counts_plot <- cell_type_counts%>%
  pivot_wider(id_cols = label, names_from = celltype, values_from = counts, values_fill = NA) %>%
  tidyr::pivot_longer(-label, names_to = "celltype", values_to = "counts") %>%
  dplyr::left_join(metadata_tsa %>% dplyr::select(label, plotting_id, disease, MBL_subtype)) %>%
  tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE) %>%
  dplyr::mutate(MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("LC_MBL", "HC_MBL", "CLL")),
    MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("LC_MBL", "HC_MBL", "CLL")),
    celltype = case_when(celltype == "LC-MBL" & MBL_CLL_subtype == "LC_MBL" ~ "LC_MBL_B_Cells",
    .default = celltype),
    counts = if_else(celltype == "B", -counts, counts),
    celltype = factor(celltype, levels = c("B", "LC_MBL_B_Cells")),
    plotting_id = factor(plotting_id, levels=rev(paste0("P", 1:24))),
    label=if_else(counts < 0, counts*-1, counts)) %>%
  ggplot(aes(y = plotting_id, x = counts, fill = celltype)) +
  geom_bar(position = "stack", stat = "identity", width = 0.7) +
  geom_text(aes(label=label), hjust=0)+
  scale_x_continuous(
     breaks = c(-2000, 0, 2000, 4000, 6000, 8000),
     labels = c("2000", "0", "2000", "4000", "6000", "8000"),
     limits = c(-2000, 8000)
  ) +
  scale_fill_manual(
    name = "Cell type", values = annotation_colours_named,
    labels = c("B" = "B cells", "LC_MBL_B_Cells" = "LC-MBL cells")
  ) +
  labs(x = "cell counts", title = "") +
  theme_pub() +
  theme(legend.position = "bottom",
    axis.text.x = element_text(angle=45, hjust=1, vjust=1),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(angle = 0),
    plot.title = element_text(margin = margin(t = 8, b = -20))
  ) 

ggsave(plot = cell_type_counts_plot, 
       file.path(plotting_dir, "Celltype_counts_TSA.pdf"),
       width=4, 
       height=4, 
       units = "in", 
       bg = "white")
```

