---
title: "ATAC_Healthy_Clustering"
author: "Anja_Rathgeber"
date: "`r Sys.Date()`"
output: html_document
---

# GOAL: The aim of this script is to find find out how clustering looks like if healthy PBMC samples are added to malignant MBL/CLL samples

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(readxl)
library(ggrepel)
library(glue)
library(ggpubr)
library(ggplot2)
library(dplyr)
library(SeuratDisk)
library(Seurat)
library(SeuratObject)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Cairo)
library(harmony)
library(patchwork)
library(BiocParallel)
library(circlize)
library(magick)
library(ArchR)
library(readr)
library(tibble)
```

```{r Run_parameters}
set.seed(1)
addArchRThreads(threads = 1)
addArchRGenome("hg38")
```

# Set paths

```{r }
source("config_paths.R")
data_dir <- file.path(project_dir_remote, "data")
library_dir <- file.path(project_dir_remote, "scratch/ATAC")
results_dir <- file.path(project_dir_remote, "results")
analysis_dir <- file.path(results_dir, "ArchR_analysis")

IGHV_analysis_dir <- file.path(results_dir, "IGHV_analysis")
IGHV_analysis_tsv_dir <- file.path(IGHV_analysis_dir, "tsv_data")

ArchR_proj <- file.path("ArchR_Healthy_35_samples")
ArchR_proj_ndb <- file.path("ArchR_Healthy_35_samples_no_doublets")
ArchR_proj_healthy_3 <- file.path("ArchR_Healthy_3")

ArchR_dir_healthy_35 <- file.path(analysis_dir, ArchR_proj)
ArchR_dir_healthy_35_no_doublets <- file.path(analysis_dir, ArchR_proj_ndb)
ArchR_dir_healthy_3 <- file.path(analysis_dir, ArchR_proj_healthy_3)
```

# Load data

```{r}
source(file.path(scripts_dir_remote, "publication_theme.R"))
```

```{r }
metadata_malignant <- read_tsv(file = file.path(data_dir, "20240723_Tidy_MBL_CLL_metadata.tsv")) %>% 
  dplyr::filter(patient_id !="MBL7dup")

metadata_healthy_atac <- read_tsv(file.path(data_dir, "20250812_Healthy_controls_mtscATACseq.tsv"))
                           #"SRR12476607", "SRR12476609", "SRR12476610" # Mimitou et al (2021) Nat Biotech healthy sample
                           #"SRR11539033", "SRR11539034"  # Laureau et al healthy sample

metadata <- metadata_malignant %>% 
  full_join(metadata_healthy)

patient_id2plotting_id <- metadata %>%
  select(patient_id, plotting_id) %>%
  unique() %>%
  deframe

label2IGHV_inferred <- read_tsv(file.path(IGHV_analysis_tsv_dir, "20251010_IGHV_status_inferred.tsv")) %>% 
  dplyr::select(label, IGHV_inferred) %>% unique() %>% deframe()

My_annotation <- read_tsv(file.path(analysis_dir, "CLLProj_35_samples_subsetted_My_annotation.tsv")) %>% 
  dplyr::mutate(My_annotation_LC=case_when(str_starts(label_cellbarcode, "MBL10|MBL11|MBL12|MBL13|MBL14|MBL15")                      & My_annotation =="MBL_B_Cells" ~ "LC_MBL_B_Cells",
                                           str_starts(label_cellbarcode, pattern="MBL10|MBL11|MBL12|MBL13|MBL14|MBL15", negate=TRUE) & My_annotation =="MBL_B_Cells" ~ "HC_MBL_B_Cells", 
                                          .default=My_annotation)) %>% 
  dplyr::select(label_cellbarcode, My_annotation_LC) 

celltype_annotation_healthy <- read_tsv(file.path(results_dir, "20250303_ATAC_celltype_annotation_healthy.tsv"), col_names = TRUE) %>%
  dplyr::filter(label != "pbmc1")
```

```{r}
# setwd(file.path(analysis_dir, "arrow_files_unprocessed"))
# 
# # do below only for new healthy controls and then pass all paths to arrowpaths
# sample_names <- metadata %>%
#   dplyr::filter(data_type == "ATAC") %>%
#   dplyr::select(label, pool) %>%
#   dplyr::arrange(pool) %>%
#   dplyr::pull(var = label, name = pool)
# 
# # Select library_file_paths based on sample_names
# library_file_paths <- c(file.path(project_dir_remote, "scratch/ATAC", names(sample_names), "fragments.tsv.gz"))
# 
# # commented out as arrow files already created from the Subset projects
# ArrowFiles <- createArrowFiles(
#   outputNames = sample_names,
#   inputFiles = library_file_paths,
#   sampleNames = sample_names,
#   minTSS = 4,
#   minFrags = 1000,
#   addTileMat = TRUE,
#   addGeneScoreMat = TRUE,
#   threads = 1,
#   subThreading = FALSE
# )

# ArrowFiles <- file.path(analysis_dir, "arrow_files_unprocessed", paste0(sample_names, ".arrow"))
```

# Doublet Inference for healthy patients 
In malignant cells already removed in other objects and filtered for them

```{r Doublet_Inference}
# setwd(file.path(analysis_dir, "arrow_files_unprocessed"))
# 
# doubScores <- addDoubletScores(
#   input = ArrowFiles[c(35,36)],
#   k = 10, # Refers to how many cells near a "pseudo-doublet" to count.
#   knnMethod = "UMAP", # Refers to the embedding to use for nearest neighbor search with doublet projection.
#   LSIMethod = 1, 
#   threads = 1
# )
```

# Create ArchR Project

```{r Make_ArchRProject}
# setwd(analysis_dir)
# 
# CLLProj_Healthy <- ArchRProject(
#   ArrowFiles = file.path(analysis_dir, "arrow_files_unprocessed", paste0(sample_names, ".arrow")),
#   outputDirectory = ArchR_proj,
#   copyArrows = TRUE,
#   threads = 1
# )
# 
# saveArchRProject(
#   ArchRProj = CLLProj_Healthy,
#   outputDirectory = ArchR_proj,
#   load = FALSE,
#   overwrite = TRUE)
# 
# CLLProj_Healthy <- loadArchRProject(path = file.path(ArchR_dir_healthy_35))
# 
# annotation <- tibble("label_cellbarcode"=CLLProj_Healthy$cellNames,
#                      "label"=CLLProj_Healthy$Sample) %>%
#   left_join(metadata %>% dplyr::filter(data_type=="ATAC") %>% 
#               dplyr::select(label, patient_id, technology, sequencing_date)) %>% 
#   left_join(My_annotation) %>% 
#   dplyr::mutate(My_annotation_LC=if_else(condition=is.na(My_annotation_LC), true=NA, false=My_annotation_LC))
```

```{r remove_doublets}
# setwd(analysis_dir)
# 
# CLLProj_Healthy <- filterDoublets(ArchRProj = CLLProj_Healthy)
# 
# saveArchRProject(
#   ArchRProj = CLLProj_Healthy,
#   outputDirectory = ArchR_proj,
#   load = FALSE,
#   overwrite = TRUE
# )
```

# Subset to no doublets
```{r}
# malignant data set has previously already been filtered for doublets, this way same filtering is kept
# no_doublet_cells <- CLLProj_Healthy$cellNames[CLLProj_Healthy$cellNames %in% My_annotation$label_cellbarcode | grepl("^(mimitou1|laureau1)", CLLProj_Healthy$cellNames)]
# 
# CLLProj_Healthy_no_db <- subsetArchRProject(
#   ArchRProj = CLLProj_Healthy,
#   cells = no_doublet_cells,
#   outputDirectory = ArchR_dir_healthy_35_no_doublets,
#   dropCells = FALSE)

CLLProj_Healthy_no_db <- loadArchRProject(ArchR_dir_healthy_35_no_doublets)
```


# LSI Dimensionality Reduction

```{r LSI_Dimensionality_Reduction}
# setwd(analysis_dir)
# 
# CLLProj_Healthy_no_db <- addIterativeLSI(
#   seed = 1,
#   ArchRProj = CLLProj_Healthy_no_db,
#   useMatrix = "TileMatrix",
#   name = "IterativeLSI",
#   iterations = 2,
#   clusterParams = list( # See Seurat::FindClusters
#     resolution = c(0.2),
#     sampleCells = 10000,
#     n.start = 10
#   ),
#   varFeatures = 25000,
#   dimsToUse = 1:30, threads = 1
# )
# 
# CLLProj_Healthy_no_db <- addIterativeLSI(
#   seed = 1,
#   ArchRProj = CLLProj_Healthy_no_db,
#   useMatrix = "TileMatrix",
#   name = "IterativeLSI_0.4",
#   iterations = 2,
#   clusterParams = list( # See Seurat::FindClusters
#     resolution = c(0.4),
#     sampleCells = 10000,
#     n.start = 10),
#   varFeatures = 25000,
#   dimsToUse = 1:30, threads = 1
# )
# 
# saveArchRProject(ArchRProj = CLLProj_Healthy_no_db,
#                  outputDirectory = ArchR_proj_ndb,
#                  load = FALSE,
#                  overwrite = TRUE)
```

# Batch correction

```{r}
# setwd(analysis_dir)
# 
# CLLProj_Healthy_no_db <- addCellColData(
#   ArchRProj = CLLProj_Healthy_no_db,
#   data = annotation %>%
#     mutate(sequencing_date=as.character(sequencing_date)) %>%
#     pull(sequencing_date),
#   cells = annotation %>% pull(label_cellbarcode),
#   name = "sequencing_date",
# )

# CLLProj_Healthy_no_db <- addHarmony(
#   ArchRProj = CLLProj_Healthy_no_db,
#   reducedDims = "IterativeLSI",
#   name = "Harmony_IterativeLSI",
#   groupBy = c("sequencing_date"), 
#   verbose = TRUE
# )
# 
# CLLProj_Healthy_no_db <- addHarmony(
#   ArchRProj = CLLProj_Healthy_no_db,
#   reducedDims = "IterativeLSI_0.4",
#   name = "Harmony_IterativeLSI_0.4",
#   groupBy = c("sequencing_date"), 
#   verbose = TRUE
# )

# saveArchRProject(ArchRProj = CLLProj_Healthy_no_db,   
#                  outputDirectory = ArchR_proj_ndb, 
#                  load = FALSE, 
#                  overwrite = TRUE)
```

# Gene Score Imputation

```{r}
CLLProj_Healthy_no_db <- addImputeWeights(CLLProj_Healthy_no_db)

markerGenes <- c(
  "CD34", # Early Progenitor
  "GATA1", # Erythroid
  "PAX5", "MS4A1", "EBF1", "MME", # B-Cell Trajectory
  "CD14", "CEBPB", "MPO", "S100A8", "S100A9", "SPI1", "LYZ",   "IRF8", # Monocytes
  "IL7R", "CD3D", "CD3E", "CD3G", # General T cell markers
  "CD4", # CD4+ T cells
  "CD8A", "CD8B", # CD8+ T cells
  "FOXP3", "IL2RA", # Tregs
  "CCR7", "SELL", # Memory/naïve T cells
  "CD69", "IFNG", "IL2", # T cell Activation markers
  "TBX21", "GATA3", "RORC", "CTLA4", # T subset-specific markers
  "CD19", "CD5", "FCER2", "CD79b", # CLL B-Cell Marker
  "GNLY", "NKG7", "NCAM1", "KLRD1", "PRF1", "KLRK1", "KLRC1", "FCGR3A", "ZBTB16", "IL2RB" # NK cells
)
```

# Seurat Clustering

```{r}
setwd(analysis_dir)

# Try combinations between different LSI resolutions and clustering resolutions
param_combinations <- tidyr::expand_grid(
  dimred = c(
    "IterativeLSI",
    "IterativeLSI_0.4",
    "Harmony_IterativeLSI", "Harmony_IterativeLSI_0.4"
  ),
  resolution = c(0.2, 0.3, 0.4, 0.6)
)

for (row in 1:nrow(param_combinations)) {
  resolution <- param_combinations$resolution[[row]]
  dimred <- param_combinations$dimred[[row]]

  CLLProj_Healthy_no_db <- addClusters(
    input = CLLProj_Healthy_no_db,
    seed = 1,
    reducedDims = dimred,
    method = "Seurat",
    name = glue("Clusters_{resolution}_{dimred}"),
    resolution = resolution
  )
}
saveArchRProject(
  ArchRProj = CLLProj_Healthy_no_db,
  outputDirectory = ArchR_proj_ndb,
  load = FALSE,
  overwrite = TRUE
)
```

# UMAP

```{r UMAP_Embedding}
setwd(analysis_dir)

dimred_df <- tidyr::expand_grid(
  dimred = c(
    "IterativeLSI",
    "IterativeLSI_0.4",
   "Harmony_IterativeLSI",
   "Harmony_IterativeLSI_0.4"
  )
)

for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  CLLProj_Healthy_no_db <- addUMAP(
    seed = 1,
    ArchRProj = CLLProj_Healthy_no_db,
    reducedDims = dimred,
    name = glue("UMAP_{dimred}"),
    nNeighbors = 30,
    minDist = 0.5,
    metric = "cosine"
  )
}

saveArchRProject(ArchRProj = CLLProj_Healthy_no_db,
                 outputDirectory = ArchR_proj_ndb,
                 load = FALSE,
                 overwrite = TRUE)
```

# Label transfer
```{r}
setwd(analysis_dir)

# Hao_seRNA <- LoadH5Seurat(file.path(data_dir, "reference_genomes/Hao_and_Hao_et_al_2021_PBMC/pbmc_multimodal.h5seurat"))
# Hao_seRNA

# rna_counts <- CreateAssayObject(Hao_seRNA@assays[["SCT"]]@counts)
# 
# NormalizeData(
#   rna_counts,
#   normalization.method = "LogNormalize",
#   scale.factor = 10000,
#   margin = 1,
#   verbose = TRUE
# )
# 
# Hao_seRNA@assays[["RNA"]] <- rna_counts

Hao_seRNA <- readRDS(file.path(data_dir, "reference_genomes/Hao_and_Hao_et_al_2021_PBMC/Hao_seRNA.RDS"))

CLLProj_Healthy_no_db <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj_Healthy_no_db,
  useMatrix = "GeneScoreMatrix",
  matrixName = "GeneIntegrationMatrix_IterativeLSI",
  reducedDims = "IterativeLSI",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "predictedCell_Un_Hao_IterativeLSI_,",
  nameGroup = "predictedGroup_Un_Hao_IterativeLSI",
  nameScore = "predictedScore_Un_Hao_IterativeLSI",
  threads = 1,
  force=TRUE)

saveArchRProject(ArchRProj = CLLProj_Healthy_no_db,
                 outputDirectory = ArchR_proj_ndb,
                 load = FALSE,
                 overwrite = TRUE)

CLLProj_Healthy_no_db <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj_Healthy_no_db,
  useMatrix = "GeneScoreMatrix",
  matrixName = "GeneIntegrationMatrix_IterativeLSI_0.4",
  reducedDims = "IterativeLSI_0.4",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "predictedCell_Un_Hao_IterativeLSI_0.4",
  nameGroup = "predictedGroup_Un_Hao_IterativeLSI_0.4",
  nameScore = "predictedScore_Un_Hao_IterativeLSI_0.4",
  threads = 1,
  force=TRUE)

saveArchRProject(ArchRProj = CLLProj_Healthy_no_db,
                 outputDirectory = ArchR_proj_ndb,
                 load = FALSE,
                 overwrite = TRUE)
```

## Visualization
### Sample annotation
```{r , fig.width=8, fig.height=8}
umap_samples <- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  name = "Sample",
  embedding = "UMAP_IterativeLSI",
  plotAs = "points",
    pal = c(my_paired_colours_named, my_unpaired_colours_named, control_colours_named)
) +
  labs(x = "UMAP1", y = "UMAP2", colour="Sample") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggsave(umap_samples, 
       filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_samples.png"), 
                              units="in",
                  width = 8,
                  height = 8)
umap_samples
```

### Patient annotation
```{r}
annotation <- tibble("label_cellbarcode"=CLLProj_Healthy_no_db$cellNames,
                     "label"=CLLProj_Healthy_no_db$Sample) %>%
  left_join(metadata %>% dplyr::filter(data_type=="ATAC") %>% 
              dplyr::select(label, patient_id, technology, sequencing_date, IGHV))%>% 
  left_join(My_annotation) %>% 
  dplyr::mutate(My_annotation_LC=if_else(condition=is.na(My_annotation_LC), true=NA, false=My_annotation_LC)) %>% 
  dplyr::mutate(IGHV_inferred=if_else(is.na(IGHV), label2IGHV_inferred[label], IGHV))


CLLProj_Healthy_no_db <- addCellColData(
  ArchRProj = CLLProj_Healthy_no_db,
  data = annotation %>% pull(patient_id),
  cells = annotation %>% pull(label_cellbarcode),
  name = "patient_id"
)
```

```{r , fig.width=5, fig.height=5}
umap_coordinates <- CLLProj_Healthy_no_db@embeddings@listData[["UMAP_IterativeLSI"]]@listData[["df"]] %>% 
  as_tibble(rownames = "label_cellbarcode")

patient_annotation <- CLLProj_Healthy_no_db@cellColData %>%
  as_tibble(rownames = "label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, patient_id)

celltype_annotation <- CLLProj_Healthy_no_db@cellColData %>%
  as_tibble(rownames = "label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, My_annotation_healthy)

umap_patient_annotation <- patient_annotation %>%
  dplyr::left_join(umap_coordinates) %>% 
  dplyr::left_join(celltype_annotation)

patient_annotation_umap <- umap_patient_annotation %>%
  dplyr::filter(patient_id != "MBL7dup") %>%
  ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = patient_id)) +
  geom_point(size = 0.5, stroke = NA) +
  coord_equal() +
  scale_colour_manual(name = "Patients", values = c(my_unpaired_colours_named, control_colours_named)) +
  labs(x = "UMAP1", y = "UMAP2", title = "Patients") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal",
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(
    keywidth = 0.5,
    keyheight = 0.5,
    override.aes = list(size = 3)
  ))

patient_annotation_umap

ggsave(patient_annotation_umap, filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_patients.png"), 
                              units="in",
                  width = 4,
                  height = 4)
```

```{r}
umap_data_no_immune <- umap_patient_annotation %>%  
  left_join(metadata %>% dplyr::select(patient_id, plotting_id)) %>% 
  dplyr::filter(patient_id != "CLL6") %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, levels=c(paste0("P", 1:24),paste0("H", 1:4)))) %>% 
  dplyr::arrange(plotting_id) %>% 
  dplyr::mutate(patient_id=factor(patient_id, levels=unique(patient_id))) %>% 
  dplyr::filter(!(My_annotation_healthy %in% c("T", "other", "Mono", "NK", "DC")))

centroids <- umap_data_no_immune %>% 
  dplyr::select(`IterativeLSI#UMAP_Dimension_1`, `IterativeLSI#UMAP_Dimension_2`, patient_id) %>% 
  group_by(patient_id) %>% 
  dplyr::summarize(across(everything(), mean)) %>% 
  dplyr::left_join(metadata %>% dplyr::filter(data_type=="ATAC") %>% dplyr::select(patient_id, plotting_id)) %>% unique()
```

```{r}
patient_annotation_umap <- umap_patient_annotation %>% 
  dplyr::filter(patient_id != "MBL7dup") %>%
  ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = patient_id)) +
  geom_point(size = 0.5, stroke = NA) +
  geom_text_repel(data=centroids, aes(x=`IterativeLSI#UMAP_Dimension_1`, y=`IterativeLSI#UMAP_Dimension_2`, label=plotting_id),colour="black",  force=300, max.overlaps = 200)+
  coord_equal() +
  scale_colour_manual(name = "Patients", values = c(my_unpaired_colours_named, control_colours_named)) +
  labs(x = "UMAP1", y = "UMAP2", title = "") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "none",
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(
    keywidth = 0.5,
    keyheight = 0.5,
    override.aes = list(size = 3)
  ))

```

```{r}
write_tsv(centroids, file = file.path(ArchR_dir_healthy_35_no_doublets, "tsv_data/20250325_ATAC_patient_annotation_healthy_centroids.tsv"), 
  col_names = TRUE )

write_tsv(patient_annotation_umap$data %>% 
  dplyr::rename("UMAP1"=`IterativeLSI#UMAP_Dimension_1`, 
                "UMAP2"=`IterativeLSI#UMAP_Dimension_2`) %>% 
    tidyr::separate(label_cellbarcode, into=c("label", "cellbarcode"), sep="#", remove = FALSE) %>% 
    dplyr::left_join(metadata %>%  dplyr::filter(data_type=="ATAC") %>% 
                       dplyr::select(label, MBL_CLL_subtype)), 
  file = file.path(ArchR_dir_healthy_35_no_doublets, "tsv_data/20250325_ATAC_celltype_annotation_healthy.tsv"), 
  col_names = TRUE )
```

```{r}
patient_annotation_split_umap <- umap_patient_annotation %>%
  dplyr::left_join(metadata %>% dplyr::select(patient_id, MBL_CLL_subtype)) %>% 
  dplyr::mutate(MBL_CLL_subtype=factor(MBL_CLL_subtype, levels=c("Healthy", "LC", "HC", "CLL"))) %>% 
  dplyr::filter(patient_id != "MBL7dup") %>%
  ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = patient_id)) +
  geom_point(size = 0.5, stroke = NA) +
  coord_equal() +
  scale_colour_manual(name = "Patients", values = c(my_unpaired_colours_named, control_colours_named)) +
  labs(x = "UMAP1", y = "UMAP2", title = "Patients") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal",
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(
    keywidth = 0.5,
    keyheight = 0.5,
    override.aes = list(size = 3)
  ))+
  facet_wrap(~MBL_CLL_subtype, nrow = 1)

write_tsv(patient_annotation_split_umap$data, file.path(ArchR_dir_healthy_35_no_doublets, "tsv_data/patient_annotation_split_umap_data.tsv"))
```


### Cluster annotation
```{r fig.width=5, fig.height=5}
# Try combinations between different LSI resolutions and clustering resolutions
param_combinations_umap <- tidyr::expand_grid(
  dimred = c("IterativeLSI", "IterativeLSI_0.4"),
  resolution = c(0.2, 0.3, 0.4, 0.6)
) %>% mutate("umap" = case_when(
  dimred == "IterativeLSI" ~ "",
  dimred == "IterativeLSI_0.4" ~ "_LSI_0.4"
))

create_clusters <- function() {
  all_plots <- list()
  for (row in 1:nrow(param_combinations_umap)) {
    resolution <- param_combinations_umap$resolution[[row]]
    dimred <- param_combinations_umap$dimred[[row]]
    umap <- param_combinations_umap$umap[[row]]

    plot <- plotEmbedding(
      ArchRProj = CLLProj_Healthy_no_db,
      colorBy = "cellColData",
      name = glue("Clusters_{resolution}_{dimred}"),
      embedding = glue("UMAP_{dimred}"),
      plotAs = "points"
    ) +
      labs(x = "UMAP1", y = "UMAP2") +
      scale_color_discrete(
        name = "Cluster",
        type = cluster_colours
      ) +
      theme_bw() +
      theme(
        strip.text = element_text(face = "bold", size = rel(1.2)),
        strip.background = element_rect(fill = "white", colour = NA),
        panel.border = element_blank(),
        axis.line = element_line(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_text(face = "bold"),
        panel.grid = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = "bottom",
        legend.direction = "horizontal"
      ) +
      guides(colour = guide_legend(override.aes = list(size = rel(2))))

    print(plot)
    all_plots[[row]] <- plot
  }

  return(all_plots)
}

all_clusters <- create_clusters()

umaps_clusters <- cowplot::plot_grid(ggpubr::ggarrange(plotlist = all_clusters, nrow = 2, ncol = 4))
ggsave(umaps_clusters, filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAPs_clusters.png"), 
                              units="in",
                  width = 16,
                  height = 15)

all_clusters[[1]]
```


### Celltype annotation
```{r}
CLLProj_Healthy_no_db <- addCellColData(
  ArchRProj = CLLProj_Healthy_no_db,
  data = annotation %>% dplyr::filter(!is.na(My_annotation_LC) & label_cellbarcode %in% CLLProj_Healthy_no_db$cellNames) %>% pull(My_annotation_LC),
  cells = annotation %>% dplyr::filter(!is.na(My_annotation_LC) & label_cellbarcode %in% CLLProj_Healthy_no_db$cellNames) %>%  pull(label_cellbarcode),
  name = "My_annotation",
)

umap_celltype <- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  name = "My_annotation",
  embedding = "UMAP_IterativeLSI",
  pal = annotation_colours_named,
  plotAs = "points") +
  labs(title="Cell type", x = "UMAP1", y = "UMAP2", colour="Cell type") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical"
  )
umap_celltype

ggsave(umap_celltype, filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_celltype.png"), 
                              units="in",
                  width = 4,
                  height = 4)
```

### Technology annotation
```{r}
CLLProj_Healthy_no_db <- addCellColData(
  ArchRProj = CLLProj_Healthy_no_db,
  data = annotation %>% pull(technology),
  cells = annotation %>% pull(label_cellbarcode),
  name = "technology")

```

```{r}
umap_technology <- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  name = "technology",
  embedding = "UMAP_IterativeLSI",
  plotAs = "points") +
  labs(title="Technology", x = "UMAP1", y = "UMAP2", colour="Technology") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical"
  )

umap_technology

ggsave(umap_technology, 
       filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_technology.png"), 
                              units="in",
                  width = 4,
                  height = 4)
```

### Sequencing date annotation
```{r}
CLLProj_Healthy_no_db <- addCellColData(
  ArchRProj = CLLProj_Healthy_no_db,
  data = annotation %>%
    mutate(sequencing_date=as.character(sequencing_date)) %>%
    pull(sequencing_date),
  cells = annotation %>% pull(label_cellbarcode),
  name = "sequencing_date",
)

umap_seqdate <- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  name = "sequencing_date",
  embedding = "UMAP_IterativeLSI",
  plotAs = "points") +
  labs(title="Sequencing date", x = "UMAP1", y = "UMAP2", colour="Sequencing date") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical"
  )
umap_seqdate

ggsave(umap_seqdate, filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_sequencingdate.png"), 
                              units="in",
                  width = 4,
                  height = 4)
```

### IGHV mutational status inferred

```{r}
CLLProj_Healthy_no_db <- addCellColData(
  ArchRProj = CLLProj_Healthy_no_db,
  data = annotation %>%
    dplyr::mutate(IGHV_inferred=as.character(IGHV_inferred)) %>%
    pull(IGHV_inferred),
  cells = annotation %>% pull(label_cellbarcode),
  name = "IGHV_inferred",
)

IGHV_inferred <- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  pal = IGHV_colours,
  name = "IGHV_inferred",
  embedding = "UMAP_IterativeLSI",
  plotAs = "points") +
  labs(title="IGHV mutational status", x = "UMAP1", y = "UMAP2", colour="IGHV") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical"
  )
cowplot::plot_grid(patient_annotation_umap+theme(legend.position = "none"), IGHV_inferred+theme(legend.position="none"), nrow=1, ncol=2)

write_tsv(IGHV_inferred$data, file.path(DCA_analysis_tsv_dir, "ATAC_healthy_IGHV_inferred_data.tsv"))
```

### GeneScores

```{r fig.width=8, fig.height=22}
setwd(analysis_dir)

UMAP_marker_genes <- plotEmbedding(
    ArchRProj = CLLProj_Healthy_no_db, 
    colorBy = "GeneScoreMatrix", 
    name = markerGenes, 
    embedding = "UMAP_IterativeLSI",
    quantCut = c(0.01, 0.95),
    imputeWeights = getImputeWeights(CLLProj_Healthy_no_db),
    plotAs = "points"
)

UMAP_marker_genes_figure <- lapply(UMAP_marker_genes, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
#cowplot::plot_grid(plotlist=UMAP_marker_genes_figure, ncol = 4)

cowplot::plot_grid(plotlist=UMAP_marker_genes_figure, ncol = 4 ) %>% 
  ggsave(filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_marker_genes_figure_batches.png"), 
                              units="in",
                  width = 8,
                  height = 22)

cowplot::plot_grid(plotlist=UMAP_marker_genes_figure, ncol = 4 ) 
```

```{r fig.width=5, fig.height=5}
B_cell_markers <- c( "CCR7", # Naïve B cells
  "CD27", "CD80", # Memory B cells
  "CD38", "PRDM1", "XBP1", "SLAMF7", # Plasma cells
  "CD40", "CD86", "HLA-DRA", "HLA-DRB1" # Activation markers
  )

UMAP_B_cells <- plotEmbedding(
    ArchRProj = CLLProj_Healthy_no_db, 
    colorBy = "GeneScoreMatrix", 
    name = B_cell_markers, 
    embedding = "UMAP_IterativeLSI",
    quantCut = c(0.01, 0.95),
    imputeWeights = getImputeWeights(CLLProj_Healthy_no_db),
    plotAs = "points")

UMAP_B_cells

UMAP_qc <- plotEmbedding(
    ArchRProj = CLLProj_Healthy_no_db,
    name = "TSSEnrichment",
    colorBy = "cellColData", 
    embedding = "UMAP_IterativeLSI",
    quantCut = c(0.01, 0.95),
    plotAs = "points")

UMAP_qc
```

### Predicted Celltype

```{r}
umap_predicted_celltype<- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  name = "predictedGroup_Un_Hao_IterativeLSI",
  embedding = "UMAP_IterativeLSI",
  pal = annotation_colours_named,
  plotAs = "points") +
  labs(title="Cell type", x = "UMAP1", y = "UMAP2", colour="Cell type") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical"
  )

umap_predicted_celltype

ggsave(umap_predicted_celltype, 
       filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_predicted_celltype.png"), 
                              units="in",
                  width = 4,
                  height = 4)
```

### Prediction Score
```{r}
umap_predicted_score<- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  name = "predictedScore_Un_Hao_IterativeLSI",
  embedding = "UMAP_IterativeLSI",
  plotAs = "points") +
  labs(title="Prediction score", x = "UMAP1", y = "UMAP2", colour="Prediction score") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical"
  )

umap_predicted_score

ggsave(umap_predicted_score, filename = file.path(ArchR_dir_healthy_35_no_doublets, "Plots/UMAP_predicted_score.png"), 
                              units="in",
                  width = 4,
                  height = 4)
```

## Celltype annotation via subclustering

### Subset to only healthy patients
```{r}
subset_healthy_pass<-which(CLLProj_Healthy_no_db@cellColData$patient_id %in% c("laureau1", "mimitou1"))
subset_healthy_cells_pass<-CLLProj_Healthy_no_db$cellNames[subset_healthy_pass]

HealthyProj <- subsetArchRProject(
  ArchRProj = CLLProj_Healthy_no_db,
  cells = subset_healthy_cells_pass,
  outputDirectory = ArchR_dir_healthy_3,
  dropCells = FALSE, 
  threads=1)
```

Subclustering and annotation via script Annotation_Healthy.Rmd

```{r}
HealthyProj <- loadArchRProject(path = file.path(ArchR_dir_healthy_3))
```

```{r}
Healthy_annotation <- tibble("label_cellbarcode"=HealthyProj$cellNames,
                             "My_annotation_new"=HealthyProj@cellColData$My_annotation_healthy)

My_annotation <- tibble("label_cellbarcode"=CLLProj_Healthy_no_db$cellNames,
                        "My_annotation_new"=CLLProj_Healthy_no_db@cellColData$My_annotation) %>%
  tidyr::separate(label_cellbarcode, into=c("label", "cellbarcode"), sep="#", remove = FALSE) %>%
  dplyr::filter(!((label %in% c("pbmc1", "mimitou1", "laureau1")))) %>%
  dplyr::select(label_cellbarcode, My_annotation_new)

My_annotation_healthy <- bind_rows(Healthy_annotation, My_annotation) %>%
  dplyr::rename("My_annotation_healthy"=My_annotation_new)

CLLProj_Healthy_no_db <- addCellColData(
  ArchRProj = CLLProj_Healthy_no_db,
  data = My_annotation_healthy %>% pull(My_annotation_healthy),
  cells = My_annotation_healthy %>% pull(label_cellbarcode),
  name = "My_annotation_healthy", 
  force=T)
```

```{r}
umap_healthy_celltype<- plotEmbedding(
  ArchRProj = CLLProj_Healthy_no_db,
  colorBy = "cellColData",
  randomize = TRUE,
  name = "My_annotation_healthy",
  embedding = "UMAP_IterativeLSI",
  pal = annotation_colours_named,
  plotAs = "points") +
  labs(title="Cell type", x = "UMAP1", y = "UMAP2", colour="Cell type") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical"
  )
```

```{r}
write_tsv(umap_healthy_celltype$data %>% 
  dplyr::rename("UMAP1"=x, "UMAP2"=y, "My_annotation_healthy"=color) %>% 
  dplyr::mutate(My_annotation_healthy=str_remove(My_annotation_healthy, pattern = "^[^-]+-"),
                My_annotation_healthy=factor(My_annotation_healthy)), file = file.path(ArchR_dir_healthy_35_no_doublets, "tsv_data/20250325_ATAC_celltype_annotation_healthy.tsv"), 
  col_names = TRUE )
```

```{r}
# # create shuffled annotation umap
# # Create a minimal expression matrix (dummy data)
# expression_matrix <- matrix(data = rnorm(208441), nrow = 1, ncol = 208441)
# rownames(expression_matrix) <- paste0("Gene1")
# colnames(expression_matrix) <- rownames(celltype_annotation_healthy)
# 
# # Create Seurat object
# seurat_obj <- CreateSeuratObject(counts = expression_matrix)
# 
# # Set the UMAP coordinates as embeddings
# umap_matrix <- celltype_annotation_healthy %>%
#   dplyr::select(`UMAP2`,`UMAP1`) %>%
#   as.matrix()
# 
# rownames(umap_matrix) <- rownames(celltype_annotation_healthy)
# seurat_obj[['umap']] <- CreateDimReducObject(embeddings = umap_matrix,
#                                              key = "UMAP_",
#                                              assay = DefaultAssay(seurat_obj))
# 
# My_annotation_healthy <- celltype_annotation_healthy %>%
#   dplyr::mutate(My_annotation_healthy=factor(My_annotation_healthy, levels=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "T", "NK", "Mono", "DC", "other"))) %>% 
#   rownames_to_column("label_cellbarcode") %>% 
#                             dplyr::select(label_cellbarcode, My_annotation_healthy) %>%
#                             deframe()
# 
# # Merge annotation_tibble with the Seurat object metadata
# seurat_obj <- AddMetaData(seurat_obj,
#                           metadata = My_annotation_healthy,
#                           col.name = "My_annotation_healthy")
# 
# metadata <- metadata %>%
#   dplyr::mutate(MBL_CLL_subtype=case_when(disease =="Healthy" ~ "Healthy",
#                                                      disease =="CLL" ~ "CLL",
#                                                      disease =="MBL" & MBL_subtype =="LC" ~ "LC",
#                                                      disease =="MBL" & MBL_subtype =="HC" ~ "HC",
#                                                      .default=NA))
# 
# MBL_CLL_subtype_annotation <- tibble("label_cellbarcode" = WhichCells(seurat_obj)) %>%
#   tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
#   dplyr::left_join(metadata %>% dplyr::select(label, MBL_CLL_subtype) %>% unique()) %>%
#   dplyr::select(label_cellbarcode, MBL_CLL_subtype) %>%
#   deframe()
# 
# seurat_obj <- AddMetaData(object=seurat_obj,
#                           metadata = MBL_CLL_subtype_annotation,
#                           col.name = "MBL_CLL_subtype")
# 
# 
# #plot
# celltype_healthy_shuffled_annotation_umap <- DimPlot(seurat_obj,
#                                                      reduction = "umap",
#                                                      group.by = "My_annotation_healthy",
#                                                      label = F,
#                                                      shuffle = TRUE,
#                                                      raster = F,
#                                                      seed = 1)+
#   scale_fill_manual(values = my_MBL_CLL_colours_named)+
#   coord_equal()+
#   scale_colour_manual(
#     name = "Cell types",
#     values = annotation_colours_named,
#     labels = c("B" = "B cells", 
#                "DC" = "Dendritic cells",
#                "Mono" = "Monocytes", 
#                "other" = "Other cell types",
#                "CLL_B_Cells" = "CLL cells", 
#                "HC_MBL_B_Cells" = "HC-MBL cells",
#                "LC_MBL_B_Cells"="LC-MBL cells",
#                "NK" = "NK cells",
#                "T" = "T cells")) +
# theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.x = element_blank(),
#     axis.text.y = element_blank(),
#     legend.position = "bottom",
#     legend.direction = "horizontal",
#     aspect.ratio = 1
#   ) +
#   labs(x = "UMAP1", y = "UMAP2", title = "Cell types") +
#   guides(colour = guide_legend(
#     keywidth = 0.5,
#     keyheight = 0.5,
#     override.aes = list(size = 3)
#   ))
# 
# celltype_healthy_shuffled_annotation_umap
# 
# 
# write_tsv(celltype_healthy_shuffled_annotation_umap[[1]]$data, file.path(ArchR_dir_healthy_35_no_doublets, "tsv_data/celltype_healthy_shuffled_annotation_umap_data.tsv"))
# 
# saveRDS(celltype_healthy_shuffled_annotation_umap, file.path(results_dir, "RDS_objects/ATAC_healthy_cell_annotation_umap.RDS"))
```


```{r}
# # no shuffeling required as separated anyway
celltype_healthy_shuffled_annotation_umap_split <- as_tibble(seurat_obj@reductions[["umap"]]@cell.embeddings, rownames = NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  left_join(tibble(
    "label_cellbarcode" = colnames(seurat_obj),
    "My_annotation_healthy" = seurat_obj@meta.data$My_annotation_healthy,
    "label" = seurat_obj@meta.data$orig.ident,
    "MBL_CLL_subtype" = seurat_obj@meta.data$MBL_CLL_subtype
  )) %>%
    dplyr::mutate(My_annotation_healthy=if_else(My_annotation_healthy=="MBL B Cell" & MBL_CLL_subtype=="LC", "LC_MBL_B_Cells", My_annotation_healthy),
                  MBL_CLL_subtype=factor(seurat_obj@meta.data$MBL_CLL_subtype, levels=c("Healthy", "LC", "HC", "CLL"))) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, colour = My_annotation_healthy)) +
  geom_point(size = 0.0001) +
  scale_colour_manual(
    name = "Cell types",
    values = annotation_colours_named,
    labels = c("B" = "B cells", "DC" = "Dendritic cells",
               "Mono" = "Monocytes", "other" = "Other cell types",
               "CLL_B_Cells" = "CLL cells", "HC_MBL_B_Cells" = "HC-MBL cells",
               "LC_MBL_B_Cells"="LC-MBL cells",
               "NK" = "Natural killer cells",
               "T" = "T cells")) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    legend.position = "bottom",
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(
    keywidth = 0.3,
    keyheight = 0.3,
    override.aes = list(size = 2)
  ))+
  facet_wrap(~MBL_CLL_subtype, nrow = 1, ncol=4)

celltype_healthy_shuffled_annotation_umap_split

write_tsv(celltype_healthy_shuffled_annotation_umap_split$data, file.path(ArchR_dir_healthy_35_no_doublets, "tsv_data/celltype_healthy_shuffled_annotation_umap_split_data.tsv"))

saveRDS(celltype_healthy_shuffled_annotation_umap_split, file.path(results_dir, "RDS_objects/ATAC_healthy_cell_annotation_umap_split.RDS"))
```


## Visualise marker genes
```{r}
# markersGS <- getMarkerFeatures(
#     ArchRProj = CLLProj_Healthy_no_db, 
#     useMatrix = "GeneScoreMatrix", 
#     groupBy = "Clusters_0.2_IterativeLSI",
#     bias = c("TSSEnrichment", "log10(nFrags)"),
#     testMethod = "wilcoxon"
# )
# 
# markerList <- getMarkers(markersGS, cutOff = "FDR <= 0.05 & Log2FC >= 1.5")
# 
# heatmapGS <- markerHeatmap(
#   seMarker = markersGS, 
#   cutOff = "FDR <= 0.05 & Log2FC >= 1.5", 
#   labelMarkers = markerGenes,
#   transpose = TRUE
# )
```

## Save ArchR project
```{r}
setwd(analysis_dir)
 
saveArchRProject(
  ArchRProj = CLLProj_Healthy_no_db,
  outputDirectory = ArchR_dir_healthy_35_no_doublets,
  load = FALSE,
  overwrite = TRUE)
```

## Quantify distance between B and (pre-)malignant cells
### Bulk
```{r}
# Retrieve LSI matrix
IterativeLSI <- getReducedDims(ArchRProj = CLLProj_Healthy_no_db, reducedDims = "IterativeLSI")

# Get all cellbarcodes of interest
Blike_bcs <- annotation %>% 
  dplyr::filter(My_annotation_LC %in% c("LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "B"),
                !str_starts(label, "lareau|mimitou")) %>% dplyr::pull(label_cellbarcode)
CLLProj_Healthy_no_db <- addCellColData(
  ArchRProj = CLLProj_Healthy_no_db,
  data = annotation %>% pull(patient_id),
  cells = annotation %>% pull(label_cellbarcode),
  name = "patient_id", force=TRUE
)
B_bcs <- annotation %>% 
  dplyr::filter(My_annotation_LC %in% c("B"),
                !str_starts(label, "lareau|mimitou")) %>% 
  pull(label_cellbarcode)

LC_MBL_bcs <- annotation %>% 
  dplyr::filter(My_annotation_LC %in% c("LC_MBL_B_Cells"),
                !str_starts(label, "lareau|mimitou")) %>% 
  pull(label_cellbarcode)

HC_MBL_bcs <- annotation %>% 
  dplyr::filter(My_annotation_LC %in% c("HC_MBL_B_Cells"),
                !str_starts(label, "lareau|mimitou")) %>% 
  pull(label_cellbarcode)

CLL_bcs <- annotation %>% 
  dplyr::filter(My_annotation_LC %in% c("CLL_B_Cells"),
                !str_starts(label, "lareau|mimitou")) %>% 
  pull(label_cellbarcode)
```


```{r}
# Subset LSI to only cells of interest and compute distances
compute_min_distance <- function(reduced_dims, bcs_celltype_from, bcs_celltype_to){
  min_dist<- proxy::dist(reduced_dims[bcs_celltype_from,],reduced_dims[bcs_celltype_to,]) %>% 
    apply(1, min)
  return(min_dist)
}

min_distances <- list("LC-MBL"=LC_MBL_bcs,
  "HC-MBL"=HC_MBL_bcs,
  "CLL"=CLL_bcs) %>% 
  map(~compute_min_distance(IterativeLSI, .x, B_bcs))
```

```{r}
min_distances <- map_dfr(min_distances, ~enframe(.x, name = "label_cellbarcode", value="dist_to_B"), .id="My_annotation_LC")
write_tsv(min_distances, file.path(DCA_analysis_tsv_dir, "min_distances.tsv"))
min_distances <- read_tsv(file.path(DCA_analysis_tsv_dir, "min_distances.tsv"))

comp <- list( c("LC-MBL", "HC-MBL"), 
              #c("HC_MBL", "CLL"), 
              c("LC-MBL", "CLL") )

distance_quantification <- min_distances %>% 
  dplyr::mutate(My_annotation_LC=factor(My_annotation_LC, levels=c("LC-MBL", "HC-MBL", "CLL"))) %>% 
  ggplot(aes(x=My_annotation_LC, y=dist_to_B, fill=My_annotation_LC))+
  geom_violin()+
  geom_boxplot(outlier.shape = NA, width=0.2, fill="white")+
  stat_compare_means(comparisons = comp, method = "wilcox", size=3)+
  scale_x_discrete(name="Cell type", expand=c(0,0))+
  scale_y_continuous(name="Distance to\nclosest B cell", expand=expansion(mult=c(0,0.05)), limits=c(0,7),oob = squish_infinite)+
  scale_fill_manual(name="Cell type", values = my_MBL_CLL_colours_named)+
  theme_pub()
distance_quantification

ggsave(plot=distance_quantification, filename=file.path(DCA_analysis_plotting_dir, "distance_quantification.pdf"), width=7, height=4, dpi=600)
write_tsv(distance_quantification$data, file.path(DCA_analysis_tsv_dir, "distance_quantification.tsv"))
```

### Per patient

```{r fig.width=3.14, fig.height=1.944}
#sorted by size of largest clone based on CD5
patient_order <- metadata %>% 
  dplyr::filter(data_type=="ATAC") %>% 
  dplyr::select(patient_id, `CD5/CD19%`) %>% 
  dplyr::filter(patient_id %in% c("MBL11", "MBL10","MBL12", "MBL15", "MBL13", "MBL14")) %>% 
  arrange(`CD5/CD19%`) %>% 
  pull(patient_id)

patient_id2_cd19cd5 <- metadata %>% 
  dplyr::filter(data_type=="ATAC") %>% 
  dplyr::select(patient_id, `CD5/CD19%`) %>% 
  dplyr::filter(patient_id %in% c("MBL11", "MBL10","MBL12", "MBL15", "MBL13", "MBL14")) %>% 
  arrange(`CD5/CD19%`) %>% dplyr::select(patient_id, `CD5/CD19%`) %>% deframe()

distance_LC_quantification <- min_distances %>% 
  dplyr::filter(My_annotation_LC =="LC-MBL") %>% 
  tidyr::separate(col = label_cellbarcode, into = c("label", "cellbarcode"), sep = "#") %>% 
  dplyr::mutate(label=factor(label, levels=patient_order)) %>% 
  ggplot(aes(x=label, y=dist_to_B, fill=My_annotation_LC, group = label))+
  geom_violin()+
  scale_x_discrete(name="Cell type", expand=c(0,0),
                   labels=label2plotting_id[c("MBL11", "MBL10","MBL12", "MBL15", "MBL13", "MBL14")])+
  scale_y_continuous(name="Distance to\nclosest B cell", expand=expansion(mult=c(0,0.05)), limits=c(0,5),oob = squish_infinite, breaks=seq(0,5,1))+
  scale_fill_manual(values = my_MBL_CLL_colours_named)+
  theme_pub()+
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x=element_blank(),
        legend.position = "none")
distance_LC_quantification
ggsave(plot = distance_LC_quantification, filename = file.path(DCA_analysis_plotting_dir, "20260204_distance_LC_quantification_figure.png"), dpi=200, width=3.14, height= 1.944, bg="white")
ggsave(plot = distance_LC_quantification, filename = file.path(DCA_analysis_plotting_dir, "20260204_distance_LC_quantification_figure.pdf"), width=3.14, height= 1.944, bg="white")
write_tsv(distance_LC_quantification$data, file.path(DCA_analysis_tsv_dir, "distance_LC_quantification.tsv"))
```

