---
title: "Compare_BCR_mtDNA_clones_MBL_CLL_longread"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
# Goal
The aim of this script is to extract mtDNA mutations which were previously identified in scATAC from scRNA data and compare them to BCR clones of the same samples to validate them.

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(Signac)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(SummarizedExperiment)
library(ggrepel)
library(Seurat)
library(circlize)
library(ComplexHeatmap)
library(Matrix)
library(ggalluvial)
library(scales)
library(reshape2)
library(vctrs)
library(cowplot)
library(numbat)
library(ggrastr)
```

# Set paths
```{r define_directories }
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "nanoranger")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "longread_mtDNA_BCR_analysis")
plotting_dir <- file.path(analysis_dir, "Plots")
RDS_dir <- file.path(analysis_dir, "RDS_objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")

ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")
GEX_analysis_tsv_dir <- file.path(result_dir, "GEX_analysis/tsv_data")
VDJ_analysis_RDS_dir <- file.path(result_dir, "VDJ_analysis/RDS_objects")
mtDNA_mutation_analysis_RDS_dir <- file.path(result_dir, "mtDNA_mutation_analysis/RDS_objects")
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r}
metadata_all <- read_tsv(file = file.path(data_dir, "20250508_Tidy_MBL_CLL_metadata.tsv")) 
metadata <- metadata_all %>% dplyr::filter(data_type=="nanoranger")

nanoranger_samples <-metadata %>%
  dplyr::select(pool, label) %>%
  deframe()

mgatk_file_paths <- file.path(library_path, paste0(names(nanoranger_samples), ".mgatk/final"))

new_BCR_paired <- readRDS(file.path(result_dir, "VDJ_analysis/RDS_Objects/new_BCR_paired.RDS"))

mt_gene_annotation <- readr::read_csv(file.path(data_dir, "reference_genomes/mt_genome_gene_annotation/NCBI RefSeq Annotation GCF_000001405.40-RS_2024_08.csv"), 
                               col_names = TRUE, 
                               skip = 1) %>% 
  dplyr::mutate(`Gene symbol`= case_when(str_starts(`Gene symbol`, pattern =  "TRN") ~ "tRNA",
                                         str_starts(`Gene symbol`, pattern =  "RNR") ~ "rRNA", .default =`Gene symbol`))

gex_scaled_10X_seurat_all <- readRDS(file.path(project_dir_ext, "results/GEX_analysis/RDS_objects/gex_scaled_10X_seurat_all.RDS"))

celltype_annotation_RNA <- gex_scaled_10X_seurat_all@meta.data %>%
  dplyr::select(orig.ident, My_annotation_LC) %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  as_tibble() %>%
  dplyr::rename(label=orig.ident) %>%
  dplyr::mutate(label_cellbarcode=str_replace(string=label_cellbarcode, pattern="_", replacement="#"))
write_tsv(celltype_annotation_RNA, file.path(result_dir, "20250513_RNA_celltype_annotation_healthy.tsv"))

celltype_annotation_RNA <- read_tsv(file.path(result_dir, "GEX_analysis/tsv_data/20250513_RNA_celltype_annotation_healthy.tsv"))

#scATAC mtDNA mutation VAFs previously called
ATAC_cln_variants_MBL <- readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset4/RDS_objects/VAFs.cln.MBL.RDS")) %>% 
  map(~ t(.x) %>% as.matrix() %>%
    .[as.logical(rowSums(. != 0)), as.logical(colSums(. != 0))] %>%
    as_tibble())

all_significant_ATAC_heteroplasmy_changes <- readRDS(file.path(mtDNA_mutation_analysis_RDS_dir, "all_significant_heteroplasmy_changes.RDS"))
all_insignificant_heteroplasmy_changes <- readRDS(file.path(mtDNA_mutation_analysis_RDS_dir, "all_insignificant_heteroplasmy_changes.RDS"))

label2pool <- metadata %>%
  dplyr::select(label, pool) %>%
  unique() %>% 
  deframe()

nb_lst <- map(metadata$label[c(1, 4, 6, 7)], ~ Numbat$new(out_dir = file.path(project_dir_ext, "numbat",label2pool[.x]))) %>% 
  purrr::set_names(metadata$label[c(1, 4, 6, 7)])

write_tsv(gex_scaled_10X_seurat_all_noIG@meta.data %>%
  dplyr::select(orig.ident, My_annotation_LC, MBL_CLL_subtype) %>%
  dplyr::rename(label=orig.ident) %>%
  rownames_to_column("label_cellbarcode") %>%
  left_join(gex_scaled_10X_seurat_all_noIG@reductions[["umap"]]@cell.embeddings %>% as_tibble(rownames = "label_cellbarcode")), file = file.path(GEX_analysis_tsv_dir, "20250626_UMAP_10X_seurat_all_no_IG_celltypes_MBL_CLL_subtype_split_noB_data.tsv"))

celltype_annotation_RNA_umap <- read_tsv(file.path(GEX_analysis_tsv_dir, "20250626_UMAP_10X_seurat_all_no_IG_celltypes_MBL_CLL_subtype_split_noB_data.tsv"))
```

```{r}
label2plotting_id <- metadata %>% 
  dplyr::select(label, plotting_id) %>% 
  deframe()

patient_id2plotting_id <- metadata %>%
  dplyr::select(patient_id, plotting_id) %>%
  unique() %>% 
  deframe()
```

# Analysis
Statistics of previously called mtDNA mutations from scATAC data
```{r}
ATAC_cln_variants_MBL %>% 
  imap(~rownames(.) %>%
         as_tibble() %>% 
         dplyr::mutate(patient_id=.y) %>% 
         group_by(patient_id) %>% 
         unique() %>% 
         dplyr::summarize(counts=n())) %>% 
  bind_rows() %>%
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id), by = join_by(patient_id)) %>% 
  dplyr::mutate(plotting_id=fct_reorder(plotting_id, -counts)) %>% 
ggplot(aes(x=plotting_id, y=counts, fill=plotting_id))+
  geom_bar(position="identity", stat="identity")+
  scale_y_continuous(expand=expansion(mult=c(0,NA)))+
  scale_fill_manual(values=rep("midnightblue", 7))+
  labs(x="Patient", title="All clonal scATAC mtDNA mutations across celltypes")+
  theme_pub()+
  theme(legend.position = "none")
```

## Calling of mtDNA mutation nanoranger data
```{r}
# Individual mgatk files per sample
mito.data.idv <- mgatk_file_paths %>%
  map(~ ReadMGATK(dir=.x, verbose=TRUE)) %>%
  purrr::set_names(nanoranger_samples)

saveRDS(mito.data.idv, file = file.path(RDS_dir, "mito.data.idv.RDS"))

mito.data.MBL <- list()
for (i in nanoranger_samples) {
  subset_names <- str_subset(names(mito.data.idv), i)
  subset_names_index <- match(subset_names, names(mito.data.idv))

  df_1 <- mito.data.idv[[subset_names_index[[1]]]]$counts
  colnames(df_1) <- paste(subset_names[1], colnames(df_1), sep = "#")
  mito.data.MBL <- append(mito.data.MBL, df_1)
}
names(mito.data.MBL) <- nanoranger_samples

saveRDS(mito.data.MBL, file = file.path(RDS_dir, "mito.data.MBL.RDS"))
mito.data.MBL <- readRDS(file = file.path(RDS_dir, "mito.data.MBL.RDS"))
```

## Variants identification
```{r Get_mt_variants}
variant.df.MBL <- mito.data.MBL %>%
  map(~ IdentifyVariants(.x, refallele = mito.data.idv[[1]]$refallele)) %>%
  imap(~dplyr::mutate(.x, sample_id = .y, .before = "nucleotide"))

variant.df.MBL %>%
  imap(~ write_tsv(.x,
    quote = "none",
    file = file.path(analysis_dir, paste0("MitochondrialVariants/Variants_", .y, ".csv"))
  ))

saveRDS(variant.df.MBL, file = file.path(RDS_dir, "variant.df.MBL.RDS"))
variant.df.MBL <- readRDS(file = file.path(RDS_dir, "variant.df.MBL.RDS"))
```

```{r Plot_mt_variants}
all.variant.df.plots <- variant.df.MBL %>%
  map(VariantPlot) %>%
  iwalk(~ ggsave(
    filename = file.path(analysis_dir, paste0("MitochondrialVariants/Plot_Variance_Mean_Ratio_", .y, ".pdf")),
    plot = .x,
    width = 5,
    height = 3
  ))
```

## Heteroplasmy calculation
```{r VAFs_clonal_mt_variants}
VAFs.cln.MBL <- mito.data.MBL%>%
  map2(intersect.cln.variants.MBL, ~ AlleleFreq(.,
    variants = .y$variant,
    assay = NULL
  ))

saveRDS(VAFs.cln.MBL, file = file.path(RDS_dir, "VAFs.cln.MBL.RDS"))
VAFs.cln.MBL <- readRDS(file = file.path(RDS_dir, "VAFs.cln.MBL.RDS"))
```

## Filtering clonal mtDNA mutations from scATAC
```{r Filter_clonal_mt_variants}
Get clonal variants
intersect.cln.variants.MBL <- VAFs.cln.MBL %>%
  imap(~ t(.x) %>% as.matrix() %>%
    .[as.logical(rowSums(. != 0)), as.logical(colSums(. != 0))] %>%
    as_tibble(rownames = "label_cellbarcode") %>%
    pivot_longer(cols=-label_cellbarcode, names_to = "variant_id", values_to = "VAF") %>%
      dplyr::filter(variant_id %in% colnames(ATAC_cln_variants_MBL[[.y]])) %>%
      tidyr::separate(col = variant_id, into = c("position", NA), sep = "(?<=\\d)(?=[A-Z])", remove=FALSE) %>%
    dplyr::mutate(position=as.numeric(position)) %>%
      dplyr::mutate(label_cellbarcode=paste0(label_cellbarcode, "-1")))

# Export_cln_mito_variants
intersect.cln.variants.MBL %>%
  imap(~ write_tsv(.x, file = paste0(analysis_dir, "/MitochondrialVariants/Clonal_Variants_", .y, ".tsv")))

saveRDS(intersect.cln.variants.MBL, file = file.path(RDS_dir, "intersect.cln.variants.MBL.RDS"))
intersect.cln.variants.MBL <- readRDS(file = file.path(RDS_dir, "intersect.cln.variants.MBL.RDS"))
```

```{r}
# all mtDNA mutations from nanoranger
ggarrange(
bind_rows(VAFs.cln.MBL %>%
  map(~ t(.x) %>% as.matrix() %>%
    .[as.logical(rowSums(. != 0)), as.logical(colSums(. != 0))] %>%
    as_tibble(rownames = "label_cellbarcode") %>% 
    pivot_longer(cols=-label_cellbarcode, names_to = "variant_id", values_to = "VAF")),
  .id = "patient_id") %>% 
  left_join(metadata_all %>% dplyr::filter(data_type=="nanoranger") %>%
              dplyr::select(plotting_id, patient_id) %>% unique, by = join_by(patient_id)) %>% 
  dplyr::select(plotting_id, variant_id) %>% 
  unique() %>% 
  group_by(plotting_id) %>% 
  dplyr::summarise(variant_count=n()) %>% 
  dplyr::mutate(plotting_id=fct_reorder(plotting_id, -variant_count)) %>% 
  ggplot2::ggplot(aes(x=plotting_id, y=variant_count, label=variant_count))+
  geom_bar(stat="identity",fill="midnightblue")+
  geom_text(vjust=-1)+
  labs(title= "All nanoranger detected mtDNA mutations across celltype", y="mtDNA mutation counts", x="")+
  scale_y_continuous(expand = expansion(mult = c(0, NA)), limits=c(0,60))+
  theme_pub(),

# intersect of nanoranger mtDNA mutations with scATAC mtDNA mutations
intersect.cln.variants.MBL %>% 
  imap(~dplyr::select(., variant_id) %>%
         dplyr::mutate(patient_id=.y) %>% 
         group_by(patient_id) %>% 
         unique() %>% 
         dplyr::summarize(counts=n())) %>% 
  bind_rows() %>%
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id), by = join_by(patient_id)) %>% 
  dplyr::mutate(plotting_id=fct_reorder(plotting_id, -counts)) %>% 
  ggplot(aes(x=plotting_id, y=counts, fill=plotting_id, label=counts))+
  geom_bar(position="identity", stat="identity", fill="midnightblue")+
  geom_text(vjust=-1)+
  scale_y_continuous(expand = expansion(mult = c(0, NA)), limits=c(0,60))+
  labs(title="All scATAC matched nanoranger mtDNA mutations", x="Patients", y="mtDNA mutation counts")+
  theme(legend.position = "none")+
  theme_pub(), nrow=1, ncol=2)
```

## FindClonotypes across all cell types
```{r, warning=FALSE, message=FALSE}
# Cluster together per patient and filter for non-zero VAF cells
mito_VAFs_MBL <- VAFs.cln.MBL %>%
  imap(~ as.matrix(.x) %>%
    as_tibble(rownames = "variant_id") %>%
    column_to_rownames(var = "variant_id") %>%
    .[unique(intersect.cln.variants.MBL[[.y]]$variant_id), ] %>% # filter for only high confidence variants from scATACseq
    .[, as.logical(colSums(. != 0))] %>% # filter out cells without heteroplasmy
    .[,colSums(is.na(.))<nrow(.)] %>% # filter without intformation
    dplyr::mutate(across(where(is.numeric), ~ ifelse(. < 10e-6 & .>0, 10e-6, .))) %>% # cap heteroplasmy at 10^-6
    as.matrix() %>%
    as("sparseMatrix") %>%
    Seurat::CreateSeuratObject(assay = "alleles"))

clonotypes_MBL_new <- mito_VAFs_MBL %>%
  map(~ FindClonotypes(.))

find_clonotype_clusters_MBL <- clonotypes_MBL_new %>%
  map(~ .@active.ident %>%
    data.frame() %>%
    rownames_to_column("label_cellbarcode") %>%
    dplyr::rename(Cluster = ".")) %>%
  imap( ~ write_tsv(.x, file = file.path(analysis_dir, paste0("MitochondrialVariants/FindClonotypesClusters_new_", .y, ".tsv"))))

saveRDS(clonotypes_MBL_new, file.path(RDS_dir, "clonotypes_MBL_new.RDS"))
clonotypes_MBL_new <- readRDS(file.path(RDS_dir, "clonotypes_MBL_new.RDS"))
```

```{r fig.width=6}
bind_rows(
  bind_rows(new_BCR_paired) %>% 
  dplyr::select(patient_id, new_common_ct_id) %>% 
  left_join(metadata %>%
              dplyr::select(plotting_id, patient_id), by = join_by(patient_id)) %>%
  dplyr::filter(!is.na(plotting_id)) %>% 
  group_by(plotting_id) %>% 
  dplyr::summarise(clone_counts=n()) %>%
  dplyr::mutate("method"="BCR"),
  
find_clonotype_clusters_MBL %>% 
  bind_rows(.id = "patient_id") %>% 
  dplyr::select(patient_id, Cluster) %>%  
  left_join(metadata %>%
              dplyr::select(plotting_id, patient_id), by = join_by(patient_id)) %>%
  group_by(plotting_id) %>% 
  dplyr::summarise(clone_counts=n())%>%
  dplyr::mutate("method"="mtDNA mutation")) %>% 
  dplyr::mutate(plotting_id=fct_reorder(plotting_id, -clone_counts)) %>% 
  ggplot2::ggplot(aes(x=plotting_id, y=clone_counts, fill=method))+
  geom_bar(stat="identity", position=position_dodge2(width = 0.9, preserve = "single"))+
  labs(y="clone counts", x="")+
  scale_y_log10(expand = expansion(mult = c(0, NA)))+
  scale_fill_manual(name="Tracing method", values = c("BCR"="lightgoldenrod3", 
                                                      "mtDNA mutation"="indianred"))+
  theme_pub()+
  theme(axis.text.x=element_text(angle=90))
```

## mtDNA mutation recurrance across patients (all celltypes)
```{r}
mutation_count_data <- intersect.cln.variants.MBL %>% 
  map(~left_join(., celltype_annotation_RNA %>%
                               dplyr::select(label_cellbarcode, My_annotation_LC), by = join_by(label_cellbarcode)) %>%
filter(!(is.na(My_annotation_LC)),
         My_annotation_LC %in% c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")))%>%
bind_rows(.id = "patient_id") %>% 
dplyr::select(patient_id, variant_id, position) %>% 
unique() %>% 
group_by(variant_id) %>% 
dplyr::mutate(patient_counts=n_distinct(patient_id)) %>% 
arrange(patient_counts) %>% 
ungroup()
```

```{r}
axis_data <- tibble("mt_genome"=c(0, 16569))

mt_count_radial_plot <- ggplot() +
  geom_line(data= axis_data, aes(x=mt_genome, y= 2.1), colour="lightgrey", linewidth = 0.5)+
  geom_line(data= axis_data, aes(x=mt_genome, y= 2.4), colour="lightgrey", linewidth = 0.5)+
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="plus"), 
            aes(xmin = Start, xmax = Stop, ymin=2, ymax=2.2, fill = `Gene symbol`), colour = "black", linewidth = 0.3) + 
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="minus"), aes(xmin = Start, xmax = Stop, ymin=2.3, ymax=2.5, fill = `Gene symbol`), colour="black", linewidth = 0.3)+
  geom_jitter(data = mutation_count_data %>% bind_rows() %>% dplyr::select(variant_id, position, patient_counts) %>% unique, aes(x = position, y = 3, colour = patient_counts), shape=16, size = 0.5) +
  geom_rect(data=tibble("position"=seq(0, 16659, 200)), aes(xmin=position,xmax=position+10, ymin=3.5, ymax=3.6), linewidth = 0.2, colour="black", fill="black")+
  geom_text(data=axis_data, aes(x=0, y=0, label="mt\ngenome"))+
  scale_x_continuous(limits=c(0, 16659), breaks=seq(0, 16569, 2000), labels = seq(0, 16569, 2000))+
  scale_y_continuous(breaks=c(3.5), limits = c(0, 3.7))+
  scale_color_distiller(name="Patients with\nmtDNA mutation", palette = "YlOrRd", direction = 1, breaks=c(1,3,5, 7, 9)) +
  scale_fill_manual(name="Gene", 
                    values=c("ATP6"="#15512A", "ATP8"="#2E8565","COX1"="#4EB1A9","COX2"="#8CB5C0","COX3"="#1A205F","CYTB"="#483591",
                             "ND1"="#885EB4","ND2"="#BB9AC4","ND3"="#6C2047","ND4"="#9C3D52","ND4L"="#B7716E","ND5"="#CAB3A8","ND6"="#757927",
                             "rRNA"="#95BB7D", "tRNA"="lightgrey"))+
  coord_radial(rotate.angle = TRUE, expand = FALSE) +
  theme_pub()+
  theme(panel.background = element_rect(fill = 'white'),
        panel.grid.major.y = element_line(colour = "black"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y =  element_blank())+
  guides(colour = guide_legend(title="Patient-shared mutations",
                              override.aes=list(size=3), ncol = 5, byrow=TRUE),
         fill = guide_legend(ncol=3, byrow=TRUE))
mt_count_radial_plot
```

```{r}
mutation_count_data %>% 
  ggplot(aes(x=patient_counts, fill=as.factor(patient_counts)))+
  geom_bar(stat="count", position = "stack")+
  scale_fill_brewer(palette = "YlOrRd")+
  scale_x_continuous(breaks = seq(0,7,1))+
  scale_y_continuous(expand = expansion(mult=c(0, 0.01)))+
  labs(x="Number of patients sharing mtDNA mutations", y="mtDNA mutation counts", fill="Number of patients sharing mutations")+
  theme_pub()
```

## Coverage whole mtDNA genome
```{r fig.height=4, fig.width=4}
#Calculates means per position across all annotated cells --> then calculates rolling mean across positions
pull_coverage<- function(SE, prefix, resolution = 5){
  zoo::rollmean(rowMeans(assays(SE)[['coverage']]), resolution)
}

coverage_whole_mt_genome <- file.path(library_path, paste0(names(nanoranger_samples), ".mgatk/final"), paste0(names(nanoranger_samples), ".rds")) %>% 
  map2(nanoranger_samples, ~data.frame(
  patient_id=.y, 
  position = zoo::rollmean(1:16569, 5),
  coverage = pull_coverage(SE=readRDS(.),
                           prefix=.y))) %>% 
  set_names(nanoranger_samples)

mt_radial_coverage_plot <- imap(coverage_whole_mt_genome,
  ~ggplot() +
  geom_line(data= axis_data, aes(x=mt_genome, y= 22), colour="lightgrey", linewidth = 0.5)+
  geom_line(data= axis_data, aes(x=mt_genome, y= 25), colour="lightgrey", linewidth = 0.5)+
  
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="plus"), 
            aes(xmin = Start, xmax = Stop, ymin=24, ymax=26, fill = `Gene symbol`), 
            colour = "black", linewidth = 0.3) + 
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="minus"), 
            aes(xmin = Start, xmax = Stop, ymin=21, ymax=23, fill = `Gene symbol`), 
            colour="black", linewidth = 0.3)+
    geom_line(data=.x,aes(x = position, y = coverage+30), 
              color = "steelblue4", linewidth = 0.5)+
  scale_x_continuous(limits=c(0, 16659), breaks=seq(0, 16569, 2000), labels = seq(0, 16569, 2000))+
  scale_y_continuous(breaks=c(3.5), limits = c(0, 110))+
  scale_fill_manual(name="Gene", 
                    values=c("ATP6"="#15512A", "ATP8"="#2E8565","COX1"="#4EB1A9","COX2"="#8CB5C0","COX3"="#1A205F","CYTB"="#483591",
                             "ND1"="#885EB4","ND2"="#BB9AC4","ND3"="#6C2047","ND4"="#9C3D52","ND4L"="#B7716E","ND5"="#CAB3A8","ND6"="#757927",
                             "rRNA"="#95BB7D", "tRNA"="lightgrey"))+
  coord_radial(rotate.angle = TRUE, expand = FALSE) +
  theme_pub()+
  theme(panel.background = element_rect(fill = 'white'),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y =  element_blank()))

mt_radial_coverage_plot

saveRDS(mt_radial_coverage_plot, file.path(RDS_dir, "mt_radial_coverage_plots.RDS"))
```

## Coverage mtDNA variants
```{r}
#Calculates means per position across all annotated cells --> then calculates rolling mean across positions
pull_coverage_subset<- function(SE, cells, prefix, resolution = 5){
  SE_subset <- SE[, colnames(SE) %in% (str_remove(cells, prefix) %>% str_remove("#") %>% str_remove("-1"))]
  zoo::rollmean(rowMeans(assays(SE_subset)[['coverage']]), resolution)

}
mtmut_positions_patient <- bind_rows(
  VAFs.cln.MBL %>%
  imap(~ as.matrix(.) %>%
    as_tibble(rownames = "variant_id") %>%
    .[, as.logical(colSums(. != 0))] %>% 
    dplyr::filter(variant_id %in% intersect.cln.variants.MBL[[.y]]$variant_id)), 
  .id = "patient_id") %>% 
  left_join(metadata_all %>% dplyr::filter(data_type=="nanoranger") %>%
              dplyr::select(plotting_id, patient_id) %>% unique, by = join_by(patient_id)) %>% 
  dplyr::select(patient_id, variant_id) %>% 
  tidyr::separate(variant_id, into = c("position", NA), sep = "(A|C|G|T)", remove = FALSE)

coverage_all_samples <- file.path(library_path, paste0(names(nanoranger_samples), ".mgatk/final"), paste0(names(nanoranger_samples), ".rds")) %>% 
  map2(nanoranger_samples, ~data.frame(
  patient_id=.y, 
  position = zoo::rollmean(1:16569, 5),
  coverage = pull_coverage_subset(SE=readRDS(.), cells=(celltype_annotation_RNA %>% 
                                                  dplyr::filter(label == .y,
                                                                !is.na(My_annotation_LC)) %>% 
                                                  pull(label_cellbarcode) ),
                           prefix=.y)) %>% 
  dplyr::left_join(mtmut_positions_patient %>% dplyr::mutate(position=as.numeric(position)), by = join_by(patient_id, position)) %>% 
  dplyr::filter(variant_id %in% intersect.cln.variants.MBL[[.y]]$variant_id)) %>% 
  set_names(nanoranger_samples)

write_tsv(bind_rows(coverage_all_samples, .id = "label"),
          file=file.path(tsv_dir, "mtDNA_coverage_nanoranger_samples.tsv"))

coverage_all_samples <- read_tsv(file=file.path(tsv_dir, "mtDNA_coverage_nanoranger_samples.tsv"))
```

```{r}
coverage_plots_variants <- coverage_all_samples %>% 
  split(.$label) %>% 
  imap(~ggplot(., aes(x = position, y = coverage, label=variant_id)) +
  geom_line(color = "black") +
  geom_label_repel(label.size = NA, 
                   fill=NA, max.overlaps = 47,
                   nudge_y = 40, 
                   colour = "darkred")+
  scale_y_continuous(expand = expansion(mult=c(0, 0.05)), limits=c(0,100), breaks=seq(0,100,10))+
  scale_x_continuous(expand = expansion(mult=c(0, 0.01)))+
  theme_pub()+
    # coord_polar()+
  labs(x = "Position on mtDNA chromosome", 
       y = "Rolling mean coverage across cells",
       title=.y))
```

```{r fig.width=10, fig.height=28}
cowplot::plot_grid(plotlist = coverage_plots_variants, ncol = 1, nrow = 7)
```

## Total depth per cell 
```{r}
depth_per_cell <- file.path(library_path, paste0(names(nanoranger_samples), ".mgatk/final"), paste0(names(nanoranger_samples), ".signac.rds")) %>% 
  map2(nanoranger_samples, 
       ~readRDS(.x)$depth%>% 
         as_tibble() %>% 
         dplyr::rename(cellbarcode=sample) %>% 
         dplyr::mutate(patient_id=.y,
                       label_cellbarcode=paste0(patient_id, "#", cellbarcode, "-1"))%>% 
        dplyr::filter(label_cellbarcode %in% (celltype_annotation_RNA %>% pull(label_cellbarcode)))) %>% 
  bind_rows()
```

```{r}
ggplot(depth_per_cell, aes(x=patient_id, y=depth, color=patient_id))+
  geom_jitter(width = 0.2, colour = "black", aes(fill = patient_id), pch = 21) +
  geom_violin(colour = "black", fill="white", alpha=0.5) +
  scale_fill_manual(name="Patient", values = my_unpaired_colours_named)+
  scale_y_continuous(expand=expansion(mult=c(0,0.01)), limits=c(0, NA))+
  labs(title="Cellular sequencing depth", x="Patient", y="Depth")+
  theme_pub()
```

## Subset scBCR and nanoranger to same barcodes of high-quality cells

```{r}
clone_matching_data <- full_join(
  bind_rows(new_BCR_paired) %>%
    left_join(metadata %>%
              dplyr::select(plotting_id, patient_id), by = join_by(patient_id)) %>%
  dplyr::select(patient_id, plotting_id, label_cellbarcode, new_common_ct_id) %>%
  dplyr::filter(!is.na(plotting_id)) %>%
  dplyr::rename("BCR_clone"=new_common_ct_id),

find_clonotype_clusters_MBL %>%
  bind_rows(.id = "patient_id") %>%
  left_join(metadata %>%
              dplyr::select(plotting_id, patient_id), by = join_by(patient_id)) %>%
  dplyr::select(patient_id, label_cellbarcode, Cluster) %>%
  dplyr::mutate(label_cellbarcode=paste0(label_cellbarcode, "-1")) %>%
  dplyr::rename("mtDNA_clone"=Cluster), by = join_by(patient_id, label_cellbarcode)) %>%
  dplyr::filter(!is.na(BCR_clone) & !is.na(mtDNA_clone),
                label_cellbarcode %in% celltype_annotation_RNA$label_cellbarcode) %>%
  unique()

write_tsv(clone_matching_data, file.path(tsv_dir, "clone_matching_data.tsv"))
clone_matching_data <- read_tsv(file.path(tsv_dir, "clone_matching_data.tsv"))
```

```{r}
clone_matching_data_complete <- clone_matching_data %>%
  dplyr::filter(!is.na(BCR_clone) & !is.na(mtDNA_clone)) %>%
  dplyr::mutate(combination=paste0(mtDNA_clone, "_", BCR_clone)) %>%
  dplyr::group_by(patient_id, combination) %>%
  dplyr::summarise(counts = n()) %>%
  dplyr::arrange(desc(counts)) %>%
  tidyr::separate(combination, into = c("mtDNA_clone", "BCR_clone"), sep = "_", remove = FALSE) %>%
  rowid_to_column("id") %>%
  group_by(patient_id) %>%
  dplyr::mutate(freq = counts / sum(counts) * 100) %>%
  dplyr::mutate(
    BCR_clone = paste0("BCR_clone_", BCR_clone))

write_tsv(clone_matching_data_complete, file.path(tsv_dir, "clone_matching_data_complete.tsv"))
clone_matching_data_complete <- read_tsv(file.path(tsv_dir, "clone_matching_data_complete.tsv"))
```

## Clone matching of mtDNA and BCR

### Alluvial plot

```{r, fig.height=12, fig.width=12, warning=FALSE, message=FALSE}
set.seed(128)
mt_colours <- clone_matching_data_complete %>%
  dplyr::mutate(mtDNA_clone=as.character(mtDNA_clone),
                BCR_clone=as.character(BCR_clone)) %>% 
  split(.$patient_id) %>% 
  map(~sample(colorRampPalette(brewer.pal(12, "Set2"))(length(unique(.x$mtDNA_clone))), replace = FALSE) %>% 
        setNames(unique(.x$mtDNA_clone)))
  
clone_matching_plots <-clone_matching_data_complete %>%
    dplyr::mutate(mtDNA_clone=factor(mtDNA_clone, levels=c(unique(.$mtDNA_clone), unique(.$BCR_clone))), # important to keep; otherwise strata and alluviae do not match up!
                BCR_clone=factor(BCR_clone, levels=c(unique(.$mtDNA_clone), unique(.$BCR_clone)))) %>% 
  split(.$patient_id) %>% 
  imap(~ggplot(.x,
              aes(y = freq, 
      axis1 = mtDNA_clone,
      axis2 = BCR_clone,
      fill=mtDNA_clone)) +
  geom_alluvium(width = 1 / 12) +
  geom_stratum(
    width = 1 / 12,
    color = "lightgrey") +
  scale_x_discrete(
    limits = c( "mtDNA clones", "BCR clones"),
    expand = c(.04, .04)  ) +
  scale_y_continuous(expand = expansion(mult=0, 0))+
  scale_fill_manual(values = mt_colours[[.y]], na.value = "white")+
  labs(y = "Frequency [%]", 
       x = "", title=paste0(.y, "\n", sum(.x$counts), " cells")) +
  theme_pub() +
  theme(legend.position = "none",
        plot.title = element_text(face="plain",
                                  hjust=0.5))) 

ggsave(plot=ggpubr::ggarrange(plotlist = clone_matching_plots, nrow=3, ncol=3), 
       file= file.path(plotting_dir, "mtDNA_clones_BCR_clone_alluvial_plots.png"), 
       width = 10, 
       height=10 )

ggpubr::ggarrange(plotlist = clone_matching_plots, nrow=3, ncol=3)
```

### Heteroplasmy heatmaps

#### All celltypes

```{r}
order_matrix_mt_clone_celltype <- function(mat, 
                                           cell_ann, 
                                           first_ann, # mtDNA_clone
                                           second_ann # condition
                                           ){
  geneOrder <- sort(rowSums(mat), decreasing = TRUE, index.return = TRUE)$ix
  
  scoreCol <- function(x) {
    score <- 0
    for (i in 1:length(x)) {
      if (x[i]) {
        score <- score + 2^(length(x) - i)
      }
    }
    return(score)
  }
  
  scores <- apply(mat[geneOrder, , drop = FALSE], 2, scoreCol)
  sampleOrder <- sort(scores, decreasing = TRUE, index.return = TRUE)$ix

  # Order cols by first annotation
  first_ann_freq <- enframe(first_ann,
                      name="label_cellbarcode",
                      value = "mtDNA_clone") %>%
  dplyr::group_by(mtDNA_clone) %>%
  dplyr::summarise(frequency_first=n()) %>%
  dplyr::arrange(desc(frequency_first)) %>%
  dplyr::mutate(mtDNA_clone=factor(mtDNA_clone, mtDNA_clone))%>%
  pull(mtDNA_clone)
  
  barcodes_celltype_first_second_ann_ordered <- tibble("label_cellbarcode"=colnames(mat),
         "scores_memosort"=scores) %>%
    dplyr::mutate(celltype=cell_ann[label_cellbarcode],
                  mtDNA_clone=first_ann[label_cellbarcode],
                  condition=second_ann[label_cellbarcode]) %>% 
    dplyr::mutate(mtDNA_clone=factor(mtDNA_clone, levels=levels(first_ann_freq)),
                  condition=factor(condition, levels=c("LC", "HC", "CLL")),
                  celltype=factor(celltype, levels=c("B", "MBL_B_Cells", "CLL_B_Cells", "T","Mono","NK", "DC","other"))) %>% 
    dplyr::group_by(mtDNA_clone) %>% 
    dplyr::arrange(mtDNA_clone, condition, celltype, desc(scores_memosort)) %>% 
    pull(label_cellbarcode)
  
      return(list(geneOrder, barcodes_celltype_first_second_ann_ordered))
}
```

```{r}
make_mtDNAclone_celltype_heteroplasmy_heatmap <- function(meta_data,
                                                          heteroplasmy_matrix, # variants in rownames; cell barcodes in column names
                                                         patient_name, # character string
                                                         celltype_annotation, # named vector
                                                         mtDNA_clone_annotation, # named vector
                                                         mtDNA_clone_colours, # named vector
                                                         BCR_clone_annotation,
                                                         BCR_clone_colours,
                                                         condition_annotation, # named vector
                                                         condition_colours # named vector
                                                         )
{
  metadata_heatmaps <- meta_data %>%
    dplyr::filter(data_type == "nanoranger") %>%
    dplyr::select(patient_id, label, disease, MBL_subtype) %>%
    dplyr::mutate(
      MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype),
      longitudinal = if_else(str_starts(label, "CLL"), "paired", "individual")
    )

  patient_id2plotting_id_intern <- metadata_all %>%
  dplyr::select(patient_id, plotting_id) %>%
  unique() %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, 
                                   levels=c(paste0("P", 1:9), "P9dup", paste0("P", 10:24)))) %>% 
  deframe()
  
  cell_matrix <- heteroplasmy_matrix  %>%
    .[rowSums(.) != 0,,drop=FALSE] %>% 
    .[, colSums(.) != 0, drop=FALSE] %>% 
    .[,colSums(is.na(.))<nrow(.),drop=FALSE]
  
  ordering <- order_matrix_mt_clone_celltype(mat = cell_matrix,
                                             cell_ann =celltype_annotation,
                                             first_ann = mtDNA_clone_annotation,
                                             second_ann = condition_annotation)
  

  cell_matrix <- -log10(cell_matrix)
  
  #reconcatenate columns and replace 0 with NA for complex heatmap
  cell_matrix <- as.matrix(cell_matrix)
  cell_matrix[cell_matrix == Inf] <- NA
  
  # order matrix in place
  cell_matrix_ordered <- cell_matrix[ordering[[1]], ordering[[2]], drop=FALSE]
  
  # prepare top celltype annotation
  celltype2colour <- c(
  "B" = "darkred",
  "CD8 T" = "azure4",
  "CD4 T" = "darkblue",
  "DC" = "darkmagenta",
  "Mono" = "tan",
  "NK" = "gold",
  "other" = "cadetblue1",
  "T" = "darkblue",
  "MBL B Cell" = "#FC8D62",
  "CLL B Cell" = "#66C2A5",
  "HC-MBL B Cell" ="#FC8D62" ,
  "LC-MBL B Cell" ="#fcbda4" ,  
  "HC_MBL_B_Cells" ="#FC8D62" ,
  "LC_MBL_B_Cells" ="#fcbda4" ,
  "MBL_B_Cells" ="#FC8D62",
  "CLL_B_Cells" ="#66C2A5")
  
  barcode2celltype <- celltype_annotation
  
  # prepare top mtDNA annotations
  mtDNAclone2colour <- mtDNA_clone_colours
  barcode2mtDNAclone <- mtDNA_clone_annotation %>% 
    as.character() %>% 
    setNames(names(mtDNA_clone_annotation))
  
    # prepare top mtDNA annotations
  BCRclone2colour <- BCR_clone_colours
  barcode2BCRclone <- BCR_clone_annotation %>% 
    as.character() %>% 
    setNames(names(BCR_clone_annotation))
  
  # prepare top condition annotation
  condition2colour <- condition_colours
  barcode2condition <- condition_annotation
  
  column_ha <- HeatmapAnnotation(
  "mtDNA clone" = anno_simple(x= barcode2mtDNAclone[colnames(cell_matrix_ordered)], 
                        col=mtDNAclone2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  "BCR clone" = anno_simple(x= barcode2BCRclone[colnames(cell_matrix_ordered)], 
                        col=BCRclone2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  "Condition" = anno_simple(x= barcode2condition[colnames(cell_matrix_ordered)], 
                        col=condition2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  "Cell type" = anno_simple(x = barcode2celltype[colnames(cell_matrix_ordered)], 
                            col = celltype2colour, 
                            border = TRUE, 
                            height = unit(0.19, "cm")),
  which = "column",
  show_annotation_name = TRUE,
  show_legend = FALSE
)

  
  # Colouring
  # edit legend manually to >6 as max cut off
  col_fun <- colorRamp2(c(0, 2), c("darkred", "yellow"))
  
  hm <- Heatmap(matrix = cell_matrix_ordered,
                top_annotation = column_ha,
                cluster_rows=FALSE,
                cluster_columns = FALSE ,
                col = col_fun,
                na_col = "white",
                border_gp = gpar(col = "black", lty = 1),
                raster_by_magick = TRUE,
                use_raster = TRUE,
                raster_quality = 6,
                
                #column parameters
                column_title = patient_id2plotting_id_intern[patient_name],
                column_title_gp = gpar(fontsize = 10),
                show_column_names = FALSE,
                
                #row parameters
                row_title = "mtDNA mutation",
                row_title_gp = gpar(fontsize = 10),
                show_row_names = TRUE,

                #legend parameters
                heatmap_legend_param = list(
                  title = "Heteroplasmy",
                  direction = "vertical",
                  legend_direction = "vertical",
                  at = c(2, 1, 0),
                  labels = c("≤ 1", "10", "100"),
                  title_gp = gpar(fontsize = 10, fontface = "plain") ,
                  labels_gp = gpar(
                       fontsize = 10,
                       fontface = "plain")))
  
  # needs to be a grob for heatmap to be placable in cowplot grid
  hm_grob <- grid::grid.grabExpr(draw(hm,
     show_heatmap_legend=FALSE))
  
  return(hm_grob)
}

```

```{r}
celltypes <- heteroplasmy_intersect.cln.variants.MBL %>% 
  map(~tibble(label_cellbarcode=colnames(.x)) %>% 
       left_join(celltype_annotation_RNA %>% 
                   dplyr::select(label_cellbarcode, My_annotation_LC), by = join_by(label_cellbarcode)) %>% 
        dplyr::select( label_cellbarcode, My_annotation_LC) %>% 
        deframe())

mtDNA_clones <- find_clonotype_clusters_MBL %>% 
  map(~dplyr::rename(.x, mtDNA_clone=Cluster) %>% 
        dplyr::mutate(label_cellbarcode=paste0(label_cellbarcode, "-1")) %>% 
        deframe())

BCR_clones <- clone_matching_data %>% 
  split(.$patient_id) %>%
  map(~dplyr::select(., label_cellbarcode, BCR_clone) %>% 
  deframe())

conditions <- heteroplasmy_intersect.cln.variants.MBL %>% 
  map(~tibble(label_cellbarcode=colnames(.x)) %>% 
      tidyr::separate(col = label_cellbarcode, into=c("label", "cellbarcode"), sep="#", remove = FALSE) %>% 
       left_join(metadata_all %>% dplyr::filter(data_type=="nanoranger") %>% 
                   dplyr::select(label, MBL_subtype), by = join_by(label)) %>% 
       dplyr::rename(condition=MBL_subtype) %>%
      dplyr::select(label_cellbarcode, condition) %>% 
       deframe())

set.seed(128)
mt_colours <- mtDNA_clones %>%
  map(~sample(colorRampPalette(brewer.pal(12, "Set3"))(length(unique(.x))), replace = FALSE) %>% 
        setNames(unique(.x)))

set.seed(466)
BCR_colours <- clone_matching_data %>%
  dplyr::mutate(BCR_clone=as.character(BCR_clone)) %>% 
  split(.$patient_id) %>%
  map(~sample(colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(.x$BCR_clone))), replace = FALSE) %>% 
        setNames(unique(.x$BCR_clone)))
```

```{r fig.height=8, fig.width=15}
nanoranger_mtDNAclone_celltype_heteroplasmy_heatmaps <- list()
for (p in names(heteroplasmy_intersect.cln.variants.MBL)){
  print(p)
  hm <- make_mtDNAclone_celltype_heteroplasmy_heatmap(meta_data=metadata_all,
                                                      heteroplasmy_matrix=heteroplasmy_intersect.cln.variants.MBL[[p]], # variants in rownames; cell barcodes in column names
                                                         patient_name=p, # character string
                                                         celltype_annotation=celltypes[[p]], # named vector
                                                         mtDNA_clone_annotation=mtDNA_clones[[p]], # named vector
                                                         mtDNA_clone_colours=mt_colours[[p]], # named vector
                                                         BCR_clone_annotation=BCR_clones[[p]],
                                                         BCR_clone_colours=BCR_colours[[p]],
                                                         condition_annotation=conditions[[p]], # named vector
                                                         condition_colours=my_MBL_CLL_colours_named # named vector
  )
  nanoranger_mtDNAclone_celltype_heteroplasmy_heatmaps[[p]] <- hm
  
}  
names(nanoranger_mtDNAclone_celltype_heteroplasmy_heatmaps) <- patient_id2plotting_id[names(nanoranger_mtDNAclone_celltype_heteroplasmy_heatmaps)]

nanoranger_mtDNAclone_celltype_heteroplasmy_heatmaps[c(paste0("P", 1:6), "P9dup")] %>% 
  map(~plot_grid(.x, nrow=1, ncol=1))

saveRDS(nanoranger_mtDNAclone_celltype_heteroplasmy_heatmaps[c(paste0("P", 1:6), "P9dup")] , file.path(RDS_dir, "nanoranger_mtDNAclone_celltype_heteroplasmy_heatmaps.RDS"))
```

```{r}
heteroplasmy_intersect.cln.variants.MBL<- intersect.cln.variants.MBL %>% 
  map(~dplyr::filter(.,label_cellbarcode %in% clone_matching_data$label_cellbarcode) %>% 
  tidyr::pivot_wider(id_cols = c(variant_id), names_from = label_cellbarcode, values_from = VAF) %>% 
  column_to_rownames("variant_id") %>% 
  as.matrix())

celltypes <- clone_matching_data %>% 
  dplyr::left_join(celltype_annotation_RNA %>% dplyr::select(label_cellbarcode, My_annotation_LC), by = join_by(label_cellbarcode)) %>% 
  dplyr::mutate(Celltype=case_when(My_annotation_LC %in% c("NK", "T", "Mono") ~ "Other immune cells",
                                   is.na(My_annotation_LC) ~ "Unannotated",
                                   .default=My_annotation_LC)) %>% 
  split(.$patient_id) %>%
  .[names(.)!="MBL13"] %>% 
  map(~dplyr::select(., label_cellbarcode, Celltype) %>% 
  deframe())

mtDNA_clones <- clone_matching_data %>% 
  split(.$patient_id) %>% 
    .[names(.)!="MBL13"] %>% 
  map(~dplyr::select(., label_cellbarcode, mtDNA_clone) %>%
        dplyr::mutate(mtDNA_clone=as.character(mtDNA_clone)) %>% 
  deframe())
                
BCR_clones <- clone_matching_data %>% 
  split(.$patient_id) %>%
    .[names(.)!="MBL13"] %>% 
  map(~dplyr::select(., label_cellbarcode, BCR_clone) %>% 
  deframe())

set.seed(466)
BCR_colours <- clone_matching_data %>%
  dplyr::mutate(BCR_clone=as.character(BCR_clone)) %>% 
  split(.$patient_id) %>%
    .[names(.)!="MBL13"] %>% 
  map(~sample(colorRampPalette(brewer.pal(8, "Set2"))(length(unique(.x$BCR_clone))), replace = FALSE) %>% 
        setNames(unique(.x$BCR_clone)))
```

#### Ordered by mtDNA clone
```{r}
order_matrix_mtDNA_clone <- function(mat, cell_ann, first_ann, second_ann, enrichment_ann){
  geneOrder <- sort(rowSums(mat), decreasing = TRUE, index.return = TRUE)$ix
  
  scoreCol <- function(x) {
    score <- 0
    for (i in 1:length(x)) {
      if (x[i]) {
        score <- score + 2^(length(x) - i)
      }
    }
    return(score)
  }
  
  scores <- apply(mat[geneOrder, , drop = FALSE], 2, scoreCol)
  sampleOrder <- sort(scores, decreasing = TRUE, index.return = TRUE)$ix

  geneSort <- sort(rownames(mat))
  
  enrich_levels <- c("B", "LC_MBL", "HC_MBL","CLL", "not significant")
  
  enrichment_annotation <- enrichment_ann %>% 
    factor(., levels=enrich_levels)
  
  geneSortEnrich <- sort(enrichment_annotation) %>% names() %>% unique()
  
  # Order cols by first annotation
  first_ann_freq <- enframe(first_ann,
                      name="label_cellbarcode",
                      value = "mtDNA_clone") %>%
  dplyr::group_by(mtDNA_clone) %>%
  dplyr::summarise(frequency_first=n()) %>%
  dplyr::arrange(desc(frequency_first)) %>%
  dplyr::mutate(mtDNA_clone=factor(mtDNA_clone, mtDNA_clone))%>%
  pull(mtDNA_clone)
  
  barcodes_first_ann_ordered <- tibble("label_cellbarcode_M"=colnames(mat),
         "scores_memosort"=scores) %>% 
    left_join(enframe(first_ann, 
                      name="label_cellbarcode", 
                      value = "mtDNA_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(mtDNA_clone=factor(mtDNA_clone, levels=levels(first_ann_freq))) %>% 
    dplyr::group_by(mtDNA_clone) %>% 
    dplyr::arrange(mtDNA_clone, desc(scores_memosort)) %>% 
    pull(label_cellbarcode_M)
  
  barcodes_first_second_ann_ordered <- tibble("label_cellbarcode_M"=colnames(mat),
         "scores_memosort"=scores) %>% 
    left_join(enframe(first_ann, 
                      name="label_cellbarcode", 
                      value = "mtDNA_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    left_join(enframe(second_ann, 
                      name="label_cellbarcode", 
                      value = "BCR_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(mtDNA_clone=factor(mtDNA_clone, levels=levels(first_ann_freq))) %>% 
    dplyr::group_by(mtDNA_clone) %>% 
    dplyr::arrange(mtDNA_clone, desc(BCR_clone), desc(scores_memosort)) %>% 
    pull(label_cellbarcode_M)
  
  barcodes_celltype_first_second_ann_ordered <- tibble("label_cellbarcode_M"=colnames(mat),
         "scores_memosort"=scores) %>%
    left_join(enframe(cell_ann, 
                      name="label_cellbarcode", 
                      value = "celltype"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    left_join(enframe(first_ann, 
                      name="label_cellbarcode", 
                      value = "mtDNA_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    left_join(enframe(second_ann, 
                      name="label_cellbarcode", 
                      value = "BCR_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(mtDNA_clone=factor(mtDNA_clone, levels=levels(first_ann_freq))) %>% 
    dplyr::group_by(mtDNA_clone) %>% 
    dplyr::arrange(celltype, mtDNA_clone, desc(BCR_clone), desc(scores_memosort)) %>% 
    pull(label_cellbarcode_M)
  
      return(list(geneSortEnrich, barcodes_celltype_first_second_ann_ordered))
}
```


```{r}
make_mtDNAclone_matching_heteroplasmy_heatmap <- function(heteroplasmy_matrix, # variants in rownames; cell barcodes in column names
                                                         patient_name, # character string
                                                         changes_oi,
                                                         celltype_annotation, # named vector
                                                         mtDNA_clone_annotation, # named vector
                                                         mtDNA_clone_colours, # named vector
                                                         BCR_clone_annotation, # named vector
                                                         BCR_clone_colours, # named vector
                                                         enrichment_annotation,
                                                         enrichment_colours)
{
  cell_matrix <-heteroplasmy_matrix  %>%
    .[rownames(heteroplasmy_matrix) %in% changes_oi$variant_id, , drop=FALSE] %>%
    .[rowSums(.) != 0,,drop=FALSE] %>% 
    .[,colSums(is.na(.))<nrow(.),drop=FALSE]
  
  ordering <- order_matrix_mtDNA_clone(cell_matrix,
                           celltype_annotation,
                           mtDNA_clone_annotation,
                           BCR_clone_annotation,
                           enrichment_ann = enrichment_annotation)
  

  cell_matrix <- -log10(cell_matrix)
  
  #reconcatenate columns and replace 0 with NA for complex heatmap
  cell_matrix <- as.matrix(cell_matrix)
  cell_matrix[cell_matrix == Inf] <- NA
  
  # order matrix in place
  cell_matrix_ordered <- cell_matrix[ordering[[1]][ordering[[1]] %in% rownames(cell_matrix)], ordering[[2]], drop=FALSE]
  
  # prepare top celltype annotation
  celltype2colour <- c("LC-MBL B Cell"="#FCBDA4","HC_MBL"="#FC8D62", "CLL" ="#66C2A5","B"="darkred", "Unannotated"="lightblue", "Other immune cells"="rosybrown")
  barcode2celltype <- celltype_annotation
  
  # prepare top mtDNA annotations
  mtDNAclone2colour <- mtDNA_clone_colours
  barcode2mtDNAclone <- mtDNA_clone_annotation
  
  # prepare top BCR annotation
  BCRclone2colour <- BCR_clone_colours
  barcode2BCRclone <- BCR_clone_annotation
  
  column_ha <- HeatmapAnnotation(
  "Cell type" = anno_simple(x = barcode2celltype[colnames(cell_matrix_ordered)], 
                            col = celltype2colour, 
                            border = TRUE, 
                            height = unit(0.19, "cm")),
  "mtDNA clone" = anno_simple(x= barcode2mtDNAclone[colnames(cell_matrix_ordered)], 
                        col=mtDNAclone2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  "BCR clone" = anno_simple(x= barcode2BCRclone[colnames(cell_matrix_ordered)], 
                        col=BCRclone2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  which = "column",
  show_annotation_name = TRUE,
  show_legend = FALSE
)

  # prepare row annotation
  enrichment2colour <- enrichment_colours
  barcode2enrichment <- enrichment_annotation

  row_ha <- rowAnnotation(
    "Enrichment" =anno_simple(
    x=barcode2enrichment[rownames(cell_matrix_ordered)],
    col = enrichment2colour, # is list of named vector
    border = TRUE,
    width = unit(0.19, "cm")
    ),
    show_legend = FALSE,
    show_annotation_name =FALSE)
  
  # Colouring
  # edit legend manually to >6 as max cut off
  col_fun <- colorRamp2(c(0, 2), c("darkred", "yellow"))
  
  hm <- Heatmap(matrix = cell_matrix_ordered,
                top_annotation = column_ha,
                right_annotation = row_ha,
                column_split = barcode2celltype[colnames(cell_matrix_ordered)],
                cluster_rows=FALSE,
                cluster_columns = FALSE ,
                col = col_fun,
                na_col = "white",
                border_gp = gpar(col = "black", lty = 1),
                raster_by_magick = TRUE,
                use_raster = TRUE,
                raster_quality = 6,
                
                #column parameters
                column_title = patient_name,
                column_title_gp = gpar(fontsize = 10),
                show_column_names = FALSE,
                
                #row parameters
                row_title = "mtDNA mutation",
                row_title_gp = gpar(fontsize = 10),
                show_row_names = TRUE,

                #legend parameters
                heatmap_legend_param = list(
                  title = "Heteroplasmy",
                  direction = "vertical",
                  legend_direction = "vertical",
                  at = c(2, 1, 0),
                  labels = c("≤ 1", "10", "100"),
                  title_gp = gpar(fontsize = 10, fontface = "plain") ,
                  labels_gp = gpar(
                       fontsize = 10,
                       fontface = "plain")))
  
  # needs to be a grob for heatmap to be placable in cowplot grid
  hm_grob <- grid::grid.grabExpr(draw(hm,
     show_heatmap_legend=FALSE))
  
  return(hm_grob)
}
```

```{r, fig.height=6, fig.width=30, warning=FALSE, message=FALSE}
# filter celltypes according to comparisons and make heatmaps
B_significant_heteroplasmy_changes <- all_significant_ATAC_heteroplasmy_changes %>%
    dplyr::filter(!(comparison %in% c("B_vs_LC_others", "B_vs_HC_others", "B_vs_CLL_others")),
                  patient_id %in% nanoranger_samples) %>%
    dplyr::mutate(comparison = factor(comparison, levels = c("B_vs_LC_MBL", "B_vs_HC_MBL", "B_vs_CLL"))) %>%
    dplyr::arrange(desc(max_blike_heteroplasmy)) %>% 
  split(.$patient_id) %>% 
    list_drop_empty

B_insignificant_heteroplasmy_changes <- all_insignificant_heteroplasmy_changes %>% 
  dplyr::filter(!(comparison %in% c("B_vs_LC_others", "B_vs_HC_others", "B_vs_CLL_others")),
                  patient_id %in% nanoranger_samples) %>%
    dplyr::mutate(comparison = factor(comparison, levels = c("B_vs_LC_MBL", "B_vs_HC_MBL", "B_vs_CLL"))) %>% 
    dplyr::arrange(desc(max_blike_heteroplasmy)) %>% 
    split(.$patient_id) %>% 
    map(~slice_head(., n = 5)) %>% 
    list_drop_empty

B_heteroplasmy_changes <- map2(.x= B_significant_heteroplasmy_changes,
                               .y= B_insignificant_heteroplasmy_changes[names(B_insignificant_heteroplasmy_changes) != "MBL13"],
                               ~bind_rows(.x, .y))

enrichment <- map2(
  .x = B_significant_heteroplasmy_changes %>%
    map(~ dplyr::filter(., comparison %in% c("B_vs_LC_MBL", "B_vs_HC_MBL", "B_vs_CLL")) %>%
      tidyr::separate(comparison, into = c("numerator", "denominator"), sep = "_vs_", remove = FALSE) %>%
      dplyr::mutate(enrichment = if_else(Change > 1, denominator, numerator)) %>%
      dplyr::select(variant_id, enrichment) %>%
      deframe()),
  .y = B_insignificant_heteroplasmy_changes[names(B_insignificant_heteroplasmy_changes) != "MBL13"] %>%
    map(~ dplyr::filter(., comparison %in% c("B_vs_LC_MBL", "B_vs_HC_MBL", "B_vs_CLL")) %>%
      dplyr::mutate(enrichment = "not significant") %>%
      dplyr::arrange(desc(max_blike_heteroplasmy)) %>% 
      dplyr::select(variant_id, enrichment) %>%
      deframe()), 
  .f = ~ c(.x, .y)
)


enrichment_cols <- setNames(
  replicate(length(paste0("MBL", 10:15)), c(my_MBL_CLL_colours_named, "not significant"="lightgrey"), simplify = FALSE),
  paste0("MBL", 10:15)
)

mtDNA_matched_clonotype_heatmaps <- list()
patient=c("MBL10", "MBL11", "MBL12",  "MBL14", "MBL15")
for (p in patient){
  print(p)
  hm <- make_mtDNAclone_matching_heteroplasmy_heatmap(
    heteroplasmy_matrix=heteroplasmy_intersect.cln.variants.MBL[[p]], 
    patient_name = p,
    changes_oi=B_heteroplasmy_changes[[p]],
    celltype_annotation= celltypes[[p]],
    mtDNA_clone_annotation = mtDNA_clones[[p]],
    mtDNA_clone_colours = mt_colours[[p]],
    BCR_clone_annotation = BCR_clones[[p]],
    BCR_clone_colours= BCR_colours[[p]],
    enrichment_annotation = enrichment[[p]],
    enrichment_colours = enrichment_cols[[p]])
  
  mtDNA_matched_clonotype_heatmaps[[p]] <- hm
}

cowplot::plot_grid(plotlist = mtDNA_matched_clonotype_heatmaps, nrow=2, ncol=3)

saveRDS(mtDNA_matched_clonotype_heatmaps, file.path(RDS_dir, "mtDNA_matched_clonotype_heatmaps.RDS"))

ggsave(plot=cowplot::plot_grid(plotlist = mtDNA_matched_clonotype_heatmaps, nrow=2, ncol=3),
       file= file.path(plotting_dir, "nanoranger_BCR_mtDNA_clone_heatmaps_mtDNA_ordered.png"), 
       width = 30, 
       height=6)
```

#### Ordered by BCR clone
```{r}
order_matrix_BCR_clone <- function(mat, cell_ann, ann, enrichment_ann){
  geneOrder <- sort(rowSums(mat), decreasing = TRUE, index.return = TRUE)$ix
  
  scoreCol <- function(x) {
    score <- 0
    for (i in 1:length(x)) {
      if (x[i]) {
        score <- score + 2^(length(x) - i)
      }
    }
    return(score)
  }
  
  geneSort <- sort(rownames(mat))
  
  enrich_levels <- c("B", "LC_MBL", "HC_MBL","CLL", "not significant")
  
  enrichment_annotation <- enrichment_ann %>% 
    factor(., levels=enrich_levels)
  
  geneSortEnrich <- sort(enrichment_annotation) %>% names() %>% unique()
  
  scores <- apply(mat[geneOrder, , drop = FALSE], 2, scoreCol)
  sampleOrder <- sort(scores, decreasing = TRUE, index.return = TRUE)$ix

  ann_freq <- enframe(ann,
                      name="label_cellbarcode",
                      value = "BCR_clone") %>%
  group_by(BCR_clone) %>%
  dplyr::summarise(freq_second=n()) %>%
  dplyr::arrange(desc(freq_second)) %>%
  dplyr::mutate(BCR_clone=factor(BCR_clone),
                BCR_clone=fct_reorder(BCR_clone, freq_second, .desc = TRUE)) %>%
  pull(BCR_clone)
  
  barcodes_celltype_ann_ordered <- tibble("label_cellbarcode_M"=colnames(mat),
         "scores_memosort"=scores) %>%
    left_join(enframe(cell_ann, 
                      name="label_cellbarcode", 
                      value = "celltype"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    left_join(enframe(ann, 
                      name="label_cellbarcode", 
                      value = "BCR_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(BCR_clone=factor(BCR_clone, levels=levels(ann_freq)))
  
      
  set.seed(123)
  barcodes_celltype_ann_ordered <- barcodes_celltype_ann_ordered[sample(nrow(barcodes_celltype_ann_ordered), replace = FALSE), ] %>% 
    dplyr::arrange(celltype, desc(BCR_clone)) %>% 
    pull(label_cellbarcode_M)
 
    return(list(geneSortEnrich, barcodes_celltype_ann_ordered))
}
```


```{r}
make_BCRclone_matching_heteroplasmy_heatmap <- function(meta_data,
                                                        heteroplasmy_matrix, # variants in rownames; cell barcodes in column names
                                                         patient_name, # character string
                                                         changes_oi,
                                                         celltype_annotation, # named vector
                                                         BCR_clone_annotation, # named vector
                                                         BCR_clone_colours, # named vector
                                                         enrichment_annotation,
                                                         enrichment_colours)
  {
  patient_id2plotting_id_intern <- meta_data %>%
  dplyr::filter(data_type == "nanoranger") %>%
  dplyr::select(patient_id, plotting_id) %>%
  unique() %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, 
                                   levels=c(paste0("P", 1:9), "P9dup", paste0("P", 10:24)))) %>% 
  deframe()
  
  cell_matrix <-heteroplasmy_matrix  %>%
    .[rownames(heteroplasmy_matrix) %in% changes_oi$variant_id, , drop=FALSE] %>%
    .[rowSums(.) != 0,,drop=FALSE] %>% 
    .[,colSums(is.na(.))<nrow(.),drop=FALSE]
  
  ordering <- order_matrix_BCR_clone(mat=cell_matrix,
                           cell_ann=celltype_annotation,
                           ann=BCR_clone_annotation,
                           enrichment_ann = enrichment_annotation)
  

  cell_matrix <- -log10(cell_matrix)
  
  #reconcatenate columns and replace 0 with NA for complex heatmap
  cell_matrix <- as.matrix(cell_matrix)
  cell_matrix[cell_matrix == Inf] <- NA
  
  # order matrix in place
  cell_matrix_ordered <- cell_matrix[ordering[[1]][ordering[[1]] %in% rownames(cell_matrix)], ordering[[2]], drop=FALSE]
  
  write_tsv(as_tibble(ordering[[1]][ordering[[1]] %in% rownames(cell_matrix)]), file.path(tsv_dir, paste0(patient_name, "_mtDNA_mutations_plotted.tsv")))
  
  # prepare top celltype annotation
  celltype2colour <- c(annotation_colours_named, "Unannotated"="lightblue", "Other immune cells"="rosybrown")
  barcode2celltype <- celltype_annotation
  
  # prepare top BCR annotation
  BCRclone2colour <- BCR_clone_colours
  barcode2BCRclone <- BCR_clone_annotation
  
  column_ha <- HeatmapAnnotation(
  "Cell type" = anno_simple(x = barcode2celltype[colnames(cell_matrix_ordered)], 
                            col = celltype2colour, 
                            border = TRUE, 
                            height = unit(0.19, "cm")),
  "BCR clone" = anno_simple(x= barcode2BCRclone[colnames(cell_matrix_ordered)], 
                        col=BCRclone2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  which = "column",
  show_annotation_name = TRUE,
  show_legend = FALSE
)
  
  # prepare row annotation
  enrichment2colour <- enrichment_colours
  barcode2enrichment <- enrichment_annotation

  row_ha <- rowAnnotation(
    "Enrichment" =anno_simple(
    x=barcode2enrichment[rownames(cell_matrix_ordered)],
    col = enrichment2colour, # is list of named vector
    border = TRUE,
    width = unit(0.19, "cm")
    ),
    show_legend = FALSE,
    show_annotation_name =FALSE)
  
  # Colouring
  # edit legend manually to >6 as max cut off
  col_fun <- colorRamp2(c(0, 2), c("darkred", "yellow"))
  
  hm <- Heatmap(matrix = cell_matrix_ordered,
                top_annotation = column_ha,
                right_annotation = row_ha,
                column_split = barcode2celltype[colnames(cell_matrix_ordered)],
                cluster_rows=FALSE,
                cluster_columns = FALSE ,
                col = col_fun,
                na_col = "white",
                border_gp = gpar(col = "black", lty = 1),
                raster_by_magick = TRUE,
                use_raster = TRUE,
                raster_quality = 6,
                
                #column parameters
                column_title = patient_id2plotting_id_intern[patient_name],
                column_title_gp = gpar(fontsize = 10),
                show_column_names = FALSE,
                
                #row parameters
                row_title = "mtDNA mutation",
                row_title_gp = gpar(fontsize = 10),
                show_row_names = TRUE,

                #legend parameters
                heatmap_legend_param = list(
                  title = "Heteroplasmy",
                  direction = "vertical",
                  legend_direction = "vertical",
                  at = c( 2, 1, 0),
                  labels = c("≤ 1", "10", "100"),
                  title_gp = gpar(fontsize = 10, fontface = "plain") ,
                  labels_gp = gpar(
                       fontsize = 10,
                       fontface = "plain")))
  
  # needs to be a grob for heatmap to be placable in cowplot grid
  hm_grob <- grid::grid.grabExpr(draw(hm,
     show_heatmap_legend=FALSE))
  
  return(hm_grob)
}

```

```{r, fig.height=6, fig.width=30, warning=FALSE, message=FALSE}
BCR_matched_clonotype_heatmaps <- list()
patient=c("MBL10", "MBL11", "MBL12",  "MBL14", "MBL15")
for (p in patient){
  print(p)
  hm <- make_BCRclone_matching_heteroplasmy_heatmap(meta_data=metadata_all,
    heteroplasmy_matrix=heteroplasmy_intersect.cln.variants.MBL[[p]], 
    patient_name = p,
    changes_oi=B_heteroplasmy_changes[[p]],
    celltype_annotation= celltypes[[p]],
    BCR_clone_annotation = BCR_clones[[p]],
    BCR_clone_colours= BCR_colours[[p]],
    enrichment_annotation = enrichment[[p]],
    enrichment_colours=enrichment_cols[[p]])
  
  BCR_matched_clonotype_heatmaps[[p]] <- hm
}

cowplot::plot_grid(plotlist = BCR_matched_clonotype_heatmaps, nrow=2, ncol=3)

saveRDS(BCR_matched_clonotype_heatmaps, file.path(RDS_dir, "BCR_matched_clonotype_heatmaps.RDS"))

ggsave(plot=cowplot::plot_grid(plotlist = BCR_matched_clonotype_heatmaps, nrow=2, ncol=3),
       file= file.path(plotting_dir, "nanoranger_BCR_mtDNA_clone_heatmaps_BCR_ordered.png"), 
       width = 30, 
       height=6)
```

#### Ordered by BCR clone and CNVs
```{r fig.width=20, fig.height=3}
order_matrix_BCR_clone_CNV <- function(mat, cell_ann, first_ann, second_ann, enrichment_ann){
  geneOrder <- sort(rowSums(mat), decreasing = TRUE, index.return = TRUE)$ix
  
  scoreCol <- function(x) {
    score <- 0
    for (i in 1:length(x)) {
      if (x[i]) {
        score <- score + 2^(length(x) - i)
      }
    }
    return(score)
  }
  
  scores <- apply(mat[geneOrder, , drop = FALSE], 2, scoreCol)
  sampleOrder <- sort(scores, decreasing = TRUE, index.return = TRUE)$ix

  geneSort <- sort(rownames(mat))
  
  enrich_levels <- c("B", "LC_MBL", "HC_MBL","CLL", "not significant")
  
  enrichment_annotation <- enrichment_ann %>% 
    factor(., levels=enrich_levels)
  
  geneSortEnrich <- sort(enrichment_annotation) %>% names() %>% unique()
  
  # Order cols by first annotation
  first_ann_freq <- enframe(first_ann,
                      name="label_cellbarcode",
                      value = "BCR_clone") %>%
  group_by(BCR_clone) %>%
  dplyr::summarise(freq_second=n()) %>%
  dplyr::arrange(desc(freq_second)) %>%
  dplyr::mutate(BCR_clone=factor(BCR_clone),
                BCR_clone=fct_reorder(BCR_clone, freq_second, .desc = TRUE)) %>%
  pull(BCR_clone)
  
  barcodes_first_ann_ordered <- tibble("label_cellbarcode_M"=colnames(mat),
         "scores_memosort"=scores) %>% 
    left_join(enframe(first_ann, 
                      name="label_cellbarcode", 
                      value = "BCR_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(BCR_clone=factor(BCR_clone, levels=levels(first_ann_freq))) %>% 
    dplyr::group_by(BCR_clone) %>% 
    dplyr::arrange(BCR_clone, desc(scores_memosort)) %>% 
    pull(label_cellbarcode_M)
  
  barcodes_first_second_ann_ordered <- tibble("label_cellbarcode_M"=colnames(mat),
         "scores_memosort"=scores) %>% 
    left_join(enframe(first_ann, 
                      name="label_cellbarcode", 
                      value = "BCR_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    left_join(enframe(second_ann, 
                      name="label_cellbarcode", 
                      value = "CNV"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(BCR_clone=factor(BCR_clone, levels=levels(first_ann_freq))) %>% 
    dplyr::group_by(BCR_clone) %>% 
    dplyr::arrange(BCR_clone, desc(CNV), desc(scores_memosort)) %>% 
    pull(label_cellbarcode_M)
  
  barcodes_celltype_first_second_ann_ordered <- tibble("label_cellbarcode_M"=colnames(mat),
         "scores_memosort"=scores) %>%
    left_join(enframe(cell_ann, 
                      name="label_cellbarcode", 
                      value = "celltype"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    left_join(enframe(first_ann, 
                      name="label_cellbarcode", 
                      value = "BCR_clone"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    left_join(enframe(second_ann, 
                      name="label_cellbarcode", 
                      value = "CNV"), 
              by = join_by("label_cellbarcode_M" == "label_cellbarcode")) %>% 
    dplyr::mutate(BCR_clone=factor(BCR_clone, levels=levels(first_ann_freq))) 
  
    set.seed(123)
    barcodes_celltype_first_second_ann_ordered <- barcodes_celltype_first_second_ann_ordered[sample(nrow(barcodes_celltype_first_second_ann_ordered), replace = FALSE), ] %>%
    dplyr::group_by(BCR_clone) %>% 
    dplyr::arrange(celltype, BCR_clone, desc(CNV)) %>% 
    pull(label_cellbarcode_M)

    return(list(geneSortEnrich, barcodes_celltype_first_second_ann_ordered))
}
```


```{r fig.width=20, fig.height=3}
make_BCRclone_CNV_matching_heteroplasmy_heatmap <- function(numbat_object,
                                                            meta_data,
                                                            heteroplasmy_matrix, # variants in rownames; cell barcodes in column names
                                                            patient_name, # character string
                                                            changes_oi,
                                                            celltype_annotation, # named vector
                                                            BCR_clone_annotation, # named vector
                                                            BCR_clone_colours, # named vector
                                                            CNV_annotation,# named vector
                                                            CNV_colours, # named vector
                                                            enrichment_annotation,
                                                            enrichment_colours)
{
  patient_id2plotting_id_intern <- meta_data %>%
  dplyr::filter(data_type == "nanoranger") %>%
  dplyr::select(patient_id, plotting_id) %>%
  unique() %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, 
                                   levels=c(paste0("P", 1:9), "P9dup", paste0("P", 10:24)))) %>% 
  deframe()
  
  cell_matrix <-heteroplasmy_matrix  %>%
    .[rownames(heteroplasmy_matrix) %in% changes_oi$variant_id, , drop=FALSE] %>%
    .[rowSums(.) != 0,,drop=FALSE] %>% 
    .[,colSums(is.na(.))<nrow(.),drop=FALSE]
  
  ordering <- order_matrix_BCR_clone_CNV(mat=cell_matrix,
                           cell_ann=celltype_annotation,
                           first_ann=BCR_clone_annotation,
                           second_ann=CNV_annotation,
                           enrichment_ann=enrichment_annotation)

  cell_matrix <- -log10(cell_matrix)
  
  #reconcatenate columns and replace 0 with NA for complex heatmap
  cell_matrix <- as.matrix(cell_matrix)
  cell_matrix[cell_matrix == Inf] <- NA
  
  # order matrix in place
  cell_matrix_ordered <- cell_matrix[ordering[[1]][ordering[[1]] %in% rownames(cell_matrix)], ordering[[2]], drop=FALSE]
  
  # prepare top celltype annotation
  celltype2colour <- c(annotation_colours_named, "Unannotated"="lightblue", "Other immune cells"="rosybrown")
  barcode2celltype <- celltype_annotation
  
  # prepare top BCR annotation
  BCRclone2colour <- BCR_clone_colours
  barcode2BCRclone <- BCR_clone_annotation
  
  # prepare top CNV annotation
  CNV2colour <- CNV_colours
  barcode2CNV <- CNV_annotation  

    column_ha <- HeatmapAnnotation(
  "Cell type" = anno_simple(x = barcode2celltype[colnames(cell_matrix_ordered)], 
                            col = celltype2colour, 
                            border = TRUE, 
                            height = unit(0.19, "cm")),
  "BCR clone" = anno_simple(x= barcode2BCRclone[colnames(cell_matrix_ordered)], 
                        col=BCRclone2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  "CNV" = anno_simple(x= barcode2CNV[colnames(cell_matrix_ordered)], 
                        col=CNV2colour,
                        border = TRUE,
                        height = unit(0.19, "cm")),
  which = "column",
  show_annotation_name = TRUE,
  show_legend = FALSE
)

  # prepare row annotation
  enrichment2colour <- enrichment_colours
  barcode2enrichment <- enrichment_annotation
  
  colours_r <- list(
    "Enrichment" = enrichment2colour)
  
  anno_df_r <- tibble(
    "Enrichment"=barcode2enrichment[rownames(cell_matrix_ordered)]
  )
  
  row_ha <- rowAnnotation(
    "Enrichment" =anno_simple(
    x=barcode2enrichment[rownames(cell_matrix_ordered)],
    col = enrichment2colour, # is list of named vector
    border = TRUE,
    width = unit(0.19, "cm")
    ),
    show_legend = FALSE,
    show_annotation_name =FALSE)
  
  # Colouring
  # edit legend manually to >2 as max cut off
  col_fun <- colorRamp2(c(0, 2), c("darkred", "yellow"))
  
  hm <- Heatmap(matrix = cell_matrix_ordered,
                top_annotation = column_ha,
                right_annotation = row_ha,
                column_split = barcode2celltype[colnames(cell_matrix_ordered)],
                cluster_rows=FALSE,
                cluster_columns = FALSE ,
                col = col_fun,
                na_col = "white",
                border_gp = gpar(col = "black", lty = 1),
                raster_by_magick = TRUE,
                use_raster = TRUE,
                raster_quality = 6,
                
                #column parameters
                column_title = patient_id2plotting_id_intern[patient_name],
                column_title_gp = gpar(fontsize = 10),
                show_column_names = FALSE,
                
                #row parameters
                row_title = "mtDNA mutation",
                row_title_gp = gpar(fontsize = 10),
                show_row_names = TRUE,

                #legend parameters
                heatmap_legend_param = list(
                  title = "Heteroplasmy",
                  direction = "vertical",
                  legend_direction = "vertical",
                  at = c( 2,1, 0),
                  labels = c("≤ 1", "10", "100"),
                  title_gp = gpar(fontsize = 10, fontface = "plain") ,
                  labels_gp = gpar(
                       fontsize = 10,
                       fontface = "plain")))
  
  # needs to be a grob for heatmap to be placable in cowplot grid
  hm_grob <- grid::grid.grabExpr(draw(hm,
     show_heatmap_legend = FALSE))
  
  return(hm_grob)
}
```


```{r fig.width=20, fig.height=3}
# Create CNV annotation per patient
CNVs <- list()
CNV_cols <- list()
for (p in c("MBL12", "MBL15")){
  print(p)
  if (p == "MBL12"){
    CNVs[[p]] <-   nb_lst[[p]]$joint_post %>% 
    dplyr::select(cell, CHROM, p_del) %>% 
    pivot_wider(id_cols = cell, names_from = CHROM, values_from = p_del) %>% 
    dplyr::mutate(cell=str_replace(string = cell, pattern = "_", replacement = "#")) %>% 
    dplyr::select(cell, `14`) %>% 
    deframe()
    
    CNV_cols[[p]] <- colorRamp2(breaks = c(0, 0.5, 1), c("white", "lightblue", "darkblue"))

    }
  
  if (p == "MBL15"){
    CNVs[[p]] <-   nb_lst[[p]]$joint_post %>% 
    dplyr::select(cell, CHROM, p_amp) %>% 
    pivot_wider(id_cols = cell, names_from = CHROM, values_from = p_amp) %>% 
    dplyr::mutate(cell=str_replace(string = cell, pattern = "_", replacement = "#")) %>% 
    dplyr::select(cell, `12`) %>% 
    deframe()
    
    CNV_cols[[p]] <- colorRamp2(breaks = c(0, 0.5, 1), c("white", "lightpink1", "darkred"))
  }
}

BCR_matched_clonotype_CNV_heatmaps <- list()
for (p in c("MBL12", "MBL15")){
  print(p)
  hm <- make_BCRclone_CNV_matching_heteroplasmy_heatmap(meta_data=metadata_all,
    heteroplasmy_matrix=heteroplasmy_intersect.cln.variants.MBL[[p]], 
    patient_name = p,
    changes_oi=B_heteroplasmy_changes[[p]],
    celltype_annotation= celltypes[[p]],
    BCR_clone_annotation = BCR_clones[[p]],
    BCR_clone_colours= BCR_colours[[p]],
    CNV_annotation = CNVs[[p]],
    CNV_colours = CNV_cols[[p]],
    enrichment_annotation=enrichment[[p]],
    enrichment_colours=enrichment_cols[[p]])
  
  BCR_matched_clonotype_CNV_heatmaps[[p]] <- hm
}

cowplot::plot_grid(plotlist = BCR_matched_clonotype_CNV_heatmaps, nrow=1, ncol=2)

saveRDS(BCR_matched_clonotype_CNV_heatmaps, file.path(RDS_dir, "BCR_matched_clonotype_CNV_heatmaps.RDS"))
```

## Zoom-in MBL12

### Cell type

```{r}
MBL12_celltype_umap <- celltype_annotation_RNA_umap %>% 
    dplyr::mutate(My_annotation_LC=if_else(label =="MBL12", My_annotation_LC, NA)) %>% 
    ggplot2::ggplot(aes(x=UMAP_1, y=UMAP_2, colour=My_annotation_LC))+
    geom_point_rast(size = 0.0001, raster.dpi = 600, dev = "cairo")+
    scale_colour_manual(name="Cell type", values = annotation_colours_named, na.value = "lightgrey")+
    labs(title="Cell types", x="UMAP1", y="UMAP2")+
    theme_pub()+
    theme(aspect.ratio = 1,
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "right")

saveRDS(MBL12_celltype_umap, file.path(RDS_dir, "MBL12_celltype_umap.RDS"))
ggsave(plot = MBL12_celltype_umap, 
       device = "pdf", 
       file=file.path(plotting_dir, "MBL12_celltype_umap.pdf"),
       width = 6, 
       height = 6)
```

### CNVs

```{r}
MBL12_CNVs <-   nb_lst[["MBL12"]]$joint_post %>% 
    dplyr::select(cell, CHROM, p_del) %>% 
    pivot_wider(id_cols = cell, names_from = CHROM, values_from = p_del) %>% 
    dplyr::select(cell, `14`) %>% 
    deframe()


MBL12_CNV_umap <- celltype_annotation_RNA_umap %>%
    dplyr::mutate(del14=MBL12_CNVs[label_cellbarcode]) %>% 
    dplyr::mutate(del14=if_else(label =="MBL12", del14, NA)) %>% 
    ggplot2::ggplot(aes(x=UMAP_1, y=UMAP_2, colour=del14))+
    geom_point_rast(size = 0.0001, raster.dpi = 600, dev = "cairo")+
    scale_colour_gradientn(name="Posterior probability\n14b(del)", 
                           colours = c("yellow", "darkseagreen4", "darkblue"),
                           limits=c(0,1), 
                           na.value = "lightgrey")+
  labs(title="del14", x="UMAP1", y="UMAP2")+
    theme_pub()+
    theme(aspect.ratio = 1,
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "right")

saveRDS(MBL12_CNV_umap, file.path(RDS_dir, "MBL12_CNV_umap.RDS"))
ggsave(plot = MBL12_CNV_umap, 
       device = "pdf", 
       file=file.path(plotting_dir, "MBL12_CNV_umap.pdf"),
       width = 6, 
       height = 6)
```

### BCR clones

```{r}
MBL12_BCRclone_umap <- celltype_annotation_RNA_umap %>%
  left_join(new_BCR_paired$individual %>% 
  dplyr::mutate(label_cellbarcode=str_replace(string = label_cellbarcode, pattern = "#", replacement = "_")) %>% 
  dplyr::select(label_cellbarcode, new_common_ct_id) %>% 
  dplyr::mutate(BCR_clone=as.character(new_common_ct_id)), by = join_by(label_cellbarcode)) %>%
  dplyr::mutate(BCR_clone=if_else(label=="MBL12", BCR_clone, NA)) %>% 
  ggplot2::ggplot(aes(x=UMAP_1, y=UMAP_2, colour=BCR_clone))+
  geom_point_rast(size = 0.0001, raster.dpi = 600, dev = "cairo")+
    scale_colour_manual(values = c("102"="#ef8fd5", BCR_colours[["MBL12"]][names(BCR_colours[["MBL12"]]) != "102"]), na.value = "lightgrey")+
    labs(title="BCR clones", x="UMAP1", y="UMAP2", colour="BCR clone")+
    theme_pub()+
    theme(aspect.ratio = 1,
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none")

saveRDS(MBL12_BCRclone_umap, file.path(RDS_dir, "MBL12_BCRclone_umap.RDS"))
ggsave(plot = MBL12_BCRclone_umap, 
       device = "pdf", 
       file=file.path(plotting_dir, "MBL12_BCRclone_umap.pdf"),
       width = 6, 
       height = 6)
```

### mtDNA mutation 2435G>A

```{r}
MBL12_2435_umap <- celltype_annotation_RNA_umap %>% 
  dplyr::left_join(heteroplasmy_intersect.cln.variants.MBL$MBL12["2435G>A",] %>% 
                     as_tibble(rownames = "label_cellbarcode") %>% 
                     dplyr::mutate(label_cellbarcode=str_replace(string = label_cellbarcode, pattern = "#", replacement = "_")) %>% 
                     dplyr::rename("2435G>A"=value)) %>% 
  ggplot2::ggplot(aes(x=UMAP_1, y=UMAP_2, colour=`2435G>A`))+
  geom_point_rast(size = 0.0001, raster.dpi = 600, dev = "cairo")+
  scale_color_gradientn(name="Heteroplasmy [%]",
    colours = c("yellow", "darkred"),
    trans = "log10",
    limits = c(1e-2, 1),
    breaks = c(1e-2, 1e-1, 1),
    labels = c("1", "10", "100"), 
    na.value = "lightgrey"
  ) +
  labs(title="2435G>A", x="UMAP1", y="UMAP2")+
  theme_pub()+
    theme(aspect.ratio = 1,
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "right"
          )
saveRDS(MBL12_2435_umap, file.path(RDS_dir, "MBL12_2435_umap.RDS"))
ggsave(plot = MBL12_2435_umap, 
       device = "pdf", 
       file=file.path(plotting_dir, "MBL12_2435_umap.pdf"),
       width = 6, 
       height = 6)
```


```{r fig.width=12, fig.height=12}
ggarrange(MBL12_celltype_umap, MBL12_CNV_umap, MBL12_BCRclone_umap, MBL12_2435_umap, 
          ncol = 2, 
          nrow=2, 
          align = "hv")
```