---
title: "Differential Chromatin Accessibility Analysis Full Cohort"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

# GOAL
The aim of this script is to find out if there are recurrent differentially accessible chromatin regions between MBL and CLL samples across patients. Also if there is differentially accessible chromatin between progressive and non-progressive MBLs.

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(ArchR)
library(readxl)
library(ggrepel)
library(ggallin)
library(tidyverse)
library(glue)
library(ggpubr)
library(SeuratDisk)
library(Seurat)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Cairo)
library(scales)
library(harmony)
library(chromVAR)
library(patchwork)
library(BiocParallel)
library(circlize)
library(ComplexHeatmap)
library(magick)
register(MulticoreParam(8))
```

```{r Run_parameters}
set.seed(1)
addArchRThreads(threads = 8)
addArchRGenome("hg38")
```

# Set paths

```{r }
# local
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "ATAC")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "DCA_analysis")
tsv_dir <- file.path(analysis_dir, "tsv_data") 
  
ArchR_project <- file.path("ArchR_35_samples/ArchR_all_samples")
ArchR_project_subset <- file.path("ArchR_35_samples/ArchR_all_samples_subsetted")
ArchR_project_LC_subset <- file.path("ArchR_35_samples/ArchR_LC_B")
ArchR_project_Blike <- file.path("ArchR_35_samples/ArchR_Blike")

ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")
ArchR_dir <- file.path(project_dir_ext, ArchR_project)
ArchR_dir_subset <- file.path(project_dir_ext, ArchR_project_subset)
ArchR_dir_LC_subset <- file.path(project_dir_ext, ArchR_project_LC_subset)
ArchR_dir_Blike <- file.path(result_dir, ArchR_project_Blike)
  
plotting_dir <- file.path(ArchR_dir, "Plots")
plotting_dir_subset <- file.path(ArchR_dir_subset, "Plots")
plotting_dir_LC_subset <- file.path(ArchR_dir_LC_subset, "Plots")
plotting_dir_Blike <- file.path(ArchR_dir_Blike, "Plots")

RDS_dir <- file.path(ArchR_dir, "RDS_objects")
RDS_dir_subset <- file.path(ArchR_dir_subset, "RDS_objects")
RDS_dir_LC_subset <- file.path(ArchR_dir_LC_subset, "RDS_objects")
RDS_dir_Blike <- file.path(ArchR_dir_Blike, "RDS_objects")
```

# Load data
```{r}
# local
source(file.path(project_dir, "scripts/publication_theme.R"))
source(file.path(project_dir, "scripts/heatmap_plotting_functions.R"))
```

```{r }
metadata <- read_tsv(file = file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata.tsv"))

patient_id2plotting_id <- metadata %>%
  dplyr::select(patient_id, plotting_id) %>%
  unique() %>%
  deframe()
```

## Reference data set Hao and Hao et al 2021

```{r Load_RNA_reference}
if (!file.exists("scRNA-PBMcs_Hao_and_Hao_et_2021.h5seurat")) {
  dir.create(file.path(data_dir, "reference_genomes"))
  download.file(
    url = "https://atlas.fredhutch.org/data/nygc/multimodal/pbmc_multimodal.h5seurat",
    destfile = file.path(data_dir, "reference_genomes/Hao_and_Hao_et_al_2021_PBMC/pbmc_multimodal.h5seurat")
  )
}

Hao_seRNA <- LoadH5Seurat(file.path(data_dir, "reference_genomes/Hao_and_Hao_et_al_2021_PBMC/pbmc_multimodal.h5seurat"))
Hao_seRNA
```

```{r Create_RNA_reference_assay, fig.width=2, fig.height=2}
# As this dataset does not have an RNA assay, the non-SCT values need to be reconstructed to an RNA assay for the reference
rna_counts <- CreateAssayObject(Hao_seRNA@assays[["SCT"]]@counts)

NormalizeData(
  rna_counts,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  margin = 1,
  verbose = TRUE
)

Hao_seRNA@assays[["RNA"]] <- rna_counts
```

## ArrowFiles

```{r Load_all_samples}
# Run this only as chuck as Arrow files are otherwise not in correct folder
setwd(file.path(project_dir_ext, "ArchR_35_samples"))

# Get all libraries files and paths
# Subset1
sample_subset <- c(
  "CLL1.1",
  "CLL1.2",
  "CLL2.1",
  "CLL2.2",
  "CLL3.1",
  "CLL3.2",
  "CLL4.1",
  "CLL4.2",
  "CLL5.1",
  "CLL5.2",
  "CLL6.1",
  "CLL6.2",
  "CLL7.1",
  "CLL7.2",
  "CLL8.1",
  "CLL8.2",
  "CLL9.1",
  "CLL9.2",
  "CLL10.1",
  "CLL10.2",
  "MBL1",
  "MBL2",
  "MBL3",
  "MBL4",
  "MBL6",
  "MBL7",
  "MBL7dup",
  "MBL8",
  "MBL9",
  "MBL10",
  "MBL11",
  "MBL12",
  "MBL13",
  "MBL14",
  "MBL15"
)

# Get all libraries and samples names
sample_names <- metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, pool) %>%
  dplyr::filter(label %in% sample_subset) %>%
  dplyr::arrange(pool) %>%
  dplyr::pull(var = label, name = pool)

# Select library_file_paths based on sample_names
library_file_paths <- purrr::map(names(sample_names), ~ file.path(library_path, .x, "fragments.tsv.gz")) %>%
  unlist()

ArrowFiles <- createArrowFiles(
  outputNames = sample_names,
  inputFiles = library_file_paths,
  sampleNames = sample_names,
  minTSS = 4,
  minFrags = 1000,
  addTileMat = TRUE,
  addGeneScoreMat = TRUE,
  threads = 1,
  subThreading = FALSE
)

ArrowFiles <- paste0(project_dir_ext, "ArchR_35_samples/", sample_names, ".arrow")
```

## Annotation data from pooled subsets

```{r}
Subset1 <- readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/Save-ArchR-Project.rds"))
Subset2 <- readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/Save-ArchR-Project.rds"))
Subset3 <- readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/Save-ArchR-Project.rds"))
Subset4 <- readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset4/Save-ArchR-Project.rds"))
```

# Doublet Inference

```{r Doublet_Inference}
setwd(paste0(project_dir_ext, "ArchR_35_samples/"))

doubScores <- addDoubletScores(
  input = ArrowFiles,
  k = 10, # Refers to how many cells near a "pseudo-doublet" to count.
  knnMethod = "UMAP", # Refers to the embedding to use for nearest neighbor search with doublet projection.
  LSIMethod = 1
)
# dir.create(paste0(ArchR_dir, "RDS_objects"))
saveRDS(doubScores, file = file.path(RDS_dir, "doubScores.RDS"))
```

# Create ArchR Project

```{r Make_ArchRProject}
setwd(file.path(project_dir_ext, "ArchR_35_samples"))

CLLProj <- ArchRProject(
  ArrowFiles = ArrowFiles,
  outputDirectory = "ArchR_all_samples",
  copyArrows = TRUE,
  threads = 1
)

saveArchRProject(
  ArchRProj = CLLProj,
  outputDirectory = ArchR_dir,
  load = FALSE,
  overwrite = TRUE
)

CLLProj <- loadArchRProject(path = ArchR_dir)
```

## Remove doublets

```{r remove_doublets}
setwd(file.path(project_dir_ext, "ArchR_35_samples"))

CLLProj <- filterDoublets(ArchRProj = CLLProj)

saveArchRProject(
  ArchRProj = CLLProj,
  outputDirectory = ArchR_dir,
  load = FALSE,
  overwrite = TRUE
)
```

## LSI Dimensionality Reduction

```{r LSI_Dimensionality_Reduction}
setwd(ArchR_dir)
CLLProj <- addIterativeLSI(
  seed = 1,
  ArchRProj = CLLProj,
  useMatrix = "TileMatrix",
  name = "IterativeLSI",
  iterations = 2,
  clusterParams = list( # See Seurat::FindClusters
    resolution = c(0.2),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30, 
  threads = 1, 
  force = TRUE
)

CLLProj <- addIterativeLSI(
  seed = 1,
  ArchRProj = CLLProj,
  useMatrix = "TileMatrix",
  name = "IterativeLSI_0.4",
  iterations = 2,
  clusterParams = list( # See Seurat::FindClusters
    resolution = c(0.4),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30, threads = 1
)

saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_dir, load = FALSE, overwrite = TRUE)
```

## Seurat Clustering

```{r}
setwd(file.path(project_dir, "results/ArchR_35_samples/"))

# Try combinations between different LSI resolutions and clustering resolutions
param_combinations <- tidyr::expand_grid(
  dimred = c(
    "IterativeLSI", "IterativeLSI_0.4"
  ),
  resolution = c(0.2, 0.3, 0.4, 0.6)
)

for (row in 1:nrow(param_combinations)) {
  resolution <- param_combinations$resolution[[row]]
  dimred <- param_combinations$dimred[[row]]

  CLLProj <- addClusters(
    input = CLLProj,
    seed = 1,
    reducedDims = dimred,
    method = "Seurat",
    name = glue("Clusters_{resolution}_{dimred}"),
    resolution = resolution
  )
}
saveArchRProject(
  ArchRProj = CLLProj,
  outputDirectory = ArchR_dir,
  load = FALSE,
  overwrite = TRUE
)
```

# UMAP

```{r UMAP_Embedding}
setwd(file.path(project_dir, "results/ArchR_35_samples"))

dimred_df <- tidyr::expand_grid(
  dimred = c(
    "IterativeLSI",
    "IterativeLSI_0.4"))

for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  CLLProj <- addUMAP(
    seed = 1,
    ArchRProj = CLLProj,
    reducedDims = dimred,
    name = glue("UMAP_{dimred}"),
    nNeighbors = 30,
    minDist = 0.5,
    metric = "cosine"
  )
}

saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_dir, load = FALSE, overwrite = TRUE)
```

```{r Plot_UMAP_Sample, fig.width=30, fig.height=10}
setwd(file.path(project_dir_ext, "ArchR_35_samples"))
umap_samples <- list()
for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  umaps <- plotEmbedding(
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = "Sample",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points",
    pal = c(my_paired_colours_named, my_unpaired_colours_named)
  ) +
    labs(x = "UMAP1", y = "UMAP2") +
    geom_point(size = 1, stroke=NA) +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    ) +
    guides(colour = guide_legend(
      keywidth = 0.5,
      keyheight = 0.5,
      override.aes = list(size = 3)
    ))

  umap_samples[[row]] <- umaps
}
```

```{r}
patientid_annotation <- tibble(
  "label_cellbarcode" = CLLProj$cellNames,
  "patient_id" = CLLProj$cellNames %>% c() %>%
    str_split("#", simplify = TRUE) %>% .[, 1] %>%
    str_split("\\.", simplify = TRUE) %>% .[, 1]
)

CLLProj <- addCellColData(
  ArchRProj = CLLProj,
  data = patientid_annotation %>% pull(patient_id),
  cells = patientid_annotation %>% pull(label_cellbarcode),
  name = "patient_id"
)

umap_patients <- plotEmbedding(
  ArchRProj = CLLProj,
  colorBy = "cellColData",
  name = "patient_id",
  embedding = "UMAP_IterativeLSI",
  plotAs = "points",
  pal = my_unpaired_colours_named
) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
```

```{r , fig.width=5, fig.height=5}
umap_coordinates <- CLLProj@embeddings@listData[["UMAP_IterativeLSI"]]@listData[["df"]] %>% as_tibble(rownames = "label_cellbarcode")

patient_annotation <- CLLProj@cellColData %>%
  as_tibble(rownames = "label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, patient_id)

umap_patient_annotation <- patient_annotation %>%
  dplyr::left_join(umap_coordinates)

patient_annotation_umap <- umap_patient_annotation %>%
  dplyr::filter(patient_id != "MBL7dup") %>%
  ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = patient_id)) +
  geom_point(size = 0.5, stroke = NA) +
  coord_equal() +
  scale_colour_manual(name = "Patients", values = c(my_unpaired_colours_named)) +
  labs(x = "UMAP1", y = "UMAP2", title = "Patients") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal",
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(
    keywidth = 0.5,
    keyheight = 0.5,
    override.aes = list(size = 3)
  ))
patient_annotation_umap
```

# Add cell type annotation

```{r}
CLLProj <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj,
  useMatrix = "GeneScoreMatrix",
  matrixName = "GeneIntegrationMatrix_IterativeLSI",
  reducedDims = "IterativeLSI",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "predictedCell_Un_Hao_IterativeLSI_,",
  nameGroup = "predictedGroup_Un_Hao_IterativeLSI",
  nameScore = "predictedScore_Un_Hao_IterativeLSI",
  threads = 1,
  force = TRUE
)

CLLProj <- addGeneIntegrationMatrix(
  ArchRProj = CLLProj,
  useMatrix = "GeneScoreMatrix",
  matrixName = "GeneIntegrationMatrix_IterativeLSI_0.4",
  reducedDims = "IterativeLSI_0.4",
  seRNA = Hao_seRNA,
  addToArrow = TRUE,
  groupRNA = "celltype.l1",
  nameCell = "predictedCell_Un_Hao_IterativeLSI_0.4",
  nameGroup = "predictedGroup_Un_Hao_IterativeLSI_0.4",
  nameScore = "predictedScore_Un_Hao_IterativeLSI_0.4",
  threads = 1,
  force = TRUE
)
```

```{r, fig.width=10, fig.height=10}
umap_annotation_no_subsets <- list()

for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  annotation_plots <- plotEmbedding(rastr=TRUE,
    randomize = TRUE,
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = "predictedGroup_Un_Hao_IterativeLSI",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points",
    pal = annotation_colours_named
  ) +
    geom_point(size = 0.01, stroke=NA) +
    labs(x = "UMAP1", y = "UMAP2") +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    ) +
    guides(colour = guide_legend(
      keywidth = 0.5,
      keyheight = 0.5,
      override.aes = list(size = 3)
    ))

  umap_annotation_no_subsets[[row]] <- annotation_plots
}

# Following plots are not ordered by celltype, currently T cells are hidden in this view! See next plot for ordering!
ggarrange(
  plotlist = umap_annotation_no_subsets,
  ncol =2,
  nrow = 1,
  common.legend = TRUE, 
  legend="bottom") %>%
  ggsave(bg = "white",
    plot = ., filename = file.path(plotting_dir, "Annotation_UMAPs_no_subsets.png"),
    width = 10,
    height = 5
  )
```

## Subset with manual annotation

```{r}
annotation_subset1 <- tibble("label_cellbarcode" = Subset1$cellNames, "My_annotation" = Subset1@cellColData$My_Annotation) %>%
  dplyr::mutate(My_annotation = if_else(str_starts(label_cellbarcode, "MBL3") & My_annotation == "CLL_B_Cells", "MBL_B_Cells", My_annotation))

annotation_subset2 <- tibble("label_cellbarcode" = Subset2$cellNames, "My_annotation" = Subset2@cellColData$My_Annotation)

annotation_subset3 <- tibble("label_cellbarcode" = Subset3$cellNames, "My_annotation" = Subset3@cellColData$My_Annotation)

annotation_subset4 <- tibble("label_cellbarcode" = Subset4$cellNames, "My_annotation" = Subset4@cellColData$My_Annotation)

My_annotation <- bind_rows(annotation_subset1, annotation_subset2, annotation_subset3, annotation_subset4) %>%
  dplyr::filter(label_cellbarcode %in% CLLProj$cellNames)


umap_annotation_My_annotation <- list()

for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  annotation_plots <- plotEmbedding(
    randomize = TRUE,
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = "My_annotation",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points",
    pal = annotation_colours_named
  ) +
    geom_point(size = 0.01, stroke="NA") +
    labs(x = "UMAP1", y = "UMAP2") +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    ) +
    guides(colour = guide_legend(
      keywidth = 0.5,
      keyheight = 0.5,
      override.aes = list(size = 3)
    ))

  umap_annotation_My_annotation[[row]] <- annotation_plots
}
```

```{r}
CLLProj <- addCellColData(
  ArchRProj = CLLProj,
  data = My_annotation[["My_annotation"]],
  cells = My_annotation[["label_cellbarcode"]],
  name = "My_annotation",
)

# drop cells which are not annotated as probably marked as doublets in subset analysis due to doublet simulation
cellindex <- BiocGenerics::which(!is.na(CLLProj@cellColData$My_annotation))
celllabel <- CLLProj$cellNames[cellindex]

CLLProj <- subsetArchRProject(
  ArchRProj = CLLProj,
  cells = celllabel,
  outputDirectory = ArchR_dir_subset,
  dropCells = TRUE,
  force = TRUE
)

My_annotation <- tibble("label_cellbarcode"=CLLProj$cellNames, "My_annotation"=CLLProj$My_annotation)

write_tsv(My_annotation, file.path(result_dir, "CLlProj_35_samples_subsetted_My_annotation.tsv"))

saveArchRProject(ArchRProj = CLLProj, 
                 outputDirectory = ArchR_dir_subset, 
                 load = FALSE, 
                 overwrite = TRUE)

CLLProj <- loadArchRProject(ArchR_dir_subset)
```

## More LSI iterations
```{r}
setwd(ArchR_dir_subset)
CLLProj <- addIterativeLSI(
  seed = 1,
  ArchRProj = CLLProj,
  useMatrix = "TileMatrix",
  name = "IterativeLSI10",
  iterations = 10,
  clusterParams = list( # See Seurat::FindClusters
    resolution = c(0.2),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30, 
  threads = 1, 
  force = TRUE
)
```

## Batch correction

```{r}
CLLProj <- loadArchRProject(path = ArchR_dir_subset)

CLLProj <- addHarmony(
  ArchRProj = CLLProj,
  reducedDims = "IterativeLSI",
  name = "Harmony_IterativeLSI",
  groupBy = "Sample", 
  verbose=TRUE
)

CLLProj <- addHarmony(
  ArchRProj = CLLProj,
  reducedDims = "IterativeLSI10",
  name = "Harmony_IterativeLSI10",
  groupBy = "Sample", 
  verbose=TRUE
)

saveArchRProject(ArchRProj = CLLProj, outputDirectory = ArchR_dir_subset, load = FALSE, overwrite = TRUE)
```

```{r, fig.width=60, fig.height=80}
param_combinations <- tidyr::expand_grid(
  dimred = c(
    "IterativeLSI", "IterativeLSI10", "Harmony_IterativeLSI", "Harmony_IterativeLSI10"),
  resolution = c(0.2, 0.3, 0.4, 0.6)
)

dimred_df <- tidyr::expand_grid(
  dimred = c(
    "IterativeLSI",
    "IterativeLSI10",
    "Harmony_IterativeLSI",
    "Harmony_IterativeLSI10"))

umap_clusters <- list()

for (row in 1:nrow(param_combinations)) {
  resolution <- param_combinations$resolution[[row]]
  dimred <- param_combinations$dimred[[row]]
  print(resolution)
  print(dimred)

  cluster_plots <- plotEmbedding(
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = glue("Clusters_{resolution}_{dimred}"),
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points"
  ) +
    geom_point(size = 0.5) +
    labs(x = "UMAP1", y = "UMAP2") +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    )

  umap_clusters[[row]] <- cluster_plots
}

ggarrange(plotlist = umap_clusters, ncol = 4, nrow = 4) %>%
  ggsave(
    plot = ., filename = file.path(plotting_dir_subset, "Cluster_UMAPs.png"),
    width = 15,
    height = 15, bg="white"  )
```

```{r}
umap_samples <- list()
for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  umaps <- plotEmbedding(rastr = TRUE,
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = "Sample",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points",
    pal = c(my_paired_colours_named, my_unpaired_colours_named)
  ) +
    labs(x = "UMAP1", y = "UMAP2") +
    geom_point(size = 0.01, stroke=NA) +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    ) +
    guides(colour = guide_legend(
      keywidth = 0.5,
      keyheight = 0.5,
      override.aes = list(size = 3)
    ))

  umap_samples[[row]] <- umaps
}

ggarrange(plotlist = umap_samples, ncol = 4,common.legend = TRUE, legend = "bottom") %>%
  ggsave(
    plot = ., filename = file.path(plotting_dir_subset, "Sample_UMAPs.png"),
    width =15,
    height = 5, 
    bg = "white"
  )
```

```{r Plot_technology, fig.width=30, fig.height=10}
label2technology <- deframe(metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, technology))
label2technology[CLLProj$Sample]
CLLProj@cellColData$technology <- unname(label2technology[CLLProj$Sample])

umap_technology <- list()
for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  umaps <- plotEmbedding(
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = "technology",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points"
  ) +    geom_point(size = 0.01, stroke=NA) +
    labs(x = "UMAP1", y = "UMAP2") +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    )

  umap_technology[[row]] <- umaps
}

ggarrange(plotlist = umap_technology, ncol = 4, common.legend = TRUE, legend = "bottom") %>%
  ggsave(
    plot = ., filename = file.path(plotting_dir_subset, "Technology_UMAPs.png"),
    width = 15, 
    bg="white",
    height = 5)
```

```{r}
seqdate_annotation <- tibble("label_cellbarcode" = CLLProj$cellNames) %>%
    tidyr::separate(label_cellbarcode, into=c("label", "cellbarcode"), sep="#", remove = F) %>% 
    dplyr::left_join(metadata %>% dplyr::filter(data_type =="ATAC") %>% dplyr::select(label, sequencing_date)) %>% 
    dplyr::mutate(sequencing_date=as.numeric(as.factor(sequencing_date))) %>% 
    dplyr::select(label_cellbarcode, sequencing_date) %>%
    deframe()

CLLProj <- addCellColData(
    ArchRProj = CLLProj,
    data = seqdate_annotation,
    cells = names(seqdate_annotation),
    name = "sequencing_date"
)

umap_seqdate <- list()
for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  umaps <- plotEmbedding(rastr = TRUE,
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = "sequencing_date",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points" ) +
    labs(x = "UMAP1", y = "UMAP2") +
    geom_point(size = 0.01, stroke=NA) +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    ) +
    guides(colour = guide_legend(
      keywidth = 0.5,
      keyheight = 0.5,
      override.aes = list(size = 3)
    ))

  umap_seqdate[[row]] <- umaps
}

ggarrange(plotlist = umap_seqdate, ncol = 4,common.legend = TRUE, legend = "bottom") %>%
  ggsave(
    plot = ., filename = file.path(plotting_dir_subset, "Sequencingdate_UMAPs.png"),
    width =15,
    height = 5, 
    bg = "white"
  )
```

```{r}
umap_celltypes <- list()
for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  umaps <- plotEmbedding(rastr = TRUE,
    ArchRProj = CLLProj,
    colorBy = "cellColData",
    name = "My_annotation",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points", randomize = TRUE,
    pal=annotation_colours_named) +
    labs(x = "UMAP1", y = "UMAP2") +
    geom_point(size = 0.01, stroke=NA) +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    ) +
    guides(colour = guide_legend(
      keywidth = 0.5,
      keyheight = 0.5,
      override.aes = list(size = 3)
    ))

  umap_celltypes[[row]] <- umaps
}

ggarrange(plotlist = umap_celltypes, ncol = 4,common.legend = TRUE, legend = "bottom") %>%
  ggsave(
    plot = ., filename = file.path(plotting_dir_subset, "Celltypes_UMAPs.png"),
    width =15,
    height = 5, 
    bg = "white"
  )
```

```{r}
umap_coordinates <- CLLProj@embeddings@listData[["UMAP_IterativeLSI"]]@listData[["df"]] %>%
  as_tibble(rownames = "label_cellbarcode")

patient_annotation <- CLLProj@cellColData %>%
  as_tibble(rownames = "label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, patient_id)

umap_patient_annotation <- patient_annotation %>%
  dplyr::left_join(umap_coordinates)

patient_annotation_umap <- umap_patient_annotation %>%
  dplyr::filter(!(patient_id %in% c("MBL7dup", "CLL6"))) %>%
  ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = patient_id)) +
  geom_point(size = 0.5, stroke = NA) +
  coord_equal() +
  scale_colour_manual(name = "Patients", values = c(my_unpaired_colours_named)) +
  labs(x = "UMAP1", y = "UMAP2", title = "Patients") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal",
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(
    keywidth = 0.5,
    keyheight = 0.5,
    override.aes = list(size = 3)
  ))
patient_annotation_umap

saveRDS(patient_annotation_umap, file.path(project_dir_subset, "RDS_Objects/UMAP_patients.RDS"))
```

```{r}
#shuffled annotation umap
umap_celltype_annotation <- umap_celltype_annotation %>% 
  dplyr::mutate(My_annotation=if_else(str_starts(label_cellbarcode, "MBL10|MBL11|MBL12|MBL13|MBL14|MBL15")& My_annotation=="MBL_B_Cells", "LC_MBL_B_Cells", My_annotation))
  
# Create a minimal expression matrix (dummy data)
expression_matrix <- matrix(data = rnorm(168533), nrow = 1, ncol = 168533)
rownames(expression_matrix) <- paste0("Gene1")
colnames(expression_matrix) <- umap_celltype_annotation$label_cellbarcode

# Create Seurat object
seurat_obj <- CreateSeuratObject(counts = expression_matrix)

# Set the UMAP coordinates as embeddings
umap_matrix <- umap_celltype_annotation %>% 
  ungroup() %>% 
  dplyr::select(`IterativeLSI#UMAP_Dimension_2`,`IterativeLSI#UMAP_Dimension_1`) %>% 
  as.matrix()

rownames(umap_matrix) <- umap_celltype_annotation$label_cellbarcode
seurat_obj[['umap']] <- CreateDimReducObject(embeddings = umap_matrix, key = "UMAP_", assay = DefaultAssay(seurat_obj))

# Merge annotation_tibble with the Seurat object metadata
seurat_obj <- AddMetaData(seurat_obj, metadata = umap_celltype_annotation$My_annotation, col.name = "My_annotation")

#plot
celltype_shuffled_annotation_umap <- DimPlot(seurat_obj, reduction = "umap", group.by = "My_annotation", label = F, shuffle = TRUE, raster = F, seed = 1)+
  scale_fill_manual(values = my_MBL_CLL_colours_named)+
  coord_equal()+
  scale_colour_manual(
    name = "Cell types",
    values = annotation_colours_named,
    labels = c("B" = "B cells", "DC" = "Dendritic cells", 
               "Mono" = "Monocytes", "other" = "Other cell types", 
               "CLL_B_Cells" = "CLL cells", "MBL_B_Cells" = "HC-MBL cells",
               "LC-LC_MBL_B_Cells"="LC-MBL cells",
               "NK" = "Natural killer cells", 
               "T" = "T cells")
  ) +
theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal",
    aspect.ratio = 1
  ) +
  labs(x = "UMAP1", y = "UMAP2", title = "Cell types") +
  guides(colour = guide_legend(
    keywidth = 0.5,
    keyheight = 0.5,
    override.aes = list(size = 3)
  ))

celltype_shuffled_annotation_umap

saveRDS(celltype_shuffled_annotation_umap, file.path(analysis_dir, "RDS_objects/Annotation_UMAP.RDS"))

saveRDS(seurat_obj, file.path(analysis_dir, "RDS_objects/Annotation_UMAP_seurat_obj.RDS"))
```

### Total cell count

```{r}
CLLProj_cell_counts_ATAC <- sum(!(CLLProj@cellColData@listData[["patient_id"]] %in% c("MBL7dup", "CLL6")))

CLLProj_cell_counts_ATAC
```


### IGHV umap
```{r}
IGHV_status <- tibble(
  "label_cellbarcode" = CLLProj@cellColData@rownames,
  "patient_id" = CLLProj@cellColData$patient_id
) %>%
  tidyr::separate(label_cellbarcode, c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type=="ATAC") %>% 
    dplyr::select(IGHV, patient_id)) %>% 
  dplyr::select(label_cellbarcode, IGHV)

IGHV_umap <- umap_celltype_annotation %>% left_join(IGHV_status) %>% 
  ggplot(aes(x = `IterativeLSI#UMAP_Dimension_1`, y = `IterativeLSI#UMAP_Dimension_2`, colour = IGHV)) +
  geom_point(size = 0.5, stroke = NA) +
  coord_equal() +
  scale_colour_manual(name = "IGHV mutational status", values = IGHV_colours, na.value = "lightgrey") +
  labs(x = "UMAP1", y = "UMAP2", title="IGHV\nmutational status") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal",
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(
    keywidth = 0.5,
    keyheight = 0.5,
    override.aes = list(size = 3)
  ))

write_tsv(IGHV_umap$data, file.path(tsv_dir, "ATAC_IGHV_umap_data.tsv"))
```

### Celltype abundance

```{r}
condition <- tibble("label_cellbarcode" = CLLProj@cellColData@rownames) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  mutate(disease = case_when(
    str_ends(label, "\\.1") ~ "MBL",
    # str_starts(label, "MBL") ~ "MBL",
    str_starts(label, paste0(c("MBL4"), collapse = "|")) ~ "CLL",
    str_starts(label, paste0(paste0("MBL", c(1:3, 6:15)), collapse = "|")) ~ "MBL",
    str_ends(label, "\\.2") ~ "CLL",
    TRUE ~ label
  ))

MBL_CLL_subtype <- tibble(
  "label_cellbarcode" = CLLProj@cellColData@rownames,
  "patient_id" = CLLProj@cellColData$patient_id
) %>%
  tidyr::separate(label_cellbarcode, c("label", NA), sep = "#", remove = FALSE) %>%
  dplyr::left_join(metadata %>%
    dplyr::select(MBL_subtype, disease, label) %>%
    tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE) %>%
    dplyr::select(label, MBL_CLL_subtype) %>% unique())


celltype_annotation <- tibble(
  "label_cellbarcode" = CLLProj$cellNames,
  "celltype_annotation" = CLLProj@cellColData$My_annotation
)

celltype_data <- left_join(celltype_annotation, MBL_CLL_subtype) %>%
  group_by(MBL_CLL_subtype, celltype_annotation) %>%
  dplyr::summarise("cell_counts" = n())

# calculate celltype abundance based on median across all samples
celltype_abundance_plot <- left_join(celltype_annotation, MBL_CLL_subtype) %>%
  tidyr::separate(label_cellbarcode, c("label", NA), sep = "#") %>%
  tidyr::separate(label, c("patient_id", NA), sep = "\\.") %>%
  group_by(MBL_CLL_subtype, celltype_annotation, patient_id) %>%
  dplyr::filter(patient_id != "CLL6") %>%
  dplyr::summarise(counts = n()) %>%
  ungroup() %>%
  group_by(MBL_CLL_subtype, celltype_annotation) %>%
  dplyr::summarise(
    sd_celltype_abundance = sd(counts, na.rm = TRUE),
    median_abundance = median(counts)
  ) %>%
  mutate(MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("LC_MBL", "HC_MBL", "CLL"))) %>%
  ggplot(., aes(fill = celltype_annotation, y = median_abundance, x = MBL_CLL_subtype)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(
    values = annotation_colours_named,
    name = "Cell type",
    labels = c("B" = "B cells", "DC" = "Dendritic cells", "Mono" = "Monocytes", "other" = "Other cell types", "CLL_B_Cells" = "CLL cells", "MBL_B_Cells" = "MBL cells", "NK" = "Natural killer cells", "T" = "T cells")
  ) +
  scale_x_discrete(labels = c("LC_MBL" = "LC MBL", "HC_MBL" = "HC MBL", "CLL" = "CLL")) +
  scale_y_continuous(labels = seq(0, 100, 25), expand = expansion(mult = c(0, NA))) +
  labs(x = "", y = "Cell type frequency [%]") +
  theme_pub()


saveRDS(celltype_abundance_plot, file = file.path(analysis_dir, "RDS_objects/celltype_abundance_plot.RDS"))

celltype_abundance_plot
```

```{r Celltype_quantification, fig.width=20, fig.height=3}
celltype_abundance_per_patient <- as_tibble(CLLProj@cellColData, rownames = NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, My_annotation, patient_id) %>%
  dplyr::left_join(condition) %>%
  dplyr::left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select(patient_id, plotting_id) %>%
    unique()) %>%
  group_by(disease, My_annotation, plotting_id) %>%
  dplyr::summarise(cell_counts = n()) %>%
  dplyr::mutate(
    disease = factor(disease, levels = c("MBL", "CLL")),
    plotting_id = factor(plotting_id, levels = c(paste0("P", 1:24), "P9dup"))
  ) %>%
  ungroup() %>%
  ggplot(., aes(fill = My_annotation, y = cell_counts, x = disease)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(
    breaks = seq(0, 1, 0.25),
    labels = c("0", "25", "50", "75", "100"),
    expand = expansion(mult = c(0, NA))
  ) +
  scale_fill_manual(
    values = annotation_colours_named,
    name = "Cell type"
  ) +
  labs(x = "", y = "Frequency [%]") +
  theme_pub() +
  theme(legend.position = "bottom") +
  facet_wrap(~plotting_id, nrow = 1, ncol = 26)

celltype_abundance_per_patient
```

# Combination cell type counts per sample

```{r fig.width=12, fig.height=5}
label2plotting_id <- metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, plotting_id) %>%
  deframe()

all_b_cell_plot <- CLLProj@cellColData %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, Sample, My_annotation) %>%
  group_by(Sample, My_annotation) %>%
  dplyr::filter((!Sample %in% c("MBL7dup", "CLL6.1", "CLL6.2"))) %>%
  dplyr::summarise(celltype_counts = n()) %>%
  dplyr::filter(My_annotation %in% c("B", "MBL_B_Cells", "CLL_B_Cells")) %>%
  pivot_wider(id_cols = Sample, names_from = My_annotation, values_from = celltype_counts, values_fill = NA) %>%
  tidyr::pivot_longer(-Sample, names_to = "celltype", values_to = "counts") %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "ATAC") %>% dplyr::select(label, plotting_id, disease, MBL_subtype), by = join_by(Sample == label)) %>%
  dplyr::arrange(desc(plotting_id)) %>%
  tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE) %>%
  dplyr::mutate(
    MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("LC_MBL", "HC_MBL", "CLL")),
    celltype = case_when(celltype == "MBL_B_Cells" & MBL_CLL_subtype == "LC_MBL" ~ "LC_MBL_B_Cells",
      celltype == "MBL_B_Cells" & MBL_CLL_subtype == "HC_MBL" ~ "HC_MBL_B_Cells",
      .default = celltype
    )
  ) %>%
  dplyr::mutate(
    counts = if_else(celltype == "B", -counts, counts),
    celltype = factor(celltype, levels = c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")),
    Sample = factor(Sample, level = c(
       "MBL12", "MBL11", "MBL14", "MBL13","MBL15", "MBL10", "MBL6", "MBL1", "MBL7", "MBL3", "MBL8", "MBL2", "MBL9",
      "CLL6.1", "CLL6.2", "CLL1.1", "CLL1.2", "CLL7.1", "CLL7.2", "CLL2.1", "CLL2.2", "CLL5.1", "CLL5.2", "CLL8.1",
      "CLL8.2", "CLL4.1", "CLL4.2", "CLL10.1", "CLL10.2", "CLL9.1", "CLL9.2", "CLL3.1", "CLL3.2", "MBL4"
    ))
  ) %>%
  ggplot(aes(x = Sample, y = counts, fill = celltype)) +
  geom_bar(position = "stack", stat = "identity", width = 0.7) +
  scale_x_discrete(labels = label2plotting_id) +
  scale_y_continuous(limits = c(-1700, 10000)) +
  # scale_y_continuous(trans = pseudolog10_trans,
  #                    breaks = c(-1000, -100 ,-10,  10, 100, 1000, 10000),
  #                    labels=c("1000", "100", "10",  "10","100" ,"1000", "10000"),
  #                    limits=c(-1000, 13000)) +
  scale_fill_manual(
    name = "Cell type", values = annotation_colours_named,
    labels = c("B" = "B cells", "CLL_B_Cells" = "CLL cells", "LC_MBL_B_Cells" = "LC-MBL cells", "HC_MBL_B_Cells" = "HC-MBL cells")
  ) +
  labs(x = "", title = "Cell type cell counts") +
  theme_pub() +
  theme(
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 90),
    # axis.line.x = element_blank(),
    # axis.ticks.x = element_blank(),
    plot.title = element_text(margin = margin(t = 8, b = -20))
  )

all_b_cell_plot

saveRDS(all_b_cell_plot, file = file.path(analysis_dir, "RDS_objects/all_b_cell_plot.RDS"))
```

```{r fig.width=5, fig.height=12}
all_b_cell_plot_vertical <- CLLProj@cellColData %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, Sample, My_annotation) %>%
  dplyr::group_by(Sample, My_annotation) %>%
  dplyr::filter((!Sample %in% c("MBL7dup", "CLL6.1", "CLL6.2"))) %>%
  dplyr::summarise(celltype_counts = n()) %>%
  dplyr::filter(My_annotation %in% c("B", "MBL_B_Cells", "CLL_B_Cells")) %>%
  pivot_wider(id_cols = Sample, names_from = My_annotation, values_from = celltype_counts, values_fill = NA) %>%
  tidyr::pivot_longer(-Sample, names_to = "celltype", values_to = "counts") %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "ATAC") %>% 
                     dplyr::select(label, plotting_id, disease, MBL_subtype), by = join_by(Sample == label)) %>%
  dplyr::arrange(desc(plotting_id)) %>%
  tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE) %>%
  dplyr::mutate(
    MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("LC_MBL", "HC_MBL", "CLL")),
    celltype = case_when(celltype == "MBL_B_Cells" & MBL_CLL_subtype == "LC_MBL" ~ "LC_MBL_B_Cells",
      celltype == "MBL_B_Cells" & MBL_CLL_subtype == "HC_MBL" ~ "HC_MBL_B_Cells",
      .default = celltype
    )
  ) %>%
  dplyr::mutate(
    counts = if_else(celltype == "B", -counts, counts),
    celltype = factor(celltype, levels = c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")),
    Sample = factor(Sample, level = rev(c(
       "MBL12", "MBL11", "MBL14", "MBL13","MBL15", "MBL10", "MBL6", "MBL1", "MBL7", "MBL3", "MBL8", "MBL2", "MBL9",
      "CLL6.1", "CLL6.2", "CLL1.1", "CLL1.2", "CLL7.1", "CLL7.2", "CLL2.1", "CLL2.2", "CLL5.1", "CLL5.2", "CLL8.1",
      "CLL8.2", "CLL4.1", "CLL4.2", "CLL10.1", "CLL10.2", "CLL9.1", "CLL9.2", "CLL3.1", "CLL3.2", "MBL4"
    )))
  ) %>%
  ggplot(aes(y = Sample, x = counts, fill = celltype)) +
  geom_bar(position = "stack", stat = "identity", width = 0.7) +
  scale_y_discrete(labels = label2plotting_id) +
  scale_x_continuous(
    breaks = c(-1250, 0, 2500, 5000, 7500, 10000),
    labels = c("1250", "0", "2500", "5000", "7500", "10000"),
    limits = c(-1700, 10000)
  ) +
  scale_fill_manual(
    name = "Cell type", values = annotation_colours_named,
    labels = c("B" = "B cells", "CLL_B_Cells" = "CLL cells", "LC_MBL_B_Cells" = "LC-MBL cells", "HC_MBL_B_Cells" = "HC-MBL cells")
  ) +
  labs(x = "", title = "") +
  theme_pub() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(angle = 0),
    # axis.line.x = element_blank(),
    # axis.ticks.x = element_blank(),
    plot.title = element_text(margin = margin(t = 8, b = -20))
  )

all_b_cell_plot_vertical


saveRDS(all_b_cell_plot_vertical, file = file.path(analysis_dir, "RDS_objects/all_b_cell_plot_vertical.RDS"))
```

```{r}
Blike_celltype_percentages <- CLLProj@cellColData %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::select(label_cellbarcode, Sample, My_annotation) %>%
  dplyr::group_by(Sample, My_annotation) %>%
  dplyr::filter((!Sample %in% c("MBL7dup", "CLL6.1", "CLL6.2"))) %>%
  dplyr::summarise(celltype_counts = n()) %>%
  dplyr::filter(My_annotation %in% c("B", "MBL_B_Cells", "CLL_B_Cells")) %>%
  pivot_wider(id_cols = Sample, names_from = My_annotation, values_from = celltype_counts, values_fill = NA) %>%
  tidyr::pivot_longer(-Sample, names_to = "celltype", values_to = "counts") %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "ATAC") %>% 
                     dplyr::select(label, plotting_id, disease, MBL_subtype), by = join_by(Sample == label)) %>%
  dplyr::arrange(desc(plotting_id)) %>%
  tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE) %>%
  dplyr::mutate(
    MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("LC_MBL", "HC_MBL", "CLL")),
    celltype = case_when(celltype == "MBL_B_Cells" & MBL_CLL_subtype == "LC_MBL" ~ "LC_MBL_B_Cells",
      celltype == "MBL_B_Cells" & MBL_CLL_subtype == "HC_MBL" ~ "HC_MBL_B_Cells",
      .default = celltype
    )
  ) %>% dplyr::select(Sample, celltype, counts, MBL_CLL_subtype) %>% 
  dplyr::filter(!is.na(counts)) %>% 
  tidyr::pivot_wider(id_cols = c(Sample, MBL_CLL_subtype), names_from = celltype, values_from = counts) %>% 
  dplyr::mutate(Total = rowSums(across(B:CLL_B_Cells), na.rm = TRUE),
                B_perc = B / Total * 100,
                HC_MBL_perc = HC_MBL_B_Cells / Total * 100,
                LC_MBL_perc = LC_MBL_B_Cells / Total * 100,
                CLL_perc = CLL_B_Cells / Total * 100
              ) 

mean_B_percentage <- Blike_celltype_percentages %>% 
  group_by(MBL_CLL_subtype) %>% 
  dplyr::summarise(Mean_B_percentage=mean(B_perc, na.rm=TRUE)) %>% deframe()
mean_B_percentage
```

# Healthy B cells comparison to malignant B cells

```{r}
b_cell_abundance_plot <- left_join(celltype_annotation, MBL_CLL_subtype) %>%
  tidyr::separate(label_cellbarcode, c("label", NA), sep = "#") %>%
  tidyr::separate(label, c("patient_id", NA), sep = "\\.") %>%
  group_by(MBL_CLL_subtype, celltype_annotation, patient_id) %>%
  dplyr::summarise(counts = n()) %>%
  ungroup() %>%
  group_by(MBL_CLL_subtype, celltype_annotation) %>%
  dplyr::summarise(
    sd_celltype_abundance = sd(counts, na.rm = TRUE),
    median_abundance = median(counts)
  ) %>%
  dplyr::filter(celltype_annotation %in% c("B", "MBL_B_Cells", "CLL_B_Cells")) %>%
  dplyr::mutate(percent = median_abundance / sum(median_abundance)) %>%
  dplyr::filter(celltype_annotation == "B") %>%
  dplyr::mutate(MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("LC_MBL", "HC_MBL", "CLL"))) %>%
  ggplot(aes(x = MBL_CLL_subtype, y = percent, fill = celltype_annotation)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = c("LC_MBL" = "LC MBL", "HC_MBL" = "HC MBL", "CLL" = "CLL")) +
  scale_fill_manual(
    values = annotation_colours_named,
    name = "Cell type",
    labels = c("B cells")
  ) +
  scale_y_continuous(
    breaks = seq(0, 0.08, 0.02),
    labels = seq(0, 8, 2),
    expand = expansion(mult = c(0, NA)),
    limits = c(0, 0.08)
  ) +
  labs(x = "", y = "B cell frequency [%]") +
  theme_pub()

b_cell_abundance_plot

saveRDS(b_cell_abundance_plot, file = file.path(analysis_dir, "RDS_objects/b_cell_abundance_plot.RDS"))
```

### Add Imputation Gene Scores
```{r}
CLLProj <- addImputeWeights(CLLProj)
```

#### Plot Gene Scores
```{r}
markerGenes <- c(
  "PAX5", "MS4A1", "EBF1", "MME", # B-Cell Trajectory
  "CD14", "CEBPB", "MPO", # Monocytes
  "CD3D", "CD8A", "TBX21", "IL7R", # TCells
  "CD19", "CD5", "FCER2", "CD79b", # CLL B-Cell Marker
  "GNLY", "NKG7", "NCAM1" # NK cells
)

gene_score_plots <- plotEmbedding(
  ArchRProj = CLLProj,
  colorBy = "GeneScoreMatrix",
  name = markerGenes,
  embedding = "UMAP_IterativeLSI",
  imputeWeights = getImputeWeights(CLLProj),
  plotAs = "points"
)

gene_score_plots %>%
  imap(~.x[["data"]] %>%
         write_tsv(file = file.path(tsv_dir, paste0("gene_scores_", .y, ".tsv"))))
```

### Add pseudobulk replicates

```{r}
CLLProj <- addGroupCoverages(
  ArchRProj = CLLProj,
  groupBy = "My_annotation"
)
```

### Peak calling

```{r}
CLLProj <- addReproduciblePeakSet(
  ArchRProj = CLLProj,
  groupBy = "My_annotation",
  pathToMacs2 = pathToMacs3)

saveArchRProject(ArchRProj = CLLProj, 
                 outputDirectory = ArchR_dir_subset, 
                 load = FALSE, 
                 overwrite = TRUE)
```

```{r}
CLLProj <- addPeakMatrix(CLLProj)

saveArchRProject(ArchRProj = CLLProj, 
                 outputDirectory = ArchR_dir_subset, 
                 load = FALSE, 
                 overwrite = TRUE)

getAvailableMatrices(CLLProj)
```

### Add TF Mofif Annotations

```{r}
CLLProj <- addMotifAnnotations(
  ArchRProj = CLLProj,
  motifSet = "cisbp",
  annoName = "Motif"
)

saveArchRProject(ArchRProj = CLLProj, 
                 outputDirectory = ArchR_dir_subset, 
                 load = FALSE, 
                 overwrite = TRUE)
```

```{r}
# background peaks added
CLLProj <- addBgdPeaks(CLLProj, method = "chromVAR")
```

```{r}
# motif deviation matrix added
CLLProj <- addDeviationsMatrix(
  ArchRProj = CLLProj,
  bgdPeaks = getBgdPeaks(CLLProj,
    method = "chromVAR"),
  peakAnnotation = "Motif",
  threads = 16
)
```

#### Plot Motifs
```{r fig.width=3, fig.height=3}
# only the ones which actually are TFs of the following will be found
motifs <- c(
  "PAX5", "EBF1", # B-Cell Trajectory
  "CEBPB",  "CEBPA",  # Monocytes
  "TBX21", "RUNX1", # T-Cells
  "CD19", "CD5", "FCER2", "CD79b", # CLL B-Cell Marker
  "EOMES" # NK cells
)

markerMotifs <- getFeatures(CLLProj, 
                            select = paste(motifs, collapse="|"), 
                            useMatrix = "MotifMatrix")

# Get only z-scores and remove SREBF1_22
markerMotifs <- grep("z:", markerMotifs, value = TRUE)
markerMotifs <- markerMotifs[markerMotifs %ni% c("z:SREBF1_22", "z:ARID2_11", "z:BCL6B_218")]

motif_zscore_plots <- plotEmbedding(
    ArchRProj = CLLProj, 
    colorBy = "MotifMatrix", 
    name = markerMotifs, 
    embedding = "UMAP_IterativeLSI",
    imputeWeights = getImputeWeights(CLLProj))

motif_zscore_plots %>%
  imap(~.x[["layers"]][[1]][["data"]] %>%
         write_tsv(file = file.path(tsv_dir, paste0("motif_zscore_", str_split_i( .y, pattern = "z:", i = 2), ".tsv"))))
```

## Gene Tracks all celltypes
```{r}
B_like_annotation <- tibble(
  "label_cellbarcode" = rownames(CLLProj@cellColData),
  "label" = str_split_i(rownames(CLLProj@cellColData), pattern = "#", i = 1),
  "My_annotation"=CLLProj@cellColData$My_annotation) %>%
  dplyr::left_join(metadata %>%
    dplyr::select(MBL_subtype, disease, label) %>%
    tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE) %>%
    dplyr::select(label, MBL_CLL_subtype) %>% unique()) %>% 
  dplyr::mutate(B_like_annotation=case_when(My_annotation == "MBL_B_Cells" & MBL_CLL_subtype == "HC_MBL" ~ "HC_MBL_B_Cells",
                                            My_annotation == "MBL_B_Cells" & MBL_CLL_subtype == "LC_MBL" ~ "LC_MBL_B_Cells"))

CLLProj <- addCellColData(
  ArchRProj = CLLProj,
  data = B_like_annotation[["B_like_annotation"]],
  cells = B_like_annotation[["label_cellbarcode"]],
  name = "B_like_annotation", force=TRUE
)

Celltype_tracks <- list()
Celltype_tracks["CD5"]<- plotBrowserTrack(
  ArchRProj = CLLProj,
  groupBy = "B_like_annotation",
  geneSymbol = c("CD5"),
  upstream = 15000,
  downstream = 30000
)

Celltype_tracks["FCER2"]  <- plotBrowserTrack(
  ArchRProj = CLLProj,
  groupBy = "B_like_annotation",
  geneSymbol = c("FCER2"),
  upstream = 15000,
  downstream = 15000
)

Celltype_tracks["MME"]  <- plotBrowserTrack(
  ArchRProj = CLLProj,
  groupBy = "B_like_annotation",
  geneSymbol = c("MME"),
  upstream = 40000,
  downstream = 170000
)

for (track in 1:length(Celltype_tracks)) {
  grid::grid.newpage()
  grid::grid.draw(Celltype_tracks[[track]])
}

saveRDS(Celltype_tracks, file.path(RDS_dir_subset, "Celltype_tracks.RDS"))
```


# Subset to only LC-MBL and corresopnding healthy B cells

```{r}
MBL_CLL_subtype_annotation <- tibble(
  "label_cellbarcode" = rownames(CLLProj@cellColData),
  "label" = str_split_i(rownames(CLLProj@cellColData), pattern = "#", i = 1)
) %>%
  dplyr::left_join(metadata %>%
    dplyr::select(MBL_subtype, disease, label) %>%
    tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE) %>%
    dplyr::select(label, MBL_CLL_subtype) %>% unique())

CLLProj <- addCellColData(
  ArchRProj = CLLProj,
  data = MBL_CLL_subtype_annotation[["MBL_CLL_subtype"]],
  cells = MBL_CLL_subtype_annotation[["label_cellbarcode"]],
  name = "MBL_CLL_subtype"
)

idxSample_2 <- BiocGenerics::which(CLLProj@cellColData$My_annotation %in% c("MBL_B_Cells", "B") &
  CLLProj@cellColData$MBL_CLL_subtype %in% c("LC_MBL"))
cellsSample_2 <- CLLProj$cellNames[idxSample_2]

# CLLProj_LC_B <- subsetArchRProject(
#   ArchRProj = CLLProj,
#   cells = cellsSample_2,
#   outputDirectory = ArchR_dir_LC_subset,
#   dropCells = TRUE
# )

CLLProj_LC_B <- loadArchRProject(path = ArchR_dir_LC_subset)
```

### Add pseudobulk replicates

```{r}
CLLProj_LC_B <- addGroupCoverages(
  ArchRProj = CLLProj_LC_B,
  groupBy = "My_annotation"
)
```

 ### Peak calling

```{r}
CLLProj_LC_B <- addReproduciblePeakSet(
  ArchRProj = CLLProj_LC_B,
  groupBy = "My_annotation",
  pathToMacs2 = pathToMacs3
)

saveArchRProject(
  ArchRProj = CLLProj_LC_B,
  outputDirectory = ArchR_dir_LC_subset,
  load = FALSE,
  overwrite = TRUE
)
```

```{r}
CLLProj_LC_B <- addPeakMatrix(CLLProj_LC_B)

saveArchRProject(
  ArchRProj = CLLProj_LC_B,
  outputDirectory = ArchR_dir_LC_subset,
  load = FALSE,
  overwrite = TRUE
)

getAvailableMatrices(CLLProj_LC_B)
```

### Marker Peaks

```{r}
# Gets peaks which are unique to cell types
markersPeaks_LC_B <- getMarkerFeatures(
  ArchRProj = CLLProj_LC_B,
  useMatrix = "PeakMatrix",
  groupBy = "My_annotation",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon",
  threads = 1
)

markerList_LC_B <- getMarkers(markersPeaks_LC_B, cutOff = "FDR <= 0.1 & Log2FC >= 1")
```

### Volcano Plots MarkerPeaks

```{r, fig.width=10,fig.height=5}
# unique peaks for MBL or CLL cells compared to all other cell types
volcanoplot_MP_MBL_LC_B <- plotMarkers(
  seMarker = markersPeaks_LC_B,
  name = "MBL_B_Cells",
  cutOff = "FDR <= 0.1 & Log2FC >= 1",
  plotAs = "Volcano"
) +
  geom_point(size = 3) +
  scale_colour_manual(
    name = "DAR",
    values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#FC8D62", "Not-Differential" = "#D3D3D3"),
    labels = c("Down-Regulated" = "B", "Up-Regulated" = "LC-MBL", "Not-Differential" = "not significant")
  ) +
  theme_pub()

volcanoplot_MP_B <- plotMarkers(
  seMarker = markersPeaks_LC_B,
  name = "B",
  cutOff = "FDR <= 0.1 & Log2FC >= 1",
  plotAs = "Volcano"
) +
  geom_point(size = 3) +
  scale_colour_manual(
    name = "DAR",
    values = c("Down-Regulated" = "#FC8D62", "Up-Regulated" = "darkred", "Not-Differential" = "#D3D3D3"),
    labels = c("Down-Regulated" = "MBL", "Up-Regulated" = "B", "Not-Differential" = "not significant")
  ) +
  theme_pub()

ggarrange(volcanoplot_MP_MBL_LC_B, volcanoplot_MP_B, ncol = 2) %>%
  ggsave(
    plot = .,
    filename = file.path(plotting_dir_LC_subset, "Unique_Markers_LC_B_Volcanoplot.pdf"),
    width = 45,
    height = 22.5,
    unit = "cm"
  )
```

### Volcano plots Pairwise testing

```{r}
# unique peaks between MBL and CLL cells only!
markerTest_LC_B <- getMarkerFeatures(
  ArchRProj = CLLProj_LC_B,
  useMatrix = "PeakMatrix",
  groupBy = "My_annotation",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  useGroups = "MBL_B_Cells",
  bgdGroups = "B"
)

volcanoplot_PWT_B <- plotMarkers(
  seMarker = markerTest_LC_B,
  name = "MBL_B_Cells",
  cutOff = "FDR <= 0.1 & abs(Log2FC) >= 1",
  plotAs = "Volcano",
  rastr = TRUE
) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
  scale_y_continuous(
    name = "-Log10(FDR)",
    breaks = seq(0, 12.5, 2.5),
    limits = c(0, 12.5),
    expand = expansion(mult = c(0, 0.01)),
    oob = scales::oob_squish_infinite
  ) +
  scale_colour_manual(
    name = "DARs",
    values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#FC8D62", "Not-Differential" = "#D3D3D3"),
    labels = c("Down-Regulated" = "B", "Up-Regulated" = "MBL", "Not-Differential" = "Not significant")
  ) +
  theme_pub() +
  labs(title = "")

volcanoplot_PWT_B[["layers"]][[2]] <- NULL

volcanoplot_PWT_B

ggarrange(volcanoplot_PWT_B, ncol = 1) %>%
  ggsave(
    plot = .,
    filename = file.path(plotting_dir_LC_subset, "Upregulation_MBL_Volcanoplot.pdf"),
    width = 22.5,
    height = 22.5,
    unit = "cm"
  )

saveRDS(volcanoplot_PWT_B, file.path(RDS_dir_LC_subset, "volcanoplot_PWT_LC_B.RDS"))
```

### Add TF Mofif Annotations

```{r}
CLLProj_LC_B <- addMotifAnnotations(
  ArchRProj = CLLProj_LC_B,
  motifSet = "cisbp",
  annoName = "Motif"
)
```

```{r}
# background peaks added
CLLProj_LC_B <- addBgdPeaks(CLLProj_LC_B, method = "chromVAR")
```

```{r}
# motif deviation matrix added
CLLProj_LC_B <- addDeviationsMatrix(
  ArchRProj = CLLProj_LC_B,
  bgdPeaks = getBgdPeaks(CLLProj_LC_B,
    method = "chromVAR"
  ),
  peakAnnotation = "Motif",
  threads = 1
)
```

### Get Matrix from project

```{r}
MotifMatrix_tmp_LC_B <- getMatrixFromProject(
  ArchRProj = CLLProj_LC_B,
  useMatrix = "MotifMatrix",
  useSeqnames = NULL,
  verbose = TRUE,
  binarize = FALSE,
  threads = 1
)

zscore_matrix_LC_B <- as.matrix(assays(MotifMatrix_tmp_LC_B)$z) %>%
  as.data.frame()

deviations_matrix_LC_B <- as.matrix(assays(MotifMatrix_tmp_LC_B)$deviations) %>%
  as.data.frame()

VarDeviation_LC_B <- getVarDeviations(CLLProj_LC_B,
  name = "MotifMatrix",
  plot = FALSE
) %>%
  as.data.frame()

Top_TF_target_LC_B <- head(VarDeviation_LC_B$name, n = 100)

zscore_part_LC_B <- zscore_matrix_LC_B[Top_TF_target_LC_B, ]
```

```{r}
condition_LC_B <- tibble(
  "label_cellbarcode" = CLLProj_LC_B$cellNames,
  "My_annotation" = CLLProj_LC_B$My_annotation
)

MBL_cells_LC_B <- dplyr::filter(condition_LC_B, My_annotation == "MBL_B_Cells") %>%
  pull(label_cellbarcode)

B_cells_LC_B <- dplyr::filter(condition_LC_B, My_annotation == "B") %>%
  pull(label_cellbarcode)

# # calculate data for volcano plot
input_zscore_matrix_LC_B <- deviations_matrix_LC_B

# If no NA values are found in zcore matrix calculate statistics
condition1_LC_B <- B_cells_LC_B
condition2_LC_B <- MBL_cells_LC_B

volcano.data_LC_B <- data.frame()
i <- 0
for (TF in rownames(input_zscore_matrix_LC_B)) {
  i <- i + 1
  print(i)
  print(TF)
  if (length(which(is.na(input_zscore_matrix_LC_B[TF, ]))) == 0) {
    p.value <- wilcox.test(as.numeric(input_zscore_matrix_LC_B[TF, condition2_LC_B]), as.numeric(input_zscore_matrix_LC_B[TF, condition1_LC_B]))$p.value
    diff.accessibility <- median(as.numeric(input_zscore_matrix_LC_B[TF, condition2_LC_B])) - median(as.numeric(input_zscore_matrix_LC_B[TF, condition1_LC_B]))
    volcano.data_LC_B <- rbind(volcano.data_LC_B, data.frame(TF = TF, p.value = p.value, diff.accessibility = diff.accessibility))
  }
}

volcano.data_LC_B$p.adj <- p.adjust(volcano.data_LC_B$p.value)
volcano.data_LC_B$p.adj[which(volcano.data_LC_B$p.adj == 0)] <- min(volcano.data_LC_B$p.adj[which(volcano.data_LC_B$p.adj != 0)])

volcano.data_LC_B$diff.accessibility <- as.numeric(volcano.data_LC_B$diff.accessibility)
volcano.data_LC_B$p.adj <- as.numeric(volcano.data_LC_B$p.adj)


volcano.data_LC_B$significant <- ifelse(abs(volcano.data_LC_B$diff.accessibility) > 1.5 & volcano.data_LC_B$p.adj <= 0.05, "significant", "not significant")
volcano.data_LC_B <- volcano.data_LC_B %>%
  dplyr::mutate(
    "significant" = case_when(
      p.adj >= 0.05 ~ "not significant",
      abs(diff.accessibility) < 1.5 ~ "not significant",
      diff.accessibility > 1.5 & p.adj < 0.05 ~ "MBL",
      diff.accessibility < -1.5 & p.adj < 0.05 ~ "B"
    ),
    "label" = if_else(significant != "not significant", stringr::str_split_fixed(TF, pattern = "_", n = 2)[, 1], NA)
  )


saveRDS(volcano.data_LC_B, file = file.path(RDS_dir_LC_subset, "volcano.data_LC_B.RDS"))
```

### Plot Differential TF enrichment

```{r}
volcano.plot_LC_B <- volcano.data_LC_B %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 2),
    limits = c(-6, 6)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 350),
    expand = expansion(mult = c(0, NA)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c(my_MBL_CLL_colours_named, "not significant" = "lightgrey")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

volcano.plot_LC_B

volcano.plot_LC_B %>%
  ggsave(
    filename = file.path(plotting_dir_LC_subset, "Differential_Motif_Enrichment_Volcanoplot_LC_MBL_B.png"),
    width = 32,
    height = 19.2,
    unit = "cm"
  )

saveRDS(volcano.plot_LC_B, file.path(RDS_dir_LC_subset, "volcanoplot_TF_enrichment_pb_LC_MBL_B.RDS"))
```

# Subset to healthy all B, MBL and CLL cells

```{r}
idxSample_Blike <- BiocGenerics::which(CLLProj@cellColData$patient_id != "MBL7dup" &
  CLLProj@cellColData$My_annotation %in% c("B", "MBL_B_Cells", "CLL_B_Cells"))

cellsSample_Blike <- CLLProj$cellNames[idxSample_Blike]

CLLProj_Blike <- subsetArchRProject(
  ArchRProj = CLLProj,
  cells = cellsSample_Blike,
  outputDirectory = ArchR_dir_Blike,
  dropCells = TRUE
)

CLLProj_Blike <- loadArchRProject(path = ArchR_dir_Blike)

# Add B_like annotation
B_like_annotation <- tibble("label_cellbarcode" = CLLProj_Blike$cellNames) %>%
  bind_cols(CLLProj_Blike@cellColData %>% as_tibble()) %>%
  dplyr::mutate(B_like_annotation = case_when(My_annotation == "MBL_B_Cells" & MBL_CLL_subtype == "LC_MBL" ~ "LC_MBL_B_Cells",
    My_annotation == "MBL_B_Cells" & MBL_CLL_subtype == "HC_MBL" ~ "HC_MBL_B_Cells",
    .default = My_annotation
  )) %>%
  dplyr::select(label_cellbarcode, B_like_annotation) %>%
  deframe()


CLLProj_Blike <- addCellColData(
  ArchRProj = CLLProj_Blike,
  data = B_like_annotation,
  cells = names(B_like_annotation),
  name = "B_like_annotation"
)

# Add sequencing date annotation
seqdate_annotation <- tibble("label_cellbarcode" = CLLProj_Blike$cellNames) %>%
  tidyr::separate(label_cellbarcode, into=c("label", "cellbarcode"), sep="#", remove = F) %>% 
  dplyr::left_join(metadata %>% dplyr::filter(data_type =="ATAC") %>% dplyr::select(label, sequencing_date)) %>% 
  dplyr::mutate(sequencing_date=as.numeric(as.factor(sequencing_date))) %>% 
  dplyr::select(label_cellbarcode, sequencing_date) %>%
  deframe()

CLLProj_Blike <- addCellColData(
  ArchRProj = CLLProj_Blike,
  data = seqdate_annotation,
  cells = names(seqdate_annotation),
  name = "sequencing_date"
)
```

```{r}
umap_seqdate <- list()
for (row in 1:nrow(dimred_df)) {
  dimred <- dimred_df$dimred[[row]]

  umaps <- plotEmbedding(
    ArchRProj = CLLProj_Blike,
    colorBy = "cellColData",
    name = "sequencing_date",
    embedding = glue("UMAP_{dimred}"),
    plotAs = "points" ) +
    labs(x = "UMAP1", y = "UMAP2") +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "bottom",
      legend.direction = "horizontal"
    ) +
    guides(colour = guide_legend(
      keywidth = 0.5,
      keyheight = 0.5,
      override.aes = list(size = 3)
    ))

  umap_seqdate[[row]] <- umaps
}
```

### Gene tracks Blike
```{r}
B_tracks <- plotBrowserTrack(
  ArchRProj = CLLProj_Blike,
  groupBy = "B_like_annotation",
  geneSymbol = c("CD5", "FCER2", "MME"),
  upstream = 40000,
  downstream = 40000
)

for (track in B_tracks) {
  grid::grid.newpage()
  grid::grid.draw(track)
}


Blike_tracks <- list()
Blike_tracks["CD5"]<- plotBrowserTrack(
  ArchRProj = CLLProj_Blike,
  groupBy = "B_like_annotation",
  geneSymbol = c("CD5"),
  upstream = 50000,
  downstream = 50000
)

Blike_tracks["FCER2"]  <- plotBrowserTrack(
  ArchRProj = CLLProj_Blike,
  groupBy = "B_like_annotation",
  geneSymbol = c("FCER2"),
  upstream = 50000,
  downstream = 50000
)

Blike_tracks["MME"]  <- plotBrowserTrack(
  ArchRProj = CLLProj_Blike,
  groupBy = "B_like_annotation",
  geneSymbol = c("MME"),
  upstream = 10000,
  downstream = 170000
)

for (track in 1:length(Blike_tracks)) {
  grid::grid.newpage()
  grid::grid.draw(Blike_tracks[[track]])
}

saveRDS(Blike_tracks, 
        file.path(RDS_dir_Blike,"Blike_tracks.RDS"))
```


### Add pseudobulk replicates

```{r}
CLLProj_Blike <- addGroupCoverages(
  ArchRProj = CLLProj_Blike,
  groupBy = "B_like_annotation"
)
```

### Peak calling

```{r}
CLLProj_Blike <- addReproduciblePeakSet(
  ArchRProj = CLLProj_Blike,
  groupBy = "B_like_annotation",
  pathToMacs2 = pathToMacs3
)

saveArchRProject(
  ArchRProj = CLLProj_Blike,
  outputDirectory = ArchR_dir_Blike,
  load = FALSE,
  overwrite = TRUE
)
```

```{r}
CLLProj_Blike <- addPeakMatrix(CLLProj_Blike)

saveArchRProject(
  ArchRProj = CLLProj_Blike,
  outputDirectory = ArchR_dir_Blike,
  load = FALSE,
  overwrite = TRUE
)

getAvailableMatrices(CLLProj_Blike)
```


### Marker Peaks

```{r}
# Gets peaks which are unique to cell types and account for sequencing date bias
markersPeaks_Blike <- getMarkerFeatures(
  ArchRProj = CLLProj_Blike,
  useMatrix = "PeakMatrix",
  groupBy = "B_like_annotation",
  bias = c("TSSEnrichment", "log10(nFrags)", "sequencing_date"),
  testMethod = "wilcoxon",
  threads = 1
)
markerList <- getMarkers(markersPeaks_Blike, cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1")
```
### Volcano Plots MarkerPeaks

```{r, fig.width=20,fig.height=5}
# unique peaks for MBL or CLL cells compared to all other cell types
volcanoplot_MP_B <- plotMarkers(
  seMarker = markersPeaks_Blike,
  name = "B",
  cutOff = "FDR <= 0.05 & Log2FC >= 1",
  plotAs = "Volcano"
) +
  geom_point(size = 1) +
  scale_colour_manual(
    name = "DAR",
    values = c("Down-Regulated" = "gray2", "Up-Regulated" = "darkred", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "LC-MBL and CLL cells",
      "Up-Regulated" = "Physiologic B cells",
      "Not-Differential" = "not significant"
    )
  ) +
  theme_pub()

volcanoplot_MP_LC_MBL <- plotMarkers(
  seMarker = markersPeaks_Blike,
  name = "LC_MBL_B_Cells",
  cutOff = "FDR <= 0.05 & Log2FC >= 1",
  plotAs = "Volcano"
) +
  geom_point(size = 1) +
  scale_colour_manual(
    name = "DAR",
    values = c("Down-Regulated" = "gray2", "Up-Regulated" = "#fcbda4", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "Physiologic and CLL cells",
      "Up-Regulated" = "LC-MBL cells",
      "Not-Differential" = "not significant"
    )
  ) +
  theme_pub()

volcanoplot_MP_HC_MBL <- plotMarkers(
  seMarker = markersPeaks_Blike,
  name = "HC_MBL_B_Cells",
  cutOff = "FDR <= 0.05 & Log2FC >= 1",
  plotAs = "Volcano"
) +
  geom_point(size = 3) +
  scale_colour_manual(
    name = "DAR",
    values = c("Down-Regulated" = "gray2", "Up-Regulated" = "#FC8D62", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "Physiologic and CLL cells",
      "Up-Regulated" = "HC-MBL cells",
      "Not-Differential" = "not significant"
    )
  ) +
  theme_pub()

volcanoplot_MP_CLL <- plotMarkers(
  seMarker = markersPeaks_Blike,
  name = "CLL_B_Cells",
  cutOff = "FDR <= 0.05 & Log2FC >= 1",
  plotAs = "Volcano"
) +
  geom_point(size = 1) +
  scale_colour_manual(
    name = "DAR",
    values = c("Down-Regulated" = "gray2", "Up-Regulated" = "#66C2A5", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "Physiologic B and LC-MBL cells",
      "Up-Regulated" = "CLL cells",
      "Not-Differential" = "not significant"
    )
  ) +
  theme_pub()

ggarrange(volcanoplot_MP_B, volcanoplot_MP_LC_MBL, volcanoplot_MP_HC_MBL, volcanoplot_MP_CLL, ncol = 4) %>%
  ggsave(
    plot = .,
    filename = file.path(plotting_dir_Blike, "Unique_Markers_Blike_Volcanoplot.pdf"),
    width = 20,
    height = 5,
    unit = "in"
  )
```

### Pairwise peaks

```{r }
markerTest_LC_MBL_HC_MBL <- getMarkerFeatures(
  ArchRProj = CLLProj_Blike,
  useMatrix = "PeakMatrix",
  groupBy = "B_like_annotation",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)", "sequencing_date"),
  useGroups = "HC_MBL_B_Cells",
  bgdGroups = "LC_MBL_B_Cells"
)

markerTest_LC_MBL_CLL <- getMarkerFeatures(
  ArchRProj = CLLProj_Blike,
  useMatrix = "PeakMatrix",
  groupBy = "B_like_annotation",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)", "sequencing_date"),
  useGroups = "CLL_B_Cells",
  bgdGroups = "LC_MBL_B_Cells"
)

markerTest_HC_MBL_CLL <- getMarkerFeatures(
  ArchRProj = CLLProj_Blike,
  useMatrix = "PeakMatrix",
  groupBy = "B_like_annotation",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)", "sequencing_date"),
  useGroups = "CLL_B_Cells",
  bgdGroups = "HC_MBL_B_Cells"
)


# unique peaks between LC-MBL and B cells only!
markerTest_B_LC_MBL <- getMarkerFeatures(
  ArchRProj = CLLProj_Blike,
  useMatrix = "PeakMatrix",
  groupBy = "B_like_annotation",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)", "sequencing_date"),
  useGroups = "LC_MBL_B_Cells",
  bgdGroups = "B"
)

markerTest_B_HC_MBL <- getMarkerFeatures(
  ArchRProj = CLLProj_Blike,
  useMatrix = "PeakMatrix",
  groupBy = "B_like_annotation",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)", "sequencing_date"),
  useGroups = "HC_MBL_B_Cells",
  bgdGroups = "B"
)

markerTest_B_CLL <- getMarkerFeatures(
  ArchRProj = CLLProj_Blike,
  useMatrix = "PeakMatrix",
  groupBy = "B_like_annotation",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)", "sequencing_date"),
  useGroups = "CLL_B_Cells",
  bgdGroups = "B"
)
```


```{r}
result <- plotMarkerHeatmap(
  plotLog2FC = T,
  transpose = T,
  seMarker = markerTest_B_CLL,
  cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1.5")
```

### Volcano plots Pairwise testing
```{r fig.width=12, fig.height=4}
volcanoplot_PWT_LC_MBL_HC_MBL <-plotMarkers(
  seMarker = markerTest_LC_MBL_HC_MBL,
  name = "HC_MBL_B_Cells",
  cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
  plotAs = "Volcano",
  rastr = TRUE
) + 
geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
  scale_y_continuous(
    name = "-log10(FDR)",
    breaks = seq(0, 30,10),
    limits = c(0, 30),
    # expand = expansion(mult = c(0, 0.01)),
    oob = scales::oob_squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak accessible in",
    values = c("Down-Regulated" ="#fcbda4" ,"Up-Regulated" ="#FC8D62" , "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "LC-MBL cells",
      "Up-Regulated" = "HC-MBL cells",
      "Not-Differential" = "Not significant"
    )
  ) +
  theme_pub() 


volcanoplot_PWT_LC_MBL_CLL <- plotMarkers(
  seMarker = markerTest_LC_MBL_CLL,
  name = "CLL_B_Cells",
  cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
  plotAs = "Volcano",
  rastr = TRUE
) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
  scale_y_continuous(
    name = "-Log10(FDR)",
    breaks = seq(0, 30,10),
    limits = c(0, 30),
    # expand = expansion(mult = c(0, 0.01)),
    oob = scales::oob_squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak accessible in",
    values = c("Down-Regulated" = "#fcbda4", "Up-Regulated" ="#66C2A5" , "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "LC-MBL cells",
      "Up-Regulated" = "CLL cells",
      "Not-Differential" = "Not significant"
    )
  ) +
  theme_pub() 

volcanoplot_PWT_HC_MBL_CLL <- plotMarkers(
  seMarker = markerTest_HC_MBL_CLL,
  name = "CLL_B_Cells",
  cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
  plotAs = "Volcano",
  rastr = TRUE
) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
  scale_y_continuous(
    name = "-Log10(FDR)",
    breaks = seq(0, 30,10),
    limits = c(0,30),
    # expand = expansion(mult = c(0, 0.01)),
    oob = scales::oob_squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak accessible in",
    values = c("Down-Regulated" = "#FC8D62", "Up-Regulated" ="#66C2A5" , "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "HC-MBL cells",
      "Up-Regulated" = "CLL cells",
      "Not-Differential" = "Not significant"
    )
  ) +
  theme_pub() 



volcanoplot_PWT_B_LC_MBL <- plotMarkers(
  seMarker = markerTest_B_LC_MBL,
  name = "LC_MBL_B_Cells",
  cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
  plotAs = "Volcano",
  rastr = TRUE
) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
  scale_y_continuous(
    name = "-Log10(FDR)",
    breaks = seq(0, 30,10),
    limits = c(0,30),
    # expand = expansion(mult = c(0, 0.01)),
    oob = scales::oob_squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak accessible in",
    values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#fcbda4", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "Physiologic B cells",
      "Up-Regulated" = "LC-MBL cells",
      "Not-Differential" = "Not significant"
    )
  ) +
  theme_pub() #+
  # labs(title = "")

volcanoplot_PWT_B_HC_MBL <- plotMarkers(
  seMarker = markerTest_B_HC_MBL,
  name = "HC_MBL_B_Cells",
  cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
  plotAs = "Volcano",
  rastr = TRUE
) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
  scale_y_continuous(
    name = "-Log10(FDR)",
    breaks = seq(0, 30,10),
    limits = c(0,30),
    # expand = expansion(mult = c(0, 0.01)),
    oob = scales::oob_squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak accessible in",
    values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#FC8D62", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "Physiologic B cells",
      "Up-Regulated" = "HC-MBL cells",
      "Not-Differential" = "Not significant"
    )
  ) +
  theme_pub() #+
  # labs(title = "")

volcanoplot_PWT_B_CLL <- plotMarkers(
  seMarker = markerTest_B_CLL,
  name = "CLL_B_Cells",
  cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1",
  plotAs = "Volcano",
  rastr = TRUE
) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
  scale_y_continuous(
    name = "-Log10(FDR)",
    breaks = seq(0, 30,10),
    limits = c(0,30),
    # expand = expansion(mult = c(0, 0.01)),
    oob = scales::oob_squish_infinite
  ) +
  scale_colour_manual(
    name = "Peak accessible in",
    values = c("Down-Regulated" = "darkred", "Up-Regulated" = "#66C2A5", "Not-Differential" = "#D3D3D3"),
    labels = c(
      "Down-Regulated" = "Physiologic B cells",
      "Up-Regulated" = "CLL cells",
      "Not-Differential" = "Not significant"
    )
  ) +
  theme_pub() #+
  # labs(title = "")

ggarrange(volcanoplot_PWT_B_LC_MBL, volcanoplot_PWT_B_HC_MBL, volcanoplot_PWT_B_CLL, nrow=1)
ggarrange(volcanoplot_PWT_LC_MBL_HC_MBL, volcanoplot_PWT_LC_MBL_CLL, volcanoplot_PWT_HC_MBL_CLL, nrow=1)

ggarrange(volcanoplot_PWT_B_LC_MBL, volcanoplot_PWT_B_HC_MBL, volcanoplot_PWT_B_CLL, nrow=1) %>% 
  ggsave(file=file.path(plotting_dir_Blike, "Peaks_volcanoplots.png"), units = "in", width=12 , height =4 )
```

```{r }
ggsave(
  plot = volcanoplot_PWT_B_LC_MBL,
  filename = file.path(plotting_dir_Blike, "Upregulation_LC_MBL_Volcanoplot.pdf"),
  width = 5,
  height = 5,
  unit = "in"
)

ggsave(
  plot = volcanoplot_PWT_B_HC_MBL,
  filename = file.path(plotting_dir_Blike, "Upregulation_HC_MBL_Volcanoplot.pdf"),
  width = 5,
  height = 5,
  unit = "in"
)

ggsave(
  plot = volcanoplot_PWT_B_CLL,
  filename = file.path(plotting_dir_Blike, "Upregulation_CLL_Volcanoplot.pdf"),
  width = 5,
  height = 5,
  unit = "in"
)

saveRDS(volcanoplot_PWT_B_LC_MBL, file.path(RDS_dir_Blike, "volcanoplot_PWT_B_LC_MBL.RDS"))
saveRDS(volcanoplot_PWT_B_HC_MBL, file.path(plotting_dir_Blike, "volcanoplot_PWT_B_HC_MBL.RDS"))
saveRDS(volcanoplot_PWT_B_CLL, file.path(plotting_dir_Blike, "volcanoplot_PWT_B_CLL.RDS"))
```

```{r}
saveArchRProject(
  ArchRProj = CLLProj_Blike,
  outputDirectory = ArchR_dir_Blike,
  load = FALSE,
  overwrite = TRUE
)
```

### Add TF Mofif Annotations

```{r}
CLLProj_Blike <- addMotifAnnotations(
  ArchRProj = CLLProj_Blike,
  motifSet = "cisbp",
  annoName = "Motif"
)
```


```{r calculate_motif_enrichment}
calculate_motif_enrichment <- function(seMarker_object){
  
  motifsUp <- peakAnnoEnrichment(
      seMarker = seMarker_object,
      ArchRProj = CLLProj_Blike,
      peakAnnotation = "Motif",
      cutOff = "FDR <= 0.05 & Log2FC >= 1" # these are the criteria for the peaks which should be compared, not the TFs
    ) 
  
  motifsDo <- peakAnnoEnrichment(
      seMarker = seMarker_object,
      ArchRProj = CLLProj_Blike,
      peakAnnotation = "Motif",
      cutOff = "FDR <= 0.05 & Log2FC <= -1" # these are the criteria for the peaks which should be compared, not the TFs
    ) 
  
  return(list(motifsUp, motifsDo) %>% 
           set_names("MotifsUp", "MotifsDo"))
}
```


```{r}
all_motif_enrichments <- list(markerTest_B_LC_MBL,
     markerTest_B_HC_MBL,
     markerTest_B_CLL,
     markerTest_LC_MBL_HC_MBL,
     markerTest_HC_MBL_CLL,
     markerTest_LC_MBL_CLL) %>% 
  set_names(c("B_LC-MBL", "B_HC-MBL", "B_CLL", "LC-MBL_HC-MBL", "HC-MBL_CLL", "LC-MBL_CLL")) %>% 
  map(~calculate_motif_enrichment(.))
```

```{r}
visualise_motif_enrichment_rank <- function(up_down_list, comparison) {
  df_up <- data.frame(TF = rownames(up_down_list$MotifsUp), mlog10Padj = assay(up_down_list$MotifsUp)[, 1])
  df_up <- df_up[order(df_up$mlog10Padj, decreasing = TRUE), ]
  df_up$rank <- seq_len(nrow(df_up))

  ggUp <- df_up %>%
    dplyr::mutate(TF_label = if_else(rank %in% 1:5, TF, NA)) %>%
    ggplot(aes(rank, mlog10Padj, color = mlog10Padj, label = TF_label)) +
    geom_point(size = 0.5) +
    geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
    ggrepel::geom_label_repel(
      label.size = NA,
      size = 1,
      # nudge_x = 2,
      color = "black",
      max.overlaps = 10
    )  +
    labs(x = "Rank", y = "", title = "Upregulated motifs") +
    scale_y_continuous(limits=c(0, 160))+
    scale_color_gradientn(colors = c("lightgrey", get_palette(palette ="YlOrRd", k = 9 )), 
                          breaks=c(0, seq(-log10(0.05), 160, length.out=9)),
                          labels = round(c(0, seq(-log10(0.05), 160, length.out = 9)), 2),
                          limits = c(0, 160))+
    theme_pub() +
    theme(legend.position = "none",
          axis.line.y=element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.y=element_blank())


  df_do <- data.frame(TF = rownames(up_down_list$MotifsDo), mlog10Padj = assay(up_down_list$MotifsDo)[, 1])
  df_do <- df_do[order(df_do$mlog10Padj, decreasing = TRUE), ]
  df_do$rank <- seq_len(nrow(df_do))

  ggDo <- df_do %>%
    dplyr::mutate(TF_label = if_else(rank %in% 1:5, TF, NA)) %>%
    ggplot(aes(rank, mlog10Padj, color = mlog10Padj, label = TF_label)) +
    geom_point(size = 0.5) +
    geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
    geom_label_repel(
      label.size = NA,
      size = 1,
      # nudge_x = 2,
      max.overlaps = 10,
      color = "black"
    )  +
    labs(x = "Rank", y = "-log10(FDR)", title = "Downregulated motifs") +
    scale_y_continuous(limits=c(0, 160))+
    scale_color_gradientn(colors = c("lightgrey", get_palette(palette ="YlOrRd", k = 9 )), 
                          breaks=c(0, seq(-log10(0.05), 160, length.out=9)),
                          labels = round(c(0, seq(-log10(0.05), 160, length.out = 9)),0),
                          limits = c(0, 160))+
    theme_pub() +
    theme(legend.position = "none")
  
  up_do_plot <- ggarrange(ggDo, ggUp,  nrow=1, ncol=2, common.legend = TRUE, legend="none") #%>% annotate_figure(top = text_grob(comparison, size=12))
  return(up_do_plot)
}
```

```{r visualise_motif_enrichment_volcano}
# visualise_motif_enrichment_volcano <- function(up_down_list, comparison) {
#   volcano_data <-tibble("TF" = up_down_list@NAMES) %>%
#       bind_cols(assay(up_down_list, "mlog10Padj"), assay(up_down_list, "Enrichment")) %>%
#       `colnames<-`(c("TF", "mlog10Padj", "Enrichment"))
#   
#   bgd <- str_split_i(comparison, "_", 1)
#   fgd <- str_split_i(comparison, "_", 2)
# 
#   colouring_values <- c("down"=case_when(bgd == "B" ~ "darkred",
#                                          bgd =="LC-MBL" ~ "#fcbda4",
#                                          bgd =="HC-MBL" ~ "#FC8D62",
#                                          bgd =="CLL" ~ "#66C2A5"),
#                         "up" = case_when(fgd == "B" ~ "darkred",
#                                          fgd =="LC-MBL" ~ "#fcbda4",
#                                          fgd =="HC-MBL" ~ "#FC8D62",
#                                          fgd =="CLL" ~ "#66C2A5"),
#                         "not significant" = "#D3D3D3")
#   
#   colouring_labels <- c("down"=bgd, "up"=fgd, "not significant"="not significant")
# 
#   volcanoplot <- volcano_data %>%
#     mutate(
#       significance = case_when(mlog10Padj >= 1.30103 & log2(Enrichment) >= 0.5 ~ "up",
#         mlog10Padj >= 1.30103 & log2(Enrichment) <= -0.5 ~ "down",
#         .default = "not significant"
#       ),
#       significance = factor(significance, levels = c("down", "up", "not significant")),
#       label = if_else(significance != "not significant", str_split_i(TF, "_", i = 1), NA)
#     ) %>%
#     ggplot(aes(x = log2(Enrichment), y = mlog10Padj, colour = significance, label = label)) +
#     geom_vline(xintercept = c(-0.5, 0.5), color = "grey", linetype = "dashed") +
#     geom_hline(yintercept = c(1.30103), color = "grey", linetype = "dashed") +
#     geom_point() +
#     ggrepel::geom_label_repel(
#       size = 2.5,
#       label.size = NA,
#       label.padding = 0.1,
#       segment.size = 0.25,
#       segment.colour = "grey",
#       min.segment.length = 0.1,
#       max.overlaps = 100,
#       colour = "black"
#     ) +
#     scale_x_continuous("Log2(Enrichment)", limits = c(-3.5, 3.5)
#                        ) +
#     scale_y_continuous("-log10(FDR)",
#       limits = c(0, 160),
#       oob = scales::squish_infinite,
#       expand = expansion(mult = c(0, NA))
#     ) +
#     scale_colour_manual(name="TF enrichment in ",
#                         values=colouring_values,
#                         labels=colouring_labels) +
#     theme_pub() +
#     theme(legend.position = "bottom")
#   return(volcanoplot)
# }
```

```{r, fig.width=16, fig.height=3}
ranks <- all_motif_enrichments %>%
  imap(~ visualise_motif_enrichment_rank(.x, .y))

ggarrange(plotlist = ranks[1:3], nrow = 1, ncol = 3, common.legend = TRUE)
saveRDS(ranks[1:3], file=file.path(RDS_dir_Blike, "rank_plots_LC_HC_CLL.rds"))

ggarrange(plotlist = ranks[4:6], nrow = 1, ncol = 3)
```

```{r}
get_significant_changes <- function(se_list, up_or_down){
   heatmap_data<- tibble("TF" = se_list[[up_or_down]]@NAMES) %>%
      bind_cols(assay(se_list[[up_or_down]], "mlog10Padj"), assay(se_list[[up_or_down]], "Enrichment")) %>%
      `colnames<-`(c("TF", "mlog10Padj", "Enrichment"))
  
  unique_signif_TFs <- heatmap_data %>% 
    dplyr::filter(mlog10Padj >= 1.30103,
                  Enrichment >= 1.5) %>% 
    pull(TF) %>% unique()
  return(unique_signif_TFs)
}

all_TS_oi <- list()
all_TS_oi[["MotifsUp"]] <- all_motif_enrichments[1:3] %>% 
  map(~get_significant_changes(., "MotifsUp")) %>% 
  unlist() %>% 
  unique()

all_TS_oi[["MotifsDo"]]  <- all_motif_enrichments[1:3] %>% 
  map(~get_significant_changes(., "MotifsDo")) %>% 
  unlist() %>% 
  unique()
```


```{r}
concatenate_rank_data <- function(list_se_list, up_or_down) {

  heatmap_data <- list_se_list %>% 
    imap(~tibble("TF" = .x[[1]]@NAMES) %>%
      bind_cols(assay(.x[[up_or_down]], "mlog10Padj"), assay(.x[[up_or_down]], "Enrichment")) %>%
      `colnames<-`(c("TF", "mlog10Padj", "Enrichment")) %>% 
        dplyr::filter(TF %in% all_TS_oi[[up_or_down]])) %>% 
    bind_rows(.id = "comparison")

  return(heatmap_data)
}
```


```{r}
enrichment_data_up <-all_motif_enrichments[1:3] %>% concatenate_rank_data(., "MotifsUp")
enrichment_data_down <-all_motif_enrichments[1:3] %>% concatenate_rank_data(., "MotifsDo")

enrichment_data_scores_up <- enrichment_data_up %>% 
  dplyr::select(-mlog10Padj) %>%
  dplyr::mutate(TF=str_split_i(TF, "_", 1),
                Enrichment=if_else(Enrichment =="NaN", NA, Enrichment),
                comparison=case_when(comparison=="B_LC-MBL" ~ "LC-MBL",
                                     comparison=="B_HC-MBL" ~ "HC-MBL",
                                     comparison=="B_CLL" ~ "CLL", .default = comparison))%>%
  dplyr::rename("B cells vs."=comparison) %>%
  tidyr::pivot_wider(id_cols=TF, names_from =`B cells vs.`, values_from = Enrichment) %>%
  tibble::column_to_rownames(var = "TF") %>% 
  as.matrix

enrichment_data_scores_down<- enrichment_data_down %>% 
  dplyr::select(-mlog10Padj) %>%
  dplyr::mutate(TF=str_split_i(TF, "_", 1),
                Enrichment=if_else(Enrichment =="NaN", NA, Enrichment),
                comparison=case_when(comparison=="B_LC-MBL" ~ "LC-MBL",
                                     comparison=="B_HC-MBL" ~ "HC-MBL",
                                     comparison=="B_CLL" ~ "CLL", .default = comparison))%>%
  dplyr::rename("B cells vs."=comparison) %>%
  tidyr::pivot_wider(id_cols=TF, names_from =`B cells vs.`, values_from = Enrichment) %>%
  tibble::column_to_rownames(var = "TF") %>% 
  as.matrix

```


```{r concatenate_volcano_data}
# concatenate_volcano_data <- function(se_list) {
#   volcano_data <- se_list %>% 
#     imap(~tibble("TF" = .x@NAMES) %>%
#       bind_cols(assay(.x, "mlog10Padj"), assay(.x, "Enrichment")) %>%
#       `colnames<-`(c("TF", "mlog10Padj", "Enrichment")) %>% 
#        dplyr::mutate(comparison=.y)) %>% 
#     bind_rows()
#   
#   # which TF motifs to change significantly over time
#   TFs_oi <- volcano_data %>% 
#     dplyr::filter(mlog10Padj >=1.30103 & log2(Enrichment) >=0.5) %>% 
#     pull(TF) %>% 
#     unique()
#   
#   return(list(volcano_data, TFs_oi))
# }
#   
# enrichment_data <-concatenate_volcano_data(all_motif_enrichments)
# 
# enrichment_data_scores <- enrichment_data[[1]] %>%
#   dplyr::filter(TF %in% enrichment_data[[2]]) %>% 
#   dplyr::select(-mlog10Padj) %>% 
#   dplyr::mutate(TF=str_split_i(TF, "_", 1),
#                 Enrichment=if_else(Enrichment =="NaN", NA, Enrichment),
#                 comparison=case_when(comparison=="B_LC-MBL" ~ "LC-MBL",
#                                      comparison=="B_HC-MBL" ~ "HC-MBL",
#                                      comparison=="B_CLL" ~ "CLL", .default = comparison)) %>% 
#   dplyr::rename("B cells vs."=comparison) %>% 
#   tidyr::pivot_wider(id_cols=TF, names_from =`B cells vs.`, values_from = Enrichment) %>% 
#   tibble::column_to_rownames(var = "TF") %>% 
#   as.matrix()
```


```{r, fig.height=10, fig.width=10}
comparison2colour <- c("LC-MBL"="#fcbda4", "HC-MBL"="#FC8D62", "CLL"="#66C2A5")
anno_df <- tibble("Background vs."=c("LC-MBL", "HC-MBL", "CLL"))
colours=list("Background vs."=comparison2colour)
      
column_ha <-HeatmapAnnotation(
  df = anno_df,
  col = colours,
  show_legend = FALSE,
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8)
)

hm_up <- Heatmap(
  matrix = enrichment_data_scores_up,
  col = colorRampPalette(brewer.pal(n = 9, name = "YlOrRd"))(100),
  top_annotation = column_ha,
  use_raster = TRUE,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  raster_by_magick = TRUE,
  show_column_names = TRUE,
  show_heatmap_legend = TRUE,
  na_col = "white",
  row_title_side = "left",
  row_names_side = "left",
  column_names_side = "top",
  row_names_gp = gpar(fontsize = 8),
  row_title_gp = gpar(fontsize = 8),
  heatmap_legend_param = list(
    title = "Enrichment",
    direction = "horizontal",
    legend_direction = "horizontal",
    legend_gp = gpar(
      fontsize = 8,
      fontface = "plain"
    )
  )
)

hm_down <- Heatmap(
  matrix = enrichment_data_scores_down,
  col = colorRampPalette(brewer.pal(n = 9, name = "YlGnBu"))(100),
  top_annotation = column_ha,
  use_raster = TRUE,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  raster_by_magick = TRUE,
  show_column_names = TRUE,
  show_heatmap_legend = TRUE,
  na_col = "white",
  row_title_side = "left",
  row_names_side = "left",
  column_names_side = "top",
  row_names_gp = gpar(fontsize = 8),
  row_title_gp = gpar(fontsize = 8),
  heatmap_legend_param = list(
    title = "Depletion",
    direction = "horizontal",
    legend_direction = "horizontal",legend.position="bottom",
    legend_gp = gpar(
      fontsize = 8,
      fontface = "plain"
    )
  )
)
draw(hm_up, heatmap_legend_side = "bottom")

hm_up_grob<- grid::grid.grabExpr(draw(hm_up, heatmap_legend_side = "bottom"))
hm_down_grob <- grid::grid.grabExpr(draw(hm_down, heatmap_legend_side = "bottom"))

saveRDS(hm_up_grob, file=file.path(RDS_dir_Blike, "TF_enrichment_LC_MBL_CLL_marker_heatmap.RDS"))
saveRDS(hm_up_grob, file=file.path(RDS_dir_Blike, "TF_depletion_LC_MBL_CLL_marker_heatmap.RDS"))

cowplot::plot_grid(hm_up_grob, hm_down_grob, nrow=1, ncol=2)
```


```{r}
# background peaks added
CLLProj_Blike <- addBgdPeaks(CLLProj_Blike, method = "chromVAR")
```

```{r}
# motif deviation matrix added
CLLProj_Blike <- addDeviationsMatrix(
  ArchRProj = CLLProj_Blike,
  bgdPeaks = getBgdPeaks(CLLProj_Blike,
    method = "chromVAR"
  ),
  peakAnnotation = "Motif",
  threads = 1
)
```

```{r}
plotVarDev <- getVarDeviations(CLLProj_Blike, 
                               name = "MotifMatrix", 
                               plot = TRUE)
```

### Get Matrix from project

```{r}
MotifMatrix_tmp_Blike <- getMatrixFromProject(
  ArchRProj = CLLProj_Blike,
  useMatrix = "MotifMatrix",
  useSeqnames = NULL,
  verbose = TRUE,
  binarize = FALSE,
  threads = 1
)

zscore_matrix_Blike <- as.matrix(assays(MotifMatrix_tmp_Blike)$z) %>%
  as.data.frame()

deviations_matrix_Blike <- as.matrix(assays(MotifMatrix_tmp_Blike)$deviations) %>%
  as.data.frame()

VarDeviation_Blike <- getVarDeviations(CLLProj_Blike,
  name = "MotifMatrix",
  plot = FALSE
) %>%
  as.data.frame()

Top_TF_target_Blike <- head(VarDeviation_Blike$name, n = 100)

zscore_part_Blike <- deviations_matrix_Blike[Top_TF_target_Blike, ]
```

```{r}
condition_Blike <- tibble(
  "label_cellbarcode" = CLLProj_Blike$cellNames,
  "B_like_annotation" = CLLProj_Blike$B_like_annotation
)

B_cells_Blike <- dplyr::filter(condition_Blike, B_like_annotation == "B") %>%
  pull(label_cellbarcode)

LC_MBL_cells_Blike <- dplyr::filter(condition_Blike, B_like_annotation == "LC_MBL_B_Cells") %>%
  pull(label_cellbarcode)

HC_MBL_cells_Blike <- dplyr::filter(condition_Blike, B_like_annotation == "HC_MBL_B_Cells") %>%
  pull(label_cellbarcode)

CLL_cells_Blike <- dplyr::filter(condition_Blike, B_like_annotation == "CLL_B_Cells") %>%
  pull(label_cellbarcode)

# # calculate data for volcano plot
input_zscore_matrix_Blike <- zscore_matrix_Blike

# If no NA values are found in zcore matrix calculate statistics
condition_list <- list(
  "B_vs_LC_MBL" = list(LC_MBL_cells_Blike, B_cells_Blike),
  "B_vs_HC_MBL" = list(HC_MBL_cells_Blike, B_cells_Blike),
  "B_vs_CLL" = list(CLL_cells_Blike, B_cells_Blike),
  "LC_MBL_vs_HC_MBL" = list(HC_MBL_cells_Blike, LC_MBL_cells_Blike),
  "HC_MBL_vs_CLL" = list(CLL_cells_Blike, HC_MBL_cells_Blike),
  "LC_MBL_vs_CLL" = list(CLL_cells_Blike, LC_MBL_cells_Blike)
)

# volcano.data_B_LC_MBL <- data.frame()
# volcano.data_B_CLL <- data.frame()
volcano.data.all.comparisons <- list(
  "B_vs_LC_MBL" = data.frame(),
  "B_vs_HC_MBL" = data.frame(),
  "B_vs_CLL" = data.frame(),
  "LC_MBL_vs_HC_MBL" = data.frame(),
  "HC_MBL_vs_CLL" = data.frame(),
  "LC_MBL_vs_CLL" = data.frame()
)

j <- 0
for (j in 1:length(condition_list)) {
  print(paste0("Current comparison: ", names(condition_list[j])))
  i <- 0
  for (TF in rownames(input_zscore_matrix_Blike)) {
    i <- i + 1
    print(i)
    print(TF)
    if (length(which(is.na(input_zscore_matrix_Blike[TF, ]))) == 0) {
      p.value <- wilcox.test(as.numeric(input_zscore_matrix_Blike[TF, condition_list[[j]][[1]]]), as.numeric(input_zscore_matrix_Blike[TF, condition_list[[j]][[2]]]))$p.value
      diff.accessibility <- median(as.numeric(input_zscore_matrix_Blike[TF, condition_list[[j]][[1]]])) - median(as.numeric(input_zscore_matrix_Blike[TF, condition_list[[j]][[2]]]))
      volcano.data.all.comparisons[[j]] <- rbind(volcano.data.all.comparisons[[j]], data.frame(TF = TF, p.value = p.value, diff.accessibility = diff.accessibility))
    }
  }
  volcano.data.all.comparisons[[j]]$p.adj <- p.adjust(volcano.data.all.comparisons[[j]]$p.value)
  volcano.data.all.comparisons[[j]]$p.adj[which(volcano.data.all.comparisons[[j]]$p.adj == 0)] <- min(volcano.data.all.comparisons[[j]]$p.adj[which(volcano.data.all.comparisons[[j]]$p.adj != 0)])
  volcano.data.all.comparisons[[j]]$p.adj <- as.numeric(volcano.data.all.comparisons[[j]]$p.adj)
  volcano.data.all.comparisons[[j]]$diff.accessibility <- as.numeric(volcano.data.all.comparisons[[j]]$diff.accessibility)
  volcano.data.all.comparisons[[j]]$significant <- ifelse(abs(volcano.data.all.comparisons[[j]]$diff.accessibility) > 1.5 & volcano.data.all.comparisons[[j]]$p.adj <= 0.05,
    "significant",
    "not significant"
  )
  volcano.data.all.comparisons[[j]] <- volcano.data.all.comparisons[[j]] %>%
    dplyr::mutate(
      "significant" = case_when(
        p.adj >= 0.05 ~ "not significant",
        abs(diff.accessibility) < 1.5 ~ "not significant",
        diff.accessibility > 1.5 & p.adj < 0.05 ~ "upregulated",
        diff.accessibility < -1.5 & p.adj < 0.05 ~ "downregulated"
      ),
      "label" = if_else(significant != "not significant", stringr::str_split_fixed(TF, pattern = "_", n = 2)[, 1], NA)
    )
}
saveRDS(volcano.data.all.comparisons, file = file.path(RDS_dir_Blike, "volcano.data.all.comparisons.RDS"))
```

### Plot Differential TF enrichment

```{r}
volcano.plot_B_LC_MBL <- volcano.data.all.comparisons$B_vs_LC_MBL %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-3, 3, 1),
    limits = c(-3, 3)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#fcbda4", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "LC-MBL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

volcano.plot_B_HC_MBL <- volcano.data.all.comparisons$B_vs_HC_MBL %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 2),
    limits = c(-6.5, 6.5)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#FC8D62", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "HC-MBL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

volcano.plot_B_CLL <- volcano.data.all.comparisons$B_vs_CLL %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-8, 8, 4),
    limits = c(-8, 8)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#66C2A5", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "CLL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

volcano.plot_LC_MBL_HC_MBL <- volcano.data.all.comparisons$LC_MBL_vs_HC_MBL %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  # geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    force = 20,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 2),
    limits = c(-6, 6)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.3)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "TF motif enrichment",
    values = c("upregulated" = "#FC8D62", "downregulated" = "#fcbda4", "not significant" = "lightgrey"),
    labels = c("upregulated" = "HC-MBL cells", "downregulated" = "LC-MBL cells")
  ) +
  labs(title="LC-MBL vs. CLL")+
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

volcano.plot_HC_MBL_CLL <- volcano.data.all.comparisons$HC_MBL_vs_CLL %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 2),
    limits = c(-6, 6)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#66C2A5", "downregulated" = "#FC8D62", "not significant" = "lightgrey"),
    labels = c("upregulated" = "CLL Cells", "downregulated" = "HC-MBL")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

volcano.plot_LC_MBL_CLL <- volcano.data.all.comparisons$LC_MBL_vs_CLL %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-6, 6, 2),
    limits = c(-7, 7)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "#66C2A5", "downregulated" = "#fcbda4", "not significant" = "lightgrey"),
    labels = c("upregulated" = "CLL Cells", "downregulated" = "LC-MBL")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

ggarrange(volcano.plot_B_LC_MBL, volcano.plot_LC_MBL_HC_MBL, volcano.plot_HC_MBL_CLL,
  volcano.plot_B_HC_MBL, volcano.plot_B_CLL, volcano.plot_LC_MBL_CLL,
  nrow = 2, ncol = 3
)

saveRDS(volcano.plot_B_LC_MBL, file.path(RDS_dir_Blike, "volcanoplot_TF_enrichment_pb_B_LC_MBL.RDS"))
saveRDS(volcano.plot_B_HC_MBL, file.path(RDS_dir_Blike, "volcanoplot_TF_enrichment_pb_B_HC_MBL.RDS"))
saveRDS(volcano.plot_B_CLL, file.path(RDS_dir_Blike, "volcanoplot_TF_enrichment_pb_B_CLL.RDS"))

saveRDS(volcano.plot_LC_MBL_HC_MBL, file.path(RDS_dir_Blike, "volcanoplot_TF_enrichment_pb_LC_MBL_HC_MBL.RDS"))
saveRDS(volcano.plot_HC_MBL_CLL, file.path(RDS_dir_Blike, "volcanoplot_TF_enrichment_pb_HC_MBL_CLL.RDS"))
saveRDS(volcano.plot_LC_MBL_CLL, file.path(RDS_dir_Blike, "volcanoplot_TF_enrichment_pb_LC_MBL_CLL.RDS"))

ggsave(plot = volcano.plot_LC_MBL_HC_MBL+theme(legend.position="none"), 
       filename = file.path(plotting_dir_Blike, "volcanoplot_TF_enrichment_pb_LC_HC_MBL.png"),
       width = 2.8, height = 2)
```

### Pull Deviation-scores for LC-MBL vs CLL differentially enrichted TF motifs

```{r}
comp_genes <- volcano.data.all.comparisons %>%
  bind_rows() %>%
  dplyr::filter(!is.na(label)) %>%
  pull(var = TF) %>%
  str_split(string = , pattern = "_") %>%
  map_chr(~ .[[1]]) %>%
  unique() %>%
  sort()
```

```{r, fig.width=20, fig.height=9}
zscore_matrix_hm_input <- zscore_matrix_Blike[, colnames(zscore_matrix_Blike) %>% keep(~ !str_starts(., "CLL6.") &
  !str_starts(., "CLL2.") &
  !str_starts(., "CLL5."))]

rownames(zscore_matrix_hm_input) <- rownames(zscore_matrix_hm_input) %>%
  str_split(string = , pattern = "_") %>%
  map_chr(~ .[[1]])

barcodes_ordered <- colnames(zscore_matrix_hm_input) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = CLLProj_Blike$cellNames,
    "B_like_annotation" = CLLProj_Blike@cellColData$B_like_annotation
  )) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
      B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
      B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
    B_like_annotation_label = paste0(B_like_annotation, "_", label),
    B_like_annotation_label = factor(B_like_annotation_label,
      levels = c(
        paste0("Physiologic B Cells_", c("CLL8.1", "CLL7.2", "MBL9", "MBL13", "CLL10.2", "MBL14", "CLL4.1", "CLL3.1", "CLL10.1", "CLL1.2", "CLL6.2", "CLL5.1", "MBL3","MBL6", "CLL8.2", "CLL9.2", "CLL9.1", "MBL7", "CLL3.2", "CLL2.1", "MBL10","CLL6.1", "MBL12", "MBL11", "MBL1", "CLL5.2", "MBL2", "CLL7.1",   "MBL8", "CLL2.2", "CLL1.1", "MBL4", "MBL15")),
        paste0("LC-MBL Cells_", c("MBL15", "MBL13", "MBL14", "MBL11", "MBL12", "MBL10")),
        paste0("HC-MBL Cells_", c("CLL8.1", "CLL9.1", "MBL9", "CLL10.1", "CLL1.1", "CLL2.1", "CLL5.1", "MBL8", "CLL4.1", "MBL7", "MBL3", "MBL1", "MBL2", "CLL7.1", "MBL6", "CLL3.1")),
        paste0("CLL Cells_", c("CLL10.2", "CLL9.2", "CLL5.2", "CLL2.2", "CLL1.2", "CLL7.2", "CLL4.2", "CLL3.2", "MBL4", "CLL8.2", "CLL6.2"))
      )
    )
  ) %>%
  arrange(B_like_annotation_label)


# map cell type annotation to this cell barcode
# map colour to barcode
barcodes <- colnames(zscore_matrix_hm_input) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = CLLProj_Blike$cellNames,
    "B_like_annotation" = CLLProj_Blike@cellColData$B_like_annotation
  )) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
      B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
      B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
    label = factor(label, levels = c("MBL13", "MBL11", "MBL12", "MBL15", "MBL14", "MBL10", "CLL6.2", "CLL1.2", "CLL7.2", "CLL2.2", "CLL5.2", "CLL8.2", "CLL4.2", "CLL10.2", "CLL9.2", "CLL3.2", "MBL4"))
  ) %>%
  arrange(B_like_annotation, label) %>%
  left_join(tibble(
    "B_like_annotation" = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"),
    "Colour" = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5")
  ))

barcodes2annotation <- barcodes %>%
  dplyr::select(label_cellbarcode, B_like_annotation) %>%
  dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
  deframe()

barcodes2plottingid <- barcodes %>%
  left_join(tibble(
    "label_cellbarcode" = colnames(zscore_matrix_hm_input),
    "plotting_id" = label2plotting_id[colnames(zscore_matrix_hm_input) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)]
  )) %>%
  dplyr::select(label_cellbarcode, plotting_id) %>%
  dplyr::mutate(plotting_id = factor(plotting_id, levels = paste0("P", 1:24))) %>%
  deframe()

colours <- list(
  "Cell type" = barcodes %>%
    dplyr::select(B_like_annotation, Colour) %>%
    unique() %>%
    deframe(),
  "Patient" = plottingid2colours
)

anno_df <- tibble(
  "Cell type" = barcodes2annotation[colnames(zscore_matrix_hm_input)],
  "Patient" = barcodes2plottingid[colnames(zscore_matrix_hm_input)]
)

column_ha <- HeatmapAnnotation(
  df = anno_df,
  col = colours,
  show_legend = FALSE,
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8), 
  border = TRUE
)

hm_LC_CLL <- Heatmap(as.matrix(zscore_matrix_hm_input[comp_genes, ]),
  column_order = barcodes_ordered %>% pull(label_cellbarcode),
  # row_order = c("CEBPZ", "PBX3", "NFATC2", "NFATC3", "NFYA", "NFYB", "NFYC", "EBF1", "BACH2", "NFE2", "MAFF", "MAFK", "FOS", "FOSB", "FOSL1", "FOSL2", "JDP2", "BATF", "JUN", "JUNB", "JUND", "BACH1", "NFE2L2", "NFKB1", "NFKB2", "REL", "RELA", "SMARCC1"),
  top_annotation = column_ha,
  col = colorRamp2(breaks = c(-quantile(as.matrix(zscore_matrix_hm_input[comp_genes, ]), 0.93), 0, quantile(as.matrix(zscore_matrix_hm_input[comp_genes, ]), 0.93)), c("#0c544c", "white", "#783f0b")),
  na_col = "black",
  use_raster = TRUE,
  cluster_rows = TRUE,
  raster_by_magick = TRUE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  show_row_names = TRUE,
  show_row_dend = TRUE,
  show_column_dend = FALSE,
  show_heatmap_legend = TRUE,
  row_title_side = "left",
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),
  row_title = "TF motif",
  row_title_gp = gpar(fontsize = 8),
  border_gp = gpar(col = "black", lty = 1),
  heatmap_legend_param = list(
    title = "Z-score",
    direction = "horizontal",
    legend_direction = "horizontal",
    legend_gp = gpar(
      fontsize = 8,
      fontface = "plain"
    )
  )
)

draw(hm_LC_CLL, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")

hm_LC_CLL_grob <- grid::grid.grabExpr(draw(hm_LC_CLL, heatmap_legend_side = "bottom"))
plot_grid(hm_LC_CLL_grob, nroow=1, ncol=1)
saveRDS(hm_LC_CLL_grob, file = file.path(RDS_dir_Blike, "TF_enrichment_B_LC_MBL_CLL_marker_heatmap.RDS"))
```

## Downsampled heatmap

```{r}
# remove low quality and multiome samples from  zscore matrix
zscore_matrix_hm_input <- zscore_matrix_Blike[, colnames(zscore_matrix_Blike) %>% 
                                                keep(~ !str_starts(., "CLL6.") & !str_starts(., "CLL2.") & !str_starts(., "CLL5."))]

rownames(zscore_matrix_hm_input) <- rownames(zscore_matrix_hm_input) %>%
  str_split(string = , pattern = "_") %>%
  map_chr(~ .[[1]])
```

```{r}
# find out how many cells there are per celltype
#4683 cells are physiologic B cells
barcodes_ordered <- colnames(zscore_matrix_hm_input) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = CLLProj_Blike$cellNames,
    "B_like_annotation" = CLLProj_Blike@cellColData$B_like_annotation
  )) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC_MBL_B_Cells" ~ "LC-MBL Cells",
      B_like_annotation == "HC_MBL_B_Cells" ~ "HC-MBL Cells",
      B_like_annotation == "CLL_B_Cells" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells")),
    B_like_annotation_label = paste0(B_like_annotation, "_", label),
    B_like_annotation_label = factor(B_like_annotation_label,
      levels = c(
        paste0("Physiologic B Cells_", c("CLL8.1", "CLL7.2", "MBL9", "MBL13", "CLL10.2", "MBL14", "CLL4.1", "CLL3.1", "CLL10.1", "CLL1.2", "CLL6.2", "CLL5.1", "MBL3","MBL6", "CLL8.2", "CLL9.2", "CLL9.1", "MBL7", "CLL3.2", "CLL2.1", "MBL10","CLL6.1", "MBL12", "MBL11", "MBL1", "CLL5.2", "MBL2", "CLL7.1",   "MBL8", "CLL2.2", "CLL1.1", "MBL4", "MBL15")),
        paste0("LC-MBL Cells_", c("MBL15", "MBL13", "MBL14", "MBL11", "MBL12", "MBL10")),
        paste0("HC-MBL Cells_", c("CLL8.1", "CLL9.1", "MBL9", "CLL10.1", "CLL1.1", "CLL2.1", "CLL5.1", "MBL8", "CLL4.1", "MBL7", "MBL3", "MBL1", "MBL2", "CLL7.1", "MBL6", "CLL3.1")),
        paste0("CLL Cells_", c("CLL10.2", "CLL9.2", "CLL5.2", "CLL2.2", "CLL1.2", "CLL7.2", "CLL4.2", "CLL3.2", "MBL4", "CLL8.2", "CLL6.2"))
      )
    )
  ) %>%
  arrange(B_like_annotation_label)
```

```{r downsample_barcodes}
# downsample cell barcodes to equal number per celltype 
downsample_cells <- function(df, max_cells) {
  set.seed(123)  # Set seed for reproducibility
  num_patients <- n_distinct(df$label)
  cells_per_patient <- max_cells %/% num_patients  # integer division to get the cells per patient
  remaining_cells <- max_cells %% num_patients  # remaining cells after even distribution

  # Sample cells per patient with even distribution
  downsampled <- df %>%
    group_by(label) %>%
    sample_n(size = min(cells_per_patient, n()), replace = FALSE) %>%
    ungroup()

  # If there are remaining cells, sample them from the already downsampled data
  if (remaining_cells > 0) {
    remaining <- df %>%
      anti_join(downsampled, by = "label_cellbarcode") %>%
      sample_n(size = min(remaining_cells, n()), replace = FALSE)

    downsampled <- bind_rows(downsampled, remaining)
  }

  return(downsampled)
}


# downsampling number for healthy B is so high, that really all cells are considered and no downsampling is performed in this case 
celltype_barcodes <- map2(c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"), c(200000, 4683,4683,4683),
      ~ barcodes_ordered %>% 
        dplyr::filter(B_like_annotation == .x) %>% 
        downsample_cells(max_cells = .y)) %>% 
  set_names(c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))
```

```{r cluster_patients}
# per patient calculate mean value per TF and run hierarchical clustering 
clustering_input <- celltype_barcodes %>%
  map(~ zscore_matrix_hm_input[comp_genes, .$label_cellbarcode] %>%
    as_tibble(rownames = NA) %>%
    rownames_to_column("TF") %>%
    pivot_longer(cols = -TF, names_to = "label_cellbarcode", values_to = "z_score") %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#") %>%
    group_by(TF, label) %>%
    summarise("mean_z_score" = mean(`z_score`)) %>%
    ungroup() %>%
    pivot_wider(id_cols = label, names_from = TF, values_from = mean_z_score) %>%
    column_to_rownames(var = "label")) %>% 
  set_names(names(celltype_barcodes))

set.seed(20)
distance_mat <- clustering_input %>% 
  map(~dist(., method="euclidean"))
clusters <- distance_mat %>% 
  map(~hclust(., method = "average"))

patients_ordered <- clustering_input %>% 
  map2(clusters, ~rownames(.x)[.y %>% as.dendrogram() %>% 
                                 order.dendrogram()])
```

```{r column2_annotations}
celltype_barcodes_ordered <- celltype_barcodes %>% 
  map2(patients_ordered, ~dplyr::mutate(.x, label=factor(label, levels=.y)) %>% 
  arrange(label))

barcodes2annotation <- celltype_barcodes %>% 
  map(~dplyr::select(., label_cellbarcode, B_like_annotation) %>%
  dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
  deframe())

barcodes2label <- celltype_barcodes %>% 
  map2(patients_ordered, ~left_join(.x, tibble(
    "label_cellbarcode" = colnames(zscore_matrix_hm_input[, .x$label_cellbarcode]),
    "label" = colnames(zscore_matrix_hm_input[, .x$label_cellbarcode]) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)
  )) %>%
  dplyr::select(label_cellbarcode, label) %>%
  dplyr::mutate(label = factor(label, levels = .y)) %>%
  deframe()
)

barcodes2IGHV <- celltype_barcodes %>% 
  map2(patients_ordered, ~left_join(.x, metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "IGHV")) %>%
  dplyr::mutate(label = factor(label, levels = .y),
                IGHV= if_else(is.na(IGHV), "NA", IGHV)) %>% 
  dplyr::select(label_cellbarcode, IGHV) %>% 
    deframe()
  )

barcode2processing_date <- celltype_barcodes %>% 
  map2(patients_ordered, ~left_join(.x, metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "processing_date")) %>%
  dplyr::mutate(label = factor(label, levels = .y),
                processing_date=as.character(processing_date)) %>% 
  dplyr::select(label_cellbarcode, processing_date) %>% 
    deframe()
  )

barcode2technology <- celltype_barcodes %>% 
  map2(patients_ordered, ~left_join(.x, metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "technology")) %>%
  dplyr::mutate(label = factor(label, levels = .y)) %>% 
  dplyr::select(label_cellbarcode, technology) %>% 
    deframe()
  )

barcode2sequencing_date <- celltype_barcodes %>% 
  map2(patients_ordered, ~left_join(.x, metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "sequencing_date")) %>%
  dplyr::mutate(label = factor(label, levels = .y),
                sequencing_date=as.character(sequencing_date)) %>% 
  dplyr::select(label_cellbarcode, sequencing_date) %>% 
    deframe()
  ) 

barcode2sex <- celltype_barcodes %>% 
  map2(patients_ordered, ~left_join(.x, metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "sex")) %>%
  dplyr::mutate(label = factor(label, levels = .y)) %>% 
  dplyr::select(label_cellbarcode, sex) %>% 
    deframe()
  ) 

"barcodes2sequencing_date_?" <- celltype_barcodes %>% 
  map2(patients_ordered, ~left_join(.x, metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "sequencing_date_?")) %>%
  dplyr::mutate(label = factor(label, levels = .y),
                sequencing_date=as.character(`sequencing_date_?`)) %>% 
  dplyr::select(label_cellbarcode, `sequencing_date_?`) %>% 
    deframe()
  ) 


label2colours <- c(my_paired_colours_named, my_unpaired_colours_named)

colours <- list(
  "Cell type" = c("Physiologic B Cells" ="darkred", "LC-MBL Cells"="#fcbda4", "HC-MBL Cells"="#FC8D62", "CLL Cells"="#66C2A5"),
  "Patient" = label2colours,
  "IGHV" = c("mutated" = "steelblue", "unmutated"="orange", "NA"="lightgrey"),
  "processing_date"=c("2022-01-10"="#1B9E77" , "2022-10-20"="#D95F02",  
                      "2022-11-04"="#7570B3",  "2023-03-24"="#66A61E" , 
                      "2024-03-05"="#A6761D"),
  "technology"=c("v2 scATAC-seq"="#666666", "multiome kit"="#E7298A"),
  "sequencing_date"=c("2022-02-26"="#E6AB02" ,  "2022-11-18"="#7FC97F", "2023-05-02"="#FFFF99", "2024-03-18"="#386CB0"),
  "sequencing_date_?"=c("2022-01-14"="midnightblue", "2022-10-20"="steelblue4", "2022-10-26"="magenta4", "2022-10-27"="purple4", "2024-03-18"="#386CB0", "2023-05-02"="violet", "2022-11-18"="plum4"),
  "sex"=c("male"="lightblue", "female"="lightpink"))

anno_df <- list(
  barcodes2annotation,
  barcodes2label,
  barcodes2IGHV,
  barcode2processing_date,
  barcode2technology,
  barcode2sequencing_date,
  barcode2sequencing_date,
  barcode2sex,
  celltype_barcodes
) %>%
  pmap(~ tibble(
    "Cell type" = ..1[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "Patient" = ..2[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "IGHV" = ..3[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "processing_date"=..4[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "technology"=..5[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "sequencing_date"=..6[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "sequencing_date_?"=..7[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "sex"=..8[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
  ))


column_ha <- anno_df %>% 
  map(~ HeatmapAnnotation(
  df = .,
  col = colours,
  show_legend = FALSE,
  annotation_name_side = "left", 
  border = TRUE,
  annotation_name_gp = gpar(fontsize = 8)
)) 
```

```{r}
hm <- list(celltype_barcodes, 
           column_ha,
           celltype_barcodes_ordered %>% map(~pull(., label_cellbarcode))) %>% 
  pmap(~Heatmap(as.matrix(zscore_matrix_hm_input[comp_genes, ..1$label_cellbarcode]),
                  col = colorRamp2(breaks = c(-quantile(as.matrix(zscore_matrix_hm_input[comp_genes, ]), 0.93), 0, quantile(as.matrix(zscore_matrix_hm_input[comp_genes, ]), 0.93)), c("#0c544c", "white", "#783f0b")),
              top_annotation = ..2,
              column_order = ..3,
              # previous row order:
              # row_order = c("CEBPZ", "PBX3", "NFATC2", "NFATC3", "NFYA", "NFYB", "NFYC", "EBF1", "BACH2",
              #               "NFE2", "MAFF", "MAFK", "FOS", "FOSB", "FOSL1", "FOSL2", "JDP2", "BATF", "JUN",
              #               "JUNB", "JUND", "BACH1", "NFE2L2", "NFKB1", "NFKB2", "REL", "RELA", "SMARCC1"),
              use_raster = TRUE,
              cluster_rows = TRUE,
              cluster_columns = FALSE,
              raster_by_magick = TRUE,
              show_column_names = FALSE, 
              show_heatmap_legend = FALSE,
              na_col = "black",
              row_title_side = "left",
              row_names_side = "left",
              row_names_gp = gpar(fontsize = 8),
              row_title_gp = gpar(fontsize = 8),                                 ,
              border_gp = gpar(col = "black", lty = 1),
              heatmap_legend_param = list(
                title = "Z-score",
                direction = "horizontal",
                legend_direction = "vertical",
                legend_gp = gpar(
                  fontsize = 8,
                  fontface = "plain"))))

```

```{r, fig.width=20, fig.height=9}
hm$`Physiologic B Cells` + hm$`LC-MBL Cells` + hm$`HC-MBL Cells` + hm$`CLL Cells`
hm_LC_CLL_grob_downsampled <- grid::grid.grabExpr(draw(hm$`Physiologic B Cells` + hm$`LC-MBL Cells` + hm$`HC-MBL Cells` + hm$`CLL Cells`))
saveRDS(hm, file = file.path(RDS_dir_Blike, "TF_enrichment_B_LC_MBL_CLL_marker_heatmap_downsampled_batch_effect.RDS"))
```

#### Without column subsetting
```{r}
celltype_barcodes_unlist <- celltype_barcodes %>% 
  bind_rows()

barcodes2annotation_unlist <- celltype_barcodes_unlist %>% 
  dplyr::select(label_cellbarcode, B_like_annotation) %>%
  dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
  deframe()

barcodes2label_unlist <- celltype_barcodes_unlist %>% 
  left_join(tibble(
    "label_cellbarcode" = colnames(zscore_matrix_hm_input[, .$label_cellbarcode]),
    "label" = colnames(zscore_matrix_hm_input[, .$label_cellbarcode]) %>%
      as_tibble() %>%
      dplyr::rename("label_cellbarcode" = value) %>%
      tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE) %>%
      pull(label)
  )) %>%
  dplyr::select(label_cellbarcode, label) %>%
  deframe()


barcodes2IGHV_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "IGHV")) %>%
  dplyr::mutate(IGHV = if_else(is.na(IGHV), "NA", IGHV)) %>%
  dplyr::select(label_cellbarcode, IGHV) %>%
  deframe()

barcode2processing_date_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "processing_date")) %>%
  dplyr::mutate(
    processing_date = as.character(processing_date)
  ) %>%
  dplyr::select(label_cellbarcode, processing_date) %>%
  deframe()


barcode2technology_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "technology")) %>%
  dplyr::select(label_cellbarcode, technology) %>%
  deframe()

barcode2sequencing_date_unlist <- celltype_barcodes_unlist %>%
  left_join(metadata %>%
    dplyr::filter(data_type == "ATAC") %>%
    dplyr::select("label", "sequencing_date")) %>%
  dplyr::mutate(
    sequencing_date = as.character(sequencing_date)
  ) %>%
  dplyr::select(label_cellbarcode, sequencing_date) %>%
  deframe()
  

barcode2sex_unlist <- celltype_barcodes_unlist %>% 
  left_join( metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "sex")) %>%
  dplyr::select(label_cellbarcode, sex) %>% 
    deframe()
 

"barcodes2sequencing_date_?_unlist" <- celltype_barcodes_unlist %>% 
  left_join( metadata %>%
                                      dplyr::filter(data_type=="ATAC") %>% 
                                      dplyr::select("label", "sequencing_date_?")) %>%
  dplyr::mutate(
                sequencing_date=as.character(`sequencing_date_?`)) %>% 
  dplyr::select(label_cellbarcode, `sequencing_date_?`) %>% 
    deframe()
  


label2colours <- c(my_paired_colours_named, my_unpaired_colours_named)

colours <- list(
  "Cell type" = c("Physiologic B Cells" ="darkred", "LC-MBL Cells"="#fcbda4", "HC-MBL Cells"="#FC8D62", "CLL Cells"="#66C2A5"),
  "Patient" = label2colours,
  "IGHV" = c("mutated" = "steelblue", "unmutated"="orange", "NA"="lightgrey"),
  "processing_date"=c("2022-01-10"="#1B9E77" , "2022-10-20"="#D95F02",  
                      "2022-11-04"="#7570B3",  "2023-03-24"="#66A61E" , 
                      "2024-03-05"="#A6761D"),
  "technology"=c("v2 scATAC-seq"="#666666", "multiome kit"="#E7298A"),
  "sequencing_date"=c("2022-02-26"="#E6AB02" ,  "2022-11-18"="#7FC97F", "2023-05-02"="#FFFF99", "2024-03-18"="#386CB0"),
  "sequencing_date_?"=c("2022-01-14"="midnightblue", "2022-10-20"="steelblue4", "2022-10-26"="magenta4", "2022-10-27"="purple4", "2024-03-18"="#386CB0", "2023-05-02"="violet", "2022-11-18"="plum4"),
  "sex"=c("male"="lightblue", "female"="lightpink"))

anno_df_unlist <- list(
  list(barcodes2annotation_unlist),
  list(barcodes2label_unlist),
  list(barcodes2IGHV_unlist),
  list(barcode2processing_date_unlist),
  list(barcode2technology_unlist),
  list(barcode2sequencing_date_unlist),
  list(barcode2sequencing_date_unlist),
  list(barcode2sex_unlist),
  list(celltype_barcodes_unlist)
) %>%
  pmap(~ tibble(
    "Cell type" = ..1[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "Patient" = ..2[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "IGHV" = ..3[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "processing_date"=..4[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "technology"=..5[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "sequencing_date"=..6[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "sequencing_date_?"=..7[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
    "sex"=..8[colnames(zscore_matrix_hm_input[, ..9$label_cellbarcode])],
  ))


column_ha_unlist <- anno_df_unlist %>% 
  map(~ HeatmapAnnotation(
  df = .,
  col = colours,
  show_legend = FALSE,
  annotation_name_side = "left",
  border = TRUE,
  annotation_name_gp = gpar(fontsize = 8)
)) 
```


```{r, fig.width=5, fig.height=9}
hm_unsort <- list(list(celltype_barcodes_unlist), 
           column_ha_unlist) %>% 
  pmap(~Heatmap(as.matrix(zscore_matrix_hm_input[comp_genes, ..1$label_cellbarcode]),
                  col = colorRamp2(breaks = c(-quantile(as.matrix(zscore_matrix_hm_input[comp_genes, ]), 0.93), 0, quantile(as.matrix(zscore_matrix_hm_input[comp_genes, ]), 0.93)), c("#0c544c", "white", "#783f0b")),
              top_annotation = ..2,
              use_raster = TRUE,
              cluster_rows = TRUE,
              cluster_columns = TRUE,
              raster_by_magick = TRUE,
              show_column_names = FALSE, 
              show_heatmap_legend = FALSE,
              na_col = "black",
              row_title_side = "left",
              row_names_side = "left",
              row_names_gp = gpar(fontsize = 8),
              row_title_gp = gpar(fontsize = 8),
                                 border_gp = gpar(col = "black", lty = 1),

              heatmap_legend_param = list(
                title = "Z-score",
                direction = "horizontal",
                legend_direction = "vertical",
                legend_gp = gpar(
                  fontsize = 8,
                  fontface = "plain"))))
```

```{r, fig.width=10, fig.height=5}
hm_unsort_grob <- grid::grid.grabExpr(draw(hm_unsort[[1]]))
plot_grid(hm_unsort_grob, nrow=1, ncol=1)
saveRDS(hm_unsort, file = file.path(RDS_dir_Blike, "TF_enrichment_B_LC_MBL_CLL_marker_heatmap_downsampled_batch_effect_unsorted.RDS"))
```

-->A huge batch effect based on sequencing date was observed, peaks will be normalised for the bias

## Downsampled heatmap of bias corrected TF motifs
Switched here to batches script to do combat across batches
```{r}
# per patient calculate mean value per TF and run hierarchical clustering 
clustering_input <- celltype_barcodes %>%
  map(~ zscore_matrix_hm_input[all_TS, .$label_cellbarcode] %>%
    as_tibble(rownames = NA) %>%
    rownames_to_column("TF") %>%
    pivot_longer(cols = -TF, names_to = "label_cellbarcode", values_to = "z_score") %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#") %>%
    group_by(TF, label) %>%
    summarise("mean_z_score" = mean(`z_score`)) %>%
    ungroup() %>%
    pivot_wider(id_cols = label, names_from = TF, values_from = mean_z_score) %>%
    column_to_rownames(var = "label")) %>% 
  set_names(names(celltype_barcodes))

set.seed(20)
distance_mat <- clustering_input %>% 
  map(~dist(., method="euclidean"))
clusters <- distance_mat %>% 
  map(~hclust(., method = "average"))

patients_ordered <- clustering_input %>% 
  map2(clusters, ~rownames(.x)[.y %>% as.dendrogram() %>% 
                                 order.dendrogram()])
```

```{r, plot_heatmap}
hm <- list(celltype_barcodes, 
           column_ha,
           celltype_barcodes_ordered %>% 
             map(~pull(., label_cellbarcode))) %>% 
  pmap(~Heatmap(as.matrix(zscore_matrix_hm_input[all_TS, ..1$label_cellbarcode]),
                  col = colorRamp2(breaks = c(-quantile(as.matrix(zscore_matrix_hm_input[all_TS, ]), 0.93), 0, quantile(as.matrix(zscore_matrix_hm_input[all_TS, ]), 0.93)), c("#0c544c", "white", "#783f0b")),
              top_annotation = ..2,
              column_order = ..3,
              use_raster = TRUE,
              cluster_rows = TRUE,
              cluster_columns = FALSE,
              raster_by_magick = TRUE,
              show_column_names = FALSE, 
              show_heatmap_legend = FALSE,
              na_col = "black",
              row_title_side = "left",
              row_names_side = "left",
              row_names_gp = gpar(fontsize = 8),
              row_title_gp = gpar(fontsize = 8), 
              heatmap_legend_param = list(
                title = "Deviation",
                direction = "horizontal",
                legend_direction = "vertical",
                legend_gp = gpar(
                  fontsize = 8,
                  fontface = "plain"))))

```

```{r, fig.width=20, fig.height=9}
hm$`Physiologic B Cells` + hm$`LC-MBL Cells` + hm$`HC-MBL Cells` + hm$`CLL Cells`
hm_LC_CLL_grob_downsampled <- grid::grid.grabExpr(draw(hm$`Physiologic B Cells` + hm$`LC-MBL Cells` + hm$`HC-MBL Cells` + hm$`CLL Cells`))
cowplot::plot_grid(hm_LC_CLL_grob_downsampled)
saveRDS(hm_LC_CLL_grob_downsampled, file = file.path(RDS_dir_Blike, "TF_enrichment_B_LC_MBL_CLL_marker_heatmap_downsampled_batch_corrected.RDS"))
```
