---
title: "Differential Chromatin Accessibility Analysis Progression Analysis per batch"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

# GOAL
The aim of this script is to find out if there are recurrent differentially accessible chromatin regions between 5-year stable HC-MBL samples and patients which progressed to CLL before. Analysis was conducted across patients with respect to their sequencing batches (batcheffect was found). 

# Load libraries
```{r Load_libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(ArchR)
library(readxl)
library(ggrepel)
library(tidyverse)
library(glue)
library(ggpubr)
library(SeuratDisk)
library(Seurat)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Cairo)
library(sva)
library(scales)
library(chromVAR)
library(patchwork)
library(circlize)
library(ComplexHeatmap)
library(magick)
register(MulticoreParam(8))
```

```{r Run_parameters}
set.seed(1)
addArchRThreads(threads = 8)
addArchRGenome("hg38")
```

## Set paths

```{r }
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "DCA_analysis")

plotting_dir <- file.path(analysis_dir, "Plots")
RDS_dir <- file.path(analysis_dir, "RDS_objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")

DESeq2_analysis_dir <- file.path(result_dir, "DESeq2_analysis")

ArchR_project_Blike <- "ArchR_35_samples/ArchR_Blike"
ArchR_dir_Blike <- file.path(result_dir, ArchR_project_Blike)

ArchR_project_MBL_Progression <- "ArchR_MBL_Progression"
ArchR_dir_MBL_Progression <- file.path(project_dir_ext, ArchR_project_MBL_Progression)
```

```{r}
# local
source(file.path(project_dir, "scripts/publication_theme.R"))
source(file.path(project_dir, "scripts/heatmap_plotting_functions.R"))
```

## Metadata

```{r }
metadata <- read_tsv(file = file.path(data_dir, "20260309_Tidy_MBL_CLL_metadata.tsv"))
cohort_overview <- read_tsv(file.path(DESeq2_analysis_dir, "tsv_data/20251017_MBL_progression_cohort_overview.tsv"))

patient_id2plotting_id <- metadata %>%
  dplyr::select(patient_id, plotting_id) %>%
  unique() %>%
  deframe()

label2plotting_id <- metadata %>%
  dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, plotting_id) %>%
  deframe()

label2IGHV_inferred <- metadata %>% 
    dplyr::filter(data_type == "ATAC") %>%
  dplyr::select(label, IGHV_inferred) %>% 
  deframe()
```

## Load ArchRproj to subset from
```{r}
CLLProj_Blike <- loadArchRProject(ArchR_dir_Blike)
```

## Subset ArchR project by batch
```{r}
# Add sequencing date annotation
five_year_MBL_stability_annotation <- tibble("label_cellbarcode" = CLLProj_Blike$cellNames) %>%
  tidyr::separate(label_cellbarcode, into = c("label", "cellbarcode"), sep = "#", remove = F) %>%
  tidyr::separate(label, into = c("patient_id", NA), sep = "\\.", remove = F) %>%
    dplyr::left_join(cohort_overview %>%
                     dplyr::select(patient_id, five_year_MBL_stability)) %>%
  dplyr::select(label_cellbarcode, five_year_MBL_stability) %>%
  deframe()

CLLProj_Blike <- addCellColData(
  ArchRProj = CLLProj_Blike,
  data = five_year_MBL_stability_annotation,
  cells = names(five_year_MBL_stability_annotation),
  name = "five_year_MBL_stability"
)

# Add sequencing date annotation
seqdate_annotation <- tibble("label_cellbarcode" = CLLProj_Blike$cellNames) %>%
  tidyr::separate(label_cellbarcode, into = c("label", "cellbarcode"), sep = "#", remove = F) %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "ATAC") %>% dplyr::select(label, sequencing_date)) %>%
  dplyr::mutate(sequencing_date = as.numeric(as.factor(sequencing_date))) %>%
  dplyr::select(label_cellbarcode, sequencing_date) %>%
  deframe()

CLLProj_Blike <- addCellColData(
  ArchRProj = CLLProj_Blike,
  data = seqdate_annotation,
  cells = names(seqdate_annotation),
  name = "sequencing_date")

# to include all patients from the scRNA pseudobulk analysis would result in two seperate batch comparisons with 2 vs 3 and 3 vs 6 samples, which is why I will only use samples from the second batch
idxBatch_MBL_Progression <- BiocGenerics::which(CLLProj_Blike@cellColData$sequencing_date == 2 & CLLProj_Blike@cellColData$patient_id %in% cohort_overview$patient_id & CLLProj_Blike@cellColData$B_like_annotation =="HC_MBL_B_Cells")

cellsSample_MBL_Progression <- CLLProj_Blike$cellNames[idxBatch_MBL_Progression]

CLLProj_MBL_Progression <-subsetArchRProject(
  ArchRProj = CLLProj_Blike,
  cells = cellsSample_MBL_Progression,
  outputDirectory = ArchR_dir_MBL_Progression,
  dropCells = TRUE)

CLLProj_MBL_Progression <- loadArchRProject(ArchR_dir_MBL_Progression)
```

### Add pseudobulk replicates

```{r}
CLLProj_MBL_Progression <-addGroupCoverages(
    ArchRProj = CLLProj_MBL_Progression,
    groupBy = "five_year_MBL_stability" )
```

### Peak calling

```{r}
CLLProj_MBL_Progression <- addReproduciblePeakSet(
    ArchRProj = CLLProj_MBL_Progression,
    groupBy = "five_year_MBL_stability",
    pathToMacs2 = pathToMacs3,
  )

saveArchRProject(
      ArchRProj = CLLProj_MBL_Progression,
      outputDirectory = ArchR_dir_MBL_Progression,
      load = FALSE,
      overwrite = TRUE)
```

```{r}
CLLProj_MBL_Progression <- addPeakMatrix(CLLProj_MBL_Progression)

saveArchRProject(
      ArchRProj = CLLProj_MBL_Progression,
      outputDirectory = ArchR_dir_MBL_Progression,
      load = FALSE,
      overwrite = TRUE)

getAvailableMatrices(CLLProj_MBL_Progression)
```

### Pairwise peaks

```{r }
markerTest_MBL_Progression <- getMarkerFeatures(
    ArchRProj = CLLProj_MBL_Progression,
    useMatrix = "PeakMatrix",
    groupBy = "five_year_MBL_stability",
    testMethod = "wilcoxon",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    useGroups = "progressive",
    bgdGroups = "unprogressive"
  )
```

### Volcano plots Pairwise testing

```{r}
volcanoplot_peaks <- plotMarkers(
    seMarker = markerTest_MBL_Progression,
    name = "progressive",
    cutOff = "FDR <= 0.1 & abs(Log2FC) >= 1",
    plotAs = "Volcano",
    rastr = TRUE
  ) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.1)), color = "grey", linetype = "dashed") +
    geom_point(size = 1) +
    scale_y_continuous(
      name = "-Log10(FDR)",
      breaks = seq(0, 5, 1),
      limits = c(0, 5),
      oob = scales::oob_squish_infinite) +
    scale_colour_manual(
      name = "Differentially accessible peaks",
      values = c("Down-Regulated" = "slategray3", 
                 "Up-Regulated" = "lightpink3", 
                 "Not-Differential" = "#D3D3D3"),
      labels = c(
        "Down-Regulated" = "unprogressive HC-MBL",
        "Up-Regulated" = "progressive HC-MBL",
        "Not-Differential" = "not significant")) +
    theme_pub()+
  theme(plot.title = element_blank(),
        legend.position = "bottom",
        legend.direction = "vertical")

volcanoplot_peaks

write_tsv(volcanoplot_peaks$data, file.path(tsv_dir, "Peaks_MBL_Progression_Volcanoplot.tsv"))
```

### Add TF Mofif Annotations

```{r}
CLLProj_MBL_Progression <- addMotifAnnotations(
    ArchRProj = CLLProj_MBL_Progression,
    motifSet = "cisbp",
    annoName = "Motif")
```

```{r}
# background peaks added
CLLProj_MBL_Progression <- addBgdPeaks(CLLProj_MBL_Progression, 
                                      method = "chromVAR")
```

```{r}
# motif deviation matrix added
CLLProj_MBL_Progression <- addDeviationsMatrix(
    ArchRProj = CLLProj_MBL_Progression,
    bgdPeaks = getBgdPeaks(CLLProj_MBL_Progression,
                           method = "chromVAR"),
    peakAnnotation = "Motif", 
    force = TRUE)
```

```{r}
plotVarDev <- getVarDeviations(CLLProj_MBL_Progression,
    name = "MotifMatrix",
    plot = TRUE)

plotVarDev
```

### Get Matrix from project

```{r}
MotifMatrix_tmp_MBL_Progression <-getMatrixFromProject(
    ArchRProj = CLLProj_MBL_Progression,
    useMatrix = "MotifMatrix",
    useSeqnames = NULL,
    verbose = TRUE,
    binarize = FALSE)

zscore_matrix_MBL_Progression <-as.matrix(assays(MotifMatrix_tmp_MBL_Progression)$z) %>%
    as.data.frame()
```

```{r}
condition <- tibble(
  "label_cellbarcode" = CLLProj_MBL_Progression$cellNames,
  "five_year_MBL_stability" = CLLProj_MBL_Progression$five_year_MBL_stability
)

progressive_cells <- dplyr::filter(condition, five_year_MBL_stability == "progressive") %>%
  pull(label_cellbarcode)

unprogressive_cells <- dplyr::filter(condition, five_year_MBL_stability == "unprogressive") %>%
  pull(label_cellbarcode)

# # calculate data for volcano plot
input_zscore_matrix <- zscore_matrix_MBL_Progression

# If no NA values are found in zcore matrix calculate statistics
condition1 <- unprogressive_cells
condition2 <- progressive_cells

volcano.data <- data.frame()
i <- 0
for (TF in rownames(input_zscore_matrix)) {
  i <- i + 1
  print(i)
  print(TF)
  if (length(which(is.na(input_zscore_matrix[TF, ]))) == 0) {
    p.value <- wilcox.test(as.numeric(input_zscore_matrix[TF, condition2]), as.numeric(input_zscore_matrix[TF, condition1]))$p.value
    diff.accessibility <- median(as.numeric(input_zscore_matrix[TF, condition2])) - median(as.numeric(input_zscore_matrix[TF, condition1]))
    volcano.data <- rbind(volcano.data, data.frame(TF = TF, p.value = p.value, diff.accessibility = diff.accessibility))
  }
}

volcano.data$p.adj <- p.adjust(volcano.data$p.value)
volcano.data$p.adj[which(volcano.data$p.adj == 0)] <- min(volcano.data$p.adj[which(volcano.data$p.adj != 0)])

volcano.data$diff.accessibility <- as.numeric(volcano.data$diff.accessibility)
volcano.data$p.adj <- as.numeric(volcano.data$p.adj)


volcano.data$significant <- ifelse(abs(volcano.data$diff.accessibility) > 1.5 & volcano.data$p.adj <= 0.05, "significant", "not significant")
volcano.data <- volcano.data %>%
  dplyr::mutate(
    "significant" = case_when(
      p.adj >= 0.05 ~ "not significant",
      abs(diff.accessibility) < 1.5 ~ "not significant",
      diff.accessibility > 1.5 & p.adj < 0.05 ~ "MBL",
      diff.accessibility < -1.5 & p.adj < 0.05 ~ "B"
    ),
    "label" = if_else(significant != "not significant", stringr::str_split_fixed(TF, pattern = "_", n = 2)[, 1], NA)
  )



saveRDS(volcano.data, file = file.path(RDS_dir, "volcano.data_MBL_Progression.RDS"))
```

### Volcanoplots Differential TF enrichment

```{r}
volcanoplot_MBL_Progression <- volcano.data %>%
  ggplot(aes(x = diff.accessibility, y = -log10(p.adj), colour = significant, label = label)) +
  geom_vline(xintercept = c(-1.5, 1.5), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(0.05), color = "grey", linetype = "dashed") +
  geom_point(size = 1) +
  geom_text_repel(
    size = 2.5,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40,
    colour = "black"
  ) +
  scale_x_continuous(
    name = "Difference median z-score",
    breaks = seq(-8, 8, 4),
    limits = c(-8, 8)
  ) +
  scale_y_continuous(
    name = "-log10(FDR)",
    limits = c(0, 300),
    expand = expansion(mult = c(0, 0.1)),
    oob = squish_infinite
  ) +
  scale_colour_manual(
    name = "Differential TF motif enrichment",
    values = c("upregulated" = "lightpink3", "downregulated" = "slategray3", "not significant" = "lightgrey"),
    labels = c("upregulated" = "progressive", "downregulated" = "unprogressive")
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )

write_tsv(volcanoplot_MBL_Progression$data, file.path(tsv_dir, "volcanoplot_MBL_Progression_data.tsv"))
```

```{r, fig.width=10, fig.height=5}
ggarrange(volcanoplot_peaks+labs(title="Peaks"), volcanoplot_MBL_Progression+labs(title="TF motif enrichment"),nrow=1, ncol=2, labels = "AUTO")
```

