---
title: "Differential Gene Expression analysis"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

#GOAL
The aim of this script is to find out if there are differentially expressed genes between LC-MBL, HC-MBL and CLL cases.

# Load libraries

```{r Load_libraries}
library(readr)
library(dplyr)
library(tibble)
library(ggpubr)
library(ggrepel)
library(Seurat)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
library(ggrastr)
```

# Set paths
```{r}
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "GEX")
result_dir <-file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "GEX_analysis")
plotting_dir <- file.path(analysis_dir, "plots")
tsv_dir <- file.path(analysis_dir, "tsv_data")
RDS_dir <- file.path(analysis_dir, "RDS_objects")
```

# Load  data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

Reference data set
```{r}
reference <- LoadH5Seurat(file.path(data_dir, "reference_genomes/pbmc_multimodal.h5seurat"))
```

Metadata

```{r Load_metadata}
metadata <- read_tsv(file = file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata_gex.tsv")) %>% 
    tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE)

patient_id2plotting_id <- metadata %>%
  dplyr::select(patient_id, plotting_id) %>%
  deframe()

label2plotting_id <- metadata %>% 
  dplyr::select(label, plotting_id) %>% 
  deframe()
```

GEX data
```{r define_readin_parameters}
samples <- metadata$label
library_names <- metadata %>%
  filter(label %in% samples) %>%
  pull(pool)
seurat_paths <- file.path(library_path, library_names, "filtered_feature_bc_matrix/")
```

```{r Create_seurat_object}
# remove indices when run all samples
gex_data <- seurat_paths %>%
  map(~ Read10X(data.dir = .)) %>%
  set_names(samples)

gex <- gex_data %>% 
  imap( ~ CreateSeuratObject(
  counts = .x, ,
  project = .y,
  min.cells = 3,
  min.features = 200
))
```

## Merge samples

```{r Merge_all_seurat_objects}
gex_merged_all <- merge(gex[[1]], 
                        y = gex[2:35], 
                        add.cell.ids = c(samples), 
                        project = "MBL_CLL_all")
```

```{r MtDNA_per_cell}
gex_merged_all[["percent.mt"]] <- PercentageFeatureSet(gex_merged_all, pattern = "^MT-")
```

```{r}
saveRDS(gex_merged_all, file = file.path(RDS_dir, "gex_merged_all.RDS"))
```

## QC plots

```{r fig.width=15}
# Visualize QC metrics as a violin plot: features = genes, counts=UMI
qc_plots <- gex_merged_all %>% 
  VlnPlot(features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
qc_plots

qc_plots %>% 
  ggsave(., filename = file.path(plotting_dir, "QC_counts_all.png"), 
         unit = "in", 
         width = 27, 
         height = 10)
```

```{r, fig.width=10,  fig.height=4}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(gex_merged_all, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(gex_merged_all, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
f_f_plot <- ggarrange(plot1, plot2, ncol = 2, nrow = 1)
f_f_plot

f_f_plot %>% 
  ggsave(., filename = file.path(plotting_dir, "QC_feature_scatterplot_all.png"), 
         unit = "in", 
         width = 10, 
         height = 4)
```

```{r Filter_cells}
gex_filtered_all <- subset(gex_merged_all, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 20)
```

## Scaling

```{r Scaling}
gex_scaled_all <- SCTransform(gex_filtered_all, verbose = TRUE)
saveRDS(gex_scaled_all, file = file.path(RDS_dir, "gex_scaled_all.RDS"))
```

```{r}
qc_counts <- left_join(gex_merged_all@meta.data[["orig.ident"]][gex_merged_all@meta.data[["orig.ident"]] !="MBL7dup"] %>% 
                         table() %>% 
                         as.tibble() %>% 
                         rename("pre_qc"=n), gex_scaled_all@meta.data[["orig.ident"]][gex_scaled_all@meta.data[["orig.ident"]] !="MBL7dup"] %>% 
                         table() %>% 
                         as.tibble() %>% 
                         rename("post_qc"=n)) %>% 
  dplyr::mutate(perc_change=100-post_qc/pre_qc*100) %>%
  rename("label"=".") %>% 
  dplyr::summarize("mean_perc_change"=mean(perc_change), 
                   "min_perc_change"=min(perc_change), 
                   "max_perc_change"=max(perc_change))
```

## LDR

```{r PCAs}
gex_scaled_all <- RunPCA(gex_scaled_all, features = VariableFeatures(object = gex_scaled_all))

# Examine and visualize PCA results a few different ways
print(gex_scaled_all[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r dot_plots}
dot_plot <- VizDimLoadings(gex_scaled_all, dims = 1:2, reduction = "pca") +
  theme_pub()
dot_plot %>% 
  ggsave(., filename = file.path(plotting_dir, "QC_PCA_DimLoadings_all.png"), 
         unit = "in", 
         width = 10, 
         height = 6)

dot_plot
```

```{r PCA_scatter}
pca <- DimPlot(gex_scaled_all, reduction = "pca", raster = FALSE) +
  theme_pub()

pca %>% 
  ggsave(., filename = file.path(plotting_dir, "PCA_all.png"), 
         unit = "in", 
         width = 5, 
         height = 4)

pca
```

```{r PCA_heatmap}
PCA_heatmap <- DimHeatmap(gex_scaled_all, dims = 1, cells = 500, balanced = TRUE) +
  theme_pub()

PCA_heatmap %>%
  ggsave(., filename = file.path(plotting_dir, "PCA_PC1_heatmap_all.png"), 
         unit = "in", 
         width = 6, 
         height = 4)
PCA_heatmap
```

```{r PCA_heatmap_pcs, fig.height=10, fig.width=7}
DimHeatmap(gex_scaled_all, dims = 1:10, cells = 500, balanced = TRUE)
```

```{r elbow_plot}
ElbowPlot(gex_scaled_all)
```

## Clustering

```{r Find_neighbors}
gex_scaled_all <- FindNeighbors(gex_scaled_all, dims = 1:10)
```

```{r Find_clusters}
gex_scaled_all <- FindClusters(gex_scaled_all, resolution = 0.5)
# Look at cluster IDs of the first 5 cells
head(Idents(gex_scaled_all), 5)
```

```{r Calculate_UMAP}
# If you haven't installed UMAP, you can do so via reticulate::py_install(packages =
# 'umap-learn')
gex_scaled_all <- RunUMAP(gex_scaled_all, dims = 1:10)
saveRDS(gex_scaled_all, file = file.path(RDS_dir, "gex_scaled_all.RDS"))
```

```{r Plot_UMAP, fig.width=6, fig.height=6}
# note that you can set `label = TRUE` or use the LabelClusters function to help label
# individual clusters
color_palette <- colorRampPalette(brewer.pal(12, "Set3"))

# Generate a palette with 40 colors
my_palette <- color_palette(40)

umap_plot <- DimPlot(gex_scaled_all, reduction = "umap", raster = FALSE, group.by = "seurat_clusters", cols=my_palette)
umap_plot

umap_plot %>%
  ggsave(., 
         filename = file.path(plotting_dir, "UMAP_all.png"), 
         unit = "in", 
         width = 6, 
         height = 6)
```

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
gex.markers <- FindAllMarkers(gex_scaled_all, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

group_by(gex.markers, cluster) %>%
  slice_max(n = 2, order_by = avg_log2FC)

saveRDS(gex.markers, file = file.path(RDS_dir, "gex.markers_all.RDS"))
```

```{r fig.width=20, fig.height=16}
marker_genes <- c(
  "PAX5", "MS4A1", "EBF1", "MME", # B-cell trajectories
  "CD14", "CEBPB", "MPO", "CD33", # Monocytes
  "CD3D", "CD8A", "TBX21", "IL7R", "CD3E", "CD3G", "CD4", "CD8B", # TCells
  "CD19", "CD5", "FCER2", "CD79B", # CLL B-Cell Marker
  "GNLY", "NKG7", "NCAM1", # NK cells
  "PF4", # platelets
  "CA1", "HBA1" # erythrocytes
)

marker_vln <- VlnPlot(gex_scaled_all, features = marker_genes) +
  theme_pub() +
  theme(legend.position = "none")

marker_vln
# CD14, MPO so far not found
marker_vln %>%
  ggsave(., file = file.path(analysis_dir, "/Plots/Marker_expression_vlnplot_all_patients.png"), 
         units = "in", 
         width = 20, 
         height = 16)
```

```{r fig.width=12, fig.height=16}
marker_umaps <- FeaturePlot(gex_scaled_all, features = marker_genes, raster = FALSE) +
  theme_pub() +
  theme(legend.position = "none")

marker_umaps %>%
  ggsave(., file = file.path(analysis_dir, "/Plots/Marker_expression_umaps_all_patients.png"), 
         units = "in", 
         width = 12, 
         height = 16)
marker_umaps
```

```{r}
gex.markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC) -> top10
```

```{r fig.height=8}
marker_heatmap <- DoHeatmap(gex_scaled_all, features = top10$gene) +
  theme_pub() +
  theme(legend.position = "none")

marker_heatmap
marker_heatmap %>%
  ggsave(., file = file.path(plotting_dir, "Marker_expression_heatmap_all.png"), 
         units = "in", 
         width = 20, 
         height = 16)
```

## Reference mapping

```{r Reference_UMAP, fig.width=4, fig.height=4}
reference_umap <- DimPlot(object = reference, 
                          reduction = "wnn.umap", 
                          group.by = "celltype.l1", 
                          label = TRUE, 
                          label.size = 3, 
                          repel = TRUE, 
                          raster = FALSE, 
                          cols = annotation_colours_named) + 
  NoLegend()

reference_umap %>%
  ggsave(., 
         file = file.path(plotting_dir, "Reference_umap_all.png"), 
         units = "in", 
         width = 4, 
         height = 4)
reference_umap
```

```{r FindTransferAnchors}
anchors <- FindTransferAnchors(
  reference = reference,
  query = gex_scaled_all,
  normalization.method = "SCT",
  reference.reduction = "spca",
  dims = 1:50
)
```

```{r MapReference_to_data}
gex_scaled_all <- MapQuery(
  anchorset = anchors,
  query = gex_scaled_all,
  reference = reference,
  refdata = list(
    celltype.l1 = "celltype.l1",
    celltype.l2 = "celltype.l2",
    predicted_ADT = "ADT"
  ),
  reference.reduction = "spca",
  reduction.model = "wnn.umap"
)

saveRDS(gex_scaled_all, file = file.path(RDS_dir, "gex_scaled_all.RDS"))
```

```{r fig.width=8, fig.height=4}
p1 <- DimPlot(gex_scaled_all,
  reduction = "umap",
  group.by = "predicted.celltype.l1",
  label = TRUE,
  label.size = 3,
  repel = TRUE,
  raster = FALSE,
  cols = annotation_colours_named
) +
  theme_pub() +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )

p2 <- DimPlot(gex_scaled_all,
  reduction = "ref.umap",
  group.by = "predicted.celltype.l1",
  label = TRUE,
  label.size = 3,
  repel = TRUE,
  raster = FALSE,
  cols = annotation_colours_named
) +
  theme_pub() +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )

ggarrange(p1, p2, nrow = 1, ncol = 2) %>% 
  ggsave(., file = file.path(plotting_dir, "Labeltransfer_Query_Ref_all.png"), 
         height = 4, 
         width = 8)

p1 + p2
```

```{r fig.width=8, fig.height=6}
prediction_score_umap <- FeaturePlot(gex_scaled_all,
  features = c("Mono", "CD8 T", "CD4 T", "other T", "other", "B", "DC", "NK"), 
  reduction = "umap", 
  cols = c("lightgrey", "darkred"), 
  ncol = 3, 
  raster = FALSE) &
  theme_pub() &
  theme(aspect.ratio = 1)

prediction_score_umap %>%
  ggsave(., file = file.path(plotting_dir, "Prediction_score_umaps_all.png"), 
         height = 6, 
         width = 8)

prediction_score_umap
```

```{r, fig.width=24}
Idents(gex_scaled_all) <- "predicted.celltype.l1"
markers_per_celltype_vlnplot <- VlnPlot(gex_scaled_all, features = c(
  "PAX5", "MS4A1", # B-cell trajectories
  "CD14", "CEBPB", "CD33", # Monocytes
  "CD3D", "CD8A", "TBX21", "IL7R", "CD3E", "CD3G", "CD4", "CD8B", # TCells
  "CD19", "CD5", "FCER2", "CD79B", # CLL B-Cell Marker
  "GNLY", "NKG7", # NK cells
  "CA1", "HBA1" # erythrocytes
), sort = TRUE) +
  theme_pub() +
  theme(legend.position = "none")

markers_per_celltype_vlnplot

markers_per_celltype_vlnplot %>%
  ggsave(., 
         file = file.path(plotting_dir, "Markers_per_celltype_vlnplot_all.png"), 
         height = 16, 
         width = 12)
```

## Add metadata to seurat objec
```{r}
# Add disease state annotation
sample_annot <- tibble("label_cellbarcode" = WhichCells(gex_scaled_all))

disease_annot <- sample_annot %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  tidyr::separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE) %>%
  dplyr::left_join(metadata %>% dplyr::select(label, disease) %>% unique()) %>% 
  dplyr::select(label_cellbarcode, disease) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = disease_annot, col.name = "disease")

## add MBL_CLL_subtype annotation
MBL_CLL_subtype_annot <- as_tibble(gex_scaled_all$orig.ident, rownames=NA) %>% 
  rownames_to_column("label_cellbarcode") %>% 
    dplyr::rename("label"="value") %>% 
  dplyr::left_join(metadata %>% dplyr::select(label, MBL_subtype)) %>% 
  dplyr::mutate(MBL_CLL_subtype=if_else(is.na(MBL_subtype), "CLL", MBL_subtype)) %>% 
  dplyr::select(label_cellbarcode, MBL_CLL_subtype) %>%
  deframe()
  
gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = MBL_CLL_subtype_annot, col.name = "MBL_CLL_subtype")

# add plotting id
plotting_annot <- as_tibble(gex_scaled_all$orig.ident, rownames=NA) %>% 
  rownames_to_column("label_cellbarcode") %>% 
  dplyr::rename("label"="value") %>% 
  left_join(metadata %>% 
              dplyr::select(label, plotting_id)) %>% 
  dplyr::select(label_cellbarcode, plotting_id) %>% 
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = plotting_annot, col.name = "plotting_id")

# add patient id
patient_id_annot <- as_tibble(gex_scaled_all$orig.ident, rownames=NA) %>% 
  rownames_to_column("label_cellbarcode") %>% 
  dplyr::rename("label"="value") %>% 
  left_join(metadata %>% 
              dplyr::select(label, patient_id)) %>% 
  dplyr::select(label_cellbarcode, patient_id) %>% 
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = patient_id_annot, col.name = "patient_id")

sequencing_annot <- tibble("label_cellbarcode" = WhichCells(gex_scaled_all)) %>% 
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
    dplyr::left_join(metadata %>% dplyr::filter(data_type=="GEX") %>% dplyr::select(label, sequencing_date) %>% unique()) %>% 
    dplyr::select(label_cellbarcode, sequencing_date) %>%
    dplyr::mutate(sequencing_date=as.factor(as.character(sequencing_date))) %>% deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = sequencing_annot, col.name = "sequencing_date")
```

## Plot UMAPs
### Patient UMAP
```{r Patient_UMAP, fig.width=5, fig.height=5}
patient_umap <- DimPlot(gex_scaled_all, 
                        group.by = "patient_id", 
                        raster = F, 
                        cols = my_unpaired_colours_named, 
                        shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", title = "All Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
    legend.position = "bottom",
      aspect.ratio=1)

patient_umap 
```

### Patient disease umap
```{r}
gex_scaled_all$disease <- factor(gex_scaled_all$disease, levels = c("MBL", "CLL"))
patient_umap_disease <- DimPlot(gex_scaled_all, 
                                group.by = "patient_id", 
                                raster = F, 
                                cols = my_unpaired_colours_named, 
                                split.by = "disease", 
                                shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1
)

patient_umap_disease
```

### Patient MBL-CLL-subtype UMAP
```{r}
gex_scaled_all$MBL_CLL_subtype <- factor(gex_scaled_all$MBL_CLL_subtype, levels = c("LC", "HC", "CLL"))

patient_umap_MBL_CLL_subtype <- DimPlot(gex_scaled_all, 
                                group.by = "patient_id", 
                                raster = F, 
                                cols = my_unpaired_colours_named, 
                                split.by = "MBL_CLL_subtype", 
                                shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1
)

patient_umap_MBL_CLL_subtype
```

### Patient progression UMAP split
```{r}
progression_annot_all <- sample_annot %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  tidyr::separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE) %>%
  dplyr::mutate(progression = if_else(str_detect(patient_id, paste0("MBL", c(1:3, 5:15), collapse = "|")), FALSE, TRUE)) %>%
  dplyr::select(label_cellbarcode, progression) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = progression_annot_all, col.name = "progression")
```

```{r}
patient_umap_disease_prog <- DimPlot(subset(gex_scaled_all, subset = progression == TRUE), 
                                     group.by = "patient_id", 
                                     raster = F, 
                                     cols = my_unpaired_colours_named, 
                                     split.by = "disease", 
                                     shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Progressive Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1
  ) +
  guides(colour = guide_legend(nrow = 2,
                            keywidth = 0.5,
                            keyheight = 0.5,
                            override.aes = list(size=2)))

patient_umap_disease_prog
```
 
### Patient non-progressive UMAP split
```{r}
patient_umap_disease_np <- DimPlot(subset(gex_scaled_all, subset = progression == FALSE), 
                                   group.by = "patient_id", 
                                   raster = F, 
                                   cols = my_unpaired_colours_named, 
                                   shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Non-progressive Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1
  ) +
  guides(fill = guide_legend(nrow = 3))

patient_umap_disease_np
```

```{r fig.width=19, fig.height=6}
cowplot::plot_grid(patient_umap, 
                   patient_umap_disease_prog, 
                   patient_umap_disease_np, 
                   nrow = 1, 
                   ncol = 3, 
                   rel_widths = c(1.1, 1.8, 1.1), 
                   align = "h") %>%
  ggsave(., 
         file = file.path(plotting_dir, "Sample_UMAPs_all.png"), 
         width = 19, 
         height = 6, 
         unit = "in")
```

### Patient UMAP split
```{r Patient_UMAP_split, fig.width=16, fig.height=16}
patient_umap_split <- DimPlot(gex_scaled_all, 
                              group.by = "patient_id", 
                              raster = F, 
                              cols = my_unpaired_colours_named, 
                              split.by = "patient_id", 
                              ncol = 3) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.line = element_line(),
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    aspect.ratio = 1
  )

patient_umap_split

patient_umap_split %>%
  ggsave(file = file.path(analysis_dir, "/Plots/Sample_UMAPs_split_all.png"), 
         width = 5.3, 
         height = 16, 
         unit = "in")
```

### Marker gene Violinplot
```{r Marker_expression_per_celltype, fig.height=12}
VlnPlot(gex_scaled_all, group.by = "predicted.celltype.l1", features = c("CD3E", "PAX5", "CD5", "CD19", "MS4A1", "CD14", "NCAM1"))

celltype_marker_vlnplot <- VlnPlot(gex_scaled_all,
  group.by = "predicted.celltype.l1", features =
    c(
      "MS4A1", # B cells
      "CD14", # Monocytes
      "IL7R", "CD3E", # T cells
      "CD19", "CD5", "FCER2", "CD79B", # CLL
      "GNLY", "NKG7", # NK cells
      "CA1", "HBA1") # Erythrocytes)
) +
  theme_pub() +
  theme(legend.position = "none")

celltype_marker_vlnplot
# DefaultAssay(gex_scaled_all) = 'SCT'

celltype_marker_vlnplot %>%
  ggsave(file = file.path(analysis_dir, "/Plots/Selected_marker_expression_celltypes_vlnplot_all.png"), 
         width = 16, 
         height = 16, 
         unit = "in")
```

### Marker gene Dotplot
```{r Marker_expression_per_celltype_dotplot, fig.width=5, fig.height=3}
celltype_marker_dotplot <- DotPlot(gex_scaled_all,
  features =
    c(
      "MS4A1", # B cells
      "CD14", # Monocytes
      "IL7R", "CD3E", # T cells
      "CD19", "CD5", "FCER2", "CD79B", # CLL
      "GNLY", "NKG7", # NK cells
      "CA1", "HBA1"
    ), # Erythrocytes)),
  group.by = "predicted.celltype.l1",
  cols = "RdYlBu"
) +
  labs(y = "Celltype", x = "Marker gene") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust=0.9))

ggsave(celltype_marker_dotplot,
  file = file.path(plotting_dir, "Selected_marker_expression_celltypes_dotplot_all.png"), 
  width = 7, 
  height = 4, 
  unit = "in"
)
```

## Manual Annotation

```{r Add_my_annotation}
refmap_annot <- tibble("label_cellbarcode" = WhichCells(gex_scaled_all), 
                       "predicted.celltype.l1" = gex_scaled_all$predicted.celltype.l1)

manual_annot <- refmap_annot %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  tidyr::separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE) %>%
  dplyr::mutate(
    My_annotation =
      case_when(
        `predicted.celltype.l1` == "B" & `sample` == "1" ~ "MBL B Cell",
        `predicted.celltype.l1` == "B" & `sample` == "2" ~ "CLL B Cell",
        `predicted.celltype.l1` == "B" & str_detect(patient_id,  "MBL4") ~ "CLL B Cell",
        `predicted.celltype.l1` == "B" & str_detect(patient_id, paste0("MBL", c(1:3, 6:15), collapse = "|")) ~ "MBL B Cell",
        `predicted.celltype.l1` %in% c("CD4 T", "CD8 T", "other T") ~ "T",
        .default = `predicted.celltype.l1`
      )
  ) %>%
  dplyr::select(My_annotation, label_cellbarcode) %>%
  with(setNames(My_annotation, label_cellbarcode))

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = manual_annot, col.name = "My_annotation")
gex_scaled_all <- SetIdent(gex_scaled_all, value = gex_scaled_all@meta.data$My_annotation)
```

## Add MBL annotation
```{r}
MBL_subtype <- tibble("label_cellbarcode"=WhichCells(gex_scaled_all),
       "label"=gex_scaled_all$orig.ident) %>% 
  left_join(metadata %>% dplyr::select(label, MBL_subtype)) %>%  
  dplyr::select(label_cellbarcode, MBL_subtype) %>% 
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = MBL_subtype, col.name = "MBL_subtype")
```

```{r}
celltype_umap <- DimPlot(gex_scaled_all, group.by = "My_annotation", raster = F, cols = annotation_colours_named,  shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2",  title="") +
  scale_colour_manual(name = "Cell type", 
                      values = annotation_colours_named, 
                      labels = c("DC" = "Dendritic cells", 
                                 "Mono" = "Monocytes", 
                                 "CLL B Cell" = "CLL cells", 
                                 "MBL B Cell" = "MBL cells", 
                                 "NK" = "Natural killer cells", 
                                 "T" = "T cells",
                                 "other"="Other cell types"))+
  theme_pub() +
  theme(axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1)
celltype_umap

celltype_umap %>%
  ggsave(file = file.path(plotting_dir, "Celltype_umap_all.png"), 
         width = 5, 
         height = 5, 
         unit = "in")
```

```{r fig.width=3, fig.height=4}
total_cell_types <- gex_scaled_all@meta.data %>%
  dplyr::select(disease, My_annotation, patient_id) %>%
  group_by(disease, My_annotation) %>%
  summarise(cell_counts = n()) %>%
  mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>%
  arrange(desc(disease), desc(cell_counts)) %>% 
  dplyr::mutate(My_annotation=factor(My_annotation, My_annotation)) %>% 
  ggplot(aes(fill = My_annotation, y = cell_counts, x = disease)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(breaks=seq(0, 1, 0.25), 
                     labels = c("0", "25", "50", "75", "100"), 
                     expand = expansion(mult = c(0, NA))) +
  scale_fill_manual(
    values = annotation_colours_named,
    name = "Cell type", 
    labels = c("DC" = "Dendritic cells", 
                                 "Mono" = "Monocytes", 
                                 "CLL B Cell" = "CLL cells", 
                                 "MBL B Cell" = "MBL cells", 
                                 "NK" = "Natural killer cells", 
                                 "T" = "T cells",
               "other"="Other cell types")
  ) +
  labs(x = "", y = "Frequency [%]") +
  theme_pub()
total_cell_types

total_cell_types %>%
  ggsave(file = file.path(plotting_dir, "Celltype_abundance_total_all.png"), 
         width = 3, 
         height = 4, 
         unit = "in")
```

```{r Celltype_quantification_median, fig.width=3, fig.height=4}
median_cell_types <- gex_scaled_all@meta.data %>%
  dplyr::select(disease, My_annotation, patient_id) %>%
  group_by(disease, My_annotation, patient_id) %>%
  dplyr::summarise(counts = n()) %>%
  ungroup() %>%
  group_by(disease, My_annotation) %>%
  dplyr::summarise(
    sd_celltype_abundance = sd(counts, na.rm = TRUE),
    mean_abundance = median(counts)
  ) %>%
  mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>%
  arrange(desc(disease), desc(mean_abundance)) %>% 
  dplyr::mutate(My_annotation=factor(My_annotation, My_annotation)) %>% 
  ggplot(., aes(fill = My_annotation, y = mean_abundance, x = disease)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(breaks=seq(0, 1, 0.25), 
                     labels = c("0", "25", "50", "75", "100"), 
                     expand = expansion(mult = c(0, NA))) +
  scale_fill_manual(
    values = annotation_colours_named,
    name = "Cell type", 
    labels = c("DC" = "Dendritic cells", 
                                 "Mono" = "Monocytes", 
                                 "CLL B Cell" = "CLL cells", 
                                 "MBL B Cell" = "MBL cells", 
                                 "NK" = "Natural killer cells", 
                                 "T" = "T cells",
               "other"="Other cell types")
  ) +
  labs(x = "", y = "Frequency [%]") +
  theme_pub()
median_cell_types

saveRDS(median_cell_types, file.path(RDS_dir, "median_cell_types_GEX_all.RDS"))

median_cell_types %>%
  ggsave(file = file.path(plotting_dir, "Celltype_abundance_median_all.png"), 
         width = 3, 
         height = 4, 
         unit = "in")
```


```{r Celltype_quantification, fig.width=20, fig.height=3}
cell_types_per_patient <- gex_scaled_all@meta.data %>%
  dplyr::select(disease, My_annotation, plotting_id) %>%
  dplyr::mutate(plotting_id=factor(plotting_id, levels=c(paste0("P", 1:25), "P9dup"))) %>% 
  group_by(disease, My_annotation, plotting_id) %>%
  summarise(cell_counts = n()) %>%
  mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>%
  ungroup() %>%
  ggplot(., aes(fill = My_annotation, y = cell_counts, x = disease)) +
    geom_bar(position = "fill", stat = "identity") +
    scale_y_continuous(breaks=seq(0, 1, 0.25), 
                     labels = c("0", "25", "50", "75", "100"), 
                     expand = expansion(mult = c(0, NA))) +
    scale_fill_manual(
      values = annotation_colours_named,
      name = "Cell type"
    ) +
    labs(x = "", y = "Frequency [%]") +
    theme_pub()+
    theme(legend.position = "bottom")+
  facet_wrap(~plotting_id, nrow=1, ncol=25)

cell_types_per_patient
```

```{r fig.width=12, fig.height=5}
all_malignant_cell_plot <-  gex_scaled_all@meta.data %>%
  dplyr::select(disease, My_annotation, patient_id) %>%
  dplyr::filter(patient_id !="MBL7dup") %>% 
  dplyr::group_by(My_annotation, patient_id, disease) %>%
  dplyr::summarise(celltype_counts = n()) %>%
  dplyr::left_join(metadata %>% dplyr::select(label, disease, patient_id, plotting_id, MBL_CLL_subtype)) %>% 
  dplyr::mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>% 
  dplyr::filter(My_annotation %in% c("B", "MBL B Cell", "CLL B Cell")) %>%
  tidyr::pivot_wider(id_cols = c(label, MBL_CLL_subtype), names_from = My_annotation, values_from = celltype_counts, values_fill = NA) %>%
  tidyr::pivot_longer(-c(label, MBL_CLL_subtype), names_to = "celltype", values_to = "counts") %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "GEX") %>% dplyr::select(label, plotting_id, disease, MBL_CLL_subtype)) %>%
  dplyr::arrange(desc(plotting_id)) %>%
  dplyr::mutate(
    counts = if_else(celltype %in% c("B"), -counts, counts),
    celltype = factor(celltype, levels = c("B", "MBL B Cell", "CLL B Cell")),
    label = factor(label, level = c(
       "MBL12", "MBL11", "MBL14", "MBL13","MBL15", "MBL10", "MBL6", "MBL1", "MBL7", "MBL3", "MBL8", "MBL2", "MBL9",
      "CLL6.1", "CLL6.2", "CLL1.1", "CLL1.2", "CLL7.1", "CLL7.2", "CLL2.1", "CLL2.2", "CLL5.1", "CLL5.2", "CLL8.1",
      "CLL8.2", "CLL4.1", "CLL4.2", "CLL10.1", "CLL10.2", "CLL9.1", "CLL9.2", "CLL3.1", "CLL3.2", "MBL4"
    ))
  ) %>%
  ggplot(aes(x = label, y = counts, fill = MBL_CLL_subtype)) +
  geom_bar(position = "stack", stat = "identity", width = 0.7) +
  scale_y_continuous(#trans = ggallin::pseudolog10_trans, 
                     #breaks = c(10, 100, 1000, 10000), 
                     #limits=c(0, 20000),
                     expand=expansion(mult = c(0, NA))) +
  scale_x_discrete(labels = label2plotting_id) +
  scale_fill_manual(name = "Cell type", values = my_MBL_CLL_colours_named, 
  labels = c("HC_MBL" = "HC MBL", "LC_MBL" = "LC MBL", "CLL" = "CLL")) +
  labs(title = "Cell counts", x = "")+
  theme_pub() +
  theme(axis.title.y=element_blank(),
        # axis.text.x=element_blank(),
        axis.text.x=element_text(angle=90),
        # axis.ticks.x=element_blank(),
        # axis.line.x=element_blank(),
    plot.title = element_text(margin = margin(t = 10, b = -20))) 
all_malignant_cell_plot

saveRDS(all_malignant_cell_plot, file.path(RDS_dir, "all_malignant_cell_plot_all.RDS"))
```

```{r fig.width=5, fig.height=12}
all_malignant_cell_plot_vertical <-  gex_scaled_all@meta.data %>%
  dplyr::select(disease, My_annotation, patient_id) %>%
  dplyr::filter(patient_id !="MBL7dup") %>% 
  dplyr::group_by(My_annotation, patient_id, disease) %>%
  dplyr::summarise(celltype_counts = n()) %>%
  dplyr::left_join(metadata %>% dplyr::select(label, disease, patient_id, plotting_id, MBL_CLL_subtype)) %>% 
  dplyr::mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>% 
  dplyr::filter(My_annotation %in% c("B", "MBL B Cell", "CLL B Cell")) %>%
  tidyr::pivot_wider(id_cols = c(label, MBL_CLL_subtype), names_from = My_annotation, values_from = celltype_counts, values_fill = NA) %>%
  tidyr::pivot_longer(-c(label, MBL_CLL_subtype), names_to = "celltype", values_to = "counts") %>%
  dplyr::left_join(metadata %>% dplyr::filter(data_type == "GEX") %>% dplyr::select(label, plotting_id, disease, MBL_CLL_subtype)) %>%
  dplyr::arrange(desc(plotting_id)) %>%
  dplyr::mutate(
    counts = if_else(celltype %in% c("B"), -counts, counts),
    celltype = factor(celltype, levels = c("B", "MBL B Cell", "CLL B Cell")),
    label = factor(label, level = rev(c(
       "MBL12", "MBL11", "MBL14", "MBL13","MBL15", "MBL10", "MBL6", "MBL1", "MBL7", "MBL3", "MBL8", "MBL2", "MBL9",
      "CLL6.1", "CLL6.2", "CLL1.1", "CLL1.2", "CLL7.1", "CLL7.2", "CLL2.1", "CLL2.2", "CLL5.1", "CLL5.2", "CLL8.1",
      "CLL8.2", "CLL4.1", "CLL4.2", "CLL10.1", "CLL10.2", "CLL9.1", "CLL9.2", "CLL3.1", "CLL3.2", "MBL4"
    )))
  ) %>%
  ggplot(aes(y = label, x = counts, fill = MBL_CLL_subtype)) +
  geom_bar(position = "stack", stat = "identity", width = 0.7) +
  scale_x_continuous(#trans = ggallin::pseudolog10_trans, 
                     #breaks = c(10, 100, 1000, 10000), 
                     #limits=c(0, 20000),
                     expand=expansion(mult = c(0, NA))) +
  scale_y_discrete(labels = label2plotting_id) +
  scale_fill_manual(name = "Cell type", values = my_MBL_CLL_colours_named, 
  labels = c("HC_MBL" = "HC MBL", "LC_MBL" = "LC MBL", "CLL" = "CLL")) +
  labs(title = "", x = "")+
  theme_pub() +
  theme(axis.title.y=element_blank(),
        # axis.text.x=element_blank(),
        axis.text.y=element_text(angle=0),
        # axis.ticks.x=element_blank(),
        # axis.line.x=element_blank(),
    plot.title = element_text(margin = margin(t = 10, b = -20))) 
all_malignant_cell_plot_vertical

saveRDS(all_malignant_cell_plot_vertical, file.path(RDS_dir, "all_malignant_cell_plot_vertical.RDS"))

```

# Sample stats

## Total cell count
```{r}
cell_counts_GEX <- sum(!(gex_scaled_all@meta.data[["patient_id"]] %in% c("MBL7dup")))
```

## Percentages polyclonal cells
```{r}
percentages_polyclonal_cells <- gex_scaled_all@meta.data %>%
  dplyr::select(orig.ident, My_annotation_B, MBL_subtype) %>%
  dplyr::mutate(MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype)) %>%
  dplyr::select(orig.ident, My_annotation_B, MBL_CLL_subtype) %>%
  dplyr::filter(orig.ident != "MBL7dup") %>%
  dplyr::group_by(My_annotation_B, orig.ident, MBL_CLL_subtype) %>%
  dplyr::summarise(celltype_counts = n()) %>%
  dplyr::left_join(metadata %>% dplyr::select(label, patient_id, plotting_id), by = join_by("orig.ident" == "label")) %>%
  tidyr::pivot_wider(id_cols = c(orig.ident, patient_id, plotting_id, MBL_CLL_subtype), names_from = My_annotation_B, values_from = celltype_counts) %>%
  dplyr::mutate(Total_cell_counts = sum(c_across(c(B, `CLL B Cell`, `MBL B Cell`)), na.rm = TRUE),
                B_perc = B / Total_cell_counts * 100) %>%
  ungroup() %>%
  group_by(MBL_CLL_subtype) %>%
  dplyr::summarize(
    min_B_perc = min(B_perc, na.rm = TRUE),
    max_B_perc = max(B_perc, na.rm = TRUE),
    mean_B_perc = mean(B_perc, na.rm = TRUE),
    median_B_perc = median(B_perc, na.rm = TRUE)
  )
percentages_polyclonal_cells
```


```{r, fig.width=6}
cellcounts <- gex_scaled_all@meta.data %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  group_by(label) %>%
  dplyr::summarise("cell_counts" = n()) %>%
  arrange(desc(cell_counts)) %>% 
  dplyr::mutate(label=factor(label, label)) %>% 
  ggplot(aes(x = label, y = cell_counts, fill = label)) +
  geom_bar(stat = "Identity") +
  scale_fill_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  scale_y_continuous(expand = expansion(mult = c(0, NA))) +
  scale_x_discrete(labels=label2plotting_id)+
  labs(x = "Sample", y = "Cells counts") +
  theme_pub() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust=1, vjust=0.9)
  )
cellcounts

cellcounts %>%
  ggsave(., file = file.path(plotting_dir, "Cellcounts_all.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

```{r, fig.width=13, fig.height=20}
ggarrange(patient_umap_disease + 
            labs(title = ""),
  ggarrange(cellcounts, 
            celltype_umap + 
              labs(title = ""), 
            ncol = 2, 
            nrow = 1, 
            labels = c("B", "C")),
  celltype_marker_dotplot,
  nrow = 3,
  ncol = 1,
  labels = c("A", "", "D")
) %>%
  ggsave(., file = file.path(plotting_dir, "Figure_patients_annotation_all.png"), 
         width = 13, 
         height = 20, 
         unit = "in")
```

Distribution of expression values for each sample and patient

```{r}
gene_expression_list <- list()
for (patient in paste0("CLL", 1:10)) {
  print(patient)
  gene_expression_patient <- GetAssayData(object = subset(gex_scaled_all, subset = patient_id == patient), slot = "scale.data")
  gene_expression_list[[patient]] <- gene_expression_patient
}

gene_expression_distribution <- gene_expression_list %>%
  imap( ~ as.data.frame(.x) %>%
    stack() %>%
    mutate(patient = .y) %>%
    dplyr::select(values, patient)) %>%
  bind_rows() %>%
  group_by(patient)

gene_expression_plot <- gene_expression_distribution %>%
  ggplot(aes(x = values, colour = patient)) +
  geom_density(linewidth = 2) +
  scale_x_continuous(limits = c(0, 20)) +
  scale_colour_manual(name = "Patient", values = my_unpaired_colours_named) +
  labs(x = "Values", "Density") +
  theme_pub() +
  theme(legend.position = "bottom")

gene_expression_plot %>% 
  ggsave(., file = file.path(plotting_dir, "Gene_expression_distribution_all.png"), 
         width = 13, 
         height = 10, 
         unit = "in")
```

# Differential Expression
## HC-MBL vs. CLL patients' cells
### All cells
```{r Differential_between_all_patients_all_cells_not_downsampled}
disease_markers <- FindMarkers(subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("HC", "CLL")), 
                               ident.1 = "CLL B Cell", ident.2 = "MBL B Cell", 
                               test.use = "wilcox") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
                                         p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
                                         .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL"), gene, NA))

saveRDS(disease_markers, file = file.path(RDS_dir, "disease_markers_all.RDS"))

# celltype_markers  <- FindAllMarkers(gex_scaled_all, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
```

```{r Volcanoplot_all_patients}
volcanoplot_patients <- disease_markers %>%
  dplyr::mutate(significance=factor(significance, levels=c("MBL", "CLL", "not significant"))) %>% 
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 1.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-5, 5)) +
  scale_y_continuous("-log10(FDR)", oob = scales::squish_infinite, expand = expansion(mult=c(0, NA))) +
  scale_colour_manual(name = "DEG", 
                      values = c("MBL" = "#FC8D62", "CLL"="#66C2A5", "not significant" = "lightgrey"),
                      labels=c("MBL"="HC-MBL", "CLL"="CLL", "not significant"="not significant")) +
  theme_pub() +
  theme(legend.position = "bottom")
volcanoplot_patients

volcanoplot_patients %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_volcanoplot_all.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

Try with patient as covariate
```{r}
disease_markers_lr <- FindMarkers(subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("HC", "CLL")), 
                                  ident.1 = "CLL B Cell", 
                                  ident.2 = "MBL B Cell", test.use = "LR", 
                                  latent.vars = c("patient_id")) %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL"), gene, NA))

saveRDS(disease_markers_lr, file.path(RDS_dir, "disease_markers_lr_all.RDS"))
```

```{r}
volcanoplot_patients_lr <- disease_markers_lr %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-5, 5)) +
  scale_y_continuous("-log10(FDR)", oob = scales::squish_infinite, expand = expansion(mult=c(0, NA))) +
  scale_colour_manual(name = "DEG", 
                      values = c("MBL" = "#FC8D62", "CLL"="#66C2A5", "not significant" = "lightgrey")) +
  theme_pub() +
  theme(legend.position = "bottom")
volcanoplot_patients_lr

volcanoplot_patients_lr %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_LR_volcanoplot_all.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

### Downsampled n=543

```{r}
# Determine number to downsample to
min_malignant_cells <- subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("HC", "CLL"))@meta.data %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  mutate(malignance = if_else(My_annotation %in% c("MBL B Cell", "CLL B Cell"), TRUE, FALSE)) %>%
  group_by(label, malignance) %>%
  dplyr::summarise("cell_counts" = n()) %>%
  dplyr::filter(malignance == TRUE) %>%
  arrange(cell_counts) %>%
  pull(cell_counts) %>%
  min()

min_malignant_cells
```

```{r Downsampling_all_patients}
malignance_annot <- gex_scaled_all@meta.data %>%
  rownames_to_column("label_cellbarcode") %>%
  mutate(malignance = if_else(My_annotation %in% c("MBL B Cell", "CLL B Cell"), TRUE, FALSE)) %>%
  dplyr::select(label_cellbarcode, malignance) %>%
  with(setNames(malignance, label_cellbarcode))

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = malignance_annot, col.name = "malignance")

gex_scaled_downsampled <- subset(gex_scaled_all, subset = malignance == TRUE, downsample = min_malignant_cells)

# Check if evey sample has now min_malignant_cells
# gex_scaled_downsampled@meta.data %>%
# rownames_to_column(var = "label_cellbarcode") %>%
#   tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
#   mutate(malignance=if_else(My_annotation %in% c("MBL B Cell", "CLL B Cell"), TRUE, FALSE)) %>%
#   group_by(label, malignance) %>%
#   dplyr::summarise("cell_counts" = n()) %>%
#   dplyr::filter(malignance ==TRUE)
```



```{r}
disease_markers_downsampled_all_progression <- FindMarkers(gex_scaled_downsampled, 
                                           ident.1 = "CLL B Cell", 
                                           ident.2 = "MBL B Cell", 
                                           test.use="wilcox") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL") & !str_starts(gene, "IG"), gene, NA))

saveRDS(disease_markers_downsampled_all_progression, file = file.path(RDS_dir, "disease_markers_downsampled_all_progression.RDS"))
```

```{r Volcanoplot_all_prog_patients_downsampled_all_progression, fig.height=5, fig.width=5}
#CLL right, MBL left
volcanoplot_patients_downsampled_all_progression <- disease_markers_downsampled_all_progression %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label=label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 150, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", #limits = c(-5, 5)
                     ) +
  scale_y_continuous("-log10(FDR)", 
                     # limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_patients_downsampled_all_progression

saveRDS(volcanoplot_patients_downsampled_all_progression, file.path(RDS_dir, "volcanoplot_patients_downsampled_all_progression.RDS"))

volcanoplot_patients_downsampled_all_progression %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_downsampled_volcanoplot_all_progression.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

```{r FindMarkers_downsampled_all_prog_patients}
# Subset only progressive patients
gex_scaled_downsampled_prog <- subset(gex_scaled_downsampled, subset = patient_id %in% paste0 ("CLL", 1:10))

#currently filtered for at least avglog2FC > 2 and non IG genes
disease_markers_downsampled <- FindMarkers(gex_scaled_downsampled_prog, 
                                           ident.1 = "CLL B Cell", 
                                           ident.2 = "MBL B Cell", 
                                           test.use="wilcox") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL") & !str_starts(gene, "IG"), gene, NA))

saveRDS(disease_markers_downsampled, file = file.path(RDS_dir, "disease_markers_downsampled_all.RDS"))

disease_markers_lr_downsampled <- FindMarkers(gex_scaled_downsampled_prog, 
                                              ident.1 = "CLL B Cell", 
                                              ident.2 = "MBL B Cell", test.use = "LR") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL") & !str_starts(gene, "IG"), gene, NA))
```

```{r Volcanoplot_all_prog_patients_downsampled, fig.height=5, fig.width=5}
#CLL right, MBL left
volcanoplot_patients_downsampled <- disease_markers_downsampled %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label=label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 150, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-5, 5)) +
  scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_patients_downsampled

saveRDS(volcanoplot_patients_downsampled, file.path(RDS_dir, "volcanoplot_patients_downsampled_all.RDS"))

volcanoplot_patients_downsampled %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_downsampled_volcanoplot_all.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

### Pseudobulk AggregateExpression()

```{r}
# Subset to only MBL and CLL cells and ensure only progressive MBLs are included
gex_scaled_malignant <- subset(gex_scaled_all, subset = malignance == TRUE & progression == TRUE)

# Aggregate cells per patient sample name (either MBL or CLL)
gex_scaled_pseudobulk <- AggregateExpression(
  gex_scaled_malignant,
  return.seurat = TRUE,
  group.by = c("orig.ident", "disease"),
  verbose = TRUE
)

# Extract from sample name if MBL or CLL sample ### needs to be adapted!! not correct
disease <- gex_scaled_pseudobulk@meta.data %>%
  rownames_to_column("label") %>%
  dplyr::select(label) %>%
  tidyr::separate(label, into = c("label", "disease"), sep = "_") %>% 
  deframe()

# Add disease state per sample as metadata
gex_scaled_pseudobulk <- AddMetaData(object = gex_scaled_pseudobulk, metadata = disease, col.name = "disease")
```

```{r}
Idents(gex_scaled_pseudobulk) <- "disease"
disease_markers_pseudobulk <- FindMarkers(gex_scaled_pseudobulk, ident.1 = "CLL", ident.2 = "MBL", test.use = "DESeq2", assay = 'RNA') %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL"), gene, NA))
saveRDS(disease_markers_pseudobulk, file = file.path(RDS_dir, "disease_markers_pseudobulk_all.RDS"))
```

```{r}
volcanoplot_pseudobulk <- disease_markers_pseudobulk %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-12, 12)) +
  scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_pseudobulk

volcanoplot_pseudobulk %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_pseudobulk_volcanoplot_all.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

 vs B cells
```{r}
# Subset to only B-like cells and ensure only progressive MBLs are included
gex_scaled_Blike_progression <- subset(gex_scaled_all, 
                                       subset = My_annotation_B %in% c("MBL B Cell", "B", "CLL B Cell") &
                                         progression == TRUE)

# Aggregate cells per celltype per patient
gex_scaled_Blike_pseudobulk <- AggregateExpression(
  gex_scaled_Blike_progression,assays = 'RNA',
  return.seurat = TRUE,
  group.by = c("orig.ident", "My_annotation_B"),
  verbose = TRUE
)

# Extract from sample name 
Blike <- gex_scaled_Blike_pseudobulk@meta.data %>%
  rownames_to_column("label") %>%
  dplyr::select(label) %>%
  tidyr::separate(label, into = c("label", "My_annotation_B"), sep = "_") %>% 
  deframe()

# Add disease state per sample as metadata
gex_scaled_Blike_pseudobulk <- AddMetaData(object = gex_scaled_Blike_pseudobulk, metadata = Blike, col.name = "My_annotation_B")

Idents(gex_scaled_Blike_pseudobulk) <- "My_annotation_B"
disease_markers_B_CLL_pseudobulk <- FindMarkers(gex_scaled_Blike_pseudobulk, ident.1 = "CLL B Cell", ident.2 = "B", test.use="DESeq2") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL B Cell",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "B",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("B", "CLL B Cell"), gene, NA))

volcanoplot_pseudobulk_B_CLL <- disease_markers_B_CLL_pseudobulk %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", #limits = c(-12, 12)
                     ) +
  scale_y_continuous("-log10(FDR)", 
                     # limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="Expression enriched in",
                      values=c("CLL B Cell"="#66C2A5", "B"="darkred", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

disease_markers_B_HC_MBL_pseudobulk <- FindMarkers(gex_scaled_Blike_pseudobulk, ident.1 = "MBL B Cell", ident.2 = "B", test.use = "DESeq2") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "MBL B Cell",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "B",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("B", "MBL B Cell"), gene, NA))

volcanoplot_pseudobulk_B_HC_MBL <- disease_markers_B_HC_MBL_pseudobulk %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", #limits = c(-12, 12)
                     ) +
  scale_y_continuous("-log10(FDR)", 
                     # limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="Expression enriched in",
                      values=c("MBL B Cell"="#FC8D62", "B"="darkred", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

gex_scaled_Blike_LC_progression <- subset(gex_scaled_all, 
                                       subset = My_annotation_B %in% c("MBL B Cell", "B") &
                                         MBL_CLL_subtype =="LC")

# Aggregate cells per celltype per patient
gex_scaled_Blike_LC_pseudobulk <- AggregateExpression(
  gex_scaled_Blike_LC_progression,
  return.seurat = TRUE,
  group.by = c("orig.ident", "My_annotation_B"),
  verbose = TRUE
)

Blike_LC <- gex_scaled_Blike_LC_pseudobulk@meta.data %>%
  rownames_to_column("label") %>%
  dplyr::select(label) %>%
  tidyr::separate(label, into = c("label", "My_annotation_B"), sep = "_") %>% 
  deframe()

gex_scaled_Blike_LC_pseudobulk <- AddMetaData(object = gex_scaled_Blike_LC_pseudobulk, metadata = Blike_LC, col.name = "My_annotation_B")

Idents(gex_scaled_Blike_LC_pseudobulk) <- "My_annotation_B"
disease_markers_B_LC_MBL_pseudobulk <- FindMarkers(gex_scaled_Blike_LC_pseudobulk, ident.1 = "MBL B Cell", ident.2 = "B") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "MBL B Cell",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "B",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("B", "MBL B Cell"), gene, NA))

volcanoplot_pseudobulk_B_LC_MBL <- disease_markers_B_LC_MBL_pseudobulk %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", #limits = c(-12, 12)
                     ) +
  scale_y_continuous("-log10(FDR)", 
                     # limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="Expression enriched in",
                      values=c("MBL B Cell"="#fcbda4", "B"="darkred", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

ggarrange(volcanoplot_pseudobulk_B_LC_MBL, volcanoplot_pseudobulk_B_HC_MBL, volcanoplot_pseudobulk_B_CLL, nrow=1, ncol=3)
```

### Individual patients
#### Logistic regression
```{r}
disease_markers_lr_patients <- list()
for (patient in paste0("CLL", seq(1:10))) {
  gex_scaled_patient <- subset(gex_scaled_all, subset = patient_id == patient)

  disease_markers_lr_patient <- FindMarkers(gex_scaled_patient, ident.1 = "CLL B Cell", ident.2 = "MBL B Cell", test.use = "LR") %>%
    rownames_to_column(var = "gene") %>%
    dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL"), gene, NA))

  disease_markers_lr_patients[[patient]] <- disease_markers_lr_patient
}
```

```{r}
volcanoplots_patients_lr <- disease_markers_lr_patients %>%
  imap( ~ 
          ggplot(., aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label=label)) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
    geom_point() +
    geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black") +
    scale_x_continuous("Log2(Fold change)", limits = c(-4, 4)) +
    scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 300), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, 0.09))) +
  scale_colour_manual(name="Differentially expressed genes",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3")) +
  labs(title = patient_id2plotting_id[.y])+
  theme_pub() +
  theme(legend.position = "bottom"))
```

```{r fig.width=22.5, fig.height=16}
volcanoplots_patients_lr %>%
  ggarrange(plotlist = ., 
            ncol = 3, 
            nrow = 4, 
            common.legend = TRUE, 
            legend = "bottom")

volcanoplots_patients_lr %>%
  ggarrange(plotlist = ., 
            ncol = 4, 
            nrow = 3, 
            common.legend = TRUE, 
            legend = "bottom") %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_LR_volcanoplots_all.png"), 
         width = 32, 
         height = 19.2, 
         unit = "cm", 
         bg="white")
```

```{r}
# Genes found in patient CLL5, CLL8 and CLL9
ATAC_genes <- c(
  "CTCF", "CTCFL", "BATF", "NFE2", "BACH2",
  "JDP2", "FOSB", "BACH1", "FOS", "FOSL2",
  "SMARCC1", "FOSL1", "JUND", "JUN", "JUNB"
)

volcanoplots_LR_ATAC_genes_patients <- disease_markers_lr_patients %>%
  imap( ~ dplyr::mutate(.x,
    new_label = if_else(gene %in% ATAC_genes,
      gene, NA
    ),
    significance_known_label = if_else(gene %in% c(ATAC_genes),
      "ATAC gene", significance
    )
  ) %>%
    ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance_known_label, label = new_label)) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
    geom_point() +
    ggrepel::geom_label_repel(
      size = 2.5,
      label.size = NA,
      label.padding = 0.1,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = 200, colour = "black"
    ) +
    scale_x_continuous("Log2(Fold change)", limits = c(-3.5, 3.5)) +
    scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 300), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3", "ATAC gene" = "steelblue")) +
    theme_pub() +
    labs(title = .y) +
    theme(legend.position = "bottom"))

volcanoplots_LR_ATAC_genes_patients
```

```{r fig.width=22.5, fig.height=16}
volcanoplots_LR_ATAC_genes_patients %>%
  ggarrange(plotlist = ., 
            ncol = 3, 
            nrow = 4, 
            common.legend = TRUE, 
            legend = "bottom")

volcanoplots_LR_ATAC_genes_patients %>%
  ggarrange(plotlist = ., ncol = 4, 
            nrow = 3, 
            common.legend = TRUE, 
            legend = "bottom") %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_LR_ATAC_genes_volcanoplots_all.png"), 
         width = 32, 
         height = 19.2, 
         unit = "cm")
```

##### Quantify DEGs
```{r}
# How many DEGs did we find in total over all patients?
total_degs_lr <- disease_markers_lr_patients %>%
  map(~ filter(., significance %in% c("MBL", "CLL")) %>%
    group_by(significance) %>%
    dplyr::summarise("DEGs" = n())) %>%
  bind_rows() %>%
  summarise("Total DEGs [#]" = sum(DEGs)) %>%
  dplyr::mutate(Fraction = 100) %>%
  ggplot(aes(x = "", y = Fraction, fill = `Total DEGs [#]`)) +
  geom_bar(stat = "identity", width = 1, fill = "#082567") +
  coord_polar("y", start = 0) +
  geom_text(aes(y = 0.0, label = disease_markers_lr_patients %>%
  map(~ filter(., significance %in% c("MBL", "CLL")) %>%
    group_by(significance) %>%
    dplyr::summarise("DEGs" = n())) %>%
  bind_rows() %>%
  summarise("Total DEGs [#]" = sum(DEGs)) %>% pull("Total DEGs [#]")), color = "white", size = 6) +
  theme_void() +
  labs(title = "Total DEG counts\nacross patients") +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 12,
      face = "bold"
    ),
    legend.position = "none"
  )

total_degs_lr

saveRDS(total_degs_lr, file.path(RDS_dir, "total_degs_lr_all.RDS"))
```

```{r}
# How many DEGs did each sample display?
degs_per_patient_lr <- disease_markers_lr_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::summarise("DEGs" = n()) %>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  arrange(desc(DEGs)) %>% 
  dplyr::mutate(patient_id=factor(patient_id, patient_id)) %>% 
  ggplot(aes(x = patient_id, y = DEGs, fill = patient_id)) +
  geom_bar(stat = "Identity") +
  scale_fill_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  scale_x_discrete(labels=patient_id2plotting_id)+
  scale_y_continuous(expand = expansion(mult = c(0, NA))) +
  labs(x = "", y = "DEG counts") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust=0.9),
    legend.position = "none")
degs_per_patient_lr

saveRDS(degs_per_patient_lr, file.path(RDS_dir, "degs_per_patient_lr_all.RDS"))
```

```{r}
# How many genes are uniquely expressed and how many genes are differentially expressed in multiple patients but do follow the same "direction"?
shared_DEG_counts_lr <- disease_markers_lr_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  group_by(gene, significance) %>%
  dplyr::summarise(DEG_in_patients = n()) %>%
  group_by(significance, DEG_in_patients) %>%
  dplyr::summarise(DEG_counts_mult_patients = n()) %>%
  ungroup() %>%
  group_by(DEG_in_patients) %>%
  dplyr::summarise(DEG_counts_mult_patients_dir = sum(DEG_counts_mult_patients)) %>%
  ggplot(aes(x = DEG_in_patients, y = DEG_counts_mult_patients_dir, fill = DEG_counts_mult_patients_dir)) +
  geom_bar(stat = "Identity", fill = "#082567") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(breaks = seq(0,2, 1))+
  labs(x = "Patient counts", y = "DEG counts") +
  theme_pub() +
  theme(legend.position = "right"
  )
shared_DEG_counts_lr

saveRDS(shared_DEG_counts_lr, file.path(RDS_dir, "shared_DEG_counts_lr_all.RDS"))
```

```{r}
common_DEGS_direction_independent_lr <- disease_markers_lr_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL") )%>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  group_by(gene) %>%
  dplyr::summarise(DEG_in_patients = n()) %>%
  dplyr::filter(DEG_in_patients > 2)
common_DEGS_direction_independent_lr

print(paste0(dim(common_DEGS_direction_independent_lr)[1], " genes are differentially expressed in >= 3 patients, without matching effect direction"))

# Which are the genes which are differentially expressed by 3-5 patients with same direction?
common_DEGs_lr <- disease_markers_lr_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  group_by(gene, significance) %>%
  dplyr::summarise(DEG_in_patients = n()) %>%
  dplyr::filter(DEG_in_patients > 2) %>%
  arrange(desc(DEG_in_patients))
common_DEGs_lr

common_DEGs_patients_lr <- disease_markers_lr_patients %>%
  imap( ~ filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::mutate(patient_id = .y)) %>%
    bind_rows() %>%
  split(.$gene)
common_DEGs_patients_lr

saveRDS(common_DEGs_patients_lr, file.path(RDS_dir, "common_DEGs_patients_lr_all.RDS"))
```

#### Wilcoxon rank sum
```{r Subset_Seurat_per_patient}
disease_markers_patients <- list()
for (patient in paste0("CLL", seq(1:10))) {
  gex_scaled_patient <- subset(gex_scaled_all, subset = patient_id == patient)

  disease_markers_patient <- FindMarkers(gex_scaled_patient, ident.1 = "CLL B Cell", ident.2 = "MBL B Cell") %>%
    rownames_to_column(var = "gene") %>%
    dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL"), gene, NA))

  disease_markers_patients[[patient]] <- disease_markers_patient
}

saveRDS(disease_markers_patients, file = file.path(RDS_dir, "disease_markers_patients_all.RDS"))
```

```{r}
volcanoplots_patients <- disease_markers_patients %>%
  imap( ~  ggplot(., aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label=label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, 
    colour = "black", 
    force = 50
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-5, 5)) +
  scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 300), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, 0.09))) +
  scale_colour_manual(name="Differentially expressed genes",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3")) +
  labs(title=patient_id2plotting_id[.y])+
  theme_pub() +
  theme(legend.position = "bottom"))
```

```{r fig.width=22.5, fig.height=16}
volcanoplots_patients %>% 
  ggarrange(plotlist = ., 
            ncol = 4, 
            nrow = 3, 
            common.legend = TRUE, 
            legend = "bottom")

volcanoplots_patients %>%
  ggarrange(plotlist = ., 
            ncol = 4, 
            nrow = 3, 
            common.legend = TRUE, 
            legend = "bottom") %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_volcanoplots_all.png"), 
         width = 32, 
         height = 19.2, 
         unit = "cm",
         bg="white")
```

```{r Volcanoplot_per_patient}
volcanoplots_ATAC_genes_patients <- disease_markers_patients %>%
  imap( ~ dplyr::mutate(.x,
    new_label = if_else(gene %in% ATAC_genes,
      gene, NA
    ),
    significance_known_label = if_else(gene %in% c(ATAC_genes),
      "ATAC gene", significance
    )
  ) %>%
    ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance_known_label, label = new_label)) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
    geom_point() +
    ggrepel::geom_label_repel(
      size = 2.5,
      label.size = NA,
      label.padding = 0.1,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = 200, colour = "black"
    ) +
    scale_x_continuous("Log2(Fold change)", limits = c(-3.5, 3.5)) +
    scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 300), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3", "ATAC gene" = "steelblue")) +
    theme_pub() +
    labs(title = .y) +
    theme(legend.position = "bottom"))

volcanoplots_ATAC_genes_patients
```

```{r fig.width=22.5, fig.height=16}
volcanoplots_ATAC_genes_patients %>%
  ggarrange(plotlist = ., 
            ncol = 4, 
            nrow = 3, 
            common.legend = TRUE, 
            legend = "bottom")

volcanoplots_ATAC_genes_patients %>%
  ggarrange(plotlist = ., 
            ncol = 4, 
            nrow = 3, 
            common.legend = TRUE, 
            legend = "bottom") %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_ATAC_genes_volcanoplots_all.png"), 
         width = 32, 
         height = 19.2, 
         unit = "cm")
```

Summaryplot of all patient specific DEGs
```{r}
volcanoplots_patients_summary <- disease_markers_patients %>%
  bind_rows(.id = "patient_id") %>% 
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id) %>% unique) %>% 
  dplyr::mutate(new_label=if_else(!is.na(label) & -log10(p_val_adj)<200, paste0(label, "_", plotting_id), NA)) %>% 
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label=new_label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
  size = 2.5,
  label.size = NA,
  label.padding = 0.1,
  segment.size = 0.25,
  segment.colour = "grey",
  min.segment.length = 0.1,
  max.overlaps = 200, colour = "black"
) +
  scale_x_continuous("Log2(Fold change)", limits = c(-3.5, 3.5)) +
  scale_y_continuous("-log10(FDR)", 
                   limits = c(0, 300), 
                   oob = scales::squish_infinite, 
                   expand = expansion(mult = c(0, 0.1))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3") )+
    theme_pub() +
    theme(legend.position = "bottom")

volcanoplots_patients_summary

saveRDS(volcanoplots_patients_summary, file = file.path(RDS_dir, "volcanoplots_patients_summary_all.RDS"))
```

```{r}
# MBL-CLL neg = CLL positiv MBL in ATAC Daten
# in ATAC daten auch differenz der mediane berechnet, hier mean fold change!
```

##### Quantify DEGs
```{r}
# How many DEGs did we find in total over all patients?
total_degs <- disease_markers_patients %>%
  map(~ filter(., significance %in% c("MBL", "CLL")) %>%
    group_by(significance) %>%
    dplyr::summarise("DEGs" = n())) %>%
  bind_rows() %>%
  summarise("Total DEGs [#]" = sum(DEGs)) %>%
  dplyr::mutate(Fraction = 100) %>%
  ggplot(aes(x = "", y = Fraction, fill = `Total DEGs [#]`)) +
  geom_bar(stat = "identity", width = 1, fill = "#082567") +
  coord_polar("y", start = 0) +
  geom_text(aes(y = 0.0, label = disease_markers_patients %>%
  map(~ filter(., significance %in% c("MBL", "CLL")) %>%
    group_by(significance) %>%
    dplyr::summarise("DEGs" = n())) %>%
  bind_rows() %>%
  summarise("Total DEGs [#]" = sum(DEGs)) %>% pull("Total DEGs [#]")), color = "white", size = 6) +
  theme_void() +
  labs(title = "Total DEG counts\nacross patients") +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 12,
      face = "bold"
    ),
    legend.position = "none"
  )

total_degs

saveRDS(total_degs, file.path(RDS_dir, "total_degs_all.RDS"))
ggsave(plot = total_degs, filename = file.path(plotting_dir, "total_degs_all.png"), width=2, height=2, unit="in", dpi=300, bg="white")
```

```{r}
# How many DEGs did each sample display?
degs_per_patient <- disease_markers_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::summarise("DEGs" = n()) %>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  arrange(desc(DEGs)) %>% 
  dplyr::mutate(patient_id=factor(patient_id, patient_id)) %>% 
  ggplot(aes(x = patient_id, y = DEGs, fill = patient_id)) +
  geom_bar(stat = "Identity") +
  scale_fill_manual(values = c(my_paired_colours_named, my_unpaired_colours_named)) +
  scale_x_discrete(labels=patient_id2plotting_id)+
  scale_y_continuous(expand = expansion(mult = c(0, NA))) +
  labs(x = "", y = "DEG counts") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust=0.9),
    legend.position = "none")
degs_per_patient

saveRDS(degs_per_patient, file.path(RDS_dir, "degs_per_patient_all.RDS"))
ggsave(plot = degs_per_patient, filename = file.path(plotting_dir, "degs_per_patient.png"), width=4, height=2, unit="in", dpi=300, bg="white")
```

```{r}
# How many genes are uniquely expressed and how many genes are differentially expressed in multiple patients but do follow the same "direction"?
shared_DEG_counts <- disease_markers_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  group_by(gene, significance) %>%
  dplyr::summarise(DEG_in_patients = n()) %>%
  group_by(significance, DEG_in_patients) %>%
  dplyr::summarise(DEG_counts_mult_patients = n()) %>%
  ungroup() %>%
  group_by(DEG_in_patients) %>%
  dplyr::summarise(DEG_counts_mult_patients_dir = sum(DEG_counts_mult_patients)) %>%
  ggplot(aes(x = DEG_in_patients, y = DEG_counts_mult_patients_dir, fill = DEG_counts_mult_patients_dir, label=DEG_counts_mult_patients_dir)) +
  geom_bar(stat = "Identity", fill = "#082567") +
  geom_text(nudge_y =4)+
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(breaks = seq(0,2, 1))+
  labs(x = "Patient counts", y = "DEG counts") +
  theme_pub() +
  theme(legend.position = "right"
  )
shared_DEG_counts

saveRDS(shared_DEG_counts, file.path(RDS_dir, "shared_DEG_counts_all.RDS"))
ggsave(plot = shared_DEG_counts, filename = file.path(plotting_dir, "shared_DEG_counts.png"), width=2, height=2, unit="in", dpi=300, bg="white")
```

```{r}
common_DEGS_direction_independent <- disease_markers_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL") )%>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  group_by(gene) %>%
  dplyr::summarise(DEG_in_patients = n()) %>%
  dplyr::filter(DEG_in_patients > 2)
common_DEGS_direction_independent

print(paste0(dim(common_DEGS_direction_independent)[1], " genes are differentially expressed in >= 3 patients, without matching effect direction"))

# Which are the genes which are differentially expressed by 3-5 patients with same direction?
common_DEGs <- disease_markers_patients %>%
  imap( ~ dplyr::filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::mutate(patient_id = .y)) %>%
  bind_rows() %>%
  group_by(gene, significance) %>%
  dplyr::summarise(DEG_in_patients = n()) %>%
  dplyr::filter(DEG_in_patients > 2) %>%
  arrange(desc(DEG_in_patients))
common_DEGs

common_DEGs_patients <- disease_markers_patients %>%
  imap( ~ filter(.x, significance %in% c("MBL", "CLL")) %>%
    dplyr::mutate(patient_id = .y)) %>%
    bind_rows() %>%
  split(.$gene)
common_DEGs_patients

saveRDS(common_DEGs_patients, file.path(RDS_dir, "common_DEGs_patients_all.RDS"))
```

```{r, fig.width=20, fig.height=45}
all_gene_plots <- list()
for (g in  common_DEGs_patients){
  gene <- g$gene %>% unique()
  patients <- g %>% pull(patient_id)
  print(gene)
  ex_values <-FetchData(object = subset(gex_scaled_all,
                                        subset = patient_id %in% patients & My_annotation %in% c("MBL B Cell", "CLL B Cell")),
                        assay="RNA", 
                        slot="data",
                        vars = gene) %>% dplyr::rename("expression"=1)
  
  gene_plot <- ex_values %>%
    rownames_to_column("label_cellbarcode") %>%
    separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_") %>%
    separate(label, c("patient_id", "sample"), sep = "\\.") %>%
    dplyr::mutate(disease = if_else(`sample` == 1, "MBL", "CLL"),
                  patient_id=factor(patient_id, levels=c(paste0("CLL", 1:10)))) %>%
    dplyr::mutate(disease = factor(disease, levels = c("MBL", "CLL"))) %>%
    group_by(disease) %>%
    ggplot(aes(x = patient_id, y = expression)) +
    geom_violin(aes(fill = disease), trim = FALSE, position = position_dodge(0.9)) +
    geom_boxplot(aes(fill = disease), width = 0.05, position = position_dodge(0.9)) +
    scale_y_continuous(expand = expansion(mult = c(0, NA))) +
    scale_x_discrete(label=patient_id2plotting_id)+
    scale_fill_manual(name = "Disease", values = my_MBL_CLL_colours_named) +
    labs(x = "", fill = "Disease", y = "Expression", title = gene) +
    theme_pub() +
    theme(legend.position = "bottom")
  
  all_gene_plots[[gene]] <- gene_plot
}

ggarrange(plotlist = all_gene_plots, common.legend = TRUE, legend = "bottom", nrow = 13, ncol=6)
saveRDS(all_gene_plots, file = file.path(RDS_dir, "DEGs_patient_expression_all.RDS"))
ggarrange(plotlist = all_gene_plots, common.legend = TRUE, legend = "bottom", nrow = 9, ncol=9)%>% 
ggsave(plot = ., file=file.path(plotting_dir, "DEGs_patient_expression_all.png"), width=31.5, height=31.5, unit="in", dpi=300, bg="white")
```


##### DEGs across modalities (scATACseq)

```{r, fig.width=8, fig.height=4}
DEGS_union_ATAC_heatmap <- disease_markers_patients %>%
  imap( ~ filter(., significance == "significant" & gene %in% ATAC_genes) %>%
    dplyr::mutate(
      patient_id = .y,
      "Counts [#]" = 1
    )) %>%
  bind_rows() %>% 
  left_join(ATAC_accessibility) %>% 
  dplyr::mutate(ATAC_GEX_relation= case_when(avg_log2FC > 0 & accessibility > 0 ~ "increased in CLL",
                                             avg_log2FC < 0 & accessibility < 0 ~ "increased in MBL",
                                             .default = "contradictory relationship"))%>%
  dplyr::select(gene, patient_id, ATAC_GEX_relation) %>%
  dplyr::mutate(gene = factor(gene, levels = sort(c("CTCF", "CTCFL", "BATF", "NFE2", "BACH2", "JDP2", "FOSB", "BACH1", "FOS", "FOSL2", "SMARCC1", "FOSL1", "JUND", "JUN", "JUNB"))),
                ATAC_GEX_relation=factor(ATAC_GEX_relation, levels=c("increased in MBL", "increased in CLL", "contradictory relationship"))) %>%
  ggplot(aes(x = patient_id, y = gene, fill = ATAC_GEX_relation)) +
  geom_tile() +
  scale_fill_manual(name = "", values = c(
    "increased in MBL" = "#FC8D62",
    "increased in CLL" = "#66C2A5",
    "contradictory relationship" = "lightgrey"
  )) +
  theme_pub() +
  labs(x = "", y = "Gene", title="Chromatin accessibility & gene expression") +
  theme(
    axis.ticks = element_blank(),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    legend.position = "right")

DEGS_union_ATAC_heatmap

DEGS_union_ATAC_heatmap %>%
  ggsave(., file = file.path(plotting_dir, "ATAC_Genes_Heatmap_all.png"), 
         width = 15, 
         height = 6, 
         unit = "cm")
```

```{r, fig.width=8, fig.height=4}
deg_quantification_1 <- ggarrange(
  total_degs,
  degs_per_patient,
  shared_DEG_counts,
  nrow = 1,
  ncol = 3,
  labels = c("A", "B", "C" ))

deg_quantification_1 %>% 
ggsave(file = file.path(plotting_dir, "DEG_quantification_figure_HC_MBL_CLL.png"), 
         width = 10, 
         height = 2, units = "in", bg="white")
```

##### TF activity analysis
```{r}
net <- get_collectri(organism='human', split_complexes=FALSE)
net

calc_activity <- function(all_patient_seurat_object, patient){
  print(patient)
  seurat_fract <- subset(gex_scaled_all, subset = patient_id == patient)
  mat <- as.matrix(seurat_fract@assays$RNA@data)
  acts <- run_wmean(mat=mat, net=net, .source='source', .target='target',
                  .mor='mor', times = 90, minsize = 5)
  #so far required too much memory
  #acts <- run_ulm(mat=mat, net=net, .source='source', .target='target',)
  
}

TF_activity <- unique(gex_scaled_all@meta.data$patient_id) %>%
  map(~calc_activity(all_patient_seurat_object=gex_scaled_all, patient=.)) %>%
  setNames(unique(gex_scaled_all@meta.data$patient_id))

saveRDS(TF_activity, file.path(RDS_dir, "TF_activity_all.RDS"))
```


```{r}
transform2seurat <- function(all_patient_seurat_object, pat, act_list){
  # all_patient_seurat_object=gex_scaled_all
  # pat=unique(gex_scaled_all@meta.data$patient_id)[6]
  # act_list=TF_activity_list
  print(pat)
  seurat_fract <- subset(all_patient_seurat_object, subset = patient_id == pat)
  act <- act_list[[pat]]
  
  tfswmean_assay <- act %>%
    dplyr::filter(statistic == 'norm_wmean') %>%
    tidyr::pivot_wider(id_cols = 'source', names_from = 'condition',
              values_from = "score") %>%
    column_to_rownames("source") %>%
    Seurat::CreateAssayObject()

  seurat_fract[['tfswmean']] <-tfswmean_assay
    
  # Change assay
  DefaultAssay(seurat_fract) <- "tfswmean"

  # Scale the data
  seurat_fract <- ScaleData(seurat_fract)
  seurat_fract@assays$tfswmean@data <- seurat_fract@assays$tfswmean@scale.data
  return(seurat_fract)
}

#CLL9 is not included yet, as required too much memory TODO rerun on cluster or with only 90 permutations
TF_activity_seurat <- unique(gex_scaled_all@meta.data$patient_id) %>% 
  map(~transform2seurat(all_patient_seurat_object=gex_scaled_all, 
                        pat=.x, 
                        act_list=TF_activity)) %>% 
  set_names(unique(gex_scaled_all@meta.data$patient_id))

#saveRDS(TF_activity_seurat, file.path(RDS_dir, "TF_activity_seurat_all.RDS"))
```

```{r fig.width=30, fig.height=90}
plot_TF_activity <- function(seurat_activity_list){
  umap_patient <- DimPlot(seurat_activity_list, 
                          reduction = "umap", 
                          label = TRUE, 
                          pt.size = 0.5, 
                          alpha=0.5,
                          cols = annotation_colours_named) + 
  NoLegend() + 
  labs(x="UMAP1", y="UMAP2")+
  ggtitle('Cell types')+
  theme_pub()+
  theme(legend.position = "top",
        legend.direction = "horizontal")
  
  boxplots_patient <- as.matrix(seurat_activity_list@assays$tfswmean@data) %>% 
    as_tibble(rownames="TF") %>% 
    tidyr::pivot_longer(- "TF", names_to = "label_cellbarcode", values_to = "TF activity" ) %>% 
    left_join(seurat_activity_list@meta.data  %>% 
                rownames_to_column("label_cellbarcode") %>% 
                dplyr::select(label_cellbarcode, My_annotation, malignance)) %>% 
    dplyr::filter(malignance ==TRUE,
                  TF %in% ATAC_genes) %>% 
    dplyr::mutate(My_annotation=factor(My_annotation, levels=c("MBL B Cell", "CLL B Cell"))) %>% 
    dplyr::group_by(My_annotation) %>% 
    ggplot(data = ., aes(y=`TF activity`, x=My_annotation, fill=My_annotation))+
    geom_violin(position = position_dodge(0.9))+
    geom_boxplot(width = 0.2, position = position_dodge(0.9))+
    ggpubr::stat_compare_means(aes(label = ..p.signif..), 
                        label.x = 1.5, 
                       label.y = 10,
                       method="wilcox.test",
                       paired = FALSE)+
    theme_pub()+
    labs(x="")+
    scale_fill_manual(name= "Celltype", values =c("MBL B Cell"= "#FC8D62", "CLL B Cell"= "#66C2A5" ))+
    theme(axis.text.x = element_text(hjust=0.5))+
    facet_wrap(~TF, nrow=1)
  
  both_plots <- cowplot::plot_grid(umap_patient, boxplots_patient, rel_widths = c(3,15), axis = "b")
  
  return(both_plots)
}

TF_activity_seurat %>% 
  map(plot_TF_activity) %>%
  iwalk(~ ggsave(
    filename = paste0(analysis_dir, "/Plots/TF_activity/TF_activities_", .y, ".png"),
    plot = .x,
    width = 30,
    height = 5,
    device="png"
  ))
  
TF_activity_seurat_figure <- TF_activity_seurat %>% 
  map(plot_TF_activity) %>%
  ggarrange(plotlist = ., nrow = 18, ncol=1) 

TF_activity_seurat_figure %>% 
ggsave(path = plotting_dir,
    filename = "TF_activities_summary_plot_all.png",
    plot = .,
    width = 30,
    height = 5*18,
    device="png",
    limitsize = FALSE)
```

Dotplot activities
```{r, fig.width=10, fig.height=4}
get_TF_activity_data <- function(seurat_activity_list){

  TF_data <- as.matrix(seurat_activity_list@assays$tfswmean@data) %>% 
    as_tibble(rownames="TF") %>% 
    tidyr::pivot_longer(- "TF", names_to = "label_cellbarcode", values_to = "TF activity" ) %>% 
    left_join(seurat_activity_list@meta.data  %>% 
                rownames_to_column("label_cellbarcode") %>% 
                dplyr::select(label_cellbarcode, My_annotation, malignance)) %>% 
    dplyr::filter(malignance ==TRUE,
                  TF %in% ATAC_genes) %>% 
    dplyr::mutate(My_annotation=factor(My_annotation, levels=c("MBL B Cell", "CLL B Cell"))) %>% 
    dplyr::group_by(My_annotation) 
  return(TF_data)
}

TF_data_pvalues <- TF_activity_seurat[-c(5,6,7,8,12,13, 16:24)] %>% 
    imap( ~get_TF_activity_data(.x) %>% 
         dplyr::mutate(patient_id=.y) %>% 
         group_by(TF) %>%
      do(p_value = wilcox.test(filter(., My_annotation =="MBL B Cell") %>% 
                                 pull(`TF activity`), 
                               filter(., My_annotation =="CLL B Cell") %>% 
                                 pull(`TF activity`), 
                               paired = FALSE)$p.value) %>%
      dplyr::mutate(
        p_value = unlist(p_value),
        patient_id = .y
      ) %>%
      dplyr::mutate("p_adjust" = p.adjust(p_value, method = "BH", n=length(.$p_value)))
  ) %>% 
  bind_rows(.id = "patient_id")

TF_data_median <- TF_activity_seurat[-c(5,6,7,8,12,13, 16:24)] %>% 
  imap( ~get_TF_activity_data(.x) %>% 
         dplyr::mutate(patient_id=.y) %>% 
         tidyr::separate(col = label_cellbarcode, into = c("label", "cellbarcode"), sep = "_")%>% 
         group_by(label, TF) %>%
    summarise_at(vars(`TF activity`), list(`TF activity median` = median)) %>% 
  tidyr::separate(label, c("patient_id", NA), sep="\\.", remove = FALSE) %>% 
  dplyr::mutate(disease=if_else(str_ends(label, "\\.1"), "MBL", "CLL")) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(-label))%>% 
  bind_rows()

statistics_TF_data<- TF_data_median %>%
  left_join(TF_data_pvalues, by = join_by("TF", "patient_id")) %>% 
  pivot_wider(names_from  =  disease, values_from = `TF activity median`)


patient_order <- statistics_TF_data %>%
    dplyr::mutate(patient_id=factor(patient_id, levels=rev(paste0("CLL", 1:10))),
                  diff_TF_act=MBL-CLL,
                  p_adjust=if_else(p_adjust >=0.05, NA, p_adjust)) %>% 
  group_by(patient_id) %>% 
  summarize(order=mean(abs(diff_TF_act))) %>% 
  arrange(order) %>% 
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id) %>% unique) %>% 
  pull(plotting_id)

TF_order <- statistics_TF_data %>%
    dplyr::mutate(patient_id=factor(patient_id, levels=rev(paste0("CLL", 1:10))),
                  diff_TF_act=MBL-CLL,
                  p_adjust=if_else(p_adjust >=0.05, NA, p_adjust)) %>% 
  group_by(TF) %>% 
  summarize(order=mean(abs(diff_TF_act))) %>% 
  arrange(order) %>% 
  pull(TF)


## if values are negative then is inactive
TF_act_dotplot <- statistics_TF_data %>%
  dplyr::left_join(metadata %>% dplyr::select(patient_id, plotting_id) %>% unique) %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, levels=patient_order),
                diff_TF_act=MBL-CLL,
                p_adjust=if_else(p_adjust >=0.05, NA, p_adjust),
                TF = factor(TF, levels=rev(TF_order))) %>% 
  {ggplot(., aes(x=TF, y=plotting_id, colour= diff_TF_act, size=-log10(`p_adjust`)))+
  geom_point()+
  scale_color_gradient2(name="TF activity difference", low="#66C2A5", mid="white", high="#FC8D62", midpoint=0, limits=c(-1.5,1.5))+
  scale_size_area(name="-log10(FDR)", limits=c(NA,20), oob=scales::squish)+
  labs(x="TF", y="")+
  theme_pub()+
  theme(axis.text.x = element_text(angle=45, hjust=1, vjust=0.9),
        panel.background = element_rect(fill="snow3"))}

TF_act_dotplot

saveRDS(TF_act_dotplot, file = file.path(project_dir, "RDS_objects/TF_act_dotplot_all.RDS"))
```

### Compare non-progressive vs. progressive MBLs
```{r Subset_only_MBL_cells}
gex_scaled_MBLs <- subset(x=gex_scaled_all, subset = 
                            MBL_CLL_subtype == "HC" & My_annotation == "MBL B Cell" & patient_id !="MBL7dup")

gex_scaled_MBLs@meta.data %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  group_by(label) %>%
  dplyr::summarise("cell_counts" = n())
```

```{r Progression_annotation}
# Add progression state annotation
progression_annot <- tibble("label_cellbarcode" = WhichCells(gex_scaled_MBLs)) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  tidyr::separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE) %>%
  dplyr::mutate(progression = if_else(str_starts(patient_id, paste0("MBL", c(1:3, 6:9), collapse = "|")), FALSE, TRUE)) %>%
  dplyr::select(label_cellbarcode, progression) %>%
 deframe()

gex_scaled_MBLs <- AddMetaData(object = gex_scaled_MBLs, metadata = progression_annot, col.name = "progression")
```

```{r Downsample_smallest_sample}
Idents(object = gex_scaled_MBLs) <- "patient_id"
gex_scaled_MBLs_downsampled <- subset(gex_scaled_MBLs, downsample = 543)
Idents(object = gex_scaled_MBLs) <- "My_annotation"
Idents(object = gex_scaled_MBLs_downsampled) <- "progression"
```

```{r Markers_MBL_prog_nonprog}
progression_markers <- FindMarkers(gex_scaled_MBLs_downsampled, ident.1 = "TRUE", ident.2 = "FALSE") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "progression",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "no progression",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("progression", "no progression"), gene, NA))

saveRDS(progression_markers, file = file.path(RDS_dir, "progression_markers_patients_all.RDS"))
```

```{r}
# Positive values indicate higher expression in first ident.1 group so progressive MBLs
volcanoplot_progression <- progression_markers %>%
  dplyr::mutate(
    new_label = if_else(gene %in% ATAC_genes, gene, NA),
    significance_known_label = if_else(gene %in% c(ATAC_genes), "ATAC gene", significance)
  ) %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance_known_label, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 1.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 200, colour = "black"
  ) +
    scale_x_continuous("Log2(Fold change)", limits = c(-5.5, 5.5)) +
    scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 300), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("progression"="#66C2A5", "no progression"="#FC8D62", "not significant"="#D3D3D3", "ATAC gene" = "steelblue")) +
    theme_pub() +
    theme(legend.position = "bottom")

volcanoplot_progression

volcanoplot_progression %>%
  ggsave(., 
         file = file.path(analysis_dir, "/Plots/Differential_expression_progression_downsampled_volcanoplot_all.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

```{r}
progression_markers_signif <- progression_markers %>%
  dplyr::filter(significance %in% c("progression", "no progression")) %>%
  dplyr::mutate(enriched_expression = if_else(avg_log2FC < 0, "non-progressive MBL", "progressive MBL")) %>%
  group_split(enriched_expression) %>%
  map(~ pull(., gene)) %>%
  set_names(c("non-progressive", "progressive"))

progression_markers_signif
```

### Compare two MBL8 Clusters
```{r Subsetting_MBL8}
# How does MBL38 differ in its two clusters? Cluster sizes are equal in size
MBL8 <- subset(gex_scaled_all, cells = gex_scaled_all@meta.data %>% 
                 filter(patient_id == "MBL8" & seurat_clusters %in% c(26, 31, 35)) %>% 
                 rownames())
Idents(MBL8) <- "My_annotation"

# Add column for clusters to compare
# Add disease state annotation
selected_clusters_annot <- MBL8@meta.data %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  dplyr::mutate(selected_clusters = if_else(seurat_clusters %in% c(31, 35), "Cluster 31,35", "Cluster 26")) %>%
  with(setNames(selected_clusters, label_cellbarcode))

MBL8 <- AddMetaData(object = MBL8, metadata = selected_clusters_annot, col.name = "selected_clusters")
Idents(MBL8) <- "selected_clusters"
```

```{r FindMarkers_MBL8}
MBL8_markers <- FindMarkers(MBL8, ident.1 = "Cluster 26", ident.2 = "Cluster 31,35") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "Cluster 26",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "Cluster 31,35",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("Cluster 26", "Cluster 31,35"), gene, NA))


saveRDS(MBL8_markers, file = file.path(RDS_dir, "MBL8_markers_all.RDS"))
```

```{r}
volcanoplot_MBL8 <- MBL8_markers %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black") +
  scale_x_continuous("Log2(Fold change)", limits = c(-5, 5)) +
  scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("Cluster 26"="lightpink", "Cluster 31,35"="darkgreen", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_MBL8


MBL8_genes <- MBL8_markers %>%
  filter(significance %in% c("Cluster 26", "Cluster 31,35")) %>%
  arrange(desc(avg_log2FC)) %>%
  pull(gene)

saveRDS(MBL8_genes, file = file.path(RDS_dir, "MBL8_genes_all.RDS"))

volcanoplot_MBL8 %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_MBL8_volcanoplot_all.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

## LC-MBL vs. CLL patients' cells
### All cells
```{r }
disease_markers_LC <- FindMarkers(subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("LC", "CLL")), 
                               ident.1 = "CLL B Cell", ident.2 = "MBL B Cell", 
                               test.use = "wilcox") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
                                         p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
                                         .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL"), gene, NA))

saveRDS(disease_markers_LC, file = file.path(RDS_dir, "disease_markers_LC.RDS"))
```

### Downsampled n=2118

```{r}
# Determine number to downsample to
min_malignant_cells_LC <- subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("LC", "CLL"))@meta.data %>%
  rownames_to_column(var = "label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  mutate(malignance = if_else(My_annotation %in% c("MBL B Cell", "CLL B Cell"), TRUE, FALSE)) %>%
  group_by(label, malignance) %>%
  dplyr::summarise("cell_counts" = n()) %>%
  dplyr::filter(malignance == TRUE) %>%
  arrange(cell_counts) %>%
  pull(cell_counts) %>%
  min()

min_malignant_cells_LC
```

```{r Downsampling_all_patients}
gex_scaled_downsampled_LC <- subset(gex_scaled_all, subset = malignance == TRUE, downsample = min_malignant_cells_LC)

# Check if evey sample has now min_malignant_cells
# gex_scaled_downsampled@meta.data %>%
# rownames_to_column(var = "label_cellbarcode") %>%
#   tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
#   mutate(malignance=if_else(My_annotation %in% c("MBL B Cell", "CLL B Cell"), TRUE, FALSE)) %>%
#   group_by(label, malignance) %>%
#   dplyr::summarise("cell_counts" = n()) %>%
#   dplyr::filter(malignance ==TRUE)
```

```{r FindMarkers_downsampled_all_prog_patients}
#currently filtered for at least avglog2FC > 2 and non IG genes
disease_markers_downsampled_LC <- FindMarkers(gex_scaled_downsampled_LC, 
                                           ident.1 = "CLL B Cell", 
                                           ident.2 = "MBL B Cell", 
                                           test.use="wilcox") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL") & !str_starts(gene, "IG"), gene, NA))

saveRDS(disease_markers_downsampled_LC, file = file.path(RDS_dir, "disease_markers_downsampled_LC.RDS"))

disease_markers_lr_downsampled_LC <- FindMarkers(gex_scaled_downsampled_LC, 
                                              ident.1 = "CLL B Cell", 
                                              ident.2 = "MBL B Cell", test.use = "LR") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL") & !str_starts(gene, "IG"), gene, NA))
```

```{r Volcanoplot_all_prog_patients_downsampled, fig.height=4, fig.width=4}
#CLL right, LC-MBL left
volcanoplot_patients_downsampled_LC <- disease_markers_downsampled_LC %>%
  dplyr::mutate(significance=factor(significance, levels=c("MBL", "CLL", "not significant"))) %>% 
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label=label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 150, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-5, 5)) +
  scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#fcbda4", "not significant"="#D3D3D3"),
                      labels=c("CLL"="CLL", "MBL"="LC-MBL")) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_patients_downsampled_LC

saveRDS(volcanoplot_patients_downsampled_LC, file.path(project_dir, "RDS_objects/volcanoplot_patients_downsampled_LC.RDS"))

volcanoplot_patients_downsampled_LC %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_downsampled_volcanoplot_LC.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```

### Pseudobulk AggregateExpression()

```{r}
# Subset to only MBL and CLL cells and ensure only progressive MBLs are included
gex_scaled_malignant_LC <- subset(subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("LC", "CLL") & 
                                        My_annotation %in% c("MBL B Cell", "CLL B Cell")))

# Aggregate cells per patient sample name (either MBL or CLL)
gex_scaled_pseudobulk_LC <- AggregateExpression(
  gex_scaled_malignant_LC,
  return.seurat = TRUE,
  group.by = c("orig.ident"),
  verbose = TRUE
)

# Extract from sample name if MBL or CLL sample
disease_LC <- gex_scaled_pseudobulk_LC@meta.data %>%
  rownames_to_column("label") %>%
  dplyr::select(-disease) %>% 
  dplyr::left_join(metadata %>% 
                     dplyr::select(label, disease)) %>% 
  dplyr::select(label, disease) %>% 
  deframe()

# Add disease state per sample as metadata
gex_scaled_pseudobulk_LC <- AddMetaData(object = gex_scaled_pseudobulk_LC, metadata = disease_LC, col.name = "disease")
```

```{r}
Idents(gex_scaled_pseudobulk_LC) <- "disease"
disease_markers_pseudobulk_LC <- FindMarkers(gex_scaled_pseudobulk_LC, ident.1 = "CLL", ident.2 = "MBL") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "CLL",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "MBL",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("MBL", "CLL"), gene, NA))
saveRDS(disease_markers_pseudobulk_LC, file = file.path(RDS_dir, "disease_markers_pseudobulk_LC.RDS"))
```

```{r}
volcanoplot_pseudobulk_LC <- disease_markers_pseudobulk_LC %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-12, 12)) +
  scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("CLL"="#66C2A5", "MBL"="#FC8D62", "not significant"="#D3D3D3")) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_pseudobulk_LC

volcanoplot_pseudobulk_LC %>%
  ggsave(., file = file.path(plotting_dir, "Differential_expression_pseudobulk_volcanoplot_LC.png"), 
         width = 12, 
         height = 10, 
         unit = "cm")
```


## LC-MBL vs HC-MBL
### All cells
```{r }
Idents(gex_scaled_all) <- "MBL_CLL_subtype"
disease_markers_LC_HC <- FindMarkers(subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("LC", "HC") & My_annotation_B %in% c("MBL B Cell")), 
                               ident.1 = "LC", ident.2 = "HC", 
                               test.use = "wilcox") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "HC",
                                         p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "LC",
                                         .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("LC", "LC"), gene, NA))

saveRDS(disease_markers_LC_HC, file = file.path(RDS_dir, "disease_markers_LC_HC.RDS"))
```


```{r}
# Determine number to downsample to
min_malignant_cells_LC_HC <- subset(x=gex_scaled_all, 
                                      subset = MBL_CLL_subtype %in% c("LC", "HC") & My_annotation_B %in% c("MBL B Cell"))@meta.data %>%
  group_by(orig.ident, MBL_CLL_subtype) %>%
  dplyr::summarise("cell_counts" = n()) %>%
  arrange(cell_counts) %>%
  dplyr::filter(orig.ident != "CLL6.1") %>% 
  pull(cell_counts) %>%
  min()

min_malignant_cells_LC_HC
```

## Downsampled
```{r}
gex_scaled_downsampled_LC_HC <- subset(gex_scaled_all,
  subset = MBL_CLL_subtype %in% c("LC", "HC") & My_annotation_B %in% c("MBL B Cell") & orig.ident %in% c("CLL1.1", "CLL10.1", "CLL2.1", "CLL3.1", "CLL4.1", "CLL5.1", "CLL7.1", "CLL8.1", "CLL9.1", "MBL1", "MBL10", "MBL11", "MBL12", "MBL13", "MBL14", "MBL15", "MBL2", "MBL3", "MBL4", "MBL6", "MBL7", "MBL8", "MBL9"),
  downsample = min_malignant_cells_LC_HC)
```

```{r}
disease_markers_downsampled_LC_HC <- FindMarkers(gex_scaled_downsampled_LC_HC, 
                                           ident.1 = "LC", 
                                           ident.2 = "HC", 
                                           test.use="wilcox") %>%
  rownames_to_column(var = "gene") %>%
  dplyr::mutate(significance = case_when(p_val_adj <= 0.05 & avg_log2FC >= 1 ~ "HC",
      p_val_adj <= 0.05 & avg_log2FC <= -1 ~ "LC",
      .default = "not significant")) %>%
  dplyr::mutate(label = if_else(significance %in% c("LC", "HC") & !str_starts(gene, "IG"), gene, NA))

saveRDS(disease_markers_downsampled_LC_HC, file = file.path(RDS_dir, "disease_markers_downsampled_LC_HC.RDS"))
```

```{r}
volcanoplot_LC_HC <- disease_markers_downsampled_LC_HC %>%
  dplyr::mutate(significance=factor(significance, levels=c("LC", "HC", "not significant")) )%>% 
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), colour = significance, label = label)) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(-log10(0.05)), color = "grey", linetype = "dashed") +
  geom_point() +
  ggrepel::geom_label_repel(
    size = 2.5,
    label.size = 0,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 25, colour = "black"
  ) +
  scale_x_continuous("Log2(Fold change)", limits = c(-4, 4)
                     ) +
  scale_y_continuous("-log10(FDR)", 
                     limits = c(0, 350), 
                     oob = scales::squish_infinite, 
                     expand = expansion(mult = c(0, NA))) +
  scale_colour_manual(name="DEG",
                      values=c("LC"="#fcbda4", "HC"="#FC8D62", "not significant"="#D3D3D3"),
                      labels=c("LC"="LC-MBL", "HC"="HC-MBL", "not significant"="not significant")) +
  theme_pub() +
  theme(legend.position = "bottom")

volcanoplot_LC_HC

saveRDS(volcanoplot_LC_HC, file.path(RDS_dir, "volcanoplot_LC_HC.RDS"))


```

# Export data for batch correction
```{r}
SaveH5Seurat(gex_scaled_all, 
             filename =file.path(RDS_dir,  "gex_scaled_all.h5Seurat"))

Convert(source=file.path(RDS_dir, "gex_scaled_all.h5Seurat"), 
        dest = file.path(RDS_dir, "gex_scale_all.h5ad"))
```

# Non integrated IG genes removed
```{r}
gex_scaled_all <- readRDS(file = file.path(RDS_dir, "gex_scaled_all.RDS"))
```

### Subset
```{r}
# identify all IG genes
features <- rownames(gex_scaled_all)
ig_pattern <- "^IG(A|D|E|G|H|K|L|M|KV|LV)"
ig_genes <- features[str_detect(features, ig_pattern)]

gex_scaled_all_noIG <- subset(x=gex_scaled_all,
                                         features=setdiff(features, ig_genes))
```

```{r}
gex_scaled_all_noIG <- SCTransform(gex_scaled_all_noIG, verbose = TRUE)

saveRDS(gex_scaled_all_noIG, file = file.path(RDS_dir, "gex_scaled_all_noIG.RDS"))
```

### LDR
```{r}
gex_scaled_all_noIG <- RunPCA(gex_scaled_all_noIG,
                                    features = VariableFeatures(object = gex_scaled_all_noIG))

ElbowPlot(gex_scaled_all_noIG) %>%
  ggsave(., filename = file.path(plotting_dir, "Elbow_plot_no_IG.png"),
         unit = "in",
         width = 6,
         height = 4)

ElbowPlot(gex_scaled_all_noIG)
```

### Clustering and UMAP
```{r}
gex_scaled_all_noIG <- FindNeighbors(gex_scaled_all_noIG, dims = 1:10) %>%
  FindClusters(resolution = 0.5)
```

```{r}
set.seed(123)
gex_scaled_all_noIG <- RunUMAP(gex_scaled_all_noIG, dims = 1:10)
```

```{r}
#saveRDS(gex_scaled_all_noIG, file = file.path(RDS_dir, "gex_scaled_all_noIG.RDS"))
gex_scaled_all_noIG <- readRDS(file = file.path(RDS_dir, "gex_scaled_all_noIG.RDS"))
```

### Plot UMAPs
```{r}
color_palette <- colorRampPalette(brewer.pal(12, "Set3"))(60)
```

#### Clusters
```{r}
umap_clusters_plot_no_IG <- DimPlot(gex_scaled_all_noIG,
                                    label = TRUE,shuffle = TRUE,
                              reduction = "umap",
                              raster = FALSE,
                              group.by = c("seurat_clusters"),
                              cols=color_palette)

umap_clusters_plot_no_IG %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_clusters.png"),
         unit = "in",
         width = 8,
         height = 6)

umap_clusters_plot_no_IG
```

#### Samples
```{r}
# add plotting id
plotting_annot_no_IG <- as_tibble(gex_scaled_all_noIG$orig.ident, rownames=NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::rename("label"="value") %>%
  left_join(metadata %>%
              dplyr::select(label, plotting_id)) %>%
  dplyr::select(label_cellbarcode, plotting_id) %>%
  deframe()

gex_scaled_all_noIG <- AddMetaData(object = gex_scaled_all_noIG,
                                         metadata = plotting_annot_no_IG,
                                         col.name = "plotting_id")
```

```{r}
umap_sample_plot_no_IG <- DimPlot(gex_scaled_all_noIG,
                            reduction = "umap",
                            raster = FALSE,
                            group.by = c("orig.ident"),
                            cols=c(my_unpaired_colours_named,my_paired_colours_named, control_colours_named)) +
  theme_pub() +
  theme(
    aspect.ratio = 1,
    legend.position = "right"
  )+
  labs(title="Samples")+
  theme_pub()+
  theme(aspect.ratio = 1,
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y=element_blank())

umap_sample_plot_no_IG %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_samples.png"),
         unit = "in",
         width = 8,
         height = 6)

umap_sample_plot_no_IG
```

#### Patients 
```{r}
gex_scaled_all_noIG$patient_id <- factor(gex_scaled_all_noIG$patient_id, levels=gex_scaled_all_noIG@meta.data %>% dplyr::select(patient_id, plotting_id) %>% unique %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, levels=c(paste0("P", 1:9), "P9dup", paste0("P", 10:24)))) %>% 
  dplyr::arrange(plotting_id) %>% 
  dplyr::mutate(patient_id=factor(patient_id, patient_id)) %>% 
  .$patient_id %>% levels)
```

```{r}
nonMBL7dup_cells <- colnames(gex_scaled_all_noIG)[-c(str_starts(colnames(gex_scaled_all_noIG), "MBL7dup"))]
```

```{r}
umap_data_no_immune <- gex_scaled_all_noIG@reductions$umap@cell.embeddings %>% as.data.frame() %>% 
  rownames_to_column("label_cellbarcode") %>% 
  left_join(gex_scaled_all_noIG@meta.data[c("orig.ident", "patient_id","MBL_CLL_subtype", "My_annotation_LC")] %>%
              rownames_to_column("label_cellbarcode")) %>% 
  left_join(metadata %>% dplyr::select(label, sequencing_date, plotting_id), by=join_by(orig.ident ==label)) %>% 
  dplyr::rename("label"="orig.ident") %>% 
  dplyr::filter(label != "MBL7dup") %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, levels=c(paste0("P", 1:24),paste0("H", 1:4)))) %>% 
  dplyr::arrange(plotting_id) %>% 
  dplyr::mutate(patient_id=factor(patient_id, levels=unique(patient_id)))%>% 
  dplyr::filter(!(My_annotation_LC %in% c("T", "other", "Mono", "NK", "DC")))

centroids <- umap_data_no_immune %>% 
  dplyr::select(UMAP_1, UMAP_2, patient_id) %>% 
  group_by(patient_id) %>% 
  dplyr::summarize(across(everything(), mean)) %>% 
  dplyr::left_join(metadata %>% dplyr::filter(data_type=="GEX") %>% dplyr::select(patient_id, plotting_id)) %>% unique()
```


```{r}
umap_patient_no_IG <- DimPlot(gex_scaled_all_noIG,
                              cells=nonMBL7dup_cells,
                        reduction = "umap",
                        group.by = "patient_id",
                        raster = F,
                        cols = my_unpaired_colours_named,
                        shuffle = TRUE) +
  geom_text_repel(data=centroids, aes(x=UMAP_1, y=UMAP_2, label=plotting_id), max.overlap=0, force=500)+
  scale_colour_manual(values=my_unpaired_colours_named, label=patient_id2plotting_id)+
  labs(x = "UMAP1", y = "UMAP2", title = "") +
  theme_pub() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "bottom",
        aspect.ratio=1)

write_tsv(umap_patient_no_IG[[1]][["data"]], file = file.path(tsv_dir, "GEX_patient_annotation_noIG_umap.tsv"))

write_tsv(centroids, file = file.path(tsv_dir, "GEX_patient_annotation_noIG_umap_centroids.tsv"))

umap_patient_no_IG %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_patients.png"),
         unit = "in",
         width = 5,
         height = 5)

umap_patient_no_IG
```

#### Patients disease-split
```{r}
umap_patient_disease_no_IG <- DimPlot(gex_scaled_all_noIG,
                                reduction = "umap",
                                group.by = "patient_id",
                                raster = F,
                                cols = c(my_unpaired_colours_named, control_colours_named),
                                split.by = "disease",
                                shuffle = TRUE) +
  scale_colour_manual(values=my_unpaired_colours_named, label=patient_id2plotting_id)+
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Disease") +
  theme_pub() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "bottom",
        aspect.ratio = 1
  )

umap_patient_disease_no_IG %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_patients_disease_split.png"),
         unit = "in",
         width = 12,
         height = 6)

umap_patient_disease_no_IG
```

#### Patients disease-subtype-split
```{r}
umap_patient_MBL_CLL_subtype_no_IG <- DimPlot(gex_scaled_all_noIG,
                                        reduction = "umap",
                                        group.by = "patient_id",
                                        raster = F,
                                        cols = c(my_unpaired_colours_named, control_colours_named),
                                        split.by = "MBL_CLL_subtype",
                                        shuffle = TRUE) +
  scale_colour_manual(values=my_unpaired_colours_named, label=patient_id2plotting_id)+
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Patients") +
  theme_pub() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "bottom",
        aspect.ratio = 1
  )

umap_patient_MBL_CLL_subtype_no_IG %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_patients_disease_subtype_split.png"),
         unit = "in",
         width = 16,
         height = 6)

umap_patient_MBL_CLL_subtype_no_IG
```

#### Celltype markers
```{r}
marker_genes <- c(
  "PAX5", "MS4A1", "EBF1", "MME", # B-cell trajectories
  "CD14", "CEBPB", "MPO", "CD33", # Monocytes
  "CD3D", "CD8A", "TBX21", "IL7R", "CD3E", "CD3G", "CD4", "CD8B", # TCells
  "CD19", "CD5", "FCER2", "CD79B", # CLL B-Cell Marker
  "GNLY", "NKG7", "NCAM1", # NK cells
  "PF4", # platelets
  "CA1", "HBA1" # erythrocytes
)
```

```{r fig.width=20, fig.height=15}
marker_vln_no_IG <- VlnPlot(gex_scaled_all_noIG, 
                            features = marker_genes, 
                            group.by = "seurat_clusters",
                            slot = "scale.data",
                            pt.size = FALSE,
                            cols = color_palette
                            ) +
  theme_pub() +
  theme(legend.position = "none")

marker_vln_no_IG %>%
  ggsave(., file = file.path(plotting_dir, "Marker_expression_vlnplot_all_no_IG.png"),
         units = "in",
         width = 20,
         height = 16)

marker_vln_no_IG
```

```{r, fig.width=20, fig.height=15}
marker_umaps_no_IG <- FeaturePlot(gex_scaled_all_noIG,
                                  reduction = "umap",
                                  features = marker_genes, 
                                  raster = FALSE) +
  theme_pub() +
  theme(legend.position = "none")

marker_umaps_no_IG %>%
  ggsave(., file = file.path(plotting_dir, "Marker_expression_umaps_all_no_IG.png"),
         units = "in",
         width = 12,
         height = 16)

marker_gene_expression_umap_data <- list()
for (marker in (1:26)){
  print(marker)
  marker_gene_expression_umap_data[[marker]] <- marker_umaps_no_IG[[marker]][["data"]]
}

marker_gene_expression_umap_data <- marker_gene_expression_umap_data %>% 
  purrr::set_names(marker_genes)

marker_gene_expression_umap_data <- marker_gene_expression_umap_data %>%
  map(~dplyr::select(.,-ident) %>% 
        rownames_to_column("label_cellbarcode")) %>%
  reduce(left_join, by = join_by(label_cellbarcode, UMAP_1, UMAP_2)) %>%
  pivot_longer(-c(label_cellbarcode, UMAP_1, UMAP_2), names_to = "marker_gene", values_to = "expression") %>% 
  dplyr::mutate(marker_gene=factor(marker_gene, levels=marker_genes)) %>% 
  arrange(marker_gene) %>% 
  split(.$marker_gene) %>% 
  purrr::set_names(marker_genes)

umap_plots <- imap(marker_gene_expression_umap_data, 
                   ~ ggplot(., aes(x = UMAP_1, y = UMAP_2, color =expression)) +
    geom_point(size = 0.5, stroke = NA) +
    scale_colour_distiller(name = "Expression", 
                           palette = "YlGnBu", 
                           direction = 1,
                           na.value = "lightgrey",
                           breaks=round(seq(min(.x$expression), max(.x$expression), length.out=4), digits = 2),
                           limits=round(c(min(.x$expression), max(.x$expression)), digits = 2)
                           ) +
    coord_equal() +
    labs(x = "UMAP1", y = "UMAP2", title=.y) +
    theme_pub() +
    theme(
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      legend.position = "right",
      legend.direction = "vertical",
      aspect.ratio = 1
    ) +
    guides(colour = guide_colourbar(
      barwidth = 0.5,
      barheight = 3,
      override.aes = list(size = 3)
    )))

umap_plots %>%
  imap(~.x[["data"]] %>%
         write_tsv(file = file.path(tsv_dir, paste0("marker_gene_expression_", .y, ".tsv"))))

marker_umaps_no_IG
```

```{r}
saveRDS(gex_scaled_all_noIG, file = file.path(RDS_dir, "gex_scaled_all_noIG.RDS"))
```

#### Marker gene Dotplot
```{r Marker_expression_per_celltype_dotplot, fig.width=5, fig.height=5}
celltype_marker_dotplot_noIG <- DotPlot(gex_scaled_all_noIG,
  features =
    c(
      "MS4A1", # B cells
      "CD14", # Monocytes
      "IL7R", "CD3E", # T cells
      "CD19", "CD5", "FCER2", "CD79B", # CLL
      "GNLY", "NKG7", # NK cells
      "CA1", "HBA1"
    ), # Erythrocytes)),
  group.by = "My_annotation_B",
  cols = "YlGnBu",
) +
  labs(y = "Cell type", x = "Marker gene") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust=0.9))

write_tsv(celltype_marker_dotplot_noIG$data, file.path(tsv_dir, "celltype_marker_dotplot.tsv"))
```


#### Celltype
```{r}
My_annotation_LC <- tibble("label_cellbarcode"=WhichCells(gex_scaled_all_noIG),
                           "label"=gex_scaled_all_noIG$orig.ident,
       "My_annotation"=gex_scaled_all_noIG$My_annotation) %>% 
  left_join(metadata %>% dplyr::select(label, MBL_CLL_subtype)) %>%  
  dplyr::mutate("My_annotation_LC"=case_when(My_annotation == "MBL B Cell" & MBL_CLL_subtype =="LC_MBL" ~ "LC-MBL B Cell",
                                             My_annotation == "MBL B Cell" & MBL_CLL_subtype =="HC_MBL" ~ "HC-MBL B Cell",
                                             .default=My_annotation),
                "My_annotation_LC"=factor(My_annotation_LC, levels=c("LC-MBL B Cell", "HC-MBL B Cell", "CLL B Cell", "T", "NK", "Mono", "DC", "other"))) %>% 
  dplyr::select(label_cellbarcode, My_annotation_LC) %>% 
  deframe()

gex_scaled_all_noIG <- AddMetaData(object = gex_scaled_all_noIG, metadata = My_annotation_LC, col.name = "My_annotation_LC")
```


```{r}
umap_celltype_no_IG <- DimPlot(gex_scaled_all_noIG,cells = nonMBL7dup_cells,
                          reduction = "umap",
                          group.by = "My_annotation_B",
                          label = FALSE,
                          repel = TRUE,
                          raster = FALSE,
                          cols = annotation_colours_named,
                          shuffle = TRUE) +
  scale_colour_manual(values=annotation_colours_named, labels=c("Mono"="Monocytes",
                               "NK"= "NK cells",
                               "DC" = "Dendritic cells",
                               "other" ="Other cell types",
                               "HC-MBL B Cell"="HC-MBL cells",
                               "T"="T cells",
                               "LC-MBL B Cell"="LC-MBL cells",
                               "CLL B Cell"="CLL cells"))+
  labs(x = "UMAP1", y = "UMAP2", colour = "Cell types", title = "Cell types") +
  theme_pub() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "right",
        aspect.ratio = 1
  )

ggsave(umap_celltype_no_IG, 
       file = file.path(plotting_dir, "UMAP_celltypes.png"), 
         height = 4, 
         width = 4)

umap_celltype_no_IG
```

```{r}
umap_celltype_no_IG_MBL_CLL_subtype_split <- DimPlot(gex_scaled_all_noIG,
                          reduction = "umap",
                          split.by = "MBL_CLL_subtype",
                          group.by = "My_annotation_B",
                          label = FALSE,
                          repel = TRUE,
                          raster = FALSE,
                          cols = annotation_colours_named,
                          shuffle = TRUE) +
  scale_colour_manual(values=annotation_colours_named, 
                      labels=c("Mono"="Monocytes",
                               "NK"= "NK cells",
                               "DC" = "Dendritic cells",
                               "other" ="Other cell types",
                               "HC-MBL B Cell"="HC-MBL cells",
                               "T"="T cells",
                               "LC-MBL B Cell"="LC-MBL cells",
                               "CLL B Cell"="CLL cells"))+
  labs(x = "UMAP1", 
       y = "UMAP2", 
       colour = "Cell types", 
       title = "Cell types") +
  theme_pub() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "none",
        aspect.ratio = 1
  )

ggsave(umap_celltype_no_IG_MBL_CLL_subtype_split, 
       file = file.path(plotting_dir, "UMAP_celltypes_MBL_CLL_subtype_split.png"), 
         height = 4, 
         width = 16)

write_tsv(umap_celltype_no_IG_MBL_CLL_subtype_split[[1]][["data"]], file = file.path(tsv_dir, "GEX_celltype_annotation_split_noIG_umap_data.tsv"))

umap_celltype_no_IG_MBL_CLL_subtype_split
```

```{r fig.height=4, fig.width=12}
ggarrange(umap_celltype_no_IG+labs(title="")+
            theme(legend.position = "right"), 
          umap_patient_no_IG+
            scale_colour_manual(values=my_unpaired_colours_named, label=label2plotting_id)+
            theme(legend.position = "right"), nrow=1) %>% 
ggsave(plot=., file=file.path(plotting_dir, "Celltype_patient_figure.pdf"), width = 12, height=4)
```

