---
title: "GEX_Healthy_Clustering"
author: "Anja_Rathgeber"
date: "`r Sys.Date()`"
output: html_document
---

# GOAL: The aim of this script is cluster my malignant samples with healthy samples to check the similarities/differences.

# Load libraries

```{r Load_libraries}
library(readxl)
library(tidyverse)
library(ggpubr)
library(Seurat)
library(SeuratData)
library(SeuratDisk)
library(RColorBrewer)
```

# Set paths
```{r Set_paths}
source("config_paths.R")
library_path <- file.path(project_dir_remote, "scratch/GEX")
library_path_10X <- file.path(project_dir_remote, "scratch/10X_GEX")
data_dir <- file.path(project_dir, "data")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "GEX_analysis")
plotting_dir <- file.path(analysis_dir, "plots")
RDS_dir <- file.path(analysis_dir, "RDS_objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")
```

# Load data
```{r}

source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r Load_data}
reference <- LoadH5Seurat(file.path(data_dir, "reference_genomes/Hao_and_Hao_et_al_2021_PBMC/pbmc_multimodal.h5seurat"))

metadata_malignant <- read_tsv(file = file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata_gex.tsv")) %>%
  tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE)

metadata_healthy <- read_tsv(file = file.path(data_dir, "20250812_Healthy_controls_scRNAseq.tsv")) # https://www.10xgenomics.com/datasets/human-pbmc-from-a-healthy-donor-10-k-cells-v-2-2-standard-5-0-0, https://www.10xgenomics.com/datasets/10-k-human-pbm-cs-5-v-2-0-chromium-controller-2-standard-6-1-0

metadata <- metadata_malignant %>%
  full_join(metadata_healthy)

patient_id2plotting_id <- metadata %>%
  dplyr::select(patient_id, plotting_id) %>%
  deframe()

label2plotting_id <- metadata %>%
  dplyr::select(label, plotting_id) %>%
  deframe()

library2label <- metadata %>%
  dplyr::select(pool, label) %>%
  deframe()

library_names <- metadata %>%
  filter(!is.na(pool)) %>%
  pull(pool)

seurat_paths <- c(
  file.path(library_path, library_names[1:35], "filtered_feature_bc_matrix"), # malignant libraries
  file.path(library_path_10X, library_names[36:37], "filtered_feature_bc_matrix") # , # 10X pbmc
)


gex_scaled_all <- readRDS(file.path(RDS_dir, "gex_scaled_all.RDS"))
```


# Analysis
## Create Seurat Object
```{r Create_seurat_object}
# # remove indices when run all samples
# gex_data_10X_seurat <- seurat_paths %>%
#  map(~ Read10X(data.dir = .)) %>%
#  set_names(library2label[library_names])
#
# gex_10X_seurat <- gex_data_10X_seurat %>%
#  imap( ~ CreateSeuratObject(
#    counts = .x, ,
#    project = .y,
#    min.cells = 3,
#    min.features = 200))
```

## Merge samples
```{r Merge_all_seurat_objects}
# gex_merged_10X_seurat_all <- merge(gex_10X_seurat[[1]],
#                                   y = c(gex_10X_seurat[2:37]),
#                                   project = "Healthy_and_malignant_MBL_CLL",
#                                   add.cell.ids = c(names(gex_data_10X_seurat[1:37])))

# gex_merged_10X_seurat_all <- RenameCells(gex_merged_10X_seurat_all, new.names=sub("^pbmc_", "", colnames(gex_merged_10X_seurat_all)))
```

## QC
```{r MtDNA_per_cell}
# gex_merged_10X_seurat_all[["percent.mt"]] <- PercentageFeatureSet(gex_merged_10X_seurat_all, pattern = "^MT-")
# saveRDS(gex_merged_10X_seurat_all, file = file.path(RDS_dir, "gex_merged_10X_seurat_all.RDS"))
```

```{r fig.width=27, fig.height=10}
# # Visualize QC metrics as a violin plot: features = genes, counts=UMI
# qc_pre_filter_plots <- gex_merged_10X_seurat_all %>%
#  VlnPlot(features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0)
#
# qc_pre_filter_plots %>%
#  ggsave(., filename = file.path(plotting_dir, "QC_GEX/QC_counts_merged_10X_seurat_all.png"),
#         unit = "in",
#         width = 27,
#         height = 10)
```

```{r, fig.width=20,  fig.height=4}
# plot1 <- FeatureScatter(gex_merged_10X_seurat_all, feature1 = "nCount_RNA", feature2 = "percent.mt")
# plot2 <- FeatureScatter(gex_merged_10X_seurat_all, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# f_f_plot <- ggarrange(plot1, plot2, ncol = 2, nrow = 1, common.legend = TRUE, legend="right")
#
# f_f_plot %>%
#   ggsave(., filename = file.path(plotting_dir, "QC_GEX/QC_feature_scatterplot_merged_10X_seurat_all.png"),
#          unit = "in",
#          width = 20,
#          height = 4)
```

```{r}
# gex_merged_10X_seurat_all <- readRDS(file = file.path(RDS_dir, "gex_merged_10X_seurat_all.RDS"))
```

```{r}
# gex_filtered_10X_seurat_all <- subset(gex_merged_10X_seurat_all,
#                                       subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 20)
```

```{r}
# rm(gex_merged_10X_seurat_all)
#
# qc_post_filter_plots <- gex_filtered_10X_seurat_all %>%
#  VlnPlot(features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)
#
# # qc_post_filter_plots
# qc_post_filter_plots %>%
# ggsave(., filename = file.path(plotting_dir, "QC_GEX/QC_counts_filtered_10X_seurat_all.png"),
#        unit = "in",
#       width = 27,
#      height = 10)
```

## Non integrated
```{r}
# gex_scaled_10X_seurat_all <- SCTransform(gex_filtered_10X_seurat_all, verbose = TRUE)
# saveRDS(gex_scaled_10X_seurat_all, file = file.path(RDS_dir, "gex_scaled_10X_seurat_all.RDS"))
```

### LDR
```{r}
# gex_scaled_10X_seurat_all <- RunPCA(gex_scaled_10X_seurat_all, features = VariableFeatures(object = gex_scaled_10X_seurat_all))
#
# dot_plot <- VizDimLoadings(gex_scaled_10X_seurat_all, dims = 1:2, reduction = "pca") +
#   theme_pub()
#
# dot_plot %>%
#  ggsave(., filename = file.path(plotting_dir, "QC_PCA_DimLoadings_all.png"),
#         unit = "in",
#         width = 10,
#         height = 6)
#
# pca <- DimPlot(gex_scaled_10X_seurat_all, reduction = "pca", raster = FALSE) +
#   theme_pub()
#
# pca %>%
#   ggsave(., filename = file.path(plotting_dir, "PCA_all.png"),
#          unit = "in",
#          width = 10,
#          height = 4)
#
# #pca
#
# PCA_heatmap <- DimHeatmap(gex_scaled_10X_seurat_all, dims = 1, cells = 500, balanced = TRUE) +
#  theme_pub()
#
# PCA_heatmap %>%
#  ggsave(., filename = file.path(plotting_dir, "PCA_PC1_heatmap_all.png"),
#         unit = "in",
#         width = 6,
#         height = 4)
#
# ElbowPlot(gex_scaled_10X_seurat_all) %>%
#   ggsave(., filename = file.path(plotting_dir, "Elbow_plot.png"),
#          unit = "in",
#          width = 6,
#          height = 4)
```

### Label transfer
Find transfer anchors
```{r}
# anchors <- FindTransferAnchors(
#   reference = reference,
#   query = gex_scaled_10X_seurat_all,
#   normalization.method = "SCT",
#   reference.reduction = "spca",
#   dims = 1:50
# )
#
# gex_scaled_10X_seurat_all <- MapQuery(
#   anchorset = anchors,
#   query = gex_scaled_10X_seurat_all,
#   reference = reference,
#   refdata = list(
#     celltype.l1 = "celltype.l1",
#     celltype.l2 = "celltype.l2",
#     predicted_ADT = "ADT"
#   ),
#   reference.reduction = "spca",
#   reduction.model = "wnn.umap"
# )
#
# saveRDS(gex_scaled_10X_seurat_all, file = file.path(RDS_dir, "gex_scaled_10X_seurat_all.RDS"))
```

### Clustering and UMAP
```{r}
# gex_scaled_10X_seurat_all <- FindNeighbors(gex_scaled_10X_seurat_all, dims = 1:10) %>%
#   FindClusters(resolution = 0.5)
```

```{r}
# gex_scaled_10X_seurat_all <- RunUMAP(gex_scaled_10X_seurat_all, dims = 1:10)
```

```{r}
# saveRDS(gex_scaled_10X_seurat_all, file = file.path(RDS_dir, "gex_scaled_10X_seurat_all.RDS"))
# gex_scaled_10X_seurat_all <- readRDS(file = file.path(RDS_dir, "gex_scaled_10X_seurat_all.RDS"))
```

### Celltype annotation
```{r}
# malignant_celltype_annotation <- tibble(
#   "label_cellbarcode" = colnames(gex_scaled_all),
#   "My_annotation_B" = gex_scaled_all$My_annotation_B
# ) %>%
#   dplyr::mutate(My_annotation_LC = case_when(My_annotation_B == "MBL B Cell" & str_starts(label_cellbarcode,
#       pattern = paste0("MBL", 10:15, collapse = "|")) ~ "LC-MBL B Cell",
#     My_annotation_B == "MBL B Cell" & str_starts(label_cellbarcode,
#       pattern = paste0("MBL", 10:15, collapse = "|"), negate = TRUE) ~ "HC-MBL B Cell",
#     .default = My_annotation_B)) %>%
#   dplyr::select(-My_annotation_B)
# 
# write_tsv(malignant_celltype_annotation, file = file.path(tsv_dir, "20250523_RNA_celltype_annotation_malignant.tsv"))
# 
# healthy_celltype_annotation <- tibble(
#   "label_cellbarcode" = colnames(gex_scaled_10X_seurat_all),
#   "My_annotation_LC" = gex_scaled_10X_seurat_all$predicted.celltype.l1
# ) %>%
#   dplyr::filter(str_starts(label_cellbarcode, "pbmc1|pbmc2|pbmc3|pbmc4")) %>%
#   dplyr::mutate(My_annotation_LC = case_when(My_annotation_LC == "CD4 T" ~ "T",
#     My_annotation_LC == "CD8 T" ~ "T",
#     My_annotation_LC == "other T" ~ "T",
#     .default = My_annotation_LC
#   ))
# 
# celltype_annotation <- bind_rows(malignant_celltype_annotation, healthy_celltype_annotation) %>% deframe()
# 
# gex_scaled_10X_seurat_all <- AddMetaData(
#   object = gex_scaled_10X_seurat_all,
#   metadata = celltype_annotation,
#   col.name = "My_annotation_LC"
# )
```

### Plot UMAPs
```{r}
# nonMBL7dup_cells <- colnames(gex_scaled_10X_seurat_all)[-c(str_starts(colnames(gex_scaled_10X_seurat_all), "MBL7dup"))]
```

#### Clusters
```{r}
# color_palette <- colorRampPalette(brewer.pal(12, "Set3"))(54)
```

```{r }
# umap_clusters_plot <- DimPlot(gex_scaled_10X_seurat_all,
#                               cells=nonMBL7dup_cells,
#   reduction = "umap",
#   raster = FALSE,
#   label = TRUE,
#   group.by = "seurat_clusters",
#   cols = color_palette
# )
# 
# umap_clusters_plot %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_clusters.png"),
#     unit = "in",
#     width = 8,
#     height = 6
#   )
# 
# umap_clusters_plot
```

#### Samples
```{r}
# add plotting id
# plotting_annot <- as_tibble(gex_scaled_10X_seurat_all$orig.ident, rownames=NA) %>%
#   rownames_to_column("label_cellbarcode") %>%
#   dplyr::rename("label"="value") %>%
#   left_join(metadata %>%
#               dplyr::select(label, plotting_id)) %>%
#   dplyr::select(label_cellbarcode, plotting_id) %>%
#   deframe()
#
# gex_scaled_10X_seurat_all <- AddMetaData(object = gex_scaled_10X_seurat_all,
#                                          metadata = plotting_annot,
#                                          col.name = "plotting_id")
```

```{r}
# umap_sample_plot <- DimPlot(gex_scaled_10X_seurat_all,
#                               cells=nonMBL7dup_cells,
#                             reduction = "umap",
#                             raster = FALSE,
#                             group.by = c("orig.ident"),
#                             cols=c(my_unpaired_colours_named,my_paired_colours_named, control_colours_named)) +
#   theme_pub() +
#   theme(
#     aspect.ratio = 1,
#     legend.position = "right"
#   )+
#   labs(title="Samples")+
#   theme_pub()+
#   theme(aspect.ratio = 1,
#         axis.ticks.x = element_blank(),
#         axis.ticks.y = element_blank(),
#         axis.text.x = element_blank(),
#         axis.text.y=element_blank())
# 
# umap_sample_plot %>%
#   ggsave(.,
#          filename = file.path(plotting_dir, "UMAP_10X_seurat_all_samples.png"),
#          unit = "in",
#          width = 8,
#          height = 6)
# 
# umap_sample_plot
```

#### Patients 
```{r}
# # add patient id
# patient_id_annot <- as_tibble(gex_scaled_10X_seurat_all$orig.ident, rownames = NA) %>%
#   rownames_to_column("label_cellbarcode") %>%
#   dplyr::rename("label" = "value") %>%
#   left_join(metadata %>%
#     dplyr::select(label, patient_id)) %>%
#   dplyr::select(label_cellbarcode, patient_id) %>%
#   deframe()
# 
# gex_scaled_10X_seurat_all <- AddMetaData(object = gex_scaled_10X_seurat_all, metadata = patient_id_annot, col.name = "patient_id")
```

```{r}
# umap_patient <- DimPlot(gex_scaled_10X_seurat_all,
#                               cells=nonMBL7dup_cells,
#   reduction = "umap",
#   group.by = "patient_id",
#   raster = F,
#   cols = c(my_unpaired_colours_named, control_colours_named),
#   shuffle = TRUE
# ) +
#   scale_colour_manual(values=c(my_unpaired_colours_named, control_colours_named),
#                         labels=patient_id2plotting_id)+
#   labs(x = "UMAP1", y = "UMAP2", title = "Patients") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "right",
#     aspect.ratio = 1
#   )
# 
# umap_patient %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_patients.png"),
#     unit = "in",
#     width = 7,
#     height = 4
#   )
# 
# umap_patient
```

#### Patients disease-split
```{r}
# sample_annot <- tibble("label_cellbarcode" = WhichCells(gex_scaled_10X_seurat_all))
# 
# disease_annot <- sample_annot %>%
#   tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
#   tidyr::separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE) %>%
#   dplyr::left_join(metadata %>% dplyr::select(label, disease) %>% unique()) %>%
#   dplyr::mutate(
#     disease = if_else(condition = is.na(disease) & patient_id %in% paste0("pbmc", 1:4), true = "Healthy", false = disease),
#     disease = factor(disease, levels = c("Healthy", "MBL", "CLL"))
#   ) %>%
#   dplyr::select(label_cellbarcode, disease) %>%
#   deframe()
# 
# gex_scaled_10X_seurat_all <- AddMetaData(
#   object = gex_scaled_10X_seurat_all,
#   metadata = disease_annot,
#   col.name = "disease"
# )
```

```{r}
# umap_patient_disease <- DimPlot(subset(gex_scaled_10X_seurat_all, cells=nonMBL7dup_cells),
#   reduction = "umap",
#   group.by = "patient_id",
#   raster = F,
#   cols = c(my_unpaired_colours_named, control_colours_named),
#   split.by = "disease",
#   shuffle = TRUE
# ) +
#   labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Disease") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "bottom",
#     aspect.ratio = 1
#   )
# 
# umap_patient_disease %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_patients_disease_split.png"),
#     unit = "in",
#     width = 24,
#     height = 6
#   )
# 
# umap_patient_disease
```

#### Patients disease-subtype-split
```{r}
# # add MBL_CLL_subtype annotation
# MBL_CLL_subtype_annot <- as_tibble(gex_scaled_10X_seurat_all$orig.ident, rownames = NA) %>%
#   rownames_to_column("label_cellbarcode") %>%
#   dplyr::rename("label" = "value") %>%
#   dplyr::left_join(metadata %>% dplyr::select(label, MBL_subtype)) %>%
#   dplyr::mutate(
#     MBL_CLL_subtype = if_else(is.na(MBL_subtype) & !(label %in% paste0("pbmc", 1:4)), "CLL", MBL_subtype),
#     MBL_CLL_subtype = if_else(condition = is.na(MBL_CLL_subtype) & label %in% paste0("pbmc", 1:4), true = "Healthy", false = MBL_CLL_subtype),
#     MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("Healthy", "LC", "HC", "CLL"))
#   ) %>%
#   dplyr::select(label_cellbarcode, MBL_CLL_subtype) %>%
#   deframe()
# 
# gex_scaled_10X_seurat_all <- AddMetaData(
#   object = gex_scaled_10X_seurat_all,
#   metadata = MBL_CLL_subtype_annot,
#   col.name = "MBL_CLL_subtype"
# )
```

```{r}
# umap_patient_MBL_CLL_subtype <- DimPlot(subset(gex_scaled_10X_seurat_all, cells=nonMBL7dup_cells),
#   reduction = "umap",
#   group.by = "patient_id",
#   raster = F,
#   cols = c(my_unpaired_colours_named, control_colours_named),
#   split.by = "MBL_CLL_subtype",
#   shuffle = TRUE
# ) +
#   labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Patients") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "bottom",
#     aspect.ratio = 1
#   )
# 
# write_tsv(umap_patient_MBL_CLL_subtype[[1]]$data, file.path(tsv_dir, "umap_patient_MBL_CLL_subtype_data.tsv" ))
# 
# umap_patient_MBL_CLL_subtype %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_patients_disease_subtype_split.png"),
#     unit = "in",
#     width = 32,
#     height = 6
#   )
# 
# umap_patient_MBL_CLL_subtype
```

#### Predicted celltypes and reference
```{r}
# umap_predcelltype <- DimPlot(gex_scaled_10X_seurat_all,
#                              cells=nonMBL7dup_cells,
#                              reduction = "umap",
#                              raster = FALSE,
#                              group.by = "predicted.celltype.l1",
#                              cols=annotation_colours_named) +
#   theme_pub() +
#   theme(
#     aspect.ratio = 1,
#     legend.position = "none"
#   )
# 
# umap_predcelltype %>%
#   ggsave(.,
#          filename = file.path(plotting_dir, "UMAP_10X_seurat_all_predicted_celltypes.png"),
#          unit = "in",
#          width = 7,
#          height = 4)
# 
# umap_predcelltype
```

```{r}
# umap_reference <- DimPlot(gex_scaled_10X_seurat_all,
#                           cells=nonMBL7dup_cells,
#                           reduction = "ref.umap",
#                           group.by = "predicted.celltype.l1",
#                           label = TRUE,
#                           label.size = 3,
#                           repel = TRUE,
#                           raster = FALSE,
#                           cols = annotation_colours_named) +
#   theme_pub() +
#   theme(
#     aspect.ratio = 1,
#     legend.position = "none"
#   )
# 
# ggarrange(umap_predcelltype, umap_reference, nrow = 1, ncol = 2) %>%
#   ggsave(., file = file.path(plotting_dir, "UMAP_10X_seurat_all_predicted_celltypes_and_reference.png"),
#          height = 4,
#          width = 8)
# 
# ggarrange(umap_predcelltype, umap_reference, nrow = 1, ncol = 2)
```

```{r}
# markers_per_predicted_celltype_vlnplot <- VlnPlot(subset(gex_scaled_10X_seurat_all, cells=nonMBL7dup_cells),
#   group.by = "predicted.celltype.l1",
#   features = marker_genes,
#   slot = "scale.data",
#   cols = annotation_colours_named,
#   pt.size = FALSE
# ) +
#   theme_pub() +
#   theme(legend.position = "none")
# 
# markers_per_predicted_celltype_vlnplot %>%
#   ggsave(.,
#     file = file.path(plotting_dir, "VlnPlots_10X_seurat_all_Markers_per_predicted_celltype_vlnplot.png"),
#     height = 16,
#     width = 12
#   )
# 
# markers_per_predicted_celltype_vlnplot
```


```{r}
# predicted_celltype_marker_dotplot <- DotPlot(subset(gex_scaled_10X_seurat_all, cells=nonMBL7dup_cells),
#   features =
#     c(
#       "MS4A1", # B cells
#       "CD19", "CD5", "FCER2", "CD79B", # CLL
#       "CD14", # Monocytes
#       "IL7R", "CD3E", # T cells
#       "GNLY", "NKG7", # NK cells
#       "CA1", "HBA1"
#     ), # Erythrocytes)),
#   group.by = "predicted.celltype.l1",
#   cols = "RdYlBu"
# ) +
#   labs(y = "Celltype", x = "Marker gene") +
#   theme_pub() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.9))
# 
# ggsave(predicted_celltype_marker_dotplot,
#   file = file.path(plotting_dir, "Dotplot_10X_seurat_all_Markers_per_predicted_celltypes.png"),
#   width = 7,
#   height = 4,
#   unit = "in"
# )
# 
# markers_per_predicted_celltype_vlnplot
```

#### Prediction scores
```{r fig.width=8, fig.height=6}
# Idents(gex_scaled_10X_seurat_all) <- "predicted.celltype.l1"
# prediction_score_umap <- FeaturePlot(gex_scaled_10X_seurat_all,
#                                      cells = nonMBL7dup_cells,
#                                      features = c("Mono", "CD8 T", "CD4 T", "other T", "other", "B", "DC", "NK"),
#                                      reduction = "umap",
#                                      cols = c("lightgrey", "darkred"),
#                                      ncol = 3,
#                                      raster = FALSE) &
#   theme_pub() &
#   theme(aspect.ratio = 1)
# 
# prediction_score_umap %>%
#   ggsave(.,
#          file = file.path(plotting_dir, "UMAP_10X_seurat_all_predicted_celltypes_scores.png"),
#          height = 6,
#          width = 8)
# 
# prediction_score_umap
```

#### Markers
```{r}
# marker_vln <- VlnPlot(subset(gex_scaled_10X_seurat_all,cells=nonMBL7dup_cells),
#   group.by = "seurat_clusters",
#   slot = "scale.data",
#   cols = color_palette,
#   pt.size = FALSE,
#   features = marker_genes
# ) +
#   theme_pub() +
#   theme(legend.position = "none")
# 
# marker_vln
# marker_vln %>%
#   ggsave(.,
#     file = file.path(plotting_dir, "Marker_expression_vlnplot_all.png"),
#     units = "in",
#     width = 27,
#     height = 16
#   )
```

```{r}
# marker_umaps <- FeaturePlot(gex_scaled_10X_seurat_all,
#                             cells = nonMBL7dup_cells,
#   features = marker_genes,
#   reduction = "umap",
#   raster = FALSE
# ) +
#   theme_pub() +
#   theme(legend.position = "none")
# 
# marker_umaps %>%
#   ggsave(.,
#     file = file.path(plotting_dir, "Marker_expression_umaps_all.png"),
#     units = "in",
#     width = 12,
#     height = 16
#   )
# 
# marker_umaps
```

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# gex.markers <- FindAllMarkers(gex_scaled_10X_seurat_all,
#                               only.pos = TRUE,
#                               min.pct = 0.25,
#                               logfc.threshold = 0.25)
#
# saveRDS(gex.markers, file = file.path(RDS_dir, "gex.markers_all.RDS"))
```

```{r fig.width=20, fig.height=16}
# gex.markers %>%
#   group_by(cluster) %>%
#   top_n(n = 10, wt = avg_log2FC) -> top10
#
# marker_heatmap <- DoHeatmap(gex_scaled_10X_seurat_all, features = top10$gene) +
#   theme_pub() +
#   theme(legend.position = "none")
#
# marker_heatmap %>%
#   ggsave(., file = file.path(plotting_dir, "Marker_expression_heatmap_all.png"),
#          units = "in",
#          width = 20,
#          height = 16)
#
# marker_heatmap
```

```{r}
# saveRDS(gex_scaled_10X_seurat_all, file = file.path(RDS_dir, "gex_scaled_10X_seurat_all.RDS"))
```

#### Celltype
```{r}
# umap_celltype <- DimPlot(gex_scaled_10X_seurat_all, 
#                           cells= nonMBL7dup_cells,
#                           reduction = "umap",
#                           group.by = "My_annotation_LC",
#                           label = FALSE,
#                           repel = TRUE,
#                           raster = FALSE,
#                           cols = annotation_colours_named,
#                           shuffle = TRUE) +
#   labs(x = "UMAP1", y = "UMAP2", colour = "Cell types", title = "Cell types") +
#   theme_pub() +
#   theme(axis.ticks = element_blank(),
#         axis.text.y = element_blank(),
#         axis.text.x = element_blank(),
#         legend.position = "none",
#         aspect.ratio = 1
#   )
# 
# ggsave(umap_celltype,
#        file = file.path(plotting_dir, "UMAP_10X_seurat_all_celltypes.png"),
#          height = 4,
#          width = 4)
# 
# umap_celltype
```

```{r}
# umap_celltype_MBL_CLL_subtype_split <- DimPlot(subset(gex_scaled_10X_seurat_all, cells=nonMBL7dup_cells),
#   reduction = "umap",
#   group.by = "My_annotation_LC",
#   split.by = "MBL_CLL_subtype",
#   label = FALSE,
#   repel = TRUE,
#   raster = FALSE,
#   cols = annotation_colours_named,
#   shuffle = TRUE
# ) +
#   labs(x = "UMAP1", y = "UMAP2", colour = "Cell types", title = "Cell type") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "none",
#     aspect.ratio = 1
#   )
# 
# write_tsv(umap_celltype_MBL_CLL_subtype_split[[1]]$data, file.path(tsv_dir, "umap_celltype_MBL_CLL_subtype_split_data.tsv" ))
# 
# ggsave(umap_celltype_MBL_CLL_subtype_split+theme(legend.position = "bottom"),
#   file = file.path(plotting_dir, "UMAP_10X_seurat_all_celltypes_MBL_CLL_subtype_split.png"),
#   height = 4,
#   width = 16
# )
# 
# umap_celltype_MBL_CLL_subtype_split
```

```{r}
# markers_per_celltype_vlnplot <- VlnPlot(subset(gex_scaled_10X_seurat_all, cells=nonMBL7dup_cells),
#   group.by = "My_annotation_LC",
#   slot = "scale.data",
#   pt.size = FALSE,
#   cols = annotation_colours_named,
#   features = c(
#     "PAX5", "MS4A1", # B-cell trajectories
#     "CD19", "CD5", "FCER2", "CD79B", # CLL B-Cell Marker
#     "CD3D", "CD8A", "TBX21", "IL7R", "CD3E", "CD3G", "CD4", "CD8B", # TCells
#     "CD14", "CEBPB", "CD33", # Monocytes
#     "GNLY", "NKG7", # NK cells
#     "CA1", "HBA1" # erythrocytes
#   ), sort = TRUE
# ) +
#   theme_pub() +
#   theme(legend.position = "none")
# 
# markers_per_celltype_vlnplot %>%
#   ggsave(.,
#     file = file.path(plotting_dir, "VlnPlots_10X_seurat_all_Markers_per_celltype_vlnplot.png"),
#     height = 16,
#     width = 12
#   )
```

```{r}
# celltype_marker_dotplot <- DotPlot(subset(gex_scaled_10X_seurat_all,cells=nonMBL7dup_cells),
#   features =
#     c(
#       "MS4A1", # B cells
#       "CD19", "CD5", "FCER2", "CD79B", # CLL
#       "CD14", # Monocytes
#       "IL7R", "CD3E", # T cells
#       "GNLY", "NKG7", # NK cells
#       "CA1", "HBA1"
#     ), # Erythrocytes)),
#   group.by = "My_annotation_LC",
#   cols = "RdYlBu"
# ) +
#   labs(y = "Celltype", x = "Marker gene") +
#   theme_pub() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.9))
# 
# ggsave(celltype_marker_dotplot,
#   file = file.path(plotting_dir, "Dotplot_10X_seurat_all_Markers_per_celltypes.png"),
#   width = 7,
#   height = 4,
#   unit = "in"
# )
```

## Non integrated IG genes removed
```{r}
# gex_scaled_10X_seurat_all <- readRDS(file = file.path(RDS_dir, "gex_scaled_10X_seurat_all.RDS"))
```

### Subset
```{r}
# # identify all IG genes
# features <- rownames(gex_scaled_10X_seurat_all)
# ig_pattern <- "^IG(A|D|E|G|H|K|L|M|KV|LV)"
# ig_genes <- features[str_detect(features, ig_pattern)]
#
# gex_10X_seurat_all_noIG <- subset(x=gex_scaled_10X_seurat_all,
#                                          features=setdiff(features, ig_genes))
```

```{r}
# gex_scaled_10X_seurat_all_noIG <- SCTransform(gex_10X_seurat_all_noIG, verbose = TRUE)
#
# saveRDS(gex_scaled_10X_seurat_all_noIG, file = file.path(RDS_dir, "gex_scaled_10X_seurat_all_noIG.RDS"))
```

### LDR
```{r}
# gex_scaled_10X_seurat_all_noIG <- RunPCA(gex_scaled_10X_seurat_all_noIG,
#                                     features = VariableFeatures(object = gex_scaled_10X_seurat_all_noIG))
#
# ElbowPlot(gex_scaled_10X_seurat_all_noIG) %>%
#   ggsave(., filename = file.path(plotting_dir, "Elbow_plot_no_IG.png"),
#          unit = "in",
#          width = 6,
#          height = 4)
#
# ElbowPlot(gex_scaled_10X_seurat_all_noIG)
```

### Clustering and UMAP
```{r}
# gex_scaled_10X_seurat_all_noIG <- FindNeighbors(gex_scaled_10X_seurat_all_noIG, dims = 1:10) %>%
#   FindClusters(resolution = 0.5)
```

```{r}
# set.seed(123)
# gex_scaled_10X_seurat_all_noIG <- RunUMAP(gex_scaled_10X_seurat_all_noIG, dims = 1:10)
```

```{r}
# saveRDS(gex_scaled_10X_seurat_all_noIG, file = file.path(RDS_dir, "gex_scaled_10X_seurat_all_noIG.RDS"))
# gex_scaled_10X_seurat_all_noIG <- readRDS(file = file.path(RDS_dir, "gex_scaled_10X_seurat_all_noIG.RDS"))
```

```{r}
# gex_scaled_10X_seurat_all_noIG <- AddMetaData(object = gex_scaled_10X_seurat_all_noIG,
#                                          metadata =celltype_annotation,
#                                          col.name = "My_annotation_LC")
```

### Plot UMAPs
```{r}
color_palette <- colorRampPalette(brewer.pal(12, "Set3"))(60)
```

#### Clusters
```{r}
nonMBL7dup_cells <- colnames(gex_scaled_10X_seurat_all_noIG)[
  !grepl("^MBL7dup", colnames(gex_scaled_10X_seurat_all_noIG))
]
```

```{r}
# umap_clusters_plot_no_IG <- DimPlot(gex_scaled_10X_seurat_all_noIG,
#   cells = nonMBL7dup_cells,
#   label = TRUE,
#   shuffle = TRUE,
#   reduction = "umap",
#   raster = FALSE,
#   group.by = c("seurat_clusters"),
#   cols = color_palette
# )
# 
# umap_clusters_plot_no_IG %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_clusters.png"),
#     unit = "in",
#     width = 8,
#     height = 6
#   )
# 
# umap_clusters_plot_no_IG
```

#### Samples
```{r}
# # add plotting id
# plotting_annot_no_IG <- as_tibble(gex_scaled_10X_seurat_all_noIG$orig.ident, rownames = NA) %>%
#   rownames_to_column("label_cellbarcode") %>%
#   dplyr::rename("label" = "value") %>%
#   left_join(metadata %>%
#     dplyr::select(label, plotting_id)) %>%
#   dplyr::select(label_cellbarcode, plotting_id) %>%
#   deframe()
# 
# gex_scaled_10X_seurat_all_noIG <- AddMetaData(
#   object = gex_scaled_10X_seurat_all_noIG,
#   metadata = plotting_annot_no_IG,
#   col.name = "plotting_id"
# )
```

```{r}
# umap_sample_plot_no_IG <- DimPlot(gex_scaled_10X_seurat_all_noIG,
#                                   cells = nonMBL7dup_cells,
#                                   shuffle = TRUE,
#   reduction = "umap",
#   raster = FALSE,
#   group.by = c("orig.ident"),
#   cols = c(my_unpaired_colours_named, my_paired_colours_named, control_colours_named)
# ) +
#   theme_pub() +
#   labs(title = "Samples") +
#   theme(
#     aspect.ratio = 1,
#     legend.position = "right",
#     axis.ticks.x = element_blank(),
#     axis.ticks.y = element_blank(),
#     axis.text.x = element_blank(),
#     axis.text.y = element_blank()
#   )
# 
# umap_sample_plot_no_IG %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_samples.png"),
#     unit = "in",
#     width = 8,
#     height = 6
#   )
# 
# umap_sample_plot_no_IG
```

#### Patients 
```{r}
# # add patient id
# patient_id_annot_no_IG <- as_tibble(gex_scaled_10X_seurat_all_noIG$orig.ident, rownames = NA) %>%
#   rownames_to_column("label_cellbarcode") %>%
#   dplyr::rename("label" = "value") %>%
#   left_join(metadata %>%
#     dplyr::select(label, patient_id)) %>%
#   dplyr::select(label_cellbarcode, patient_id) %>%
#   deframe()
# 
# gex_scaled_10X_seurat_all_noIG <- AddMetaData(
#   object = gex_scaled_10X_seurat_all_noIG,
#   metadata = patient_id_annot_no_IG,
#   col.name = "patient_id"
# )
```

```{r}
# umap_patient_no_IG <- DimPlot(gex_scaled_10X_seurat_all_noIG,
#                               cells = nonMBL7dup_cells,
#   reduction = "umap",
#   group.by = "patient_id",
#   raster = F,
#   cols = c(my_unpaired_colours_named, control_colours_named),
#   shuffle = TRUE
# ) +
#   scale_colour_manual(values=c(my_unpaired_colours_named, control_colours_named),
#                         labels=patient_id2plotting_id)+
#   labs(x = "UMAP1", y = "UMAP2", title = "Patients") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "right",
#     aspect.ratio = 1
#   )
# 
# write_tsv(umap_patient_no_IG[[1]]$data, file.path(tsv_dir, "umap_patient_no_IG_data.tsv"))
# 
# umap_patient_no_IG %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_patients.png"),
#     unit = "in",
#     width = 7,
#     height = 4
#   )
# 
# umap_patient_no_IG
```

```{r}
# umap_data <- gex_scaled_10X_seurat_all_noIG@reductions$umap@cell.embeddings %>% as.data.frame() %>%
#   rownames_to_column("label_cellbarcode") %>%
#   left_join(gex_scaled_10X_seurat_all_noIG@meta.data[c("orig.ident", "patient_id","MBL_CLL_subtype", "My_annotation_LC", "plotting_id")] %>%
#               rownames_to_column("label_cellbarcode")) %>%
#   left_join(metadata %>% dplyr::select(label, sequencing_date), by=join_by(orig.ident ==label)) %>%
#   dplyr::rename("label"="orig.ident") %>%
#   dplyr::filter(label != "MBL7dup") %>%
#   dplyr::mutate(plotting_id=factor(plotting_id, levels=c(paste0("P", 1:24),paste0("H", 1:4)))) %>%
#   dplyr::arrange(plotting_id) %>%
#   dplyr::mutate(patient_id=factor(patient_id, levels=unique(patient_id)))
```

```{r fig.height=5, fig.width=5}
# umap_density_sample_plot_no_IG <- umap_data %>% 
#   ggplot(aes(x=UMAP_1, y=UMAP_2, colour=patient_id))+
#   geom_density2d()+
#   scale_colour_manual(name="Patient", 
#                       values=c(my_unpaired_colours_named, my_paired_colours_named, control_colours_named), 
#                       labels=patient_id2plotting_id)+
#   theme_pub()+
#   theme(aspect.ratio = 1,
#     legend.position = "bottom",
#     axis.ticks.x = element_blank(),
#     axis.ticks.y = element_blank(),
#     axis.text.x = element_blank(),
#     axis.text.y = element_blank())
```

#### Patients disease-split
```{r}
# sample_annot_no_IG <- tibble("label_cellbarcode" = WhichCells(gex_scaled_10X_seurat_all_noIG))
# 
# disease_annot_no_IG <- sample_annot_no_IG %>%
#   tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
#   tidyr::separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE) %>%
#   dplyr::left_join(metadata %>% dplyr::select(label, disease) %>% unique()) %>%
#   dplyr::mutate(
#     disease = if_else(condition = is.na(disease) & patient_id %in% paste0("pbmc", 1:4), true = "Healthy", false = disease),
#     disease = factor(disease, levels = c("Healthy", "MBL", "CLL"))
#   ) %>%
#   dplyr::select(label_cellbarcode, disease) %>%
#   deframe()
# 
# gex_scaled_10X_seurat_all_noIG <- AddMetaData(
#   object = gex_scaled_10X_seurat_all_noIG,
#   metadata = disease_annot_no_IG,
#   col.name = "disease"
# )
```

```{r}
# umap_patient_disease_no_IG <- DimPlot(subset(x = gex_scaled_10X_seurat_all_noIG, 
#                                              cells = nonMBL7dup_cells),
#   reduction = "umap",
#   group.by = "patient_id",
#   raster = F,
#   cols = c(my_unpaired_colours_named, control_colours_named),
#   split.by = "disease",
#   shuffle = TRUE
# ) +
#   labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Disease") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "bottom",
#     aspect.ratio = 1
#   )
# 
# umap_patient_disease_no_IG %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_patients_disease_split.png"),
#     unit = "in",
#     width = 12,
#     height = 6
#   )
# 
# umap_patient_disease_no_IG
```

```{r fig.height=4, fig.width=20}
# umap_density_sample_plot_no_IG_split <- umap_data %>% 
#   ggplot(aes(x=UMAP_1, y=UMAP_2, colour=patient_id))+
#   geom_density2d()+
#   scale_colour_manual(name="Sample", values=c(my_unpaired_colours_named, my_paired_colours_named, control_colours_named), labels=patient_id2plotting_id)+
#   theme(aspect.ratio = 1,
#     legend.position = "right",
#     axis.ticks.x = element_blank(),
#     axis.ticks.y = element_blank(),
#     axis.text.x = element_blank(),
#     axis.text.y = element_blank())+
#   theme_pub()+
#   facet_wrap(~MBL_CLL_subtype, nrow=1)
```

```{r, fig.width=12, fig.height=4}
# #hide density for other immune cells
# umap_data_no_immune <- umap_data %>% 
#   dplyr::filter(!(My_annotation_LC %in% c("T", "other", "Mono", "NK", "DC")))
# 
# umap_density_sample_plot_no_IG_no_immune_split <- ggplot()+
#   geom_point(data = umap_data, aes(x=UMAP_1, y=UMAP_2, colour=patient_id), alpha=0.05, size=0.000000001)+
#   geom_density2d(data = umap_data_no_immune, aes(x=UMAP_1, y=UMAP_2, colour=patient_id), linewidth = 0.5)+
#   scale_colour_manual(name="Patient", values=c(my_unpaired_colours_named, my_paired_colours_named, control_colours_named), 
#                       labels=patient_id2plotting_id)+
#   labs(x="UMAP1", y="UMAP2")+
#   theme_pub()+
#   theme(aspect.ratio = 1,
#     legend.position = "right",
#     axis.ticks = element_blank(),
#     axis.text = element_blank())+
#   facet_wrap(~ MBL_CLL_subtype, nrow = 1)+
#   guides(colour=list(size=1))
# 
# ggsave(plot=umap_density_sample_plot_no_IG_no_immune_split,
#        file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_patients_disease_split_density.png"),
#        width=16,
#        height=4,
#        unit="in",
#        dpi=600)
```

```{r}
centroids <- umap_data_no_immune %>% 
  dplyr::select(UMAP_1, UMAP_2, patient_id) %>% 
  group_by(patient_id) %>% 
  dplyr::summarize(across(everything(), mean)) %>% 
  dplyr::left_join(metadata %>% dplyr::filter(data_type=="GEX") %>% dplyr::select(patient_id, plotting_id)) %>% unique()

umap_patient_no_IG_labelled <- DimPlot(gex_scaled_10X_seurat_all_noIG,
                              cells=nonMBL7dup_cells,
                        reduction = "umap",
                        group.by = "patient_id",
                        raster = F,
                        cols = my_unpaired_colours_named,
                        shuffle = TRUE) +
  geom_text_repel(data=centroids, aes(x=UMAP_1, y=UMAP_2, label=plotting_id), max.overlap=0, force=500)+
  scale_colour_manual(values=my_unpaired_colours_named, label=patient_id2plotting_id)+
  labs(x = "UMAP1", y = "UMAP2", title = "") +
  theme_pub() +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "bottom",
        aspect.ratio=1)

write_tsv(umap_patient_no_IG_labelled[[1]][["data"]], file.path(tsv_dir, "20250325_GEX_healthy_patient_annotation_noIG_umap_data.tsv"))
write_tsv(centroids, file.path(tsv_dir, "20250325_GEX_healthy_patient_annotation_noIG_umap_centroids.tsv"))
```

```{r}
# umap_data_no_immune %>% 
#   dplyr::mutate(sequencing_date=as.character(sequencing_date)) %>% 
#   ggplot(aes(x=UMAP_1, y=UMAP_2, colour=sequencing_date))+
#   geom_point(size=0.000000001)+
#   scale_colour_manual(name="Sequencing_date",
#                       values=c("2022-02-28" = "#E6AB02", 
#                         "2022-11-18" = "#7FC97F", 
#                         "2023-05-02" = "#FFFF99", 
#                         "2024-03-18" = "#386CB0",
#                         "2022-02-26" = "#FFEF40"))+
#   theme_pub()+
#   theme(aspect.ratio = 1,
#     legend.position = "bottom",
#     axis.ticks.x = element_blank(),
#     axis.ticks.y = element_blank(),
#     axis.text.x = element_blank(),
#     axis.text.y = element_blank())+
#   facet_wrap(~ MBL_CLL_subtype, nrow = 1)+
#   guides(colour=list(size=1))
```

#### Patients disease-subtype-split
```{r}
# # add MBL_CLL_subtype annotation
# MBL_CLL_subtype_annot_no_IG <- as_tibble(gex_scaled_10X_seurat_all_noIG$orig.ident, rownames = NA) %>%
#   rownames_to_column("label_cellbarcode") %>%
#   dplyr::rename("label" = "value") %>%
#   dplyr::left_join(metadata %>% dplyr::select(label, MBL_subtype)) %>%
#   dplyr::mutate(
#     MBL_CLL_subtype = if_else(is.na(MBL_subtype) & !(label %in% paste0("pbmc", 1:4)), "CLL", MBL_subtype),
#     MBL_CLL_subtype = if_else(condition = is.na(MBL_CLL_subtype) & label %in% paste0("pbmc", 1:4), true = "Healthy", false = MBL_CLL_subtype),
#     MBL_CLL_subtype = factor(MBL_CLL_subtype, levels = c("Healthy", "LC", "HC", "CLL"))
#   ) %>%
#   dplyr::select(label_cellbarcode, MBL_CLL_subtype) %>%
#   deframe()
# 
# gex_scaled_10X_seurat_all_noIG <- AddMetaData(
#   object = gex_scaled_10X_seurat_all_noIG,
#   metadata = MBL_CLL_subtype_annot_no_IG,
#   col.name = "MBL_CLL_subtype"
# )
```

```{r}
# umap_patient_MBL_CLL_subtype_no_IG <- DimPlot(subset(x=gex_scaled_10X_seurat_all_noIG,
#                                               cells = nonMBL7dup_cells),
#   reduction = "umap",
#   group.by = "patient_id",
#   raster = F,
#   cols = c(my_unpaired_colours_named, control_colours_named),
#   split.by = "MBL_CLL_subtype",
#   shuffle = TRUE
# ) +
#   labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Patients") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "bottom",
#     aspect.ratio = 1
#   )
# 
# umap_patient_MBL_CLL_subtype_no_IG %>%
#   ggsave(.,
#     filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_patients_disease_subtype_split.png"),
#     unit = "in",
#     width = 16,
#     height = 6
#   )
# 
# umap_patient_MBL_CLL_subtype_no_IG
```

#### Celltype markers
```{r}
marker_genes <- c(
  "PAX5", "MS4A1", "EBF1", "MME", # B-cell trajectories
  "CD14", "CEBPB", "MPO", "CD33", # Monocytes
  "CD3D", "CD8A", "TBX21", "IL7R", "CD3E", "CD3G", "CD4", "CD8B", # TCells
  "CD19", "CD5", "FCER2", "CD79B", # CLL B-Cell Marker
  "GNLY", "NKG7", "NCAM1", # NK cells
  "PF4", # platelets
  "CA1", "HBA1" # erythrocytes
)
```

```{r fig.width=20, fig.height=15}
# marker_vln_no_IG <- VlnPlot(subset(gex_scaled_10X_seurat_all_noIG, cells=nonMBL7dup_cells),
#   features = marker_genes,
#   group.by = "seurat_clusters",
#   slot = "scale.data",
#   pt.size = FALSE,
#   cols = color_palette
# ) +
#   theme_pub() +
#   theme(legend.position = "none")
# 
# marker_vln_no_IG %>%
#   ggsave(.,
#     file = file.path(plotting_dir, "Marker_expression_vlnplot_all_no_IG.png"),
#     units = "in",
#     width = 20,
#     height = 16
#   )
# 
# marker_vln_no_IG
```

```{r, fig.width=20, fig.height=15}
# marker_umaps_no_IG <- FeaturePlot(gex_scaled_10X_seurat_all_noIG,
#                                   cells = nonMBL7dup_cells,
#   reduction = "umap",
#   features = marker_genes,
#   raster = FALSE
# ) +
#   theme_pub() +
#   theme(legend.position = "none")
# 
# marker_umaps_no_IG %>%
#   ggsave(.,
#     file = file.path(plotting_dir, "Marker_expression_umaps_all_no_IG.png"),
#     units = "in",
#     width = 12,
#     height = 16
#   )
# 
# marker_umaps_no_IG
```

```{r}
# saveRDS(gex_scaled_10X_seurat_all_noIG, file = file.path(RDS_dir, "gex_scaled_10X_seurat_all_noIG.RDS"))
```

#### Celltype annotation
```{r}
malignant_celltype_annotation <- tibble(
  "label_cellbarcode" = colnames(gex_scaled_all),
  "My_annotation" = gex_scaled_all$My_annotation
) %>%
  dplyr::mutate(My_annotation = case_when(
    My_annotation == "MBL B Cell" & str_starts(label_cellbarcode,
      pattern = paste0("MBL", 10:15, collapse = "|")
    ) ~ "LC-MBL B Cell",
    My_annotation == "MBL B Cell" & str_starts(label_cellbarcode,
      pattern = paste0("MBL", 10:15, collapse = "|"),
      negate = TRUE
    ) ~ "HC-MBL B Cell",
    .default = My_annotation
  ))

healthy_celltype_annotation <- tibble(
  "label_cellbarcode" = colnames(gex_scaled_10X_seurat_all_noIG),
  "My_annotation" = gex_scaled_10X_seurat_all_noIG$predicted.celltype.l1
) %>%
  dplyr::filter(str_starts(label_cellbarcode, "pbmc1|pbmc2|pbmc3|pbmc4")) %>%
  dplyr::mutate(My_annotation = case_when(My_annotation == "CD4 T" ~ "T",
    My_annotation == "CD8 T" ~ "T",
    My_annotation == "other T" ~ "T",
    .default = My_annotation
  ))

celltype_annotation <- bind_rows(malignant_celltype_annotation, healthy_celltype_annotation) %>% deframe()

gex_scaled_10X_seurat_all_noIG <- AddMetaData(
  object = gex_scaled_10X_seurat_all_noIG,
  metadata = celltype_annotation,
  col.name = "My_annotation"
)
```

#### Celltype
```{r}
# umap_celltype_no_IG <- DimPlot(gex_scaled_10X_seurat_all_noIG,
#                                cells=nonMBL7dup_cells,
#   reduction = "umap",
#   group.by = "My_annotation_LC",
#   label = FALSE,
#   repel = TRUE,
#   raster = FALSE,
#   cols = annotation_colours_named,
#   shuffle = TRUE
# ) +
#   labs(x = "UMAP1", y = "UMAP2", colour = "Cell types", title = "Cell types") +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "none",
#     aspect.ratio = 1
#   )
# 
# ggsave(umap_celltype_no_IG,
#   file = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_celltypes.png"),
#   height = 4,
#   width = 4
# )
# 
# umap_celltype_no_IG
```

```{r}
# umap_celltype_no_IG_MBL_CLL_subtype_split <- DimPlot(subset(x=gex_scaled_10X_seurat_all_noIG, cells=nonMBL7dup_cells),
#   reduction = "umap",
#   split.by = "MBL_CLL_subtype",
#   group.by = "My_annotation_LC",
#   label = FALSE,
#   repel = TRUE,
#   raster = FALSE,
#   cols = annotation_colours_named,
#   shuffle = TRUE
# ) +
#   labs(
#     x = "UMAP1",
#     y = "UMAP2",
#     colour = "Cell type",
#     title = "Cell type"
#   ) +
#   theme_pub() +
#   theme(
#     axis.ticks = element_blank(),
#     axis.text.y = element_blank(),
#     axis.text.x = element_blank(),
#     legend.position = "none",
#     aspect.ratio = 1
#   )
# 
# ggsave(umap_celltype_no_IG_MBL_CLL_subtype_split,
#   file = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_celltypes_MBL_CLL_subtype_split.png"),
#   height = 4,
#   width = 16
# )
# 
# write_tsv(umap_celltype_no_IG_MBL_CLL_subtype_split[[1]]$data, file = file.path(tsv_dir, "UMAP_10X_seurat_all_no_IG_celltypes_MBL_CLL_subtype_split_data.tsv"))
# 
# umap_celltype_no_IG_MBL_CLL_subtype_split
```

```{r}
umap_celltype_no_IG_MBL_CLL_subtype_split_noB <- DimPlot(subset(x=gex_scaled_10X_seurat_all_noIG, cells=nonMBL7dup_cells),
  reduction = "umap",
  split.by = "MBL_CLL_subtype",
  group.by = "My_annotation",
  label = FALSE,
  repel = TRUE,
  raster = FALSE, 
  cols = annotation_colours_named,
  shuffle = TRUE
) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    colour = "Cell type",
    title = "Cell type"
  ) +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "none",
    aspect.ratio = 1
  )

write_tsv(umap_celltype_no_IG_MBL_CLL_subtype_split_noB[[1]]$data, file = file.path(tsv_dir, "20250326_UMAP_10X_seurat_all_no_IG_celltypes_MBL_CLL_subtype_split_noB_data.tsv"))

umap_celltype_no_IG_MBL_CLL_subtype_split_noB
```

## Combined IG genes removal with scVI technology integration

```{r}
gex_scVI_10X_seurat_all_noIG <- readRDS(file.path(RDS_dir, "gex_scaled_10X_seurat_all_scvi_MBL_CLL_healthy_no_IG.RDS"))
```

### Plot UMAPs

```{r}
color_palette <- colorRampPalette(brewer.pal(12, "Set3"))(60)
```

#### Clusters

```{r}
umap_clusters_plot_no_IG_scVI <- DimPlot(gex_scVI_10X_seurat_all_noIG,
                              reduction = "scvi",
                              raster = FALSE,
                              group.by = c("seurat_clusters"),
                              cols=color_palette)

umap_clusters_plot_no_IG_scVI %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_scVI_clusters.png"),
         unit = "in",
         width = 8,
         height = 6)
```

#### Samples
```{r}
# add plotting id
plotting_annot_no_IG_scVI <- as_tibble(gex_scVI_10X_seurat_all_noIG$orig.ident, rownames=NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::rename("label"="value") %>%
  left_join(metadata %>%
              dplyr::select(label, plotting_id)) %>%
  dplyr::select(label_cellbarcode, plotting_id) %>%
  deframe()

gex_scVI_10X_seurat_all_noIG <- AddMetaData(object = gex_scVI_10X_seurat_all_noIG,
                                         metadata = plotting_annot_no_IG_scVI,
                                         col.name = "plotting_id")
```

```{r}
umap_sample_plot_no_IG_scVI <- DimPlot(gex_scVI_10X_seurat_all_noIG,
                              reduction = "scvi",
                            raster = FALSE,
                            group.by = c("orig.ident"),
                            cols=c(my_unpaired_colours_named,my_paired_colours_named, control_colours_named)) +
  theme_pub() +
  theme(
    aspect.ratio = 1,
    legend.position = "right"
  )+
  labs(title="Samples")+
  theme_pub()+
  theme(aspect.ratio = 1,
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y=element_blank())

umap_sample_plot_no_IG_scVI %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_scVI_samples.png"),
         unit = "in",
         width = 8,
         height = 6)
```

#### Patients 
```{r}
# add patient id
patient_id_annot_no_IG_scVI <- as_tibble(gex_scVI_10X_seurat_all_noIG$orig.ident, rownames=NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::rename("label"="value") %>%
  left_join(metadata %>%
              dplyr::select(label, patient_id)) %>%
  dplyr::select(label_cellbarcode, patient_id) %>%
  deframe()

gex_scVI_10X_seurat_all_noIG <- AddMetaData(object = gex_scVI_10X_seurat_all_noIG,
                                              metadata = patient_id_annot_no_IG_scVI,
                                              col.name = "patient_id")
```

```{r}
umap_patient_no_IG_scVI <- DimPlot(gex_scVI_10X_seurat_all_noIG,
                        reduction = "scvi",
                        group.by = "patient_id",
                        raster = F,
                        cols = c(my_unpaired_colours_named, control_colours_named),
                        shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", title = "All Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "bottom",
        aspect.ratio=1)

umap_patient_no_IG_scVI %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_scVI_patients.png"),
         unit = "in",
         width = 7,
         height = 4)
```

#### Patients disease-split
```{r}
sample_annot_no_IG_scVI <- tibble("label_cellbarcode" = WhichCells(gex_scVI_10X_seurat_all_noIG))

disease_annot_no_IG_scVI <- sample_annot_no_IG_scVI %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  tidyr::separate(label, c("patient_id", "sample"), sep = "\\.", remove = FALSE) %>%
  dplyr::left_join(metadata %>% dplyr::select(label, disease) %>% unique()) %>%
  dplyr::mutate(disease=if_else(condition=is.na(disease) & patient_id %in% paste0("pbmc", 1:4),true= "Healthy", false=disease),
                disease=factor(disease, levels=c("Healthy", "MBL", "CLL"))) %>%
  dplyr::select(label_cellbarcode, disease) %>%
  deframe()

gex_scVI_10X_seurat_all_noIG <- AddMetaData(object = gex_scVI_10X_seurat_all_noIG,
                                         metadata = disease_annot_no_IG_scVI,
                                         col.name = "disease")
```

```{r}
umap_patient_disease_no_IG_scVI <- DimPlot(gex_scVI_10X_seurat_all_noIG,
                                reduction = "scvi",
                                group.by = "patient_id",
                                raster = F,
                                cols = c(my_unpaired_colours_named, control_colours_named),
                                split.by = "disease",
                                shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Disease") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "bottom",
        aspect.ratio = 1
  )

umap_patient_disease_no_IG_scVI %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_scVI_patients_disease_split.png"),
         unit = "in",
         width = 24,
         height = 6)
```

#### Patients disease-subtype-split
```{r}
# add MBL_CLL_subtype annotation
MBL_CLL_subtype_annot_no_IG_scVI <- as_tibble(gex_scVI_10X_seurat_all_noIG$orig.ident, rownames=NA) %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::rename("label"="value") %>%
  dplyr::left_join(metadata %>% dplyr::select(label, MBL_subtype)) %>%
  dplyr::mutate(MBL_CLL_subtype=if_else(is.na(MBL_subtype) & !(label %in% paste0("pbmc", 1:4)), "CLL", MBL_subtype),
                MBL_CLL_subtype=if_else(condition=is.na(MBL_CLL_subtype) & label %in% paste0("pbmc", 1:4), true= "Healthy", false=MBL_CLL_subtype),
                MBL_CLL_subtype=factor(MBL_CLL_subtype, levels=c("Healthy", "LC", "HC", "CLL"))) %>%
  dplyr::select(label_cellbarcode, MBL_CLL_subtype) %>%
  deframe()

gex_scVI_10X_seurat_all_noIG <- AddMetaData(object = gex_scVI_10X_seurat_all_noIG,
                                         metadata = MBL_CLL_subtype_annot_no_IG_scVI,
                                         col.name = "MBL_CLL_subtype")
```

```{r}
umap_patient_MBL_CLL_subtype_no_IG_scVI <- DimPlot(gex_scVI_10X_seurat_all_noIG,
                                                   reduction = "scvi",
                                                   group.by = "patient_id",
                                        raster = F,
                                        cols = c(my_unpaired_colours_named, control_colours_named),
                                        split.by = "MBL_CLL_subtype",
                                        shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Patient", title = "Patients") +
  theme_pub() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "bottom",
        aspect.ratio = 1
  )

umap_patient_MBL_CLL_subtype_no_IG_scVI %>%
  ggsave(.,
         filename = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_scVI_patients_disease_subtype_split.png"),
         unit = "in",
         width = 32,
         height = 6)
```

#### Celltype markers
```{r}
marker_genes <- c(
  "PAX5", "MS4A1", "EBF1", "MME", # B-cell trajectories
  "CD14", "CEBPB", "MPO", "CD33", # Monocytes
  "CD3D", "CD8A", "TBX21", "IL7R", "CD3E", "CD3G", "CD4", "CD8B", # TCells
  "CD19", "CD5", "FCER2", "CD79B", # CLL B-Cell Marker
  "GNLY", "NKG7", "NCAM1", # NK cells
  "PF4", # platelets
  "CA1", "HBA1" # erythrocytes
)
```

```{r}
marker_vln_no_IG_scVI <- VlnPlot(gex_scVI_10X_seurat_all_noIG, features = marker_genes) +
  theme_pub() +
  theme(legend.position = "none")

marker_vln_no_IG_scVI %>%
  ggsave(., file = file.path(plotting_dir, "Marker_expression_vlnplot_all_no_IG_scVI.png"),
         units = "in",
         width = 20,
         height = 16)
```

```{r}
marker_umaps_no_IG_scVI <- FeaturePlot(gex_scVI_10X_seurat_all_noIG,
                                  reduction = "scvi",
                                  features = marker_genes, raster = FALSE) +
                                  theme_pub() +
                                  theme(legend.position = "none")

marker_umaps_no_IG_scVI %>%
  ggsave(., file = file.path(plotting_dir, "Marker_expression_umaps_all_no_IG_scVI.png"),
         units = "in",
         width = 12,
         height = 16)
```

```{r}
# saveRDS(gex_scVI_10X_seurat_all_noIG,
#         file = file.path(RDS_dir, "gex_scaled_10X_seurat_all_scvi_MBL_CLL_healthy_no_IG.RDS"))
```

#### Celltype

```{r}
gex_scVI_10X_seurat_all_noIG <- AddMetaData(
  object = gex_scVI_10X_seurat_all_noIG,
  metadata = celltype_annotation,
  col.name = "My_annotation_LC")

umap_celltype_no_IG_scVI <- DimPlot(gex_scVI_10X_seurat_all_noIG,
  reduction = "scvi",
  group.by = "My_annotation_LC",
  label = FALSE,
  repel = TRUE,
  raster = FALSE,
  cols = annotation_colours_named,
  shuffle = TRUE) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Cell types", title = "Cell types") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "none",
    aspect.ratio = 1)

ggsave(umap_celltype_no_IG_scVI,
  file = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_scVI_celltypes.png"),
  height = 4,
  width = 4)

umap_celltype_no_IG_scVI
```

```{r}
umap_celltype_no_IG_integrated_MBL_CLL_subtype_split <- DimPlot(gex_scVI_10X_seurat_all_noIG,
  reduction = "scvi",
  split.by = "MBL_CLL_subtype",
  group.by = "My_annotation_LC",
  label = FALSE,
  repel = TRUE,
  raster = FALSE,
  cols = annotation_colours_named,
  shuffle = TRUE) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    colour = "Cell type",
    title = "Cell type") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "none",
    aspect.ratio = 1)

ggsave(umap_celltype_no_IG_integrated_MBL_CLL_subtype_split,
  file = file.path(plotting_dir, "UMAP_10X_seurat_all_no_IG_integrated_celltypes_MBL_CLL_subtype_split.png"),
  height = 4,
  width = 16
)

write_tsv(umap_celltype_no_IG_integrated_MBL_CLL_subtype_split[[1]]$data, file = file.path(tsv_dir, "UMAP_10X_seurat_all_no_IG_integrated_celltypes_MBL_CLL_subtype_split_data.tsv"))

umap_celltype_no_IG_integrated_MBL_CLL_subtype_split
```

#### Technology
```{r}
sequencing_date_annotation <- tibble(
  "label_cellbarcode" = colnames(gex_scVI_10X_seurat_all_noIG),
  "label" = gex_scVI_10X_seurat_all_noIG$orig.ident
) %>%
  left_join(metadata %>% dplyr::select(label, sequencing_date)) %>%
  dplyr::mutate(sequencing_date=as.character(sequencing_date)) %>%
  dplyr::select(label_cellbarcode, sequencing_date) %>% deframe

gex_scVI_10X_seurat_all_noIG <- AddMetaData(object = gex_scVI_10X_seurat_all_noIG,metadata = sequencing_date_annotation, col.name = "sequencing_date")

umap_sequencing_date_no_IG_scVI <- DimPlot(gex_scVI_10X_seurat_all_noIG,
  reduction = "scvi",
  split.by = "MBL_CLL_subtype",
  group.by = "sequencing_date",
  label = FALSE,
  repel = TRUE,
  raster = FALSE,
  cols = c("2022-02-28" = "#E6AB02",
          "2022-11-18" = "#7FC97F",
          "2023-05-02" = "#FFFF99",
          "2024-03-18" = "#386CB0",
          "2022-02-26" = "#FFEF40"),
  shuffle = TRUE
) +
  labs(x = "UMAP1", y = "UMAP2", colour = "Cell types", title = "Sequencing date") +
  theme_pub() +
  theme(
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "none",
    aspect.ratio = 1)
umap_sequencing_date_no_IG_scVI
```

