---
title: "Mitochondrial mutation statistics"
author: "Anja_Rathgeber"
date: "`r Sys.Date()`"
output: html_document
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---
#GOAL
The aim of this script is to investigate how many mt mutations we have per cell and how deep the mutations have been covered.

# Load libraries
```{r}
library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(ggsignif)
```

# Set paths
```{r}
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
library_path <- file.path(data_dir, "ATAC")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "mtDNA_mutation_analysis")
plotting_dir <- file.path(analysis_dir, "Plots")
tsv_dir <- file.path(analysis_dir, "tsv_data")
RDS_dir <- file.path(analysis_dir, "RDS_Objects")
ArchR_dir_Annotation <- file.path(project_dir_ext, "ArchR_Projects_4_Annotation")
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
```

```{r}
metadata <- read_tsv(file = file.path(data_dir, "20250403_Tidy_MBL_CLL_metadata.tsv"))%>% 
  tidyr::unite("MBL_CLL_subtype", c(MBL_subtype, disease), sep = "_", remove = FALSE, na.rm = TRUE)

variant.df <- c(
    readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/variant.df.RDS")),
    readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/variant.df.RDS")),
    readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/variant.df.RDS")),
    readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/variant.df.MBL.RDS")),
    readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/variant.df.MBL.RDS")),
    readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/variant.df.MBL.RDS")))

clonal_mt_VAFs <- list(readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/umap_embedding_mito_cln.RDS")),
                       readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset1/RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
                       readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/umap_embedding_mito_cln.RDS")),
                       readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset2/RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
                       readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/umap_embedding_mito_cln.RDS")),
                       readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset3/RDS_objects/umap_embedding_mito_cln_MBL.RDS")),
                       readRDS(file.path(ArchR_dir_Annotation, "ArchR_Subset4/RDS_objects/umap_embedding_mito_cln_MBL.RDS"))
                       ) %>% bind_rows() %>% dplyr::select(-c(UMAP1, UMAP2, sample)) %>% 
  tidyr::unite("label_cellbarcode", label:cell_barcode, sep = "#", remove = FALSE ) %>% 
  tidyr::separate(col = variant_id, into = c("position", "variant"), sep = "(?<=\\d)(?=[A-Z])", remove=FALSE) %>% 
  dplyr::filter(patient_id != "MBL7dup",
                variant_id != "310T>C")

ref_all <- read_tsv(file.path(data_dir, "reference_genomes/chrM_refAllele.txt"), col_names = c("V1", "V2"))

mt_gene_annotation <- readr::read_csv(file.path(data_dir, "reference_genomes/mt_genome_gene_annotation/NCBI RefSeq Annotation GCF_000001405.40-RS_2024_08.csv"), 
                               col_names = TRUE, 
                               skip = 1) %>% 
  dplyr::mutate(`Gene symbol`= case_when(str_starts(`Gene symbol`, pattern =  "TRN") ~ "tRNA",
                                         str_starts(`Gene symbol`, pattern =  "RNR") ~ "rRNA", .default =`Gene symbol`))

celltype_annotation <- read_tsv(file.path(result_dir, "20250129_ATAC_celltype_annotation_healthy.tsv"))

label2pool <- metadata %>% 
  dplyr::filter(data_type =="ATAC",
                !(patient_id %in% c("CLL6", "MBL7dup"))) %>% 
  dplyr::select(label, pool) %>% 
  deframe()

label2plotting_id <- metadata %>% 
  dplyr::filter(data_type =="ATAC",
                !(patient_id %in% c("CLL6", "MBL7dup"))) %>% 
  dplyr::select(label, plotting_id) %>% 
  deframe()

mtDNA_depth <- file.path(library_path, paste0(label2pool, ".mgatk"), paste0(label2pool, ".depthTable.txt")) %>% 
  purrr::set_names(names(label2pool)) %>% 
  imap(~read_tsv(., col_names = c("cellbarcode", "depth")) %>% 
        dplyr::mutate(label_cellbarcode=paste0(.y, "#", cellbarcode)))

mt_gene_annotation <- readr::read_csv(file.path(data_dir, "reference_genomes/mt_genome_gene_annotation/NCBI RefSeq Annotation GCF_000001405.40-RS_2024_08.csv"), 
                               col_names = TRUE, 
                               skip = 1) %>% 
  dplyr::mutate(`Gene symbol`= case_when(str_starts(`Gene symbol`, pattern =  "TRN") ~ "tRNA",
                                         str_starts(`Gene symbol`, pattern =  "RNR") ~ "rRNA", .default =`Gene symbol`))

```

```{r}
patient_id2plotting_id <- metadata %>%
  dplyr::filter(data_type=="ATAC") %>% 
  dplyr::select(patient_id, plotting_id) %>%
  deframe()

patient2IGHV <- metadata %>% 
  dplyr::select(patient_id, IGHV_inferred) %>% 
  unique() %>% 
  deframe
```

# Analysis

## Total number of mtDNA mutation profiles
```{r}
cell_count_mtDNA_profiles <- clonal_mt_VAFs %>% 
  dplyr::filter(!(patient_id %in% c("MBL6", "MBL7dup"))) %>% 
  dplyr::select(label_cellbarcode) %>% 
  pull() %>% 
  unique() %>% 
  length()
cell_count_mtDNA_profiles
```


```{r}
mutation_count_data <- clonal_mt_VAFs %>% 
  dplyr::mutate(position=as.numeric(position) )%>%
  left_join(celltype_annotation %>% 
                               dplyr::select(label_cellbarcode, My_annotation_healthy), by = join_by(label_cellbarcode)) %>% 
  filter(!(is.na(My_annotation_healthy)),
         My_annotation_healthy %in% c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")) %>% 
  dplyr::select(patient_id, variant_id, position) %>% 
  unique() %>% 
  group_by(variant_id) %>% 
  dplyr::mutate(patient_counts=n_distinct(patient_id)) %>% 
  arrange(patient_counts) %>% 
  ungroup()
```

## Mean mtDNA mutation number per case

```{r}
mean_mtDNA_variant_counts <- mutation_count_data %>% 
  dplyr::group_by(patient_id) %>% 
  dplyr::summarize(variant_counts=n()) %>% 
  arrange(variant_counts) %>% 
  dplyr::summarize(mean_mtDNA_counts=mean(variant_counts),
                   min_mtDNA_counts=min(variant_counts),
                   max_mtDNA_counts=max(variant_counts))

mean_mtDNA_variant_counts
```

## Total and non recurrent mtDNA mutations

```{r}
non_recurrent_mtDNA_mutations <- mutation_count_data %>% 
  dplyr::select(variant_id, patient_counts) %>% 
  unique() %>% 
  dplyr::summarize(total_mtDNA_mutation_count=n_distinct(variant_id),
                   non_recurrent_mtDNA_mutations=sum(patient_counts==1))
non_recurrent_mtDNA_mutations
```

## mtDNA mutation recurrance

### Top 3 recurrent mutations
```{r}
top3_recurrent_mtDNA_mutations <- mutation_count_data %>% 
  dplyr::select(variant_id, patient_counts) %>% 
  unique() %>% 
  dplyr::arrange(desc(patient_counts)) %>% 
  slice_head(n=3) %>% 
  deframe

top3_recurrent_mtDNA_mutations
```

```{r}
axis_data <- tibble("mt_genome"=c(0, 16569))

mt_count_radial_plot <- ggplot() +
  geom_line(data= axis_data, aes(x=mt_genome, y= 2.1), colour="lightgrey", linewidth = 0.5)+
  geom_line(data= axis_data, aes(x=mt_genome, y= 2.4), colour="lightgrey", linewidth = 0.5)+
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="plus"), 
            aes(xmin = Start, xmax = Stop, ymin=2, ymax=2.2, fill = `Gene symbol`), colour = "black", linewidth = 0.3) + 
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="minus"), aes(xmin = Start, xmax = Stop, ymin=2.3, ymax=2.5, fill = `Gene symbol`), colour="black", linewidth = 0.3)+
  geom_jitter(data = mutation_count_data %>% dplyr::select(variant_id, position, patient_counts) %>% unique, aes(x = position, y = 3, colour = patient_counts), shape=16, size = 0.5) +
  geom_rect(data=tibble("position"=seq(0, 16659, 200)), aes(xmin=position,xmax=position+10, ymin=3.5, ymax=3.6), linewidth = 0.2, colour="black", fill="black")+
  geom_text(data=axis_data, aes(x=0, y=0, label="mt\ngenome"))+
  scale_x_continuous(limits=c(0, 16659), breaks=seq(0, 16569, 2000), labels = seq(0, 16569, 2000))+
  scale_y_continuous(breaks=c(3.5), limits = c(0, 3.7))+
  scale_color_distiller(name="Patients with\nmtDNA mutation", palette = "YlOrRd", direction = 1, breaks=c(1, 5, 10, 15, 20)) +
  scale_fill_manual(name="Gene", 
                    values=c("ATP6"="#15512A", "ATP8"="#2E8565","COX1"="#4EB1A9","COX2"="#8CB5C0","COX3"="#1A205F","CYTB"="#483591",
                             "ND1"="#885EB4","ND2"="#BB9AC4","ND3"="#6C2047","ND4"="#9C3D52","ND4L"="#B7716E","ND5"="#CAB3A8","ND6"="#757927",
                             "rRNA"="#95BB7D", "tRNA"="lightgrey"))+
  coord_radial(rotate_angle = TRUE, expand = FALSE) +
  theme_pub()+
  theme(panel.background = element_rect(fill = 'white'),
        panel.grid.major.y = element_line(colour = "black"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y =  element_blank())+
  guides(colour = guide_legend(title="Patient-shared mutations",
                              override.aes=list(size=3), ncol = 5, byrow=TRUE),
         fill = guide_legend(ncol=3, byrow=TRUE))
mt_count_radial_plot
saveRDS(mt_count_radial_plot, file.path(RDS_dir, "mt_count_radial_plot.RDS"))
```

### All recurrent mutations
```{r}
recurrant_mutations <- mutation_count_data %>% 
  dplyr::select(variant_id, position, patient_counts) %>% 
  unique() %>% 
  dplyr::arrange(desc(patient_counts)) %>% 
  dplyr::filter(patient_counts >1) %>% 
  dplyr::select(variant_id, patient_counts)

write_excel_csv(recurrant_mutations, file.path(tsv_dir, "20250703_all_recurrent_mtDNA_mutations.tsv"))
```


### Top 20 recurrent mutations
```{r}
top20_mutations <- mutation_count_data %>% 
  dplyr::select(variant_id, position, patient_counts) %>% 
  unique() %>% 
  dplyr::arrange(desc(patient_counts)) %>% 
  slice_head(n=20)

write_excel_csv(top20_mutations, file.path(tsv_dir, "20250417_top20_recurrent_mtDNA_mutations.tsv"))

top20_mutations_heteroplasmies <- clonal_mt_VAFs %>% 
  dplyr::filter(variant_id %in% top20_mutations$variant_id) %>% 
  dplyr::group_by(variant_id, patient_id) %>% 
  dplyr::summarize(mean_heteroplasmy_perc=round(mean(VAF)*100, digits = 2),
                   median_heteroplasmy_perc=round(median(VAF)*100), digits=2)

write_excel_csv(top20_mutations_heteroplasmies %>% 
                  dplyr::mutate(plotting_id=factor(patient_id2plotting_id[patient_id], levels=paste0("P", 1:24))) %>%
                  dplyr::select(-c(patient_id,median_heteroplasmy_perc)) %>%
                  arrange(plotting_id) %>% 
                  tidyr::pivot_wider(id_cols = variant_id, names_from = plotting_id, values_from = mean_heteroplasmy_perc), file.path(tsv_dir, "20250716_top20_recurrent_mtDNA_mutations_heteroplasmies.tsv"))
```


## Patient exclusive mtDNA mutations
```{r, fig.width=3, fig.height=4}
percentage_malignant_cells <- clonal_mt_VAFs %>% 
  left_join(celltype_annotation %>% 
            dplyr::select(label_cellbarcode, My_annotation_healthy)) %>% 
  dplyr::filter(!is.na(My_annotation_healthy),
                My_annotation_healthy %in% c("LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")) %>% 
  dplyr::select(patient_id, variant_id, label_cellbarcode, VAF) %>% 
    group_by(patient_id) %>%
  mutate(total_cells = n_distinct(label_cellbarcode)) %>% 
  group_by(variant_id, patient_id, total_cells) %>% 
  dplyr::summarise(
    variant_cell_count = sum(VAF != 0),  
    percent_cells = 100 * variant_cell_count / total_cells) %>%
  ungroup() %>% 
  unique() %>% 
  pivot_wider(id_cols = variant_id, names_from = patient_id, values_from = percent_cells) %>% 
  column_to_rownames("variant_id") %>% 
  as.matrix()

percentage_malignant_cells[is.na(percentage_malignant_cells)] <- 0

patient_order <- metadata %>%
  dplyr::filter(data_type=="ATAC") %>%  
  dplyr::filter(patient_id !="MBL7dup") %>% 
  dplyr::select(patient_id, plotting_id) %>% 
  unique %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, levels=paste0("P", 1:24))) %>% 
  arrange(plotting_id) %>% 
  pull(patient_id)

mutation_order <- c() 
for (patient in patient_order) {
  ordered_variants <- percentage_malignant_cells %>%
    as_tibble(rownames = "variant_id") %>%
    dplyr::select(variant_id, all_of(patient)) %>%
    filter(.data[[patient]] > 0) %>%  
    arrange(.data[[patient]]) %>%  
    pull(variant_id)  

  # Append unique values while keeping order
  mutation_order <- unique(c(mutation_order, ordered_variants))
}

# In case we want to annotate by condition as well
patient_condition <- metadata %>% 
  dplyr::filter(data_type=="ATAC") %>%
  dplyr::select(patient_id, MBL_CLL_subtype) %>%
  tidyr::pivot_wider(id_cols = patient_id, names_from=MBL_CLL_subtype, values_from = MBL_CLL_subtype) %>%
  unique() %>%
  dplyr::mutate(Condition=
                 case_when(!is.na(LC_MBL) ~ "LC-MBL",
                           !is.na(HC_MBL) & !is.na(CLL) ~ "CLL",
                            is.na(HC_MBL) & !is.na(CLL) ~ "CLL",
                           !is.na(HC_MBL) & is.na(CLL) ~ "HC-MBL",
                           .default=NA)) %>%
  dplyr::select(patient_id, Condition) %>% 
  deframe()

patient2condition <- my_MBL_CLL_colours_named[patient_condition]
names(patient2condition) <- names(patient_condition)

# Create a column annotation
top_annotation_col_anno <- HeatmapAnnotation(
  Condition= anno_simple(patient_order, col=patient2condition, border=TRUE),
  Patient = anno_simple(patient_order, col=c(my_unpaired_colours_named, my_paired_colours_named),
  border = TRUE),
  IGHV = anno_simple(patient2IGHV[patient_order], col=IGHV_colours, border=TRUE),
  annotation_height = unit(4, "mm"),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 10,
                            fontfamily ="Arial"))


percentage_malignant_cells <- -log10(percentage_malignant_cells)
percentage_malignant_cells[percentage_malignant_cells == Inf] <- NA

hm <- Heatmap(matrix = percentage_malignant_cells[mutation_order, patient_order], # values from -2 to 1 --> 0.1 - 100%
              col=colorRamp2(c(0, 2), c("darkred", "yellow")), # 0.01- >=1 %
              top_annotation = top_annotation_col_anno,
              cluster_rows = FALSE,
              cluster_columns = FALSE,
              na_col = "white",
                border_gp = gpar(col = "black", lty = 1),
                raster_by_magick = TRUE,
                use_raster = TRUE,
                raster_quality = 6,
                
                #column parameters
                column_title = "",
                column_title_gp = gpar(fontsize = 5),
                show_column_names = FALSE,
                
                #row parameters
                row_title = "677 mtDNA mutations",
                row_title_gp = gpar(fontsize = 10),
                show_row_names = FALSE,
                heatmap_legend_param = list(
                  title = "% (pre-)malignant cells",
                  legend_direction = "horizontal",
                  at = c(2, 1,0), # 0.01, 0.1, ≥1 % 
                  labels = c("0.01","0.1", "≥ 1"),
                  title_gp = gpar(fontsize = 10, fontface = "plain") ,
                  legend_width = unit(1.7, "cm"),
                  labels_gp = gpar(
                       fontsize = 10,
                       fontface = "plain"
                  )
                ))               

draw(hm, heatmap_legend_side = "bottom")

hm_grob <- grid::grid.grabExpr(draw(hm, heatmap_legend_side = "bottom"))
cowplot::plot_grid(hm_grob, nrow=1, ncol=1)
saveRDS(hm_grob, file.path(RDS_dir, "mtDNA_mutation_patient_heatmap.RDS"))
```

## mtDNA mutations per celltype

```{r}
#Which mutations have pseudobulk heteroplasmy >0.1% per cell type
mean_clonal_mt_VAFs <- clonal_mt_VAFs %>% 
  left_join(celltype_annotation %>% 
                               dplyr::select(label_cellbarcode, My_annotation_healthy), by = join_by(label_cellbarcode)) %>% 
  filter(!(is.na(My_annotation_healthy))) %>% 
  group_by(variant_id, patient_id, My_annotation_healthy) %>% 
  dplyr::summarize(mean_VAF=mean(VAF)) %>% 
  dplyr::filter(mean_VAF > 0.001)

#Filtering for >0.1% pseudobulk heteroplasmy of mt variant in cell type 
variant_counts_celltype <- clonal_mt_VAFs %>%
  left_join(celltype_annotation %>% 
                               dplyr::select(label_cellbarcode, My_annotation_healthy),
            by = join_by(label_cellbarcode)) %>% 
  filter(!(is.na(My_annotation_healthy))) %>% 
  dplyr::select(variant_id, patient_id, My_annotation_healthy, label_cellbarcode) %>% 
  semi_join(mean_clonal_mt_VAFs, by = c("variant_id", "patient_id", "My_annotation_healthy")) %>% 
  unique() %>% 
  group_by(patient_id, My_annotation_healthy) %>% 
  dplyr::summarize(variant_counts=n()) %>% 
  dplyr::mutate(My_annotation_healthy=factor(My_annotation_healthy, levels=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "T", "NK", "Mono", "DC", "other")))

celltype_counts_per_sample <- clonal_mt_VAFs %>%
  left_join(celltype_annotation %>% 
                               dplyr::select(label_cellbarcode, My_annotation_healthy)) %>% 
  filter(!(is.na(My_annotation_healthy))) %>% 
  dplyr::select(label_cellbarcode, patient_id, My_annotation_healthy) %>% 
  unique() %>% 
  dplyr::group_by(My_annotation_healthy, patient_id) %>% 
  dplyr::summarize(cell_counts=n())
```

```{r}
# normalize by cell counts per sample
variant_counts_celltype_norm <- variant_counts_celltype %>%
  dplyr::left_join(celltype_counts_per_sample) %>% 
  dplyr::filter(My_annotation_healthy != "other") %>% 
  dplyr::mutate(variant_counts_norm=variant_counts/cell_counts)

variant_counts_celltype_norm_plot <- variant_counts_celltype_norm %>%
dplyr::filter(!(My_annotation_healthy %in% c("other"))) %>%
dplyr::mutate(My_annotation_healthy=factor(My_annotation_healthy, levels=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "T", "NK", "Mono", "DC", "other"))) %>% 
ggplot(aes(x=My_annotation_healthy, y=variant_counts_norm, colour=My_annotation_healthy))+
  geom_boxplot(colour = "black", aes(fill = My_annotation_healthy), outliers = FALSE)+
  geom_jitter(width = 0.2, colour = "black", aes(fill = My_annotation_healthy), pch = 21, size=0.5) +
  # geom_signif(comparisons = list(c('B', 'LC_MBL_B_Cells'),
  #                                c('B', 'HC_MBL_B_Cells'),
  #                                c('B', "CLL_B_Cells"),
  #                                c("B", "T")),
  #             step_increase = 0.1, textsize = 3, colour="black") +
  scale_colour_manual("Cell type", 
                      values=c(annotation_colours_named),
                      labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL", "T"="T", "NK"= "NK", "Mono" = "Monocytes", "DC"="Dendritic", "other"="Other cell types"))+
  scale_fill_manual("Cell type", 
                    values=c(annotation_colours_named),
                    labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL", "T"="T", "NK"= "NK", "Mono" = "Monocytes", "DC"="Dendritic", "other"="Other cell types"))+
  scale_x_discrete(labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL", "T"="T", "NK"= "NK", "Mono" = "Monocytes", "DC"="Dendritic", "other"="Other cell types"), 
                   breaks=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "T", "NK", "Mono", "DC", "other"))+
  # scale_y_continuous(breaks=seq(0, 150, 50), limits = c(0, 150))+
  labs(x="Cell type", y= "Normalized\nmtDNA mutations counts\n> 0.1% mean heteroplasmy")+
  theme_pub()+
  theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1),
        axis.title.x = element_blank(),
        legend.position = "none")
variant_counts_celltype_norm_plot

write_tsv(variant_counts_celltype_norm_plot$data, file.path(tsv_dir, "variant_counts_celltype_data.tsv"))
```

## Downsampling of cells per celltype depth norm

```{r}
# how many cells are available per patient and cell type?
patient_celltype_counts <- clonal_mt_VAFs %>%
  left_join(celltype_annotation %>% 
                               dplyr::select(label_cellbarcode, My_annotation_healthy),
            by = join_by(label_cellbarcode)) %>% 
  filter(!(is.na(My_annotation_healthy))) %>% 
  dplyr::select(variant_id, patient_id, My_annotation_healthy, label_cellbarcode) %>% 
  semi_join(mean_clonal_mt_VAFs, by = c("variant_id", "patient_id", "My_annotation_healthy")) %>% 
  group_by(patient_id, My_annotation_healthy) %>% 
  dplyr::select(-variant_id) %>% 
  unique() %>% 
  dplyr::summarize(cell_counts=n()) 

patient_celltype_counts %>% 
  dplyr::mutate(My_annotation_healthy=factor(My_annotation_healthy, levels=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "T", "NK", "Mono", "DC", "other"))) %>% 
  ggplot(aes(x=My_annotation_healthy, y=cell_counts, fill=patient_id))+
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single"))+
  geom_hline(yintercept = 100, linetype = "dashed")+
  scale_y_log10(expand=c(0,0))+
  scale_fill_manual(name="", values = c(my_paired_colours_named, my_unpaired_colours_named), 
                    labels=patient_id2plotting_id)+
  theme_pub()+
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x = element_blank(
          ))
```

```{r}
# Which patients have more than 100 cells available and for which cell types does a comparison make sense?
patient_celltype_counts <- patient_celltype_counts %>% 
  dplyr::mutate(My_annotation_healthy=factor(My_annotation_healthy, levels=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "T", "NK", "Mono", "DC", "other"))) %>% 
  dplyr::filter(cell_counts >=100 & My_annotation_healthy %in% c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells"))

set.seed(123)
bcs_sampled <- clonal_mt_VAFs %>%
  left_join(celltype_annotation %>% 
                               dplyr::select(label_cellbarcode, My_annotation_healthy),
            by = join_by(label_cellbarcode)) %>% 
  dplyr::filter(My_annotation_healthy %in% c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")) %>% 
  dplyr::select(variant_id, patient_id, My_annotation_healthy, label_cellbarcode) %>% 
  semi_join(mean_clonal_mt_VAFs, by = c("variant_id", "patient_id", "My_annotation_healthy")) %>% 
  unique() %>% 
  right_join(patient_celltype_counts) %>%
  distinct(patient_id, My_annotation_healthy, label_cellbarcode) %>%
  group_by(patient_id, My_annotation_healthy) %>%
  slice_sample(n = 100) %>%  
  ungroup()
```

```{r}
downsampled_variant_counts_norm_celltype <- clonal_mt_VAFs %>%
  left_join(celltype_annotation %>% 
                               dplyr::select(label_cellbarcode, My_annotation_healthy),
            by = join_by(label_cellbarcode)) %>% 
  dplyr::filter(My_annotation_healthy %in% c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")) %>% 
  dplyr::select(variant_id, patient_id, My_annotation_healthy, label_cellbarcode) %>% 
  semi_join(mean_clonal_mt_VAFs, by = c("variant_id", "patient_id", "My_annotation_healthy")) %>% 
  distinct() %>% 
  semi_join(bcs_sampled, by = c("patient_id", "My_annotation_healthy", "label_cellbarcode")) %>%
  group_by(label_cellbarcode, patient_id, My_annotation_healthy) %>%
  dplyr::summarize(n=n(), .groups = "drop") %>%
  left_join(mtDNA_depth %>% bind_rows()) %>%
  dplyr::mutate(norm_n=n/depth) %>% 
  group_by(patient_id, My_annotation_healthy) %>% 
  dplyr::summarize(norm_n_celltype=mean(norm_n)) %>% 
  dplyr::mutate(My_annotation_healthy=factor(My_annotation_healthy, levels=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells")))
```

```{r}
downsampled_variant_counts_depthnorm_celltype_plot <- downsampled_variant_counts_norm_celltype %>%
ggplot(aes(x=My_annotation_healthy, y=norm_n_celltype, colour=My_annotation_healthy))+
  geom_boxplot(colour = "black", aes(fill = My_annotation_healthy), outliers = FALSE)+
  geom_jitter(width = 0.2, colour = "black", aes(fill = My_annotation_healthy), pch = 21, size=0.5) +
  geom_signif(comparisons = list(c('B', 'LC_MBL_B_Cells'),
                                 c('B', 'HC_MBL_B_Cells'),
                                 c('B', "CLL_B_Cells")),
             step_increase = 0.1, textsize = 3, colour="black", test = "wilcox.test") +
  scale_colour_manual("Cell type", 
                      values=c(annotation_colours_named),
                      labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL"))+
  scale_fill_manual(name="Cell type", 
                    values=c(annotation_colours_named),
                    labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL"))+
  scale_x_discrete(labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL"), 
                   breaks=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells"))+
  scale_y_continuous(expand=c(0,0),breaks=seq(0, 2, 1), limits = c(0, 2.6))+
  labs(x="Cell type", y= "mtDNA mutation counts\n> 0.1% mean heteroplasmy")+
  theme_pub()+
  theme(axis.text.x = element_text(angle=0, hjust=1, vjust=1),
        axis.title.x = element_blank(),
        legend.position = "none")

write_tsv(downsampled_variant_counts_depthnorm_celltype_plot$data, file.path(tsv_dir, "downsampled_variant_counts_depthnorm_celltype_data.tsv"))
```

## mtDNA mutation coverage per celltype
```{r}
mtDNA_depth_annotated <- mtDNA_depth %>% 
  map(~left_join(.x, celltype_annotation, by = join_by(cellbarcode, label_cellbarcode)) %>% 
        dplyr::filter(!is.na(My_annotation_healthy)) %>% 
        dplyr::select(label_cellbarcode, My_annotation_healthy, depth)) %>% 
  bind_rows() %>% 
  tidyr::separate(label_cellbarcode, into = c("label", "cellbarcode"), sep="#", remove=FALSE) %>% 
  dplyr::group_by(label, My_annotation_healthy) %>% 
  dplyr::summarize(mean_depth=mean(depth))

coverage_celltype_plot <- mtDNA_depth_annotated %>% 
  dplyr::mutate(My_annotation_healthy=factor(My_annotation_healthy, levels=c("B", "LC_MBL_B_Cells", "HC_MBL_B_Cells", "CLL_B_Cells", "T", "NK", "Mono", "DC", "other"))) %>% 
  ggplot(aes(x=My_annotation_healthy, y=mean_depth, colour=My_annotation_healthy))+
  geom_boxplot(colour = "black", aes(fill = My_annotation_healthy), outlier.shape = NA)+
  geom_jitter(width = 0.2, colour = "black", aes(fill = My_annotation_healthy), pch = 21, size=0.5) +
  geom_signif(comparisons = list(c('B', 'LC_MBL_B_Cells'),
                               c('B', 'HC_MBL_B_Cells'),
                               c('B', "CLL_B_Cells"),
                               c("B", "T")),
            step_increase = 0.1, textsize = 3, colour="black") +
  scale_colour_manual("Cell type", values=c(annotation_colours_named),labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL", "T"="T", "NK"= "NK", "Mono" = "Monocytes", "DC"="DC", "other"="Unannotated"))+
  scale_fill_manual("Cell type", values=c(annotation_colours_named),labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL", "T"="T", "NK"= "NK", "Mono" = "Monocytes", "DC"="DC", "other"="Unannotated"))+
  scale_x_discrete(labels=c("B"="B", "LC_MBL_B_Cells"="LC-MBL", "HC_MBL_B_Cells"="HC-MBL", "CLL_B_Cells"="CLL", "T"="T", "NK"= "NK", "Mono" = "Monocytes", "DC"="DC", "other"="Unannotated"))+
  scale_y_continuous(breaks=seq(0, 100, 20), limits = c(0, 100))+
  labs(x="Cell type", y= "mtDNA coverage")+
  theme_pub()+
  theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1),
        axis.title.x = element_blank(),
        legend.position = "none")

write_tsv(coverage_celltype_plot$data, file.path(tsv_dir, "coverage_celltype_data.tsv"))
ggsave(plot = coverage_celltype_plot, filename = file.path(plotting_dir, "coverage_celltype_plot.png"), width = 3, height = 3)
```

## mtDNA general coverage
```{r}
axis_data <- tibble("mt_genome"=c(0, 16569))

pull_coverage_subset<- function(SE, cells, resolution = 5){
  SE_subset <- SE[, colnames(SE) %in% cells]
  zoo::rollmean(rowMeans(assays(SE_subset)[['coverage']]), resolution)
}

pull_precise_coverage_subset <-  function(SE, cells){
  SE_subset <- SE[, colnames(SE) %in% cells]
  apply(assays(SE_subset)[["coverage"]] , 1, max, na.rm = TRUE)
}

mtmut_cells_patient <- clonal_mt_VAFs %>% 
  dplyr::select(label, cell_barcode)

# filtering to only qc accepted cells from ATAC data
coverage_rm_all_samples <- file.path(library_path, paste0(label2pool, ".mgatk"), paste0(label2pool, ".rds")) %>% 
  map2(names(label2pool), ~data.frame(
  patient_id=.y, 
  position = zoo::rollmean(1:16569, 5),
  coverage = pull_coverage_subset(SE=readRDS(.x), cells=dplyr::filter(mtmut_cells_patient, label == .y) %>% pull(cell_barcode)))) %>% 
  set_names(names(label2pool))

coverage_statistics_all_samples <- file.path(library_path, paste0(label2pool, ".mgatk"), paste0(label2pool, ".rds")) %>% 
  map2(names(label2pool), ~data.frame(
  patient_id=.y, 
  position = c(1:16569),
  max_coverage = pull_precise_coverage_subset(readRDS(.x), cells=dplyr::filter(mtmut_cells_patient, label == .y) %>% pull(cell_barcode)))) %>% 
  set_names(names(label2pool)) %>% 
  imap(~dplyr::group_by(.x, patient_id) %>% 
         dplyr::summarize(max_coverage=max(max_coverage))) %>% 
  bind_rows()

saveRDS(coverage_statistics_all_samples, file.path(RDS_dir, "coverage_statistics_all_samples.RDS"))
```


```{r fig.width=20, fig.height=20}
mt_linear_coverage_plot <- imap(coverage_rm_all_samples,
  ~ggplot() +
  geom_line(data= axis_data, aes(x=mt_genome, y= -2), colour="lightgrey", linewidth = 0.5)+
  geom_line(data= axis_data, aes(x=mt_genome, y= -5), colour="lightgrey", linewidth = 0.5)+
  
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="plus"), 
            aes(xmin = Start, xmax = Stop, ymin=-1, ymax=-3, fill = `Gene symbol`), 
            colour = "black", linewidth = 0.3) + 
  geom_rect(data = mt_gene_annotation %>% dplyr::filter(Strand=="minus"), 
            aes(xmin = Start, xmax = Stop, ymin=-4, ymax=-6, fill = `Gene symbol`), 
            colour="black", linewidth = 0.3)+
    
  geom_line(data=.x,aes(x = position, y = coverage), 
              color = "steelblue4", linewidth = 0.5)+
  
  geom_smooth(data=.x,aes(x = position, y = coverage), colour="darkred", linewidth=0.5, method = "lm")+
  scale_x_continuous(limits=c(0, 16659), breaks=seq(0, 16569, 2000), labels = seq(0, 16569, 2000))+
  scale_y_continuous(limits=c(-10, 100))+
  scale_fill_manual(name="Gene", 
                    values=c("ATP6"="#15512A", "ATP8"="#2E8565","COX1"="#4EB1A9","COX2"="#8CB5C0","COX3"="#1A205F","CYTB"="#483591",
                             "ND1"="#885EB4","ND2"="#BB9AC4","ND3"="#6C2047","ND4"="#9C3D52","ND4L"="#B7716E","ND5"="#CAB3A8","ND6"="#757927",
                             "rRNA"="#95BB7D", "tRNA"="lightgrey"))+
    labs(x="Position mt genome [bp]", y="Coverage", title=.y)+
  # coord_radial(rotate.angle = TRUE, expand = FALSE) +
  theme_pub()+
  theme(panel.background = element_rect(fill = 'white'),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none"))

ggarrange(plotlist=mt_linear_coverage_plot, nrow=8, ncol=4)

saveRDS(mt_linear_coverage_plot, file.path(RDS_dir, "mt_linear_coverage_plot.RDS"))
```

```{r}
clonal_mt_VAF_statistics <- 
  clonal_mt_VAFs %>% 
  dplyr::group_by(label) %>%
  dplyr::filter(VAF !=0) %>% 
  dplyr::summarize(min_VAF=min(VAF),
            max_VAF=max(VAF))

summary_statistics <- coverage_statistics_all_samples %>% 
  left_join(clonal_mt_VAF_statistics, by = join_by("patient_id"=="label"))

summary_statistics %>% ggplot(aes(x=min_VAF, y=max_coverage))+
  geom_point()+
  stat_smooth(method = "lm", colour="darkred" )+
  stat_cor(method="pearson", label.x = 0.007, label.y=600, )+
  scale_x_continuous(limits=c(NA, 0.011))+
  theme_pub()
```

## Transitions/Transversions
```{r fig.width=10, fig.height=8}
ti_tv_data <- clonal_mt_VAFs %>%
  dplyr::filter(!(patient_id %in% c("CLL6", "MBL7dup"))) %>%
  left_join(celltype_annotation) %>% 
        dplyr::filter(!is.na(My_annotation_healthy)) %>% 
  dplyr::select(label, patient_id, variant_id, variant, position, My_annotation_healthy) %>% 
  unique() %>% 
  dplyr::mutate(ti_tv=case_when(variant %in% c("G>A", "A>G", "C>T", "T>C") ~ "transition",
                                variant %in% c("G>C", "C>G", "A>T", "T>A", "G>T", "T>G", "C>A", "A>C") ~ "transversion"))

ti_tv_data %>%
  dplyr::mutate(plotting_id=factor(label2plotting_id[label], levels=paste0("P", 1:24))) %>%
  arrange(plotting_id) %>% 
  dplyr::mutate(label=factor(label, levels=unique(label)),
                variant=factor(variant, levels=c("A>G", "C>T", "G>A", "T>C", 
                                                 "A>C", "A>T", "C>A", "C>G", "G>C", "G>T","T>A","T>G"))) %>% 
  ggplot(aes(x=variant, fill=label))+
  geom_bar(position = position_dodge(preserve = "single"))+
  scale_fill_manual(name="Patient", values = c(my_paired_colours_named, my_unpaired_colours_named), labels=label2plotting_id)+
  scale_y_continuous(expand = c(0,0.05))+
  labs(y="unique mtDNA mutation counts", x="Transitions/Transversions")+
  theme_pub()+
  theme(legend.position = "bottom", legend.byrow = TRUE)+
  guides(fill=guide_legend(nrow = 3))#+
  # facet_wrap(~My_annotation_healthy, scales="free_x")
```

```{r}
called_variants <- clonal_mt_VAFs %>%
  dplyr::filter(!(patient_id %in% c("CLL6", "MBL7dup"))) %>%
  left_join(celltype_annotation, by = join_by(label_cellbarcode, label)) %>% 
        dplyr::filter(!is.na(My_annotation_healthy)) %>% 
  dplyr::select(label, patient_id, variant_id, variant, position, My_annotation_healthy) %>% 
  unique()

# Simple reverse complement function
reverse_complement <- function(s){
  chartr("ATGC","TACG",s)
}

# Process 3 digit signature based on letters
colnames(ref_all) <- c("pos", "ref")
ref_all$ref <- toupper(ref_all$ref)
l <- as.character(ref_all$ref)

# Gs happen to be at the first and last position
ref_all$three <- paste0(c("G", l[-length(l)]), l, c(l[-1], "G"))

# Remove Ns
ref_all <- ref_all[!grepl("N", ref_all$three),]

# Make every possible mutation
ref_all_long <- rbind(ref_all,ref_all, ref_all,ref_all)
ref_all_long$alt <- rep(c("A", "C", "G", "T"), each = dim(ref_all)[1])
ref_all_long <- ref_all_long[ref_all_long$ref != ref_all_long$alt,]

# add some meta data
ref_all_long$variant <- paste0(as.character(ref_all_long$pos), ref_all_long$ref, ">", ref_all_long$alt)
ref_all_long$change <- paste0(ref_all_long$ref, ref_all_long$alt)
ref_all_long$change_rc <- reverse_complement(paste0(ref_all_long$ref, ref_all_long$alt))

# A/G rich strand is "heavy" -- https://en.wikipedia.org/wiki/Heavy_strand
table(ref_all$ref) # so the reference strand is light (more C/T)
ref_all_long$strand <- ifelse(ref_all_long$ref %in% c("C","T"), "L", "H")

# Change to C/T as ref allele
ref_all_long$rc3 <- reverse_complement(ref_all_long$three)
ref_all_long$three_plot <- ifelse(ref_all_long$strand == "L", ref_all_long$three, ref_all_long$rc3)
ref_all_long$group_change <- ifelse(ref_all_long$strand == "L", ref_all_long$change, ref_all_long$change_rc)

# Annotate with called variants
my_variants <- called_variants$variant_id
ref_all_long$my_variants <- ref_all_long$variant %in% my_variants

total <- dim(ref_all_long)[1]
total_my_variants <- sum(ref_all_long$my_variants)
prop_df <- ref_all_long %>% 
  group_by(three_plot, group_change, strand) %>%
  dplyr::summarize(observed_prop_my_variants = sum(my_variants)/total_my_variants, expected_prop = n()/total, n = n()) %>%
  dplyr::mutate(fc_my_variants = observed_prop_my_variants/expected_prop)

prop_df$change_plot <- paste0(prop_df$group_change, "_", prop_df$three_plot)

write_tsv(prop_df, file.path(tsv_dir, "transition_transversion_data.tsv"))
```


```{r fig.width=20, fig.height=5}
# Visualize
ti_tv_plot <- ggplot(prop_df, aes(x = change_plot, fill = strand, y = fc_my_variants)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 1, linetype =2, color = "black") +
  scale_fill_manual(values= c("firebrick", "dodgerblue3")) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "Change in nucleotide", y = "Substitution Rate (Observed / Expected)")+
  theme_pub()+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_text(angle=90),
        legend.position = "bottom")
```


## Clonal variants non-default filtering
```{r}
cln.variants <- variant.df %>%
  map(~ subset(., subset = n_cells_conf_detected >= 1 & strand_correlation >= 0.65 & vmr > 0.01))
```


### Calculate all VAFs.cln
```{r}
VAFs.cln <- mito.data %>%
  map2(cln.variants, ~ AlleleFreq(.,
    variants = .y$variant,
    assay = NULL
  ))
```

### Transform data
```{r}
heteroplasmy <- VAFs.cln %>%
  map(~ as.matrix(.) %>%
    t() %>%
    as_tibble(rownames = "label_cellbarcode"))

heteroplasmy_long <- heteroplasmy %>%
  map(~ tidyr::pivot_longer(., -c(label_cellbarcode), names_to = "Variant", values_to = "Heteroplasmy") %>%
    tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "#", remove = FALSE))
```

### mtCounts per cell
```{r}
mean_mt_cts <- heteroplasmy_long %>% 
  bind_rows() %>% 
  split(.$label) %>% 
         map_dbl(~ filter(., Heteroplasmy > 0) %>% 
        nrow()/length(unique(.$label_cellbarcode))) %>% 
  enframe(name = "patient_id", value = "mean_ct_mtSNV") %>% 
  tidyr::separate(patient_id, c("patient", "sample")) %>% 
  group_by(patient) %>% 
  dplyr::summarize(patient_ct=mean(mean_ct_mtSNV))

```

### mtdepth per cell
```{r}
mean_depth <- names(variant.df) %>% 
  map_dbl(~variant.df[[.x]] %>% 
  dplyr::filter(variant %in% pull(heteroplasmy_long[[.x]], Variant)) %>% 
  summarize(mean_coverage=mean(mean_coverage)) %>% 
    pull(mean_coverage)) %>% 
  set_names(names(variant.df)) %>% 
  enframe(name ="patient", value = "mean_depth")
```

correlation plot
```{r}
scatterplot <- mean_mt_cts %>% left_join(mean_depth) %>% 
  ggplot(aes(x=mean_depth,  y=patient_ct, label=patient))+
  geom_point()+
  geom_smooth(color = "black", method = lm, se = FALSE, show.legend = FALSE) +
  geom_text_repel(aes(label = patient), size = 2.5, fontface = "bold", ) +
  labs(x="mean sequencing depth per cell", y="mean mtSNVs per cell [#]")+
  stat_cor(method = "pearson", size=2.5) +
  theme_pub()

ggsave(scatterplot, file.path(plotting_dir, "mtSNVcounts_vs_depth.png"))
```
