---
title: "Differential Gene Pseudobulk Expression analysis"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

# GOAL 
The aim of this script is to find out if there are differentially expressed genes between HC-MBLs which remained as such within 5 years post diagnosis and HC-MBLs which progressed to CLL. We use pseudobulk analysis here adapted from https://hbctraining.github.io/scRNA-seq/lessons/pseudobulk_DESeq2_scrnaseq.html 

# Load libraries

```{r Load_libraries}
library(DESeq2)
library(magrittr)
library(Matrix.utils)
library(readxl)
library(tidyverse)
library(ggpubr)
library(ggrepel)
library(Cairo)
library(Seurat)
library(SeuratDisk)
library(RColorBrewer)
library("org.Hs.eg.db")
library(pheatmap)
library(circlize)
library(SingleCellExperiment)
library(ggvenn)
library(disgenet2r)
library(ComplexHeatmap)
```

# Set paths

```{r}
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "DESeq2_analysis")
plotting_dir <- file.path(analysis_dir, "Plots")
RDS_dir <- file.path(analysis_dir, "RDS_Objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")
GEX_RDS_dir <- file.path(result_dir, "GEX_Analysis/RDS_Objects")
```

# Load data

```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
source(file.path(project_dir, "scripts/heatmap_plotting_functions.R"))
```

```{r}
gex_scaled_all <- readRDS(file.path(GEX_RDS_dir, "gex_scaled_all.RDS"))
metadata_gex <- read_tsv(file = file.path(data_dir, "20260309_Tidy_MBL_CLL_metadata_gex.tsv")) %>%
  dplyr::mutate(MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype))
disgenet_MBL_results <- readRDS(file = file.path(RDS_dir, "disgenet_MBL_results.RDS"))
disgenet_CLL_results <- readRDS(file = file.path(RDS_dir, "disgenet_CLL_results.RDS"))
disgenet_genes <- c(disgenet_MBL_results@qresult$gene_symbol, disgenet_CLL_results@qresult$gene_symbol) %>% unique()
B_markers <- c("MS4A1", "EBF1", "PAX5", "MME", "CD19")
CLL_markers <- c("CD19", "CD5", "FCER2", "CD79b")
our_genes_with_CLL_association <- read_csv(file.path(tsv_dir, "20251015_MBL_progression_genes.csv")) %>% 
  dplyr::filter(`CLL related`==TRUE) %>% 
  pull(gene)
```

## Add Seurat metadata

```{r}
sequencing_annot <- tibble("label_cellbarcode" = WhichCells(gex_scaled_all)) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  dplyr::left_join(metadata_gex %>% dplyr::select(label, sequencing_date) %>% unique()) %>%
  dplyr::select(label_cellbarcode, sequencing_date) %>%
  dplyr::mutate(sequencing_date = as.factor(as.character(sequencing_date))) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = sequencing_annot, col.name = "sequencing_date")

LC_celltype_annot <- tibble(
  "label_cellbarcode" = WhichCells(gex_scaled_all),
  "My_annotation_B" = gex_scaled_all$My_annotation_B
) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  dplyr::left_join(metadata_gex %>%
    dplyr::select(label, MBL_subtype) %>% unique()) %>%
  dplyr::mutate(My_annotation_LC = case_when(My_annotation_B == "MBL B Cell" & MBL_subtype == "HC" ~ "HC-MBL B Cell",
    My_annotation_B == "MBL B Cell" & MBL_subtype == "LC" ~ "LC-MBL B Cell",
    .default = My_annotation_B
  )) %>%
  dplyr::select(label_cellbarcode, My_annotation_LC) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = LC_celltype_annot, col.name = "My_annotation_LC")
```

# Analysis

## Pick samples

```{r}
cohort_overview <- metadata_gex %>% 
dplyr::filter(!(label %in% c("MBL7dup", "CLL6.1", "CLL6.2"))) %>%
  dplyr::rename("date_of_death" = death) %>%
  dplyr::select(patient_id, plotting_id, disease, MBL_CLL_subtype, date_of_birth, sample_date, date_of_MBL_diagnosis, date_of_CLL_diagnosis, date_of_death, last_status, sequencing_date) %>%
  unique() %>%
  dplyr::mutate(
    "date_of_MBL_diagnosis" = if_else(condition = disease == "MBL" & sample_date < date_of_MBL_diagnosis, true = sample_date, false = date_of_MBL_diagnosis)) %>% # if MBL sample was taken before official diagnosis, sample date is set as diagnosis date
  dplyr::mutate(
    "MBL diagnosis" = lubridate::time_length(difftime(date_of_MBL_diagnosis, date_of_MBL_diagnosis), "years"),
    "CLL diagnosis" = lubridate::time_length(difftime(date_of_CLL_diagnosis, date_of_MBL_diagnosis), "years"),
    "last check-up" = lubridate::time_length(difftime(last_status, date_of_MBL_diagnosis), "years"),
    "decease" = lubridate::time_length(difftime(date_of_death, date_of_MBL_diagnosis), "years")) %>%
  pivot_wider(names_from = disease, values_from = sample_date) %>%
  dplyr::rename(
    "sample_date_MBL" = MBL,
    "sample_date_CLL" = CLL) %>%
  dplyr::mutate(
    "MBL sample" = lubridate::time_length(difftime(sample_date_MBL, date_of_MBL_diagnosis), "years"),
    "CLL sample" = lubridate::time_length(difftime(sample_date_CLL, date_of_MBL_diagnosis), "years")) %>% 
  dplyr::filter(MBL_CLL_subtype=="HC",
                #`last check-up` >= 5
                ) %>% 
  dplyr::mutate(five_year_MBL_stability=case_when(`CLL diagnosis` <= 5 ~ "progressive", 
                                                  is.na(`CLL diagnosis`)~ "unprogressive",
                                                  .default="progressive")) %>% 
  dplyr::select(-c(date_of_birth, date_of_CLL_diagnosis, date_of_MBL_diagnosis, date_of_death, last_status, sample_date_MBL, sample_date_CLL)) %>% 
  dplyr::group_by(five_year_MBL_stability) %>% 
  dplyr::arrange(five_year_MBL_stability, (`CLL diagnosis`))
# select equal number of samples of good sample quality with either longest persistence of MBL or quickest progression to CLL
write_tsv(cohort_overview, file.path(tsv_dir, "20251017_MBL_progression_cohort_overview.tsv"))

cohort_overview_gex <- cohort_overview %>% 
  dplyr::filter(!(patient_id %in% c("CLL2", "CLL5")))
```

## Swimmersplot of progression

```{r}
swimmersplot_input <- cohort_overview %>% 
  dplyr::select(-c(MBL_CLL_subtype, sequencing_date, decease)) %>% 
  dplyr::mutate(MBL=if_else(!is.na(`CLL diagnosis`),`CLL diagnosis`-`MBL diagnosis`,`last check-up`-`MBL diagnosis`),
                CLL=if_else(!is.na(`CLL diagnosis`), `last check-up`- `CLL diagnosis`, NA),
                plotting_id=factor(plotting_id, levels=paste0("P", 1:24))) %>% 
  dplyr::select(-c(`last check-up`,`CLL diagnosis`, `MBL diagnosis`)) %>% 
  tidyr::pivot_longer(-c(patient_id, plotting_id, five_year_MBL_stability), names_to = "event", values_to = "time [years]") %>% 
  dplyr::group_by(patient_id, plotting_id)
```

```{r}
diagnosis_segments <- swimmersplot_input %>%
  group_by(patient_id, plotting_id) %>%
  dplyr::summarise(
    MBL_time = max(`time [years]`[event == "MBL"], na.rm = TRUE),
    CLL_time = ifelse(all(is.na(`time [years]`[event == "CLL"])),
                      NA, max(`time [years]`[event == "CLL"], na.rm = TRUE)),
    .groups = "drop") %>%
  mutate(seg_start_1 = 0,
    seg_end_1   = MBL_time,
    seg_start_2 = MBL_time,
    seg_end_2   = CLL_time) %>%
  pivot_longer(cols = starts_with("seg_"),
    names_to = c(".value", "segment"),
    names_pattern = "(seg_[a-z]+)_(\\d)") %>%
  dplyr::mutate(phase = ifelse(segment == "1", "MBL", "CLL"),
                phase=factor(phase, levels=c("MBL", "CLL"))) %>%
  dplyr::filter(!is.na(seg_end)) %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, levels=paste0("P", 1:24))) %>% 
  arrange((plotting_id))

write_tsv(diagnosis_segments, file.path(tsv_dir, "diagnosis_segments_swimmersplot_progression.tsv"))

condition_samples <- swimmersplot_input %>%
  dplyr::filter(event %in%c("MBL sample") & !is.na(`time [years]`)) %>% 
  dplyr::mutate(plotting_id=factor(plotting_id, levels=paste0("P", 1:24))) %>% 
  ungroup() %>% 
  arrange((plotting_id))

write_tsv(condition_samples, file.path(tsv_dir, "condition_samples_swimmersplot_progression.tsv"))

five_year_MBL_stability <- c("unprogressive"="slategray3", "progressive"="lightpink3")

condition <- cohort_overview %>% 
  dplyr::select(plotting_id, five_year_MBL_stability) %>% 
  unique() %>% 
  dplyr::mutate(five_year_MBL_stability=factor(five_year_MBL_stability, levels=c("unprogressive", "progressive")),
                plotting_id=factor(plotting_id, levels=rev(paste0("P", c(1:9, 13,11, 15:18, 19, 12,20, 23, 22, 21, 10))))) %>% 
  ungroup() %>% 
  arrange((plotting_id))

write_tsv(condition, file.path(tsv_dir, "condition_swimmersplot_progression.tsv"))

swimmersplot_progression <- ggplot() +
  geom_tile(data = condition, aes(y =plotting_id, x =7, width = 16, fill=five_year_MBL_stability),  alpha=0.3, inherit.aes = FALSE) +  
  geom_vline(xintercept = c(5), color = "grey", linetype = "dashed") +
  geom_segment(data = diagnosis_segments, aes(x = seg_start, xend = seg_end, y = plotting_id, yend = plotting_id, color = phase), size = 1.5) +
  geom_point(data = condition_samples, aes(x = `time [years]`, y = plotting_id), shape = 21, fill = "#FC8D62", colour="#FC8D62", size = 4, stroke = 0.8) +
  scale_y_discrete(expand = c(0,0))+
  scale_x_continuous(expand = c(0, 0))+
  scale_color_manual(values = my_MBL_CLL_colours_named) +
  scale_fill_manual(name="Progression status", 
                    values=five_year_MBL_stability, 
                    labels=c("unprogressive"="stable", "progressive"="progressive"))+
  labs(x = "Time since MBL diagnosis\n[years]", y = "", color = "Condition") +
  theme_pub() 

swimmersplot_progression
```

  ## Pseudobulk data

```{r Calculate_celltype_condition_numbers}
celltype_condition_counts <- gex_scaled_all@meta.data %>%
  dplyr::select(orig.ident, patient_id, MBL_CLL_subtype, My_annotation_LC) %>%
  dplyr::filter(My_annotation_LC %in% c("HC-MBL B Cell"),
                patient_id %in% cohort_overview_gex$patient_id) %>%
  left_join(cohort_overview_gex %>% dplyr::select(patient_id, five_year_MBL_stability)) %>% 
  dplyr::group_by(patient_id, orig.ident, five_year_MBL_stability) %>%
  summarise(cell_counts = n()) %>%
  dplyr::select(patient_id, orig.ident, five_year_MBL_stability, cell_counts)


five_year_MBL_stability_annot <- tibble(
  "label_cellbarcode" = WhichCells(gex_scaled_all)) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  tidyr::separate(label, c("patient_id", NA), sep="\\.", remove = FALSE) %>% 
  dplyr::left_join(cohort_overview_gex %>%
                     dplyr::select(patient_id, five_year_MBL_stability)) %>%
  dplyr::select(label_cellbarcode, five_year_MBL_stability) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = five_year_MBL_stability_annot, col.name = "five_year_MBL_stability")
```

```{r}
# Filter out MBL7dup
Idents(gex_scaled_all) <- "orig.ident"
gex_scaled_all_oi <- subset(gex_scaled_all, subset = orig.ident %in% celltype_condition_counts$orig.ident & My_annotation_LC %in% c("HC-MBL B Cell"))

Idents(gex_scaled_all_oi) <- "five_year_MBL_stability"
# Extract raw counts and metadata to create SingleCellExperiment object
counts <- gex_scaled_all_oi@assays$RNA@counts

metadata <- gex_scaled_all_oi@meta.data

# Set up metadata as desired for aggregation and DE analysis
metadata$cluster_id <- factor(gex_scaled_all_oi@active.ident)
metadata$orig.ident <- factor(gex_scaled_all_oi$orig.ident)

# Create single cell experiment object
sce <- SingleCellExperiment(
  assays = list(counts = counts),
  colData = metadata
)

# Identify groups for aggregation of counts
groups <- colData(sce)[, c("cluster_id", "orig.ident")]
```


```{r Check_aggregation_info}
kids <- purrr::set_names(levels(sce$cluster_id))

nk <- length(kids)

sids <- purrr::set_names(levels(sce$`orig.ident`))

ns <- length(sids)
```

```{r Create_sample_level_metadata}
table(sce$orig.ident)

n_cells <- as.numeric(table(sce$orig.ident))

## Determine how to reorder the samples (rows) of the metadata to match the order of sample names in sids vector
m <- match(sids, sce$orig.ident)

## Create the sample level metadata by combining the reordered metadata with the number of cells corresponding to each sample.
ei_sample_res <- data.frame(colData(sce)[m, ],
  n_cells,
  row.names = NULL
) %>%
  dplyr::select(-"cluster_id")

ei <- celltype_condition_counts %>%
  dplyr::mutate(orig.ident = factor(orig.ident, levels = sids)) %>%
  dplyr::left_join(ei_sample_res %>% 
                     dplyr::select(-c(five_year_MBL_stability, My_annotation_LC, My_annotation_B, My_annotation)), by = c("patient_id", "orig.ident")) %>%
  dplyr::relocate(patient_id:orig.ident, .before = MBL_CLL_subtype) %>%
  dplyr::mutate(orig.ident = factor(orig.ident, levels = sids)) %>%
  dplyr::rename("celltype_cell_counts" = cell_counts)
```

```{r Count_aggregation_to_sample_leve}
# Aggregate the counts per sample_id and cluster_id

# Subset metadata to only include the cluster and sample IDs to aggregate across
groups <- colData(sce)[, c("cluster_id", "orig.ident")]

# Aggregate across cluster-sample groups
pb <- aggregate.Matrix(t(counts(sce)),
  groupings = groups, fun = "sum"
)

class(pb)

dim(pb)

pb[1:6, 1:6]

# splitf <- sapply(stringr::str_split(rownames(pb), pattern = "_", n = 2),
#                 `[`, 1)

# Turn into a list and split the list into components for each cluster/celltype and transform, so rows are genes and columns are samples and make rownames as the sample IDs
pb <- t(pb)


# Explore the different components of list
str(pb)

# Print out the table of cells in each cluster-sample group
options(width = 100)
table(sce$cluster_id, sce$orig.ident)
```

```{r}
# Get sample names for each of the cell type clusters

# prep. data.frame for plotting
de_samples <- colnames(pb)
```

```{r}
# Get cluster IDs for each of the samples
samples_list <- colnames(pb)
de_orig_idents <- colnames(pb) %>% str_split_i(pattern = "_", i = 2)
de_cluster_ids <- colnames(pb) %>% str_split_i(pattern="_", i=1)
```

```{r}
# Create a data frame with the sample IDs, cluster IDs and condition
gg_df <- data.frame(
  cluster_id = as.factor(de_cluster_ids),
  orig.ident = as.factor(de_orig_idents),
  sample_id = as.factor(de_samples)
)

gg_df <- left_join(gg_df, ei[, c("orig.ident", "sequencing_date", "five_year_MBL_stability", "celltype_cell_counts")], by = join_by(orig.ident, cluster_id == five_year_MBL_stability))

metadata <- gg_df %>%
  dplyr::select(cluster_id, sample_id, sequencing_date, celltype_cell_counts) %>%
  dplyr::mutate(celltype_cell_counts = log(celltype_cell_counts)) %>%
  dplyr::mutate(cluster_id = factor(cluster_id, levels = c("unprogressive", "progressive")))
```

```{r Subsetting_dataset_to_celltypes_oi}
# Generate vector of cluster IDs
clusters <- levels(metadata$cluster_id)
clusters
```


## Run DESeq2

```{r}
# Create matrix for all comparisons
combinations_ordered <- tibble(
  "Downregulated" = c(
    "unprogressive"),
  "Upregulated" = c(
    "progressive")
)

create.deseq <- function(condition_1, condition_2, complete_count_data, meta) {
  cl_metadata <- meta[which(meta$cluster_id %in% c(condition_1, condition_2)), ]
  rownames(cl_metadata) <- cl_metadata$sample_id

  count_data <- complete_count_data[, cl_metadata$sample_id]

  dds <- DESeqDataSetFromMatrix(count_data,
      colData = cl_metadata,
      design = ~ sequencing_date + cluster_id)
  
  return(dds)
}

dds_list <- list()
for (n in 1:nrow(combinations_ordered)) {
  dds_list[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])] <- create.deseq(
    condition_1 = combinations_ordered$Downregulated[n],
    condition_2 = combinations_ordered$Upregulated[n],
    complete_count_data = pb,
    meta = metadata
  )
}
```

```{r PC}
# Transform counts for data visualization
rld <- dds_list %>% 
  map(~ rlog(., blind = TRUE))

# Plot PCA
rld %>%
  map(~ DESeq2::plotPCA(., intgroup = "cluster_id")) %>%
  ggarrange(plotlist = ., nrow = 1, ncol = 1)

rld %>%
  map(~ DESeq2::plotPCA(., intgroup = "sequencing_date")) %>%
  ggarrange(plotlist = ., nrow = 1, ncol = 1)
```

```{r Hierarchical_Clustering, fig.height=12, fig.width=12}
# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- rld %>% map(~ assay(.))
rld_cor <- rld_mat %>% map(~ cor(.))

annotation_colours_seqdate <- c("2022-02-28" = "#E6AB02", "2022-11-18" = "#7FC97F", "2023-05-02" = "#FFFF99", "2024-03-18" = "#386CB0") %>% as.factor()

# Plot heatmap
for (n in 1:nrow(combinations_ordered)) {
  print(n)
  cluster_metadata <- metadata[which(metadata$cluster_id %in% c(combinations_ordered$Downregulated[n], combinations_ordered$Upregulated[n])), ] %>%
    dplyr::mutate(cluster_id = as.character(cluster_id)) %>%
    tidyr::separate(cluster_id, into = c("cluster_id", NA), sep = "_")
  rownames(cluster_metadata) <- cluster_metadata$sample_id

  rld_cor[[n]] %>% pheatmap(
    annotation = cluster_metadata[, c("cluster_id", "sequencing_date"), drop = F],
    annotation_colors = list(
      cluster_id = annotation_colours_named[c(1, 10, 11, 12)],
      sequencing_date = annotation_colours_seqdate
    )
  )
}
```

```{r Run_DESeq}
dds_result <- dds_list %>% 
  map(~ DESeq(.))
# saveRDS(dds_result, file.path(RDS_dir, "dds_result_MBL_progression.RDS"))
# dds_result <- readRDS(file.path(RDS_dir, "dds_result_MBL_progression.RDS"))
```

```{r fig.width=6, fig.height=6}
# Plot dispersion estimates
dds_dispersionplots <- dds_result %>% 
  map(~ plotDispEsts(.))
```
```{r}
# Output results of Wald test 
res_lst <- list()
for (n in 1:nrow(combinations_ordered)) {
  print(n)
  cluster_metadata <- metadata[which(metadata$cluster_id %in% c(combinations_ordered$Downregulated[n], combinations_ordered$Upregulated[n])), ]
  rownames(cluster_metadata) <- cluster_metadata$sample_id

  contrast <- c("cluster_id", combinations_ordered$Upregulated[n], combinations_ordered$Downregulated[n])

  resultsNames(dds_result[[n]])
  res_lst[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]] <- results(dds_result[[n]],
    contrast = contrast,
    alpha = 0.05
  )

  res_lst[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]] <- lfcShrink(
    dds = dds_result[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]],
    contrast = contrast, type = "ashr",
    res = res_lst[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]]
  )
}
```

```{r}
res_tbl <- res_lst %>% 
  map(~ data.frame(.) %>%
  rownames_to_column(var = "gene") %>%
  as_tibble())

# saveRDS(res_tbl, file = file.path(RDS_dir, "res_tbl_MBL_progression.RDS"))
# res_tbl <- readRDS(file = file.path(RDS_dir, "res_tbl_MBL_progression.RDS"))
```

## Volcanoplots

### All genes -log10(padj) >= 1 labelled
```{r Volcanoplot, fig.height=6, fig.width=6}
five_year_MBL_stability_colours <- c("unprogressive HC-MBL"="slategray3","progressive HC-MBL"="lightpink3" )

res_table_thres <- res_tbl %>% 
  map(~mutate(., threshold = if_else(-log10(padj) >= 10 & abs(log2FoldChange) >= 1, TRUE, NA),
              
              threshold_1 = case_when(-log10(padj) >= 1 & log2FoldChange >= 1~ "progressive HC-MBL", 
                                      -log10(padj) >= 1 & log2FoldChange <=- 1 ~ "unprogressive HC-MBL", 
                                      .default=NA),
              threshold_1=factor(threshold_1, levels=c("unprogressive HC-MBL", "progressive HC-MBL"))) %>%
            dplyr::arrange(desc(threshold_1), padj) %>%
    dplyr::mutate(label = if_else(!is.na(threshold_1) , gene, NA
    )))

progression_volcano <- res_table_thres$unprogressive_vs_progressive %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj), colour = threshold_1, label = label)) +
      geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
      geom_hline(yintercept = c(1), color = "grey", linetype = "dashed") +
      geom_point(size = 1) +
      geom_text_repel(
      size = 2.5, box.padding = 0.5,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = Inf,
      force_pull = 150,
      force = 10,
      colour="black")+      
      labs(y = "-log10(FDR)", x = "log2(Fold change)") +
      scale_y_continuous(limits=c(0,10), expand = expansion(mult = c(0, 0.01))) +
      scale_x_continuous(limits=c(-30, 30), expand = expansion(mult = c(0, 0))) +
      scale_colour_manual(name = "Differentially expressed genes", values = five_year_MBL_stability_colours, na.value = "lightgrey") +
      theme_pub()+ 
      theme(legend.position = "bottom", legend.direction = "horizontal") 

progression_volcano
write_tsv(progression_volcano$data, file.path(tsv_dir, "volcanoplot_HC_MBL_progression_data.tsv"))
```

### Selected genes labelled without MBL9
```{r , fig.height=4, fig.width=4}
res_table_thres <- res_table_thres %>%  
  map(~mutate(.x, label_selected = if_else(!is.na(threshold_1) & !grepl("^IG", gene) & gene %in% c(disgenet_genes,CLL_markers, B_markers, our_genes_with_CLL_association) , gene, NA
    )))

progression_volcano_selected <- res_table_thres$unprogressive_vs_progressive %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj), colour = threshold_1, label = label_selected)) +
      geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
      geom_hline(yintercept = c(1), color = "grey", linetype = "dashed") +
      geom_point(size = 1) +
      geom_text_repel(
      size = 4, 
      box.padding = 0.5,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = Inf,
      force = 50,
      colour="black")+      
      labs(title = "Unprogressive vs. progessive HC-MBLs", y = "-log10(FDR)", x = "log2(Fold change)") +
      scale_y_continuous(limits=c(0,15), expand = expansion(mult = c(0, 0.01))) +
      scale_x_continuous(limits=c(-25, 25), expand = expansion(mult = c(0, 0))) +
      scale_colour_manual(name = "Differentially expressed genes", values = five_year_MBL_stability_colours, na.value = "lightgrey") +
      theme_pub()+ 
      theme(legend.position = "bottom", legend.direction = "horizontal", plot.title = element_blank()) 

progression_volcano_selected
```



```{r, fig.width=10, fig.height=5}
ggarrange(swimmersplot_progression, progression_volcano,
          nrow=1, ncol=2, labels = "AUTO")
```


