---
title: "Differential Gene Pseudobulk Expression analysis"
author: "Anja Rathgeber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 4
email: anja.rathgeber@mdc-berlin.de
---

# GOAL
The aim of this script is to find out if there are differentially expressed genes between LC-MBL, HC-MBL and CLL cases across patients in comparison to physiologic B cells. We use pseudobulk analysis here adapted from https://hbctraining.github.io/scRNA-seq/lessons/pseudobulk_DESeq2_scrnaseq.html 

# Load libraries

```{r Load_libraries}
library(DESeq2)
library(magrittr)
library(Matrix.utils)
library(readxl)
library(tidyverse)
library(ggpubr)
library(ggrepel)
library(Cairo)
library(Seurat)
library(SeuratDisk)
library(RColorBrewer)
library("org.Hs.eg.db")
library(pheatmap)
library(circlize)
library(SingleCellExperiment)
library(ggvenn)
library(disgenet2r)
library(ComplexHeatmap)
```

# Set paths

```{r}
source("config_paths.R")
data_dir <- file.path(project_dir, "data")
result_dir <- file.path(project_dir, "results")
analysis_dir <- file.path(result_dir, "DESeq2_analysis")
plotting_dir <- file.path(analysis_dir, "Plots")
RDS_dir <- file.path(analysis_dir, "RDS_Objects")
tsv_dir <- file.path(analysis_dir, "tsv_data")
GEX_RDS_dir <- file.path(result_dir, "GEX_Analysis/RDS_Objects")
```

# Load data
```{r}
source(file.path(project_dir, "scripts/publication_theme.R"))
source(file.path(project_dir, "scripts/heatmap_plotting_functions.R"))
```

```{r}
gex_scaled_all <- readRDS(file.path(GEX_RDS_dir, "gex_scaled_all.RDS"))
metadata_all <- read_tsv(file = file.path(data_dir, "20260309_Tidy_MBL_CLL_metadata.tsv")) %>%
  dplyr::mutate(MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype))
gex_data_scvi <- readRDS(file = file.path(project_dir_ext, "scVI\ expression\ data/scvi_expression.RDS"))
c2_data <- read_tsv(file.path(c2_data_dir, "c2.all.v2024.1.Hs.symbols.gmt.txt"), col_names = FALSE) %>%
  dplyr::filter(str_starts(
    X1,
    "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR|REACTOME_CD22_MEDIATED_BCR_REGULATION|REACTOME_FCERI_MEDIATED_CA_2_MOBILIZATION|REACTOME_FCERI_MEDIATED_MAPK_ACTIVATION|REACTOME_FCERI_MEDIATED_NF_KB_ACTIVATION|REACTOME_ANTIGEN_ACTIVATES_B_CELL_RECEPTOR_BCR_LEADING_TO_GENERATION_OF_SECOND_MESSENGERS"
  ))
```

## Add Seurat metadata

```{r}
sequencing_annot <- tibble("label_cellbarcode" = WhichCells(gex_scaled_all)) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  dplyr::left_join(metadata_all %>% dplyr::filter(data_type == "GEX") %>% dplyr::select(label, sequencing_date) %>% unique()) %>%
  dplyr::select(label_cellbarcode, sequencing_date) %>%
  dplyr::mutate(sequencing_date = as.factor(as.character(sequencing_date))) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = sequencing_annot, col.name = "sequencing_date")

LC_celltype_annot <- tibble(
  "label_cellbarcode" = WhichCells(gex_scaled_all),
  "My_annotation_B" = gex_scaled_all$My_annotation_B
) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  dplyr::left_join(metadata_all %>% dplyr::filter(data_type == "GEX") %>%
    dplyr::select(label, MBL_subtype) %>% unique()) %>%
  dplyr::mutate(My_annotation_LC = case_when(My_annotation_B == "MBL B Cell" & MBL_subtype == "HC" ~ "HC-MBL B Cell",
    My_annotation_B == "MBL B Cell" & MBL_subtype == "LC" ~ "LC-MBL B Cell",
    .default = My_annotation_B
  )) %>%
  dplyr::select(label_cellbarcode, My_annotation_LC) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = LC_celltype_annot, col.name = "My_annotation_LC")

celltype_subtype_annot <- tibble(
  "label_cellbarcode" = WhichCells(gex_scaled_all),
  "My_annotation_LC" = gex_scaled_all$My_annotation_LC
) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  dplyr::left_join(metadata_all %>% dplyr::filter(data_type == "GEX") %>%
    dplyr::select(label, MBL_CLL_subtype) %>% unique()) %>%
  dplyr::mutate(celltype_subtype = paste0(My_annotation_LC, "_", MBL_CLL_subtype)) %>%
  dplyr::mutate(celltype_subtype = if_else(celltype_subtype %in% c("B_HC", "B_CLL"), "B_HCCLL", celltype_subtype)) %>%
  dplyr::select(label_cellbarcode, celltype_subtype) %>%
  deframe()

gex_scaled_all <- AddMetaData(object = gex_scaled_all, metadata = celltype_subtype_annot, col.name = "celltype_subtype")
```

# Analysis

```{r Calculate_celltype_condition_numbers}
celltype_condition_counts <- gex_scaled_all@meta.data %>%
  dplyr::select(orig.ident, patient_id, MBL_CLL_subtype, My_annotation_LC, celltype_subtype) %>%
  dplyr::filter(
    My_annotation_LC %in% c("B", "LC-MBL B Cell", "HC-MBL B Cell", "CLL B Cell"),
    patient_id != "MBL7dup"
  ) %>%
  dplyr::group_by(patient_id, orig.ident, MBL_CLL_subtype, My_annotation_LC, celltype_subtype) %>%
  summarise(cell_counts = n()) %>%
  dplyr::select(patient_id, orig.ident, celltype_subtype, cell_counts)
```

```{r}
# Filter out MBL7dup
Idents(gex_scaled_all) <- "orig.ident"
gex_scaled_all_oi <- subset(gex_scaled_all, subset = orig.ident != "MBL7dup" & My_annotation_LC %in% c("B", "LC-MBL B Cell", "HC-MBL B Cell", "CLL B Cell"))

Idents(gex_scaled_all_oi) <- "celltype_subtype"
# Extract raw counts and metadata to create SingleCellExperiment object
counts <- gex_scaled_all_oi@assays$RNA@counts

metadata <- gex_scaled_all_oi@meta.data

# Set up metadata as desired for aggregation and DE analysis
metadata$cluster_id <- factor(gex_scaled_all_oi@active.ident)
metadata$orig.ident <- factor(gex_scaled_all_oi$orig.ident)

# Create single cell experiment object
sce <- SingleCellExperiment(
  assays = list(counts = counts),
  colData = metadata
)

# Identify groups for aggregation of counts
groups <- colData(sce)[, c("cluster_id", "orig.ident")]
```

```{r Check_aggregation_info}
kids <- purrr::set_names(levels(sce$cluster_id))

nk <- length(kids)

sids <- purrr::set_names(levels(sce$`orig.ident`))

ns <- length(sids)
```

```{r Create_sample_level_metadata}
table(sce$orig.ident)

n_cells <- as.numeric(table(sce$orig.ident))

## Determine how to reoder the samples (rows) of the metadata to match the order of sample names in sids vector
m <- match(sids, sce$orig.ident)

## Create the sample level metadata by combining the reordered metadata with the number of cells corresponding to each sample.
ei_sample_res <- data.frame(colData(sce)[m, ],
  n_cells,
  row.names = NULL
) %>%
  dplyr::select(-"cluster_id")

ei <- celltype_condition_counts %>%
  dplyr::mutate(orig.ident = factor(orig.ident, levels = sids)) %>%
  dplyr::left_join(ei_sample_res %>% dplyr::select(-c(celltype_subtype, My_annotation_LC, My_annotation_B, My_annotation)), by = c("patient_id", "orig.ident", "MBL_CLL_subtype")) %>%
  dplyr::relocate(patient_id:orig.ident, .before = MBL_CLL_subtype) %>%
  dplyr::mutate(orig.ident = factor(orig.ident, levels = sids)) %>%
  dplyr::rename("celltype_cell_counts" = cell_counts)
```

```{r Count_aggregation_to_sample_leve}
# Aggregate the counts per sample_id and cluster_id

# Subset metadata to only include the cluster and sample IDs to aggregate across
groups <- colData(sce)[, c("cluster_id", "orig.ident")]

# Aggregate across cluster-sample groups
pb <- aggregate.Matrix(t(counts(sce)),
  groupings = groups, fun = "sum"
)

class(pb)

dim(pb)

pb[1:6, 1:6]

# splitf <- sapply(stringr::str_split(rownames(pb), pattern = "_", n = 2),
#                 `[`, 1)

# Turn into a list and split the list into components for each cluster/celltype and transform, so rows are genes and columns are samples and make rownames as the sample IDs
pb <- t(pb)


# Explore the different components of list
str(pb)

# Print out the table of cells in each cluster-sample group
options(width = 100)
table(sce$cluster_id, sce$orig.ident)
```

```{r}
# Get sample names for each of the cell type clusters

# prep. data.frame for plotting
de_samples <- colnames(pb)
```

```{r}
# Get cluster IDs for each of the samples
samples_list <- colnames(pb)
de_orig_idents <- colnames(pb) %>% str_split_i(pattern = "_", i = 3)
de_cluster_ids <- colnames(pb) %>% str_extract("^[^_]*_[^_]*")
```

```{r}
# Create a data frame with the sample IDs, cluster IDs and condition
gg_df <- data.frame(
  cluster_id = as.factor(de_cluster_ids),
  orig.ident = as.factor(de_orig_idents),
  sample_id = as.factor(de_samples)
)

gg_df <- left_join(gg_df, ei[, c("orig.ident", "sequencing_date", "celltype_subtype", "celltype_cell_counts")], by = join_by(orig.ident, cluster_id == celltype_subtype))


metadata <- gg_df %>%
  dplyr::select(cluster_id, sample_id, sequencing_date, celltype_cell_counts) %>%
  dplyr::mutate(celltype_cell_counts = log(celltype_cell_counts)) %>%
  dplyr::mutate(cluster_id = factor(cluster_id, levels = c("B_LC", "B_HCCLL", "LC-MBL B Cell_LC", "HC-MBL B Cell_HC", "CLL B Cell_CLL")))
```

```{r Subsetting_dataset_to_celltypes_oi}
# Generate vector of cluster IDs
clusters <- levels(metadata$cluster_id)
clusters
```

```{r}
# Subset the metadata to only the B like cells
# cluster_metadata <- metadata[which(metadata$cluster_id == clusters[c(1,2,4,5)]), ]
# head(cluster_metadata)
#
# # Assign the rownames of the metadata to be the sample IDs
# rownames(cluster_metadata) <- cluster_metadata$sample_id
# head(cluster_metadata)
#
# # Subset the counts to only the B cells
# counts <- pb[[clusters[1]]]
#
# cluster_counts <- data.frame(counts[, which(colnames(counts) %in% rownames(cluster_metadata))])
#
# # Check that all of the row names of the metadata are the same and in the same order as the column names of the counts in order to use as input to DESeq2
# all(rownames(cluster_metadata) == colnames(cluster_counts))
```

```{r}
# Create matrix for all comparisons
combinations_ordered <- tibble(
  "Downregulated" = c(
    "B_LC", "B_HCCLL", "B_HCCLL",
    "HC-MBL B Cell_HC" # , "LC-MBL B Cell_LC"
  ),
  "Upregulated" = c(
    "LC-MBL B Cell_LC", "HC-MBL B Cell_HC", "CLL B Cell_CLL",
    "CLL B Cell_CLL" # , "HC-MBL B Cell_HC"
  )
)

create.deseq <- function(condition_1, condition_2, complete_count_data, meta) {
  cl_metadata <- meta[which(meta$cluster_id %in% c(condition_1, condition_2)), ]
  rownames(cl_metadata) <- cl_metadata$sample_id

  count_data <- complete_count_data[, cl_metadata$sample_id]

  if (condition_1 == "B_LC" & condition_2 == "LC-MBL B Cell_LC") {
    dds <- DESeqDataSetFromMatrix(count_data,
      colData = cl_metadata,
      design = ~cluster_id
    )
  } else {
    dds <- DESeqDataSetFromMatrix(count_data,
      colData = cl_metadata,
      design = ~ sequencing_date + cluster_id
    )
  }
  return(dds)
}

dds_list <- list()
for (n in 1:nrow(combinations_ordered)) {
  dds_list[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])] <- create.deseq(
    condition_1 = combinations_ordered$Downregulated[n],
    condition_2 = combinations_ordered$Upregulated[n],
    complete_count_data = pb,
    meta = metadata
  )
}
```

```{r PC, fig.width=20, fig.height=12}
# Transform counts for data visualization
rld <- dds_list %>% 
  map(~ rlog(., blind = TRUE))

# Plot PCA
rld %>%
  map(~ DESeq2::plotPCA(., intgroup = "cluster_id")) %>%
  ggarrange(plotlist = ., nrow = 2, ncol = 3)

rld %>%
  map(~ DESeq2::plotPCA(., intgroup = "sequencing_date")) %>%
  ggarrange(plotlist = ., nrow = 2, ncol = 3)
```

```{r Hierarchical_Clustering, fig.height=12, fig.width=12}
# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- rld %>% map(~ assay(.))
rld_cor <- rld_mat %>% map(~ cor(.))

annotation_colours_seqdate <- c("2022-02-28" = "#E6AB02", "2022-11-18" = "#7FC97F", "2023-05-02" = "#FFFF99", "2024-03-18" = "#386CB0") %>% as.factor()

# Plot heatmap
for (n in 1:nrow(combinations_ordered)) {
  print(n)
  cluster_metadata <- metadata[which(metadata$cluster_id %in% c(combinations_ordered$Downregulated[n], combinations_ordered$Upregulated[n])), ] %>%
    dplyr::mutate(cluster_id = as.character(cluster_id)) %>%
    tidyr::separate(cluster_id, into = c("cluster_id", NA), sep = "_")
  rownames(cluster_metadata) <- cluster_metadata$sample_id

  rld_cor[[n]] %>% pheatmap(
    annotation = cluster_metadata[, c("cluster_id", "sequencing_date"), drop = F],
    annotation_colors = list(
      cluster_id = annotation_colours_named[c(1, 10, 11, 12)],
      sequencing_date = annotation_colours_seqdate
    )
  )
}
```

```{r Run_DESeq}
dds_result <- dds_list %>% 
  map(~ DESeq(.))
# saveRDS(dds_result, file.path(RDS_dir, "dds_result.RDS"))
dds_result <- readRDS(file.path(RDS_dir, "dds_result.RDS"))
```

```{r fig.width=6, fig.height=6}
# Plot dispersion estimates
dds_dispersionplots <- dds_result %>% 
  map(~ plotDispEsts(.))
```

```{r}
# Output results of Wald test for contrast for stim vs ctrl
res_lst <- list()
for (n in 1:nrow(combinations_ordered)) {
  print(n)
  cluster_metadata <- metadata[which(metadata$cluster_id %in% c(combinations_ordered$Downregulated[n], combinations_ordered$Upregulated[n])), ]
  rownames(cluster_metadata) <- cluster_metadata$sample_id

  contrast <- c("cluster_id", combinations_ordered$Upregulated[n], combinations_ordered$Downregulated[n])

  resultsNames(dds_result[[n]])
  res_lst[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]] <- results(dds_result[[n]],
    contrast = contrast,
    alpha = 0.05
  )

  res_lst[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]] <- lfcShrink(
    dds = dds_result[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]],
    contrast = contrast, type = "ashr",
    res = res_lst[[paste0(combinations_ordered$Downregulated[n], "_vs_", combinations_ordered$Upregulated[n])]]
  )
}
```

```{r}
res_tbl <- res_lst %>% 
  map(~ data.frame(.) %>%
  rownames_to_column(var = "gene") %>%
  as_tibble())

# saveRDS(res_tbl, file = file.path(RDS_dir, "res_tbl.RDS"))
res_tbl <- readRDS(file = file.path(RDS_dir, "res_tbl.RDS"))
```

### Volcanoplots

```{r Volcanoplot, fig.width=15, fig.height=5}
res_table_thres <- res_tbl %>% 
  map(~mutate(., threshold = if_else(-log10(padj) >= 10 & abs(log2FoldChange) >= 1, TRUE, NA)) %>%
    dplyr::arrange(desc(threshold), padj) %>%
    dplyr::mutate(label = if_else(threshold == TRUE & row_number() <= 20 & !grepl("^IG", gene),
      gene, NA
    )))

res_table_thres %>%
  imap(~
    ggplot(.x, aes(x = log2FoldChange, y = -log10(padj), colour = threshold, label = label)) +
      geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
      geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
      geom_point(size = 0.5) +
      geom_label_repel(
      size = 2.5,
      label.size = NA,
      label.padding = 0.1,
      segment.size = 0.25,
      segment.colour = "grey",
      min.segment.length = 0.1,
      max.overlaps = 40, colour="black")+      
      labs(title = .y, y = "-log10(FDR)", x = "log2(Fold change)") +
      scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
      scale_x_continuous(limits = c(-20, 20), expand = expansion(mult = c(0, 0.01))) +
      scale_colour_manual(name = "Significance", values = c("TRUE" = "darkred", "FALSE" = "lightgrey"), na.value = "lightgrey") +
      theme(legend.position = "bottom") +
      theme_pub()) %>%
  ggarrange(plotlist = ., nrow = 1, ncol = 4, common.legend = TRUE, legend = "bottom")
```

### Venn Diagram between conditions

```{r fig.width=6, fig.height=6}
venn_input <- res_table_thres[1:3] %>%
  map(~ dplyr::filter(., threshold == "TRUE") %>%
    dplyr::mutate(gene = if_else(log2FoldChange >= 0, gene, paste0("-", gene))) %>%
    pull(gene))
names(venn_input) <- c("LC-MBL vs. B", "HC-MBL vs. B", "CLL vs. B")

ggvenn(
  data = venn_input,
  fill_color = c("#fcbda4", "#FC8D62", "#66C2A5"),
  stroke_size = 0.5,
  set_name_size = 4,
)
```

### Highlight Disgenet genes

```{r}
# register at disgnet and retrive api key https://disgenet.com/Profile-area#apiKey

# api_key <- "a1df2b60-8ac8-451c-9454-1ab487e56606"
# Sys.setenv(DISGENET_API_KEY= api_key)
# disgenet_CLL_results <- disease2gene( disease  = "UMLS_C0023434",
#                            database = "CURATED",
#                            score    = c( 0.1,1 ), n_pags=100 )
#
# saveRDS(disgenet_CLL_results, file = file.path(RDS_dir, "disgenet_CLL_results.RDS"))
disgenet_CLL_results <- readRDS(file = file.path(RDS_dir, "disgenet_CLL_results.RDS"))

# disgenet_MBL_results <- disease2gene( disease  = "UMLS_C2698259",
#                           database = "CURATED",
#                           score    = c( 0.1,1 ), n_pags=100 )

# saveRDS(disgenet_MBL_results, file = file.path(RDS_dir, "disgenet_MBL_results.RDS"))
disgenet_MBL_results <- readRDS(file = file.path(RDS_dir, "disgenet_MBL_results.RDS"))

genes2label <- c(disgenet_CLL_results@qresult[["gene_symbol"]], disgenet_MBL_results@qresult[["gene_symbol"]])
```

disgenet CLL/MBL associated genes and top10 significant genes
```{r fig.width=10, fig.height=5}
res_table_thres_disgenet <- res_table_thres %>%
  map(~ mutate(., disgenet = if_else(gene %in% genes2label & threshold == TRUE, gene, NA)) %>%
    dplyr::arrange(desc(disgenet))) %>%
  map(~ arrange(., desc(threshold), padj) %>%
    dplyr::mutate(top10_threshold = if_else(threshold == TRUE & row_number() <= 10,
      gene, NA
    )))

# show disgnet genes associated with CLL or MBL
res_table_thres_disgenet %>%
  imap(~
    ggplot(.x, aes(x = log2FoldChange, y = -log10(padj), colour = threshold, label = disgenet)) +
      geom_point(size = 0.5) +
      geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
      geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
      geom_label_repel(
          size = 2.5,
          label.size = NA,
          label.padding = 0.1,
          segment.size = 0.25,
          segment.colour = "grey",
          min.segment.length = 0.1,
          max.overlaps = 40, colour="black")+      
      labs(title = .y, y = "-log10(FDR)", x = "log2(Fold change)") +
      scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
      scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
      scale_colour_manual(name = "Significance", values = c("TRUE" = "darkred", "FALSE" = "lightgrey"), na.value = "lightgrey") +
      theme(legend.position = "bottom") +
      theme_pub()) %>%
  ggarrange(plotlist = ., nrow = 1, ncol = 4, common.legend = TRUE, legend = "bottom")

# additionally show top10 significant genes
res_table_thres_disgenet %>%
  map(~ mutate(., disgenet_and_top10 = if_else(gene %in% top10_threshold | !is.na(disgenet), gene, NA))) %>%
  imap(~ ggplot(.x, aes(x = log2FoldChange, y = -log10(padj), colour = threshold, label = disgenet_and_top10)) +
    geom_point(size = 0.5) +
    geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
    geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
    geom_label_repel(
        size = 2.5,
        label.size = NA,
        label.padding = 0.1,
        segment.size = 0.25,
        segment.colour = "grey",
        min.segment.length = 0.1,
        max.overlaps = 40, colour="black")+    
    labs(title = .y, y = "-log10(FDR)", x = "log2(Fold change)") +
    scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
    scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
    scale_colour_manual(name = "Significance", values = c("TRUE" = "darkred", "FALSE" = "lightgrey"), na.value = "lightgrey") +
    theme(legend.position = "bottom") +
    theme_pub()) %>%
  ggarrange(plotlist = ., nrow = 1, ncol = 4, common.legend = TRUE, legend = "bottom")
```

### Highlight Reactome genes in addition

```{r fig.width=10, fig.height=5}
# Reoccuring pathways and genes in disease progression from GSEA of Pseudobulk_GeneSetEnrichmenb_MBL_CLL.Rmd
c2 <- c2_data %>%
  tidyr::pivot_longer(cols = X3:X30, names_to = NULL, values_to = "gene") %>%
  dplyr::rename("gene_set" = X1, "URL" = X2) %>%
  dplyr::select(gene_set, gene) %>%
  separate_rows(gene, sep = "\t")

res_table_reactome <- res_table_thres_disgenet %>%
  map(~ mutate(.,
    reactome = if_else(gene %in% c2$gene & threshold == TRUE, gene, NA),
    disgenet_reactome = if_else((!is.na(disgenet) | !is.na(reactome)) & !grepl("^IG", gene), gene, NA),
    disgenet_reactome_IG = if_else((!is.na(disgenet) | !is.na(reactome)) , gene, NA),
    
  ) %>%
    dplyr::arrange(desc(reactome), desc(disgenet)))

res_table_reactome %>%
  imap(~
    ggplot(.x, aes(x = log2FoldChange, y = -log10(padj), colour = threshold, label = disgenet_reactome)) +
      geom_point(size = 0.5) +
      geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
      geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
      geom_label_repel(
          size = 2.5,
          label.size = NA,
          label.padding = 0.1,
          segment.size = 0.25,
          segment.colour = "grey",
          min.segment.length = 0.1,
          max.overlaps = 40, colour="black")+      
      labs(title = .y, y = "-log10(FDR)", x = "log2(Fold change)") +
      scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
      scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
      scale_colour_manual(name = "Significance", values = c("TRUE" = "darkred", "FALSE" = "lightgrey"), na.value = "lightgrey") +
      theme(legend.position = "bottom") +
      theme_pub()) %>%
  ggarrange(plotlist = ., nrow = 1, ncol = 4, common.legend = TRUE, legend = "bottom")
```


### Highlight B and CLL cell marker genes in addition

```{r fig.width=10, fig.height=5}
B_markers <- c("MS4A1", "EBF1", "PAX5", "MME", "CD19")
CLL_markers <- c("CD19", "CD5", "FCER2", "CD79b")

res_table_marker <- res_table_reactome %>%
  map(~ mutate(.,
    marker_gene = if_else(gene %in% c(B_markers, CLL_markers) & threshold == TRUE, gene, NA),
    disgenet_reactome_marker_gene = if_else((!is.na(disgenet) | !is.na(reactome) | !is.na(marker_gene)) & !grepl("^IG", gene), gene, NA),
    disgenet_reactome_marker_gene_IG = if_else((!is.na(disgenet_reactome_IG) | !is.na(marker_gene)),  gene, NA),
    
  ) %>%
    arrange(desc(reactome), desc(disgenet), desc(marker_gene)))

res_table_marker %>%
  imap(~
    ggplot(.x, aes(x = log2FoldChange, y = -log10(padj), colour = threshold, label = disgenet_reactome_marker_gene)) +
      geom_point(size = 0.5) +
      geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
      geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
      geom_label_repel(
        size = 2.5,
        label.size = NA,
        label.padding = 0.1,
        segment.size = 0.25,
        segment.colour = "grey",
        min.segment.length = 0.1,
        max.overlaps = 40, colour = "black"
      ) +
      labs(title = .y, y = "-log10(FDR)", x = "log2(Fold change)") +
      scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
      scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
      scale_colour_manual(name = "Significance", values = c("TRUE" = "darkred", "FALSE" = "lightgrey"), na.value = "lightgrey") +
      theme(legend.position = "bottom") +
      theme_pub()) %>%
  ggarrange(plotlist = ., nrow = 1, ncol = 4, common.legend = TRUE, legend = "bottom")
```

### Individual volcanoplots correct colouring

```{r}
LC_MBL_pb_volcano <- res_table_marker$`B_LC_vs_LC-MBL B Cell_LC` %>%
  dplyr::mutate(significant = case_when(threshold == TRUE & log2FoldChange > 0 ~ "upregulated",
    threshold == TRUE & log2FoldChange < 0 ~ "downregulated",
    .default = "not significant"
  )) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj), colour = significant, label = disgenet_reactome_marker_gene)) +
  geom_point(size = 0.5) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40, colour = "black"
  ) +
  labs(title = "B vs. LC-MBL", y = "-log10(FDR)", x = "log2(Fold change)") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
  scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
  scale_colour_manual(
    name = "DEGs",
    values = c("upregulated" = "#fcbda4", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "LC-MBL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub()+
  theme(legend.position = "bottom")

saveRDS(LC_MBL_pb_volcano, file.path(RDS_dir, "LC_MBL_pb_volcano.RDS"))


HC_MBL_pb_volcano <- res_table_marker$`B_HCCLL_vs_HC-MBL B Cell_HC` %>%
  dplyr::mutate(significant = case_when(threshold == TRUE & log2FoldChange > 0 ~ "upregulated",
    threshold == TRUE & log2FoldChange < 0 ~ "downregulated",
    .default = "not significant"
  )) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj), colour = significant, label = disgenet_reactome_marker_gene)) +
  geom_point(size = 0.5) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40, colour = "black"
  ) +
  labs(title = "B vs. HC-MBL", y = "-log10(FDR)", x = "log2(Fold change)") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
  scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
  scale_colour_manual(
    name = "DEGs",
    values = c("upregulated" = "#FC8D62", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "HC-MBL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub()+
  theme(legend.position = "bottom")

saveRDS(HC_MBL_pb_volcano, file.path(RDS_dir, "HC_MBL_pb_volcano.RDS"))

CLL_pb_volcano <- res_table_marker$`B_HCCLL_vs_CLL B Cell_CLL` %>%
  dplyr::mutate(significant = case_when(threshold == TRUE & log2FoldChange > 0 ~ "upregulated",
    threshold == TRUE & log2FoldChange < 0 ~ "downregulated",
    .default = "not significant"
  )) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj), colour = significant, label = disgenet_reactome_marker_gene)) +
  geom_point(size = 0.5) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40, colour = "black"
  ) +
  labs(title = "B vs. CLL", y = "-log10(FDR)", x = "log2(Fold change)") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
  scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
  scale_colour_manual(
    name = "DEGs",
    values = c("upregulated" = "#66C2A5", "downregulated" = "darkred", "not significant" = "lightgrey"),
    labels = c("upregulated" = "CLL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub()+
  theme(legend.position = "bottom")

saveRDS(CLL_pb_volcano, file.path(RDS_dir, "CLL_pb_volcano.RDS"))

HC_CLL_pb_volcano <- res_table_marker$`HC-MBL B Cell_HC_vs_CLL B Cell_CLL` %>%
  dplyr::mutate(significant = case_when(threshold == TRUE & log2FoldChange > 0 ~ "upregulated",
    threshold == TRUE & log2FoldChange < 0 ~ "downregulated",
    .default = "not significant"
  )) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj), colour = significant, label = disgenet_reactome_marker_gene)) +
  geom_point(size = 0.5) +
  geom_vline(xintercept = c(-1, 1), color = "grey", linetype = "dashed") +
  geom_hline(yintercept = c(10), color = "grey", linetype = "dashed") +
  geom_label_repel(
    size = 2.5,
    label.size = NA,
    label.padding = 0.1,
    segment.size = 0.25,
    segment.colour = "grey",
    min.segment.length = 0.1,
    max.overlaps = 40, colour = "black"
  ) +
  labs(title = "HC-MBL vs. CLL", y = "-log10(FDR)", x = "log2(Fold change)") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.01))) +
  scale_x_continuous(limits = c(-10, 10), expand = expansion(mult = c(0, 0.01))) +
  scale_colour_manual(
    name = "DEGs",
    values = c("upregulated" = "#66C2A5", "downregulated" = "#FC8D62", "not significant" = "lightgrey"),
    labels = c("upregulated" = "CLL Cells", "downregulated" = "Physiologic B Cells")
  ) +
  theme_pub()+
  theme(legend.position = "bottom")


ggsave(plot = HC_CLL_pb_volcano+theme(legend.position = "none"), 
       filename = file.path(plotting_dir, "HC_CLL_pb_volcano_DEGs.png"),
       width = 2.8, 
       height = 2)

saveRDS(HC_CLL_pb_volcano, file.path(RDS_dir, "HC_CLL_pb_volcano.RDS"))
```

```{r, fig.width=10, fig.height=4}
ggarrange(LC_MBL_pb_volcano, HC_MBL_pb_volcano, CLL_pb_volcano, nrow=1, ncol=3)
```

```{r}
# all significant genes
DEGs_oi <- res_table_thres %>%
  map(~ dplyr::filter(., !is.na(threshold) & !grepl("^IG", gene)) %>%
    pull(gene)) %>%
  unlist() %>%
  unique()

# saveRDS(DEGs_oi, file = file.path(RDS_dir , "DEGs_pseudobulk.RDS"))
DEGs_oi <- readRDS(file.path(RDS_dir, "DEGs_pseudobulk.RDS"))

# all significant genes including immunoglobulinas
DEGs_oi_IG <- res_table_thres %>%
  map(~ dplyr::filter(., !is.na(threshold)) %>%
    pull(gene)) %>%
  unlist() %>%
  unique()

# saveRDS(DEGs_oi_IG, file = file.path(RDS_dir , "DEGs_IG_pseudobulk.RDS"))


# union of all significant disgenet, reactome and marker genes without immunoglobulins
DEGs_disgenet_reactome_marker <- res_table_marker %>%
  map(~ dplyr::filter(., !is.na(threshold) & !grepl("^IG", gene) &!is.na(disgenet_reactome_marker_gene)) %>%
    pull(gene)) %>%
  unlist() %>%
  unique()

saveRDS(DEGs_disgenet_reactome_marker, file = file.path(RDS_dir , "DEGs_disgenet_reactome_marker_pseudobulk.RDS"))

# union of all significant disgenet, reactome and marker genes WITH immunoglobulins
DEGs_disgenet_reactome_marker_IG <- res_table_marker %>%
  map(~ dplyr::filter(., !is.na(threshold) & !is.na(disgenet_reactome_marker_gene_IG)) %>%
    pull(gene)) %>%
  unlist() %>%
  unique()

saveRDS(DEGs_disgenet_reactome_marker_IG, file = file.path(RDS_dir , "DEGs_disgenet_reactome_marker_IG_pseudobulk.RDS"))
```

### Downsampled heatmap
Plot heatmap of downsampled cell number for DEGs on batch corrected expression values
```{r}
celltype_barcode_counts <- gex_scaled_all@meta.data %>%
  dplyr::select(patient_id, My_annotation_LC) %>%
  dplyr::filter(patient_id != "MBL7dup") %>%
  rownames_to_column("label_cellbarcode") %>%
  dplyr::filter(My_annotation_LC %in% c("B", "LC-MBL B Cell", "HC-MBL B Cell", "CLL B Cell")) %>%
  dplyr::mutate(B_like_annotation = case_when(My_annotation_LC == "LC-MBL B Cell" ~ "LC-MBL Cells",
    My_annotation_LC == "HC-MBL B Cell" ~ "HC-MBL Cells",
    My_annotation_LC == "CLL B Cell" ~ "CLL Cells",
    My_annotation_LC == "B" ~ "Physiologic B Cells",
    .default = My_annotation_LC
  )) %>%
  group_by(B_like_annotation, patient_id) %>%
  dplyr::summarise(celltype_count = n())
# 20263 total physiologic B cells across patients

barcodes <- gex_scaled_all@meta.data %>%
  dplyr::select(patient_id, My_annotation_LC) %>%
  dplyr::filter(patient_id != "MBL7dup") %>%
  rownames_to_column("label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  dplyr::filter(My_annotation_LC %in% c("B", "LC-MBL B Cell", "HC-MBL B Cell", "CLL B Cell")) %>%
  dplyr::mutate(B_like_annotation = case_when(My_annotation_LC == "LC-MBL B Cell" ~ "LC-MBL Cells",
    My_annotation_LC == "HC-MBL B Cell" ~ "HC-MBL Cells",
    My_annotation_LC == "CLL B Cell" ~ "CLL Cells",
    My_annotation_LC == "B" ~ "Physiologic B Cells",
    .default = My_annotation_LC
  ))

# max cell count possible c(400000, 32900,21460 ,20263)
celltype_barcodes <- map2(
  c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"), c(5000, 5000, 5000, 5000),
  ~ barcodes %>%
    dplyr::filter(B_like_annotation == .x) %>%
    downsample_cells(max_cells = .y)
) %>%
  set_names(c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))

heatmap_input_b_like_scaled <- t(scale(gex_data_scvi[barcodes$label_cellbarcode, DEGs_disgenet_reactome_marker]))
saveRDS(heatmap_input_b_like_scaled, file.path(RDS_dir, "expression_heatmap_DEGs_batch_corrected_input.RDS"))
```

```{r fig.width=20, fig.height=10}
hm_downsampled <- build_celltype_heatmap(
  bcs_per_celltype = celltype_barcodes,
  data_matrix = heatmap_input_b_like_scaled,
  modality = "GEX",
  include_all_metadata = FALSE,
  show_heatmap_legend = FALSE,
  show_annotation_legend = FALSE,
  absolute_capping = c(-3, 3),
  quantile_capping = FALSE,
  high_score_colour = "darkred",
  mid_score_colour = "white",
  low_score_colour = "midnightblue",
  heatmap_legend_name = "Expression"
)

hm_downsampled$`Physiologic B Cells` + hm_downsampled$`LC-MBL Cells` + hm_downsampled$`HC-MBL Cells` + hm_downsampled$`CLL Cells`

hm_downsampled_grob <- grid.grabExpr(draw(hm_downsampled$`Physiologic B Cells` + hm_downsampled$`LC-MBL Cells` + hm_downsampled$`HC-MBL Cells` + hm_downsampled$`CLL Cells`, heatmap_legend_side = "right"))

cowplot::plot_grid(hm_downsampled_grob)
saveRDS(hm_downsampled_grob, file.path(RDS_dir, "expression_heatmaps_DEGs_batch_corrected.RDS"))
```

#### Clustered heatmap

```{r fig.width=20, fig.height=10}
hm_downsampled_clustered <- build_clustered_heatmap(
  bcs_cells_df = bind_rows(celltype_barcodes),
  data_matrix = heatmap_input_b_like_scaled,
  modality = "GEX",
  include_all_metadata = TRUE,
  show_heatmap_legend = TRUE,
  show_annotation_legend = TRUE,
  absolute_capping = c(-3, 3),
  quantile_capping = FALSE,
  high_score_colour = "darkred",
  mid_score_colour = "white",
  low_score_colour = "midnightblue",
  heatmap_legend_name = "Expression"
)

saveRDS(hm_downsampled_clustered, file = file.path(RDS_dir, "expression_heatmaps_DEGs_batch_corrected_clustered.RDS"))

hm_downsampled_grob_clustered <- grid.grabExpr(draw(hm_downsampled_clustered[[1]]))
#cowplot::plot_grid(hm_downsampled_grob_clustered)
saveRDS(hm_downsampled_grob_clustered, file.path(RDS_dir, "expression_heatmaps_DEGs_batch_corrected_clustered_grob.RDS"))
# do not try to display, session crashes, rather export
```

## DEG Scheme

```{r, fig.width=9, fig.height=4}
WBC_summary <- metadata_all %>%
  dplyr::filter(data_type == "GEX") %>%
  dplyr::select(label, patient_id, `WBC_[cells/nl]`, MBL_subtype) %>%
  dplyr::mutate(MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype)) %>%
  dplyr::select(patient_id, `WBC_[cells/nl]`, MBL_CLL_subtype) %>%
  group_by(MBL_CLL_subtype) %>%
  dplyr::summarise(
    mean_WBC = mean(`WBC_[cells/nl]`),
    median_WBC = median(`WBC_[cells/nl]`)
  )

LC_DEGs <- res_table_thres[["B_LC_vs_LC-MBL B Cell_LC"]] %>%
  dplyr::filter(!is.na(threshold)) %>%
  pull(var = gene)

HC_DEGs <- res_table_thres[["B_HCCLL_vs_HC-MBL B Cell_HC"]] %>%
  bind_rows() %>%
  dplyr::filter(!is.na(threshold)) %>%
  pull(var = gene)

CLL_DEGs <- res_table_thres[["B_HCCLL_vs_CLL B Cell_CLL"]] %>%
  bind_rows() %>%
  dplyr::filter(!is.na(threshold)) %>%
  pull(var = gene)

barcodes_scheme <- rownames(gex_data_scvi) %>%
  as_tibble() %>%
  dplyr::rename("label_cellbarcode" = value) %>%
  dplyr::filter(label_cellbarcode %in% barcodes$label_cellbarcode) %>%
  tidyr::separate(label_cellbarcode, c("label", "cellbarcode"), sep = "_", remove = FALSE) %>%
  dplyr::left_join(tibble(
    "label_cellbarcode" = colnames(gex_scaled_all),
    "B_like_annotation" = gex_scaled_all@meta.data$My_annotation_LC
  )) %>%
  dplyr::filter(!str_starts(label_cellbarcode, pattern = "MBL7dup")) %>%
  dplyr::mutate(
    B_like_annotation = case_when(B_like_annotation == "B" ~ "Physiologic B Cells",
      B_like_annotation == "LC-MBL B Cell" ~ "LC-MBL Cells",
      B_like_annotation == "HC-MBL B Cell" ~ "HC-MBL Cells",
      B_like_annotation == "CLL B Cell" ~ "CLL Cells",
      .default = B_like_annotation
    ),
    B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))
  ) %>%
  arrange(B_like_annotation, label) %>%
  left_join(tibble(
    "B_like_annotation" = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"),
    "Colour" = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5")
  ))

barcodes2annotation <- barcodes_scheme %>%
  dplyr::select(label_cellbarcode, B_like_annotation) %>%
  dplyr::mutate(B_like_annotation = factor(B_like_annotation, levels = c("Physiologic B Cells", "LC-MBL Cells", "HC-MBL Cells", "CLL Cells"))) %>%
  deframe()

# plots schematic now for all CLL related DEGs including IGs
gex_data_scvi_DEGs <- t(gex_data_scvi[barcodes_scheme$label_cellbarcode, DEGs_disgenet_reactome_marker_IG])
saveRDS(gex_data_scvi_DEGs, file.path(RDS_dir, "gex_data_scvi_DEGs.RDS"))

deviation_summary_tmp <- scale(gex_data_scvi_DEGs) %>%
  t() %>%
  as_tibble(rownames = "label_cellbarcode") %>%
  tidyr::separate(label_cellbarcode, into = c("label", NA), sep = "_", remove = FALSE) %>%
  dplyr::left_join(metadata_all %>%
    dplyr::filter(data_type == "GEX") %>%
    dplyr::select(label, patient_id, MBL_subtype)) %>%
  dplyr::mutate(
    celltype = barcodes2annotation[label_cellbarcode],
    MBL_CLL_subtype = if_else(is.na(MBL_subtype), "CLL", MBL_subtype)
  ) %>%
  pivot_longer(-c(label, label_cellbarcode, patient_id, MBL_subtype, celltype, MBL_CLL_subtype), names_to = "DEG", values_to = "deviation") %>%
  dplyr::group_by(DEG, celltype) %>%
  dplyr::summarise(
    mean_deviation = mean(deviation),
    median_deviation = median(deviation)
  ) %>%
  dplyr::mutate(direction = case_when(mean_deviation & median_deviation >= 0 ~ "positive",
    mean_deviation & median_deviation < 0 ~ "negative",
    .default = "ambiguous"
  )) %>%
  ungroup() %>%
  group_by(celltype, direction) %>%
  dplyr::summarise(
    celltype_mean_deviation = mean(mean_deviation),
    celltype_median_deviation = median(median_deviation)
  )

saveRDS(deviation_summary_tmp, file = file.path(RDS_dir, "deviation_summary_tmp.RDS"))

deviation_summary <- deviation_summary_tmp %>%
  group_by(direction) %>%
  dplyr::filter(celltype != "Physiologic B Cells") %>%
  # Join with Physiologic B Cells on the same direction
  left_join(
    deviation_summary_tmp %>%
      filter(celltype == "Physiologic B Cells"),
    by = "direction",
    suffix = c("_current", "_reference")
  ) %>%
  # Calculate ratios
  dplyr::mutate( # mean_deviation_ratio = celltype_mean_deviation_current / celltype_mean_deviation_reference,
    mean_deviation_difference = celltype_mean_deviation_reference - celltype_mean_deviation_current,
    # median_deviation_ratio = celltype_median_deviation_current / celltype_median_deviation_reference,
    median_deviation_difference = celltype_median_deviation_reference - celltype_median_deviation_current
  ) %>%
  # Select relevant columns for output
  dplyr::select(
    celltype_current, direction, # mean_deviation_ratio, median_deviation_ratio,
    mean_deviation_difference, median_deviation_difference
  )

scheme_input <- deviation_summary %>%
  dplyr::mutate(MBL_CLL_subtype = case_when(
    celltype_current == "LC-MBL Cells" ~ "LC",
    celltype_current == "HC-MBL Cells" ~ "HC",
    celltype_current == "CLL Cells" ~ "CLL"
  )) %>%
  left_join(WBC_summary) %>%
  bind_rows(tibble(
    celltype_current = c("Physiologic B Cells", "Physiologic B Cells", "Physiologic B Cells", "Physiologic B Cells"),
    MBL_CLL_subtype = c(NA, NA, NA, NA),
    mean_WBC = c(0, 0, 5, 5),
    median_WBC = c(0, 0, 5, 5),
    mean_deviation_difference = c(0, 0, 0, 0),
    median_deviation_difference = c(0, 0, 0, 0),
    direction = c("positive", "negative", "positive", "negative")
  ))

rect_data <- tibble(
  ymin = rep(-0.4, 4),
  ymax = rep(0.4, 4),
  xmin = c(0, 5.48, 9.96, 18.35),
  xmax = c(5.48, 9.96, 18.35, 25)
)

write_tsv(scheme_input, file = file.path(tsv_dir, "GEX_motif_scheme_input.tsv"))
```

```{r, fig.width=9, fig.height=4}
scheme <- scheme_input %>%
  ggplot(aes(x = median_WBC, y = median_deviation_difference, group = direction)) +
  geom_rect(
    data = rect_data,
    mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    alpha = 0.2,
    fill = c("darkred", "#fcbda4", "#FC8D62", "#66C2A5"),
    inherit.aes = FALSE
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "lightgrey") +
  geom_line(colour = "olivedrab", linewidth = 1.5) +
  geom_point(colour = "olivedrab", size = 3) +
  scale_y_continuous(limits = c(-0.4, 0.4), expand = expansion(mult = c(0, 0))) +
  scale_x_continuous(limits = c(0, 25), expand = expansion(mult = c(0, 0))) +
  labs(title = "Gene expression", x = "Median WBC/nl", y = "Median difference") +
  theme_pub() +
  theme(plot.title = element_text(size = 10))
scheme

saveRDS(object = scheme, file = file.path(RDS_dir, "GEX_scheme.RDS"))
## values which increase in plot depict genes which were blue in heatmap at B cell stage and then turned red in disease progresion
## values which decrease vice versa
```
